### 2015-11-19 main.tex
initial commit

% sage_latex_guidelines.tex V1.01, 11 June 2015

\documentclass[shortAfour,sageh]{sagej}

\setcounter{secnumdepth}{3}
\usepackage{moreverb,url}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2015}

\begin{document}

\runninghead{Tonneau et al.}

\title{A demonstration of the \LaTeXe\ class file for
\itshape{SAGE Publications}}

\author{Steve Tonneau\affilnum{1}, Andrea Del Prete\affilnum{1}, Julien Pettr\'e\affilnum{2}, Franck Multon\affilnum{3}, Nicolas Mansard\affilnum{1}}

\affiliation{\affilnum{1}LAAS-CNRS, Toulouse, France\\
\affilnum{2}Inria, Rennes, France \\
\affilnum{3}Universit\'e Rennes 2, France}

\corrauth{Steve Tonneau, LAAS-CNRS
7 av. Colonel Roche, \\
BP 54200,
31031 Toulouse cedex 4,
France.}

\email{pro@stevetonneau.fr}

\begin{abstract}
	\input{body/abstract}
\end{abstract}

\keywords{Class file, \LaTeXe, \textit{SAGE Publications}}

\maketitle

\section{Introduction}
\input{body/intro}

\end{document}

### 2015-11-24 body/abstract.tex
abstract v1

We address the issue of automatically planning discrete sequences of contacts
that characterize a multiped motion in an arbitrary cluttered environment.
To overcome the inherent combinatorial difficulty of this problem, we separate it
in two sub-problems: first, planning a guide path for the root of the robot and
then, generating relevant contacts along this path.
We propose theoritical contributions for each sub-problem, and implement them
as a complete contact planner. 
First, we propose a geometrical characterization of a guide trajectory, the "reachability condition", that guarantees that a
guide can be mapped into the contact manifold of the robot. With this condition, asserting the "true feasibility" 
of a guide path boils down to performing collision tests, thus avoiding the costful of explicit computation of balanced contact configurations, as opposed to previous approaches.
Then, we propose a sequential algorithm to extend a truly feasible path into a sequence of contacts. For each limb, an offline sampling approach,
independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
The best candidate is chosen according to user created heuristics. We propose two such heuristics, used in all our experiments, and 
designed for generic robotic applications: a manipulatibily based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
These contributions are implemented within the open source software Hpp. Our framework allows us to generate complex multi contact sequences for a large
variety of robots, including a HRP-2, Hyq, and a dextrous hand, with interactive performances.

### 2015-11-24 body/abstract.tex
revised beg intro

@@
 as a complete contact planner. 
-First, we propose a geometrical characterization of a guide trajectory, the "reachability condition", that guarantees that a
+First, we propose a geometrical characterization of a guide trajectory, the ``reachability condition", that guarantees that a
 guide can be mapped into the contact manifold of the robot. With this condition, asserting the "true feasibility" 

### 2015-11-24 body/intro.tex
revised beg intro

We consider the problem of planning the acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment. Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to support the balance of the system.

Most multipedal locomotion focus on cyclic walking gaits \citep{Kajita03a}. But executing
this behaviour on cluttered environments is dangerous, if not impossible.
in an analysis of their participation to the darpa challenge, \citeauthor{atkensondarpa}
noted: ``Except for egress, no robots in the DRC Finals used
the  stair  railings  or  any  form  of  bracing.   Even  drunk  people  are  smart  enough  to  use  nearby  supports.
Full body locomotion (handholds,  bracing,  leaning against a wall or obstacles) should be easier than our
current high performance minimum contact locomotion approaches."

Indeed, the current approach to locomotion planning is to avoid obstacles as much as possible, instead of using them
to faciliate locomotion. The reason for this, as the authors state, is that ``More contacts make tasks
mechanically easier, but algorithmically more complicated for planning, and the transitions are difficult to
both plan and control[...].  We have seen very few robot planners that are  capable of  generating this  behavior."

The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of valid configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.

\subsection{State of the art}

\newcommand{\Pa}{$\mathcal{P}_1$ }
\newcommand{\Pb}{$\mathcal{P}_2$ }

Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences and biomechanics, and virtual character animation.
Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}. These approaches can intrinsically not adapt to new situations or discover complex behaviors in unforeseen contexts.

The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04}, where  it is proven to require the handling of two simultaneous problems, $\mathcal{P}_1$: a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic, balanced contact configurations along the trajectory\endnote{A third non trivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence. Noticebly, \citeauthor{Carpentier2016} combine
our planner with such a method to generate the complete motion.}.  A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories. This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not scalable to arbitrary environments. Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

Most of the papers that followed the work of \citeauthor{conf/iser/BretlRLKA04} have explored alternative formulations to handle the combinatorial issue. Two main directions have been explored. \textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence. A complete example of the potential offered by such approaches is given by \cite{Mordatch:2012:DCB:2185520.2185539}, followed by a successful adaptation to a real robot \citep{mordatch2015}. To keep reasonable computation times, the method uses a simplified dynamic model for the avatar. Still, the computation time is far from interactive  (about 1 minute of computation for a sequence of 20 contacts). \cite{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed integer problem, but only cyclic, bipedal locomotion is considered. Aside from the computation cost, a major drawback of these optimization based approaches is thus that they only offer local convergence when applied to acyclic contact planning.

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough guide trajectory (i.e. an ad-hoc solution to $\mathcal{P}_1$). \Pb  is then addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution can then be found, at the cost of prohibitive computation times (several hours). Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the relevance of the guide trajectory,  which is not evaluated \textit{a priori}. \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body, balanced contact configurations. Randomly sampled configurations are projected into the contact submanifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide trajectory in the examples presented). Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the trajectory.

\subsection{Paper contribution and organization}
Although the optimization approach is promising, we choose to focus on the sample-based methodology, more able to find complex trajectories in cluttered environments. While the theoretical structure of the problem is well understood, there is currently no scalable method to solve it. The combinatorial of the original problem (as described by \citeauthor{conf/iser/BretlRLKA04}) is too high to have any hope of tackling it directly. Alternative formulations are necessary to obtain practical solutions. We believe that the separation between the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}. However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
\begin{itemize}
\item The guide trajectory must satisfy a property guaranteeing the existence of a contact sequence to actuate it\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties.}. This property has not been studied yet: the only way to validate a trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}.
\item There is an infinite combination of possible contact sequences for a given root trajectory. The selection of one particular contact sequence with interesting properties (minimum number of contact change, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly pick one contact sequence, leading to possibly very tedious contact sequences).

\end{itemize}
\begin{figure*}[t]
  \centering
  \begin{overpic}[width=1\linewidth]{figures/workflow_new}
    \put (6.4,1.8) {\normalsize{Request}} 
    \put (22,11.3) {\large{A} }
    \put (31.6,2) {\scriptsize{\color{white}RB-PRM}} 
    \put (46.4,11.3) {\large{B} }
    \put (72.5,2) {\tiny{\color{white}EFORT}} 
  \end{overpic}
  \vspace{-1em}
  \caption{
    Overview of our 2-stage framework. (A) Given a path request between the yellow and blue positions, a guide trajectory is computed in $C_{reach}$ using RB-PRM. (B) The trajectory is extended into a discrete sequence of contact configurations using EFORT.}
  \label{fig:framework}
\end{figure*}

We claim that the desirable contact properties of a guide trajectory, proposed by \citeauthor{Bouyarmane2009}, can be formulated in a space of lower dimension, which we call $C_{reach}$. This formulation can make the planning of a guide trajectory more efficient computationally, while providing equivalent guarantees to planning directly in the configuration space. Among the particular properties obtained when planning in $C_{reach}$, we would like to guarantee that any reduced trajectory can actually lead to a feasible sequence of contacts, in which case we say that the reduced trajectory is truly feasible. It is possible in theory to guarantee that any reduced trajectory is truly feasible, even if it is more efficient in practice to approximate this property. The true-feasibility of the guide trajectory then allows us to focus on the selection of one particular sequence of contacts, for example one that minimizes the number of contacts in the sequence or maximizes the robot efficiency or style.

Based on these fundamental observations, we implement a very efficient acyclic contact planner. Our method is based on a probabilistic roadmap (PRM), that computes offline guide trajectories that are approximately truly feasible. The planner then resolves online the contact sequence by refining a guide trajectory computed from the PRM. Our planner is able to compute physically-consistent contact sequences for very complex systems (a humanoid, 28 joints; and an insectoid, 48 joints) in a few seconds for classical scenarios like climbing, and less than a minute for very complex problems like egress from a damaged truck. The planner also generalizes to planning dexterous manipulation movements, as demonstrated by preliminary results.

The contributions of the paper are twofold. We propose the first theoretical characterization of today's most efficient practical approach to sampled-based planning of acyclic contacts. And based on this characterization, we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with interactive applications.

We propose a framework to address the motion planning problem for multiped robots in cluttered environments: given a start and a goal configuration, the objective is to compute a sequence of contact configurations allowing to achieve the motion. For instance, we can consider the task of standing up, illustrated in Figure~\ref{fig:framework}--right. The problem is decoupled into two sequential phases: 1) the computation of a guide trajectory for the root of the robot; 2) the computation of a discrete sequence of contact configurations allowing to achieve the motion along the trajectory.
%
The remainder of this section presents the general organization of our method in Section~2. The two following sections 2 and 3 present respectively our answer to problems \Pa and \Pb. Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (humanoid, insectoid and three-finger manipulator) in various scenarios.

### 2015-11-24 body/intro.tex
fixed template, introdone up 1.2 not included

@@
 
-Most multipedal locomotion focus on cyclic walking gaits \citep{Kajita03a}. But executing
+Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. But executing
 this behaviour on cluttered environments is dangerous, if not impossible.
-in an analysis of their participation to the darpa challenge, \citeauthor{atkensondarpa}
+In an analysis of their participation to the darpa challenge, \citeauthor{atkensondarpa}
 noted: ``Except for egress, no robots in the DRC Finals used
@@
 both plan and control[...].  We have seen very few robot planners that are  capable of  generating this  behavior."
+The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of valid configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
 
-The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of valid configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
+Previous contributions that embrace the combinatorial provide complete approaches, not applicable in practice because they require hours of computations \citep{conf/iser/BretlRLKA04}.
+More recent successes use a local solution, resulting in more reasonnable performances (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
+Neither global or local methods presents satifying performances, because planning simultaneously the robot trajectory and the contacts that allow
+its execution is too costly.
+
+As suggested by \citeauthor{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner, while preserving the completeness of the search.
+We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot trajectory.
+This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
+This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely rapidly,
+while preserving a success rate superior to TODO \% in the demonstrated scenarios.
+
+In the remainder of this introduction, we discuss further the current state of the art. This allows us to then situate more precisely our contributions.
 
@@
 \textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence.
-A complete example of the potential offered by such approaches is given by \cite{Mordatch:2012:DCB:2185520.2185539}, followed by a successful adaptation to a real robot \citep{mordatch2015}.
+A complete example of the potential offered by such approaches was propose \citep{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
 To keep reasonable computation times, the method uses a simplified dynamic model for the avatar.
 Still, the computation time is far from interactive  (about 1 minute of computation for a sequence of 20 contacts).
-\cite{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed integer problem, but only cyclic, bipedal locomotion is considered.
+\citeauthor{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed integer problem, but only cyclic, bipedal locomotion is considered.
 Aside from the computation cost, a major drawback of these optimization based approaches is thus that they only offer local convergence when applied to acyclic contact planning.

### 2015-11-30 body/intro.tex
intro + overview no fig

@@
 
-The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04}, where  it is proven to require the handling of two simultaneous problems, $\mathcal{P}_1$: a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic, balanced contact configurations along the trajectory\endnote{A third non trivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence.
-Noticebly, \citeauthor{Carpentier2016} combine
-our planner with such a method to generate the complete motion.}.
+The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper.
+The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic, balanced contact configurations along the trajectory (A third non trivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories.
@@
 \textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence.
-A complete example of the potential offered by such approaches was propose \citep{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
+A complete example of the potential offered by such approaches was proposed \citep{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
 To keep reasonable computation times, the method uses a simplified dynamic model for the avatar.
@@
 
+It thus appears that an integrated optimization based approach \citep{Mordatch:2012:DCB:2185520.2185539}, able to generate a locally optimal, complete trajectory within minutes, currently outperforms
+existing planners, decoupled or not. However, recent contributions are able to interpolate dynamically executable 
+trajectories between contact configurations \citep{herzog2015trajectory, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real time performances.
+To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly. 
+Such a planner holds the promise of a complete real time multi contact locomotion system.
+
 \subsection{Paper contribution and organization}
-Although the optimization approach is promising, we choose to focus on the sample-based methodology, more able to find complex trajectories in cluttered environments.
-While the theoretical structure of the problem is well understood, there is currently no scalable method to solve it.
-The combinatorial of the original problem (as described by \citeauthor{conf/iser/BretlRLKA04}) is too high to have any hope of tackling it directly.
-Alternative formulations are necessary to obtain practical solutions.
-We believe that the separation between the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
+This paper presents a pragmatic approach to break the complexity of multi contact planning. With that objective in mind,
+we believe that the separation between the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
 However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
@@
 \item The guide trajectory must satisfy a property guaranteeing the existence of a contact sequence to actuate it\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties.}.
-This property has not been studied yet: the only way to validate a trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}.
+We call this property \textit{true feasibility}.
+This property has not been studied yet: the only way to validate a waypoint in the trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}.
 \item There is an infinite combination of possible contact sequences for a given root trajectory.
 The selection of one particular contact sequence with interesting properties (minimum number of contact change, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly pick one contact sequence, leading to possibly very tedious contact sequences).
+\end{itemize}
 
-\end{itemize}
-\begin{figure*}[t]
-  \centering
-  \begin{overpic}[width=1\linewidth]{figures/workflow_new}
-    \put (6.4,1.8) {\normalsize{Request}} 
-    \put (22,11.3) {\large{A} }
-    \put (31.6,2) {\scriptsize{\color{white}RB-PRM}} 
-    \put (46.4,11.3) {\large{B} }
-    \put (72.5,2) {\tiny{\color{white}EFORT}} 
-  \end{overpic}
-  \vspace{-1em}
-  \caption{
-    Overview of our 2-stage framework.
-(A) Given a path request between the yellow and blue positions, a guide trajectory is computed in $C_{reach}$ using RB-PRM.
-(B) The trajectory is extended into a discrete sequence of contact configurations using EFORT.}
-  \label{fig:framework}
-\end{figure*}
+By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
+For each problem, we introduce an original formulation of the related theoritical issues.
+Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
+The relaxation is justified empirically, with statistics presented for each robot and scenario presented. The complete framework is implemented within
+the open source platform Hpp.
 
-We claim that the desirable contact properties of a guide trajectory, proposed by \citeauthor{Bouyarmane2009}, can be formulated in a space of lower dimension, which we call $C_{reach}$.
+To address \Pa we rigorously formulate a geometric condition for the true feasibility of a path, the reachability condition. We claim that this 
+condition can be verified in a low dimensional space, that we call $C_{reach}$.
 This formulation can make the planning of a guide trajectory more efficient computationally, while providing equivalent guarantees to planning directly in the configuration space.
-Among the particular properties obtained when planning in $C_{reach}$, we would like to guarantee that any reduced trajectory can actually lead to a feasible sequence of contacts, in which case we say that the reduced trajectory is truly feasible.
-It is possible in theory to guarantee that any reduced trajectory is truly feasible, even if it is more efficient in practice to approximate this property.
-The true-feasibility of the guide trajectory then allows us to focus on the selection of one particular sequence of contacts, for example one that minimizes the number of contacts in the sequence or maximizes the robot efficiency or style.
+In practice, for each considered robot, we compute an approximation of $C_{reach}$. From this approximation,
+an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained.
+This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly feasible paths for the root of the robot within a few milliseconds to a few seconds depending on the scenario.
 
-Based on these fundamental observations, we implement a very efficient acyclic contact planner.
-Our method is based on a probabilistic roadmap (PRM), that computes offline guide trajectories that are approximately truly feasible.
-The planner then resolves online the contact sequence by refining a guide trajectory computed from the PRM.
-Our planner is able to compute physically-consistent contact sequences for very complex systems (a humanoid, 28 joints; and an insectoid, 48 joints) in a few seconds for classical scenarios like climbing, and less than a minute for very complex problems like egress from a damaged truck.
-The planner also generalizes to planning dexterous manipulation movements, as demonstrated by preliminary results.
+To address \Pb, we consider a truly feasible path as an input, and generate new contact configurations along it using a sampling approach.
+The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed).
+The theorical completeness of the approach is partially traded for efficiency: instead of sampling online
+limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
+for proximity queries. At runtime, this framework allows to obtain extremly rapidly a set of candidate configurations for contact.
+These candidates are sorted using a criterion to evaluate the quasi static balance of the configuration, and ranked according to user defined heuristics.
+In our implementation we use two such heuristics, used in all our experiments: a robustness criterion regarding the balance of a configuration, and a manipulability based heuristic,
+the Extended FORce Transmission ratio (EFORT).
 
-The contributions of the paper are twofold.
-We propose the first theoretical characterization of today's most efficient practical approach to sampled-based planning of acyclic contacts.
-And based on this characterization, we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with interactive applications.
+The contributions of the paper are thus twofold.
+We propose a theoretical characterization of today's most efficient practical approach to sampled-based planning of acyclic contacts.
+Based on this characterization, we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with interactive applications.
 
-We propose a framework to address the motion planning problem for multiped robots in cluttered environments: given a start and a goal configuration, the objective is to compute a sequence of contact configurations allowing to achieve the motion.
-For instance, we can consider the task of standing up, illustrated in Figure~\ref{fig:framework}--right.
-The problem is decoupled into two sequential phases: 1) the computation of a guide trajectory for the root of the robot; 2) the computation of a discrete sequence of contact configurations allowing to achieve the motion along the trajectory.
-%
-The remainder of this section presents the general organization of our method in Section~2.
-The two following sections 2 and 3 present respectively our answer to problems \Pa and \Pb.
-Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (humanoid, insectoid and three-finger manipulator) in various scenarios.
+In comparison with previous work, our planner does not produce a complete motion, but a discrete sequence of contacts.
+Because we showed that the interpolation of a complete trajectory between such configurations can be solved within real time constraints \citep{Carpentier2016}, we claim that our planner
+is the fastest multi-contact planning solution to our knowlege.
+
+The present paper is an extension of a conference paper to appear in the proceedings of the ISRR '15 conference TODO CITE.
+The conference paper focuses on the theoritical formulation of the problem, and presents results obtained with virtual avatars.
+This extension completes this work with an argumented discussion on the implementation of this theory to real world robots and problems.
+In particular, we detail the robust balance criteria designed to ensure the equilibrium of the robot. Using HRP-2 as a running example, we describe how $C_{reach}$ is approximated to address
+\Pa, and provide the pseudo code of the contact generation algorithm implemented to find balanced contact sequences.
+
+In Section~\ref{overview}, we present the general organization of our method.
+Section~\ref{rbprm} and Section~\ref{sec:efort} present respectively our answer to problems \Pa and \Pb.
+In Section~\ref{heuristics}, we present two heuristics used for the selection of the best contact configuration.
+Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and Hyq robots, and a three-finger manipulator) in various scenarios,
+that complete those presented in the conference paper.

### 2015-11-30 body/overview.tex
intro + overview no fig

\section{Overview}
\label{overview}

%
\begin{figure}
  \begin{overpic}[width=0.9\linewidth]{figures/contact_gen}
		\put (3,3) {1)} 
		\put (25,3) {2)} 
		\put (44,3) {3)} 
		\put (63,3) {4)} 
		\put (82,3) {5)} 
	\end{overpic}
  \caption{Generation of a contact configuration for the right arm of a humanoid robot. 1) Selection of reachable obstacles. 2) A request is performed on a database of configurations. 3) Configurations too far from contact are eliminated. 4) The best candidate according to EFORT is chosen. 5) The final contact is achieved using inverse kinematics.}
  \label{fig:contact_gen}
\end{figure}
\subsection{Computation of a guide path}
We first consider the problem of planning a relevant guide path. The objective is to compute a path of root placements which will allow contact creation. A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide trajectory, but at the same time, any computed guide trajectory must be truly feasible, i.e. must lead to a valid sequence of contacts.

 An intuitive description of such placements is "close, but not too close": close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Fig.~\ref{fig:contact_gen}--1); not too close, because the robot must avoid collision (which is represented by the hull including the torso in Fig.~\ref{fig:contact_gen}--1). We define formally $C_{reach}$, the set of interesting root placements, in which we compute a guide trajectory with a sampling based planner,  the Reachability-Based RRT --RB-RRT-- (Figure~\ref{fig:framework}--A). Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
%
Details are presented in Section~\ref{rbprm}.

\subsection{Generating a discrete sequence of contact configurations}

The second stage extends the guide path into a sequence of contact configurations (Fig.~\ref{fig:framework}--B). The true feasibility of the input path is guaranteed by the first step.
This provides in turn a reasonnable guarantee that a contact configuration can be obtained for any root placement along the path. With a fixed root position, the dimensionality of the 
contact generation problem is reduced to the number of degrees of freedom of the considered limb. In other words, we consider each limb as a manipulator attached to the root, and select the most relevant contact from a database of precomputed configurations, independent from the environment (Figure~\ref{fig:contact_gen}).

From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root placement, an inverse kinematics solver 
is used to maintain the contacts that were previously existing (TODO Figure). Often, these contacts cannot be maintained, because of joint limits or collisions (TODO Figure).
In such a case the contacts are broken. Conversely, new contacts are created to ensure the quasi static balance of the robot (TODO FIGURE).
The algorithm is designed so that only one contact can be created or broken between two successive configurations. This can be seen as a heuristic to ensure
that the interpolation between two configurations is achievable.
Details are presented in Section~\ref{sec:efort}.

To select the "most relevant" contact configurations, user defined heuristics can be chosen. In Section~\ref{heuristics} we present extensively two such heuristics, that have been
used in our experiments.

%~ To create a contact, we consider each limb of the robot as a manipulator arm attached to the root. We store a database of configurations for each manipulator. Figure~\ref{fig:contact_gen} -- 2 presents a few configurations for the right arm of the robot. The configurations in the database which are close to contact are considered (Figure~\ref{fig:contact_gen} -- 2 and 3). Among the candidates, 3 criteria are considered to choose the most relevant: the ideal candidate is free of collision, allows to maintain static balance if necessary, and  contributes to the motion in an optimal way. We call the measure of this contribution task efficiency: it represents the ability the limb has to exert a force compatible with the direction of motion. In Figure~\ref{fig:contact_gen} -- 4, the black arrow represents the direction of motion. Among the candidates, the most relevant to achieve a vertical motion is chosen, and projected on the contact surface using an inverse kinematics solver (Figure~\ref{fig:contact_gen} -- 5). The task efficiency is measured using the Extended FORce Transmission ratio (EFORT), proposed by ~\cite{Tonneau:2014:TEC:2619648.2619652}.

\subsection{Notation conventions and definitions} \label{notations}

A vector  $\mathbf{x}$ is denoted with a bold lower case letter.
A matrix $\mathbf{A}$ is denoted with a bold upper case letter.
A set $C$ is denoted with a upper case italic letter.
Scalar variables and functions are denoted with lower case italic letters, such as
$r$ or $f(\textbf{x})$.

%\subsubsection{Robot description}
%\label{def}

\begin{figure}[][t]
  \centering
  \begin{overpic}[width=0.46\linewidth]{figures/character}
    \put (7,29) {\tiny{$\in \mathbf{q}^1$}}
    \put (7,25) {\tiny{effector}}
  \end{overpic}
  \caption{
    Left: Robot in a rest configuration. The right arm is denoted as the limb $R^1$. Each colored dot represents a degree of freedom around an axis. Right: Volumes of the robot. The red geometry denotes $W^0$ and must remain collision-free. The green spheres are the  $W^k$.}
  \label{fig:character}
\end{figure}

\medskip
\textbf{A robot} is a kinematic chain $R$, comprising $n + 6$ degrees of freedom (DOFs).
$R$ is composed of $l$ limbs $R^k, 1 \leq k \leq l$, attached to a root.
It is described by a configuration $\mathbf{q} \in SE(3) \times \mathbb{R}^n$.
We define some relevant projections of $\mathbf{q}$:
\begin{itemize}
	\item $\mathbf{q}^k$ denotes the configuration (a vector of joint values) of the limb $R^k$ (Fig.~\ref{fig:character});
	\item $\mathbf{q}^{\overline{k}}$ denotes the vector of joint values of R \textbf{not} related to $R^k$. We define for convenience $\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
	\item $\mathbf{q}^{0}\in SE(3)$ denotes the position and orientation of the root of the robot $R$.
\end{itemize}
%Finally, $ \mathbf{q}_{start}$ and  $\mathbf{q}_{goal}$ are the start and goal configurations given as an input of our problem.

%\subsubsection{Environment and other volumes} 
%\label{sec:A}
\medskip
\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ it contains.
The volume encompassing the trunk of the robot is denoted $W^0$ (Fig.~\ref{fig:character}-right: central cylinder). The reachable workspace of a limb $R^k$ is denoted $W^k$ (Fig.~\ref{fig:character}-right: the four ellipses). 
\begin{equation}
  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \mathbf{q}^k, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\}
\end{equation}
where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0$ the null displacement. We also define $W = \bigcup_{k=1}^{l}W^k$.
%
Finally, we define $W^k(\mathbf{q}^{0}), 0 < k \leq l$ as the volume $W^k$ translated and rotated by the rigid displacement~$\mathbf{q}^{0}$.

### 2015-12-01 body/contact.tex
debut planner

\section{From a guide path to a discrete sequence of contact configurations (Stage 2)}
\label{sec:contact}
As an input of this stage, we consider a truly feasible guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$. We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$. From true feasibility we know that such a trajectory exists.

In this Section we first describe a single contact generation process, that is how to generate a contact configuration for a given limb 
and root location.
Then, on this method we build an iterative algorithm to generate a discrete sequence of balanced contact configurations along a truly feasible path.

\subsection{Single contact generation}
We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$ truly feasible regarding $R^k$. FIGURE
From true feasibility we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify Eq.~\eqref{eq:pi} for $\mathbf{q}^0$.
The issue is then to evaluate locally any element of $P$ at $\mathbf{q}^0$.
This can be achieved as follows:
\begin{enumerate}
\item Generate randomly a valid limb configuration;
\item Project the end-effector onto the closest surface with inverse kinematics;
\item If a feasible solution is found, stop. Otherwise repeat from step 1.
\end{enumerate}

We partially trade the completeness property of this approach for a more efficient solution \citep{Tonneau2014}:
\begin{enumerate}
\item Generate offline $N$ valid sample limb configurations;
\item using the end-effector position expressed in $R$ coordinates as an index, store each sample in an octree stata structure;
\item Figure TODO, 2 and 3: At runtime, when contact creation is required, retrieve from the octree the list of samples $S = [\mathbf{q}^k_0, ..., \mathbf{q}^k_i, ...]$ such that $dist(\mathbf{p}^k(\mathbf{q_k}^k),O) < \epsilon$, with $\epsilon $ small ;
\item According to a user-defined heuristic $h$, sort the elements of $S$;
\item FIGURE TODO, 4 and 5: Select the first element of $S$, and project the configuration to contact with inverse kinematics;
\item If a feasible solution is found, stop. Otherwise remove the element from $S$ and go from step 5, or fail if $S$ is empty.
\end{enumerate}

This approach has two main advantages:
First, this allows us to select a number of candidates at a small cost, (a single proximity request), and to sort them according to a heuristic;
Secondly, thanks to the use of proximity requests, the number of iterations necessary for the inverse kinematics to converge is reduced.
Finally, it is easy to see that as $N$ grows to infinity, the completeness of the method is restored.
$N$ is a parameter that allows us to specify the trade-off between completeness and efficiency.
The reader is invited to refer to our previous work for an extensive discussion on the optimal value of $N$ \citep{Tonneau2014}.

\subsubsection{Efficient contact generation }

Concretely, any, we can use a random sampling approach combined with an inverse kinematics projector.

given a root configuration $\mathbf{q}^0$ and a limb $R^k$, we locally identify a countable set of mappings $P = [\pi_0,..,\pi_i] $ that verify Eq.~\eqref{eq:pi} for $\mathbf{q}^0$. and select the most relevant according to a heuristic $h$.

%We compute the image of a mapping $f$ along $\mathbf{q}^0(t)$ into a discrete sequence of contacts $ \in C_{contact}$. We first describe the discretization of $\mathbf{q}^0(t)$ and the algorithm used to create or break contacts along it. Then we describe how contacts are explicitely generated along the discretized trajectory.

\subsection{Extension of the guide trajectory} 
The guide trajectory $\mathbf{q}^0(t)$ is first discretized into a sequence of $j$ key placements:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations.  To ensure continuity in the contact transition phases, we rewrite $\pi$ under the following recursive form for any $0<i<j$:
\begin{equation*}
    \pi\colon\left\{
    \begin{aligned}		
        \mathbf{Q}^0 \in C_{reach} & \longrightarrow C_{contact} \\
        %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
        \mathbf{q}^{0}_i &  \longrightarrow  \mathbf{q}^{0}_i \oplus g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
    \end{aligned}
    \right.
\end{equation*} 
We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
The function $g$ is defined independently by $g^k$ for each limb $R^k$. In defining $g^k$, two aspects must be considered. Is the limb $R^k$ in contact? And which criteria is it optimizing?

\medskip%\subsection{Maintaining and breaking contact in the sequence}
\noindent\textbf{Maintaining a contact:} If possible, a limb in contact at time $i-1$ remains in contact at $i$. The contact is broken if an inverse kinematics solver fails to find a collision free limb configuration which satisfies joint limits.
The solver is a direct implementation of the inverse kinematics solver with task priorities framework proposed in \cite{Baerlocher2004}, which
handles explicitely joint limits, and allows to integrate collision avoidance as a secondary criterion.

If the solver fails, the contact is broken and a collision free configuration is assigned to the limb.
%~ The configuration is selected to be the closest to the previous configuration (by minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$).

Once a first candidate configuration is taken for all limbs, the quasi-static balance is tested by whether the weight wrench is in the gravito-inertial cone (i.e. there exists valid contact forces that compensate for the weight of the robot), using the geometric approach described in \cite{qiu:dhm:2011}. If the balance is not obtained, new contacts are randomly generated using the following procedure.

%\begin{itemize} 
%\item If the robot is not statically balanced and $\mathbf{q}_{i-1}^k$ is not in contact, $g$ computes a configuration $\mathbf{q}_{i}^k$ so that a contact is taken with the environment (Fig.~\ref{fig:generate_contact}).
%\item Else, $g$ computes a configuration $\mathbf{q}_{i}^k$ minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$, subject to $R^k(\mathbf{q_i}) \cap O = \emptyset$ and $\mathbf{q_i}^k$ respects the  joint limits of $R^k$.
%
%If $\mathbf{q}_{i - 1}^k$ is in contact, this results in trying to maintain the contact for $\mathbf{q}_{i}^k$ (Figure~\ref{fig:break_contact}--1 and 2). The minimum is different than zero if $W^k(\mathbf{q}_i^0) \cap  \mathbf{p}^k(\mathbf{q_{i-1}}) = \emptyset$ (Figure~\ref{fig:break_contact}--3), or collision can not be  prevented (Figure~\ref{fig:break_contact}--4).
%\end{itemize}

\begin{figure}[t]
\centering
  \begin{overpic}[width=0.6\linewidth]{figures/break_contact}
		\put (3,15) {1)} 
		\put (27,15) {2)} 
		\put (50,15) {3)} 
		\put (75,15) {4)} 
		%~ \put (68,58) {3.a)} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	\end{overpic}
\caption{Contacts are maintained unless their position is too far, or the environment prevents it.}
		   \label{fig:break_contact}
\end{figure}

\begin{figure}[t]
\centering
  \begin{overpic}[width=0.6\linewidth]{figures/generate_contact}
		%~ \put (5,58) {1)} 
		%~ \put (37,58) {2)} 
		%~ \put (68,58) {3.a)} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	\end{overpic}
\caption{Contacts are generated when the configuration is not balanced.}
		   \label{fig:generate_contact}
\end{figure}

%\subsection{Generation of a contact configuration}  
\medskip\noindent\textbf{Creating a contact:} We consider a configuration where some limbs are in contact, some are free and quasi-static balance is not enforced. To enforce balance, we proceed in the following 
manner:
we randomly select a contact free limb; if there is no contact free limb, we select
the limb that made contact first.
Using the contact generator introduced in~\cite{Tonneau2014}, we project the configuration of this limb into a contact that enhances balance, if it exists (Figure~\ref{fig:generate_contact}); If balance is not achievable and a contact is possible, it is generated anyway;
If balance is not achieved, the next limb is selected and projected into a contact configuration, and so on.
This approach can lead to the repositioning of existing contacts, in which case intermediate states are inserted to reposition the contacts.
This current implementation does not guarantee that the planner will succeed in generating a balanced configuration, because
true feasibility is not fully guaranteed.
However in practice the planner is successful in the large majority of cases, as discussed in section~\ref{sec:stat1}.
%~ A new contact configuration is obtained by projecting a random configuration to the nearest contacts. The sampling is biased by selecting configurations so that the limbs previously in contact are already close to contact and at least one additional limb is close to the environment. The random configuration is then projected onto the contact by iteratively minimizing the distance between the end effectors and the environment, under the constraint of respecting the joint limits and avoiding the collision.

\subsection{Contribution to the global movement: the EFORT criteria}
\noindent\textbf{EFORT criterion:} If only relying on the random sampling to select new contacts, the planner produces inefficient postures. The resulting contact sequence is then poorly efficient and unnatural. Moreover, the limbs are not well configured and are not able to efficiently follow the general movement: contacts break frequently.

When creating additional contacts, we therefore propose to select particular configurations that allow to exert a force compatible with the direction of motion. This task efficiency is measured  with the Extended FORce Transmission ratio (EFORT)~\cite{Tonneau2014}.
%
The measure of EFORT is given by
\begin{equation}
\alpha_{EFORT}(\mathbf{q}^k, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \nu_0 \mathbf{n}^T \mathbf{m})
\end{equation}
where $\mathbf{J}$ is the Jacobian matrix of the limb $R^k$ in configuration $\mathbf{q}^k$; $\nu_0$ is the friction coefficient of the contact surface; $\mathbf{n}$ is the normal of the contact surface; and $\mathbf{m}$ is the direction opposite to the motion,
given by the 3D vector connecting $\mathbf{q}_{i}^0$ and $\mathbf{q}_{i+1}^0$.
%
The first part of the equation measures the ratio between the joint  torques and the resulting force applied along $\mathbf{m}$. The second part quantifies the odds of slipping while applying a force along $\mathbf{m}$.

\noindent\textbf{Optimization at creation:} In practice, a database of configurations is stored for each limb, which can be considered as manipulator arms. The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations (and additionally storing $\mathbf{J}$). 
%Fig.~\ref{fig:contact_gen}-2 presents a few configurations for the right arm of the robot.  
Upon request, the octree returns a set of configurations close to contact (Fig.~\ref{fig:contact_gen}-3). These candidates are sorted based on their task efficiency, given by $\alpha_{EFORT}$. The first candidate in this list satisfying the balance criterion and is collision free is selected and projected on the contact surface using our inverse kinematics solver.

%Therefore to generate a contact, we want to find a configuration which is collision free, allows to maintain balance, and maximizes EFORT. We choose this configuration among a set of possible candidates, rapidly computed thanks to a contact generator, also proposed in~\cite{Tonneau2014}. We recall the overall idea here. We consider each limb of the robot as a manipulator arm attached to the root. We store a database of configurations for each manipulator. The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations. The jacobian matrix $\mathbf{J}$ of each configuration is also stored. Figure~\ref{fig:contact_gen} -- 2 presents a few configurations for the right arm of the robot.

%To summarize: given a feasible trajectory for the root of the robot $R$, we are able to generate a discrete sequence of contact configurations  describing a feasible motion for $R$, between a start an goal configurations. The contact configurations are chosen to be statically balanced and task efficient, which means that they can generate a force  contributing to the motion of the root along the trajectory.

### 2015-12-01 body/efort.tex
debut planner

--- 2015-12-01 body/contact.tex
@@
-\section{From a guide path to a discrete sequence of contact configurations (Stage 2)}
-\label{sec:contact}
-As an input of this stage, we consider a truly feasible guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$.
-We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.
-From true feasibility we know that such a trajectory exists.
+\section{From a guide trajectory to a discrete sequence of contact configurations (Stage 2)}
+\label{sec:efort}
+As an input of this stage, we consider a truly feasible guide trajectory $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$.
+We now consider the second problem of computing a trajectory $\mathbf{q}^{\overline{0}}(t)$ for the limbs of the robot.
+Since we assume true feasibility, we know that such a trajectory exists.
+Contrary to previous works \cite{DBLP:conf/iser/EscandeKMG08,Bouyarmane2009}, the goal here is not to find any such trajectory but rather to select one with particular properties.
+Specifically, we show here how to build a contact sequence with a small number of contact variations and  good-efficiency and naturalness of the postures.
 
-In this Section we first describe a single contact generation process, that is how to generate a contact configuration for a given limb 
-and root location.
-Then, on this method we build an iterative algorithm to generate a discrete sequence of balanced contact configurations along a truly feasible path.
-
-\subsection{Single contact generation}
-We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$ truly feasible regarding $R^k$. FIGURE
-From true feasibility we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify Eq.~\eqref{eq:pi} for $\mathbf{q}^0$.
-The issue is then to evaluate locally any element of $P$ at $\mathbf{q}^0$.
-This can be achieved as follows:
-\begin{enumerate}
-\item Generate randomly a valid limb configuration;
-\item Project the end-effector onto the closest surface with inverse kinematics;
-\item If a feasible solution is found, stop. Otherwise repeat from step 1.
-\end{enumerate}
-
-We partially trade the completeness property of this approach for a more efficient solution \citep{Tonneau2014}:
-\begin{enumerate}
-\item Generate offline $N$ valid sample limb configurations;
-\item using the end-effector position expressed in $R$ coordinates as an index, store each sample in an octree stata structure;
-\item Figure TODO, 2 and 3: At runtime, when contact creation is required, retrieve from the octree the list of samples $S = [\mathbf{q}^k_0, ..., \mathbf{q}^k_i, ...]$ such that $dist(\mathbf{p}^k(\mathbf{q_k}^k),O) < \epsilon$, with $\epsilon $ small ;
-\item According to a user-defined heuristic $h$, sort the elements of $S$;
-\item FIGURE TODO, 4 and 5: Select the first element of $S$, and project the configuration to contact with inverse kinematics;
-\item If a feasible solution is found, stop. Otherwise remove the element from $S$ and go from step 5, or fail if $S$ is empty.
-\end{enumerate}
-
-This approach has two main advantages:
-First, this allows us to select a number of candidates at a small cost, (a single proximity request), and to sort them according to a heuristic;
-Secondly, thanks to the use of proximity requests, the number of iterations necessary for the inverse kinematics to converge is reduced.
-Finally, it is easy to see that as $N$ grows to infinity, the completeness of the method is restored.
-$N$ is a parameter that allows us to specify the trade-off between completeness and efficiency.
-The reader is invited to refer to our previous work for an extensive discussion on the optimal value of $N$ \citep{Tonneau2014}.
-
-\subsubsection{Efficient contact generation }
-
-Concretely, any, we can use a random sampling approach combined with an inverse kinematics projector.
-
-given a root configuration $\mathbf{q}^0$ and a limb $R^k$, we locally identify a countable set of mappings $P = [\pi_0,..,\pi_i] $ that verify Eq.~\eqref{eq:pi} for $\mathbf{q}^0$.
-and select the most relevant according to a heuristic $h$.
+More precisely, any mapping $\pi$ introduced in \eqref{eq:pi} can be used to expand $\mathbf{q}^0(t)$ into a whole-body trajectory.
+We propose here a particular construction of $\pi$ leading to interesting contact sequences.
 

### 2015-12-01 body/intro.tex
debut planner

@@
 We call this property \textit{true feasibility}.
-This property has not been studied yet: the only way to validate a waypoint in the trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}.
+This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi flat cases \citep{zucker2010optimization}.
 \item There is an infinite combination of possible contact sequences for a given root trajectory.
@@
 This extension completes this work with an argumented discussion on the implementation of this theory to real world robots and problems.
-In particular, we detail the robust balance criteria designed to ensure the equilibrium of the robot. Using HRP-2 as a running example, we describe how $C_{reach}$ is approximated to address
-\Pa, and provide the pseudo code of the contact generation algorithm implemented to find balanced contact sequences.
+In particular, we detail the robust balance criteria designed to ensure the equilibrium of the robot and provide details on the hpp implementation
+of our contact generation algorithm. Furthermore, our experiments now concern real robots, namely HRP-2 and Hyq. TODO Appendix describes the generation process
+for the robot.
 

### 2015-12-01 body/overview.tex
debut planner

@@
 \medskip
-\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ it contains.
 The volume encompassing the trunk of the robot is denoted $W^0$ (Fig.~\ref{fig:character}-right: central cylinder).
@@
 %
-Finally, we define $W^k(\mathbf{q}^{0}), 0 < k \leq l$ as the volume $W^k$ translated and rotated by the rigid displacement~$\mathbf{q}^{0}$.
+$W^k(\mathbf{q}^{0}), 0 < k \leq l$ is defined as the volume $W^k$ translated and rotated by the rigid displacement~$\mathbf{q}^{0}$.
+
+\medskip
+\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ it contains.
+Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.

### 2015-12-01 body/rbprm.tex
debut planner

\section{Computation of a guide path in $C_{reach}$ (Stage 1)}
\label{rbprm}

We consider the problem of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the geometrical root of a multiped robot, connecting start and goal configurations. As said in the previous section, the goal is to enforce that any configuration $\mathbf{q}^0$ of the guide is truly feasible, i.e. can be mapped to a configuration in contact. We denote by $C_{contact} \subset SE(3) \times \mathbb{R}^n$ the contact submanifold of the robot.
%~ We say that a root placement $\mathbf{q}^{0}$ is \textit{truly feasible} if there exists $\mathbf{q}^{\overline{0}}$ such that 
%~ $$\mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact} $$
%~ The set of all truly-feasible root placements is denoted by $C_{reach}$. By extension, a trajectory $\mathbf{q}^0(t)$ is truly feasible if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{reach}$. 
%~ We define a bijective mapping $\pi$ between $C_{reach}$ and $C_{contact}$:
%~ \begin{equation}
    %~ \pi\colon \quad \mathbf{q}^{0} \in C_{reach} \longrightarrow  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact} 
    %~ \label{eq:pi}
%~ \end{equation}

We say that a root placement $\mathbf{q}^{0}$ is \textit{truly feasible} if there exists at least one bijective mapping $\pi$
such that
\begin{equation}
    \pi\colon \quad \mathbf{q}^{0} \in  SE(3) \longrightarrow  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact} 
    \label{eq:pi}
\end{equation}

The set of all truly feasible root placements is denoted by $C_{reach}$. By extension, a path $\mathbf{q}^0(t)$ is truly feasible if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{reach}$.

For a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide trajectory generator: all the generated trajectories must be \textit{truly feasible} (sufficient condition); the guide planner must be complete, i.e. it must be able to discover any truly feasible trajectory (necessary condition).

In this Section we characterize $C_{reach}$ theoritically with a necessary and a sufficient condition for belonging to the set.
We then describe our implementation of an efficient RRT planner in $C_{reach}$.

%~ We want all the configurations of $q^0(t)$ to belong to a set $C_{reach}$ defined as:
%~ 
%~ \begin{eqnarray}
%~ C_{reach}= \{ \mathbf{q}^{0} & : & \exists f, f(\mathbf{q}^{0}) \in C_{contact} \}
%~ \end{eqnarray}

\subsection{Conditions for true feasibility}
By default, the true feasibility implies a constructive demonstration by exhibiting a valid $\pi$. This is the approach chosen by \cite{Bouyarmane2009}. However, computing a valid  $\mathbf{q}^{\overline{0}}$ is costly in practice. In this section we rather define a necessary condition and a sufficient condition for true feasibility that do not require this explicit computation.

% and  present the parametrization of a compromise condition used to approximate $C_{reach} \subset SE(3)$, the set of truly feasible configurations. This compromise condition, which we call the reachability condition, is used to compute a relevant guide trajectory using a sampling based motion planner, the Reachability-Based PRM (RB-PRM).

\subsubsection*{True feasibility: necessary condition}
For a contact to be possible, a volume $O_i \in O$ necessarily intersects with the reachable workspace $W(\mathbf{q}^{0})$ (Fig.~\ref{fig:contact_gen}--1). Furthermore, if $\mathbf{q}^{0}$ is truly feasible, then the trunk of the robot $W^0(\mathbf{q}^{0})$ is necessarily not colliding  with the environment $O$.

Therefore we can approximate $C_{reach}$ with a set $C_{reach} \subset C_{reach}^1$ with the reachability condition defined as: 
\begin{equation}
C_{reach}^1 = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
 %~ & \wedge & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}
\end{equation}
It is straighforward to prove that  $C_{reach} \subset C_{reach}^1$ (by construction of the included set). 
This inclusion is very important: it directly implies that any motion-planning algorithm with a guaranty of completeness in $SE(3) \times \mathbb{R}^n$ is complete  in $C_{reach} \times \mathbb{R}^n$. This is a strong reduction of the search space, which can be directly applied to any existing method.

The condition $C_{reach}^1$ is only necessary which means that one such root placement might not be truly feasible: in practice it is not guaranteed to find a valid sequence of contacts for every guide trajectory in $C_{reach}^1$.

\subsubsection*{Sufficient condition for true feasibility}
%~ In building $C_{reach}^1$, we directly considered the including hull of the root body $W^{0}$, obtaining a necessary condition. 
A trivial sufficient condition for true feasibility can be constructed as a variation of $C_{reach}^1$, by replacing $W^0$ with a bounding volume $B^{max}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact (This condition is trivial in the sense that the resulting $W$ has a zero measure. For the need of the proof, the trivial sufficient condition is enough. In practice, the construction of a non-trivial including shape $W^0$ was possible for all the robot structures we considered). We denote by $C_{reach}^\infty \subset C_{reach}$ the set of root placements corresponding to the sufficient condition.

In general, the inclusion is strict, which means that we lose the completeness of the two-stage contact planner (i.e. the planner is not able to discover a trajectory inside $C_{reach} \setminus C_{reach}^\infty$). However, the sufficient condition guarantees that any such trajectories leads to a valid sequence of contacts (i.e. $\pi$ is defined).

\subsection{Reachability: a compromise condition}
The sufficient condition is not  interesting in practice since it leads the solver to lose too many interesting trajectories. The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not truly feasible in practice. It might be possible to find a shape $B$ that is necessary and sufficient; however, it seems intuitively very unlikely in general. The construction of a shape $W^0 \subset B \subset B^{max}$ leading to a necessary and sufficient condition (or the proof of its inexistence) is out of the scope of this work.

However between $W^0$ and $B^{max}$, a trade-off can be found between a necessary and a sufficient condition. We define $W^0_s$ as the volume $W^0$ subject to a scaling transformation by a factor $s \in \mathbb{R}^+$. 
%
We then consider the spaces $C_{reach}^s$
 \begin{equation}
C_{reach}^s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
 %~ & \wedge & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}
\end{equation}
%
The higher $s$ is, the closer this "reachability condition" is to being sufficient, and if $s=1$, the planner is complete. The parametrization of $s$ allows to find a trade-off between these two desirable properties.
Section~\ref{sec:results} shows that in practice, it is easy to adjust $s$ to keep most of the interesting guides without introducing incorrect guides.

\subsection{Computing the guide trajectory in $C_{reach}^s$}
Once a value of $s$ has been fixed, any sampling-based motion planner can be used to plan a trajectory in $C_{reach}^s$. 
Indeed, contrary to $C_{Contact}$,  $C_{reach}^s$ has a non zero measure in the configuration space $C$. Therefore a standard uniform sampling approach
can work, in spite of a high rejection rate. 
Thus, the only significant change regarding a classical planner is to replace the classical collision checking method with a test of appartenance to $C_{reach}^s$ when verifying
that drawn configurations and associated local paths are valid.

However, to improve the sampling efficiency while remaining probabilistically complete, we bias the sampling process to generate near obstacle configurations, similarly to~\cite{Amato98choosinggood}.
First, a configuration is set to a random point on the surface of one randomly chosen obstacle. The root location is then translated and rotated randomly until the reachability condition is satisfied.
Our current implementation of these modifications is based on the Bi-RRT planner provided by the hpp software.

Thanks to this small set of modifications, the high dimensional truly feasible path planning problem is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for Hyq, 8 for HRP-2 that has a flexible torso). In so doing we make the problem tractable for sampling based approaches, with extremely efficient computation times.

%~ Algorithm~\ref{alg:generate} describe the sampling process for one configuration.

%~ Two user defined numeric values, $LIMIT1$ and $LIMIT2$ are used to ensure that the generation will not loop
%~ infinitely around areas of the workspace where the reachability condition cannot be met.
%~ Two configurations are connected if a linearly interpolated path between them also belongs to 
%~ $C_{reach}$.

%~ \begin{algorithm}
%~ \caption{Generation of a configuration $\mathbf{q}^{0} \in C_{reach}^s$} \label{euclid}
	%~ \begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
		%~ \State $\mathbf{q}^{0} \gets [0,...,0]^T$ /*Set all DOFs to 0*/
		%~ \State $\textit{ntry1} \gets 0; \textit{ntry2} \gets 0;$
		%~ \While {$\textit{ntry1} < \textit{LIMIT1}$}
			%~ \State Select an obstacle $O_i \in O$
			%~ \While {$\textit{ntry2} < \textit{LIMIT2}$}				
				%~ \State Translate $\mathbf{q}^{0}$ to a random point on the surface of $O_i$
				%~ \State $\mathbf{d} \gets $ Random unit direction
				%~ \While {$W(\mathbf{q}^{0}) \cap O_i \ne \emptyset$}
					%~ \State $\epsilon \gets$ random number, $ 0 < \epsilon \leq 1$
					%~ \State Translate $\mathbf{q}^{0}$ by $\mathbf{u} = \epsilon \mathbf{d}$
					%~ \If {$W^0_s(\mathbf{q}^{0}) \cap O = \emptyset$}
						%~ \State \textbf{return} $\mathbf{q}^{0}$
					%~ \Else
						%~ \State Rotate $\mathbf{q}^{0}$ randomly
						%~ \State $\textit{ntry2} \gets \textit{ntry2} + 1$
					%~ \EndIf
				%~ \EndWhile
			%~ \EndWhile
			%~ \State $\textit{ntry1} \gets \textit{ntry1} + 1$
		%~ \EndWhile
		%~ \State \textbf{return} failure
	%~ \EndFunction
%~ \end{algorithmic}
%~ \label{alg:generate}
%~ \end{algorithm}

%To summarize, in this section we define a class of spaces $C_{reach}^s$.We  propose to use RB-PRM, a sampling based motion planner, to compute guide trajectories entirely contained in $C_{reach}^s$. Such trajectories provide a guaranty that a robot will be able to create contacts along them with a probability depending on a parameter $s$.

### 2015-12-02 body/appendix_algo.tex
finished contact generation now heuristics

\section{Algorithms for discretizing of a path}
\label{app:contact}

First, we define an abtract structure State,
that describes a contact configuration.
The use of queues allows a FIFO approach regarding the order 
in which contacts are tested: we try to replace older contacts first when necessary.
Thus the algorithm is deterministic even though it can handle acyclic motions.

\begin{lstlisting}]
Struct Limb
{
    // Limb Configuration
    Configuration qk;
    // Effector position in
    // world coordinates
    vector6 pk;
};

Struct State
{
    // root location
    Configuration q0;
    // List of limbs not in contact
    queue<Limb> freeLimbs;
    // List of limbs in contact
    queue<Limb> contactLimbs;
};
\end{lstlisting}

From the start configuration, given as an input by the user,
we create the initial state $s0$.
Algorithm~\ref{alg:interpolate}  is then called with $s0$, as well as the discretized path 
$\mathbf{Q}^0$, as input parameters.

\begin{algorithm}
\caption{Discretization of a path} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{Interpolate}{$s0$,$\mathbf{Q}^0$, $MAX\_TRIES$}
		\State $list<$State$>$ $states = [s0]$
		\State $nb\_fail = 0$ 
		\State $i = 1;$ /*Current index in the list*/
		\While {$i < length(\mathbf{Q}^0)$}
			\State State $previous = last\_element(states)$
			\State State $s = \pi(previous, \mathbf{Q}^0[i])$
			\If {$s != NULL $}
				\State $nb\_fail = 0$
				\State $i += 1$
				\State \textbf{return} $\mathbf{q}^{0}$
			\Else
				\State $nb\_fail += 1$
				\If {$nb\_fail == MAX\_TRIES$}
					\State \textbf{return} $FAILURE$
				\EndIf				
				\State $s = $\textsc{RepositionContacts}$(previous)$
			\EndIf
			\State $push\_back(states, s)$
		\EndWhile
		\State \textbf{return} $states$
	\EndFunction
\end{algorithmic}
\label{alg:interpolate}
\end{algorithm}

At each step, $g$ is called with the previous state as a parameter, as well
as a new root placement. $g$ returns a new contact configuration, if it suceeded
in computing a balance configuration with only one contact switch occuring.
If $g$ failed in achieving this, the method \textsc{RepositionContacts} is called.
It repositions one end effector (either a free limb, or the oldest active contact) towards a new balanced contact position if possible.
This repositionning allows to increase the odds that the contact can be maintained at the next step.

The pseudo code for the method $g$ is given by Algorithm~\ref{alg:pi}.

\begin{algorithm}
\caption{Full body contact generation method} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{$g$}{$pState$,$\mathbf{q}^0$}
		\State State $newState$
		\State $newState.q0 = \mathbf{q}^0$
		\State $newState.freeLimbs = pState.freeLimbs$
		\State /*First try to maintain previous contacts*/
		\State $nbContactsBroken = 0$
		\For {\textbf{each} Limb $k$ in $previous.contactLimbs$}
			\If {$!$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
				\State $nbContactsBroken += 1$
				\If {$nbContactsBroken > 1$}				
					\State \textbf{return} $NULL$
				\EndIf				
				\State $push(newState.freeLimbs,k)$
			\Else 					
				\State $push(newState.contactLimbs,k)$
			\EndIf
		\EndFor
		\For {\textbf{each} Limb $k$ in $previous.freeLimbs$}
			\State /*Generate a balanced contact*/
			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
				\State $push(newState.contactLimbs,k)$			
				\State \textbf{return} $newState$
			\EndIf
		\EndFor
		\State \textbf{return} $NULL$
	\EndFunction
\end{algorithmic}
\label{alg:pi}
\end{algorithm}

The method \textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$ performs inverse kinematics to reach the previous contact position for the Limb.
If it succeeds, the new limb configuration is assigned to $k$. If it fails, a random collision free configuration is assigned to $k$.

\textsc{GenerateContact}$(\mathbf{q}^0,k)$ is a call to the contact generator presented in Section~\ref{sec:single_contact}. It generates a balanced contact configuration, and assigns the corresponding configuration to $k$.
If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.

The pseudo code for the method \textsc{RepositionContacts} is given by Algorithm~\ref{alg:repo}.

\begin{algorithm}
\caption{Performs contact repositioning for one limb} \label{interpolate}
	\begin{algorithmic}[1]
	\Function{RepositionContacts}{$state$}
		\State $i=0$
		\While {$i<length(states.freeLimbs)$}
			\State Limb $k = pop(states.freeLimbs)$
			\If {\textsc{GenerateContact}$(state.q0,k)$}	
				\State $push(newState.contactLimbs,k)$			
				\State \textbf{return}
			\Else
				\State $i+=1$
				\State $push(states.freeLimbs,k)$		
			\EndIf
		\EndWhile
		\State $i=0$
		\While {$i<length(states.contactLimbs)$}
			\State Limb $k = pop(states.contactLimbs)$
			\State Limb $copy = k$
			\State $i+=1$
			\If {\textsc{GenerateContact}$(state.q0,k)$}	
				\State $push(newState.contactLimbs,k)$			
				\State \textbf{return}
			\Else
				\State $push(newState.contactLimbs,copy)$	
			\EndIf
		\EndWhile
		/*Fails if impossible to relocate any effector*/
		\State \textbf{return} $FAILURE$
	\EndFunction
\end{algorithmic}
\label{alg:repo}
\end{algorithm}

### 2015-12-02 body/contact.tex
finished contact generation now heuristics

@@
-\section{From a guide path to a discrete sequence of contact configurations (Stage 2)}
+\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
 \label{sec:contact}
@@
 
-\subsection{Single contact generation}
+In what follows a \textit{valid} contact configuration is understood as being collision free, respecting joint limits, and allowing to maintain the quasi static balance
+of the robot. The first two tests are trivial; our criterion to assert efficiently the quasi static balance of the system
+is described in Section TODO.
+
+\subsection{Definition of a contact generator}
+\label{sec:single_contact}
 We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$ truly feasible regarding $R^k$. FIGURE
 From true feasibility we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify Eq.~\eqref{eq:pi} for $\mathbf{q}^0$.
-The issue is then to evaluate locally any element of $P$ at $\mathbf{q}^0$.
-This can be achieved as follows:
+The issue is then to evaluate locally at least one element of $P$ at $\mathbf{q}^0$.
+
+While exhibiting analytically a $\pi_i$ does not seem tractable, we can iteratively evaluate $\pi_i(\mathbf{q}^0)$ as follows:
 \begin{enumerate}
@@
 \item Project the end-effector onto the closest surface with inverse kinematics;
-\item If a feasible solution is found, stop. Otherwise repeat from step 1.
+\item If a valid solution is found, stop. Otherwise repeat from step 1.
 \end{enumerate}
 
-We partially trade the completeness property of this approach for a more efficient solution \citep{Tonneau2014}:
+We partially trade the completeness property of this approach for a more efficient solution previously introduced \citep{Tonneau2014}, that we describe
+here for completeness.
+
+We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimal distance 
+between an effector and a surface of the environment is lesser than a small value $\epsilon$.
+We then apply the following steps:
 \begin{enumerate}
-\item Generate offline $N$ valid sample limb configurations;
-\item using the end-effector position expressed in $R$ coordinates as an index, store each sample in an octree stata structure;
-\item Figure TODO, 2 and 3: At runtime, when contact creation is required, retrieve from the octree the list of samples $S = [\mathbf{q}^k_0, ..., \mathbf{q}^k_i, ...]$ such that $dist(\mathbf{p}^k(\mathbf{q_k}^k),O) < \epsilon$, with $\epsilon $ small ;
+\item Generate offline $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
+\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree stata structure;
+\item Figure~\ref{fig:contact_gen}--2 and 3--: At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
 \item According to a user-defined heuristic $h$, sort the elements of $S$;
-\item FIGURE TODO, 4 and 5: Select the first element of $S$, and project the configuration to contact with inverse kinematics;
-\item If a feasible solution is found, stop. Otherwise remove the element from $S$ and go from step 5, or fail if $S$ is empty.
+\item Figure~\ref{fig:contact_gen}--4 and 5--: Select the first configuration of $S$ (or fail if $S$ is empty). Project the configuration onto contact with inverse kinematics;
+\item If a valid solution is found, stop. Otherwise remove the element from $S$ and go back step 5.
 \end{enumerate}
@@
 This approach has two main advantages:
-First, this allows us to select a number of candidates at a small cost, (a single proximity request), and to sort them according to a heuristic;
-Secondly, thanks to the use of proximity requests, the number of iterations necessary for the inverse kinematics to converge is reduced.
-Finally, it is easy to see that as $N$ grows to infinity, the completeness of the method is restored.
+First, this allows us to select a large number of candidates in a single proximity request.
+Having several candidates is interesting, because it allows to compare them regarding a user selected heuristic $h$, thus obtaining
+a locally optimal candidate.
+Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution,
+and this in a small number of iterations.
+Regarding completeness, it is immediate to verify that as $N$ and $\epsilon$ grow, the completeness of the method is restored.
 $N$ is a parameter that allows us to specify the trade-off between completeness and efficiency.
@@
 
-\subsubsection{Efficient contact generation }
-
-Concretely, any, we can use a random sampling approach combined with an inverse kinematics projector.
-
-given a root configuration $\mathbf{q}^0$ and a limb $R^k$, we locally identify a countable set of mappings $P = [\pi_0,..,\pi_i] $ that verify Eq.~\eqref{eq:pi} for $\mathbf{q}^0$.
-and select the most relevant according to a heuristic $h$.
-
+We thus have defined a real time, extremely efficient contact generator, that can be as complete as required by the context.
 %We compute the image of a mapping $f$ along $\mathbf{q}^0(t)$ into a discrete sequence of contacts $ \in C_{contact}$.
@@
 \subsection{Extension of the guide trajectory}
-The guide trajectory $\mathbf{q}^0(t)$ is first discretized into a sequence of $j$ key placements:  
+Using the contact generator, we define an iterative algorithm to generate the contact
+sequence along the guide path.
+Although we address acyclic contact sequences, the algorithm is deterministic in the order in which 
+the contacts are created, allowing it to break the combinatorial.
+The complete algorithm can be found in Appendix~\ref{app:contact}.
+In the remainder of the section we provide an intuition of it.
+
+As an input, we consider the truly feasible guide path $\mathbf{q}^0(t)$, automatically discretized into a sequence of $j$ key placements:  
 \begin{equation*}
@@
 We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
-The function $g$ is defined independently by $g^k$ for each limb $R^k$. In defining $g^k$, two aspects must be considered. Is the limb $R^k$ in contact? And which criteria is it optimizing?
 
-\medskip%\subsection{Maintaining and breaking contact in the sequence}
-\noindent\textbf{Maintaining a contact:} If possible, a limb in contact at time $i-1$ remains in contact at $i$.
-The contact is broken if an inverse kinematics solver fails to find a collision free limb configuration which satisfies joint limits.
-The solver is a direct implementation of the inverse kinematics solver with task priorities framework proposed in \cite{Baerlocher2004}, which
-handles explicitely joint limits, and allows to integrate collision avoidance as a secondary criterion.
+At each step, we consider a new root placement $\mathbf{q}^0_{i}$ and try to generate a balanced contact configuration that preserves as much as possible
+the previous contacts.
+
+%~ The function $g$ is defined independently by $g^k$ for each limb $R^k$. In defining $g^k$, two aspects must be considered. Is the limb $R^k$ in contact? And which criteria is it optimizing?
+
+\subsubsection{Maintaining a contact in the sequence.}
+
+Figure~\ref{fig:break_contact} illustrates the contact persistence method.
+If possible, a limb in contact at time $i-1$ remains in contact at $i$.
+The contact is broken if an inverse kinematics solver fails to find a collision free limb configuration that satisfies joint limits.
+The solver is directy provided by the Hpp software.
 
 If the solver fails, the contact is broken and a collision free configuration is assigned to the limb.
-%~ The configuration is selected to be the closest to the previous configuration (by minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$).
-
-Once a first candidate configuration is taken for all limbs, the quasi-static balance is tested by whether the weight wrench is in the gravito-inertial cone (i.e.
-there exists valid contact forces that compensate for the weight of the robot), using the geometric approach described in \cite{qiu:dhm:2011}.
-If the balance is not obtained, new contacts are randomly generated using the following procedure.
-
-%\begin{itemize} 
-%\item If the robot is not statically balanced and $\mathbf{q}_{i-1}^k$ is not in contact, $g$ computes a configuration $\mathbf{q}_{i}^k$ so that a contact is taken with the environment (Fig.~\ref{fig:generate_contact}).
-%\item Else, $g$ computes a configuration $\mathbf{q}_{i}^k$ minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$, subject to $R^k(\mathbf{q_i}) \cap O = \emptyset$ and $\mathbf{q_i}^k$ respects the  joint limits of $R^k$.
-%
-%If $\mathbf{q}_{i - 1}^k$ is in contact, this results in trying to maintain the contact for $\mathbf{q}_{i}^k$ (Figure~\ref{fig:break_contact}--1 and 2).
-The minimum is different than zero if $W^k(\mathbf{q}_i^0) \cap  \mathbf{p}^k(\mathbf{q_{i-1}}) = \emptyset$ (Figure~\ref{fig:break_contact}--3), or collision can not be  prevented (Figure~\ref{fig:break_contact}--4).
-%\end{itemize}
+If more than one contacts are broken for a single step, intermediate steps are added to reposition, one at a time,
+the contacts, before moving.
 
@@
 \centering
-  \begin{overpic}[width=0.6\linewidth]{figures/break_contact}
-		\put (3,15) {1)} 
-		\put (27,15) {2)} 
-		\put (50,15) {3)} 
-		\put (75,15) {4)} 
+  \begin{overpic}[width=0.9\linewidth]{figures/break_contact}
+		\put (0,4) {1} 
+		\put (25,4) {2} 
+		\put (50,4) {3} 
+		\put (76,4) {4} 
 		%~ \put (68,58) {3.a)} 
@@
 	\end{overpic}
-\caption{Contacts are maintained unless their position is too far, or the environment prevents it.}
+\caption{Contacts are maintained if joint limits and collisions constraints are respected (2). They are broken otherwise(3,4).}
 		   \label{fig:break_contact}
@@
 
-\begin{figure}[t]
-\centering
-  \begin{overpic}[width=0.6\linewidth]{figures/generate_contact}
+%~ \begin{figure}[t]
+%~ \centering
+  %~ \begin{overpic}[width=0.6\linewidth]{figures/generate_contact}
 		%~ \put (5,58) {1)} 
@@
 		%~ \put (68,27) {4.b)} 
-	\end{overpic}
-\caption{Contacts are generated when the configuration is not balanced.}
-		   \label{fig:generate_contact}
-\end{figure}
+	%~ \end{overpic}
+%~ \caption{Contacts are generated when the configuration is not balanced.}
+		   %~ \label{fig:generate_contact}
+%~ \end{figure}
 
 %\subsection{Generation of a contact configuration}  
-\medskip\noindent\textbf{Creating a contact:} We consider a configuration where some limbs are in contact, some are free and quasi-static balance is not enforced.
-To enforce balance, we proceed in the following
-manner:
-we randomly select a contact free limb; if there is no contact free limb, we select
-the limb that made contact first.
-Using the contact generator introduced in~\cite{Tonneau2014}, we project the configuration of this limb into a contact that enhances balance, if it exists (Figure~\ref{fig:generate_contact}); If balance is not achievable and a contact is possible, it is generated anyway;
-If balance is not achieved, the next limb is selected and projected into a contact configuration, and so on.
-This approach can lead to the repositioning of existing contacts, in which case intermediate states are inserted to reposition the contacts.
-This current implementation does not guarantee that the planner will succeed in generating a balanced configuration, because
-true feasibility is not fully guaranteed.
-However in practice the planner is successful in the large majority of cases, as discussed in section~\ref{sec:stat1}.
-%~ A new contact configuration is obtained by projecting a random configuration to the nearest contacts.
-The sampling is biased by selecting configurations so that the limbs previously in contact are already close to contact and at least one additional limb is close to the environment.
-The random configuration is then projected onto the contact by iteratively minimizing the distance between the end effectors and the environment, under the constraint of respecting the joint limits and avoiding the collision.
+\subsection{Creating contacts.}
+Contacts are created using three rules:
+\begin{enumerate}
+\item Only one contact creation can happen between two consecutive states. This increases the odds that the interpolation between the two states will be feasible; 
+\item A contact is validated if and only if the resulting configuration is statically balanced, no matter what user-defined heuristics are used; 
+\item We use a FIFO approach:  we always try to first create a contact with the limb that has been free for the longest time. If the contact creation
+was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others have been treated.
+\end{enumerate}
 
-\subsection{Contribution to the global movement: the EFORT criteria}
-\noindent\textbf{EFORT criterion:} If only relying on the random sampling to select new contacts, the planner produces inefficient postures.
-The resulting contact sequence is then poorly efficient and unnatural.
-Moreover, the limbs are not well configured and are not able to efficiently follow the general movement: contacts break frequently.
-
-When creating additional contacts, we therefore propose to select particular configurations that allow to exert a force compatible with the direction of motion.
-This task efficiency is measured  with the Extended FORce Transmission ratio (EFORT)~\cite{Tonneau2014}.
-%
-The measure of EFORT is given by
-\begin{equation}
-\alpha_{EFORT}(\mathbf{q}^k, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \nu_0 \mathbf{n}^T \mathbf{m})
-\end{equation}
-where $\mathbf{J}$ is the Jacobian matrix of the limb $R^k$ in configuration $\mathbf{q}^k$; $\nu_0$ is the friction coefficient of the contact surface; $\mathbf{n}$ is the normal of the contact surface; and $\mathbf{m}$ is the direction opposite to the motion,
-given by the 3D vector connecting $\mathbf{q}_{i}^0$ and $\mathbf{q}_{i+1}^0$.
-%
-The first part of the equation measures the ratio between the joint  torques and the resulting force applied along $\mathbf{m}$.
-The second part quantifies the odds of slipping while applying a force along $\mathbf{m}$.
-
-\noindent\textbf{Optimization at creation:} In practice, a database of configurations is stored for each limb, which can be considered as manipulator arms.
-The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations (and additionally storing $\mathbf{J}$).
-%Fig.~\ref{fig:contact_gen}-2 presents a few configurations for the right arm of the robot.  
-Upon request, the octree returns a set of configurations close to contact (Fig.~\ref{fig:contact_gen}-3).
-These candidates are sorted based on their task efficiency, given by $\alpha_{EFORT}$.
-The first candidate in this list satisfying the balance criterion and is collision free is selected and projected on the contact surface using our inverse kinematics solver.
-
-%Therefore to generate a contact, we want to find a configuration which is collision free, allows to maintain balance, and maximizes EFORT.
-We choose this configuration among a set of possible candidates, rapidly computed thanks to a contact generator, also proposed in~\cite{Tonneau2014}.
-We recall the overall idea here.
-We consider each limb of the robot as a manipulator arm attached to the root.
-We store a database of configurations for each manipulator.
-The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations.
-The jacobian matrix $\mathbf{J}$ of each configuration is also stored.
-Figure~\ref{fig:contact_gen} -- 2 presents a few configurations for the right arm of the robot.
-
-%To summarize: given a feasible trajectory for the root of the robot $R$, we are able to generate a discrete sequence of contact configurations  describing a feasible motion for $R$, between a start an goal configurations.
-The contact configurations are chosen to be statically balanced and task efficient, which means that they can generate a force  contributing to the motion of the root along the trajectory.
+These three rules provide a deterministic manner to generate contacts along the discretized guide path.
+FIGURE. For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a balanced configuration, because
+of the deterministic approach used to break the combinatorial.
+However, in practice the planner is successful in the large majority of cases, as discussed in section~\ref{sec:stat1}, thanks to 
+the heuristics we use for relevant contact selection.

### 2015-12-02 body/equilibrium.tex
finished contact generation now heuristics

--- 2015-12-01 body/efort.tex
@@
-\section{From a guide trajectory to a discrete sequence of contact configurations (Stage 2)}
-\label{sec:efort}
-As an input of this stage, we consider a truly feasible guide trajectory $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$.
-We now consider the second problem of computing a trajectory $\mathbf{q}^{\overline{0}}(t)$ for the limbs of the robot.
-Since we assume true feasibility, we know that such a trajectory exists.
-Contrary to previous works \cite{DBLP:conf/iser/EscandeKMG08,Bouyarmane2009}, the goal here is not to find any such trajectory but rather to select one with particular properties.
-Specifically, we show here how to build a contact sequence with a small number of contact variations and  good-efficiency and naturalness of the postures.
+\section{Heuristics for relevant contact selection}
 
-More precisely, any mapping $\pi$ introduced in \eqref{eq:pi} can be used to expand $\mathbf{q}^0(t)$ into a whole-body trajectory.
-We propose here a particular construction of $\pi$ leading to interesting contact sequences.
-
-%We compute the image of a mapping $f$ along $\mathbf{q}^0(t)$ into a discrete sequence of contacts $ \in C_{contact}$.
-We first describe the discretization of $\mathbf{q}^0(t)$ and the algorithm used to create or break contacts along it.
-Then we describe how contacts are explicitely generated along the discretized trajectory.
-
-\subsection{Extension of the guide trajectory} 
-The guide trajectory $\mathbf{q}^0(t)$ is first discretized into a sequence of $j$ key placements:  
-\begin{equation*}
-	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
-\end{equation*} 
-where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations.
-To ensure continuity in the contact transition phases, we rewrite $\pi$ under the following recursive form for any $0<i<j$:
-\begin{equation*}
-    \pi\colon\left\{
-    \begin{aligned}		
-        \mathbf{Q}^0 \in C_{reach} & \longrightarrow C_{contact} \\
-        %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
-        \mathbf{q}^{0}_i &  \longrightarrow  \mathbf{q}^{0}_i \oplus g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
-    \end{aligned}
-    \right.
-\end{equation*} 
-We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
-The function $g$ is defined independently by $g^k$ for each limb $R^k$. In defining $g^k$, two aspects must be considered. Is the limb $R^k$ in contact? And which criteria is it optimizing?
-
-\medskip%\subsection{Maintaining and breaking contact in the sequence}
-\noindent\textbf{Maintaining a contact:} If possible, a limb in contact at time $i-1$ remains in contact at $i$.
-The contact is broken if an inverse kinematics solver fails to find a collision free limb configuration which satisfies joint limits.
-The solver is a direct implementation of the inverse kinematics solver with task priorities framework proposed in \cite{Baerlocher2004}, which
-handles explicitely joint limits, and allows to integrate collision avoidance as a secondary criterion.
-
-If the solver fails, the contact is broken and a collision free configuration is assigned to the limb.
-%~ The configuration is selected to be the closest to the previous configuration (by minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$).
-
-Once a first candidate configuration is taken for all limbs, the quasi-static balance is tested by whether the weight wrench is in the gravito-inertial cone (i.e.
-there exists valid contact forces that compensate for the weight of the robot), using the geometric approach described in \cite{qiu:dhm:2011}.
-If the balance is not obtained, new contacts are randomly generated using the following procedure.
-
-%\begin{itemize} 
-%\item If the robot is not statically balanced and $\mathbf{q}_{i-1}^k$ is not in contact, $g$ computes a configuration $\mathbf{q}_{i}^k$ so that a contact is taken with the environment (Fig.~\ref{fig:generate_contact}).
-%\item Else, $g$ computes a configuration $\mathbf{q}_{i}^k$ minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$, subject to $R^k(\mathbf{q_i}) \cap O = \emptyset$ and $\mathbf{q_i}^k$ respects the  joint limits of $R^k$.
-%
-%If $\mathbf{q}_{i - 1}^k$ is in contact, this results in trying to maintain the contact for $\mathbf{q}_{i}^k$ (Figure~\ref{fig:break_contact}--1 and 2).
-The minimum is different than zero if $W^k(\mathbf{q}_i^0) \cap  \mathbf{p}^k(\mathbf{q_{i-1}}) = \emptyset$ (Figure~\ref{fig:break_contact}--3), or collision can not be  prevented (Figure~\ref{fig:break_contact}--4).
-%\end{itemize}
-
-\begin{figure}[t]
-\centering
-  \begin{overpic}[width=0.6\linewidth]{figures/break_contact}
-		\put (3,15) {1)} 
-		\put (27,15) {2)} 
-		\put (50,15) {3)} 
-		\put (75,15) {4)} 
-		%~ \put (68,58) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
-	\end{overpic}
-\caption{Contacts are maintained unless their position is too far, or the environment prevents it.}
-		   \label{fig:break_contact}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-  \begin{overpic}[width=0.6\linewidth]{figures/generate_contact}
-		%~ \put (5,58) {1)} 
-		%~ \put (37,58) {2)} 
-		%~ \put (68,58) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
-	\end{overpic}
-\caption{Contacts are generated when the configuration is not balanced.}
-		   \label{fig:generate_contact}
-\end{figure}
-
-%\subsection{Generation of a contact configuration}  
-\medskip\noindent\textbf{Creating a contact:} We consider a configuration where some limbs are in contact, some are free and quasi-static balance is not enforced.
-To enforce balance, we proceed in the following
-manner:
-we randomly select a contact free limb; if there is no contact free limb, we select
-the limb that made contact first.
-Using the contact generator introduced in~\cite{Tonneau2014}, we project the configuration of this limb into a contact that enhances balance, if it exists (Figure~\ref{fig:generate_contact}); If balance is not achievable and a contact is possible, it is generated anyway;
-If balance is not achieved, the next limb is selected and projected into a contact configuration, and so on.
-This approach can lead to the repositioning of existing contacts, in which case intermediate states are inserted to reposition the contacts.
-This current implementation does not guarantee that the planner will succeed in generating a balanced configuration, because
-true feasibility is not fully guaranteed.
-However in practice the planner is successful in the large majority of cases, as discussed in section~\ref{sec:stat1}.
-%~ A new contact configuration is obtained by projecting a random configuration to the nearest contacts.
-The sampling is biased by selecting configurations so that the limbs previously in contact are already close to contact and at least one additional limb is close to the environment.
-The random configuration is then projected onto the contact by iteratively minimizing the distance between the end effectors and the environment, under the constraint of respecting the joint limits and avoiding the collision.
-
-\subsection{Contribution to the global movement: the EFORT criteria}
+\subsection{A heuristic for robust quasi static balance}
 \noindent\textbf{EFORT criterion:} If only relying on the random sampling to select new contacts, the planner produces inefficient postures.

### 2015-12-02 body/intro.tex
finished contact generation now heuristics

@@
 
+\begin{figure*}
+  \centering
+  \begin{overpic}[width=0.8\linewidth]{figures/workflow}
+    %~ \put (6.4,1.8) {\normalsize{Request}} 
+    \put (30,10.8) {\large{\color{white}\Pa} }
+    %~ \put (31.6,2) {\scriptsize{\color{white}RB-RRT}} 
+    \put (66.4,10.8) {\large{\color{white}\Pb} }
+    %~ \put (72.5,2) {\tiny{\color{white}Contact}} 
+  \end{overpic}
+  \vspace{-1em}
+  \caption{
+    Overview of our 2-stage framework.
+\Pa: Given a path request between the yellow and blue positions, a guide path is computed in the space of truly feasible guides $C_{reach}$.
+This is achieved by defining a geometric condition, the reachability condition (abstracted here with the transparent cylinders).
+\Pb: The trajectory is extended into a discrete sequence of contact configurations using an iterative algorithm.}
+  \label{fig:framework}
+\end{figure*}
+
 To address \Pa we rigorously formulate a geometric condition for the true feasibility of a path, the reachability condition. We claim that this 

### 2015-12-02 body/overview.tex
finished contact generation now heuristics

@@
 
+Figure~\ref{fig:framework} illustrates our workflow.
+\Pa and \Pb are addressed in a sequential fashion.
+The guarantees provided by the truly feasible guide path allow
+us to avoid the combinatorial by generating the contacts in a straightforward manner.
+
 %
-\begin{figure}
-  \begin{overpic}[width=0.9\linewidth]{figures/contact_gen}
-		\put (3,3) {1)} 
-		\put (25,3) {2)} 
-		\put (44,3) {3)} 
-		\put (63,3) {4)} 
-		\put (82,3) {5)} 
+\begin{figure*}
+  \centering
+  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
+		\put (1,1) {1} 
+		\put (22,1) {2} 
+		\put (42,1) {3} 
+		\put (62,1) {4} 
+		\put (83,1) {5} 
 	\end{overpic}
-  \caption{Generation of a contact configuration for the right arm of a humanoid robot.
-1) Selection of reachable obstacles.
-2) A request is performed on a database of configurations.
-3) Configurations too far from contact are eliminated.
-4) The best candidate according to EFORT is chosen.
-5) The final contact is achieved using inverse kinematics.}
+  \caption{Generation of a contact configuration for the right leg of HRP-2.
+1: Selection of reachable obstacles.
+2: Entries of the limb samples database (with $N = 4$).
+3: With a proximity query on the octree database, configurations too far from obstacles are eliminated.
+4: The best candidate according to a user-defined heuristic $h$ is chosen.
+5: The final contact is achieved using inverse kinematics.}
   \label{fig:contact_gen}
-\end{figure}
-\subsection{Computation of a guide path}
+\end{figure*}
+\subsection{Computation of a guide path -- \Pa}
 We first consider the problem of planning a relevant guide path.
@@
  An intuitive description of such placements is "close, but not too close": close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Fig.~\ref{fig:contact_gen}--1); not too close, because the robot must avoid collision (which is represented by the hull including the torso in Fig.~\ref{fig:contact_gen}--1).
-We define formally $C_{reach}$, the set of interesting root placements, in which we compute a guide trajectory with a sampling based planner,  the Reachability-Based RRT --RB-RRT-- (Figure~\ref{fig:framework}--A).
+We define formally $C_{reach}$, the set of interesting root placements, in which we compute a guide trajectory with a sampling based planner,  the Reachability-Based RRT --RB-RRT-- (Figure~\ref{fig:framework}--\Pa).
 Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
@@
 
-\subsection{Generating a discrete sequence of contact configurations}
+\subsection{Generating a discrete sequence of contact configurations -- \Pb}
 
-The second stage extends the guide path into a sequence of contact configurations (Fig.~\ref{fig:framework}--B). The true feasibility of the input path is guaranteed by the first step.
+The second stage extends the guide path into a sequence of contact configurations (Fig.~\ref{fig:framework}--\Pb). The true feasibility of the input path is guaranteed by the first step.
 This provides in turn a reasonnable guarantee that a contact configuration can be obtained for any root placement along the path. With a fixed root position, the dimensionality of the 
@@
 
-\begin{figure}[][t]
+\begin{figure}
   \centering
-  \begin{overpic}[width=0.46\linewidth]{figures/character}
-    \put (7,29) {\tiny{$\in \mathbf{q}^1$}}
-    \put (7,25) {\tiny{effector}}
+  \begin{overpic}[width=0.8\linewidth]{figures/character}
+    \put (7,29) {\small{$\in \mathbf{q}^1$}}
+    \put (7,25) {\small{effector}}
   \end{overpic}
@@
 The red geometry denotes $W^0$ and must remain collision-free.
-The green spheres are the  $W^k$.}
+The green spheres are the reachable workspace of each limb, the  $W^k$.}
   \label{fig:character}

### 2015-12-02 body/rbprm.tex
finished contact generation now heuristics

@@
-\section{Computation of a guide path in $C_{reach}$ (Stage 1)}
+\section{Computing a guide path in $C_{reach}$ ($\mathcal{P}_1$) }
 \label{rbprm}
 
-We consider the problem of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the geometrical root of a multiped robot, connecting start and goal configurations.
+We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the geometrical root of a multiped robot, connecting start and goal configurations.
 As said in the previous section, the goal is to enforce that any configuration $\mathbf{q}^0$ of the guide is truly feasible, i.e.

### 2015-12-04 body/efort.tex
robust equilibrium

@@
-\section{From a guide trajectory to a discrete sequence of contact configurations (Stage 2)}
-\label{sec:efort}
-As an input of this stage, we consider a truly feasible guide trajectory $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$.
-We now consider the second problem of computing a trajectory $\mathbf{q}^{\overline{0}}(t)$ for the limbs of the robot.
-Since we assume true feasibility, we know that such a trajectory exists.
-Contrary to previous works \cite{DBLP:conf/iser/EscandeKMG08,Bouyarmane2009}, the goal here is not to find any such trajectory but rather to select one with particular properties.
-Specifically, we show here how to build a contact sequence with a small number of contact variations and  good-efficiency and naturalness of the postures.
+\subsection{Manipulability-based heuristics for locally optimal contact selection}
+The other kind of heuristics we use are based on the analysis of the motion capabilities
+provided by a candidate configuration regarding the task being achieved.
+These heuristics allow to locally optimize a contact location regarding a particular task.
+For instance, once can be interested in configurations that allow to efficienctly exert a force compatible with the global direction of motion,
+a high velocity in a given direction, or to stay away from singular configurations.
+In these section, three such heuristics are derived in this section from the work on manipulability by \cite{Yoshikawa1984}, that we present first.
 
-More precisely, any mapping $\pi$ introduced in \eqref{eq:pi} can be used to expand $\mathbf{q}^0(t)$ into a whole-body trajectory.
-We propose here a particular construction of $\pi$ leading to interesting contact sequences.
+\subsubsection{The force and velocity ellipsoids.}
+For a limb configuration $\mathbf{q}^k$, the jacobian matrix
+ $\mathbf{J}(\mathbf{q}^k)$   defines the relation:
 
-%We compute the image of a mapping $f$ along $\mathbf{q}^0(t)$ into a discrete sequence of contacts $ \in C_{contact}$.
-We first describe the discretization of $\mathbf{q}^0(t)$ and the algorithm used to create or break contacts along it.
-Then we describe how contacts are explicitely generated along the discretized trajectory.
+\begin{equation} \label{jac}
+\mathbf{\dot{p}^k} = \mathbf{J}(\mathbf{q}^k) \dot{\mathbf{q}^k}
+\end{equation}
 
-\subsection{Extension of the guide trajectory} 
-The guide trajectory $\mathbf{q}^0(t)$ is first discretized into a sequence of $j$ key placements:  
-\begin{equation*}
-	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
-\end{equation*} 
-where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations.
-To ensure continuity in the contact transition phases, we rewrite $\pi$ under the following recursive form for any $0<i<j$:
-\begin{equation*}
-    \pi\colon\left\{
-    \begin{aligned}		
-        \mathbf{Q}^0 \in C_{reach} & \longrightarrow C_{contact} \\
-        %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
-        \mathbf{q}^{0}_i &  \longrightarrow  \mathbf{q}^{0}_i \oplus g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
-    \end{aligned}
-    \right.
-\end{equation*} 
-We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
-The function $g$ is defined independently by $g^k$ for each limb $R^k$. In defining $g^k$, two aspects must be considered. Is the limb $R^k$ in contact? And which criteria is it optimizing?
+For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}(\mathbf{q}^k)$ as $\mathbf{J}$.
 
-\medskip%\subsection{Maintaining and breaking contact in the sequence}
-\noindent\textbf{Maintaining a contact:} If possible, a limb in contact at time $i-1$ remains in contact at $i$.
-The contact is broken if an inverse kinematics solver fails to find a collision free limb configuration which satisfies joint limits.
-The solver is a direct implementation of the inverse kinematics solver with task priorities framework proposed in \cite{Baerlocher2004}, which
-handles explicitely joint limits, and allows to integrate collision avoidance as a secondary criterion.
+As a linear approximation of the forward kinematics function $f$ , $\mathbf{J}$ describes how small
+variations from the configuration $\mathbf{q}$ affect the position vector $\mathbf{p}$.
+It can be seen as the transformation mapping a velocity in the configuration space into a velocity in
+the euclidian space.
 
-If the solver fails, the contact is broken and a collision free configuration is assigned to the limb.
-%~ The configuration is selected to be the closest to the previous configuration (by minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$).
+Now we consider the unit ball in the configuration space $C$ defined by the set of joint velocities
+for which the norm is inferior or equal to 1:
+\begin{equation} \label{ball}
+||\dot{\mathbf{q}}||^2 \leq 1
+\end{equation}
 
-Once a first candidate configuration is taken for all limbs, the quasi-static balance is tested by whether the weight wrench is in the gravito-inertial cone (i.e.
-there exists valid contact forces that compensate for the weight of the robot), using the geometric approach described in \cite{qiu:dhm:2011}.
-If the balance is not obtained, new contacts are randomly generated using the following procedure.
+From Equation~\ref{jac} we can obtain the following equality:
+\begin{equation} \label{jaceq}
+\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} = \dot{\mathbf{q}}^T \dot{\mathbf{q}}
+\end{equation}
 
-%\begin{itemize} 
-%\item If the robot is not statically balanced and $\mathbf{q}_{i-1}^k$ is not in contact, $g$ computes a configuration $\mathbf{q}_{i}^k$ so that a contact is taken with the environment (Fig.~\ref{fig:generate_contact}).
-%\item Else, $g$ computes a configuration $\mathbf{q}_{i}^k$ minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$, subject to $R^k(\mathbf{q_i}) \cap O = \emptyset$ and $\mathbf{q_i}^k$ respects the  joint limits of $R^k$.
-%
-%If $\mathbf{q}_{i - 1}^k$ is in contact, this results in trying to maintain the contact for $\mathbf{q}_{i}^k$ (Figure~\ref{fig:break_contact}--1 and 2).
-The minimum is different than zero if $W^k(\mathbf{q}_i^0) \cap  \mathbf{p}^k(\mathbf{q_{i-1}}) = \emptyset$ (Figure~\ref{fig:break_contact}--3), or collision can not be  prevented (Figure~\ref{fig:break_contact}--4).
-%\end{itemize}
+We can use Equation~\ref{jaceq} to map the ball into an ellipsoid in the euclidian space $\mathbb{R}^m$:
 
-\begin{figure}[t]
+\begin{equation} \label{ellipsoid}
+\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} \leq 1
+\end{equation}
+This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of velocities that can
+be reached under the constraints of Equation~\ref{ball} for the current configuration.
+The longer an axis of the ellipsoid is, the faster the configuration can move along the direction of the axis.
+Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.
+
+\begin{figure}[htb]
   \centering
-  \begin{overpic}[width=0.6\linewidth]{figures/break_contact}
-		\put (3,15) {1)} 
-		\put (27,15) {2)} 
-		\put (50,15) {3)} 
-		\put (75,15) {4)} 
-		%~ \put (68,58) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
+	\begin{overpic}[width=1\linewidth]{figures/EFORT/ellipsoid}
+		\put (8.5,1.2) {\small{Velocity ellipsoid}}
+		\put (62.5,1.2) {\small{Force ellipsoid} \tiny{(scale 0.5)}}
 	\end{overpic}
-\caption{Contacts are maintained unless their position is too far, or the environment prevents it.}
-		   \label{fig:break_contact}
+  % \includegraphics[width=1\linewidth]{ressources/screens/sampling/canap/bd2_cmyk}
+  \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 dofs and 2 segments.
+Only the horizontal and vertical speeds are shown (not the rotation speeds), since it would require being able to draw in four dimensions.}
+		   \label{sec:efort_ellipsoid}
 \end{figure}
 
-\begin{figure}[t]
-\centering
-  \begin{overpic}[width=0.6\linewidth]{figures/generate_contact}
-		%~ \put (5,58) {1)} 
-		%~ \put (37,58) {2)} 
-		%~ \put (68,58) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
-	\end{overpic}
-\caption{Contacts are generated when the configuration is not balanced.}
-		   \label{fig:generate_contact}
-\end{figure}
+Similarly to the velocity ellipsoid, \citeauthor{Yoshikawa1984} also defines the force ellipsoid.
+Considering: a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
+the equivalent joint torque vector $\mathbf{\tau}$;
+we can define the mechanical work in both spaces:
+\begin{equation*} \label{power}
+\dot{\mathbf{q}}^T \mathbf{\tau} = \dot{\mathbf{p}}^T \mathbf{f}
+\end{equation*}
 
-%\subsection{Generation of a contact configuration}  
-\medskip\noindent\textbf{Creating a contact:} We consider a configuration where some limbs are in contact, some are free and quasi-static balance is not enforced.
-To enforce balance, we proceed in the following
-manner:
-we randomly select a contact free limb; if there is no contact free limb, we select
-the limb that made contact first.
-Using the contact generator introduced in~\cite{Tonneau2014}, we project the configuration of this limb into a contact that enhances balance, if it exists (Figure~\ref{fig:generate_contact}); If balance is not achievable and a contact is possible, it is generated anyway;
-If balance is not achieved, the next limb is selected and projected into a contact configuration, and so on.
-This approach can lead to the repositioning of existing contacts, in which case intermediate states are inserted to reposition the contacts.
-This current implementation does not guarantee that the planner will succeed in generating a balanced configuration, because
-true feasibility is not fully guaranteed.
-However in practice the planner is successful in the large majority of cases, as discussed in section~\ref{sec:stat1}.
-%~ A new contact configuration is obtained by projecting a random configuration to the nearest contacts.
-The sampling is biased by selecting configurations so that the limbs previously in contact are already close to contact and at least one additional limb is close to the environment.
-The random configuration is then projected onto the contact by iteratively minimizing the distance between the end effectors and the environment, under the constraint of respecting the joint limits and avoiding the collision.
+Deriving from Equation~\ref{jac}, we define that the set of achievable force in $\mathbb{R}^m$  subject to the constraint:
+\begin{equation*} \label{ballforce}
+||\mathbf{\tau}||^2 \leq 1
+\end{equation*}
+is the force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right) defined by:
+\begin{equation} \label{ellipsoidforce}
+\mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
+\end{equation}
 
-\subsection{Contribution to the global movement: the EFORT criteria}
-\noindent\textbf{EFORT criterion:} If only relying on the random sampling to select new contacts, the planner produces inefficient postures.
-The resulting contact sequence is then poorly efficient and unnatural.
-Moreover, the limbs are not well configured and are not able to efficiently follow the general movement: contacts break frequently.
+\subsubsection{Manipulability based heuristics.}
+From these definitions, we can derive three useful heuristics, that all account from the environment and the task being performed.
+The first one, EFORT, is introduced by \cite{Tonneau2014}; the other two are new minor contributions, derived from these previous works.
 
-When creating additional contacts, we therefore propose to select particular configurations that allow to exert a force compatible with the direction of motion.
-This task efficiency is measured  with the Extended FORce Transmission ratio (EFORT)~\cite{Tonneau2014}.
-%
-The measure of EFORT is given by
+With EFORT, we define the efficiency of a configuration as the ability a limb has to exert a force in a given direction.
+We thus consider the force ellipsoid as a basis for our heuristic.
+In a given direction $\mathbf{m}$, the length of the ellispoid is given by the force transmission ratio \citep{1087795}:
+\begin{equation*}
+f_\mathsf{T}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(JJ^{T})\mathbf{m}]^{-\frac{1}{2}}
+\end{equation*}
+
+In our problem, to compare candidate configurations, we include the quality of the contact surface, and choose $\mathbf{m}$ as the direction
+opposite to the local motion (thus given by the difference between two consecutive root positions):
+
 \begin{equation}
-\alpha_{EFORT}(\mathbf{q}^k, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \nu_0 \mathbf{n}^T \mathbf{m})
+h_{EFORT}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \mu_0 \mathbf{n}^T \mathbf{m})
 \end{equation}
-where $\mathbf{J}$ is the Jacobian matrix of the limb $R^k$ in configuration $\mathbf{q}^k$; $\nu_0$ is the friction coefficient of the contact surface; $\mathbf{n}$ is the normal of the contact surface; and $\mathbf{m}$ is the direction opposite to the motion,
-given by the 3D vector connecting $\mathbf{q}_{i}^0$ and $\mathbf{q}_{i+1}^0$.
-%
-The first part of the equation measures the ratio between the joint  torques and the resulting force applied along $\mathbf{m}$.
-The second part quantifies the odds of slipping while applying a force along $\mathbf{m}$.
+where $\mu_0$ and $\mathbf{n}$ are respectively the friction coefficient and the normal vector of the contact surface.
 
-\noindent\textbf{Optimization at creation:} In practice, a database of configurations is stored for each limb, which can be considered as manipulator arms.
-The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations (and additionally storing $\mathbf{J}$).
-%Fig.~\ref{fig:contact_gen}-2 presents a few configurations for the right arm of the robot.  
-Upon request, the octree returns a set of configurations close to contact (Fig.~\ref{fig:contact_gen}-3).
-These candidates are sorted based on their task efficiency, given by $\alpha_{EFORT}$.
-The first candidate in this list satisfying the balance criterion and is collision free is selected and projected on the contact surface using our inverse kinematics solver.
+If the ability to apply important velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasonning on the velocity ellispoid:
+\begin{equation}
+h_{vel}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{m}]^{-\frac{1}{2}} ( \mu_0 \mathbf{n}^T \mathbf{m})
+\end{equation}
 
-%Therefore to generate a contact, we want to find a configuration which is collision free, allows to maintain balance, and maximizes EFORT.
-We choose this configuration among a set of possible candidates, rapidly computed thanks to a contact generator, also proposed in~\cite{Tonneau2014}.
-We recall the overall idea here.
-We consider each limb of the robot as a manipulator arm attached to the root.
-We store a database of configurations for each manipulator.
-The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations.
-The jacobian matrix $\mathbf{J}$ of each configuration is also stored.
-Figure~\ref{fig:contact_gen} -- 2 presents a few configurations for the right arm of the robot.
+$h_{EFORT}$ and $h_{vel}$ will favor contacts that allow important efforts or fast modifications in the velocity.
+$EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
+In other less demanding cases, manipulability can also be considered regarding its original purpose, that means to avoid singularties.
+To do so, we can consider the manipulability measure $w$, also given by \citeauthor{Yoshikawa1984}:
 
-%To summarize: given a feasible trajectory for the root of the robot $R$, we are able to generate a discrete sequence of contact configurations  describing a feasible motion for $R$, between a start an goal configurations.
-The contact configurations are chosen to be statically balanced and task efficient, which means that they can generate a force  contributing to the motion of the root along the trajectory.
+\begin{equation*} \label{ellipsoid}
+w(\mathbf{q}) = \sqrt{\mathbf{J}\mathbf{J}^T}
+\end{equation*}
+$w$ measures the ``distance" between a given configuration and a singular configuration. When $w$ is equal to 0, the configuration is in a singular state;
+the greater $w$ is, the further away the configuration is from singularity.
+We then derive:
+\begin{equation}
+h_{w}(\mathbf{q}, \mathbf{m}) = \sqrt{\mathbf{J}\mathbf{J}^T} ( \mu_0 \mathbf{n}^T \mathbf{m})
+\end{equation}
+
+$h_{EFORT}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling based approach, the jacobian an inverse products
+can be precomputed offline), that allow the planner to select the more relevant
+candidates regarding the local motion task along the guide path.
+This result in faster computation times, and a reduced number of contact switches overall.

### 2015-12-04 body/equilibrium.tex
robust equilibrium

@@
 \subsection{A heuristic for robust quasi static balance}
-\noindent\textbf{EFORT criterion:} If only relying on the random sampling to select new contacts, the planner produces inefficient postures.
-The resulting contact sequence is then poorly efficient and unnatural.
-Moreover, the limbs are not well configured and are not able to efficiently follow the general movement: contacts break frequently.
+The planner is designed so that any contact configuration it outputs is quasi statically balanced.
+Balance is of course critical in multipedal locomotion. For this reason we are interested in a robust 
+criterion, that ensures that the robot remains balanced in a real world application, regardless of the errors
+that can result from peception and control uncertainties.
 
-When creating additional contacts, we therefore propose to select particular configurations that allow to exert a force compatible with the direction of motion.
-This task efficiency is measured  with the Extended FORce Transmission ratio (EFORT)~\cite{Tonneau2014}.
-%
-The measure of EFORT is given by
-\begin{equation}
-\alpha_{EFORT}(\mathbf{q}^k, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \nu_0 \mathbf{n}^T \mathbf{m})
+We first give a classic linear program (LP) that asserts that a contact configuration is statically balanced.
+From this formulation we derive a new LP that returns a quantification of the robustness of the configuration regarding balance.
+From this value we can in turn either define a heuristic to choose the more robust candidate, or set a treshold on the required balance robustness
+for any configuration. While the LP we propose is new, it is based on an analysis of the problem we propose in \cite{Prete2016}, where the interested reader
+can find more details.
+
+\subsubsection{Conditions for static balance.}
+We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:
+\begin{itemize}
+\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
+\item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
+\item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
+\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
+\item $\mu$ is the friction coefficient;
+\item for the i-th contact $1 \leq i \leq e$:
+	\begin{itemize}
+	\item $\mathbf{p_i}$ is the contact position;
+	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
+	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
+	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
+	\end{itemize}
+\end{itemize}
+
+According to Coulomb's law, the robot is statically balanced if all contact forces lie in the friction cone defined by the surface.
+As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
+\begin{equation*}
+\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{t}_{i1} & \mathbf{n}_{i} -\mu \mathbf{t}_{i1} & \mathbf{n}_{i} + \mu \mathbf{t}_{i2} & \mathbf{n}_{i} - \mu \mathbf{t}_{i2}}^T
+\end{equation*}
+
+Any force belonging to the linearized cone
+can thus be express as a linear combination of its four generating rays.
+Therefore a constraint on the $\mathbf{f}_i$ is given by the relation:
+
+\begin{equation*}
+\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{+4} , \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i \qquad 
+\end{equation*}
+
+we can stack the constraints to obtain:
+\begin{equation}\label{eq:gen}
+\exists \bm{\beta} \in \mathbb{R}^{+4e} , \mathbf{f} = \mathbf{V} \bm{\beta} \qquad 
 \end{equation}
-where $\mathbf{J}$ is the Jacobian matrix of the limb $R^k$ in configuration $\mathbf{q}^k$; $\nu_0$ is the friction coefficient of the contact surface; $\mathbf{n}$ is the normal of the contact surface; and $\mathbf{m}$ is the direction opposite to the motion,
-given by the 3D vector connecting $\mathbf{q}_{i}^0$ and $\mathbf{q}_{i+1}^0$.
-%
-The first part of the equation measures the ratio between the joint  torques and the resulting force applied along $\mathbf{m}$.
-The second part quantifies the odds of slipping while applying a force along $\mathbf{m}$.
+where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = [\mathbf{f}_0,...,\mathbf{f}_e]^T$.
 
-\noindent\textbf{Optimization at creation:} In practice, a database of configurations is stored for each limb, which can be considered as manipulator arms.
-The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations (and additionally storing $\mathbf{J}$).
-%Fig.~\ref{fig:contact_gen}-2 presents a few configurations for the right arm of the robot.  
-Upon request, the octree returns a set of configurations close to contact (Fig.~\ref{fig:contact_gen}-3).
-These candidates are sorted based on their task efficiency, given by $\alpha_{EFORT}$.
-The first candidate in this list satisfying the balance criterion and is collision free is selected and projected on the contact surface using our inverse kinematics solver.
+Additional constraints on the forces result from Newton-Euler equations:
 
-%Therefore to generate a contact, we want to find a configuration which is collision free, allows to maintain balance, and maximizes EFORT.
-We choose this configuration among a set of possible candidates, rapidly computed thanks to a contact generator, also proposed in~\cite{Tonneau2014}.
-We recall the overall idea here.
-We consider each limb of the robot as a manipulator arm attached to the root.
-We store a database of configurations for each manipulator.
-The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations.
-The jacobian matrix $\mathbf{J}$ of each configuration is also stored.
-Figure~\ref{fig:contact_gen} -- 2 presents a few configurations for the right arm of the robot.
+\begin{align}
+\underbrace{\mat{m (\ddot{\mathbf{c}} - \mathbf{g})  \\ m \mathbf{c} \times (\ddot{\mathbf{c}} - \mathbf{g}) + \dot{\mathbf{L}}}}_\mathbf{w}
+= 
+\underbrace{
+\mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
+\hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
+}_\mathbf{G}
+\bm{\beta},
+\end{align}
+where $\mathbf{w}\in \Rv{6}$ is the so-called gravito-inertial wrench (GIW) \citep{qiu:dhm:2011, Caron2015} and $\hat{\mathbf{p}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{p}$.
 
-%To summarize: given a feasible trajectory for the root of the robot $R$, we are able to generate a discrete sequence of contact configurations  describing a feasible motion for $R$, between a start an goal configurations.
-The contact configurations are chosen to be statically balanced and task efficient, which means that they can generate a force  contributing to the motion of the root along the trajectory.
+Static balance assumes zero acceleration thus we can write:
+\begin{align} \label{eq:new_eul}
+\mathbf{w}_0 = 
+\underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
+\underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
+\end{align}
+The wrench $\mathbf{w}_0$ is independent of $c^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite \eqref{eq:new_eul} as:
+\begin{align} \label{eq:new_eul2d}
+\mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
+\end{align}
+
+If there exists a $\bm{\beta}^*$ satisfying equations \ref{eq:gen} and \ref{eq:new_eul2d}, it means that the configuration is statically balanced.
+The problem can then be formulated as an LP:
+
+\begin{equation} \label{eq:lin_prog} \begin{aligned}
+\find \quad & \bm{\beta} \in \Rv{+4e} \\
+\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
+& \bm{\beta} \ge 0 \\
+\end{aligned} \end{equation}
+
+\subsubsection{Formulation of a robust LP:}
+Let $b_0 \in \mathbb{R}$ a scalar value. We now define the following LP:
+
+\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
+\find \quad & \bm{\beta} \in \Rv{+4e}, b_0 \in \Rv{} \\
+\minimize  \quad & -b_0 \\
+\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
+& \bm{\beta} \ge b_0 \\
+\end{aligned} \end{equation}
+
+We observe that if $b_0$ is positive, LP~\ref{eq:lin_prog} admits a solution.
+The higher $b_0$, the further away from the cone limits admissible forces can be found.
+If $b_0$ is negative, the configuration is not balanced, and the value $b_0$ indicates "how far" from static balance
+the configuration is. $b_0$ can thus be seen as a measure of robustness.
+
+In our implementation, we solve the dual of LP~\ref{eq:lin_prog_rob}.
+after a trivial elimination of Lagrange multipliers the dual problem has lower dimensionality:
+\begin{equation} \label{eq:dual} \begin{aligned}
+\find \quad & \bm{\nu} \in \Rv{6}\\
+\minimize  \quad & (\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
+\st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
+& \mathbf{1}^T \mathbf{G}^T \bm{\nu} = 1 \\
+\end{aligned} \end{equation}
+
+Indeed, from Slater's condition, we know that the optimal values of a LP and its dual are equal. Thus the optimal value of LP~\ref{eq:dual} is indeed the optimal $b_0$.
+
+In conclusion, LP~\ref{eq:dual} provides a method to assert rapidly quasi static balance, and returns a robustness value, that can be used as a heuristic for robust static balance.
+Another option to enforce robustness is to set a minimum acceptable value for $b_0$; this avoids ponderating balance robustness with other heuristics,
+but as a result this conservative approach is not complete.

### 2015-12-04 body/intro.tex
robust equilibrium

@@
 This extension completes this work with an argumented discussion on the implementation of this theory to real world robots and problems.
-In particular, we detail the robust balance criteria designed to ensure the equilibrium of the robot and provide details on the hpp implementation
-of our contact generation algorithm. Furthermore, our experiments now concern real robots, namely HRP-2 and Hyq. TODO Appendix describes the generation process
+In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot. We also provide details on the hpp implementation
+of our contact generation algorithm. Furthermore, our experiments are now applied to real robot models, namely HRP-2 and Hyq. TODO Appendix describes the generation process
 for the robot.

### 2015-12-04 body/appendix_algo.tex
corrections andrea

@@
 as a new root placement. $g$ returns a new contact configuration, if it suceeded
-in computing a balance configuration with only one contact switch occuring.
+in computing a configuration with only one contact switch occuring.
 If $g$ failed in achieving this, the method \textsc{RepositionContacts} is called.
-It repositions one end effector (either a free limb, or the oldest active contact) towards a new balanced contact position if possible.
+It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
 This repositionning allows to increase the odds that the contact can be maintained at the next step.
@@
 		\For {\textbf{each} Limb $k$ in $previous.freeLimbs$}
-			\State /*Generate a balanced contact*/
 			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
@@
 \textsc{GenerateContact}$(\mathbf{q}^0,k)$ is a call to the contact generator presented in Section~\ref{sec:single_contact}.
-It generates a balanced contact configuration, and assigns the corresponding configuration to $k$.
+It generates a contact configuration in static equilibrium, and assigns the corresponding configuration to $k$.
 If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.

### 2015-12-04 body/contact.tex
corrections andrea

@@
 and root location.
-Then, on this method we build an iterative algorithm to generate a discrete sequence of balanced contact configurations along a truly feasible path.
+Then, on this method we build an iterative algorithm to generate a discrete sequence of contact configurations in equilibrium, along a truly feasible path.
 
-In what follows a \textit{valid} contact configuration is understood as being collision free, respecting joint limits, and allowing to maintain the quasi static balance
-of the robot. The first two tests are trivial; our criterion to assert efficiently the quasi static balance of the system
+In what follows a \textit{valid} contact configuration is understood as being collision free, respecting joint limits, and allowing to maintain the static equilibrium
+of the robot. The first two tests are trivial; our criterion to assert efficiently the static equilibrium of the system
 is described in Section TODO.
@@
 
-At each step, we consider a new root placement $\mathbf{q}^0_{i}$ and try to generate a balanced contact configuration that preserves as much as possible
-the previous contacts.
+At each step, we consider a new root placement $\mathbf{q}^0_{i}$ and try to generate a contact configuration that preserves as much as possible
+the previous contacts, and is in static equilibrium.
 
@@
 
-Figure~\ref{fig:break_contact} illustrates the contact persistence method.
+Figure~\ref{fig:break_contact} illustrates the contact persistence strategy.
 If possible, a limb in contact at time $i-1$ remains in contact at $i$.
@@
 \item Only one contact creation can happen between two consecutive states. This increases the odds that the interpolation between the two states will be feasible; 
-\item A contact is validated if and only if the resulting configuration is statically balanced, no matter what user-defined heuristics are used; 
+\item A contact is validated if and only if the resulting configuration is in static equilibrium, no matter what user-defined heuristics are used; 
 \item We use a FIFO approach:  we always try to first create a contact with the limb that has been free for the longest time. If the contact creation
@@
 These three rules provide a deterministic manner to generate contacts along the discretized guide path.
-FIGURE. For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a balanced configuration, because
+FIGURE. For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
 of the deterministic approach used to break the combinatorial.

### 2015-12-04 body/equilibrium.tex
corrections andrea

@@
 
-\subsection{A heuristic for robust quasi static balance}
-The planner is designed so that any contact configuration it outputs is quasi statically balanced.
-Balance is of course critical in multipedal locomotion. For this reason we are interested in a robust 
+\subsection{A heuristic for robust static equilibrium}
+The planner is designed so that any contact configuration it outputs is in static equilibrium.
+Equilibrium is of course critical in multipedal locomotion. For this reason we are interested in a robust 
 criterion, that ensures that the robot remains balanced in a real world application, regardless of the errors
-that can result from peception and control uncertainties.
+that can result from perception and control uncertainties.
 
-We first give a classic linear program (LP) that asserts that a contact configuration is statically balanced.
-From this formulation we derive a new LP that returns a quantification of the robustness of the configuration regarding balance.
-From this value we can in turn either define a heuristic to choose the more robust candidate, or set a treshold on the required balance robustness
-for any configuration. While the LP we propose is new, it is based on an analysis of the problem we propose in \cite{Prete2016}, where the interested reader
+We first give a linear program (LP) that asserts that a contact configuration is in static equilibrium.
+From this formulation we derive a new LP that quantifies the robustness of the configuration regarding equilibrium.
+From this value we can in turn either define a heuristic to choose the most robust candidate, or set a treshold on the required robustness
+for any configuration. While the LP we propose is original, it is based on an analysis of the problem we proposed in \cite{Prete2016}, where the interested reader
 can find more details.
@@
 
-According to Coulomb's law, the robot is statically balanced if all contact forces lie in the friction cone defined by the surface.
+According to Coulomb's law, the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
 As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
@@
 Any force belonging to the linearized cone
-can thus be express as a linear combination of its four generating rays.
-Therefore a constraint on the $\mathbf{f}_i$ is given by the relation:
+can thus be express as a convex combination of its four generating rays.
+Therefore we can express the non slipping constraint on $\mathbf{f}_i$ as:
 
 \begin{equation*}
-\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{+4} , \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i \qquad 
+\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} , \bm{\beta}_i \ge 0 \wedge \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i \qquad 
 \end{equation*}
@@
 \begin{equation}\label{eq:gen}
-\exists \bm{\beta} \in \mathbb{R}^{+4e} , \mathbf{f} = \mathbf{V} \bm{\beta} \qquad 
+\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \wedge \mathbf{f} = \mathbf{V} \bm{\beta} \qquad 
 \end{equation}
@@
 
-Additional constraints on the forces result from Newton-Euler equations:
+From the Newton-Euler equations, and assuming a null acceleration, we also constrain the sum of contact forces to compensate the gravity at the center of mass:
 
-\begin{align}
-\underbrace{\mat{m (\ddot{\mathbf{c}} - \mathbf{g})  \\ m \mathbf{c} \times (\ddot{\mathbf{c}} - \mathbf{g}) + \dot{\mathbf{L}}}}_\mathbf{w}
-= 
+%~ \begin{align}
+%~ \underbrace{\mat{m (\ddot{\mathbf{c}} - \mathbf{g})  \\ m \mathbf{c} \times (\ddot{\mathbf{c}} - \mathbf{g}) + \dot{\mathbf{L}}}}_\mathbf{w}
+%~ = 
+%~ \underbrace{
+%~ \mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
+%~ \hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
+%~ }_\mathbf{G}
+%~ \bm{\beta},
+%~ \end{align}
+%~ where $\mathbf{w}\in \Rv{6}$ is the so-called gravito-inertial wrench (GIW) \citep{qiu:dhm:2011, Caron2015} and $\hat{\mathbf{p}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{p}$.
+
+%~ Static balance assumes zero acceleration thus we can write:
+\begin{align} \label{eq:new_eul}
 \underbrace{
@@
 \hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
-}_\mathbf{G}
-\bm{\beta},
-\end{align}
-where $\mathbf{w}\in \Rv{6}$ is the so-called gravito-inertial wrench (GIW) \citep{qiu:dhm:2011, Caron2015} and $\hat{\mathbf{p}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{p}$.
-
-Static balance assumes zero acceleration thus we can write:
-\begin{align} \label{eq:new_eul}
-\mathbf{w}_0 = 
+}_\mathbf{G} \bm{\beta}, = 
 \underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
@@
 \end{align}
-The wrench $\mathbf{w}_0$ is independent of $c^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite \eqref{eq:new_eul} as:
+where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.
+The right side of \ref{eq:new_eul} is independent from $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as:
 \begin{align} \label{eq:new_eul2d}
@@
 \begin{equation} \label{eq:lin_prog} \begin{aligned}
-\find \quad & \bm{\beta} \in \Rv{+4e} \\
+\find \quad & \bm{\beta} \in \Rv{4e} \\
 \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
@@
 \subsubsection{Formulation of a robust LP:}
-Let $b_0 \in \mathbb{R}$ a scalar value. We now define the following LP:
+Let $b_0 \in \mathbb{R}$ be a scalar value. We now define the following LP:
 
 \begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
-\find \quad & \bm{\beta} \in \Rv{+4e}, b_0 \in \Rv{} \\
+\find \quad & \bm{\beta} \in \Rv{4e}, b_0 \in \Rv{} \\
 \minimize  \quad & -b_0 \\
@@
 
-We observe that if $b_0$ is positive, LP~\ref{eq:lin_prog} admits a solution.
-The higher $b_0$, the further away from the cone limits admissible forces can be found.
-If $b_0$ is negative, the configuration is not balanced, and the value $b_0$ indicates "how far" from static balance
+We observe that if $b_0$ is positive, LP~\ref{eq:lin_prog} admits a solution, and $b_0$ is proportional to the minimum distance of the equilibrium contact forces to the boundaries of the friction cones.
+If $b_0$ is negative, the configuration is not in static equilibrium, and the value $b_0$ indicates "how far" from equilibrium
 the configuration is. $b_0$ can thus be seen as a measure of robustness.
@@
 In our implementation, we solve the dual of LP~\ref{eq:lin_prog_rob}.
-after a trivial elimination of Lagrange multipliers the dual problem has lower dimensionality:
+after a trivial elimination of the Lagrange multipliers associated to the inequality constraints, the dual problem has a lower dimensionality:
 \begin{equation} \label{eq:dual} \begin{aligned}
 \find \quad & \bm{\nu} \in \Rv{6}\\
-\minimize  \quad & (\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
+\maximize  \quad & -(\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
 \st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
@@
 
-Indeed, from Slater's condition, we know that the optimal values of a LP and its dual are equal. Thus the optimal value of LP~\ref{eq:dual} is indeed the optimal $b_0$.
+Indeed, from Slater's condition \citep{Boyd:2004:CO:993483}, we know that the optimal values of an LP and its dual are equal. Thus the optimal value of LP~\ref{eq:dual} is indeed the optimal $b_0$.
 
-In conclusion, LP~\ref{eq:dual} provides a method to assert rapidly quasi static balance, and returns a robustness value, that can be used as a heuristic for robust static balance.
-Another option to enforce robustness is to set a minimum acceptable value for $b_0$; this avoids ponderating balance robustness with other heuristics,
-but as a result this conservative approach is not complete.
+In conclusion, LP~\ref{eq:dual} provides a method to assert rapidly static equilibrium, and returns a robustness value.
+This robustness criterion can be immediately transformed into a heuristic, and eventually be ponderated with others.
+Another option to enforce robustness is to set a minimum acceptable value for $b_0$, but as a result this conservative approach is not complete.

### 2015-12-04 body/intro.tex
corrections andrea

@@
 We consider the problem of planning the acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment.
-Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to support the balance of the system.
+Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to support the equilibrium of the system.
 
@@
 The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper.
-The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic, balanced contact configurations along the trajectory (A third non trivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
+The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third non trivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories.
@@
 Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the relevance of the guide trajectory,  which is not evaluated \textit{a priori}.
-\citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body, balanced contact configurations.
+\citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body contact configurations in static equilibrium.
 Randomly sampled configurations are projected into the contact submanifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide trajectory in the examples presented).
@@
 for proximity queries. At runtime, this framework allows to obtain extremly rapidly a set of candidate configurations for contact.
-These candidates are sorted using a criterion to evaluate the quasi static balance of the configuration, and ranked according to user defined heuristics.
-In our implementation we use two such heuristics, used in all our experiments: a robustness criterion regarding the balance of a configuration, and a manipulability based heuristic,
+These candidates are sorted using a criterion to evaluate the quasi static equilibrium of the configuration, and ranked according to user defined heuristics.
+In our implementation we use two such heuristics, used in all our experiments: a robustness criterion regarding the static equilibrium of a configuration, and a manipulability based heuristic,
 the Extended FORce Transmission ratio (EFORT).
@@
 
-The present paper is an extension of a conference paper to appear in the proceedings of the ISRR '15 conference TODO CITE.
+The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep(tonneauisrr15).
 The conference paper focuses on the theoritical formulation of the problem, and presents results obtained with virtual avatars.

### 2015-12-04 body/results.tex
corrections andrea

 \section{Results}
\label{sec:results}
In this Section we present some of the results obtained with our planner. The complete sequences computed are shown in the companion video, TODO.
Specifically, we demonstrate the planner for two really different robots, in a large vartiety of environments: the humanoid HRP-2 and the quadriped Hyq.
For each scenario we indicate the heuristics chosen. We also provide a performance analysis, that shows that the planner is compatible with interactive applications,
and present the success rates obtained in each scenario.
We also demonstrate the interest of our robustness criterion in the different computed poses.
Finally, a last example suggests possible applications to dextrous manipulation.

In our ISRR conference paper~\citep{tonneauisrr15}, additional results are demonstrated with various virtual avatars (Figure~\ref{fig:robots_old}).
In this extension we choose to focus on results obtained with actual robots. We invite the interested reader to watch the ISRR video\endnote{https://www.youtube.com/watch?v=LmLAHgGQJGA}, and 
to refer to the previous paper for a discussion on these results.

\begin{figure}[t]
\centering
  \begin{overpic}[width=1\linewidth]{figures/robots_old}
		%~ \put (5,58) {1)} 
		%~ \put (37,58) {2)} 
		%~ \put (68,58) {3.a)} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	\end{overpic}
\caption{Virtual avatars and associated volumes used in our conference paper: in red $W^0_{s}$; in green the range of motion of each limb.}
		   \label{fig:robots_old}
\end{figure}

\subsection{Description of the scenarios}
In all the scenarios considered, the formulation of the problem is always the same:
a start and goal root placements are provided as an input of the scenario.
The framework computes the initial balanced contact configuration, and outputs a sequence of statically balanced contact configurations connecting it to the goal.
A companion video available at TODO \url{http://youtu.be/LmLAHgGQJGA} (anonymous link) displays the complete contact sequence obtained in all these scenarios. The video only renders the contact configurations. Again, the interpolation between contact configurations is out of the scope of this paper~\citep{Carpentier2016}.

% \subsubsection*{Truck egress (Figure~\ref{res_truck_pres} and Figure~\ref{res_truck_bd}) -- Humanoid and insectoid robots.}
%\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}
\noindent\textbf{Truck egress (humanoid and insectoid):}
The robot must leave a truck the doors of which are blocked: it has to crawl through the front window.
Figure~\ref{res_truck_bd} presents the sequence of contacts obtained for both robots: RB-PRM can find solutions in highly cluttered environments with narrow passages.

\begin{figure}[h!]
  \centering
  \includegraphics[width=1\linewidth]{figures/climballlow}
  \caption{
           The computed contact sequences for the truck egress scenario. Only selected postures are shown for the insect.}
		   \label{res_truck_bd}
\end{figure}

% \subsubsection*{Climbing (Figure~\ref{res_climb}) -- Humanoid and insectoid robots.}
%\subsubsection*{Climbing -- Humanoid and insectoid robots.}
\noindent\textbf{Climbing (humanoid and insectoid):}
The robot has to climb on a wall with several grasps disposed along it. In this scenario, we give stronger conditions for the sampled root placements: we require that more than one range of motion $W^k$ collide with obstacles of the environment. Fig.~\ref{res_climb} presents the contact sequence obtained for the 
humanoid robot.

\begin{figure}[h!]
  \centering
  %\includegraphics[width=0.6\linewidth]{figures/climb}
  \includegraphics[width=1\linewidth]{figures/climb_line}
  \caption{The computed sequence for the climbing scenario.}
		   \label{res_climb}
\end{figure}

%~ \subsection{Implementation details}
%~ The test application was developed using the C++ language.
%~ The volumes describing the Range Of Motions of the robots $A_{ROM}$ are computed using the Matlab mpt toolbox~\footnote{http://people.ee.ethz.ch/\~mpt/3/}.
%~ They are exported to the obj format using a custom Matlab script.
%~ 
%~ The robot is described using the urdf file format~\footnote{http://wiki.ros.org/urdf}, used as a standard by the popular ROS platform.
%~ The scenarios where described using a custom text file format.
%~ Environments are described in the obj format. Rendering is achieved using Blender~\footnote{http://www.blender.org}.
%~ The runs were performed on a laptop with an Intel Core i7-2760QM 2.40GHz processor and 4 GB of memory. The application is not multi-threaded.

\noindent\textbf{Manipulation of a pen (3-finger hand):}
This scenario is proposed to illustrate the genericity of our approach: we consider a manipulation task for a robotic hand and use
our contact planner to compute a guide trajectory for the fingers, considered as effectors (Figure \ref{fig:penrot}).
Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences, this in less than 5 seconds.

\begin{figure}[t]
\centering
  \begin{overpic}[width=1\linewidth]{figures/penrot}
	\end{overpic}
\caption{Contact sequence found for a pen manipulation in a zero gravity environment.}
		   \label{fig:penrot}
\end{figure}

%\subsubsection*{Other scenarios}
\noindent\textbf{Other scenarios (humanoids): }
%Two additional scenarios were considered for the humanoid robot. A crouching scenario, and a standing up scenario (already presented in Figure~\ref{fig:framework} -- right) -- Humanoid robot). 
The standing-up scenario (already presented in Fig.~\ref{fig:framework}) is a setup taken from \cite{DBLP:conf/iser/EscandeKMG08}: it corresponds to a long narrow passage in the configuration space.
In the crouching scenario, demonstrated in the companion video, the character automatically goes from a standing to a crouching position to crawl under an obstacle.

\subsection{Parametrization of the reachability condition} \label{sec:stat1}

To find the appropriate $C_{reach}^s$ in which to sample the guide trajectory,
we computed the rejection rate for various values of $s$ for each robot in the most cluttered truck scenario.
For a given value of $s$, $10^6$ root positions and orientations are computed
in $C_{reach}^s$. In each case we try to generate a collision free contact configuration, with a database
comprising $N=10^5$ sample configurations for each limb. The rejection rate
is the ratio between the number of failures  and the number of trials. From Figure~\ref{fig:rejection} $s$ is empirically chosen
as the smallest value for which the rejection rate is minimal.
For the humanoid, we thus chose $s=2.2$, and for the insect, $s=2.8$.

\begin{figure}[t]
\centering
  \begin{overpic}[width=0.6\linewidth]{figures/rejectionnotext}
		 \put (1.1,24){\rotatebox{90}{\tiny{Rejection rate}}}
		 \put (4.8,52) {\tiny{20} }
		 \put (4.8,48) {\tiny{18} }
		 \put (4.8,43) {\tiny{16} }
		 \put (4.8,38) {\tiny{14} }
		 \put (4.8,34) {\tiny{12} }
		 \put (4.8,29) {\tiny{10} }
		 \put (4.8,24.5) {\tiny{8} }
		 \put (4.8,20) {\tiny{6} }
		 \put (4.8,15) {\tiny{4} }
		 \put (4.8,10) {\tiny{2} }
		 \put (11.5,5) {\tiny{1} }
		 \put (24.5,5) {\tiny{1.5} }
		 \put (39,5) {\tiny{2} }
		 \put (52,5) {\tiny{2.5} }
		 \put (45,3) {\tiny{s} }
		 \put (66,5) {\tiny{3} }
		 \put (79,5) {\tiny{3.5} }
		%~ \put (37,58) {2)} 
		%~ \put (68,58) {3.a)} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	\end{overpic}
\caption{Truck scenario rejection rates(\%) for the humanoid(orange) and insectoid(blue), given $s$.}
		   \label{fig:rejection}
\end{figure}

\subsection{Performance}\label{sec:stat2}
The number of samples used for generating the contacts of each limb is 10000.
Table~\ref{tab:requestime} presents the average time (seconds) spent in the phases of the planner, for 
each phase and each scenario, and the number of contact phases of the sequence.

\begin{table}[b]
\centering
\begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | c}
  &  Generate RB-PRM (offline) & Generating the contact sequence & Number of contact states\\
 \hline
   Truck egress (humanoid) & $ 73 $ & 15 & 10 \\
   Truck egress (insectoid) & $70$ & 23 & 48\\
   Climbing (humanoid)& $25$ &  5 & 15\\
   Climbing (insectoid)  & $ 21 $ & 27 & 51\\
   Crouching (humanoid)& $5$ & 6 & 22 \\
 \end{tabular}
\caption{\textbf{Average time} (in seconds) spent in RB-PRM generation, and the online generation of the contact sequence.}
\label{tab:requestime}
\quad
 \end{table}

 We observe that many contacts are required for the insect, which can be explained by its restricted range of motion.
 The time spent generating the navigation graph is about one minute. The time spent in generating the graph of the climbing scenario,
 despite the relatively open environment, is explained by the additional restrictions imposed on the reachability condition.
 The difficulty to find a balanced configuration essentially influences the time spent generating the contacts.

The number of contacts in the sequence gives a rough estimation of its duration in seconds. Except for the robots crawling out of the truck, all the contact generation are real-time. Additionaly to the quality of the generated trajectories shown in the video, these computation times are a major practical achievement.

%~ \subsection{Early results: manipulating scenario in zero gravity}
%~ In this scenario, we suggest an extension of our framework to manipulation.
%~ We consider the task of rotating a pen using a robotic hand.
%~ A start and goal configurations for the pen are expressed in the world coordinate system.
%~ The problem is automatically reformulated into an equivalent motion planning one for the robotic hand, by expressing its start
%~ and goal configurations in the coordinate system of the pen. In other words, for the planning the pen is considered as a static object, and a feasible motion for the hand
%~ is computed along the pen. Figure {penrot} presents the results obtained with this approach.
%~ Additional work is required to generate feasible manipulation motions.

### 2015-12-06 body/results.tex
half results figures + full text

@@
 	\end{overpic}
-\caption{Virtual avatars and associated volumes used in our conference paper: in red $W^0_{s}$; in green the range of motion of each limb.}
+\caption{Virtual avatars in various scenarios demonstrated in our conference paper.}
 		   \label{fig:robots_old}
@@
 a start and goal root placements are provided as an input of the scenario.
-The framework computes the initial balanced contact configuration, and outputs a sequence of statically balanced contact configurations connecting it to the goal.
+The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
+In each scenario we detail the parameters chosen: the heuristics, and the constraints on the reachable workspaces (for instance in all the scenarios,
+the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 
 A companion video available at TODO \url{http://youtu.be/LmLAHgGQJGA} (anonymous link) displays the complete contact sequence obtained in all these scenarios.
-The video only renders the contact configurations.
-Again, the interpolation between contact configurations is out of the scope of this paper~\citep{Carpentier2016}.
 
@@
 %\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}
-\noindent\textbf{Truck egress (humanoid and insectoid):}
-The robot must leave a truck the doors of which are blocked: it has to crawl through the front window.
-Figure~\ref{res_truck_bd} presents the sequence of contacts obtained for both robots: RB-PRM can find solutions in highly cluttered environments with narrow passages.
+\subsubsection{HRP-2 -- Steep staircase (Figure~\ref{fig:standing}):}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=0.5\linewidth]{figures/stair}
+  \caption{
+           HRP-2 in the steep stair climbing scenario. }
+		   \label{fig:stair_robust}
+\end{figure*}
+
+The goal is to cross a set of stairs of a height of 20 cm. This height requires HRP-2 to use a ramp to be able to perform the task.
+
+\noindent\textbf{Contacts involved:} Both feet and right arm.
+
+\noindent\textbf{Heuristics:} The extended manipulability $h_w$ is chosen for the feet; $h_{EFORT}$ is chosen for the right arm.
+Regarding equilibrium, the video demonstrates two sequences computed for two different treshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
+
+\noindent\textbf{Observations:}
+This scenario illustrates best the importance of the equilibrium robustness criterion.
+With a robust approach, more states are required to reach the last step (15 against 13 in average).
+However, when the last step is reached by both feet, in the non robust case, the contacts are extremely close to 
+the cone limits, and we believe this can be seen visually (Figure~\ref{fig:stair_comp}).
+
+The geometry of the environment is easily addressed by our planner, and the contact planning is several times faster than real time in this scenario.
+
+Again, the interpolation motion between the contact stances is out of the scope of this paper.
+However it should be noted that the computed plan in this scenario has been executed successfully on the robot~\citep{Carpentier2016}.
 
@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/climballlow}
+  %~ \includegraphics[width=0.6\linewidth]{figures/stair_robust}
+  \begin{overpic}[width=0.5\linewidth]{figures/stair_robust}
+		\put (17,5) {\small{\color{red}$b_0 = 0.23$}} 
+		\put (79,5) {\small{\color{green}$b_0 = 6.16$}} 
+		%~ \put (68,58) {3.a)} 
+		%~ \put (5,27) {3.b)} 
+		%~ \put (37,27) {4.a)} 
+		%~ \put (68,27) {4.b)} 
+	\end{overpic}
   \caption{
-           The computed contact sequences for the truck egress scenario. Only selected postures are shown for the insect.}
-		   \label{res_truck_bd}
+           Evaluation of the robustness $b_0$ of two contact configurations. Although in equilibrium, the left configuration is on the verge of slipping.}
+		   \label{fig:stair_comp}
 \end{figure}
 
-% \subsubsection*{Climbing (Figure~\ref{res_climb}) -- Humanoid and insectoid robots.}
-%\subsubsection*{Climbing -- Humanoid and insectoid robots.}
-\noindent\textbf{Climbing (humanoid and insectoid):}
-The robot has to climb on a wall with several grasps disposed along it.
-In this scenario, we give stronger conditions for the sampled root placements: we require that more than one range of motion $W^k$ collide with obstacles of the environment.
-Fig.~\ref{res_climb} presents the contact sequence obtained for the
-humanoid robot.
+\subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing}):}
+From a bent configuration, a standing up motion is computed in a cluttered environment.
+The resulting motion involves using a wall as support, and crossing a 25 cm high stair.
 
-\begin{figure}[h!]
+\begin{figure*}
   \centering
-  %\includegraphics[width=0.6\linewidth]{figures/climb}
-  \includegraphics[width=1\linewidth]{figures/climb_line}
-  \caption{The computed sequence for the climbing scenario.}
-		   \label{res_climb}
+  \includegraphics[width=0.5\linewidth]{figures/standing}
+  \caption{
+           HRP-2 in the standing scenario. }
+		   \label{fig:standing}
+\end{figure*}
+
+\noindent\textbf{Contacts involved:} All (both feet and hands).
+
+\noindent\textbf{Heuristics:} $h_w$ is used for the feet, $h_{EFORT}$ is used for the hands.
+
+\noindent\textbf{Observations:} The scenario illustrates well the acyclic aspect of the planning. For instance, in the 4 first frames of Figure~\ref{fig:standing}, we can see that the right foot
+is moved twice, with the left foot in between, before the configuration allows HRP-2 to move its hand.
+Because the contacts are tried in a FIFO manner, the fact that the output contact sequence is acyclic shows that a cyclic approach (with a finite state machine for instance) is not sufficient
+for the computed trajectory. The reason for this is not reachability, but equilibrium. The planning is slower than for the stair scenario (because the contact generation fails more),
+though it remains compatible with interactive performances.
+
+\subsubsection{HRP-2 -- Truck egress (Figure~\ref{fig:TODO}):}
+HRP-2 stands half sitted in a truck, and is required to exit through the front window.
+
+%~ \begin{figure*}
+  %~ \centering
+  %~ \includegraphics[width=0.5\linewidth]{figures/standing}
+  %~ \caption{
+           %~ HRP-2 in the standing scenario. }
+		   %~ \label{fig:standing}
+%~ \end{figure*}
+
+\noindent\textbf{Contacts involved:} All (both feet and hands).
+
+\noindent\textbf{Heuristics:} $h_w$ is used for the feet, $h_{EFORT}$ is used for the hands.
+
+\noindent\textbf{Observations:} This scenario demonstrates the ability of the planner to compute a path in an extremely constrained environment. In this case,
+the scenario is not always real time (TODO or is it ?): while the planner is often able to compute a guide path rapidly, in the worst case up to one minute can be spent
+in the planning.
+
+\subsubsection{HyQ -- Darpa style rubble (Figure~\ref{fig:darpa})}
+The quadruped robot is given the task to cross a rubble composed of bricks rotated at different angles and directions.
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=0.5\linewidth]{figures/darpa}
+  \caption{
+           Robust crossing of rubbles by HyQ ($b_0 > 20$). }
+		   \label{fig:darpa}
+\end{figure*}
+
+\noindent\textbf{Contacts involved:} All (the 4 legs).
+
+\noindent\textbf{Heuristics:} $h_w$ is used for the all legs. The robustness treshold $b_0$ is set to $20$.
+
+\noindent\textbf{Observations:} In this context, setting up a really important minimum value for $b_0$ is possible due to the high
+stability of the HyQ robot, and results in more contact switches, in exchange for safety.
+The path planning time is higher than for the previous HRP-2 robots due to the constraint that the 4 reachable workspace of all legs must
+collide with the environment at all times. Again, the computed time remains however interactive.
+
+\subsubsection{HyQ -- Obstacle race:}
+In this long scene, HyQ is first required to cross a 55 cm large hole; then, to cross a narrow "bridge", only large of 25 cm,
+while maintaining equilibrium
+
+%~ \begin{figure*}
+  %~ \centering
+  %~ \includegraphics[width=0.5\linewidth]{figures/stair}
+  %~ \caption{
+           %~ HRP-2 in the steep stair climbing scenario. }
+		   %~ \label{fig:stair_robust}
+%~ \end{figure*}
+
+\noindent\textbf{Contacts involved:} All (the 4 legs).
+
+\noindent\textbf{Heuristics:} $h_w$ is used for the all legs. The robustness treshold $b_0$ is set to $4$.
+
+\noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is the harder to compute for our planner.
+While finding a guide path above the hole is easy for the planner, finding a relevant sequence of contacts that remain in equilibrium is not trivial.
+Secondly, the narrow bridge is hard both for the planner and the contact generator: To make sure that equilibrium is preserved along the traversal,
+the bridge must be approached with the right angle.
+This can be seen in Figure (TODO), where several feet repositionning are required to cross both obstacles (although the video shows this best).
+The planner however suceeds in finding a feasible sequence in the end, again with interactive computation times.
+
+\subsubsection{HRP-2 -- Path replanning(Figure~\ref{figures/replanning}):}
+In this long scene, HRP-2 plans a path through through several obstacles. the scene is edited during the execution of the motion: a stair is added,
+some stepping stones are removed, and part of the final staircase is deleted. All these modifications involve replanning.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\linewidth]{figures/replanning}
+  \caption{
+           HRP-2 in the replanning scenario. After the red step stones are removed, a new sequence of contacts is replanned. Hand contacts
+           are not presented here for readability.}
+		   \label{fig:replanning}
 \end{figure}
 
-%~ \subsection{Implementation details}
-%~ The test application was developed using the C++ language.
-%~ The volumes describing the Range Of Motions of the robots $A_{ROM}$ are computed using the Matlab mpt toolbox~\footnote{http://people.ee.ethz.ch/\~mpt/3/}.
-%~ They are exported to the obj format using a custom Matlab script.
-%~ 
-%~ The robot is described using the urdf file format~\footnote{http://wiki.ros.org/urdf}, used as a standard by the popular ROS platform.
-%~ The scenarios where described using a custom text file format.
-%~ Environments are described in the obj format. Rendering is achieved using Blender~\footnote{http://www.blender.org}.
-%~ The runs were performed on a laptop with an Intel Core i7-2760QM 2.40GHz processor and 4 GB of memory. The application is not multi-threaded.
+\noindent\textbf{Contacts involved:} Both feet and the right arm.
 
-\noindent\textbf{Manipulation of a pen (3-finger hand):}
+\noindent\textbf{Heuristics:} $h_w$ is used for the all legs. $h_{EFORT}$ is used for the right arm. The robustness treshold is set to $2$.
+
+\noindent\textbf{Observations:} This scenario is designed to illustrate concretely the computation times of the planner.
+In the video, the footsteps indicating the contact sequence appear at the average speed of their computation (including the guide path planning).
+
+\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{figures/penrot}):}
 This scenario is proposed to illustrate the genericity of our approach: we consider a manipulation task for a robotic hand and use
@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
 \centering
@@
 		   \label{fig:penrot}
-\end{figure}
+\end{figure*}
 
-%\subsubsection*{Other scenarios}
-\noindent\textbf{Other scenarios (humanoids): }
-%Two additional scenarios were considered for the humanoid robot. A crouching scenario, and a standing up scenario (already presented in Figure~\ref{fig:framework} -- right) -- Humanoid robot). 
-The standing-up scenario (already presented in Fig.~\ref{fig:framework}) is a setup taken from \cite{DBLP:conf/iser/EscandeKMG08}: it corresponds to a long narrow passage in the configuration space.
-In the crouching scenario, demonstrated in the companion video, the character automatically goes from a standing to a crouching position to crawl under an obstacle.
+\noindent\textbf{Contacts involved:} Three finger-tips.
 
-\subsection{Parametrization of the reachability condition} \label{sec:stat1}
+\noindent\textbf{Heuristics:} $h_{EFORT}$ is used for the all fingers.
 
-To find the appropriate $C_{reach}^s$ in which to sample the guide trajectory,
-we computed the rejection rate for various values of $s$ for each robot in the most cluttered truck scenario.
-For a given value of $s$, $10^6$ root positions and orientations are computed
-in $C_{reach}^s$. In each case we try to generate a collision free contact configuration, with a database
-comprising $N=10^5$ sample configurations for each limb. The rejection rate
-is the ratio between the number of failures  and the number of trials. From Figure~\ref{fig:rejection} $s$ is empirically chosen
-as the smallest value for which the rejection rate is minimal.
-For the humanoid, we thus chose $s=2.2$, and for the insect, $s=2.8$.
+\subsection{Performance analysis} \label{sec:perf}
+To analyze performance, for each considered scenario, we ran the simulation 100 times.
+We measured the computation time spent in each aspect of the algorithm, and also analyzed the success
+rate obtained for each scenario.
 
-\begin{figure}[t]
-\centering
-  \begin{overpic}[width=0.6\linewidth]{figures/rejectionnotext}
-		 \put (1.1,24){\rotatebox{90}{\tiny{Rejection rate}}}
-		 \put (4.8,52) {\tiny{20} }
-		 \put (4.8,48) {\tiny{18} }
-		 \put (4.8,43) {\tiny{16} }
-		 \put (4.8,38) {\tiny{14} }
-		 \put (4.8,34) {\tiny{12} }
-		 \put (4.8,29) {\tiny{10} }
-		 \put (4.8,24.5) {\tiny{8} }
-		 \put (4.8,20) {\tiny{6} }
-		 \put (4.8,15) {\tiny{4} }
-		 \put (4.8,10) {\tiny{2} }
-		 \put (11.5,5) {\tiny{1} }
-		 \put (24.5,5) {\tiny{1.5} }
-		 \put (39,5) {\tiny{2} }
-		 \put (52,5) {\tiny{2.5} }
-		 \put (45,3) {\tiny{s} }
-		 \put (66,5) {\tiny{3} }
-		 \put (79,5) {\tiny{3.5} }
-		%~ \put (37,58) {2)} 
-		%~ \put (68,58) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
-	\end{overpic}
-\caption{Truck scenario rejection rates(\%) for the humanoid(orange) and insectoid(blue), given $s$.}
-		   \label{fig:rejection}
-\end{figure}
+Table TODO summarizes the performance measurement obtained, in terms of computation times.
 
-\subsection{Performance}\label{sec:stat2}
-The number of samples used for generating the contacts of each limb is 10000.
-Table~\ref{tab:requestime} presents the average time (seconds) spent in the phases of the planner, for 
-each phase and each scenario, and the number of contact phases of the sequence.
+For each scenario, one can observe that average computation time for one single step is largely below the second,
+thus allowing to consider interactive applications. In some cases such as the stair climbing, this computation time is actually negligeable. 
+Considering the repartition of the computation time, it appears that the majority of the time is spent performing inverse kinematics.
+This is not surprising considering the number of calls to the methods: IK is used intensively to maintain contact continuity between two postures; 
+it is also applied every time a new candidate needs to be evaluated. However regarding state of the art IK solutions, we believe there is room
+for improving the performances of the algorithm.
+
+The other important piece of information to consider is the success rate obtained for each scenario.
+With the exception of the truck egress scenario, this rate is really high.
+From a pragmatic point of view, these results allow us to claim that our approach is the best current compromise between completeness and efficiency:
+Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to replan a new contact sequence also rapidly with 
+a reasonable chance of success.
 

### 2015-12-07 body/appendix_roms.tex
appendix

\section{Generating the $W$ volumes for HRP-2}
\label{app:rom}

In this Appendix,
we detail our generic method to generate the volumes $W$ used
in RB-RRT; with the example of HRP-2.

### 2015-12-07 body/conclusion.tex
appendix

 \section{Discussion and future work} 
\label{sec:conclusion}

In this paper we consider acyclic contact planning in cluttered environments, formulated as two sub problems that we address sequentially:
\Pa: the computation of a guide trajectory for the root of the robot that can be extended ; \Pb: the computation
of a discrete sequence of contacts along this trajectory.
Our contribution to \Pa is a generic characterization of the properties that the guide trajectory must staisfy, in particular to enforce the completeness of the acyclic contact planner. We introduced a low dimensional space  $C_{reach}$ that can be mapped 
into the contact submanifold of the robot, approximated and efficiently sampled by our Reachability-Based planner.
Our contribution to \Pb is a pragmatic contact generation scheme that can take into
account criteria to enforce interesting properties on the generated contacts (such as robustness, energy efficiency or naturalness).

Aside from the theoretical contributions, our results demonstrate that our method allows a very interesting compromise between three 
criteria that are hard to conciliate: generality, performance, and quality of the solution, making it the first acyclic contact
planner compatible with interactive applications.
%
\textbf{Regarding generality}, the reachability condition, coupled with an approach based on limb decomposition, 
allows the method to address arbitrary multiped robots. The only pre-requisite is the specification 
of the volumes $W^0$.
%
\textbf{Regarding performance}, our framework outperforms existing methods in addressing either \Pa or \Pb, leading to computation costs close to real-time in
statically known environments.
%
\textbf{Regarding the quality of the trajectories}, a parametrization of the reachability condition allows us to compute
relevant trajectories in all the scenarios presented, with low rejection rates.
As for \cite{Bouyarmane2009}, failures can still occur, due to the compromise criterion used in computing the guide trajectory.
The low computational burden of our framework however allows for fast replanning when the solution fails.

One direction for future work is to focus on a more accurate formulation of $C_{reach}$ to address 
this limitation.
Our main objective is of course to generate the complete interpolation between the contact sequences.
We have already been able to achieve this for HRP-2 for some of the sequences computed \citep{Carpentier2016}, but we concede that additional
work is required on the planning side to obtain a seemless workflow. To achieve this, we are currently working on the notion of transition certificate, ie formulating
conditions that guarantee that the interpolation between two contact configurations is dynamically feasible. We are also
working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).

### 2015-12-07 body/rbprm.tex
appendix

@@
 The parametrization of $s$ allows to find a trade-off between these two desirable properties.
-Section~\ref{sec:results} shows that in practice, it is easy to adjust $s$ to keep most of the interesting guides without introducing incorrect guides.
+As an example, Figure~\ref{fig:hyq_roms} presents the volumes $W$ computed for the HyQ robot.
+In Appendix~\ref{app:rom}, we give a generic method to compute these volumes, with the example of HRP-2.
+%~ Section~\ref{sec:results} shows that in practice, it is easy to adjust $s$ to keep most of the interesting guides without introducing incorrect guides.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.95\linewidth]{figures/hyq_roms}
+  \caption{
+           $W$ volumes for HyQ. The green shapes represent the reachable workspace $W^k$ of each limb. The red shape is $W^0$, a scaling of the convex hull
+           of the torso of the robot.}
+		   \label{fig:hyq_roms}
+\end{figure}
 
@@
 In so doing we make the problem tractable for sampling based approaches, with extremely efficient computation times.
-
-%~ Algorithm~\ref{alg:generate} describe the sampling process for one configuration.
-
-%~ Two user defined numeric values, $LIMIT1$ and $LIMIT2$ are used to ensure that the generation will not loop
-%~ infinitely around areas of the workspace where the reachability condition cannot be met.
-%~ Two configurations are connected if a linearly interpolated path between them also belongs to 
-%~ $C_{reach}$.
-
-%~ \begin{algorithm}
-%~ \caption{Generation of a configuration $\mathbf{q}^{0} \in C_{reach}^s$} \label{euclid}
-	%~ \begin{algorithmic}[1]
-	%~ \Function{GenerateConfiguration}{}
-		%~ \State $\mathbf{q}^{0} \gets [0,...,0]^T$ /*Set all DOFs to 0*/
-		%~ \State $\textit{ntry1} \gets 0; \textit{ntry2} \gets 0;$
-		%~ \While {$\textit{ntry1} < \textit{LIMIT1}$}
-			%~ \State Select an obstacle $O_i \in O$
-			%~ \While {$\textit{ntry2} < \textit{LIMIT2}$}				
-				%~ \State Translate $\mathbf{q}^{0}$ to a random point on the surface of $O_i$
-				%~ \State $\mathbf{d} \gets $ Random unit direction
-				%~ \While {$W(\mathbf{q}^{0}) \cap O_i \ne \emptyset$}
-					%~ \State $\epsilon \gets$ random number, $ 0 < \epsilon \leq 1$
-					%~ \State Translate $\mathbf{q}^{0}$ by $\mathbf{u} = \epsilon \mathbf{d}$
-					%~ \If {$W^0_s(\mathbf{q}^{0}) \cap O = \emptyset$}
-						%~ \State \textbf{return} $\mathbf{q}^{0}$
-					%~ \Else
-						%~ \State Rotate $\mathbf{q}^{0}$ randomly
-						%~ \State $\textit{ntry2} \gets \textit{ntry2} + 1$
-					%~ \EndIf
-				%~ \EndWhile
-			%~ \EndWhile
-			%~ \State $\textit{ntry1} \gets \textit{ntry1} + 1$
-		%~ \EndWhile
-		%~ \State \textbf{return} failure
-	%~ \EndFunction
-%~ \end{algorithmic}
-%~ \label{alg:generate}
-%~ \end{algorithm}
-
-%To summarize, in this section we define a class of spaces $C_{reach}^s$.We  propose to use RB-PRM, a sampling based motion planner, to compute guide trajectories entirely contained in $C_{reach}^s$.
-Such trajectories provide a guaranty that a robot will be able to create contacts along them with a probability depending on a parameter $s$.

### 2015-12-07 body/results.tex
appendix

@@
 
-\subsubsection{HyQ -- Obstacle race:}
+\subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:hyq_bridge} and~\ref{fig:hyq_obs}):}
 In this long scene, HyQ is first required to cross a 55 cm large hole; then, to cross a narrow "bridge", only large of 25 cm,
-while maintaining equilibrium
+while maintaining equilibrium.
 
-%~ \begin{figure*}
-  %~ \centering
-  %~ \includegraphics[width=0.5\linewidth]{figures/stair}
-  %~ \caption{
-           %~ HRP-2 in the steep stair climbing scenario. }
-		   %~ \label{fig:stair_robust}
-%~ \end{figure*}
+\begin{figure}
+  \centering
+  \includegraphics[width=0.4\linewidth]{figures/hyq_bridge}
+  \caption{
+           HyQ crossing a narrow bridge. }
+		   \label{fig:hyq_bridge}
+\end{figure}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=0.5\linewidth]{figures/hyq_obs}
+  \caption{
+           Crossing a hole contact sequence for HyQ ($b_0 > 4$). }
+		   \label{fig:hyq_obs}
+\end{figure*}
 
@@
 
-\subsubsection{HRP-2 -- Path replanning(Figure~\ref{figures/replanning}):}
+\subsubsection{HRP-2 -- Path replanning(Figure~\ref{fig:replanning}):}
 In this long scene, HRP-2 plans a path through through several obstacles. the scene is edited during the execution of the motion: a stair is added,
@@
 
-For each scenario, one can observe that average computation time for one single step is largely below the second,
+For each scenario, one can observe that the average computation time for one single step is largely below the second,
 thus allowing to consider interactive applications. In some cases such as the stair climbing, this computation time is actually negligeable. 
@@
 
-\begin{table}[b]
+\begin{table*}
 \centering
@@
 \quad
- \end{table}
-
- We observe that many contacts are required for the insect, which can be explained by its restricted range of motion.
- The time spent generating the navigation graph is about one minute. The time spent in generating the graph of the climbing scenario,
- despite the relatively open environment, is explained by the additional restrictions imposed on the reachability condition.
- The difficulty to find a balanced configuration essentially influences the time spent generating the contacts.
-
-The number of contacts in the sequence gives a rough estimation of its duration in seconds.
-Except for the robots crawling out of the truck, all the contact generation are real-time.
-Additionaly to the quality of the generated trajectories shown in the video, these computation times are a major practical achievement.
-
-%~ \subsection{Early results: manipulating scenario in zero gravity}
-%~ In this scenario, we suggest an extension of our framework to manipulation.
-%~ We consider the task of rotating a pen using a robotic hand.
-%~ A start and goal configurations for the pen are expressed in the world coordinate system.
-%~ The problem is automatically reformulated into an equivalent motion planning one for the robotic hand, by expressing its start
-%~ and goal configurations in the coordinate system of the pen. In other words, for the planning the pen is considered as a static object, and a feasible motion for the hand
-%~ is computed along the pen. Figure {penrot} presents the results obtained with this approach.
-%~ Additional work is required to generate feasible manipulation motions.
+ \end{table*}

### 2015-12-08 body/abstract.tex
Add my comments to abstract

--- 2015-11-24 body/abstract.tex
@@
 then, generating relevant contacts along this path.
-We propose theoritical contributions for each sub-problem, and implement them
+We propose theoretical contributions for each sub-problem, and implement them
 as a complete contact planner. 
-First, we propose a geometrical characterization of a guide trajectory, the "reachability condition", that guarantees that a
-guide can be mapped into the contact manifold of the robot. With this condition, asserting the "true feasibility" 
-of a guide path boils down to performing collision tests, thus avoiding the costful of explicit computation of balanced contact configurations, as opposed to previous approaches.
+First, we propose a geometrical characterization of a guide trajectory, the ``reachability condition", which guarantees that a
+guide can be mapped into the contact manifold of the robot. With this condition, asserting the``true feasibility" 
+of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
 Then, we propose a sequential algorithm to extend a truly feasible path into a sequence of contacts. For each limb, an offline sampling approach,
 independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
-The best candidate is chosen according to user created heuristics. We propose two such heuristics, used in all our experiments, and 
-designed for generic robotic applications: a manipulatibily based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
-These contributions are implemented within the open source software Hpp. Our framework allows us to generate complex multi contact sequences for a large
+The best candidate is chosen according to user-created \adnote{user-defined?} heuristics. We propose two such heuristics, used in all our experiments, and 
+designed for generic robotic applications: a manipulatibily-based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
+These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
 variety of robots, including a HRP-2, Hyq, and a dextrous hand, with interactive performances.

### 2015-12-08 body/results.tex
results updated

@@
 \subsection{Performance analysis} \label{sec:perf}
-To analyze performance, for each considered scenario, we ran the simulation 100 times.
+To analyze performance, for each considered scenario, we ran the simulation 1000 times.
 We measured the computation time spent in each aspect of the algorithm, and also analyzed the success
@@
 
+\subsection{Computation times}
 For each scenario, one can observe that the average computation time for one single step is largely below the second,
@@
 for improving the performances of the algorithm.
+
+\subsection{Success rates times}
+TODO contact generation 95 %.
 

### 2015-12-08 body/results.tex
update results

@@
 %\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}
-\subsubsection{HRP-2 -- Steep staircase (Figure~\ref{fig:standing}):}
+\subsubsection{HRP-2 -- Steep staircase (Figure~\ref{fig:stair_robust}):}
 
@@
 
-\begin{figure}[h!]
+\begin{figure}
   \centering
@@
 
-\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{figures/penrot}):}
+\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot}):}
 This scenario is proposed to illustrate the genericity of our approach: we consider a manipulation task for a robotic hand and use
-our contact planner to compute a guide trajectory for the fingers, considered as effectors (Figure \ref{fig:penrot}).
+our contact planner to compute a guide trajectory for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
 Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences, this in less than 5 seconds.
 
-\begin{figure*}[t]
+\begin{figure*}
 \centering
@@
 
-Table TODO summarizes the performance measurement obtained, in terms of computation times.
+\subsection{Computation times}
+Table~\ref{tab:requestime} summarizes the performance measurement obtained, in terms of computation times.
 
-\subsection{Computation times}
-For each scenario, one can observe that the average computation time for one single step is largely below the second,
-thus allowing to consider interactive applications. In some cases such as the stair climbing, this computation time is actually negligeable. 
+\begin{table*}
+\centering
+\begin{tabular}{ l | >{\centering\arraybackslash}m{60pt} | >{\centering\arraybackslash}m{60pt} | >{\centering\arraybackslash}m{60pt} | >{\centering\arraybackslash}m{60pt} | >{\centering\arraybackslash}m{60pt} | c}
+  &  Complete guide generation & Collision & Static equilibrium & ik  & Total generation time & Time per step\\
+ \hline
+   Steep stairs & 22 -- \textbf{33} --  76 & 1 --  \textbf{4} -- 61   & 22 --  \textbf{52} -- 534 & 108 --  \textbf{1069} -- 10550 & 267 --  \textbf{1337} -- 11484 & \textbf{66} \\
+   %Standing up & $70$ & 23 & 48\\
+ \end{tabular}
+\caption{minimum, \textbf{average} and worst time (in ms) spent in the generation process for each scenario and each critical part of the generation process. The last
+column indicates the average time necessary to compute one contact transition.}
+\label{tab:requestime}
+\quad
+\end{table*}
+
+In most scenarios, one can observe that the average computation time for one single step is largely below the second,
+thus allowing to consider interactive applications. 
 Considering the repartition of the computation time, it appears that the majority of the time is spent performing inverse kinematics.
@@
 
-\subsection{Success rates times}
-TODO contact generation 95 %.
+\subsection{Success rates}
+Table~\ref{tab:requestpercent} summarizes the success rates obtained for each scenario.
+\begin{table*}
+\centering
+\begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | c}
+  &  Path planning & True feasibility & Kinematic failure & Equilibrium failure \\
+ \hline
+   Steep stairs & 100\%  & 99.5\% & 0.11\% & 0.39\% \\
+   Standing up & 100\% & 89\% & 7\% & 4\% \\
+   Truck egress & \% & \% & \% &  \% \\
+   Rubble & \% & \% & \% & \% \\
+   Obstacle race & \% & \% & \% & \% \\
+ \end{tabular}
+\caption{Success rates for each scenario. The first value indicates the percentage of sucessfull complete path plannings for 1000 tests; The second value
+indicates the percentage of truly feasible root configurations: considering each limb individually, indicates the percentage of root placements of the guide trajectory that
+led to a feasible contact. Kinematic failure is the percentage of contact generations that failed because no collision-free candidate was found. Equilibrium failure is the percentage of contact
+generations that failed because no candidate respected the static equilibrium condition.}
+\label{tab:requestpercent}
+\quad
+\end{table*}
 
-The other important piece of information to consider is the success rate obtained for each scenario.
-With the exception of the truck egress scenario, this rate is really high.
-From a pragmatic point of view, these results allow us to claim that our approach is the best current compromise between completeness and efficiency:
+We observe that, as expected, our planner does not succeed systematically, because of the approximations made in the implementation
+of our theoritical formulation. 
+From a pragmatic point of view, regarding the computation times, we claim that our approach is the best current compromise between completeness and efficiency:
 Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to replan a new contact sequence also rapidly with 
 a reasonable chance of success.
-
-\begin{table*}
-\centering
-\begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | c}
-  &  Generate RB-PRM (offline) & Generating the contact sequence & Number of contact states\\
- \hline
-   Truck egress (humanoid) & $ 73 $ & 15 & 10 \\
-   Truck egress (insectoid) & $70$ & 23 & 48\\
-   Climbing (humanoid)& $25$ &  5 & 15\\
-   Climbing (insectoid)  & $ 21 $ & 27 & 51\\
-   Crouching (humanoid)& $5$ & 6 & 22 \\
- \end{tabular}
-\caption{\textbf{Average time} (in seconds) spent in RB-PRM generation, and the online generation of the contact sequence.}
-\label{tab:requestime}
-\quad
-\end{table*}

### 2015-12-08 body/appendix_hpp.tex
appendix done, missing all stats and truck

\section{Source code of our planner}
\label{app:hpp}
Our planner is implemented using the Humanoid Path Planner (HPP) software.
HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS, that can be found at \url{http://projects.laas.fr/gepetto/index.php/Software/Main}.
The HPP modules implement the standard tools and algorithms used in motion planning,
such as the Bi-RRT planner from which RB-RRT is derived.

The robot models used in our experiments are described using the standard urdf file format, that is 
compatible with HPP.

Our implementation of the planner is also open source, and can be found at \url{https://github.com/stonneau/hpp-rbprm}. 
Contrary to the mature HPP, at the time of submitting this article,
our planner is not officialy released, mostly for lack of a complete documentation and a tutorial.

However a reader familiar with HPP should be able to use the library.
Our goal is to make our code useful for the community, and we are working on releasing a stable 
version of our planner as soon as possible.

### 2015-12-08 body/appendix_roms.tex
appendix done, missing all stats and truck

@@
 In this Appendix,
-we detail our generic method to generate the volumes $W$ used
-in RB-RRT; with the example of HRP-2.
+we detail our method to generate the volumes $W$ used
+in RB-RRT, with the example of HRP-2.
+
+We consider the robot HRP-2, and proceed to the decomposition 
+of its kinematic chain into four limbs $R^k$.
+The arms are connected to the shoulders, and the legs to the root.
+The obtained volumes $W$ are shown in Figure~\ref{fig:hrp2_w}.
+
+\begin{figure}
+\centering
+  \begin{overpic}[width=1\linewidth]{figures/hrp2_w}
+	\end{overpic}
+\caption{The $W$ volumes computed for HRP-2. The red shapes are $W^0$. The green shapes represent the $W^k$.}
+		   \label{fig:hrp2_w}
+\end{figure}
+
+\subsection{Step 1: computing the reachable workspace $W^k$ of a limb}
+
+To generate a volume $W^k$, we proceed as follows:
+\begin{enumerate}
+\item Generate randomly $N$ valid limb configurations for $R^k$, for $N$ really large (say $100000$);
+\item For each configuration, store the 3D position of the end effector joint relatively to the root of $R^k$; then compute the convex hull of the resulting point cloud;
+\item The resulting polytope can contain a very large number of faces.
+A last step is thus to simplify it in a conservative way with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}).
+For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
+\end{enumerate}
+
+\begin{figure}
+\centering
+  \begin{overpic}[width=1\linewidth]{figures/roms}
+	\end{overpic}
+\caption{Different approximations of the range of motion of the right arm HRP-2. Left: non convex-hull, computed with the powercrust algorithm~\citep{Amenta:2001:PC:376957.376986}. Middle:
+convex hull of the reachable workspace. Right: Simplified hull used in our experiments.}
+		   \label{fig:hrp2_roms}
+\end{figure}
+
+Figure~\ref{fig:hrp2_roms} illustrate the obtained $W^k$ for HRP-2.
+Regarding the procedure, we can see that step 2 is conservative (Figure~\ref{fig:hrp2_roms}--right), which 
+is acceptable, especially because the lost set essentially relates to configurations close to singularity (they are close to the boundaries of the reachable workspace).
+We choose again to be less complete
+but more efficient, regarding the number of collision tests to be performed by RB-RRT.
+In step 1 on the other hand, selecting the convex hull (Figure~\ref{fig:hrp2_roms}--middle) instead of a minimum encompassing shape (Figure~\ref{fig:hrp2_roms}--left) may introduce false positives.
+Concretely, because the false positive set intersects with $W^0$, the scaling volume of the robot torso, the induced error is compensated,
+as verified by the date of Table~\ref{tab:requestpercent}.
+
+\subsection{Step 2: computing the trunk scaling workspace $W^0$ of the robot}
+To define the volume $W^0$ of HRP-2, we proceed in an empirical manner.
+First, we compute the bounding boxes of the robot trunk, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
+Then, we perform a scaling of these boxes by a factor $s$. For different values of $s$, we verify the true feasibility condition:
+For a large number of configurati
+ons that verify the reachability condition, we try to generate contact configurations. We plot 
+the success rates obtained depending on $s$ (Figure~\ref{hrp2_w}). We then choose the minimal satisfying value of $s$.

### 2015-12-08 body/intro.tex
appendix done, missing all stats and truck

@@
 The relaxation is justified empirically, with statistics presented for each robot and scenario presented. The complete framework is implemented within
-the open source platform Hpp.
+the open source platform HPP.
 
@@
 This extension completes this work with an argumented discussion on the implementation of this theory to real world robots and problems.
-In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot. We also provide details on the hpp implementation
+In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot. We also provide details on the HPP implementation
 of our contact generation algorithm. Furthermore, our experiments are now applied to real robot models, namely HRP-2 and Hyq. TODO Appendix describes the generation process

### 2015-12-09 body/intro.tex
Add my comments to Section I and II

--- 2015-12-04 body/intro.tex
@@
-We consider the problem of planning the acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment.
-Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to support the equilibrium of the system.
+% !TEX root =  ../main.tex
+We consider the problem of planning the \adnote{an} acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment.
+Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to support \adnote{maintain} the equilibrium of the system.
 
-Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. But executing
-this behaviour on cluttered environments is dangerous, if not impossible.
-In an analysis of their participation to the darpa challenge, \citeauthor{atkensondarpa}
+Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. But \adnote{don't start sentences with 'but', you can start with 'however' followed by a comma} executing
+this behavior on cluttered environments is dangerous, if not impossible.
+In an analysis of their participation to the Darpa Robotics Challenge, \citeauthor{atkensondarpa}
 noted: ``Except for egress, no robots in the DRC Finals used
@@
 Indeed, the current approach to locomotion planning is to avoid obstacles as much as possible, instead of using them
-to faciliate locomotion. The reason for this, as the authors state, is that ``More contacts make tasks
+to facilitate locomotion. The reason for this, as the authors state, is that ``More contacts make tasks
 mechanically easier, but algorithmically more complicated for planning, and the transitions are difficult to
 both plan and control[...].  We have seen very few robot planners that are  capable of  generating this  behavior."
-The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of valid configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
+The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of valid \adnote{'valid' is not super clear to me here, I'd rather say 'collision-free'} configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
 
 Previous contributions that embrace the combinatorial provide complete approaches, not applicable in practice because they require hours of computations \citep{conf/iser/BretlRLKA04}.
-More recent successes use a local solution, resulting in more reasonnable performances (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
-Neither global or local methods presents satifying performances, because planning simultaneously the robot trajectory and the contacts that allow
+More recent successes use a local solution, resulting in more reasonable performances \adnote{'performances' is ambiguous, I'd rather use 'computation times'} (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
+Neither global nor local methods present satisfying performances, because planning simultaneously the robot trajectory and the contacts that allow
 its execution is too costly.
@@
 This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
-This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely rapidly,
-while preserving a success rate superior to TODO \% in the demonstrated scenarios.
+This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely rapidly, \adnote{I don't like having two adverbs ending in 'ly' one after the other, 'extremely fast' sounds better to me}
+while preserving a success rate greater than TODO \% in the demonstrated scenarios.
 
-In the remainder of this introduction, we discuss further the current state of the art. This allows us to then situate more precisely our contributions.
+In the remainder of this introduction, we discuss further the current state of the art. This allows us to then \adnote{this 'then' sounds bad here} situate more precisely our contributions.
 
@@
 
-Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences and biomechanics, and virtual character animation.
+Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences, biomechanics, and virtual character animation.
 Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}.
-These approaches can intrinsically not adapt to new situations or discover complex behaviors in unforeseen contexts.
+These approaches can intrinsically not \adnote{can not intrinsically?
+are intrinsically not able to?} adapt to new situations or discover complex behaviors in unforeseen contexts.
 
 The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper.
-The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third non trivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
+The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: \adnote{planning?
+finding?
+computing?} a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third non trivial \adnote{http://www.thefreedictionary.com/nontrivial} problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories.
-This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not scalable to arbitrary environments.
+This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not scalable \adnote{'scalable' means 'capable of being scaled'} to arbitrary environments.
 Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.
@@
 A solution can then be found, at the cost of prohibitive computation times (several hours).
-Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the relevance of the guide trajectory,  which is not evaluated \textit{a priori}.
+Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the relevance \adnote{I don't like 'relevance', what about 'quality'?} of the guide trajectory,  which is not evaluated \textit{a priori}.
 \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body contact configurations in static equilibrium.
@@
 
-It thus appears that an integrated optimization based approach \citep{Mordatch:2012:DCB:2185520.2185539}, able to generate a locally optimal, complete trajectory within minutes, currently outperforms
-existing planners, decoupled or not. However, recent contributions are able to interpolate dynamically executable 
-trajectories between contact configurations \citep{herzog2015trajectory, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real time performances.
+It thus appears that an integrated optimization-based approach \citep{Mordatch:2012:DCB:2185520.2185539}, able to generate a locally-optimal, complete trajectory within minutes, currently outperforms
+existing planners, decoupled or not. However, recent contributions are able to interpolate dynamically-executable \adnote{dynamically-feasible?}
+trajectories between contact configurations \citep{herzog2015trajectory, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real-time performances.
 To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly. 
-Such a planner holds the promise of a complete real time multi contact locomotion system.
+Such a planner holds the promise of a complete real-time multi-contact locomotion system.
 
 \subsection{Paper contribution and organization}
-This paper presents a pragmatic approach to break the complexity of multi contact planning. With that objective in mind,
-we believe that the separation between the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
+This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
+we believe that the separation between the \adnote{generation of the} guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
 However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
 \begin{itemize}
-\item The guide trajectory must satisfy a property guaranteeing the existence of a contact sequence to actuate it\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties.}.
+\item The guide trajectory must satisfy a property guaranteeing \adnote{replace 'satisfy a property guaranteeing' with 'guarantee'} the existence of a contact sequence to actuate it\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties.}.
 We call this property \textit{true feasibility}.
-This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi flat cases \citep{zucker2010optimization}.
-\item There is an infinite combination of possible contact sequences for a given root trajectory.
-The selection of one particular contact sequence with interesting properties (minimum number of contact change, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly pick one contact sequence, leading to possibly very tedious contact sequences).
+This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
+\item There is an infinite combination \adnote{There are infinite combinations} of possible contact sequences for a given root trajectory.
+The selection of one particular contact sequence with interesting properties (minimum number of contact change, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious contact sequences \adnote{repetition of 'contact sequence'}).
 \end{itemize}
@@
 By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
-For each problem, we introduce an original formulation of the related theoritical issues.
+For each problem, we introduce an original formulation of the related theoretical issues.
 Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
-The relaxation is justified empirically, with statistics presented for each robot and scenario presented. The complete framework is implemented within
+The relaxation is justified empirically, with statistics presented for each robot and scenario presented\adnote{several robots and scenario, or even better, for N robots and M scenarios}.
+The complete framework is implemented within
 the open source platform Hpp.
@@
 To address \Pa we rigorously formulate a geometric condition for the true feasibility of a path, the reachability condition. We claim that this 
-condition can be verified in a low dimensional space, that we call $C_{reach}$.
+condition can be verified in a low-dimensional space, that we call $C_{reach}$.
 This formulation can make the planning of a guide trajectory more efficient computationally, while providing equivalent guarantees to planning directly in the configuration space.
@@
 an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained.
-This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly feasible paths for the root of the robot within a few milliseconds to a few seconds depending on the scenario.
+This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot within a few milliseconds to a few seconds depending on the scenario.
 
-To address \Pb, we consider a truly feasible path as an input, and generate new contact configurations along it using a sampling approach.
+To address \Pb, we consider a truly-feasible path as an input, and generate new contact configurations along it using a sampling approach.
 The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed).
-The theorical completeness of the approach is partially traded for efficiency: instead of sampling online
+The theoretical completeness of the approach is partially traded for efficiency: instead of sampling online
 limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
-for proximity queries. At runtime, this framework allows to obtain extremly rapidly a set of candidate configurations for contact.
-These candidates are sorted using a criterion to evaluate the quasi static equilibrium of the configuration, and ranked according to user defined heuristics.
-In our implementation we use two such heuristics, used in all our experiments: a robustness criterion regarding the static equilibrium of a configuration, and a manipulability based heuristic,
+for proximity queries. At runtime, this framework allows to obtain extremely rapidly \adnote{again, extremely rapidly} a set of candidate configurations for contact.
+These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, and ranked according to user-defined heuristics.
+In all our experiments we used two such heuristics: a measurement of robustness of a configuration to errors in the contact forces, and a manipulability-based heuristic,
 the Extended FORce Transmission ratio (EFORT).
 
-The contributions of the paper are thus twofold.
+The contribution of the paper is thus twofold.
 We propose a theoretical characterization of today's most efficient practical approach to sampled-based planning of acyclic contacts.
@@
 
-In comparison with previous work, our planner does not produce a complete motion, but a discrete sequence of contacts.
-Because we showed that the interpolation of a complete trajectory between such configurations can be solved within real time constraints \citep{Carpentier2016}, we claim that our planner
-is the fastest multi-contact planning solution to our knowlege.
+In comparison with previous work, \adnote{Compared to previous work/Contrary to previous work [cite Mordatch,...]} our planner does not produce a complete motion, but a discrete sequence of contacts.
+Since the interpolation of a complete trajectory between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
+is the fastest multi-contact planning solution to our knowledge.
 
 The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep(tonneauisrr15).
-The conference paper focuses on the theoritical formulation of the problem, and presents results obtained with virtual avatars.
-This extension completes this work with an argumented discussion on the implementation of this theory to real world robots and problems.
-In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot. We also provide details on the hpp implementation
-of our contact generation algorithm. Furthermore, our experiments are now applied to real robot models, namely HRP-2 and Hyq. TODO Appendix describes the generation process
+The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
+This extension completes this work with a \adnote{I remove the word 'argumented' because it does not exist and I've not been able to find a good replacement} discussion on the implementation of this theory to real-world robots and problems.
+In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot despite bounded error in the contact forces. We also provide details on the hpp implementation
+of our contact generation algorithm. Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ\adnote{cite Semini}. TODO Appendix describes the generation process
 for the robot.
@@
 In Section~\ref{overview}, we present the general organization of our method.
-Section~\ref{rbprm} and Section~\ref{sec:efort} present respectively our answer to problems \Pa and \Pb.
-In Section~\ref{heuristics}, we present two heuristics used for the selection of the best contact configuration.
-Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and Hyq robots, and a three-finger manipulator) in various scenarios,
+Section~\ref{rbprm} and Section~\ref{sec:efort} present respectively our answer to \adnote{I remove 'problems'} \Pa and \Pb.
+In Section~\ref{heuristics}, we present two heuristics used for the selection of \adnote{I remove 'the best'} a contact configuration.
+Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
 that complete those presented in the conference paper.

### 2015-12-09 body/overview.tex
corrections andrea
Add my comments to Section I and II

@@
+% !TEX root =  ../main.tex
 \section{Overview}
@@
 \Pa and \Pb are addressed in a sequential fashion.
-The guarantees provided by the truly feasible guide path allow
+The guarantees provided by the truly-feasible guide path allow
 us to avoid the combinatorial by generating the contacts in a straightforward manner.
@@
 \end{figure*}
-\subsection{Computation of a guide path -- \Pa}
+\subsection{Computation of a guide path --- \Pa}
 We first consider the problem of planning a relevant guide path.
@@
 
- An intuitive description of such placements is "close, but not too close": close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Fig.~\ref{fig:contact_gen}--1); not too close, because the robot must avoid collision (which is represented by the hull including the torso in Fig.~\ref{fig:contact_gen}--1).
-We define formally $C_{reach}$, the set of interesting root placements, in which we compute a guide trajectory with a sampling based planner,  the Reachability-Based RRT --RB-RRT-- (Figure~\ref{fig:framework}--\Pa).
+ An intuitive description of such placements is ``close, but not too close'': close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Fig.~\ref{fig:contact_gen}--1); not too close, because the robot must avoid collision (which is represented by the hull including the torso in Fig.~\ref{fig:contact_gen}--1).
+We define formally $C_{reach}$, the set of interesting \adnote{'interesting' is vague, consider replacing with 'the set of root placements that allow the robot not to be in collision and be in contact with the environment'; alternatively you can even be more precise I write this in math, but maybe that's not the right place to be formal}  root placements, in which we compute a guide trajectory with a sampling based planner,  the Reachability-Based RRT --RB-RRT-- (Figure~\ref{fig:framework}--\Pa).
 Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
@@
 
-\subsection{Generating a discrete sequence of contact configurations -- \Pb}
+\subsection{Generating a discrete sequence of contact configurations --- \Pb}
 
 The second stage extends the guide path into a sequence of contact configurations (Fig.~\ref{fig:framework}--\Pb). The true feasibility of the input path is guaranteed by the first step.
-This provides in turn a reasonnable guarantee that a contact configuration can be obtained for any root placement along the path. With a fixed root position, the dimensionality of the 
+This provides in turn a reasonable \adnote{FYI: you always wrote 'reasonable' as 'reasonnable'} guarantee that a contact configuration can be obtained for any root placement along the path.
+With a fixed root position, the dimensionality of the
 contact generation problem is reduced to the number of degrees of freedom of the considered limb.
-In other words, we consider each limb as a manipulator attached to the root, and select the most relevant contact from a database of precomputed configurations, independent from the environment (Figure~\ref{fig:contact_gen}).
+In other words, we consider each limb as a manipulator attached to the root, and select the most relevant contact from a database of precomputed configurations, which is independent from the environment (Figure~\ref{fig:contact_gen}).
 
-From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root placement, an inverse kinematics solver 
+From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root placement, an inverse-kinematics solver 
 is used to maintain the contacts that were previously existing (TODO Figure). Often, these contacts cannot be maintained, because of joint limits or collisions (TODO Figure).
-In such a case the contacts are broken. Conversely, new contacts are created to ensure the quasi static balance of the robot (TODO FIGURE).
-The algorithm is designed so that only one contact can be created or broken between two successive configurations. This can be seen as a heuristic to ensure
+In such a case the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot (TODO FIGURE).
+The algorithm is designed so that only one contact can be created or broken between two successive configurations. This can be seen as a heuristic \adnote{why heuristic?} to ensure
 that the interpolation between two configurations is achievable.
@@
 
-To select the "most relevant" contact configurations, user defined heuristics can be chosen. In Section~\ref{heuristics} we present extensively two such heuristics, that have been
+To select the ``most relevant'' contact configurations, user-defined heuristics can be chosen. In Section~\ref{heuristics} we present extensively two such heuristics, that have been
 used in our experiments.
@@
 
-A vector  $\mathbf{x}$ is denoted with a bold lower case letter.
-A matrix $\mathbf{A}$ is denoted with a bold upper case letter.
-A set $C$ is denoted with a upper case italic letter.
-Scalar variables and functions are denoted with lower case italic letters, such as
+A vector  $\mathbf{x}$ is denoted with a bold lower-case letter.
+A matrix $\mathbf{A}$ is denoted with a bold upper-case letter.
+A set $C$ is denoted with an upper-case italic letter.
+Scalar variables and functions are denoted with lower-case italic letters, such as
 $r$ or $f(\textbf{x})$.
@@
 \medskip
-\textbf{A robot} is a kinematic chain $R$, comprising $n + 6$ degrees of freedom (DOFs).
+\textbf{A robot} is a kinematic chain $R$, comprising \mbox{$n + 6$} degrees of freedom (DOFs).
 $R$ is composed of $l$ limbs $R^k, 1 \leq k \leq l$, attached to a root.
@@
 	\item $\mathbf{q}^{\overline{k}}$ denotes the vector of joint values of R \textbf{not} related to $R^k$.
-We define for convenience $\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
+We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
 	\item $\mathbf{q}^{0}\in SE(3)$ denotes the position and orientation of the root of the robot $R$.
@@
 \begin{equation}
-  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \mathbf{q}^k, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\}
+  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\},
 \end{equation}
-where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0$ the null displacement. We also define $W = \bigcup_{k=1}^{l}W^k$.
-%
-$W^k(\mathbf{q}^{0}), 0 < k \leq l$ is defined as the volume $W^k$ translated and rotated by the rigid displacement~$\mathbf{q}^{0}$.
+\adnote{in this definition you don't account for joint limits}
+where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0$ being the null displacement. We also define $W = \bigcup_{k=1}^{l}W^k$, and
+$W^k(\mathbf{q}^{0})$ (for $1 \leq k \leq l$) as the volume $W^k$ translated and rotated by the rigid displacement~$\mathbf{q}^{0}$.
 
 \medskip
-\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ it contains.
+\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains.
 Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.

### 2015-12-09 body/intro.tex

@@
 The complete framework is implemented within
-the open source platform Hpp.
+the open source platform HPP.
 
@@
 This extension completes this work with a \adnote{I remove the word 'argumented' because it does not exist and I've not been able to find a good replacement} discussion on the implementation of this theory to real-world robots and problems.
-In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot despite bounded error in the contact forces. We also provide details on the hpp implementation
+In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot despite bounded error in the contact forces. We also provide details on the HPP implementation
 of our contact generation algorithm. Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ\adnote{cite Semini}. TODO Appendix describes the generation process

### 2015-12-14 body/conclusion.tex
Andrea's first complete review

@@
 
-In this paper we consider acyclic contact planning in cluttered environments, formulated as two sub problems that we address sequentially:
-\Pa: the computation of a guide trajectory for the root of the robot that can be extended ; \Pb: the computation
-of a discrete sequence of contacts along this trajectory.
-Our contribution to \Pa is a generic characterization of the properties that the guide trajectory must staisfy, in particular to enforce the completeness of the acyclic contact planner.
-We introduced a low dimensional space  $C_{reach}$ that can be mapped
-into the contact submanifold of the robot, approximated and efficiently sampled by our Reachability-Based planner.
+In this paper we consider acyclic contact planning in cluttered environments, formulated as two sub problems that we address sequentially.
+The first problem is \Pa, that consists in computing a guide path for the root of the robot\deleted{ that can be extended}; the second problem is \Pb, that is the computation of a discrete sequence of contacts along the root path.
+Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner.
+We introduced a low-dimensional space $C_{reach}$ that can be mapped
+into the contact submanifold of the robot, approximated, and efficiently sampled by our Reachability-Based planner.
 Our contribution to \Pb is a pragmatic contact generation scheme that can take into
@@
 %
-\textbf{Regarding performance}, our framework outperforms existing methods in addressing either \Pa or \Pb, leading to computation costs close to real-time in
-statically known environments.
+\textbf{Regarding performance}, our framework outperforms existing methods in addressing either \Pa or \Pb, \adnote{saying you outperform existing methods when you do not show any comparison is risky, the reviewers might ask for comparisons} leading to computation costs close to real-time in
+\replaced{known static}{statically known} environments.
 %
-\textbf{Regarding the quality of the trajectories}, a parametrization of the reachability condition allows us to compute
-relevant trajectories in all the scenarios presented, with low rejection rates.
-As for \cite{Bouyarmane2009}, failures can still occur, due to the compromise criterion used in computing the guide trajectory.
-The low computational burden of our framework however allows for fast replanning when the solution fails.
+\textbf{Regarding the quality of the paths}, a parametrization of the reachability condition allows us to compute
+relevant paths in all the presented scenarios, with low rejection rates.
+As for \cite{Bouyarmane2009}, failures can still occur, due to the compromise criterion used in computing the guide path.
+The low computational burden of our framework however allows for fast replanning \replaced{in case of failure}{when the solution fails}.
 
@@
 Our main objective is of course to generate the complete interpolation between the contact sequences.
-We have already been able to achieve this for HRP-2 for some of the sequences computed \citep{Carpentier2016}, but we concede that additional
-work is required on the planning side to obtain a seemless workflow. To achieve this, we are currently working on the notion of transition certificate, ie formulating
+We have already been able to achieve this for HRP-2 for some of the computed sequences~\citep{Carpentier2016}, but we \replaced{admit}{concede} that additional
+work is required on the planning side to obtain a seamless workflow. To achieve this, we are currently working on the notion of transition certificate, i.e. formulating
 conditions that guarantee that the interpolation between two contact configurations is dynamically feasible. We are also
 working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
+\adnote{The 'future work' part seems a bit poor to me}

### 2015-12-14 body/contact.tex
appendix done, missing all stats and truck
Andrea's first complete review

@@
+% !TEX root =  ../main.tex
 \section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
@@
 We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.
-From true feasibility we know that such a trajectory exists.
+\deleted[remark={the input is not truly feasible, actually it is not even geometrically feasible, it is only approximately geometrically feasible}]{From true feasibility we know that such a trajectory exists.}
 
-In this Section we first describe a single contact generation process, that is how to generate a contact configuration for a given limb 
+In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a given limb 
 and root location.
-Then, on this method we build an iterative algorithm to generate a discrete sequence of contact configurations in equilibrium, along a truly feasible path.
+Then, on this method we build an iterative algorithm to generate a discrete sequence of contact configurations in equilibrium\deleted{, along a truly feasible path}.
 
-In what follows a \textit{valid} contact configuration is understood as being collision free, respecting joint limits, and allowing to maintain the static equilibrium
+In what follows a \textit{valid} contact configuration is understood as being collision free, satisfying joint limits, and allowing to maintain the static equilibrium
 of the robot. The first two tests are trivial; our criterion to assert efficiently the static equilibrium of the system
@@
 \label{sec:single_contact}
-We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$ truly feasible regarding $R^k$. FIGURE
-From true feasibility we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify Eq.~\eqref{eq:pi} for $\mathbf{q}^0$.
+We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$\deleted{ truly feasible regarding $R^k$}. FIGURE
+From true feasibility we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify \eqref{eq:pi} for $\mathbf{q}^0$.
 The issue is then to evaluate locally at least one element of $P$ at $\mathbf{q}^0$.
@@
 \begin{enumerate}
-\item Generate randomly a valid limb configuration;
+\item Generate randomly a \replaced{collision-free}{valid} limb configuration;
 \item Project the end-effector onto the closest surface with inverse kinematics;
@@
 
-We partially trade the completeness property of this approach for a more efficient solution previously introduced \citep{Tonneau2014}, that we describe
+We partially trade the completeness property of this approach for a more efficient solution introduced in our previous work~\citep{Tonneau2014}, which we describe
 here for completeness.
@@
 We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimal distance 
-between an effector and a surface of the environment is lesser than a small value $\epsilon$.
+between an effector and a surface of the environment is less than a small value $\epsilon$.
 We then apply the following steps:
@@
 \item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree stata structure;
-\item Figure~\ref{fig:contact_gen}--2 and 3--: At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
+\item Figure~\ref{fig:contact_gen}--2 and 3--:\adnote{This notation for subfigures is not clear, consider using subfigures, e.g.
+2a, 2b} At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
 \item According to a user-defined heuristic $h$, sort the elements of $S$;
 \item Figure~\ref{fig:contact_gen}--4 and 5--: Select the first configuration of $S$ (or fail if $S$ is empty). Project the configuration onto contact with inverse kinematics;
-\item If a valid solution is found, stop. Otherwise remove the element from $S$ and go back step 5.
+\item If a valid \adnote{being valid depends also on the other contacts because it regards also equilibrium, doesn't it?
+Then this assessment depends on the other contacts previously generated.
+Doesn't this have implications on the completeness?
+Even if it doesn't, isn't it worth to talk about it?} solution is found, stop.
+Otherwise remove the element from $S$ and go back step 5.
 \end{enumerate}
 
-This approach has two main advantages:
+This approach has two main advantages.
 First, this allows us to select a large number of candidates in a single proximity request.
-Having several candidates is interesting, because it allows to compare them regarding a user selected heuristic $h$, thus obtaining
-a locally optimal candidate.
-Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution,
-and this in a small number of iterations.
+Having several candidates is interesting, because it allows to compare them \replaced{using}{regarding} a user-selected heuristic $h$, thus obtaining
+a locally-optimal candidate.
+Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
 Regarding completeness, it is immediate to verify that as $N$ and $\epsilon$ grow, the completeness of the method is restored.
@@
 
-We thus have defined a real time, extremely efficient contact generator, that can be as complete as required by the context.
+We thus have defined a real-time\deleted{, extremely efficient} contact generator that can be as complete as required by the context.
 %We compute the image of a mapping $f$ along $\mathbf{q}^0(t)$ into a discrete sequence of contacts $ \in C_{contact}$.
@@
 \end{equation*} 
+\adnote{Not clear, what is g?}
 We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
@@
 At each step, we consider a new root placement $\mathbf{q}^0_{i}$ and try to generate a contact configuration that preserves as much as possible
-the previous contacts, and is in static equilibrium.
+the previous contacts while allowing for static equilibrium.
 
@@
 
-Figure~\ref{fig:break_contact} illustrates the contact persistence strategy.
-If possible, a limb in contact at time $i-1$ remains in contact at $i$.
-The contact is broken if an inverse kinematics solver fails to find a collision free limb configuration that satisfies joint limits.
-The solver is directy provided by the Hpp software.
+Fig.~\ref{fig:break_contact} illustrates the contact-persistence strategy.
+If possible, a limb in contact at time $i-1$ remains in contact at time $i$.
+The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits.
+The solver is directly provided by the HPP software.
 
-If the solver fails, the contact is broken and a collision free configuration is assigned to the limb.
-If more than one contacts are broken for a single step, intermediate steps are added to reposition, one at a time,
-the contacts, before moving.
+If the solver fails, the contact is broken and a collision-free configuration is assigned to the limb.
+If two or more contacts are broken in a single time step, intermediate steps are added to reposition the contacts one at a time, before moving.
 
@@
 \begin{enumerate}
-\item Only one contact creation can happen between two consecutive states. This increases the odds that the interpolation between the two states will be feasible; 
+\item Only one contact creation can happen between two consecutive states.
+This increases the odds that the interpolation between the two states \deleted[remark={not 100\% sure}]{will} be feasible; \adnote{Before you said that you ensured that transition between two states is feasible...}
 \item A contact is validated if and only if the resulting configuration is in static equilibrium, no matter what user-defined heuristics are used; 
-\item We use a FIFO approach:  we always try to first create a contact with the limb that has been free for the longest time. If the contact creation
+\item We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
 was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others have been treated.
@@
 of the deterministic approach used to break the combinatorial.
-However, in practice the planner is successful in the large majority of cases, as discussed in section~\ref{sec:stat1}, thanks to 
-the heuristics we use for relevant contact selection.
+However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:stat1}, thanks to 
+the heuristics that we use for contact selection.

### 2015-12-14 body/efort.tex
Andrea's first complete review

@@
-\subsection{Manipulability-based heuristics for locally optimal contact selection}
-The other kind of heuristics we use are based on the analysis of the motion capabilities
+% !TEX root =  ../main.tex
+\subsection{Manipulability-based heuristics for locally-optimal contact selection}
+The other kind of heuristics that we use are based on the analysis of the motion capabilities
 provided by a candidate configuration regarding the task being achieved.
-These heuristics allow to locally optimize a contact location regarding a particular task.
-For instance, once can be interested in configurations that allow to efficienctly exert a force compatible with the global direction of motion,
+These heuristics allow to locally optimize a contact location with respect to a particular task.
+For instance, one can be interested in configurations that allow to efficiently exert a force compatible with the global direction of motion,
 a high velocity in a given direction, or to stay away from singular configurations.
-In these section, three such heuristics are derived in this section from the work on manipulability by \cite{Yoshikawa1984}, that we present first.
+In this section, we derive three such heuristics from the work on manipulability by \cite{Yoshikawa1984}, that we present first.
 
 \subsubsection{The force and velocity ellipsoids.}
-For a limb configuration $\mathbf{q}^k$, the jacobian matrix
+For a limb configuration $\mathbf{q}^k$, the Jacobian matrix
  $\mathbf{J}(\mathbf{q}^k)$   defines the relation:
@@
 \begin{equation} \label{jac}
-\mathbf{\dot{p}^k} = \mathbf{J}(\mathbf{q}^k) \dot{\mathbf{q}^k}
+\mathbf{\dot{p}^k} = \mathbf{J}(\mathbf{q}^k) \dot{\mathbf{q}}^k
 \end{equation}
@@
 
-As a linear approximation of the forward kinematics function $f$ , $\mathbf{J}$ describes how small
+As a linear approximation of the forward-kinematics function $f$ , $\mathbf{J}$ describes how small
 variations from the configuration $\mathbf{q}$ affect the position vector $\mathbf{p}$.
 It can be seen as the transformation mapping a velocity in the configuration space into a velocity in
-the euclidian space.
+the Euclidian space.
 
 Now we consider the unit ball in the configuration space $C$ defined by the set of joint velocities
-for which the norm is inferior or equal to 1:
+for which the norm is at most 1:
 \begin{equation} \label{ball}
@@
 
-From Equation~\ref{jac} we can obtain the following equality:
+From \eqref{jac} we can obtain the following equality: \adnote{how?}
 \begin{equation} \label{jaceq}
@@
 
-We can use Equation~\ref{jaceq} to map the ball into an ellipsoid in the euclidian space $\mathbb{R}^m$:
+We can use \eqref{jaceq} to map the ball into an ellipsoid in the Euclidian space $\mathbb{R}^m$:
 
@@
 This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of velocities that can
-be reached under the constraints of Equation~\ref{ball} for the current configuration.
+be reached under the constraint \eqref{ball} for the current configuration.
 The longer an axis of the ellipsoid is, the faster the configuration can move along the direction of the axis.
-Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.
+Fig.~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.
 
@@
 
-Deriving from Equation~\ref{jac}, we define that the set of achievable force in $\mathbb{R}^m$  subject to the constraint:
+Exploiting \eqref{jac}, we can easily see that the set of achievable forces in $\mathbb{R}^m$ subject to the constraint:
 \begin{equation*} \label{ballforce}
@@
 \end{equation*}
-is the force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right) defined by:
+is the so-called force ellipsoid (Fig.~\ref{sec:efort_ellipsoid} - right), defined by:
 \begin{equation} \label{ellipsoidforce}
@@
 
-\subsubsection{Manipulability based heuristics.}
-From these definitions, we can derive three useful heuristics, that all account from the environment and the task being performed.
-The first one, EFORT, is introduced by \cite{Tonneau2014}; the other two are new minor contributions, derived from these previous works.
+\subsubsection{Manipulability-based heuristics.}
+From these definitions, we can derive three useful heuristics, that all account \replaced{for}{from} the environment and the task being performed.
+The first one, EFORT, was introduced by \cite{Tonneau2014}; the other two are new minor contributions, derived from these previous works.
 
-With EFORT, we define the efficiency of a configuration as the ability a limb has to exert a force in a given direction.
+With EFORT, we define the efficiency of a configuration as the ability of a limb to exert a force in a given direction.
 We thus consider the force ellipsoid as a basis for our heuristic.
-In a given direction $\mathbf{m}$, the length of the ellispoid is given by the force transmission ratio \citep{1087795}:
+In a given direction $\mathbf{m}$, the length of the ellipsoid is given by the force-transmission ratio \citep{1087795}:
 \begin{equation*}
@@
 
-If the ability to apply important velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasonning on the velocity ellispoid:
+If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
 \begin{equation}
@@
 
-$h_{EFORT}$ and $h_{vel}$ will favor contacts that allow important efforts or fast modifications in the velocity.
+$h_{EFORT}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
 $EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
-In other less demanding cases, manipulability can also be considered regarding its original purpose, that means to avoid singularties.
+In other less demanding cases, manipulability can also be considered \deleted{regarding its original purpose, that means} to avoid singularities.
 To do so, we can consider the manipulability measure $w$, also given by \citeauthor{Yoshikawa1984}:
@@
 w(\mathbf{q}) = \sqrt{\mathbf{J}\mathbf{J}^T}
-\end{equation*}
-$w$ measures the ``distance" between a given configuration and a singular configuration. When $w$ is equal to 0, the configuration is in a singular state;
+\end{equation*} \adnote{Didn't you forget something in this equation? I guess $w$ is a scalar, but the right side is a matrix.}
+$w$ measures the ``distance'' between a given configuration and a singular configuration. When $w$ is equal to 0, the configuration is in a singular state;
 the greater $w$ is, the further away the configuration is from singularity.
@@
 
-$h_{EFORT}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling based approach, the jacobian an inverse products
-can be precomputed offline), that allow the planner to select the more relevant
-candidates regarding the local motion task along the guide path.
-This result in faster computation times, and a reduced number of contact switches overall.
+$h_{EFORT}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
+can be precomputed offline), that allow the planner to select the \replaced{best}{more relevant}
+candidates \replaced{with respect to}{regarding} the local motion task along the guide path.
+This results in faster computation times, and a reduced number of overall contact switches.

### 2015-12-14 body/equilibrium.tex
Andrea's first complete review

@@
+% !TEX root =  ../main.tex
 \section{Heuristics for relevant contact selection}
@@
 \subsection{A heuristic for robust static equilibrium}
-The planner is designed so that any contact configuration it outputs is in static equilibrium.
+The planner is designed so that any generated contact configuration is in static equilibrium.
 Equilibrium is of course critical in multipedal locomotion. For this reason we are interested in a robust 
-criterion, that ensures that the robot remains balanced in a real world application, regardless of the errors
-that can result from perception and control uncertainties.
+criterion, that ensures that the robot remains balanced in a real-world application, regardless of perception and control uncertainties.
 
@@
 From this formulation we derive a new LP that quantifies the robustness of the configuration regarding equilibrium.
-From this value we can in turn either define a heuristic to choose the most robust candidate, or set a treshold on the required robustness
-for any configuration. While the LP we propose is original, it is based on an analysis of the problem we proposed in \cite{Prete2016}, where the interested reader
-can find more details.
+In turn, from this value we can either define a heuristic to choose the most robust candidate, or set a threshold on the required robustness\deleted{for any configuration}.
+While the presented LP is original, it is based on an analysis of the problem that we proposed in \cite{Prete2016}, where the interested reader can find more details.
 
@@
 \item $\mu$ is the friction coefficient;
-\item for the i-th contact $1 \leq i \leq e$:
+\item for the i-th contact point $1 \leq i \leq e$:
 	\begin{itemize}
@@
 
-According to Coulomb's law, the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
+According to Coulomb's law, \replaced{slipping occurs only if a contact force is on the boundaries of the friction cone.}{ the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.}
 As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
@@
 Any force belonging to the linearized cone
-can thus be express as a convex combination of its four generating rays.
-Therefore we can express the non slipping constraint on $\mathbf{f}_i$ as:
+can thus be expressed as a convex combination of its four generating rays.
+\deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}
 
 \begin{equation*}
-\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} , \bm{\beta}_i \ge 0 \wedge \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i \qquad 
+\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} : \bm{\beta}_i \ge 0 \wedge \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i,
 \end{equation*}
+\added{where $\bm{\beta}_i$ contains the coefficient of the cone generators.}
+We can then stack all the constraints to obtain:
+\begin{equation}\label{eq:gen}
+\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \wedge \mathbf{f} = \mathbf{V} \bm{\beta},
+\end{equation}
+where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = (\mathbf{f}_0,...,\mathbf{f}_e)$.
 
-we can stack the constraints to obtain:
-\begin{equation}\label{eq:gen}
-\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \wedge \mathbf{f} = \mathbf{V} \bm{\beta} \qquad 
-\end{equation}
-where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = [\mathbf{f}_0,...,\mathbf{f}_e]^T$.
-
-From the Newton-Euler equations, and assuming a null acceleration, we also constrain the sum of contact forces to compensate the gravity at the center of mass:
+From the Newton-Euler equations, and assuming a null acceleration of the COM, we want the contact forces to compensate the gravity at the center of mass:
 
@@
 where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.
-The right side of \ref{eq:new_eul} is independent from $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as:
+The right hand side of \eqref{eq:new_eul} is independent of $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as: \adnote{It is not really necessary to show that equilibrium is independent of the com altitude here, anyway the com is given, not a variable, so it does not change the size of the problem}
 \begin{align} \label{eq:new_eul2d}
@@
 
-If there exists a $\bm{\beta}^*$ satisfying equations \ref{eq:gen} and \ref{eq:new_eul2d}, it means that the configuration is statically balanced.
+If there exists a $\bm{\beta}^*$ satisfying \eqref{eq:gen} and \eqref{eq:new_eul2d}, it means that the configuration is statically balanced.
 The problem can then be formulated as an LP:
@@
 \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
-& \bm{\beta} \ge b_0 \\
+& \bm{\beta} \ge b_0 \bm{1}\\
 \end{aligned} \end{equation}
 
-We observe that if $b_0$ is positive, LP~\ref{eq:lin_prog} admits a solution, and $b_0$ is proportional to the minimum distance of the equilibrium contact forces to the boundaries of the friction cones.
-If $b_0$ is negative, the configuration is not in static equilibrium, and the value $b_0$ indicates "how far" from equilibrium
-the configuration is. $b_0$ can thus be seen as a measure of robustness.
+We observe that if $b_0$ is positive then \eqref{eq:lin_prog} admits a solution, and $b_0$ is proportional to the minimum distance of the equilibrium contact forces to the boundaries of the friction cones.
+If $b_0$ is negative, the configuration is not in static equilibrium, and $b_0$ indicates ``how far'' from equilibrium the configuration is. We thus use $b_0$ as a measure of robustness.
 
-In our implementation, we solve the dual of LP~\ref{eq:lin_prog_rob}.
-after a trivial elimination of the Lagrange multipliers associated to the inequality constraints, the dual problem has a lower dimensionality:
+In our implementation, rather than solving directly \eqref{eq:lin_prog_rob}, we solve an equivalent problem of smaller dimension that we get by taking the dual of \eqref{eq:lin_prog_rob} and eliminating the Lagrange multipliers associated to the inequality constraints:
 \begin{equation} \label{eq:dual} \begin{aligned}
@@
 
-Indeed, from Slater's condition \citep{Boyd:2004:CO:993483}, we know that the optimal values of an LP and its dual are equal. Thus the optimal value of LP~\ref{eq:dual} is indeed the optimal $b_0$.
+Indeed, from Slater's conditions \citep{Boyd:2004:CO:993483}, we know that the optimal values of an LP and its dual are equal.
+Thus the optimal value of the LP~\eqref{eq:dual} is indeed the optimal $b_0$.
 
-In conclusion, LP~\ref{eq:dual} provides a method to assert rapidly static equilibrium, and returns a robustness value.
+In conclusion, the LP~\eqref{eq:dual} provides a method to assert rapidly static equilibrium, and returns a robustness value.
 This robustness criterion can be immediately transformed into a heuristic, and eventually be ponderated with others.

### 2015-12-14 body/intro.tex
Andrea's first complete review

@@
 \begin{itemize}
-\item The guide trajectory must satisfy a property guaranteeing \adnote{replace 'satisfy a property guaranteeing' with 'guarantee'} the existence of a contact sequence to actuate it\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties.}.
+\item The guide trajectory must satisfy a property guaranteeing \adnote{replace 'satisfy a property guaranteeing' with 'guarantee'} the existence of a contact sequence to actuate it\adnote{Are you sure that you wanna use an 'endnote' here rather than a 'footnote'?}\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties.
+\adnote{not clear what 'discrete bounded properties' means}}.
 We call this property \textit{true feasibility}.
-This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
+This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contacts\adnote{ forces?}, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
 \item There is an infinite combination \adnote{There are infinite combinations} of possible contact sequences for a given root trajectory.
-The selection of one particular contact sequence with interesting properties (minimum number of contact change, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious contact sequences \adnote{repetition of 'contact sequence'}).
+The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious contact sequences \adnote{repetition of 'contact sequence'}).
 \end{itemize}
@@
 Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
-The relaxation is justified empirically, with statistics presented for each robot and scenario presented\adnote{several robots and scenario, or even better, for N robots and M scenarios}.
+The relaxation is justified empirically, with statistics presented for each robot and scenario presented\adnote{several robots and scenarios, or even better, for N robots and M scenarios}.
 The complete framework is implemented within
-the open source platform HPP.
+the open-source platform HPP.
 
@@
 
-To address \Pa we rigorously formulate a geometric condition for the true feasibility of a path, the reachability condition. We claim that this 
-condition can be verified in a low-dimensional space, that we call $C_{reach}$.
-This formulation can make the planning of a guide trajectory more efficient computationally, while providing equivalent guarantees to planning directly in the configuration space.
+To address \Pa we rigorously formulate a geometric condition \replaced{that approximates}{for} the true feasibility of a path, the reachability condition. We claim that this 
+condition can be verified in a low-dimensional space, which we call $C_{reach}$.
+This formulation can make the planning of a guide path more efficient computationally, while providing \replaced[remark={planning in $C_{reach}$ you don't account for equilibrium}]{simular}{equivalent} guarantees to planning directly in the configuration space.
 In practice, for each considered robot, we compute an approximation of $C_{reach}$. From this approximation,
 an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained.
-This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot within a few milliseconds to a few seconds depending on the scenario.
+This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot \replaced{in a timespan ranging from}{within} a few milliseconds to a few seconds, depending on the scenario.
 
@@
 These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, and ranked according to user-defined heuristics.
+\adnote{Not clear, first you say the candidates are sorted based on static equilibrium, then you say they are ranked based on user-defined heuristics.
+What's the difference between sorting and ranking?}
 In all our experiments we used two such heuristics: a measurement of robustness of a configuration to errors in the contact forces, and a manipulability-based heuristic,

### 2015-12-14 body/overview.tex
Andrea's first complete review

@@
 We first consider the problem of planning a relevant guide path.
-The objective is to compute a path of root placements which will allow contact creation.
+The objective is to compute a path of root placements that will allow contact creation.
 A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide trajectory, but at the same time, any computed guide trajectory must be truly feasible, i.e.
@@
 
- An intuitive description of such placements is ``close, but not too close'': close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Fig.~\ref{fig:contact_gen}--1); not too close, because the robot must avoid collision (which is represented by the hull including the torso in Fig.~\ref{fig:contact_gen}--1).
-We define formally $C_{reach}$, the set of interesting \adnote{'interesting' is vague, consider replacing with 'the set of root placements that allow the robot not to be in collision and be in contact with the environment'; alternatively you can even be more precise I write this in math, but maybe that's not the right place to be formal}  root placements, in which we compute a guide trajectory with a sampling based planner,  the Reachability-Based RRT --RB-RRT-- (Figure~\ref{fig:framework}--\Pa).
+\added
+{
+More precisely, a root placement is truly feasible if and only if there exist at least one joint configuration such that the robot satisfies three conditions: }
+\begin{description}
+\item[C1] is not in collision with the environment, 
+\item[C2] has at least one end-effector in contact with the environment, 
+\item[C3] is in static equilibrium. 
+\end{description}
+\added{
+The conditions C1 and C2 are purely geometrical and can be efficiently approximated even without computing the contact points.
+Instead C3 is difficult to approximate and it requires the computation of the contact points, and the resolution of a linear program to compute the contact forces~\cite{DelPrete2016}.
+Moreover our intuition suggests us that, for cluttered environment, C3 is not critical, meaning that, most of the times, root placements satisfying C1 and C2 would satisfy also C3.
+For this reason we neglect C3 during the resolution of \Pa, and we worry about it only when solving \Pb.
+}
+
+An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Fig.~\ref{fig:contact_gen}--1); not too close, because the robot must avoid collision (which is represented by the hull including the torso in Fig.~\ref{fig:contact_gen}--1).
+We define formally $C_{reach}$, the set of \replaced{root placements satisfying C1 and C2}{interesting root placements}, in which we compute a guide trajectory with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Fig.~\ref{fig:framework}---\Pa).
 Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
@@
 
-The second stage extends the guide path into a sequence of contact configurations (Fig.~\ref{fig:framework}--\Pb). The true feasibility of the input path is guaranteed by the first step.
-This provides in turn a reasonable \adnote{FYI: you always wrote 'reasonable' as 'reasonnable'} guarantee that a contact configuration can be obtained for any root placement along the path.
+The second stage extends the guide path into a sequence of contact configurations (Fig.~\ref{fig:framework}--\Pb).
+\deleted{The true feasibility of the input path is guaranteed by the first step.
+This provides in turn a reasonable guarantee that a contact configuration can be obtained for any root placement along the path.
+}
 With a fixed root position, the dimensionality of the 
@@
 In such a case the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot (TODO FIGURE).
-The algorithm is designed so that only one contact can be created or broken between two successive configurations. This can be seen as a heuristic \adnote{why heuristic?} to ensure
-that the interpolation between two configurations is achievable.
+The algorithm is designed so that only one contact can be created or broken between two successive configurations.
+This \replaced{ensures}{can be seen as a heuristic to ensure} that the interpolation between two configurations is achievable.
 Details are presented in Section~\ref{sec:efort}.
 
-To select the ``most relevant'' contact configurations, user-defined heuristics can be chosen. In Section~\ref{heuristics} we present extensively two such heuristics, that have been
-used in our experiments.
+To select the ``most relevant'' contact configurations, user-defined heuristics can be chosen.
+In Section~\ref{heuristics} we present extensively two such heuristics, which have been used in our experiments.
 

### 2015-12-14 body/rbprm.tex
appendix done, missing all stats and truck
Andrea's first complete review

@@
+% !TEX root =  ../main.tex
 \section{Computing a guide path in $C_{reach}$ ($\mathcal{P}_1$) }
@@
 
-We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the geometrical root of a multiped robot, connecting start and goal configurations.
-As said in the previous section, the goal is to enforce that any configuration $\mathbf{q}^0$ of the guide is truly feasible, i.e.
+We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the geometrical \adnote{why the root is geometrical?
+Isn't the path to be geometrical?} root of a multiped robot, connecting \adnote{user-defined} start and goal configurations.
+As said in the previous section, the goal is to enforce that any configuration $\mathbf{q}^0$ of the guide is \replaced{geometrically}{truly} feasible, i.e.
 can be mapped to a configuration in contact.
-We denote by $C_{contact} \subset SE(3) \times \mathbb{R}^n$ the contact submanifold of the robot.
+We denote by $C_{contact} \subset SE(3) \times \mathbb{R}^n$ the contact submanifold of the robot\added{, that is the set of configurations in which the robot is in contact with the environment}.
 %~ We say that a root placement $\mathbf{q}^{0}$ is \textit{truly feasible} if there exists $\mathbf{q}^{\overline{0}}$ such that 
@@
 
-We say that a root placement $\mathbf{q}^{0}$ is \textit{truly feasible} if there exists at least one bijective mapping $\pi$
+We say that a root placement $\mathbf{q}^{0}$ is \textit{\replaced[remark={I don't replace 'truly feasible' with 'geometrically feasible' from now on because I don't know whether you like it}]{geometrically}{truly} feasible} if there exists at least one bijective mapping $\pi$
 such that
@@
 
-In this Section we characterize $C_{reach}$ theoritically with a necessary and a sufficient condition for belonging to the set.
+In this Section we characterize $C_{reach}$ theoretically with a necessary and a sufficient condition for belonging to the set.
 We then describe our implementation of an efficient RRT planner in $C_{reach}$.
@@
 
-\subsubsection*{Sufficient condition for true feasibility}
+\subsubsection*{\replaced[remark={for coherence with previous subsection title}]{True feasibility: sufficient condition}{Sufficient condition for true feasibility}}
 %~ In building $C_{reach}^1$, we directly considered the including hull of the root body $W^{0}$, obtaining a necessary condition. 
@@
 In practice, the construction of a non-trivial including shape $W^0$ was possible for all the robot structures we considered).
-We denote by $C_{reach}^\infty \subset C_{reach}$ the set of root placements corresponding to the sufficient condition.
+\adnote{Isn't this parenthesis a bit too long?} We denote by \mbox{$C_{reach}^\infty \subset C_{reach}$} the set of root placements corresponding to the sufficient condition.
 
 In general, the inclusion is strict, which means that we lose the completeness of the two-stage contact planner (i.e.
-the planner is not able to discover a trajectory inside $C_{reach} \setminus C_{reach}^\infty$).
+the planner is not able to discover a trajectory inside \mbox{$C_{reach} \setminus C_{reach}^\infty$}).
 However, the sufficient condition guarantees that any such trajectories leads to a valid sequence of contacts (i.e.
@@
 
-However between $W^0$ and $B^{max}$, a trade-off can be found between a necessary and a sufficient condition.
+However, between $W^0$ and $B^{max}$, a trade-off can be found between a necessary and a sufficient condition.
 We define $W^0_s$ as the volume $W^0$ subject to a scaling transformation by a factor $s \in \mathbb{R}^+$.
@@
 %
-The higher $s$ is, the closer this "reachability condition" is to being sufficient, and if $s=1$, the planner is complete.
+The higher $s$, the closer this ``reachability condition'' is to being sufficient, and if $s=1$, the planner is complete.
 The parametrization of $s$ allows to find a trade-off between these two desirable properties.
@@
 First, a configuration is set to a random point on the surface of one randomly chosen obstacle. The root location is then translated and rotated randomly until the reachability condition is satisfied.
-Our current implementation of these modifications is based on the Bi-RRT planner provided by the hpp software.
+Our current implementation of these modifications is based on the Bi-RRT planner provided by the HPP software.
 
-Thanks to this small set of modifications, the high dimensional truly feasible path planning problem is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for Hyq, 8 for HRP-2 that has a flexible torso).
-In so doing we make the problem tractable for sampling based approaches, with extremely efficient computation times.
+Thanks to this small set of modifications, the high dimensional truly feasible path planning problem is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for Hyq, 8 for HRP-2 that has \replaced{joints in the}{a flexible} torso).
+\replaced{By doing do}{In so doing} we make the problem tractable for sampling-based approaches, with extremely efficient computation times.

### 2015-12-14 body/results.tex
Andrea's first complete review

@@
+% !TEX root =  ../main.tex
 \section{Results}
@@
 In this Section we present some of the results obtained with our planner. The complete sequences computed are shown in the companion video, TODO.
-Specifically, we demonstrate the planner for two really different robots, in a large vartiety of environments: the humanoid HRP-2 and the quadriped Hyq.
-For each scenario we indicate the heuristics chosen. We also provide a performance analysis, that shows that the planner is compatible with interactive applications,
+Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ~\cite{Semini}.
+For each scenario we indicate the chosen heuristics. We also provide a performance analysis, which shows that the planner is compatible with interactive applications,
 and present the success rates obtained in each scenario.
-We also demonstrate the interest of our robustness criterion in the different computed poses.
+Moreover, we demonstrate the interest of our robustness criterion in the different computed poses.
 Finally, a last example suggests possible applications to dextrous manipulation.
 
-In our ISRR conference paper~\citep{tonneauisrr15}, additional results are demonstrated with various virtual avatars (Figure~\ref{fig:robots_old}).
-In this extension we choose to focus on results obtained with actual robots. We invite the interested reader to watch the ISRR video\endnote{https://www.youtube.com/watch?v=LmLAHgGQJGA}, and 
+In our previous work~\citep{tonneauisrr15} additional results are demonstrated with various virtual avatars (Fig.~\ref{fig:robots_old}).
+In this extension we choose to focus on \deleted{results obtained with} actual robots.
+We invite the interested reader to watch the ISRR video\endnote{https://www.youtube.com/watch?v=LmLAHgGQJGA}, and
 to refer to the previous paper for a discussion on these results.
@@
 
-The goal is to cross a set of stairs of a height of 20 cm. This height requires HRP-2 to use a ramp to be able to perform the task.
+The goal is to climb three 20-cm high steps. This height requires HRP-2 to use a ramp to \deleted{be able to} perform the task.
 
-\noindent\textbf{Contacts involved:} Both feet and right arm.
+\noindent\textbf{Contacts involved:} \deleted{Both} Feet and right arm.
 
 \noindent\textbf{Heuristics:} The extended manipulability $h_w$ is chosen for the feet; $h_{EFORT}$ is chosen for the right arm.
-Regarding equilibrium, the video demonstrates two sequences computed for two different treshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
+Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Fig.~\ref{fig:stair_robust}).
 
 \noindent\textbf{Observations:}
-This scenario illustrates best the importance of the equilibrium robustness criterion.
-With a robust approach, more states are required to reach the last step (15 against 13 in average).
-However, when the last step is reached by both feet, in the non robust case, the contacts are extremely close to 
-the cone limits, and we believe this can be seen visually (Figure~\ref{fig:stair_comp}).
+This scenario illustrates best the importance of the equilibrium-robustness criterion.
+With a robust approach, more states are required to reach the last step (15 \replaced{rather than}{against} 13 in average).
+However, when the last step is reached by both feet, in the nonrobust case the contacts are extremely close to 
+the cone limits\deleted{, and we believe this can be seen visually} (Fig.~\ref{fig:stair_comp}).
 
@@
 
-\subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing}):}
-From a bent configuration, a standing up motion is computed in a cluttered environment.
-The resulting motion involves using a wall as support, and crossing a 25 cm high stair.
+\subsubsection{HRP-2 -- Standing up (Fig.~\ref{fig:standing}):}
+From a bent configuration, a standing-up motion is computed in a cluttered environment.
+The resulting motion involves using a wall as support, and climbing a 25-cm high step.
 
@@
 
-\noindent\textbf{Heuristics:} $h_w$ is used for the feet, $h_{EFORT}$ is used for the hands.
+\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for the feet, $h_{EFORT}$ \deleted{is used} for the hands.
 
-\noindent\textbf{Observations:} The scenario illustrates well the acyclic aspect of the planning. For instance, in the 4 first frames of Figure~\ref{fig:standing}, we can see that the right foot
+\noindent\textbf{Observations:} The scenario illustrates well the acyclic aspect of the planning. For instance, in the four first frames of Fig.~\ref{fig:standing}, we can see that the right foot
 is moved twice, with the left foot in between, before the configuration allows HRP-2 to move its hand.
 Because the contacts are tried in a FIFO manner, the fact that the output contact sequence is acyclic shows that a cyclic approach (with a finite state machine for instance) is not sufficient
-for the computed trajectory. The reason for this is not reachability, but equilibrium. The planning is slower than for the stair scenario (because the contact generation fails more),
-though it remains compatible with interactive performances.
+for the computed path. The reason for this is not reachability, but equilibrium. The planning is slower than for the stair scenario (because the contact generation fails more),
+though it remains compatible with interactive performances. \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
 
-\subsubsection{HRP-2 -- Truck egress (Figure~\ref{fig:TODO}):}
-HRP-2 stands half sitted in a truck, and is required to exit through the front window.
+\subsubsection{HRP-2 -- Truck egress (Fig.~\ref{fig:TODO}):}
+HRP-2 stands half seated in a truck, and \replaced{is required to}{it must} \replaced{get out}{exit} through the front window.
 
@@
 
-\noindent\textbf{Heuristics:} $h_w$ is used for the feet, $h_{EFORT}$ is used for the hands.
+\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for the feet, $h_{EFORT}$ \deleted{is used} for the hands.
 
-\noindent\textbf{Observations:} This scenario demonstrates the ability of the planner to compute a path in an extremely constrained environment. In this case,
+\noindent\textbf{Observations:} This scenario demonstrates the ability of the planner to compute a path in an extremely-constrained environment. In this case,
 the scenario is not always real time (TODO or is it ?): while the planner is often able to compute a guide path rapidly, in the worst case up to one minute can be spent
@@
 
-\subsubsection{HyQ -- Darpa style rubble (Figure~\ref{fig:darpa})}
+\subsubsection{HyQ -- Darpa-style rubble (Fig.~\ref{fig:darpa})}
 The quadruped robot is given the task to cross a rubble composed of bricks rotated at different angles and directions.
@@
 
-\noindent\textbf{Heuristics:} $h_w$ is used for the all legs. The robustness treshold $b_0$ is set to $20$.
+\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for all legs. The robustness threshold $b_0$ is set to $20$.
 
@@
 stability of the HyQ robot, and results in more contact switches, in exchange for safety.
-The path planning time is higher than for the previous HRP-2 robots due to the constraint that the 4 reachable workspace of all legs must
-collide with the environment at all times. Again, the computed time remains however interactive.
+The path-planning time is higher than for the previous HRP-2 robots due to the constraint that the 4 reachable workspace of all legs must
+collide with the environment at all times. \adnote{Why do not relax this constraint and require only 3 legs?} Again, the computed time remains however interactive.
 
 \subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:hyq_bridge} and~\ref{fig:hyq_obs}):}
-In this long scene, HyQ is first required to cross a 55 cm large hole; then, to cross a narrow "bridge", only large of 25 cm,
-while maintaining equilibrium.
+In this long scene, HyQ is first required to cross a 55-cm large hole; then, to cross a narrow "bridge",  only 25-cm large\deleted{,
+while maintaining equilibrium}.
 
@@
 
-\noindent\textbf{Heuristics:} $h_w$ is used for the all legs. The robustness treshold $b_0$ is set to $4$.
+\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for all legs. The robustness threshold $b_0$ is set to $4$.
 
-\noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is the harder to compute for our planner.
-While finding a guide path above the hole is easy for the planner, finding a relevant sequence of contacts that remain in equilibrium is not trivial.
-Secondly, the narrow bridge is hard both for the planner and the contact generator: To make sure that equilibrium is preserved along the traversal,
+\noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is the hardest \deleted{to compute} for our planner.
+While finding a guide path above the hole is easy for the planner, finding a \deleted{relevant} sequence of contacts that \replaced{allows for}{remain in} equilibrium is not trivial.
+Second, the narrow bridge is hard both for the planner and the contact generator: to make sure that equilibrium is preserved along the traversal,
 the bridge must be approached with the right angle.
-This can be seen in Figure (TODO), where several feet repositionning are required to cross both obstacles (although the video shows this best).
-The planner however suceeds in finding a feasible sequence in the end, again with interactive computation times.
+This can be seen in Fig.~(TODO), where several feet \replaced{rearrangements}{repositioning} are required to cross both obstacles (although the video shows this best).
+The planner however succeeds in finding a feasible sequence in the end, again with interactive computation times.
 
@@
 In this long scene, HRP-2 plans a path through through several obstacles. the scene is edited during the execution of the motion: a stair is added,
-some stepping stones are removed, and part of the final staircase is deleted. All these modifications involve replanning.
+some stepping stones are removed, and part of the final staircase is deleted. All these modifications \replaced{demands}{involve} replanning.
 
@@
 
-\noindent\textbf{Contacts involved:} Both feet and the right arm.
+\noindent\textbf{Contacts involved:} \replaced{F}{Both f}eet and the right arm.
 
-\noindent\textbf{Heuristics:} $h_w$ is used for the all legs. $h_{EFORT}$ is used for the right arm. The robustness treshold is set to $2$.
+\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for all legs. $h_{EFORT}$ \deleted{is used} for the right arm. The robustness threshold is set to $2$.
 
 \noindent\textbf{Observations:} This scenario is designed to illustrate concretely the computation times of the planner.
-In the video, the footsteps indicating the contact sequence appear at the average speed of their computation (including the guide path planning).
+In the video, the footsteps indicating the contact sequence appear at the average speed of their computation (including the guide-path planning).
 
-\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot}):}
-This scenario is proposed to illustrate the genericity of our approach: we consider a manipulation task for a robotic hand and use
-our contact planner to compute a guide trajectory for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
-Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences, this in less than 5 seconds.
+\subsubsection{3-fingered hand -- Manipulation of a pen (Fig.~\ref{fig:penrot}):}
+This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
+our contact planner to compute a guide trajectory for the fingers, considered as effectors (Fig.~\ref{fig:penrot}).
+Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences\deleted{, this} in less than 5 seconds.
 
@@
 
-\noindent\textbf{Heuristics:} $h_{EFORT}$ is used for the all fingers.
+\noindent\textbf{Heuristics:} $h_{EFORT}$ \deleted{is used} for all fingers.
 
@@
 thus allowing to consider interactive applications. 
-Considering the repartition of the computation time, it appears that the majority of the time is spent performing inverse kinematics.
+Considering the repartition of the computation time, it appears that \replaced{most}{the majority} of the time is spent performing inverse kinematics.
 This is not surprising considering the number of calls to the methods: IK is used intensively to maintain contact continuity between two postures; 
-it is also applied every time a new candidate needs to be evaluated. However regarding state of the art IK solutions, we believe there is room
+it is also applied every time a new candidate needs to be evaluated. However regarding state-of-the-art IK solutions, we believe there is room
 for improving the performances of the algorithm.
@@
 We observe that, as expected, our planner does not succeed systematically, because of the approximations made in the implementation
-of our theoritical formulation. 
+of our theoretical formulation. 
 From a pragmatic point of view, regarding the computation times, we claim that our approach is the best current compromise between completeness and efficiency:
-Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to replan a new contact sequence also rapidly with 
+Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to rapidly replan a new contact sequence with 
 a reasonable chance of success.

### 2015-12-22 body/appendix_algo.tex
andrea's modifications

@@
 
-First, we define an abtract structure State,
+First, we define an abstract structure State,
 that describes a contact configuration.
@@
 \end{algorithm}
+
+\section{Derivation of the manipulability ellipsoid}
+\label{app:manipulability}
+
+Again, we assume that $\mathbf{J}$ is full rank. We discard the $k$ indices, and write the pseudo-inverse of $\mathbf{J}$ as $\mathbf{J}^{\dagger}$.
+\begin{eqnarray*}
+\mathbf{\dot{p}} & =  & \mathbf{J} \dot{\mathbf{q}} \\ 
+\mathbf{\dot{q}} & =  & \mathbf{J}^{\dagger} \dot{\mathbf{p}} \\ 
+\mathbf{\dot{q}}^T & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \\ 
+\mathbf{\dot{q}}^T\mathbf{\dot{q}} & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} \dot{\mathbf{p}}\\ 
+\end{eqnarray*}
+
+Then, proving that $\mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} = (\mathbf{J}\mathbf{J}^T)^{-1}$ is straight-forward,
+using the SVD decomposition of each term.

### 2015-12-22 body/conclusion.tex
andrea's modifications

@@
 In this paper we consider acyclic contact planning in cluttered environments, formulated as two sub problems that we address sequentially.
-The first problem is \Pa, that consists in computing a guide path for the root of the robot\deleted{ that can be extended}; the second problem is \Pb, that is the computation of a discrete sequence of contacts along the root path.
+The first problem is \Pa, that consists in computing a guide path for the root of the robot that can be extended; the second problem is \Pb, that is the computation of a discrete sequence of contacts along the root path.
 Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner.
@@
 %
-\textbf{Regarding performance}, our framework outperforms existing methods in addressing either \Pa or \Pb, \adnote{saying you outperform existing methods when you do not show any comparison is risky, the reviewers might ask for comparisons} leading to computation costs close to real-time in
-\replaced{known static}{statically known} environments.
+\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and \Pb. This results in computation costs close to real-time in
+known static environments.
 %
@@
 As for \cite{Bouyarmane2009}, failures can still occur, due to the compromise criterion used in computing the guide path.
-The low computational burden of our framework however allows for fast replanning \replaced{in case of failure}{when the solution fails}.
+The low computational burden of our framework however allows for fast replanning in case of failure.
 
@@
 Our main objective is of course to generate the complete interpolation between the contact sequences.
-We have already been able to achieve this for HRP-2 for some of the computed sequences~\citep{Carpentier2016}, but we \replaced{admit}{concede} that additional
+We have already been able to achieve this for HRP-2 for some of the computed sequences~\citep{Carpentier2016}, but we admit that additional
 work is required on the planning side to obtain a seamless workflow. To achieve this, we are currently working on the notion of transition certificate, i.e. formulating

### 2015-12-22 body/contact.tex
andrea's modifications

@@
 \label{sec:contact}
-As an input of this stage, we consider a truly feasible guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$.
+As an input of this stage, we consider a geometrically feasible guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$.
 We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.
-\deleted[remark={the input is not truly feasible, actually it is not even geometrically feasible, it is only approximately geometrically feasible}]{From true feasibility we know that such a trajectory exists.}
 
@@
 and root location.
-Then, on this method we build an iterative algorithm to generate a discrete sequence of contact configurations in equilibrium\deleted{, along a truly feasible path}.
+Then, on this method we build an iterative algorithm to generate a discrete sequence of contact configurations in equilibrium.
 
@@
 \label{sec:single_contact}
-We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$\deleted{ truly feasible regarding $R^k$}. FIGURE
+We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$. TODO FIGURE
 From true feasibility we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify \eqref{eq:pi} for $\mathbf{q}^0$.
@@
 \begin{enumerate}
-\item Generate randomly a \replaced{collision-free}{valid} limb configuration;
+\item Generate randomly a collision-free limb configuration;
 \item Project the end-effector onto the closest surface with inverse kinematics;
@@
 \item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree stata structure;
-\item Figure~\ref{fig:contact_gen}--2 and 3--:\adnote{This notation for subfigures is not clear, consider using subfigures, e.g.
-2a, 2b} At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
+\item Figure~\ref{fig:contact_gen}--b and c--: At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
 \item According to a user-defined heuristic $h$, sort the elements of $S$;
-\item Figure~\ref{fig:contact_gen}--4 and 5--: Select the first configuration of $S$ (or fail if $S$ is empty). Project the configuration onto contact with inverse kinematics;
-\item If a valid \adnote{being valid depends also on the other contacts because it regards also equilibrium, doesn't it?
-Then this assessment depends on the other contacts previously generated.
-Doesn't this have implications on the completeness?
-Even if it doesn't, isn't it worth to talk about it?} solution is found, stop.
-Otherwise remove the element from $S$ and go back step 5.
+\item Figure~\ref{fig:contact_gen}--d and e--: Select the first configuration of $S$ (or fail if $S$ is empty). Project the configuration onto contact with inverse kinematics;
+\item If a valid solution is found, stop. Otherwise remove the element from $S$ and go back step 5.
 \end{enumerate}
@@
 First, this allows us to select a large number of candidates in a single proximity request.
-Having several candidates is interesting, because it allows to compare them \replaced{using}{regarding} a user-selected heuristic $h$, thus obtaining
+Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
 a locally-optimal candidate.
@@
 
-We thus have defined a real-time\deleted{, extremely efficient} contact generator that can be as complete as required by the context.
-%We compute the image of a mapping $f$ along $\mathbf{q}^0(t)$ into a discrete sequence of contacts $ \in C_{contact}$.
-We first describe the discretization of $\mathbf{q}^0(t)$ and the algorithm used to create or break contacts along it.
-Then we describe how contacts are explicitely generated along the discretized trajectory.
+We thus have defined a real-time contact generator. For a given configuration, the generator can be as complete as required by the context (by increasing the number of samples).
 
@@
 
-As an input, we consider the truly feasible guide path $\mathbf{q}^0(t)$, automatically discretized into a sequence of $j$ key placements:  
+As an input, we consider the guide path $\mathbf{q}^0(t)$, automatically discretized into a sequence of $j$ key placements:  
 \begin{equation*}
@@
 \end{equation*} 
-\adnote{Not clear, what is g?}
+$g$ is the method that extends a root configuration into a full-body configuration. At each step, it tries to generate a contact configuration that preserves as much as possible
+the previous contacts while allowing for static equilibrium. The objective is thus to characterize $g$.
 We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
-
-At each step, we consider a new root placement $\mathbf{q}^0_{i}$ and try to generate a contact configuration that preserves as much as possible
-the previous contacts while allowing for static equilibrium.
 
@@
 
-Fig.~\ref{fig:break_contact} illustrates the contact-persistence strategy.
+Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
 If possible, a limb in contact at time $i-1$ remains in contact at time $i$.
@@
 \begin{enumerate}
-\item Only one contact creation can happen between two consecutive states.
-This increases the odds that the interpolation between the two states \deleted[remark={not 100\% sure}]{will} be feasible; \adnote{Before you said that you ensured that transition between two states is feasible...}
+\item Only one contact creation can happen between two consecutive states. This increases the odds that the interpolation between the two states be feasible;
+%\adnote{Before you said that you ensured that transition between two states is feasible...}
 \item A contact is validated if and only if the resulting configuration is in static equilibrium, no matter what user-defined heuristics are used; 
@@
 These three rules provide a deterministic manner to generate contacts along the discretized guide path.
-FIGURE. For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
+TODO FIGURE. For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
 of the deterministic approach used to break the combinatorial.
-However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:stat1}, thanks to 
+However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}, thanks to 
 the heuristics that we use for contact selection.

### 2015-12-22 body/efort.tex
andrea's modifications

@@
 
-From \eqref{jac} we can obtain the following equality: \adnote{how?}
+We assume that $J$ is full rank (this is reasonable since singular configurations are discarded by our algorithm).
+From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
 \begin{equation} \label{jaceq}
@@
 The longer an axis of the ellipsoid is, the faster the configuration can move along the direction of the axis.
-Fig.~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.
+Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.
 
@@
 \end{equation*}
-is the so-called force ellipsoid (Fig.~\ref{sec:efort_ellipsoid} - right), defined by:
+is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right), defined by:
 \begin{equation} \label{ellipsoidforce}
@@
 \subsubsection{Manipulability-based heuristics.}
-From these definitions, we can derive three useful heuristics, that all account \replaced{for}{from} the environment and the task being performed.
+From these definitions, we can derive three useful heuristics, that all account for the environment and the task being performed.
 The first one, EFORT, was introduced by \cite{Tonneau2014}; the other two are new minor contributions, derived from these previous works.
@@
 $EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
-In other less demanding cases, manipulability can also be considered \deleted{regarding its original purpose, that means} to avoid singularities.
+In other less demanding cases, manipulability can also be considered to avoid singularities.
 To do so, we can consider the manipulability measure $w$, also given by \citeauthor{Yoshikawa1984}:
@@
 \begin{equation*} \label{ellipsoid}
-w(\mathbf{q}) = \sqrt{\mathbf{J}\mathbf{J}^T}
-\end{equation*} \adnote{Didn't you forget something in this equation? I guess $w$ is a scalar, but the right side is a matrix.}
+w(\mathbf{q}) = \sqrt{det(\mathbf{J}\mathbf{J}^T)}
+\end{equation*}
 $w$ measures the ``distance'' between a given configuration and a singular configuration. When $w$ is equal to 0, the configuration is in a singular state;
@@
 \begin{equation}
-h_{w}(\mathbf{q}, \mathbf{m}) = \sqrt{\mathbf{J}\mathbf{J}^T} ( \mu_0 \mathbf{n}^T \mathbf{m})
+h_{w}(\mathbf{q}, \mathbf{m}) = w(\mathbf{q})  (\mu_0 \mathbf{n}^T \mathbf{m})
 \end{equation}
@@
 $h_{EFORT}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
-can be precomputed offline), that allow the planner to select the \replaced{best}{more relevant}
-candidates \replaced{with respect to}{regarding} the local motion task along the guide path.
+can be precomputed offline), that allow the planner to select the best
+candidates with respect to the local motion task along the guide path.
 This results in faster computation times, and a reduced number of overall contact switches.

### 2015-12-22 body/equilibrium.tex
andrea's modifications

@@
 \section{Heuristics for relevant contact selection}
-
+\label{sec:heuristics}
 \subsection{A heuristic for robust static equilibrium}
@@
 From this formulation we derive a new LP that quantifies the robustness of the configuration regarding equilibrium.
-In turn, from this value we can either define a heuristic to choose the most robust candidate, or set a threshold on the required robustness\deleted{for any configuration}.
-While the presented LP is original, it is based on an analysis of the problem that we proposed in \cite{Prete2016}, where the interested reader can find more details.
+In turn, from this value we can either define a heuristic to choose the most robust candidate, or set a threshold on the required robustness.
+While the presented LP is original, it is based on an analysis of the problem that we proposed in \citep{Prete2016}, where the interested reader can find more details.
 
@@
 
-According to Coulomb's law, \replaced{slipping occurs only if a contact force is on the boundaries of the friction cone.}{ the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.}
+According to Coulomb's law, the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
 As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
@@
 can thus be expressed as a convex combination of its four generating rays.
-\deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}
+%~ \deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}
 
@@
 \end{equation*}
-\added{where $\bm{\beta}_i$ contains the coefficient of the cone generators.}
+where $\bm{\beta}_i$ contains the coefficient of the cone generators.
 We can then stack all the constraints to obtain:
@@
 where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.
-The right hand side of \eqref{eq:new_eul} is independent of $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as: \adnote{It is not really necessary to show that equilibrium is independent of the com altitude here, anyway the com is given, not a variable, so it does not change the size of the problem}
-\begin{align} \label{eq:new_eul2d}
-\mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
-\end{align}
+%~ The right hand side of \eqref{eq:new_eul} is independent of $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as: \adnote{It is not really necessary to show that equilibrium is independent of the com altitude here, anyway the com is given, not a variable, so it does not change the size of the problem}
+%~ \begin{align} \label{eq:new_eul2d}
+%~ \mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
+%~ \end{align}
 
-If there exists a $\bm{\beta}^*$ satisfying \eqref{eq:gen} and \eqref{eq:new_eul2d}, it means that the configuration is statically balanced.
-The problem can then be formulated as an LP:
+If there exists a $\bm{\beta}^*$ satisfying \eqref{eq:gen} and \eqref{eq:new_eul}, it means that the configuration is statically balanced.
+The problem can then be formulated as a LP:
 
@@
 \find \quad & \bm{\beta} \in \Rv{4e} \\
-\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
+%~ \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
+\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D} \mathbf{c} + \mathbf{d} \\
 & \bm{\beta} \ge 0 \\
@@
 \minimize  \quad & -b_0 \\
-\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
+\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D} \mathbf{c} + \mathbf{d} \\
+%~ \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
 & \bm{\beta} \ge b_0 \bm{1}\\
@@
 \find \quad & \bm{\nu} \in \Rv{6}\\
-\maximize  \quad & -(\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
+\maximize  \quad & -(\mathbf{D} \mathbf{c} + \mathbf{d})^T \nu \\
 \st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\

### 2015-12-22 body/intro.tex
andrea's modifications

--- 2015-12-09 body/intro.tex
@@
 % !TEX root =  ../main.tex
-We consider the problem of planning the \adnote{an} acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment.
-Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to support \adnote{maintain} the equilibrium of the system.
+We consider the problem of planning an acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment.
+Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to maintain the equilibrium of the system.
 
-Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. But \adnote{don't start sentences with 'but', you can start with 'however' followed by a comma} executing
+Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. However, executing
 this behavior on cluttered environments is dangerous, if not impossible.
@@
 both plan and control[...].  We have seen very few robot planners that are  capable of  generating this  behavior."
-The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of valid \adnote{'valid' is not super clear to me here, I'd rather say 'collision-free'} configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
+The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of collision-free configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
 
 Previous contributions that embrace the combinatorial provide complete approaches, not applicable in practice because they require hours of computations \citep{conf/iser/BretlRLKA04}.
-More recent successes use a local solution, resulting in more reasonable performances \adnote{'performances' is ambiguous, I'd rather use 'computation times'} (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
+More recent successes use a local solution, resulting in more reasonable computation times (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
 Neither global nor local methods present satisfying performances, because planning simultaneously the robot trajectory and the contacts that allow
@@
 This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
-This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely rapidly, \adnote{I don't like having two adverbs ending in 'ly' one after the other, 'extremely fast' sounds better to me}
+This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely fast.
 while preserving a success rate greater than TODO \% in the demonstrated scenarios.
 
-In the remainder of this introduction, we discuss further the current state of the art. This allows us to then \adnote{this 'then' sounds bad here} situate more precisely our contributions.
+In the remainder of this introduction, we discuss further the current state of the art. This allows us to situate more precisely our contributions.
 
@@
 Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}.
-These approaches can intrinsically not \adnote{can not intrinsically?
-are intrinsically not able to?} adapt to new situations or discover complex behaviors in unforeseen contexts.
+These approaches are intrinsically not able to adapt to new situations or discover complex behaviors in unforeseen contexts.
 
 The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper.
-The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: \adnote{planning?
-finding?
-computing?} a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third non trivial \adnote{http://www.thefreedictionary.com/nontrivial} problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
+The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories.
-This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not scalable \adnote{'scalable' means 'capable of being scaled'} to arbitrary environments.
+This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments.
 Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.
@@
 A solution can then be found, at the cost of prohibitive computation times (several hours).
-Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the relevance \adnote{I don't like 'relevance', what about 'quality'?} of the guide trajectory,  which is not evaluated \textit{a priori}.
+Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the quality of the guide trajectory,  which is not evaluated \textit{a priori}.
 \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body contact configurations in static equilibrium.
@@
 It thus appears that an integrated optimization-based approach \citep{Mordatch:2012:DCB:2185520.2185539}, able to generate a locally-optimal, complete trajectory within minutes, currently outperforms
-existing planners, decoupled or not. However, recent contributions are able to interpolate dynamically-executable \adnote{dynamically-feasible?}
+existing planners, decoupled or not. However, recent contributions are able to interpolate dynamically-feasible
 trajectories between contact configurations \citep{herzog2015trajectory, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real-time performances.
@@
 This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
-we believe that the separation between the \adnote{generation of the} guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
+we believe that the separation between the generation of the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
 However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
 \begin{itemize}
-\item The guide trajectory must satisfy a property guaranteeing \adnote{replace 'satisfy a property guaranteeing' with 'guarantee'} the existence of a contact sequence to actuate it\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties.}.
+\item The guide trajectory must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces).
 We call this property \textit{true feasibility}.
-This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
-\item There is an infinite combination \adnote{There are infinite combinations} of possible contact sequences for a given root trajectory.
-The selection of one particular contact sequence with interesting properties (minimum number of contact change, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious contact sequences \adnote{repetition of 'contact sequence'}).
+This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
+\item There are infinite combinations of possible contact sequences for a given root trajectory.
+The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious transitions.
 \end{itemize}
@@
 Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
-The relaxation is justified empirically, with statistics presented for each robot and scenario presented\adnote{several robots and scenario, or even better, for N robots and M scenarios}.
-The complete framework is implemented within
-the open source platform HPP.
+The relaxation is justified empirically, with statistics presented for 2 robots and 4 scenarios. The complete framework is implemented within
+the open-source platform HPP.
 
@@
 
-To address \Pa we rigorously formulate a geometric condition for the true feasibility of a path, the reachability condition. We claim that this 
-condition can be verified in a low-dimensional space, that we call $C_{reach}$.
-This formulation can make the planning of a guide trajectory more efficient computationally, while providing equivalent guarantees to planning directly in the configuration space.
+To address \Pa we rigorously formulate a geometric condition that approximates the true feasibility of a path, the reachability condition. We claim that this 
+condition can be verified in a low-dimensional space, which we call $C_{reach}$.
+This formulation can make the planning of a guide path more efficient computationally, while providing similar guarantees to planning directly in the configuration space.
 In practice, for each considered robot, we compute an approximation of $C_{reach}$. From this approximation,
 an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained.
-This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot within a few milliseconds to a few seconds depending on the scenario.
+This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot in a timespan ranging from a few milliseconds to a few seconds, depending on the scenario.
 
@@
 limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
-for proximity queries. At runtime, this framework allows to obtain extremely rapidly \adnote{again, extremely rapidly} a set of candidate configurations for contact.
-These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, and ranked according to user-defined heuristics.
+for proximity queries. At runtime, this framework allows to obtain rapidly a set of candidate configurations for contact.
+These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, as well as a set of user-defined heuristics.
 In all our experiments we used two such heuristics: a measurement of robustness of a configuration to errors in the contact forces, and a manipulability-based heuristic,
@@
 
-In comparison with previous work, \adnote{Compared to previous work/Contrary to previous work [cite Mordatch,...]} our planner does not produce a complete motion, but a discrete sequence of contacts.
+Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
 Since the interpolation of a complete trajectory between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
@@
 The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
-This extension completes this work with a \adnote{I remove the word 'argumented' because it does not exist and I've not been able to find a good replacement} discussion on the implementation of this theory to real-world robots and problems.
+This extension completes this work with a discussion on the implementation of this theory to real-world robots and problems.
 In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot despite bounded error in the contact forces. We also provide details on the HPP implementation
-of our contact generation algorithm. Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ\adnote{cite Semini}. TODO Appendix describes the generation process
+of our contact generation algorithm.
+Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ\adnote{cite Semini}.
+Appendix~\ref{app:rom} describes the generation process
 for the robot.
@@
 In Section~\ref{overview}, we present the general organization of our method.
-Section~\ref{rbprm} and Section~\ref{sec:efort} present respectively our answer to \adnote{I remove 'problems'} \Pa and \Pb.
-In Section~\ref{heuristics}, we present two heuristics used for the selection of \adnote{I remove 'the best'} a contact configuration.
+Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb.
+In Section~\ref{sec:heuristics}, we present two heuristics used for the selection of a contact configuration.
 Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,

### 2015-12-22 body/overview.tex
andrea's modifications

@@
   \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
-		\put (1,1) {1} 
-		\put (22,1) {2} 
-		\put (42,1) {3} 
-		\put (62,1) {4} 
-		\put (83,1) {5} 
+		\put (1,1) {a} 
+		\put (22,1) {b} 
+		\put (42,1) {c} 
+		\put (62,1) {d} 
+		\put (83,1) {e} 
 	\end{overpic}
   \caption{Generation of a contact configuration for the right leg of HRP-2.
-1: Selection of reachable obstacles.
-2: Entries of the limb samples database (with $N = 4$).
-3: With a proximity query on the octree database, configurations too far from obstacles are eliminated.
-4: The best candidate according to a user-defined heuristic $h$ is chosen.
-5: The final contact is achieved using inverse kinematics.}
+a: Selection of reachable obstacles.
+b: Entries of the limb samples database (with $N = 4$).
+c: With a proximity query on the octree database, configurations too far from obstacles are eliminated.
+d: The best candidate according to a user-defined heuristic $h$ is chosen.
+e: The final contact is achieved using inverse kinematics.}
   \label{fig:contact_gen}
@@
 
-\added
-{
-More precisely, a root placement is truly feasible if and only if there exist at least one joint configuration such that the robot satisfies three conditions: }
+More precisely, a root placement is truly feasible if and only if there exists at least one joint configuration such that the robot satisfies three conditions:
 \begin{description}
@@
 \end{description}
-\added{
 The conditions C1 and C2 are purely geometrical and can be efficiently approximated even without computing the contact points.
-Instead C3 is difficult to approximate and it requires the computation of the contact points, and the resolution of a linear program to compute the contact forces~\cite{DelPrete2016}.
+Instead C3 is difficult to approximate and it requires the computation of the contact points, and the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
 Moreover our intuition suggests us that, for cluttered environment, C3 is not critical, meaning that, most of the times, root placements satisfying C1 and C2 would satisfy also C3.
 For this reason we neglect C3 during the resolution of \Pa, and we worry about it only when solving \Pb.
-}
 
-An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Fig.~\ref{fig:contact_gen}--1); not too close, because the robot must avoid collision (which is represented by the hull including the torso in Fig.~\ref{fig:contact_gen}--1).
-We define formally $C_{reach}$, the set of \replaced{root placements satisfying C1 and C2}{interesting root placements}, in which we compute a guide trajectory with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Fig.~\ref{fig:framework}---\Pa).
+An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
+We define formally $C_{reach}$, the set of root placements satisfying C1 and C2, in which we compute a guide trajectory with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
 Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
@@
 
-The second stage extends the guide path into a sequence of contact configurations (Fig.~\ref{fig:framework}--\Pb).
-\deleted{The true feasibility of the input path is guaranteed by the first step.
-This provides in turn a reasonable guarantee that a contact configuration can be obtained for any root placement along the path.
-}
+The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--\Pb). In this
+step we assume that the input guide path is truly feasible.
 With a fixed root position, the dimensionality of the 
@@
 In such a case the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot (TODO FIGURE).
-The algorithm is designed so that only one contact can be created or broken between two successive configurations.
-This \replaced{ensures}{can be seen as a heuristic to ensure} that the interpolation between two configurations is achievable.
-Details are presented in Section~\ref{sec:efort}.
+The algorithm is designed so that only one contact can be created or broken between two successive configurations. This ensures that the interpolation between two configurations is achievable.
+Details are presented in Section~\ref{sec:contact}.
 
 To select the ``most relevant'' contact configurations, user-defined heuristics can be chosen.
-In Section~\ref{heuristics} we present extensively two such heuristics, which have been used in our experiments.
+In Section~\ref{sec:heuristics} we present extensively two such heuristics, which have been used in our experiments.
 
@@
 \begin{itemize}
-	\item $\mathbf{q}^k$ denotes the configuration (a vector of joint values) of the limb $R^k$ (Fig.~\ref{fig:character});
+	\item $\mathbf{q}^k$ denotes the configuration (a vector of joint values) of the limb $R^k$ (Figure~\ref{fig:character});
 	\item $\mathbf{q}^{\overline{k}}$ denotes the vector of joint values of R \textbf{not} related to $R^k$.
@@
 \medskip
-The volume encompassing the trunk of the robot is denoted $W^0$ (Fig.~\ref{fig:character}-right: central cylinder).
-The reachable workspace of a limb $R^k$ is denoted $W^k$ (Fig.~\ref{fig:character}-right: the four ellipses).
+The volume encompassing the trunk of the robot is denoted $W^0$ (Figure~\ref{fig:character}-right: central cylinder).
+The reachable workspace of a limb $R^k$ is denoted $W^k$ (Figure~\ref{fig:character}-right: the four ellipses).
 \begin{equation}
-  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\},
+  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k \in C^k_{jl}, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\},
 \end{equation}
-\adnote{in this definition you don't account for joint limits}
-where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0$ being the null displacement. We also define $W = \bigcup_{k=1}^{l}W^k$, and
+%~ \adnote{in this definition you don't account for joint limits}
+where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0$ being the null displacement, and  $C^k_{jl}$ is the space
+of admissible limb joint configurations. We also define $W = \bigcup_{k=1}^{l}W^k$, and
 $W^k(\mathbf{q}^{0})$ (for $1 \leq k \leq l$) as the volume $W^k$ translated and rotated by the rigid displacement~$\mathbf{q}^{0}$.

### 2015-12-22 body/rbprm.tex
andrea's modifications

@@
 
-We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the geometrical \adnote{why the root is geometrical?
-Isn't the path to be geometrical?} root of a multiped robot, connecting \adnote{user-defined} start and goal configurations.
-As said in the previous section, the goal is to enforce that any configuration $\mathbf{q}^0$ of the guide is \replaced{geometrically}{truly} feasible, i.e.
+We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of a multiped robot, connecting user-defined start and goal configurations.
+As said in the previous section, the goal is to enforce that any configuration $\mathbf{q}^0$ of the guide is geometrically feasible, i.e.
 can be mapped to a configuration in contact.
-We denote by $C_{contact} \subset SE(3) \times \mathbb{R}^n$ the contact submanifold of the robot\added{, that is the set of configurations in which the robot is in contact with the environment}.
+We denote by $C_{contact} \subset SE(3) \times \mathbb{R}^n$ the contact submanifold of the robot, that is the set of configurations for which the robot is in contact with the environment.
 %~ We say that a root placement $\mathbf{q}^{0}$ is \textit{truly feasible} if there exists $\mathbf{q}^{\overline{0}}$ such that 
@@
 
-We say that a root placement $\mathbf{q}^{0}$ is \textit{\replaced[remark={I don't replace 'truly feasible' with 'geometrically feasible' from now on because I don't know whether you like it}]{geometrically}{truly} feasible} if there exists at least one bijective mapping $\pi$
+We say that a root placement $\mathbf{q}^{0}$ is geometrically feasible if there exists at least one bijective mapping $\pi$
 such that
@@
 
-The set of all truly feasible root placements is denoted by $C_{reach}$. By extension, a path $\mathbf{q}^0(t)$ is truly feasible if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{reach}$.
+The set of all geometrically feasible root placements is denoted by $C_{reach}$. By extension, a path $\mathbf{q}^0(t)$ is truly feasible if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{reach}$.
 
-For a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide trajectory generator: all the generated trajectories must be \textit{truly feasible} (sufficient condition); the guide planner must be complete, i.e.
+For a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide trajectory generator: all the generated trajectories must be \textit{geometrically feasible} (sufficient condition); the guide planner must be complete, i.e.
 it must be able to discover any truly feasible trajectory (necessary condition).
@@
 
-\subsection{Conditions for true feasibility}
-By default, the true feasibility implies a constructive demonstration by exhibiting a valid $\pi$.
+\subsection{Conditions for geometric feasibility}
+By default, the geometric feasibility implies a constructive demonstration by exhibiting a valid $\pi$.
 This is the approach chosen by \cite{Bouyarmane2009}.
 However, computing a valid  $\mathbf{q}^{\overline{0}}$ is costly in practice.
-In this section we rather define a necessary condition and a sufficient condition for true feasibility that do not require this explicit computation.
+In this section we rather define a necessary condition and a sufficient condition for geometric feasibility that do not require this explicit computation.
 
@@
 
-\subsubsection*{True feasibility: necessary condition}
-For a contact to be possible, a volume $O_i \in O$ necessarily intersects with the reachable workspace $W(\mathbf{q}^{0})$ (Fig.~\ref{fig:contact_gen}--1).
-Furthermore, if $\mathbf{q}^{0}$ is truly feasible, then the trunk of the robot $W^0(\mathbf{q}^{0})$ is necessarily not colliding  with the environment $O$.
+\subsubsection*{Geometric feasibility: necessary condition}
+For a contact to be possible, a volume $O_i \in O$ necessarily intersects with the reachable workspace $W(\mathbf{q}^{0})$ (Figure~\ref{fig:contact_gen}--1).
+Furthermore, if $\mathbf{q}^{0}$ is  feasible, then the trunk of the robot $W^0(\mathbf{q}^{0})$ is necessarily not colliding  with the environment $O$.
 
@@
 
-\subsubsection*{\replaced[remark={for coherence with previous subsection title}]{True feasibility: sufficient condition}{Sufficient condition for true feasibility}}
+\subsubsection*{Geometric feasibility: sufficient condition}
 %~ In building $C_{reach}^1$, we directly considered the including hull of the root body $W^{0}$, obtaining a necessary condition. 
-A trivial sufficient condition for true feasibility can be constructed as a variation of $C_{reach}^1$, by replacing $W^0$ with a bounding volume $B^{max}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact (This condition is trivial in the sense that the resulting $W$ has a zero measure.
-For the need of the proof, the trivial sufficient condition is enough.
-In practice, the construction of a non-trivial including shape $W^0$ was possible for all the robot structures we considered).
-\adnote{Isn't this parenthesis a bit too long?} We denote by \mbox{$C_{reach}^\infty \subset C_{reach}$} the set of root placements corresponding to the sufficient condition.
+A trivial sufficient condition for geometric feasibility can be constructed as a variation of $C_{reach}^1$, by replacing $W^0$ with a bounding volume $B^{max}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
+We denote by \mbox{$C_{reach}^\infty \subset C_{reach}$} the set of root placements corresponding to the sufficient condition.
 
@@
 
-Thanks to this small set of modifications, the high dimensional truly feasible path planning problem is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for Hyq, 8 for HRP-2 that has \replaced{joints in the}{a flexible} torso).
-\replaced{By doing do}{In so doing} we make the problem tractable for sampling-based approaches, with extremely efficient computation times.
+Thanks to this small set of modifications, the high dimensional truly feasible path planning problem is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for Hyq, 8 for HRP-2 that has joints in the torso).
+In so doing we make the problem tractable for sampling-based approaches, with extremely efficient computation times.

### 2015-12-22 body/results.tex
andrea's modifications

@@
 In this Section we present some of the results obtained with our planner. The complete sequences computed are shown in the companion video, TODO.
-Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ~\cite{Semini}.
+Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ TODO~\cite{Semini}.
 For each scenario we indicate the chosen heuristics. We also provide a performance analysis, which shows that the planner is compatible with interactive applications,
-and present the success rates obtained in each scenario.
-Moreover, we demonstrate the interest of our robustness criterion in the different computed poses.
+and present the success rates obtained in each scenario. Moreover, we demonstrate the interest of our robustness criterion in the different computed poses.
 Finally, a last example suggests possible applications to dextrous manipulation.
 
-In our previous work~\citep{tonneauisrr15} additional results are demonstrated with various virtual avatars (Fig.~\ref{fig:robots_old}).
-In this extension we choose to focus on \deleted{results obtained with} actual robots.
-We invite the interested reader to watch the ISRR video\endnote{https://www.youtube.com/watch?v=LmLAHgGQJGA}, and
+In our previous work~\citep{tonneauisrr15} additional results are demonstrated with various virtual avatars (Figure~\ref{fig:robots_old}).
+In this extension we choose to focus on actual robots. We invite the interested reader to watch the ISRR video\endnote{https://www.youtube.com/watch?v=LmLAHgGQJGA}, and 
 to refer to the previous paper for a discussion on these results.
+
+We say that the planning is interactive when the computation time for one step is lesser than the
+time to execute it. We arbitrarily approximate this time to one second.
 
@@
 
-The goal is to climb three 20-cm high steps. This height requires HRP-2 to use a ramp to \deleted{be able to} perform the task.
+The goal is to climb three 20-cm high steps. This height requires HRP-2 to use a ramp to perform the task.
 
-\noindent\textbf{Contacts involved:} \deleted{Both} Feet and right arm.
+\noindent\textbf{Contacts involved:} Feet and right arm.
 
 \noindent\textbf{Heuristics:} The extended manipulability $h_w$ is chosen for the feet; $h_{EFORT}$ is chosen for the right arm.
-Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Fig.~\ref{fig:stair_robust}).
+Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
 
@@
 This scenario illustrates best the importance of the equilibrium-robustness criterion.
-With a robust approach, more states are required to reach the last step (15 \replaced{rather than}{against} 13 in average).
+With a robust approach, more states are required to reach the last step (15 rather than 13 in average).
 However, when the last step is reached by both feet, in the nonrobust case the contacts are extremely close to 
-the cone limits\deleted{, and we believe this can be seen visually} (Fig.~\ref{fig:stair_comp}).
+the cone limits (Figure~\ref{fig:stair_comp}).
 
@@
 
-\subsubsection{HRP-2 -- Standing up (Fig.~\ref{fig:standing}):}
+\subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing}):}
 From a bent configuration, a standing-up motion is computed in a cluttered environment.
@@
 
-\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for the feet, $h_{EFORT}$ \deleted{is used} for the hands.
+\noindent\textbf{Heuristics:} $h_w$ for the feet, $h_{EFORT}$  for the hands.
 
-\noindent\textbf{Observations:} The scenario illustrates well the acyclic aspect of the planning. For instance, in the four first frames of Fig.~\ref{fig:standing}, we can see that the right foot
+\noindent\textbf{Observations:} The scenario illustrates well the acyclic aspect of the planning. For instance, in the four first frames of Figure~\ref{fig:standing}, we can see that the right foot
 is moved twice, with the left foot in between, before the configuration allows HRP-2 to move its hand.
@@
 for the computed path. The reason for this is not reachability, but equilibrium. The planning is slower than for the stair scenario (because the contact generation fails more),
-though it remains compatible with interactive performances. \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
+though it remains compatible with interactive performances. % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
 
-\subsubsection{HRP-2 -- Truck egress (Fig.~\ref{fig:TODO}):}
-HRP-2 stands half seated in a truck, and \replaced{is required to}{it must} \replaced{get out}{exit} through the front window.
+%~ \subsubsection{HRP-2 -- Truck egress (Figure~\ref{fig:TODO}):}
+%~ HRP-2 stands half seated in a truck, and \replaced{is required to}{it must} \replaced{get out}{exit} through the front window.
 
@@
 
-\noindent\textbf{Contacts involved:} All (both feet and hands).
+%~ 
+%~ \noindent\textbf{Contacts involved:} All (both feet and hands).
+%~ 
+%~ \noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for the feet, $h_{EFORT}$ \deleted{is used} for the hands.
+%~ 
+%~ \noindent\textbf{Observations:} This scenario demonstrates the ability of the planner to compute a path in an extremely-constrained environment. In this case,
+%~ the scenario is not always real time (TODO or is it ?): while the planner is often able to compute a guide path rapidly, in the worst case up to one minute can be spent
+%~ in the planning.
 
-\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for the feet, $h_{EFORT}$ \deleted{is used} for the hands.
-
-\noindent\textbf{Observations:} This scenario demonstrates the ability of the planner to compute a path in an extremely-constrained environment. In this case,
-the scenario is not always real time (TODO or is it ?): while the planner is often able to compute a guide path rapidly, in the worst case up to one minute can be spent
-in the planning.
-
-\subsubsection{HyQ -- Darpa-style rubble (Fig.~\ref{fig:darpa})}
+\subsubsection{HyQ -- Darpa-style rubble (Figure~\ref{fig:darpa})}
 The quadruped robot is given the task to cross a rubble composed of bricks rotated at different angles and directions.
@@
 
-\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for all legs. The robustness threshold $b_0$ is set to $20$.
+\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $20$.
 
@@
 The path-planning time is higher than for the previous HRP-2 robots due to the constraint that the 4 reachable workspace of all legs must
-collide with the environment at all times. \adnote{Why do not relax this constraint and require only 3 legs?} Again, the computed time remains however interactive.
+collide with the environment at all times. %\adnote{Why do not relax this constraint and require only 3 legs?} 
+Again, the computed time remains however interactive.
 
 \subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:hyq_bridge} and~\ref{fig:hyq_obs}):}
-In this long scene, HyQ is first required to cross a 55-cm large hole; then, to cross a narrow "bridge",  only 25-cm large\deleted{,
-while maintaining equilibrium}.
+In this long scene, HyQ is first required to cross a 55-cm large hole; then, to cross a narrow "bridge",  only 25-cm large.
 
@@
 
-\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for all legs. The robustness threshold $b_0$ is set to $4$.
+\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $4$.
 
-\noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is the hardest \deleted{to compute} for our planner.
-While finding a guide path above the hole is easy for the planner, finding a \deleted{relevant} sequence of contacts that \replaced{allows for}{remain in} equilibrium is not trivial.
+\noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is the hardest for our planner.
+While finding a guide path above the hole is easy for the planner, finding a sequence of contacts that allows for equilibrium is not trivial.
 Second, the narrow bridge is hard both for the planner and the contact generator: to make sure that equilibrium is preserved along the traversal,
 the bridge must be approached with the right angle.
-This can be seen in Fig.~(TODO), where several feet \replaced{rearrangements}{repositioning} are required to cross both obstacles (although the video shows this best).
+This can is illustrated in Figure~\ref{fig:hyq_obs}, where several feet rearrangements are required to cross both obstacles (although the video shows this best).
 The planner however succeeds in finding a feasible sequence in the end, again with interactive computation times.
@@
 In this long scene, HRP-2 plans a path through through several obstacles. the scene is edited during the execution of the motion: a stair is added,
-some stepping stones are removed, and part of the final staircase is deleted. All these modifications \replaced{demands}{involve} replanning.
+some stepping stones are removed, and part of the final staircase is deleted. All these modifications require replanning.
 
@@
 
-\noindent\textbf{Contacts involved:} \replaced{F}{Both f}eet and the right arm.
+\noindent\textbf{Contacts involved:} Feet and the right arm.
 
-\noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for all legs. $h_{EFORT}$ \deleted{is used} for the right arm. The robustness threshold is set to $2$.
+\noindent\textbf{Heuristics:} $h_w$  for all legs. $h_{EFORT}$  for the right arm. The robustness threshold is set to $2$.
 
@@
 
-\subsubsection{3-fingered hand -- Manipulation of a pen (Fig.~\ref{fig:penrot}):}
+\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot}):}
 This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
-our contact planner to compute a guide trajectory for the fingers, considered as effectors (Fig.~\ref{fig:penrot}).
-Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences\deleted{, this} in less than 5 seconds.
+our contact planner to compute a guide trajectory for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
+Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences in less than 5 seconds.
 
@@
 
-\noindent\textbf{Heuristics:} $h_{EFORT}$ \deleted{is used} for all fingers.
+\noindent\textbf{Heuristics:} $h_{EFORT}$ for all fingers.
 
@@
 
-\subsection{Computation times}
+\subsubsection{Computation times}
 Table~\ref{tab:requestime} summarizes the performance measurement obtained, in terms of computation times.
@@
 thus allowing to consider interactive applications. 
-Considering the repartition of the computation time, it appears that \replaced{most}{the majority} of the time is spent performing inverse kinematics.
+Considering the repartition of the computation time, it appears that most of the time is spent performing inverse kinematics.
 This is not surprising considering the number of calls to the methods: IK is used intensively to maintain contact continuity between two postures; 
@@
 
-\subsection{Success rates}
+\subsubsection{Success rates}
 Table~\ref{tab:requestpercent} summarizes the success rates obtained for each scenario.

### 2015-12-22 body/abstract.tex
andrea's modifications
typos

@@
 of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
-Then, we propose a sequential algorithm to extend a truly feasible path into a sequence of contacts. For each limb, an offline sampling approach,
+Then, we propose a sequential algorithm to extend a truly feasible path into a sequence of contacts. For each limb, an off-line sampling approach,
 independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
-The best candidate is chosen according to user-created \adnote{user-defined?} heuristics. We propose two such heuristics, used in all our experiments, and 
+The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
 designed for generic robotic applications: a manipulatibily-based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
 These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
-variety of robots, including a HRP-2, Hyq, and a dextrous hand, with interactive performances.
+variety of robots, including a HRP-2, Hyq, and a dexterous hand, with interactive performances.

### 2015-12-22 body/appendix_hpp.tex
typos

@@
 Contrary to the mature HPP, at the time of submitting this article,
-our planner is not officialy released, mostly for lack of a complete documentation and a tutorial.
+our planner is not officially released, mostly for lack of a complete documentation and a tutorial.
 

### 2015-12-22 body/conclusion.tex
typos

@@
 We introduced a low-dimensional space $C_{reach}$ that can be mapped
-into the contact submanifold of the robot, approximated, and efficiently sampled by our Reachability-Based planner.
+into the contact sub-manifold of the robot, approximated, and efficiently sampled by our Reachability-Based planner.
 Our contribution to \Pb is a pragmatic contact generation scheme that can take into
@@
 As for \cite{Bouyarmane2009}, failures can still occur, due to the compromise criterion used in computing the guide path.
-The low computational burden of our framework however allows for fast replanning in case of failure.
+The low computational burden of our framework however allows for fast re-planning in case of failure.
 

### 2015-12-22 body/efort.tex
typos

@@
 $h_{EFORT}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
-can be precomputed offline), that allow the planner to select the best
+can be precomputed off-line), that allow the planner to select the best
 candidates with respect to the local motion task along the guide path.

### 2015-12-22 body/intro.tex
typos

@@
 The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper.
-The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
+The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories.
@@
 \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body contact configurations in static equilibrium.
-Randomly sampled configurations are projected into the contact submanifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide trajectory in the examples presented).
+Randomly sampled configurations are projected into the contact sub-manifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide trajectory in the examples presented).
 Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the trajectory.
@@
 The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed).
-The theoretical completeness of the approach is partially traded for efficiency: instead of sampling online
-limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
+The theoretical completeness of the approach is partially traded for efficiency: instead of sampling on-line
+limb configurations and projecting them onto contact, we precompute off-line a large database of limb configurations, stored in a spatial structure optimized
 for proximity queries. At runtime, this framework allows to obtain rapidly a set of candidate configurations for contact.

### 2015-12-22 body/rbprm.tex
typos

@@
 can be mapped to a configuration in contact.
-We denote by $C_{contact} \subset SE(3) \times \mathbb{R}^n$ the contact submanifold of the robot, that is the set of configurations for which the robot is in contact with the environment.
+We denote by $C_{contact} \subset SE(3) \times \mathbb{R}^n$ the contact sub-manifold of the robot, that is the set of configurations for which the robot is in contact with the environment.
 %~ We say that a root placement $\mathbf{q}^{0}$ is \textit{truly feasible} if there exists $\mathbf{q}^{\overline{0}}$ such that 
@@
 It might be possible to find a shape $B$ that is necessary and sufficient; however, it seems intuitively very unlikely in general.
-The construction of a shape $W^0 \subset B \subset B^{max}$ leading to a necessary and sufficient condition (or the proof of its inexistence) is out of the scope of this work.
+The construction of a shape $W^0 \subset B \subset B^{max}$ leading to a necessary and sufficient condition (or the proof of its nonexistence) is out of the scope of this work.
 

### 2015-12-22 body/results.tex
typos

@@
 and present the success rates obtained in each scenario. Moreover, we demonstrate the interest of our robustness criterion in the different computed poses.
-Finally, a last example suggests possible applications to dextrous manipulation.
+Finally, a last example suggests possible applications to dexterous manipulation.
 
@@
 
-\subsubsection{HRP-2 -- Path replanning(Figure~\ref{fig:replanning}):}
+\subsubsection{HRP-2 -- Path re-planning(Figure~\ref{fig:re-planning}):}
 In this long scene, HRP-2 plans a path through through several obstacles. the scene is edited during the execution of the motion: a stair is added,
-some stepping stones are removed, and part of the final staircase is deleted. All these modifications require replanning.
+some stepping stones are removed, and part of the final staircase is deleted. All these modifications require re-planning.
 
@@
   \caption{
-           HRP-2 in the replanning scenario. After the red step stones are removed, a new sequence of contacts is replanned. Hand contacts
+           HRP-2 in the re-planning scenario. After the red step stones are removed, a new sequence of contacts is re-planned. Hand contacts
            are not presented here for readability.}
-		   \label{fig:replanning}
+		   \label{fig:re-planning}
 \end{figure}
@@
  \end{tabular}
-\caption{Success rates for each scenario. The first value indicates the percentage of sucessfull complete path plannings for 1000 tests; The second value
+\caption{Success rates for each scenario. The first value indicates the percentage of successfull complete path plannings for 1000 tests; The second value
 indicates the percentage of truly feasible root configurations: considering each limb individually, indicates the percentage of root placements of the guide trajectory that
@@
 From a pragmatic point of view, regarding the computation times, we claim that our approach is the best current compromise between completeness and efficiency:
-Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to rapidly replan a new contact sequence with 
+Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan a new contact sequence with 
 a reasonable chance of success.

### 2015-12-23 body/abstract.tex
geometrical feasibility

@@
 These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
-variety of robots, including a HRP-2, Hyq, and a dexterous hand, with interactive performances.
+variety of robots, including HRP-2, Hyq, and a dexterous hand, with interactive performances.

### 2015-12-23 body/contact.tex
typos
geometrical feasibility

@@
 \label{sec:contact}
-As an input of this stage, we consider a geometrically feasible guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$.
+As an input of this stage, we consider a \textit{geometrically feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$.
 We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.
@@
 of the robot. The first two tests are trivial; our criterion to assert efficiently the static equilibrium of the system
-is described in Section TODO.
+is described in Section~\ref{sec:heuristics}.
 
@@
 \label{sec:single_contact}
-We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$. TODO FIGURE
-From true feasibility we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify \eqref{eq:pi} for $\mathbf{q}^0$.
+We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$.
+From \textit{geometrical feasibility} we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify \eqref{eq:pi} for $\mathbf{q}^0$.
 The issue is then to evaluate locally at least one element of $P$ at $\mathbf{q}^0$.
@@
 \begin{enumerate}
-\item Generate offline $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
+\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
 \item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree stata structure;
@@
 Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
-Regarding completeness, it is immediate to verify that as $N$ and $\epsilon$ grow, the completeness of the method is restored.
+Regarding completeness, it is immediate to verify that as $N$ grows, the completeness of the method is restored.
 $N$ is a parameter that allows us to specify the trade-off between completeness and efficiency.
@@
 These three rules provide a deterministic manner to generate contacts along the discretized guide path.
-TODO FIGURE. For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
+For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
 of the deterministic approach used to break the combinatorial.

### 2015-12-23 body/efort.tex
geometrical feasibility

@@
 
-We assume that $J$ is full rank (this is reasonable since singular configurations are discarded by our algorithm).
+We assume that $\mathbf{J}$ is full rank (we are not interested in singular configurations that we discard).
 From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
@@
 \end{equation*}
-is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right), defined by:
+is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right):
 \begin{equation} \label{ellipsoidforce}

### 2015-12-23 body/equilibrium.tex
typos
geometrical feasibility

@@
 	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
-	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
+	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local Cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
 	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
@@
 Any force belonging to the linearized cone
-can thus be expressed as a convex combination of its four generating rays.
+can thus be expressed as a positive combination of its four generating rays.
 %~ \deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}

### 2015-12-23 body/intro.tex
geometrical feasibility

@@
 The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper.
-The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
+The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: Planning a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories.
@@
 
-\subsection{Paper contribution and organization}
+\subsection{Rationale}
 This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
 we believe that the separation between the generation of the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
-However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
-\begin{itemize}
-\item The guide trajectory must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces).
+However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated. \\
+
+\noindent \textbf{Regarding $\mathcal{P}_1$}, the guide trajectory must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces).
 We call this property \textit{true feasibility}.
 This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
-\item There are infinite combinations of possible contact sequences for a given root trajectory.
-The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious transitions.
+\\
+
+\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinite combinations of possible contact sequences for a given root trajectory.
+The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious transitions).
+\\
+
+%~ In this paper, we address both questions.
+%~ For each problem, we introduce an original formulation of the related theoretical issues.
+%~ Then, we propose a concrete implementation of a solution, where the constraints of the problem are relaxed in favor of efficiency.
+
+The key idea of the paper is the theoritical and practical characterization of the space of \textit{truly feasible} configurations to address both issues.
+We first exhibit a low dimensional over-approximation of this space. This approximation is defined geometrically. 
+We make the strong assumption that these \textit{geometrically feasible} paths can be efficiently transformed into \textit{truly feasible} paths
+in cluttered environments. 
+Under this assumption we can take advantage of the decoupling approach.
+
+We first address \Pa by planning \textit{geometrically feasible} paths. We achieve this efficiently thanks to the \textit{reachability condition} that we introduce, that 
+does not require explicit contact computation.
+
+Then we address \Pb by extending the path into a \textit{truly feasible} sequence of contact configurations.
+This second step requires computing contact configurations verifying a condition for \textit{static equilibrium feasibility}, that requires
+explicit contact computation.
+
+This sequential approach is the key to the efficiency of our method, though it can
+can result in failures because our assumption is not always true.
+However, we demonstrate empirically the validity of the approach: the high success rate of our method, combined with the low computation times we obtain, allows us to plan (and re-plan upon failure) multi contact sequences at interactive rates.
+
+\subsection{Paper contribution and organization}
+
+By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
+\begin{itemize}
+\item First, we propose a theoretical characterization of an efficient approach to sampling-based planning of acyclic contacts;
+\item Then we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with interactive applications;
+\item Finally we propose three heuristics for contact generation. They bias the planner towards configurations that are robust regarding static equilibrium, or 
+efficient regarding the task being performed.
 \end{itemize}
 
-By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
-For each problem, we introduce an original formulation of the related theoretical issues.
-Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
-The relaxation is justified empirically, with statistics presented for 2 robots and 4 scenarios. The complete framework is implemented within
-the open-source platform HPP.
+Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
+Since the interpolation of a complete trajectory between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
+is the fastest multi-contact planning solution to our knowledge.
 
@@
 
-To address \Pa we rigorously formulate a geometric condition that approximates the true feasibility of a path, the reachability condition. We claim that this 
-condition can be verified in a low-dimensional space, which we call $C_{reach}$.
-This formulation can make the planning of a guide path more efficient computationally, while providing similar guarantees to planning directly in the configuration space.
-In practice, for each considered robot, we compute an approximation of $C_{reach}$. From this approximation,
-an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained.
-This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot in a timespan ranging from a few milliseconds to a few seconds, depending on the scenario.
-
-To address \Pb, we consider a truly-feasible path as an input, and generate new contact configurations along it using a sampling approach.
-The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed).
-The theoretical completeness of the approach is partially traded for efficiency: instead of sampling on-line
-limb configurations and projecting them onto contact, we precompute off-line a large database of limb configurations, stored in a spatial structure optimized
-for proximity queries. At runtime, this framework allows to obtain rapidly a set of candidate configurations for contact.
-These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, as well as a set of user-defined heuristics.
-In all our experiments we used two such heuristics: a measurement of robustness of a configuration to errors in the contact forces, and a manipulability-based heuristic,
-the Extended FORce Transmission ratio (EFORT).
-
-The contribution of the paper is thus twofold.
-We propose a theoretical characterization of today's most efficient practical approach to sampled-based planning of acyclic contacts.
-Based on this characterization, we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with interactive applications.
-
-Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
-Since the interpolation of a complete trajectory between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
-is the fastest multi-contact planning solution to our knowledge.
-
-The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep(tonneauisrr15).
+In Section~\ref{overview}, we present the general organization of our method.
+Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb.
+In Section~\ref{sec:heuristics}, we present two heuristics used for the selection of a contact configuration.
+Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
+that complete those presented in the conference paper.
+\subsection*{Comparison with our previous work.}
+The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep{tonneauisrr15}.
 The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
@@
 for the robot.
-
-In Section~\ref{overview}, we present the general organization of our method.
-Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb.
-In Section~\ref{sec:heuristics}, we present two heuristics used for the selection of a contact configuration.
-Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
-that complete those presented in the conference paper.

### 2015-12-23 body/overview.tex
geometrical feasibility

@@
 \Pa and \Pb are addressed in a sequential fashion.
-The guarantees provided by the truly-feasible guide path allow
+The guarantees provided by the \textit{geometrically feasible} guide path allow
 us to avoid the combinatorial by generating the contacts in a straightforward manner.
@@
 The objective is to compute a path of root placements that will allow contact creation.
-A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide trajectory, but at the same time, any computed guide trajectory must be truly feasible, i.e.
+A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide trajectory, but at the same time, any computed guide trajectory must be \textit{truly feasible}, i.e.
 must lead to a valid sequence of contacts.
 
-More precisely, a root placement is truly feasible if and only if there exists at least one joint configuration such that the robot satisfies three conditions:
+More precisely, a root placement is \textit{truly feasible} if and only if there exists at least one joint configuration such that the robot satisfies three conditions:
 \begin{description}
@@
 \end{description}
-The conditions C1 and C2 are purely geometrical and can be efficiently approximated even without computing the contact points.
-Instead C3 is difficult to approximate and it requires the computation of the contact points, and the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
-Moreover our intuition suggests us that, for cluttered environment, C3 is not critical, meaning that, most of the times, root placements satisfying C1 and C2 would satisfy also C3.
+C1 and C2 define \textit{geometrical feasibility}, and can be efficiently approximated even without computing the contact points.
+C3 defines \textit{equilibrium feasibility}.
+C3 is difficult to approximate and requires the computation of the contact points, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
+However our intuition suggests us that, for cluttered environments, verifying C3 is not critical for $\mathcal{P}_1$, meaning that, most of the times, root placements satisfying C1 and C2 would satisfy also C3.
 For this reason we neglect C3 during the resolution of \Pa, and we worry about it only when solving \Pb.
 
-An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
+An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right arm in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
 We define formally $C_{reach}$, the set of root placements satisfying C1 and C2, in which we compute a guide trajectory with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
@@
 The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--\Pb). In this
-step we assume that the input guide path is truly feasible.
+step the input guide path is \textit{geometrically feasible}.
 With a fixed root position, the dimensionality of the 

### 2015-12-23 body/rbprm.tex
geometrical feasibility

@@
 
-We say that a root placement $\mathbf{q}^{0}$ is geometrically feasible if there exists at least one bijective mapping $\pi$
+We say that a root placement $\mathbf{q}^{0}$ is \textit{geometrically feasible} if there exists at least one bijective mapping $\pi$
 such that
@@
 
-The set of all geometrically feasible root placements is denoted by $C_{reach}$. By extension, a path $\mathbf{q}^0(t)$ is truly feasible if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{reach}$.
+The set of all \textit{geometrically feasible} root placements is denoted by $C_{reach}$.
+By extension, a path $\mathbf{q}^0(t)$ is \textit{geometrically feasible} if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{reach}$.
 
-For a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide trajectory generator: all the generated trajectories must be \textit{geometrically feasible} (sufficient condition); the guide planner must be complete, i.e.
-it must be able to discover any truly feasible trajectory (necessary condition).
+For this stage, and for this stage only, we make the assumption that in cluttered environments, a \textit{geometrically feasible} path can be transformed into a \textit{truly feasible} path.
+Under this assumption, for a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide trajectory generator: all the generated trajectories must be \textit{geometrically feasible} (sufficient condition); the guide planner must be complete, i.e.
+it must be able to discover any \textit{geometrically feasible} trajectory (necessary condition).
 
@@
 
-\subsection{Conditions for geometric feasibility}
-By default, the geometric feasibility implies a constructive demonstration by exhibiting a valid $\pi$.
+\subsection{Conditions for geometrical feasibility}
+By default, \textit{geometrical feasibility} implies a constructive demonstration by exhibiting a valid $\pi$.
 This is the approach chosen by \cite{Bouyarmane2009}.
 However, computing a valid  $\mathbf{q}^{\overline{0}}$ is costly in practice.
-In this section we rather define a necessary condition and a sufficient condition for geometric feasibility that do not require this explicit computation.
+In this section we rather define a necessary condition and a sufficient condition for \textit{geometrical feasibility} that do not require this explicit computation.
 
@@
 
-\subsubsection*{Geometric feasibility: necessary condition}
+\subsubsection*{Geometrical feasibility: necessary condition}
 For a contact to be possible, a volume $O_i \in O$ necessarily intersects with the reachable workspace $W(\mathbf{q}^{0})$ (Figure~\ref{fig:contact_gen}--1).
@@
 
-\subsubsection*{Geometric feasibility: sufficient condition}
+\subsubsection*{Geometrical feasibility: sufficient condition}
 %~ In building $C_{reach}^1$, we directly considered the including hull of the root body $W^{0}$, obtaining a necessary condition. 
-A trivial sufficient condition for geometric feasibility can be constructed as a variation of $C_{reach}^1$, by replacing $W^0$ with a bounding volume $B^{max}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
+A trivial sufficient condition for \textit{geometrical feasibility} can be constructed as a variation of $C_{reach}^1$, by replacing $W^0$ with a bounding volume $B^{max}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
 We denote by \mbox{$C_{reach}^\infty \subset C_{reach}$} the set of root placements corresponding to the sufficient condition.
@@
 
-\subsection{Reachability: a compromise condition}
+\subsection{Geometrical feasibility: a compromise reachability condition}
 The sufficient condition is not interesting in practice since it leads the solver to lose too many interesting trajectories.
-The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not truly feasible in practice.
+The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \textit{geometrical feasible} in practice.
 It might be possible to find a shape $B$ that is necessary and sufficient; however, it seems intuitively very unlikely in general.

### 2015-12-23 body/results.tex
geometrical feasibility

@@
 In this Section we present some of the results obtained with our planner. The complete sequences computed are shown in the companion video, TODO.
-Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ TODO~\cite{Semini}.
+Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
 For each scenario we indicate the chosen heuristics. We also provide a performance analysis, which shows that the planner is compatible with interactive applications,

### 2015-12-28 body/results.tex
figures + all stats

@@
 \subsubsection{HRP-2 -- Path re-planning(Figure~\ref{fig:re-planning}):}
-In this long scene, HRP-2 plans a path through through several obstacles. the scene is edited during the execution of the motion: a stair is added,
+In this long scene, HRP-2 plans a path through several obstacles. the scene is edited during the execution of the motion: a stair is added,
 some stepping stones are removed, and part of the final staircase is deleted. All these modifications require re-planning.
@@
 \centering
-\begin{tabular}{ l | >{\centering\arraybackslash}m{60pt} | >{\centering\arraybackslash}m{60pt} | >{\centering\arraybackslash}m{60pt} | >{\centering\arraybackslash}m{60pt} | >{\centering\arraybackslash}m{60pt} | c}
-  &  Complete guide generation & Collision & Static equilibrium & ik  & Total generation time & Time per step\\
+\small
+\begin{tabular}{ l | >{\centering\arraybackslash}m{57pt} | >{\centering\arraybackslash}m{70pt} | >{\centering\arraybackslash}m{75pt} | >{\centering\arraybackslash}m{77pt} | >{\centering\arraybackslash}m{85pt} | c}
+  &  Complete guide generation & Static equilibrium & Collision & ik  & Total generation time & 1 step\\
  \hline
-   Steep stairs & 22 -- \textbf{33} --  76 & 1 --  \textbf{4} -- 61   & 22 --  \textbf{52} -- 534 & 108 --  \textbf{1069} -- 10550 & 267 --  \textbf{1337} -- 11484 & \textbf{66} \\
+   Stairs & 6 -- \textbf{13} --  24 & 13 --  \textbf{32} -- 329   & 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{14} \\
+   Standing & 4 -- \textbf{451} --  985 & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 310 --  \textbf{1622} -- 3429 & \textbf{66} \\
+   Rubble & 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{3058} -- 13126 & \textbf{43} \\
+   Race & 1 -- \textbf{501} --  995 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2530 --  \textbf{3722} -- 10030 & \textbf{33} \\
    %Standing up & $70$ & 23 & 48\\
@@
    Steep stairs & 100\%  & 99.5\% & 0.11\% & 0.39\% \\
-   Standing up & 100\% & 89\% & 7\% & 4\% \\
-   Truck egress & \% & \% & \% &  \% \\
-   Rubble & \% & \% & \% & \% \\
-   Obstacle race & \% & \% & \% & \% \\
+   Standing up & 68\% & 88\% & 6\% & 6\% \\
+   %~ Truck egress & \% & \% & \% &  \% \\
+   Rubble & 74\% & 97.87\% & 0.13\% & 2\% \\
+   Obstacle race & 58\% & 95.7\% & 1.8\% & 2.5\% \\
  \end{tabular}
-\caption{Success rates for each scenario. The first value indicates the percentage of successfull complete path plannings for 1000 tests; The second value
+\caption{Success rates for each scenario. The first value indicates the percentage of successfull complete contact plannings for 1000 tests; The second value
 indicates the percentage of truly feasible root configurations: considering each limb individually, indicates the percentage of root placements of the guide trajectory that

### 2015-12-31 body/abstract.tex
cosmetics, added title, extended conclusion

@@
 The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
-designed for generic robotic applications: a manipulatibily-based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
+designed for generic robotic applications: a manipulability-based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
 These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
-variety of robots, including HRP-2, Hyq, and a dexterous hand, with interactive performances.
+variety of robots, including HRP-2, HyQ, and a dexterous hand, with interactive performances.

### 2015-12-31 body/appendix_algo.tex
typos
cosmetics, added title, extended conclusion

@@
-\section{Algorithms for discretizing of a path}
+\section{Algorithms for the discretization of a path}
 \label{app:contact}
@@
 At each step, $g$ is called with the previous state as a parameter, as well
-as a new root placement. $g$ returns a new contact configuration, if it suceeded
-in computing a configuration with only one contact switch occuring.
+as a new root placement. $g$ returns a new contact configuration, if it succeeded
+in computing a configuration with only one contact switch occurring.
 If $g$ failed in achieving this, the method \textsc{RepositionContacts} is called.
 It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
-This repositionning allows to increase the odds that the contact can be maintained at the next step.
+This repositioning allows to increase the odds that the contact can be maintained at the next step.
 

### 2015-12-31 body/conclusion.tex
cosmetics, added title, extended conclusion

@@
 work is required on the planning side to obtain a seamless workflow. To achieve this, we are currently working on the notion of transition certificate, i.e. formulating
-conditions that guarantee that the interpolation between two contact configurations is dynamically feasible. We are also
+conditions that guarantee that the interpolation between two contact configurations is dynamically feasible.
+To achieve this, our objective is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
+at the center of mass of the robot.
+ We are also
 working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
-\adnote{The 'future work' part seems a bit poor to me}
+These two points illustrate the need to reduce the distance between the planning and control aspects of multiped locomotion. We believe that one way 
+to bridge the gap is to introduce robustness while addressing these two issues.

### 2015-12-31 body/intro.tex
cosmetics, added title, extended conclusion

@@
 the  stair  railings  or  any  form  of  bracing.   Even  drunk  people  are  smart  enough  to  use  nearby  supports.
-Full body locomotion (handholds,  bracing,  leaning against a wall or obstacles) should be easier than our
+Full body locomotion (hand holds,  bracing,  leaning against a wall or obstacles) should be easier than our
 current high performance minimum contact locomotion approaches."
@@
 This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely fast.
-while preserving a success rate greater than TODO \% in the demonstrated scenarios.
+while preserving a high success rate in the demonstrated scenarios.
 

### 2015-12-31 body/overview.tex
cosmetics, added title, extended conclusion

@@
 From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root placement, an inverse-kinematics solver 
-is used to maintain the contacts that were previously existing (TODO Figure). Often, these contacts cannot be maintained, because of joint limits or collisions (TODO Figure).
-In such a case the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot (TODO FIGURE).
+is used to maintain the contacts that were previously existing. Often, these contacts cannot be maintained, because of joint limits or collisions.
+In such a case the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
 The algorithm is designed so that only one contact can be created or broken between two successive configurations. This ensures that the interpolation between two configurations is achievable.

### 2015-12-31 body/rbprm.tex
cosmetics, added title, extended conclusion

@@
 \end{equation}
-It is straighforward to prove that  $C_{reach} \subset C_{reach}^1$ (by construction of the included set). 
+It is straightforward to prove that  $C_{reach} \subset C_{reach}^1$ (by construction of the included set). 
 This inclusion is very important: it directly implies that any motion-planning algorithm with a guaranty of completeness in $SE(3) \times \mathbb{R}^n$ is complete  in $C_{reach} \times \mathbb{R}^n$.
@@
 The parametrization of $s$ allows to find a trade-off between these two desirable properties.
-As an example, Figure~\ref{fig:hyq_roms} presents the volumes $W$ computed for the HyQ robot.
+As an example, Figure~\ref{fig:HyQ_roms} presents the volumes $W$ computed for the HyQ robot.
 In Appendix~\ref{app:rom}, we give a generic method to compute these volumes, with the example of HRP-2.
@@
   \centering
-  \includegraphics[width=0.95\linewidth]{figures/hyq_roms}
+  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
   \caption{
@@
            of the torso of the robot.}
-		   \label{fig:hyq_roms}
+		   \label{fig:HyQ_roms}
 \end{figure}
@@
 can work, in spite of a high rejection rate. 
-Thus, the only significant change regarding a classical planner is to replace the classical collision checking method with a test of appartenance to $C_{reach}^s$ when verifying
+Thus, the only significant change regarding a classical planner is to replace the classical collision checking method with a test of belonging to $C_{reach}^s$ when verifying
 that drawn configurations and associated local paths are valid.
@@
 
-Thanks to this small set of modifications, the high dimensional truly feasible path planning problem is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for Hyq, 8 for HRP-2 that has joints in the torso).
+Thanks to this small set of modifications, the high dimensional truly feasible path planning problem is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for HyQ, 8 for HRP-2 that has joints in the torso).
 In so doing we make the problem tractable for sampling-based approaches, with extremely efficient computation times.

### 2015-12-31 body/results.tex
cosmetics, added title, extended conclusion

@@
 
-%~ \subsubsection{HRP-2 -- Truck egress (Figure~\ref{fig:TODO}):}
-%~ HRP-2 stands half seated in a truck, and \replaced{is required to}{it must} \replaced{get out}{exit} through the front window.
-
-%~ \begin{figure*}
-  %~ \centering
-  %~ \includegraphics[width=0.5\linewidth]{figures/standing}
-  %~ \caption{
-           %~ HRP-2 in the standing scenario. }
-		   %~ \label{fig:standing}
-%~ \end{figure*}
-
-%~ 
-%~ \noindent\textbf{Contacts involved:} All (both feet and hands).
-%~ 
-%~ \noindent\textbf{Heuristics:} $h_w$ \deleted{is used} for the feet, $h_{EFORT}$ \deleted{is used} for the hands.
-%~ 
-%~ \noindent\textbf{Observations:} This scenario demonstrates the ability of the planner to compute a path in an extremely-constrained environment. In this case,
-%~ the scenario is not always real time (TODO or is it ?): while the planner is often able to compute a guide path rapidly, in the worst case up to one minute can be spent
-%~ in the planning.
-
 \subsubsection{HyQ -- Darpa-style rubble (Figure~\ref{fig:darpa})}
@@
 
-\subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:hyq_bridge} and~\ref{fig:hyq_obs}):}
+\subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs}):}
 In this long scene, HyQ is first required to cross a 55-cm large hole; then, to cross a narrow "bridge",  only 25-cm large.
@@
   \centering
-  \includegraphics[width=0.4\linewidth]{figures/hyq_bridge}
+  \includegraphics[width=0.4\linewidth]{figures/HyQ_bridge}
   \caption{
            HyQ crossing a narrow bridge. }
-		   \label{fig:hyq_bridge}
+		   \label{fig:HyQ_bridge}
 \end{figure}
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{figures/hyq_obs}
+  \includegraphics[width=0.5\linewidth]{figures/HyQ_obs}
   \caption{
            Crossing a hole contact sequence for HyQ ($b_0 > 4$). }
-		   \label{fig:hyq_obs}
+		   \label{fig:HyQ_obs}
 \end{figure*}
@@
 the bridge must be approached with the right angle.
-This can is illustrated in Figure~\ref{fig:hyq_obs}, where several feet rearrangements are required to cross both obstacles (although the video shows this best).
+This can is illustrated in Figure~\ref{fig:HyQ_obs}, where several feet rearrangements are required to cross both obstacles (although the video shows this best).
 The planner however succeeds in finding a feasible sequence in the end, again with interactive computation times.
@@
 \centering
-\small
-\begin{tabular}{ l | >{\centering\arraybackslash}m{57pt} | >{\centering\arraybackslash}m{70pt} | >{\centering\arraybackslash}m{75pt} | >{\centering\arraybackslash}m{77pt} | >{\centering\arraybackslash}m{85pt} | c}
-  &  Complete guide generation & Static equilibrium & Collision & ik  & Total generation time & 1 step\\
+\footnotesize
+\begin{tabular}{ >{\centering\arraybackslash}m{37pt} | >{\centering\arraybackslash}m{57pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{70pt} | >{\centering\arraybackslash}m{73pt} | >{\centering\arraybackslash}m{80pt} | >{\centering\arraybackslash}m{10pt}}
+  Scenario (nb steps) &  Complete guide generation & Static equilibrium & Collision & Inverse Kinematics  & Total generation time & Time per step\\
  \hline
-   Stairs & 6 -- \textbf{13} --  24 & 13 --  \textbf{32} -- 329   & 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{14} \\
-   Standing & 4 -- \textbf{451} --  985 & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 310 --  \textbf{1622} -- 3429 & \textbf{66} \\
-   Rubble & 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{3058} -- 13126 & \textbf{43} \\
-   Race & 1 -- \textbf{501} --  995 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2530 --  \textbf{3722} -- 10030 & \textbf{33} \\
+   Stairs (18) & 6 -- \textbf{13} --  24 & 13 --  \textbf{32} -- 329   & 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{14} \\
+   Standing (24)& 4 -- \textbf{451} --  985 & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 310 --  \textbf{1622} -- 3429 & \textbf{66}  \\
+   Rubble (71)& 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{3058} -- 13126 & \textbf{43} \\
+   Race (112)& 1 -- \textbf{501} --  995 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2530 --  \textbf{3722} -- 10030 & \textbf{33} \\
    %Standing up & $70$ & 23 & 48\\
@@
 
-In most scenarios, one can observe that the average computation time for one single step is largely below the second,
+Considering the repartition of the computation time, for HRP-2, most of the time is spent performing inverse kinematics.
+This is not surprising considering the number of calls to the methods: IK projection is used intensively to maintain contact continuity between two postures; 
+it is also applied every time a new candidate needs to be evaluated. In particular for the standing up scenario,
+the kinematic constraints are really strong. Therefore a large number of projections fail because of joint limits violation.
+
+For HyQ, there is a more regular repartition of the performances for the rubble scenario. For the obstacle scenario,
+we observe a large increase of the time spent performing collision checking and inverse kinematics. This is explained
+by the complexity of crossing the bridge: the robot has to stand high and create contacts close to each other because
+the bridge is narrow. In many cases this results in collisions, thus a large number of contacts have to be evaluated.
+
+In all scenarios, one can observe that the average computation time for one single step is largely below the second,
 thus allowing to consider interactive applications.
-Considering the repartition of the computation time, it appears that most of the time is spent performing inverse kinematics.
-This is not surprising considering the number of calls to the methods: IK is used intensively to maintain contact continuity between two postures; 
-it is also applied every time a new candidate needs to be evaluated. However regarding state-of-the-art IK solutions, we believe there is room
-for improving the performances of the algorithm.
 
@@
  \end{tabular}
-\caption{Success rates for each scenario. The first value indicates the percentage of successfull complete contact plannings for 1000 tests; The second value
+\caption{Success rates for each scenario. The first value indicates the percentage of successful complete contact plannings for 1000 tests; The second value
 indicates the percentage of truly feasible root configurations: considering each limb individually, indicates the percentage of root placements of the guide trajectory that
@@
 of our theoretical formulation. 
-From a pragmatic point of view, regarding the computation times, we claim that our approach is the best current compromise between completeness and efficiency:
-Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan a new contact sequence with 
-a reasonable chance of success.
+From a pragmatic point of view, regarding the computation times, we claim that our approach is the best current compromise between completeness and efficiency.
+Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan a new contact sequence with a reasonable chance of success.
+The most efficient (and immediate) approach is probably to launch in parallel several instances of the planner for a given problem (our current implementation is single threaded) and to apply any successful result.

### 2015-12-31 main.tex
revised beg intro
intro + overview no fig
debut planner
finished contact generation now heuristics
robust equilibrium
corrections andrea
appendix
Add my comments to abstract
Andrea's first complete review
cosmetics, added title, extended conclusion

@@
 \usepackage{moreverb,url}
+\usepackage{makeidx}         % allows index generation
+\usepackage[percent]{overpic}
+\usepackage{array} 
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
+\usepackage{listings}
+\usepackage{color}
+\usepackage{bm}
+
+\input{./math_commands.tex}
+
+\newcommand{\commentt}[2]{\textcolor{#1}{\textbf{\textit{[#2]}}}} 	% make sure no comments left in the paper
+%\newcommand{\commentt}[2]{}							% using this line rather than the previous one you remove all comments
+\newcommand{\adnote}[1]{\commentt{blue}{AD: #1}}
+\newcommand{\nmnote}[1]{\commentt{OliveGreen}{NM: #1}}
+
+% Another method to track changes
+% Examples of usage:
+% - This is \added[id=per,remark={we need this}]{new} text.
+% - This is \deleted[id=per,remark=obsolete]{unnecessary}text.
+% - This is \replaced[id=per]{nice}{bad} text.
+% To print the list of change use: \listofchanges
+\usepackage{changes}	% use this for the working version
+%\usepackage[final]{changes} % use this for the final version
+\definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
 
@@
 
-\title{A demonstration of the \LaTeXe\ class file for
-\itshape{SAGE Publications}}
+\title{An efficient acyclic contact planner for multiped robots}
 
-\author{Steve Tonneau\affilnum{1}, Andrea Del Prete\affilnum{1}, Julien Pettr\'e\affilnum{2}, Franck Multon\affilnum{3}, Nicolas Mansard\affilnum{1}}
-
+\author{Steve Tonneau\affilnum{1}, Andrea Del Prete\affilnum{1}, Julien Pettr\'e\affilnum{2}, Nicolas Mansard\affilnum{1}}
+ %~ Franck Multon\affilnum{3},
 \affiliation{\affilnum{1}LAAS-CNRS, Toulouse, France\\
-\affilnum{2}Inria, Rennes, France \\
-\affilnum{3}Universit\'e Rennes 2, France}
+\affilnum{2}Inria, Rennes, France } % \\
+%~ \affilnum{3}Universit\'e Rennes 2, France}
 
@@
 
-\keywords{Class file, \LaTeXe, \textit{SAGE Publications}}
+\keywords{Motion planning, Contact planning, Humanoid Robot, Legged Locomotion, Rough Terrain}
 
@@
 \input{body/intro}
-
+\input{body/overview}
+\input{body/rbprm}
+\input{body/contact}
+\input{body/equilibrium}
+\input{body/efort}
+\input{body/results}
+\input{body/conclusion}
+\theendnotes
+\bibliographystyle{SageH}
+\bibliography{biblio}
+\appendix
+\input{body/appendix_roms}
+\input{body/appendix_algo}
+\input{body/appendix_hpp}
 \end{document}

### 2015-12-31 body/intro.tex
citation tedrake

@@
 \citeauthor{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed integer problem, but only cyclic, bipedal locomotion is considered.
-Aside from the computation cost, a major drawback of these optimization based approaches is thus that they only offer local convergence when applied to acyclic contact planning.
+\citeauthor{dai2014whole} extend the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the short contact sequence for landing motions, but specify
+the contacts manually for more complex interactions.
+A major drawback of these optimization based approaches is thus that they only offer local convergence when applied to acyclic contact planning.
 

### 2016-01-03 body/appendix_roms.tex
final version before video inclusion

@@
 Concretely, because the false positive set intersects with $W^0$, the scaling volume of the robot torso, the induced error is compensated,
-as verified by the date of Table~\ref{tab:requestpercent}.
+as verified by the results shown by Table~\ref{tab:requestpercent}.
 
@@
 First, we compute the bounding boxes of the robot trunk, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
-Then, we perform a scaling of these boxes by a factor $s$. For different values of $s$, we verify the true feasibility condition:
-For a large number of configurati
-ons that verify the reachability condition, we try to generate contact configurations. We plot 
-the success rates obtained depending on $s$ (Figure~\ref{hrp2_w}). We then choose the minimal satisfying value of $s$.
+Then, we perform a scaling of these boxes by a factor $s$. For different values of $s$, we verify the geometrical feasibility condition:
+For a large number of configurations that verify the reachability condition, we try to generate contact configurations. We then
+empirically select the lowest value of $s$ that provides a reasonnable success rate in all scenarios.
+For HRP-2, we choose $s=1.2$. This leads to a success rate higher than 86 \% for all scenarios.

### 2016-01-03 body/conclusion.tex
final version before video inclusion

@@
 Our main objective is of course to generate the complete interpolation between the contact sequences.
-We have already been able to achieve this for HRP-2 for some of the computed sequences~\citep{Carpentier2016}, but we admit that additional
+We have already obtained some successes for some of the computed sequences~\citep{Carpentier2016}, but we admit that additional
 work is required on the planning side to obtain a seamless workflow. To achieve this, we are currently working on the notion of transition certificate, i.e. formulating
 conditions that guarantee that the interpolation between two contact configurations is dynamically feasible.
-To achieve this, our objective is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
+Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
 at the center of mass of the robot.

### 2016-01-03 body/intro.tex
final version before video inclusion

@@
 existing planners, decoupled or not. However, recent contributions are able to interpolate dynamically-feasible
-trajectories between contact configurations \citep{herzog2015trajectory, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real-time performances.
+trajectories between contact configurations \citep{herzog2015trajectory, Park116, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real-time performances on the real robot.
 To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly. 
@@
 
-This sequential approach is the key to the efficiency of our method, though it can
+This sequential approach is the key to the efficiency of our method, though it
 can result in failures because our assumption is not always true.
@@
 of our contact generation algorithm.
-Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ\adnote{cite Semini}.
+Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ~\citep{semini11hyqdesignjsce}.
 Appendix~\ref{app:rom} describes the generation process

### 2016-01-03 body/rbprm.tex
final version before video inclusion

@@
 The sufficient condition is not interesting in practice since it leads the solver to lose too many interesting trajectories.
-The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \textit{geometrical feasible} in practice.
+The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \textit{geometrically feasible} in practice.
 It might be possible to find a shape $B$ that is necessary and sufficient; however, it seems intuitively very unlikely in general.

### 2016-01-03 body/results.tex
final version before video inclusion

@@
   \begin{overpic}[width=1\linewidth]{figures/robots_old}
-		%~ \put (5,58) {1)} 
-		%~ \put (37,58) {2)} 
-		%~ \put (68,58) {3.a)} 
+		%~ \put (5,) {1)} 
+		%~ \put (37,) {2)} 
+		%~ \put (68,) {3.a)} 
 		%~ \put (5,27) {3.b)} 
@@
 		\put (79,5) {\small{\color{green}$b_0 = 6.16$}} 
-		%~ \put (68,58) {3.a)} 
+		%~ \put (68,) {3.a)} 
 		%~ \put (5,27) {3.b)} 
@@
 
+\subsubsection{HRP-2 -- Car egress (Figure~\ref{fig:car}):}
+This scenario is inspired from the Darpa challenge car egress scenario (\url{http://cpc.cx/edH}). HRP-2 has 
+to find the contact sequence that allows it to step out of a car.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{figures/car}
+  \caption{
+           Selected frames the car egress scenario. }
+		   \label{fig:car}
+\end{figure}
+
+\noindent\textbf{Contacts involved:} All (both feet and hands).
+
+\noindent\textbf{Heuristics:} $h_w$.
+
+\noindent\textbf{Observations:} The difficulty of this scenario lies in the strong reduction of the reachable workspace induced 
+by the extreme proximity of all obstacles. The planner is able to find a sequence, that consists in many stances.
+The proximity of the obstacles invalidate a large number of contact candidates because of collisions. To respect the condition
+that only one contact must be broken between each phase, the motion has to be decomposed into a large number of stances (61 in average).
+While this scenario is the slower to solve, the planner still computes a solution interactively.
+ % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
+
 \subsubsection{HyQ -- Darpa-style rubble (Figure~\ref{fig:darpa})}
@@
 \subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs}):}
-In this long scene, HyQ is first required to cross a 55-cm large hole; then, to cross a narrow "bridge",  only 25-cm large.
+In this long scene, HyQ is first required to cross a 55-cm large hole; then, to cross a narrow ``bridge'', large of only 25-cm.
 
@@
 
-\noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is the hardest for our planner.
-While finding a guide path above the hole is easy for the planner, finding a sequence of contacts that allows for equilibrium is not trivial.
+\noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is a hard case for a contact planner.
+While finding a guide path above the hole is easy for the guide planner, finding a sequence of contacts that allows for equilibrium is not trivial.
 Second, the narrow bridge is hard both for the planner and the contact generator: to make sure that equilibrium is preserved along the traversal,
 the bridge must be approached with the right angle.
-This can is illustrated in Figure~\ref{fig:HyQ_obs}, where several feet rearrangements are required to cross both obstacles (although the video shows this best).
+The difficulty is illustrated in Figure~\ref{fig:HyQ_obs}, where several feet rearrangements are required to cross the hole (although the video shows this best).
 The planner however succeeds in finding a feasible sequence in the end, again with interactive computation times.
 
-\subsubsection{HRP-2 -- Path re-planning(Figure~\ref{fig:re-planning}):}
+\subsubsection{HRP-2 -- Path re-planning (Figure~\ref{fig:re-planning}):}
 In this long scene, HRP-2 plans a path through several obstacles. the scene is edited during the execution of the motion: a stair is added,
@@
 
-\subsubsection{Computation times}
-Table~\ref{tab:requestime} summarizes the performance measurement obtained, in terms of computation times.
-
 \begin{table*}
@@
    Standing (24)& 4 -- \textbf{451} --  985 & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 310 --  \textbf{1622} -- 3429 & \textbf{66}  \\
-   Rubble (71)& 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{3058} -- 13126 & \textbf{43} \\
-   Race (112)& 1 -- \textbf{501} --  995 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2530 --  \textbf{3722} -- 10030 & \textbf{33} \\
+   Car (61)& 1 -- \textbf{529} --  995 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6775 --  \textbf{20890} -- 52963 & \textbf{342} \\
+   Rubble (71)& 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{30} -- 13126 & \textbf{43} \\
+   Race (112)& 1 -- \textbf{501} --  995 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2530 --  \textbf{3722} -- 10030 & \textbf{33}
    %Standing up & $70$ & 23 & 48\\
@@
 
-Considering the repartition of the computation time, for HRP-2, most of the time is spent performing inverse kinematics.
-This is not surprising considering the number of calls to the methods: IK projection is used intensively to maintain contact continuity between two postures; 
-it is also applied every time a new candidate needs to be evaluated. In particular for the standing up scenario,
-the kinematic constraints are really strong. Therefore a large number of projections fail because of joint limits violation.
-
-For HyQ, there is a more regular repartition of the performances for the rubble scenario. For the obstacle scenario,
-we observe a large increase of the time spent performing collision checking and inverse kinematics. This is explained
-by the complexity of crossing the bridge: the robot has to stand high and create contacts close to each other because
-the bridge is narrow. In many cases this results in collisions, thus a large number of contacts have to be evaluated.
-
-In all scenarios, one can observe that the average computation time for one single step is largely below the second,
-thus allowing to consider interactive applications.
-
-\subsubsection{Success rates}
-Table~\ref{tab:requestpercent} summarizes the success rates obtained for each scenario.
 \begin{table*}
@@
    Standing up & 68\% & 88\% & 6\% & 6\% \\
-   %~ Truck egress & \% & \% & \% &  \% \\
+   Car egress & 39\% & 77\% & 21\% &  2\% \\
    Rubble & 74\% & 97.87\% & 0.13\% & 2\% \\
@@
 
+\subsubsection{Computation times}
+Table~\ref{tab:requestime} summarizes the performance measurement obtained, in terms of computation times.
+
+Considering the repartition of the computation time, for HRP-2, most of the time is spent performing inverse kinematics.
+This is not surprising considering the number of calls to the methods: IK projection is used intensively to maintain contact continuity between two postures; 
+it is also applied every time a new candidate needs to be evaluated. In particular for the car egress scenario,
+the kinematic constraints are really strong. Therefore a large number of projections fail because of joint limits violation.
+
+For HyQ, there is a more regular repartition of the performances for the rubble scenario. For the obstacle scenario,
+we observe a large increase of the time spent performing collision checking and inverse kinematics. This is explained
+by the complexity of crossing the bridge: the robot has to stand high and create contacts close to each other because
+the bridge is narrow. In many cases this results in collisions, thus a large number of contacts have to be evaluated.
+
+In all scenarios, one can observe that the average computation time for one single step is largely below the second,
+thus allowing to consider interactive applications.
+
+Table~\ref{tab:requestpercent} summarizes the success rates obtained for each scenario.
+
+\subsubsection{Success rates}
 We observe that, as expected, our planner does not succeed systematically, because of the approximations made in the implementation
-of our theoretical formulation. 
-From a pragmatic point of view, regarding the computation times, we claim that our approach is the best current compromise between completeness and efficiency.
+of our theoretical formulation. The extreme situation of the car egress scenario provides the less satisfying results in terms of kinematic failure. This is not
+suprising considering the narrowness of the scene.
+From a pragmatic point of view, regarding the computation times, we claim that our approach provides a satisfying compromise between completeness and efficiency.
 Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan a new contact sequence with a reasonable chance of success.

### 2016-01-03 main.tex
citation tedrake
final version before video inclusion

@@
 
-\author{Steve Tonneau\affilnum{1}, Andrea Del Prete\affilnum{1}, Julien Pettr\'e\affilnum{2}, Nicolas Mansard\affilnum{1}}
+\author{Steve Tonneau\affilnum{1}, Andrea Del Prete\affilnum{1}, Julien Pettr\'e\affilnum{2}, Chonhyon Park\affilnum{3}, Dinesh Manocha\affilnum{3}, Nicolas Mansard\affilnum{1}}
  %~ Franck Multon\affilnum{3},
 \affiliation{\affilnum{1}LAAS-CNRS, Toulouse, France\\
-\affilnum{2}Inria, Rennes, France } % \\
+\affilnum{2}Inria, Rennes, France\\
+\affilnum{3}UNC, Chapel Hill, USA  } % \\
 %~ \affilnum{3}Universit\'e Rennes 2, France}
@@
 \input{body/conclusion}
+\section*{Acknowledgements}
+This research is supported by Euroc (project under FP7 Grant Agreement  608849),  Entracte (ANR  grant  agreement  13-CORD-002-01),
+the ARO Contract W911NF-14-1-0437,
+the NSF award 1305286, and a grant from Boeing.
 \theendnotes

### 2016-01-04 body/intro.tex
adding cover letter

@@
 \citeauthor{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed integer problem, but only cyclic, bipedal locomotion is considered.
-\citeauthor{dai2014whole} extend the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the short contact sequence for landing motions, but specify
+\citeauthor{dai2014whole} extend the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
 the contacts manually for more complex interactions.
@@
 
-It thus appears that an integrated optimization-based approach \citep{Mordatch:2012:DCB:2185520.2185539}, able to generate a locally-optimal, complete trajectory within minutes, currently outperforms
-existing planners, decoupled or not. However, recent contributions are able to interpolate dynamically-feasible
+For completeness, we lastly mention a new kind of approach, recently proposed from the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
+use black box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
+at interactive frame rates. While this new approach provides an exciting direction of research, currently the resulting motions
+look unnatural, and do not seem applicable to real robots.
+
+It thus appears that, regarding robotics applications, an integrated optimization-based approach \citep{Mordatch:2012:DCB:2185520.2185539}, able to generate a locally-optimal, complete trajectory within minutes, currently outperforms existing planners, decoupled or not.
+However, recent contributions are able to interpolate dynamically-feasible
 trajectories between contact configurations \citep{herzog2015trajectory, Park116, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real-time performances on the real robot.

### 2016-01-04 body/results.tex
adding cover letter

@@
 \label{sec:results}
-In this Section we present some of the results obtained with our planner. The complete sequences computed are shown in the companion video, TODO.
+In this Section we present some of the results obtained with our planner. The complete sequences computed are shown in the companion video.
 Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
@@
 the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 
-A companion video available at TODO \url{http://youtu.be/LmLAHgGQJGA} (anonymous link) displays the complete contact sequence obtained in all these scenarios.
+The companion video presents the complete contact sequence obtained in all these scenarios.
 

### 2016-01-04 cover_letter.tex
adding cover letter

\documentclass[a4paper]{article}
 \usepackage{fullpage}
\usepackage{amsfonts} % pour les lettres maths creuses \mathbb
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[frenchb]{babel}
\usepackage{aeguill}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\usepackage{listings}
\usepackage{pdfpages}

\newcommand\dx{\dot{x}}
\newcommand\dy{\dot{y}}
\newcommand\ddx{\ddot{x}}
\newcommand\ddy{\ddot{y}}
\def\real{{\mathbb R}}
\frenchbsetup%
{%
StandardItemLabels=true,%
ItemLabels=\ding{43},%
}%

\author {}
\title {Cover letter for the IJRR submission \\ ``An efficient acyclic contact planner for multiped robots''}
\date {}
\begin{document}
\maketitle

Dear editor and reviewers, \\ \\

The current submission is an extension to our conference paper ``A reachability-based planner for sequences of acyclic contacts in
cluttered environments'', accepted to ISRR 15 (\url{http://stevetonneau.fr/files/publications/isrr15/isrr15.pdf}).

In these two papers, our objective is to propose automatic methods
to compute the contact sequence that characterize a robot motion in
cluttered environment.

We joined to the present submission the reviews that were made on the
conference paper. They follow the cover letter.

The extension of our work addresses the main comments of the reviewers:

\begin{itemize}
\item They were concerned by the fact that our demonstrations were
considering virtual avatars and not actual robots. In the present
submission, all our examples are applied to the HRP-2 and HyQ robots (Section 6).
To achieve this, a new contribution is proposed in our paper:
We introduce a robustness criterion for asserting the static equilibrium
of a configuration (Section 5.1). We use it in combination with new heuristics that we
propose to bias the contact generation process towards configurations
relevant for the motion being performed (Section 5.2).
With the exception of the dextrous hand, all the demonstrations shown in
the paper and the video are entirely new.

\item They also judged that the conference paper provided a theoretical
description of our approach, but lacked the details on how to apply
the method on an actual robot. We did not have the room to describe
this process in the conference paper, and explain the complete process
for the HRP-2 in the present submission (Appendix A).

\item Concerning remarks made on the reproducibility of the method,
we now provide the complete pseudo code of the algorithm, as well as the
source code of our application (Appendix B).

\item The paper has been largely re written: we provide a clearer
comparison to the state of the art (Section 1.1); we clarify our statement regarding
our computation times (Section 6 and 6.1.6); we make a clear distinction between the
theoretical concepts we propose and their approximation (Section 2).

\item Regarding the performances, compared to our conference paper, the planner is much more efficient.
A fundamental change is the reimplementation of the algorithm as a Bi-RRT planner instead of the PRM
approach. The code is also more efficient thanks to its complete rewriting into the HPP platform.

\item Finally, we propose an analysis of the performances of our method and
the success rates we obtained for each scenario (Section 6.2). This analysis provides
a support for our argumentation on the strong choices made in the design
of our planner (Section 1.2 and 2).
\end{itemize}

We are convinced that these points not only provide a complete answer to
the reviews of the conference paper, but also bring new contributions to
the community, that justify a journal publication.

We thank the editor and the reviewers for their time.

Sincerely, \\

The authors

\includepdf[pages=-]{reviewisrr.pdf}

\end{document}

### 2016-01-04 body/conclusion.tex
last biblio inclusion

@@
 working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
-These two points illustrate the need to reduce the distance between the planning and control aspects of multiped locomotion. We believe that one way 
-to bridge the gap is to introduce robustness while addressing these two issues.
+Lastly, we aim at going beyond static equilibrium and to perform kynodynamic planning in the configuration space. We believe that the most promising direction in this regard is to integrate
+the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
+Addressing these three issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.

### 2016-01-04 body/intro.tex
last biblio inclusion

@@
 Still, the computation time is far from interactive  (about 1 minute of computation for a sequence of 20 contacts).
+A similar approach has been considered for manipulation by \cite{gabicciniisrr15}.
 \citeauthor{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed integer problem, but only cyclic, bipedal locomotion is considered.
@@
 \Pb  is then addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
-A solution can then be found, at the cost of prohibitive computation times (several hours).
-Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the quality of the guide trajectory,  which is not evaluated \textit{a priori}.
+A solution can then be found, efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, but at the cost of prohibitive computation times (several hours) for constrained scenarios for humanoid.
+In the latter case, this approach is suboptimal because the quality of the motion is conditioned by the quality of the guide trajectory,  which is not evaluated \textit{a priori}.
 \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body contact configurations in static equilibrium.
@@
 Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the trajectory.
+\citeauthor{7140082} also propose a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, that characterize the ability to create a contact with
+a discretized environment. This planning phase does not account for collisions, involving that replanning is required in case of failure. This approach is efficient 
+in the scenarios demonstrated. In extremely constrained cases such 
+as the car egress scenario we address here, we believe that including collision constraints in the planning is a requirement.
 
@@
 However, recent contributions are able to interpolate dynamically-feasible
-trajectories between contact configurations \citep{herzog2015trajectory, Park116, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real-time performances on the real robot.
+trajectories between contact configurations \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
+\citeauthor{Carpentier2016} are able to achieve this with real-time performances on the real robot.
 To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly. 

### 2016-01-04 cover_letter.tex
last biblio inclusion

@@
 In these two papers, our objective is to propose automatic methods
-to compute the contact sequence that characterize a robot motion in
+to compute the contact sequence that characterizes a robot motion in
 cluttered environment.
@@
 description of our approach, but lacked the details on how to apply
-the method on an actual robot. We did not have the room to describe
-this process in the conference paper, and explain the complete process
+the method on an actual robot. We now explain the complete process
 for the HRP-2 in the present submission (Appendix A).
@@
 
-\item The paper has been largely re written: we provide a clearer
+\item The paper has been largely re-written: we provide a clearer
 comparison to the state of the art (Section 1.1); we clarify our statement regarding
@@
 
-\item Regarding the performances, compared to our conference paper, the planner is much more efficient.
+\item Regarding the computation time performances, compared to our conference paper, the planner is much more efficient.
 A fundamental change is the reimplementation of the algorithm as a Bi-RRT planner instead of the PRM
@@
 
+%~ On a last note, as required by the IJRR submission form,
+%~ we proposed a list of editors and reviewers for the submission.
+%~ We affirm that we don't have conflicts of interest with the people we propose.
+%~ For complete clarity however, we want to mention that some of the authors
+%~ are proposing two workshops for the ICRA conference this year.
+%~ Among the names we proposed for reviews, Karen Liu, kris Hauser, and Quang-Cuong Pham have accepted to be invited speakers.
+
 We thank the editor and the reviewers for their time.

### 2016-01-12 body/intro.tex
m

@@
 
-Previous contributions that embrace the combinatorial provide complete approaches, not applicable in practice because they require hours of computations \citep{conf/iser/BretlRLKA04}.
+Previous contributions that embrace the combinatorial provide complete approaches, not applicable in practice because they require hours of computations \citep{Bretl:2006:MPM:1124573.1124585}.
 More recent successes use a local solution, resulting in more reasonable computation times (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
@@
 
-The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper.
+The issue of planning acyclic contacts was first completely described by \citeauthor{Bretl:2006:MPM:1124573.1124585} in their seminal paper.
 The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: Planning a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
@@
 
-Most of the papers that followed the work of \citeauthor{conf/iser/BretlRLKA04} have explored alternative formulations to handle the combinatorial issue.
+Most of the papers that followed the work of \citeauthor{Bretl:2006:MPM:1124573.1124585} have explored alternative formulations to handle the combinatorial issue.
 Two main directions have been explored.

### 2016-01-17 body/abstract.tex
trajectory removed

@@
 as a complete contact planner. 
-First, we propose a geometrical characterization of a guide trajectory, the ``reachability condition", which guarantees that a
+First, we propose a geometrical characterization of a guide path, the ``reachability condition", which guarantees that a
 guide can be mapped into the contact manifold of the robot. With this condition, asserting the``true feasibility" 

### 2016-01-17 body/contact.tex
cosmetics, added title, extended conclusion
trajectory removed

@@
 \item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
-\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree stata structure;
+\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
 \item Figure~\ref{fig:contact_gen}--b and c--: At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
@@
 
-\subsection{Extension of the guide trajectory}
+\subsection{Extension of the guide path}
 Using the contact generator, we define an iterative algorithm to generate the contact

### 2016-01-17 body/intro.tex
trajectory removed

@@
 As suggested by \citeauthor{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner, while preserving the completeness of the search.
-We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot trajectory.
+We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot path.
 This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
@@
 The issue of planning acyclic contacts was first completely described by \citeauthor{Bretl:2006:MPM:1124573.1124585} in their seminal paper.
-The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: Planning a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
-A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories.
+The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: Planning a discrete sequence of acyclic contact configurations along the path (A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
+A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential paths.
 This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments.
@@
 \textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity.
-The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough guide trajectory (i.e.
+The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough guide path (i.e.
 an ad-hoc solution to $\mathcal{P}_1$).
@@
 A solution can then be found, efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, but at the cost of prohibitive computation times (several hours) for constrained scenarios for humanoid.
-In the latter case, this approach is suboptimal because the quality of the motion is conditioned by the quality of the guide trajectory,  which is not evaluated \textit{a priori}.
-\citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body contact configurations in static equilibrium.
-Randomly sampled configurations are projected into the contact sub-manifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide trajectory in the examples presented).
-Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the trajectory.
+In the latter case, this approach is suboptimal because the quality of the motion is conditioned by the quality of the guide path,  which is not evaluated \textit{a priori}.
+\citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide path with guarantees of contact feasibility, by extending key frames of the path into whole-body contact configurations in static equilibrium.
+Randomly sampled configurations are projected into the contact sub-manifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide path in the examples presented).
+Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the path.
 \citeauthor{7140082} also propose a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, that characterize the ability to create a contact with
@@
 
-It thus appears that, regarding robotics applications, an integrated optimization-based approach \citep{Mordatch:2012:DCB:2185520.2185539}, able to generate a locally-optimal, complete trajectory within minutes, currently outperforms existing planners, decoupled or not.
-However, recent contributions are able to interpolate dynamically-feasible
-trajectories between contact configurations \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
+Interpolation between contact poses ($\mathcal{P}_3$) remains an unsolved problem,
+though recent contributions have been able to bring partial solutions to it \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
 \citeauthor{Carpentier2016} are able to achieve this with real-time performances on the real robot.
@@
 This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
-we believe that the separation between the generation of the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
+we believe that the separation between the generation of the guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
 However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated. \\
 
-\noindent \textbf{Regarding $\mathcal{P}_1$}, the guide trajectory must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces).
+\noindent \textbf{Regarding $\mathcal{P}_1$}, the guide path must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces).
 We call this property \textit{true feasibility}.
-This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
+This property has not been studied yet; the only way to validate a waypoint in the path is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
 \\
 
-\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinite combinations of possible contact sequences for a given root trajectory.
+\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinite combinations of possible contact sequences for a given root path.
 The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious transitions).
@@
 Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
-Since the interpolation of a complete trajectory between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
+Since the interpolation of a complete path between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
 is the fastest multi-contact planning solution to our knowledge.
@@
 This is achieved by defining a geometric condition, the reachability condition (abstracted here with the transparent cylinders).
-\Pb: The trajectory is extended into a discrete sequence of contact configurations using an iterative algorithm.}
+\Pb: The path is extended into a discrete sequence of contact configurations using an iterative algorithm.}
   \label{fig:framework}

### 2016-01-17 body/overview.tex
trajectory removed

@@
 The objective is to compute a path of root placements that will allow contact creation.
-A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide trajectory, but at the same time, any computed guide trajectory must be \textit{truly feasible}, i.e.
+A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide path, but at the same time, any computed guide path must be \textit{truly feasible}, i.e.
 must lead to a valid sequence of contacts.
@@
 An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right arm in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
-We define formally $C_{reach}$, the set of root placements satisfying C1 and C2, in which we compute a guide trajectory with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
+We define formally $C_{reach}$, the set of root placements satisfying C1 and C2, in which we compute a guide path with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
 Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.

### 2016-01-17 body/rbprm.tex
trajectory removed

@@
 For this stage, and for this stage only, we make the assumption that in cluttered environments, a \textit{geometrically feasible} path can be transformed into a \textit{truly feasible} path.
-Under this assumption, for a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide trajectory generator: all the generated trajectories must be \textit{geometrically feasible} (sufficient condition); the guide planner must be complete, i.e.
-it must be able to discover any \textit{geometrically feasible} trajectory (necessary condition).
+Under this assumption, for a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide path generator: all the generated paths must be \textit{geometrically feasible} (sufficient condition); the guide planner must be complete, i.e.
+it must be able to discover any \textit{geometrically feasible} path (necessary condition).
 
@@
 
-The condition $C_{reach}^1$ is only necessary which means that one such root placement might not be truly feasible: in practice it is not guaranteed to find a valid sequence of contacts for every guide trajectory in $C_{reach}^1$.
+The condition $C_{reach}^1$ is only necessary which means that one such root placement might not be truly feasible: in practice it is not guaranteed to find a valid sequence of contacts for every guide path in $C_{reach}^1$.
 
@@
 In general, the inclusion is strict, which means that we lose the completeness of the two-stage contact planner (i.e.
-the planner is not able to discover a trajectory inside \mbox{$C_{reach} \setminus C_{reach}^\infty$}).
-However, the sufficient condition guarantees that any such trajectories leads to a valid sequence of contacts (i.e.
+the planner is not able to discover a path inside \mbox{$C_{reach} \setminus C_{reach}^\infty$}).
+However, the sufficient condition guarantees that any such path leads to a valid sequence of contacts (i.e.
 $\pi$ is defined).
@@
 \subsection{Geometrical feasibility: a compromise reachability condition}
-The sufficient condition is not interesting in practice since it leads the solver to lose too many interesting trajectories.
+The sufficient condition is not interesting in practice since it leads the solver to lose too many interesting paths.
 The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \textit{geometrically feasible} in practice.
@@
 
-\subsection{Computing the guide trajectory in $C_{reach}^s$}
-Once a value of $s$ has been fixed, any sampling-based motion planner can be used to plan a trajectory in $C_{reach}^s$. 
+\subsection{Computing the guide path in $C_{reach}^s$}
+Once a value of $s$ has been fixed, any sampling-based motion planner can be used to plan a path in $C_{reach}^s$. 
 Indeed, contrary to $C_{Contact}$,  $C_{reach}^s$ has a non zero measure in the configuration space $C$. Therefore a standard uniform sampling approach

### 2016-01-18 body/abstract.tex
nico + glossary

@@
 First, we propose a geometrical characterization of a guide path, the ``reachability condition", which guarantees that a
-guide can be mapped into the contact manifold of the robot. With this condition, asserting the``true feasibility" 
+guide can be mapped into the contact manifold of the robot. With this condition, asserting the feasibility
 of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
-Then, we propose a sequential algorithm to extend a truly feasible path into a sequence of contacts. For each limb, an off-line sampling approach,
+Then, we propose a sequential algorithm to extend a feasible guide path into a sequence of contacts. For each limb, an off-line sampling approach,
 independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
 The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
-designed for generic robotic applications: a manipulability-based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
+designed for generic robotic applications: a manipulability-based heuristic, EFORT, and a robustness criterion regarding the equilibrium of the robot.
 These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large

### 2016-01-18 body/appendix_roms.tex
nico + glossary

@@
 Then, we perform a scaling of these boxes by a factor $s$. For different values of $s$, we verify the geometrical feasibility condition:
-For a large number of configurations that verify the reachability condition, we try to generate contact configurations. We then
+For a large number of configurations that verify the \gls{reachc}, we try to generate contact configurations. We then
 empirically select the lowest value of $s$ that provides a reasonnable success rate in all scenarios.

### 2016-01-18 body/conclusion.tex
nico + glossary

@@
 criteria that are hard to conciliate: generality, performance, and quality of the solution, making it the first acyclic contact
-planner compatible with interactive applications.
+planner compatible with \gls{interactive} applications.
 %
-\textbf{Regarding generality}, the reachability condition, coupled with an approach based on limb decomposition, 
+\textbf{Regarding generality}, the \gls{reachc}, coupled with an approach based on limb decomposition, 
 allows the method to address arbitrary multiped robots. The only pre-requisite is the specification 
@@
 %
-\textbf{Regarding the quality of the paths}, a parametrization of the reachability condition allows us to compute
+\textbf{Regarding the quality of the paths}, a parametrization of the \gls{reachc} allows us to compute
 relevant paths in all the presented scenarios, with low rejection rates.

### 2016-01-18 body/contact.tex
nico + glossary

@@
 \label{sec:contact}
-As an input of this stage, we consider a \textit{geometrically feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$.
+Our planner computes guide paths in \gls{creach0}, an over-approximation of the set of \gls{equilibrium feasible} paths, \gls{ccequil0}.
+As an input of this stage, we however assume an \gls{equilibrium feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow \mathbb{R}^r$ for the root of the robot $R$.
+If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section TODO.
 We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.
@@
 and root location.
-Then, on this method we build an iterative algorithm to generate a discrete sequence of contact configurations in equilibrium.
+Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.
 
-In what follows a \textit{valid} contact configuration is understood as being collision free, satisfying joint limits, and allowing to maintain the static equilibrium
-of the robot. The first two tests are trivial; our criterion to assert efficiently the static equilibrium of the system
+Our criterion to assert efficiently the static equilibrium of the system
 is described in Section~\ref{sec:heuristics}.
@@
 We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$.
-From \textit{geometrical feasibility} we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify \eqref{eq:pi} for $\mathbf{q}^0$.
-The issue is then to evaluate locally at least one element of $P$ at $\mathbf{q}^0$.
+From \glslink{contact feasible}{\textit{contact feasibility}} we know that there exists at least one $\mathbf{q}^{\overline{0}}$ that verifies $\mathbf{q}^k \oplus \mathbf{q}^{\overline{k}} \in$ \gls{ccontact}.
+The issue is then to exhibit locally this $\mathbf{q}^{\overline{0}}$.
 
-While exhibiting analytically a $\pi_i$ does not seem tractable, we can iteratively evaluate $\pi_i(\mathbf{q}^0)$ as follows:
+While exhibiting analytically a $\mathbf{q}^{\overline{0}}$ does not seem tractable, we can iteratively try to generate one as follows:
 \begin{enumerate}
@@
 \item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
-\item Figure~\ref{fig:contact_gen}--b and c--: At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
+\item At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
 \item According to a user-defined heuristic $h$, sort the elements of $S$;
-\item Figure~\ref{fig:contact_gen}--d and e--: Select the first configuration of $S$ (or fail if $S$ is empty). Project the configuration onto contact with inverse kinematics;
-\item If a valid solution is found, stop. Otherwise remove the element from $S$ and go back step 5.
+\item Select the first configuration of $S$. Project the configuration onto contact with inverse kinematics. If $S$ is empty, stop (failure case) (Figure~\ref{fig:contact_gen} (d) and (e));
+\item If a valid solution is found, stop (success case). Otherwise remove the element from $S$ and go back step 5.
 \end{enumerate}
@@
 Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
-Regarding completeness, it is immediate to verify that as $N$ grows, the completeness of the method is restored.
-$N$ is a parameter that allows us to specify the trade-off between completeness and efficiency.
+Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
+$N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
 The reader is invited to refer to our previous work for an extensive discussion on the optimal value of $N$ \citep{Tonneau2014}.
+
+It should be noted that a failure case for our contact generator does not necessarily mean that the planner will globally fail. Indeed,
+for a given root configuration, if contact configurations are found for other limbs, resulting in static equilibrium, the configuration will be accepted despite
+the failure to generate a contact.
 
@@
 
-As an input, we consider the guide path $\mathbf{q}^0(t)$, automatically discretized into a sequence of $j$ key placements:  
+As an input, we consider the guide path $\mathbf{q}^0(t)$, automatically discretized into a sequence of $j$ key configurations:  
 \begin{equation*}
@@
 where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations.
-To ensure continuity in the contact transition phases, we rewrite $\pi$ under the following recursive form for any $0<i<j$:
+To ensure continuity in the contact transition phases, we propose a recursive mapping $\pi$, for any $0<i<j$:
 \begin{equation*}
@@
     \begin{aligned}		
-        \mathbf{Q}^0 \in C_{reach} & \longrightarrow C_{contact} \\
+        \mathbf{Q}^0 \in C_{equil}^0 & \longrightarrow C_{equil} \\
         %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
@@
 These three rules provide a deterministic manner to generate contacts along the discretized guide path.
-For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
+For a given \gls{equilibrium feasible} path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
 of the deterministic approach used to break the combinatorial.

### 2016-01-18 body/equilibrium.tex
nico + glossary

@@
 
-\subsubsection{Conditions for static balance.}
+\subsubsection{Conditions for static equilibrium.}
 We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:

### 2016-01-18 body/glossary.tex
nico + glossary

\newglossaryentry{quasi-flat}
{
  name=\textit{quasi-flat},
  description={refers to a contact surface for which the friction cone includes the direction opposite to the gravity}
}

\newglossaryentry{cluttered}
{
  name=cluttered,
  description={A class of contact planning problems that admit as a solution a sequence of contact configurations for which at least one contact occurs with a \gls{quasi-flat} surface}
}

\newglossaryentry{ccontact}
{
  name=$C_{Contact}$,
  description={Set of collision free full body configurations such that at least one end-effector is in contact with the environment}
}

\newglossaryentry{ccontact0}
{
  name=$C_{Contact}^0$,
  description={Set of \textit{contact feasible} root configurations $\mathbf{q}^{0}$, that verify $\exists \mathbf{q}^{\overline{0}},  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact}$}
}

\newglossaryentry{contact feasible}
{
  name=\textit{contact feasible},
  description={Refers to a root configuration that can be extended into a collision free contact configuration}
}

\newglossaryentry{equilibrium feasible}
{
  name=\textit{equilibrium feasible},
  description={Refers to a root configuration that can be extended into a collision free configuration in static equilibrium}
}

\newglossaryentry{ccequil}
{
  name=$C_{equil}$,
  description={Set of collision free full body configurations that are in static equilibrium}
}

\newglossaryentry{ccequil0}
{
  name=$C_{equil}^0$,
  description={Set of \textit{equilibrium feasible} root configurations $\mathbf{q}^{0}$, that verify $\exists \mathbf{q}^{\overline{0}},  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{equil}$}
}

\newglossaryentry{reachc}
{
  name=\textit{reachability condition},
  description={A root configuration $\mathbf{q}^{0}$ that verifies :\\ $W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0_{s^*}(\mathbf{q}^{0}) \cap O = \emptyset$,
  where $s^*$ is a user-defined value}
}

\newglossaryentry{creach0}
{
  name=$C_{reach}^0$,
  description={Set of root configurations  $\mathbf{q}^{0}$ that verify the \gls{reachc}}
}

\newglossaryentry{interactive}
{
  name=interactive,
  description={We say that a planning method is \gls{interactive} when the computation time for one step is lesser than the
time to execute it. We arbitrarily approximate this time to one second.}
}

### 2016-01-18 body/intro.tex
nico + glossary

@@
 % !TEX root =  ../main.tex
-We consider the problem of planning an acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment.
+We consider the problem of planning an acyclic sequence of contacts describing the motion of a multiped robot in a constrained environment.
 Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to maintain the equilibrium of the system.
@@
 Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. However, executing
-this behavior on cluttered environments is dangerous, if not impossible.
+this behavior on constrained environments is dangerous, if not impossible.
 In an analysis of their participation to the Darpa Robotics Challenge, \citeauthor{atkensondarpa}
@@
 Neither global nor local methods present satisfying performances, because planning simultaneously the robot trajectory and the contacts that allow
-its execution is too costly.
+its execution is a very hard problem.
 
-As suggested by \citeauthor{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner, while preserving the completeness of the search.
+As suggested in \cite{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner, while preserving the completeness of the search.
 We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot path.
 This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
-This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely fast.
+This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely rapidly.
 while preserving a high success rate in the demonstrated scenarios.
@@
 
-The issue of planning acyclic contacts was first completely described by \citeauthor{Bretl:2006:MPM:1124573.1124585} in their seminal paper.
-The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: Planning a discrete sequence of acyclic contact configurations along the path (A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).
+The issue of planning acyclic contacts was first completely described by Bretl et al.
+in their seminal paper  \citep{Bretl:2006:MPM:1124573.1124585}.
+The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: Planning a discrete sequence of acyclic contact configurations along the path.
+A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence.
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential paths.
@@
 the contacts manually for more complex interactions.
-A major drawback of these optimization based approaches is thus that they only offer local convergence when applied to acyclic contact planning.
+In addition to the practical limits of the current implementations, a major drawback of these optimization based approaches is thus that they only offer local convergence when applied to acyclic contact planning.
 
@@
 a discretized environment. This planning phase does not account for collisions, involving that replanning is required in case of failure. This approach is efficient 
-in the scenarios demonstrated. In extremely constrained cases such 
+in the scenarios demonstrated. In highly constrained cases such 
 as the car egress scenario we address here, we believe that including collision constraints in the planning is a requirement.
@@
 use black box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
-at interactive frame rates. While this new approach provides an exciting direction of research, currently the resulting motions
+at \gls{interactive} frame rates. While this new approach provides an exciting direction of research, currently the resulting motions
 look unnatural, and do not seem applicable to real robots.
 
-Interpolation between contact poses ($\mathcal{P}_3$) remains an unsolved problem,
-though recent contributions have been able to bring partial solutions to it \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
-\citeauthor{Carpentier2016} are able to achieve this with real-time performances on the real robot.
+Regarding robotics applications, it thus appears that optimization based approaches outperform existing planners,
+partly because the interpolation between contact poses ($\mathcal{P}_3$) remains an unsolved problem.
+However recent contributions have been able to bring partial solutions to it \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
+\citeauthor{Carpentier2016} are able to achieve this with \gls{interactive} performances on the real robot.
 To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly.
@@
 \subsection{Rationale}
-This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
-we believe that the separation between the generation of the guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
-However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated. \\
-
-\noindent \textbf{Regarding $\mathcal{P}_1$}, the guide path must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces).
-We call this property \textit{true feasibility}.
-This property has not been studied yet; the only way to validate a waypoint in the path is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
-\\
-
-\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinite combinations of possible contact sequences for a given root path.
-The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious transitions).
-\\
-
-%~ In this paper, we address both questions.
-%~ For each problem, we introduce an original formulation of the related theoretical issues.
-%~ Then, we propose a concrete implementation of a solution, where the constraints of the problem are relaxed in favor of efficiency.
-
-The key idea of the paper is the theoritical and practical characterization of the space of \textit{truly feasible} configurations to address both issues.
-We first exhibit a low dimensional over-approximation of this space. This approximation is defined geometrically. 
-We make the strong assumption that these \textit{geometrically feasible} paths can be efficiently transformed into \textit{truly feasible} paths
-in cluttered environments. 
-Under this assumption we can take advantage of the decoupling approach.
-
-We first address \Pa by planning \textit{geometrically feasible} paths. We achieve this efficiently thanks to the \textit{reachability condition} that we introduce, that 
-does not require explicit contact computation.
-
-Then we address \Pb by extending the path into a \textit{truly feasible} sequence of contact configurations.
-This second step requires computing contact configurations verifying a condition for \textit{static equilibrium feasibility}, that requires
-explicit contact computation.
-
-This sequential approach is the key to the efficiency of our method, though it
-can result in failures because our assumption is not always true.
-However, we demonstrate empirically the validity of the approach: the high success rate of our method, combined with the low computation times we obtain, allows us to plan (and re-plan upon failure) multi contact sequences at interactive rates.
-
-\subsection{Paper contribution and organization}
-
-By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
-\begin{itemize}
-\item First, we propose a theoretical characterization of an efficient approach to sampling-based planning of acyclic contacts;
-\item Then we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with interactive applications;
-\item Finally we propose three heuristics for contact generation. They bias the planner towards configurations that are robust regarding static equilibrium, or 
-efficient regarding the task being performed.
-\end{itemize}
-
-Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
-Since the interpolation of a complete path between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
-is the fastest multi-contact planning solution to our knowledge.
 
@@
   \caption{
-    Overview of our 2-stage framework.
-\Pa: Given a path request between the yellow and blue positions, a guide path is computed in the space of truly feasible guides $C_{reach}$.
-This is achieved by defining a geometric condition, the reachability condition (abstracted here with the transparent cylinders).
-\Pb: The path is extended into a discrete sequence of contact configurations using an iterative algorithm.}
+    Overview of our two-stage framework. Given a path request between start and goal positions (left image), \Pa is the problem of computing a guide path in the space
+    of \textit{equilibrium feasible} root configurations.
+We achieve this by defining a geometric condition, the \textit{reachability condition} (abstracted with the transparent cylinders on the middle image).
+\Pb is then the problem of extending the path into a discrete sequence of contact configurations using an iterative algorithm (right image).}
   \label{fig:framework}
 \end{figure*}
+This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
+we believe that the separation between the generation of the guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
+However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated. \\
+
+\noindent \textbf{Regarding $\mathcal{P}_1$}, the guide path must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces).
+We call this property \textit{equilibrium feasibility}.
+This property has not been studied yet; the only way to validate a waypoint in the path is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
+\\
+
+\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinite combinations of possible contact sequences for a given root path.
+The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to constrained environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, possibly leading to very tedious transitions).
+\\
+
+\textbf{The key idea of the paper is the \textit{reachability condition}}, an efficient approximation of the space of \gls{equilibrium feasible} configurations.
+We make the strong assumption that the root configurations in this approximation can be efficiently extended into full body configurations in static equilibrium.
+We know that this assumption is not always verified, but we provide empirical evidence that it is often
+the case in the class of problems we consider. We define this class as the set of \textit{\gls{cluttered} contact planning problems}.
+We consider that such problems can be solved with a contact sequence 
+where every configuration contains at least one \gls{quasi-flat} contact support \citep{Prete2016}. A \textit{quasi-flat} contact support denotes a contact
+with a surface where the friction cone contains the direction opposite to the gravity. 
+%~ This definition includes motions on flat or uneven terrains, but also more complex ones such as stair climbing, car egress, or standing up scenarios, that might require noncoplanar contacts.
+Under this assumption we can take advantage of the decoupling approach to find a solution rapidly.
+
+To address \Pa, we first use the \textit{reachability condition} to plan the guide path. This step is fast because the planning happens
+in a low dimensional space, and does not require explicit contact computation.
+
+Then we address \Pb by extending the path into a sequence of \gls{equilibrium feasible} contact configurations.
+This second step requires computing contact configurations verifying a condition for \textit{static equilibrium feasibility}.
+
+This sequential approach is the key to the efficiency of our method, though it
+can result in failures because our hypothesis is not always true. Similarly, in this paper we do not 
+prove that all the contact sequences produced by our planner can be interpolated ($\mathcal{P}_3$), although we have been able to do this
+for some of the sequences demonstrated \citep{Carpentier2016}.
+However, we demonstrate empirically the validity of the approach: the high success rate of our method, combined with the low computation times we obtain, allows us to plan (and re-plan upon failure) multi contact sequences at \gls{interactive} rates.
+
+\subsection{Paper contribution and organization}
+
+In this context, our method proposes contributions to both problems \Pa and \Pb.
+\begin{itemize}
+\item First, we propose a low dimensional, efficient sampling based approach to plan guide paths ;
+\item Then we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with \gls{interactive} applications;
+\item We propose three heuristics for contact generation. They bias the planner towards configurations that are robust regarding static equilibrium, or 
+efficient regarding the task being performed.
+\item Finally, we present statistical tests that empirically demonstrate the validity of our approach in the scenarios we consider. 
+\end{itemize}
+
+Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
+Since the interpolation of a complete path between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
+is the fastest multi-contact planning solution to our knowledge.
 
@@
 Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb.
-In Section~\ref{sec:heuristics}, we present two heuristics used for the selection of a contact configuration.
+In Section~\ref{sec:heuristics}, we present heuristics used for the selection of a contact configuration.
 Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
@@
 The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
-This extension completes this work with a discussion on the implementation of this theory to real-world robots and problems.
-In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot despite bounded error in the contact forces. We also provide details on the HPP implementation
-of our contact generation algorithm.
-Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ~\citep{semini11hyqdesignjsce}.
-Appendix~\ref{app:rom} describes the generation process
-for the robot.
+This extension completes this work and brings new contributions, resulting from the implementation of our approach real-world robots and problems.
+In particular, a strong contribution is the introduction of a robust equilibrium criterion, designed to ensure the equilibrium of the robot despite bounded errors in the contact forces, coupled
+with heuristics for relevant contact selection. These two contributions were required to apply our method to real-robot models, namely HRP-2 and HyQ~\citep{semini11hyqdesignjsce}.
+As opposed to the conference paper, the complete algorithm for the contact planner is provided, and an access to the source code of our implementation is provided.
+Furthermore, our experiments are supported by a an in-depth analysis of our performance, as well as a discussion on the influence of the parameters to the method.

### 2016-01-18 body/overview.tex
nico + glossary

@@
 
+%~ \begin{table*}
+%~ \centering
+%~ \begin{tabular}{ c | >{\arraybackslash}m{200pt} | c}
+   %~ Robot configuration set &  \centering Definition: $\mathbf{q} \in$ set $\Rightarrow$?  & Root configuration set\\
+ %~ \hline
+ %~ \hline
+   %~ $C_{contact}$ & $\mathbf{q}$ collision free; at least one end effector in contact. &  $\mathbf{q}^{0} \in C_{contact}^0$ is ``contact feasible''.  \\
+ %~ \hline
+   %~ $C_{equil} \subset C_{contact} $ & $\mathbf{q}$  in static equilibrium &  $\mathbf{q}^{0} \in C_{equil}^0$ is ``equilibrium feasible''. \\
+ %~ \end{tabular}
+%~ \caption{TODO.}
+%~ \label{tab:glossary}
+%~ \quad
+%~ \end{table*}
+
 Figure~\ref{fig:framework} illustrates our workflow.
-\Pa and \Pb are addressed in a sequential fashion.
-The guarantees provided by the \textit{geometrically feasible} guide path allow
-us to avoid the combinatorial by generating the contacts in a straightforward manner.
+\Pa and \Pb are addressed in a sequential fashion: we first plan a guide path with the \textit{reachability condition}, before
+extending it into a sequence of contact configurations in static equilibrium.
+In this Section we describe this workflow and give some notations used throughout the paper. Some key concepts are introduced,
+the definitions of which can be found in the glossary at the end of this paper.
 
@@
   \caption{Generation of a contact configuration for the right leg of HRP-2.
-a: Selection of reachable obstacles.
-b: Entries of the limb samples database (with $N = 4$).
-c: With a proximity query on the octree database, configurations too far from obstacles are eliminated.
-d: The best candidate according to a user-defined heuristic $h$ is chosen.
-e: The final contact is achieved using inverse kinematics.}
+(a): Selection of reachable obstacles.
+(b): Entries of the limb samples database (with $N = 4$).
+(c): With a proximity query on the octree database, configurations too far from obstacles are eliminated.
+(d): The best candidate according to a user-defined heuristic $h$ is chosen.
+(e): The final contact is achieved using inverse kinematics.}
   \label{fig:contact_gen}
 \end{figure*}
+
 \subsection{Computation of a guide path --- \Pa}
 We first consider the problem of planning a relevant guide path.
-The objective is to compute a path of root placements that will allow contact creation.
-A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide path, but at the same time, any computed guide path must be \textit{truly feasible}, i.e.
-must lead to a valid sequence of contacts.
+The objective is to compute a path of root configurations that will allow contact creation.
+A desired requirement is to preserve the completeness of the planner: it should be able to explore any valid guide path, but at the same time, any computed guide path must lead to a sequence of contact configurations in static equilibrium.
 
-More precisely, a root placement is \textit{truly feasible} if and only if there exists at least one joint configuration such that the robot satisfies three conditions:
+More precisely, a root configuration is \gls{equilibrium feasible} if and only if there exists at least one joint configuration such that the robot satisfies two conditions:
 \begin{description}
-\item[C1] is not in collision with the environment, 
-\item[C2] has at least one end-effector in contact with the environment, 
-\item[C3] is in static equilibrium. 
+\item[C1]: is not in collision with the environment, 
+%~ \item[C2] has at least one end-effector in contact with the environment, 
+\item[C2]: is in static equilibrium. 
 \end{description}
-C1 and C2 define \textit{geometrical feasibility}, and can be efficiently approximated even without computing the contact points.
-C3 defines \textit{equilibrium feasibility}.
-C3 is difficult to approximate and requires the computation of the contact points, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
-However our intuition suggests us that, for cluttered environments, verifying C3 is not critical for $\mathcal{P}_1$, meaning that, most of the times, root placements satisfying C1 and C2 would satisfy also C3.
-For this reason we neglect C3 during the resolution of \Pa, and we worry about it only when solving \Pb.
+A requirement for C2 is the condition C3: at least one end-effector must be in contact with the environment.
+C1 and C3 define \glslink{contact feasible}{\textit{contact feasibility}}, and can be efficiently approximated even without computing the contact points.
+C2 defines \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
+C2 is difficult to approximate and requires the computation of the contact points, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
 
-An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right arm in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
-We define formally $C_{reach}$, the set of root placements satisfying C1 and C2, in which we compute a guide path with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
-Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
-%
-Details are presented in Section~\ref{rbprm}.
+We make the assumption that for the \textit{\gls{cluttered} contact planning problem} (that we define as admitting a solution where every configuration contains at least one contact
+is \gls{quasi-flat}) verifying C2 is not critical for $\mathcal{P}_1$.
+This means that root configurations satisfying C1 and C3 also satisfy C2. For this reason we neglect C2 during the resolution of \Pa, and we worry about it only when solving \Pb.
+
+We define $C_{Contact}^0$ as the set of \gls{contact feasible} root configurations (that can satisfy C1 and C2).
+Planning in $C_{Contact}^0$ boils down to planning in $\mathbb{R}^r$, which has an acceptable practical complexity.
+
+An intuitive description of \gls{contact feasible} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
+We approximate $C_{Contact}^0$ with a set $C_{reach}^0$, defined by the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
+In Section~\ref{rbprm} we detail how  $C_{reach}^0$ is sampled to compute a guide path with a dedicated planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
 
@@
 
-The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--\Pb). In this
-step the input guide path is \textit{geometrically feasible}.
+The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--\Pb).
 With a fixed root position, the dimensionality of the 
@@
 
-From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root placement, an inverse-kinematics solver 
-is used to maintain the contacts that were previously existing. Often, these contacts cannot be maintained, because of joint limits or collisions.
-In such a case the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
+From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root configuration, an inverse-kinematics solver 
+is used to maintain the contacts that were previously existing. Possibly, some of these contacts cannot be maintained (because of joint limits or collisions).
+Then the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
 The algorithm is designed so that only one contact can be created or broken between two successive configurations. This ensures that the interpolation between two configurations is achievable.
@@
 
-To select the ``most relevant'' contact configurations, user-defined heuristics can be chosen.
-In Section~\ref{sec:heuristics} we present extensively two such heuristics, which have been used in our experiments.
+There is a combinatorial issue in the choice of which contact to create,
+and where to create it.
+This complexity is avoided by selecting the ``most relevant'' contact configurations with user-defined heuristics.
+In Section~\ref{sec:heuristics} we present extensively those heuristics that have been used in our experiments.
 
@@
 
-\begin{figure}
-  \centering
-  \begin{overpic}[width=0.8\linewidth]{figures/character}
-    \put (7,29) {\small{$\in \mathbf{q}^1$}}
-    \put (7,25) {\small{effector}}
-  \end{overpic}
-  \caption{
-    Left: Robot in a rest configuration.
+%~ \begin{figure}
+  %~ \centering
+  %~ \begin{overpic}[width=0.8\linewidth]{figures/character}
+    %~ \put (7,29) {\small{$\in \mathbf{q}^1$}}
+    %~ \put (7,25) {\small{effector}}
+  %~ \end{overpic}
+  %~ \caption{
+    %~ Left: Robot in a rest configuration.
 The right arm is denoted as the limb $R^1$.
@@
 The green spheres are the reachable workspace of each limb, the  $W^k$.}
-  \label{fig:character}
-\end{figure}
+  %~ \label{fig:character}
+%~ \end{figure}
 
 \medskip
-\textbf{A robot} is a kinematic chain $R$, comprising \mbox{$n + 6$} degrees of freedom (DOFs).
+\textbf{A robot} is a kinematic chain $R$, comprising \mbox{$n + r$} degrees of freedom (DOFs), with $r \geq 6$. $r$ denotes the world coordinates of the root of the robot,
+as well as extra DOFs not related to the limbs (For instance, HRP-2 has two extra DOFS in the torso, such that we have $r=8$. 
 $R$ is composed of $l$ limbs $R^k, 1 \leq k \leq l$, attached to a root.
-It is described by a configuration $\mathbf{q} \in SE(3) \times \mathbb{R}^n$.
+It is described by a configuration $\mathbf{q} \in  \mathbb{R}^r \times \mathbb{R}^n$.
 We define some relevant projections of $\mathbf{q}$:
 \begin{itemize}
-	\item $\mathbf{q}^k$ denotes the configuration (a vector of joint values) of the limb $R^k$ (Figure~\ref{fig:character});
+	\item $\mathbf{q}^k$ denotes the configuration (a vector of joint values) of the limb $R^k$; % (Figure~\ref{fig:character});
 	\item $\mathbf{q}^{\overline{k}}$ denotes the vector of joint values of R \textbf{not} related to $R^k$.
 We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
-	\item $\mathbf{q}^{0}\in SE(3)$ denotes the position and orientation of the root of the robot $R$.
+	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ denotes the position and orientation of the root of the robot $R$.
 \end{itemize}
@@
 \medskip
-The volume encompassing the trunk of the robot is denoted $W^0$ (Figure~\ref{fig:character}-right: central cylinder).
-The reachable workspace of a limb $R^k$ is denoted $W^k$ (Figure~\ref{fig:character}-right: the four ellipses).
+The volume encompassing the trunk of the robot is denoted $W^0$
+ %~ (Figure~\ref{fig:character}-right: central cylinder). 
+ The reachable workspace of a limb $R^k$ is denoted $W^k$ %(Figure~\ref{fig:character}-right: the four ellipses). 
 \begin{equation}

### 2016-01-18 body/rbprm.tex
nico + glossary

@@
 
-We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of a multiped robot, connecting user-defined start and goal configurations.
-As said in the previous section, the goal is to enforce that any configuration $\mathbf{q}^0$ of the guide is geometrically feasible, i.e.
-can be mapped to a configuration in contact.
-We denote by $C_{contact} \subset SE(3) \times \mathbb{R}^n$ the contact sub-manifold of the robot, that is the set of configurations for which the robot is in contact with the environment.
-%~ We say that a root placement $\mathbf{q}^{0}$ is \textit{truly feasible} if there exists $\mathbf{q}^{\overline{0}}$ such that 
-%~ $$\mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact} $$
-%~ The set of all truly-feasible root placements is denoted by $C_{reach}$. By extension, a trajectory $\mathbf{q}^0(t)$ is truly feasible if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{reach}$. 
-%~ We define a bijective mapping $\pi$ between $C_{reach}$ and $C_{contact}$:
-%~ \begin{equation}
-    %~ \pi\colon \quad \mathbf{q}^{0} \in C_{reach} \longrightarrow  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact} 
-    %~ \label{eq:pi}
-%~ \end{equation}
+We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow \mathbb{R}^r$ for the root of a multiped robot, connecting user-defined start and goal configurations (.
+As said in the previous section, the goal is to enforce that the guide can be discretized into a sequence of \gls{contact feasible} configurations.
+We denote by \gls{ccontact} the contact sub-manifold of the robot.
 
-We say that a root placement $\mathbf{q}^{0}$ is \textit{geometrically feasible} if there exists at least one bijective mapping $\pi$
-such that
+We say that a root configuration $\mathbf{q}^{0}$ is \gls{contact feasible} if it belongs to the set:
 \begin{equation}
-    \pi\colon \quad \mathbf{q}^{0} \in  SE(3) \longrightarrow  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact} 
+      C_{contact}^0= \{ \mathbf{q}^{0}: \exists \mathbf{q}^{\overline{0}}, \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact} \}
     \label{eq:pi}
@@
 
-The set of all \textit{geometrically feasible} root placements is denoted by $C_{reach}$.
-By extension, a path $\mathbf{q}^0(t)$ is \textit{geometrically feasible} if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{reach}$.
+%~ The set of all \textit{geometrically feasible} root configurations is denoted by $C_{reach}$.
+By extension, a path $\mathbf{q}^0(t)$ is \gls{contact feasible} if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{contact}^0$.
 
-For this stage, and for this stage only, we make the assumption that in cluttered environments, a \textit{geometrically feasible} path can be transformed into a \textit{truly feasible} path.
-Under this assumption, for a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide path generator: all the generated paths must be \textit{geometrically feasible} (sufficient condition); the guide planner must be complete, i.e.
-it must be able to discover any \textit{geometrically feasible} path (necessary condition).
+For this stage, and for this stage only, we make the assumption that for \gls{cluttered} problems, a \gls{contact feasible} path can be transformed into a \gls{equilibrium feasible} path.
+Under this assumption, for a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide path generator: all the generated paths must be \gls{contact feasible} (sufficient condition); the guide planner must be complete, i.e.
+it must be able to discover any \textit{contact feasible} path (necessary condition).
 
-In this Section we characterize $C_{reach}$ theoretically with a necessary and a sufficient condition for belonging to the set.
-We then describe our implementation of an efficient RRT planner in $C_{reach}$.
+In this Section we characterize \gls{ccontact0} theoretically with a necessary and a sufficient condition for belonging to the set.
+We then approximate \gls{ccontact0} as a space $C_{reach}$, where we can efficiently sample configurations.
+We finally describe our implementation of an efficient RRT planner in $C_{reach}$.
 
@@
 
-\subsection{Conditions for geometrical feasibility}
-By default, \textit{geometrical feasibility} implies a constructive demonstration by exhibiting a valid $\pi$.
+\subsection{Conditions for contact feasibility}
+By default, \glslink{contact feasible}{\textit{contact feasibility}} implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$.
 This is the approach chosen by \cite{Bouyarmane2009}.
-However, computing a valid  $\mathbf{q}^{\overline{0}}$ is costly in practice.
-In this section we rather define a necessary condition and a sufficient condition for \textit{geometrical feasibility} that do not require this explicit computation.
+However, computing $\mathbf{q}^{\overline{0}}$ is expensive in terms of computation time.
+In this section we rather define a necessary condition and a sufficient condition for \glslink{contact feasible}{\textit{contact feasibility}} that do not require this explicit computation.
 
-% and  present the parametrization of a compromise condition used to approximate $C_{reach} \subset SE(3)$, the set of truly feasible configurations.
-This compromise condition, which we call the reachability condition, is used to compute a relevant guide trajectory using a sampling based motion planner, the Reachability-Based PRM (RB-PRM).
-
-\subsubsection*{Geometrical feasibility: necessary condition}
+\subsubsection*{Contact feasibility: necessary condition}
 For a contact to be possible, a volume $O_i \in O$ necessarily intersects with the reachable workspace $W(\mathbf{q}^{0})$ (Figure~\ref{fig:contact_gen}--1).
@@
 
-Therefore we can approximate $C_{reach}$ with a set $C_{reach} \subset C_{reach}^1$ with the reachability condition defined as: 
+Therefore we can define an optimistic approximation  $C^0_{Nec} \supset$ \gls{ccontact0} defined as: 
 \begin{equation}
-C_{reach}^1 = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+C_{Nec}^0 = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \wedge & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}
 \end{equation}
-It is straightforward to prove that  $C_{reach} \subset C_{reach}^1$ (by construction of the included set). 
-This inclusion is very important: it directly implies that any motion-planning algorithm with a guaranty of completeness in $SE(3) \times \mathbb{R}^n$ is complete  in $C_{reach} \times \mathbb{R}^n$.
-This is a strong reduction of the search space, which can be directly applied to any existing method.
+The inclusion \gls{ccontact0} $\subset C^0_{Nec}$ is straightforward, but it is very important.
+Planning in $C^0_{Nec}$ allows a strong reduction of the search space, while not discarding 
+any root configuration that could lead to a solution.
 
-The condition $C_{reach}^1$ is only necessary which means that one such root placement might not be truly feasible: in practice it is not guaranteed to find a valid sequence of contacts for every guide path in $C_{reach}^1$.
+The condition defining $C^0_{Nec}$ is only necessary. This means that it might not be possible to extend a guide path planned 
+in $C^0_{Nec}$ into a sequence of contact configurations.
 
-\subsubsection*{Geometrical feasibility: sufficient condition}
+\subsubsection*{Contact feasibility: sufficient condition}
 %~ In building $C_{reach}^1$, we directly considered the including hull of the root body $W^{0}$, obtaining a necessary condition. 
-A trivial sufficient condition for \textit{geometrical feasibility} can be constructed as a variation of $C_{reach}^1$, by replacing $W^0$ with a bounding volume $B^{max}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
-We denote by \mbox{$C_{reach}^\infty \subset C_{reach}$} the set of root placements corresponding to the sufficient condition.
+A trivial sufficient condition for \glslink{contact feasible}{\textit{contact feasibility}} can be constructed as a variation of $C^0_{Nec}$, by replacing $W^0$ with a bounding volume $B^{Suf}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
+We denote by \mbox{$C_{Suf}^0 \supset $ \gls{ccontact0}} the set of root configurations corresponding to this sufficient condition.
 
 In general, the inclusion is strict, which means that we lose the completeness of the two-stage contact planner (i.e.
-the planner is not able to discover a path inside \mbox{$C_{reach} \setminus C_{reach}^\infty$}).
-However, the sufficient condition guarantees that any such path leads to a valid sequence of contacts (i.e.
-$\pi$ is defined).
+the planner is not able to discover a path inside \mbox{$C_{reach} \setminus C_{Suf}^0$}).
+However, the sufficient condition guarantees that any such path leads to a valid sequence of contacts.
 
-\subsection{Geometrical feasibility: a compromise reachability condition}
+\subsection{Contact feasibility: a compromise \gls{reachc}}
 The sufficient condition is not interesting in practice since it leads the solver to lose too many interesting paths.
-The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \textit{geometrically feasible} in practice.
-It might be possible to find a shape $B$ that is necessary and sufficient; however, it seems intuitively very unlikely in general.
-The construction of a shape $W^0 \subset B \subset B^{max}$ leading to a necessary and sufficient condition (or the proof of its nonexistence) is out of the scope of this work.
+The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \gls{contact feasible} in practice.
+It might be possible to find an ideal shape $B$ that is necessary and sufficient; however, it seems intuitively very unlikely in general.
+The construction of a shape $W^0 \subset B \subset B^{Suf}$ leading to a necessary and sufficient condition (or the proof of its nonexistence) is out of the scope of this work.
 
-However, between $W^0$ and $B^{max}$, a trade-off can be found between a necessary and a sufficient condition.
+However, between $W^0$ and $B^{Suf}$, a trade-off can be found between a necessary and a sufficient condition.
 We define $W^0_s$ as the volume $W^0$ subject to a scaling transformation by a factor $s \in \mathbb{R}^+$.
 %
-We then consider the spaces $C_{reach}^s$
+We then consider the spaces $C_{s}^0$
  \begin{equation}
-C_{reach}^s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
- %~ & \wedge & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}
+C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+ %~ & \text{ and } & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}
 \end{equation}
 %
-The higher $s$, the closer this ``reachability condition'' is to being sufficient, and if $s=1$, the planner is complete.
-The parametrization of $s$ allows to find a trade-off between these two desirable properties.
+If $s=1$, then $W^0_s$ = $W^0$, such that $C_1^0$ = $C_{Nec}$. We thus define $s \in \mathbb{R}^+$, since values inferior to $1$ can only introduce more errors.
+The higher $s$, the closer this \gls{reachc} is to being sufficient, and if $s=1$, the search is complete (in the sense that all \gls{contact feasible} configurations can be found).
+The parametrization of $s$ defines a trade-off between these two desirable properties.
+$s$ can be chosen by hand, or automatically as explained in section TODO.
+In the rest of the paper, we consider that an appropriate value $s^*$ of $s$ is chosen, and define \gls{creach0} $= C^0_{s^*}$.
+
 As an example, Figure~\ref{fig:HyQ_roms} presents the volumes $W$ computed for the HyQ robot.
@@
 
-\subsection{Computing the guide path in $C_{reach}^s$}
-Once a value of $s$ has been fixed, any sampling-based motion planner can be used to plan a path in $C_{reach}^s$. 
-Indeed, contrary to $C_{Contact}$,  $C_{reach}^s$ has a non zero measure in the configuration space $C$. Therefore a standard uniform sampling approach
+\subsection{Computing the guide path in $C_{reach}^0$}
+Any sampling-based motion planner can be used to plan a path in \gls{creach0}. 
+Indeed, contrary to \gls{ccontact}, \gls{creach0} has a non zero measure in the configuration space $C$. Therefore a standard uniform sampling approach
 can work, in spite of a high rejection rate. 
-Thus, the only significant change regarding a classical planner is to replace the classical collision checking method with a test of belonging to $C_{reach}^s$ when verifying
+Thus, the only significant change regarding a classical planner is to replace the classical collision checking method with a test of belonging to \gls{creach0} when verifying
 that drawn configurations and associated local paths are valid.
@@
 However, to improve the sampling efficiency while remaining probabilistically complete, we bias the sampling process to generate near obstacle configurations, similarly to~\cite{Amato98choosinggood}.
-First, a configuration is set to a random point on the surface of one randomly chosen obstacle. The root location is then translated and rotated randomly until the reachability condition is satisfied.
-Our current implementation of these modifications is based on the Bi-RRT planner provided by the HPP software.
+First, a configuration is set to a random point on the surface of one randomly chosen obstacle. The root location is then translated and rotated randomly until the \gls{reachc} is satisfied.
+Our current implementation of these modifications is based on the Bi-RRT planner \citep{770022} provided by the HPP software.
 
-Thanks to this small set of modifications, the high dimensional truly feasible path planning problem is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for HyQ, 8 for HRP-2 that has joints in the torso).
-In so doing we make the problem tractable for sampling-based approaches, with extremely efficient computation times.
+Thanks to this set of modifications, the problem of planning an \gls{contact feasible} path in a high dimensional space is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for HyQ, 8 for HRP-2 that has joints in the torso).
+In so doing we make the problem tractable for sampling-based approaches, with \gls{interactive} computation times.

### 2016-01-18 body/results.tex
last biblio inclusion
m
nico + glossary

@@
 Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
-For each scenario we indicate the chosen heuristics. We also provide a performance analysis, which shows that the planner is compatible with interactive applications,
+For each scenario we indicate the chosen heuristics. We also provide a performance analysis, which shows that the planner is compatible with \gls{interactive} applications,
 and present the success rates obtained in each scenario. Moreover, we demonstrate the interest of our robustness criterion in the different computed poses.
@@
 
-We say that the planning is interactive when the computation time for one step is lesser than the
+We say that the planning is \gls{interactive} when the computation time for one step is lesser than the
 time to execute it. We arbitrarily approximate this time to one second.
@@
 In all the scenarios considered, the formulation of the problem is always the same:
-a start and goal root placements are provided as an input of the scenario.
+a start and goal root configurations are provided as an input of the scenario.
 The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
@@
 
-The goal is to climb three 20-cm high steps. This height requires HRP-2 to use a ramp to perform the task.
+The goal is to climb three 15-cm high steps. This height requires HRP-2 to use a ramp to perform the task.
 
@@
 for the computed path. The reason for this is not reachability, but equilibrium. The planning is slower than for the stair scenario (because the contact generation fails more),
-though it remains compatible with interactive performances. % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
+though it remains compatible with \gls{interactive} performances. % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
 
@@
 collide with the environment at all times. %\adnote{Why do not relax this constraint and require only 3 legs?} 
-Again, the computed time remains however interactive.
+Again, the computed time remains however \gls{interactive}.
 
@@
 The difficulty is illustrated in Figure~\ref{fig:HyQ_obs}, where several feet rearrangements are required to cross the hole (although the video shows this best).
-The planner however succeeds in finding a feasible sequence in the end, again with interactive computation times.
+The planner however succeeds in finding a feasible sequence in the end, again with \gls{interactive} computation times.
 
@@
 This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
-our contact planner to compute a guide trajectory for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
+our contact planner to compute a guide path for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
 Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences in less than 5 seconds.
@@
    Car (61)& 1 -- \textbf{529} --  995 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6775 --  \textbf{20890} -- 52963 & \textbf{342} \\
-   Rubble (71)& 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{30} -- 13126 & \textbf{43} \\
+   Rubble (71)& 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{3058} -- 13126 & \textbf{43} \\
    Race (112)& 1 -- \textbf{501} --  995 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2530 --  \textbf{3722} -- 10030 & \textbf{33}
@@
 \begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | c}
-  &  Path planning & True feasibility & Kinematic failure & Equilibrium failure \\
+  &  Path planning & Equilibrium feasibility & Kinematic failure & Equilibrium failure \\
  \hline
@@
 \caption{Success rates for each scenario. The first value indicates the percentage of successful complete contact plannings for 1000 tests; The second value
-indicates the percentage of truly feasible root configurations: considering each limb individually, indicates the percentage of root placements of the guide trajectory that
+indicates the percentage of equilibrium feasible root configurations: considering each limb individually, indicates the percentage of root configurations of the guide path that
 led to a feasible contact. Kinematic failure is the percentage of contact generations that failed because no collision-free candidate was found. Equilibrium failure is the percentage of contact
@@
 In all scenarios, one can observe that the average computation time for one single step is largely below the second,
-thus allowing to consider interactive applications.
+thus allowing to consider \gls{interactive} applications.
 

### 2016-01-18 body/overview.tex
andrea sauf abstract

@@
 We approximate $C_{Contact}^0$ with a set $C_{reach}^0$, defined by the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
-In Section~\ref{rbprm} we detail how  $C_{reach}^0$ is sampled to compute a guide path with a dedicated planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
+Our objective is to find a root path such that the root (scaled by a factor $s>1$) is not in collision, while the reachable workspace is in collision with the environment.
+In Section~\ref{rbprm} we detail how  $C_{reach}^0$ is designed and sampled to compute a guide path with a dedicated planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
 

### 2016-01-19 body/appendix_roms.tex
complete, need abstract and section 4

@@
 First, we compute the bounding boxes of the robot trunk, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
-Then, we perform a scaling of these boxes by a factor $s$. For different values of $s$, we verify the geometrical feasibility condition:
-For a large number of configurations that verify the \gls{reachc}, we try to generate contact configurations. We then
-empirically select the lowest value of $s$ that provides a reasonnable success rate in all scenarios.
-For HRP-2, we choose $s=1.2$. This leads to a success rate higher than 86 \% for all scenarios.
+Then, we perform a scaling of these boxes by a factor $s$. 
+The higher $s$ is, the more likely sampled configurations are to be feasible, but the less complete is the approach.
+To compute the appropriate value of $s$, we proceed as described in Section~\ref{sec:params}, and choose empirically
+$s^*=1.2$ as the appropriate value for HRP-2.

### 2016-01-19 body/glossary.tex
complete, need abstract and section 4

@@
 {
-  name=cluttered,
+  name=\textit{cluttered},
   description={A class of contact planning problems that admit as a solution a sequence of contact configurations for which at least one contact occurs with a \gls{quasi-flat} surface}
@@
 {
-  name=interactive,
-  description={We say that a planning method is \gls{interactive} when the computation time for one step is lesser than the
-time to execute it. We arbitrarily approximate this time to one second.}
+  name=\textit{interactive},
+  description={We say that a planning method is interactive when the computation time for one step is lesser than the
+time to execute it. We arbitrarily approximate this time to one second}
 }

### 2016-01-19 body/results.tex
complete, need abstract and section 4

@@
 
+\subsection{Influence of the parameters} \label{sec:influence}
+We discuss the several factors can influence the outcome of our planner: the root scaling factor $s$, the choice of heuristics for
+contact generation, the parametrization of the static equilibrium robustness, and lastly, the discretization 
+of the guide path.
+
+\subsubsection{Choosing the scaling factor $s$} \label{sec:params}
+To find a convenient value $s^*$ of $s$, we proceeded as follows. 
+For several values of $s$, we generated $10 000$ configurations. 
+We then computed the rates of false positives (ie the resulting configuration satisfies
+the reachability condition, but does not belong to \gls{ccontact0}) and false negatives (ie the resulting configuration does not satisfy the 
+reachability condition, but belongs to \gls{ccequil0}).
+
+The obtained results for HRP-2 are summarized in Table~\ref{tab:scale}, averaged over all scenes (except for the car egress: in this scenario, 
+statistical tests are not really conclusive since we are only interested in a small area of the environment).
+As it can be expected, the scaling results in a high increase of the false negative, while the false positive decreases.
+For HRP-2 we decided to set $s^*=1.2$, that results in less than 3 \% of false positives, and was also effective
+for the car egress scenario.
+
+\begin{table}
+\centering
+\footnotesize
+\begin{tabular}{c | c | c}
+   Value of $s$ &  False positive & False negative\\
+ \hline
+   1   & 24\% & 0 \%\\
+   1.1& 12\% & 4\% \\
+   1.15& 07\% & 6\%\\
+   1.2 & 03\% & 7.5\%\\
+   1.25& 02\% & 8.3\%\\
+   1.5 & 01\% & 9.5\%\\
+ \end{tabular}
+\caption{Percentage of true and false positive in the reachability condition, depending on the scaling value of $W^0$.}
+\label{tab:scale}
+\quad
+\end{table}
+
+\subsubsection{On the choice of heuristics} \label{sec:heuristics}
+In our conference paper, the computed motions were only generated using the EFORT heuristic.
+EFORT is designed for tasks requiring to exert important forces (such as pushing / pulling / climbing). 
+Regarding locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
+configurations close to singularities (and joint limits). While this does not significantly impact
+the generation of the plan, the resulting interpolation on the real robot turned out to be harder \citep{Carpentier2016}.
+For this reason, we now prefer to use our manipulability based heuristic for the legs of the robot, but we still
+use EFORT for the arm, that results in lesser contact repositionning.
+
+\subsubsection{parametrization of the robustness equilibrium criterion} \label{sec:parrob}
+Robustness is really important when considering practical applications on the robot, to account 
+for the various uncertainties that result from environment and state estimator. However,
+maximizing the robustness criterion is not necessarily optimal, because the resulting contacts will stay close to the center
+of mass, thus not favoring the motion. In our experiments, we choose not to maximize the robustness,
+but to empirically set a robustness treshold value under which a configuration is not considered to be in
+static equilibrium (If no candidate reaches the treshold, instead of failing, the algorithm can eventually return the ``more robust'' configuration
+found). Currently for HRP-2, a treshold value of 2 subjectively gives the best results in the considered scenarios, while 10 seems to be a good choice for HyQ. Both values do not
+significantly slow down the planning times.
+
+\subsubsection{Discretization of the guide path} \label{sec:disc}
+Currently, to discretize the path, the user defines a fixed step size. The step size
+has an influence on the output of the planner: if too large steps are taken,
+the planner may fail since we impose the constraint that only one contact change might occur
+between two consecutive stances. For instance, in most scenarios the torso of HRP-2 translates by about 15 cm between two postures, but about only 3 cm
+for the car egress scenario for instance, where we had to lower the step size because of the many contact breaks, induced by the extremely constrained geometry.
+For future work, we would like to automatically adapt the size of the discretization step to the complexity of the planning, for instance
+by reducing the step size when in the areas where failure occur.
+
 \subsection{Performance analysis} \label{sec:perf}
@@
  \hline
-   Steep stairs & 100\%  & 99.5\% & 0.11\% & 0.39\% \\
-   Standing up & 68\% & 88\% & 6\% & 6\% \\
-   Car egress & 39\% & 77\% & 21\% &  2\% \\
-   Rubble & 74\% & 97.87\% & 0.13\% & 2\% \\
-   Obstacle race & 58\% & 95.7\% & 1.8\% & 2.5\% \\
+   Steep stairs & 100\%  & 99.5\% & 0.1\% & 0.4\% \\
+   Standing up & 68\% & 87.8\% & 6.1\% & 6.1\% \\
+   Car egress & 39\% & 77.0\% & 21.0\% &  2.0\% \\
+   Rubble & 74\% & 97.9\% & 0.1\% & 2.0\% \\
+   Obstacle race & 58.0\% & 95.7\% & 1.8\% & 2.5\% \\
  \end{tabular}
-\caption{Success rates for each scenario. The first value indicates the percentage of successful complete contact plannings for 1000 tests; The second value
+\caption{Success rates for each scenario, rounded to the first decimal. The first value indicates the percentage of successful complete contact plannings for 1000 tests; The second value
 indicates the percentage of equilibrium feasible root configurations: considering each limb individually, indicates the percentage of root configurations of the guide path that
@@
 The most efficient (and immediate) approach is probably to launch in parallel several instances of the planner for a given problem (our current implementation is single threaded) and to apply any successful result.
+
+\begin{table*}
+\centering
+\begin{tabular}{ c | c | c | c | c }
+ Scenario & Method  & Guide Path & Contact sequence & Interpolation \\
+ \hline
+   \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{5.42 min}  \\
+							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{2 to 10 min} \\
+							 & \textbf{Ours} + \citeauthor{Carpentier2016}  & $\mathbf{<}$ \textbf{2 ms} & $\mathbf{<}$ \textbf{14 ms} & $ <$ 2s \\
+ \hline
+   \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{4.08 min}  \\
+							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{2 to 10 min} \\
+							 & \textbf{Ours}  & $\mathbf{<}$ \textbf{2 ms} & $\mathbf{<}$ \textbf{14 ms} & X \\
+ \hline
+   \multirow{3}{*}{Stair 40 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{10.08 min}  \\
+							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{2 to 10 min} \\
+							 & \textbf{Ours}  & \textbf{5 s} & \textbf{3 s} & X \\
+ \hline
+   \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 10 min & \multicolumn{2}{c}{3.5 hours}  \\
+							 & \textbf{Ours}  & \textbf{529 ms} & \textbf{$<$ 21 s} & X \\
+ \end{tabular}
+\caption{TODO VERIFIE STATS A LA MAISON.}
+\label{tab:compprev}
+\quad
+\end{table*}
+\subsection{Comparison with previous work} \label{sec:compa}
+Comparing our method with others is not trivial. Indeed, we focus on the contact planning aspect of the problem, and do not address
+the interpolation in this paper, nor provide a strong guaranty that our sequence can be interpolated.
+Papers in the state of the art, when they indicate the computation times, often only provide an overall computation
+time, and we can not afford the important development resources required to reimplement them.
+
+Furthermore, the interpolation itself is a problem that is far from being solved.
+For instance in \cite{DBLP:conf/iser/EscandeKMG08}, the end-effector trajectories are computed automatically with splines, but are manually corrected if they enter in collision.
+In \citep{Mordatch:2012:DCB:2185520.2185539}, internal joint collisions can occur in the resulting motion.
+Recent methods suffer from similar issues \citep{Carpentier2016}.
+
+Regardless, we indicate in Table~\ref{tab:compprev} what is actually computed by the method and in how much time.
+We compare motions obtained in previous works with HRP-2 (or a humanoid avatar) with ours: the two scenarios we selected consist in:  climbing one step, high of 20, 30 or 40 centimeters; 
+moving out of a table, that we consider of a complexity similar to the car egress scenario (or inferior since the feet location remain on the ground in the former case).
+For scenarios that apply, we also indicate the interpolation time that we obtain by feeding the plan to our interpolation method presented in \cite{Carpentier2016}.
+
+Despite the difficulty of comparing the methods, we believe the results presented in Table~\ref{tab:compprev} indicate that our planner holds the promise of planning complete trajectories significantly faster than previous works.

### 2016-01-20 body/equilibrium.tex
andrea sauf abstract
added missing figure on feasibility

@@
 According to Coulomb's law, the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
-As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
+As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays of unit length. We choose for instance:
 \begin{equation*}
@@
 In conclusion, the LP~\eqref{eq:dual} provides a method to assert rapidly static equilibrium, and returns a robustness value.
-This robustness criterion can be immediately transformed into a heuristic, and eventually be ponderated with others.
+This robustness criterion can be immediately transformed into a heuristic, and eventually be pondered with others.
 Another option to enforce robustness is to set a minimum acceptable value for $b_0$, but as a result this conservative approach is not complete.

### 2016-01-20 body/glossary.tex
added missing figure on feasibility

@@
 {
-  name=$C_{equil}$,
+  name=$C_{Equil}$,
   description={Set of collision free full body configurations that are in static equilibrium}
@@
 {
-  name=$C_{equil}^0$,
-  description={Set of \textit{equilibrium feasible} root configurations $\mathbf{q}^{0}$, that verify $\exists \mathbf{q}^{\overline{0}},  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{equil}$}
+  name=$C_{Equil}^0$,
+  description={Set of \textit{equilibrium feasible} root configurations $\mathbf{q}^{0}$, that verify $\exists \mathbf{q}^{\overline{0}},  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{Equil}$}
 }

### 2016-01-20 body/overview.tex
added missing figure on feasibility

@@
  %~ \hline
-   %~ $C_{equil} \subset C_{contact} $ & $\mathbf{q}$  in static equilibrium &  $\mathbf{q}^{0} \in C_{equil}^0$ is ``equilibrium feasible''. \\
+   %~ $C_{Equil} \subset C_{contact} $ & $\mathbf{q}$  in static equilibrium &  $\mathbf{q}^{0} \in C_{Equil}^0$ is ``equilibrium feasible''. \\
  %~ \end{tabular}
@@
 A desired requirement is to preserve the completeness of the planner: it should be able to explore any valid guide path, but at the same time, any computed guide path must lead to a sequence of contact configurations in static equilibrium.
+The sets we introduce are illustrated in Figure~\ref{fig:dedefeas}.
+
+\begin{figure}[t]
+\centering
+  \begin{overpic}[width=1\linewidth]{figures/2D_feas}
+		\put (15,) {$C_{Equil}^0$      $\subset$} 
+		\put (47,) {$C_{Contact}^0$ $\approx$ } 
+		\put (76,) {$C_{Reach}^0$} 
+		%~ \put (5,27) {3.b)} 
+		%~ \put (37,27) {4.a)} 
+		%~ \put (68,27) {4.b)} 
+	\end{overpic}
+\caption{Illustration of several configurations sets used in this paper in a 2D scene.
+Obstacles are violet, and units are in meters.
+To show the sets in a 2D representation, all the rotational joints of HRP-2 are locked in the shown configuration, such that a torso configuration
+is only described by two positional parameters (x and y).
+$C_{Equil}^0$ is included in $C_{Contact}^0$.
+$C_{Reach}^0$ approximates $C_{Contact}^0$.
+Depending on a parametrization, we can obtain $C_{Contact}^0 \subset C_{Reach}^0$.
+Considering the configurations around the top obstacle, we can observe a dramatic
+divergence between  $C_{Equil}^0$  and $C_{Contact}^0$ when the problem is not \textit{\gls{cluttered}}.}
+		   \label{fig:dedefeas}
+\end{figure}
 
@@
 C1 and C3 define \glslink{contact feasible}{\textit{contact feasibility}}, and can be efficiently approximated even without computing the contact points.
-C2 defines \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
+C2 defines \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}, that we describe with the set $C_{Equil}^0$ (Figure~\ref{fig:dedefeas} -- Green).
 C2 is difficult to approximate and requires the computation of the contact points, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
@@
 
-We define $C_{Contact}^0$ as the set of \gls{contact feasible} root configurations (that can satisfy C1 and C2).
+We define $C_{Contact}^0$ (Figure~\ref{fig:dedefeas} -- Orange) as the set of \gls{contact feasible} root configurations (that can satisfy C1 and C2).
 Planning in $C_{Contact}^0$ boils down to planning in $\mathbb{R}^r$, which has an acceptable practical complexity.
@@
 An intuitive description of \gls{contact feasible} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
-We approximate $C_{Contact}^0$ with a set $C_{reach}^0$, defined by the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
+We approximate $C_{Contact}^0$ with a set $C_{reach}^0$, defined by the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:dedefeas} -- Red).
 Our objective is to find a root path such that the root (scaled by a factor $s>1$) is not in collision, while the reachable workspace is in collision with the environment.
@@
 Then the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
-The algorithm is designed so that only one contact can be created or broken between two successive configurations. This ensures that the interpolation between two configurations is achievable.
+The algorithm is designed so that only one contact can be created or broken between two successive configurations.
+While does not provide guarantees that the interpolation between two configurations is achievable, it appears empirically as a reasonable heuristic.
 Details are presented in Section~\ref{sec:contact}.

### 2016-01-20 body/rbprm.tex
complete, need abstract and section 4
added missing figure on feasibility

@@
 
-We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow \mathbb{R}^r$ for the root of a multiped robot, connecting user-defined start and goal configurations (.
+We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow \mathbb{R}^r$ for the root of a multiped robot, connecting user-defined start and goal configurations.
 As said in the previous section, the goal is to enforce that the guide can be discretized into a sequence of \gls{contact feasible} configurations.
@@
 The parametrization of $s$ defines a trade-off between these two desirable properties.
-$s$ can be chosen by hand, or automatically as explained in section TODO.
+$s$ can be chosen by hand, or automatically as explained in Section~\ref{sec:params}.
 In the rest of the paper, we consider that an appropriate value $s^*$ of $s$ is chosen, and define \gls{creach0} $= C^0_{s^*}$.

### 2016-01-20 body/results.tex
added missing figure on feasibility

@@
   \caption{
-           Selected frames the car egress scenario. }
+           Selected frames from the car egress scenario. }
 		   \label{fig:car}
@@
 of mass, thus not favoring the motion. In our experiments, we choose not to maximize the robustness,
-but to empirically set a robustness treshold value under which a configuration is not considered to be in
-static equilibrium (If no candidate reaches the treshold, instead of failing, the algorithm can eventually return the ``more robust'' configuration
-found). Currently for HRP-2, a treshold value of 2 subjectively gives the best results in the considered scenarios, while 10 seems to be a good choice for HyQ. Both values do not
+but to empirically set a robustness threshold value under which a configuration is not considered to be in
+static equilibrium (If no candidate reaches the threshold, instead of failing, the algorithm can eventually return the ``more robust'' configuration
+found). Currently for HRP-2, a threshold value of 2 subjectively gives the best results in the considered scenarios, while 10 seems to be a good choice for HyQ. Both values do not
 significantly slow down the planning times.
@@
 \begin{tabular}{ >{\centering\arraybackslash}m{37pt} | >{\centering\arraybackslash}m{57pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{70pt} | >{\centering\arraybackslash}m{73pt} | >{\centering\arraybackslash}m{80pt} | >{\centering\arraybackslash}m{10pt}}
-  Scenario (nb steps) &  Complete guide generation & Static equilibrium & Collision & Inverse Kinematics  & Total generation time & Time per step\\
+  Scenario (nb steps) &  Complete guide generation (ms) & Static equilibrium (ms) & Collision (ms) & Inverse Kinematics (ms) & Total generation time (ms) & Time per step (ms)\\
  \hline
@@
 of our theoretical formulation. The extreme situation of the car egress scenario provides the less satisfying results in terms of kinematic failure. This is not
-suprising considering the narrowness of the scene.
+surprising considering the narrowness of the scene.
 From a pragmatic point of view, regarding the computation times, we claim that our approach provides a satisfying compromise between completeness and efficiency.
@@
 Comparing our method with others is not trivial. Indeed, we focus on the contact planning aspect of the problem, and do not address
-the interpolation in this paper, nor provide a strong guaranty that our sequence can be interpolated.
+the interpolation in this paper, nor provide a strong guarantee that our sequence can be interpolated.
 Papers in the state of the art, when they indicate the computation times, often only provide an overall computation

### 2016-01-20 body/abstract.tex
typos

@@
 We address the issue of automatically planning discrete sequences of contacts
-that characterize a multiped motion in an arbitrary cluttered environment.
+that characterize a multiped motion in a cluttered environment.
 To overcome the inherent combinatorial difficulty of this problem, we separate it
@@
 of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
-Then, we propose a sequential algorithm to extend a feasible guide path into a sequence of contacts. For each limb, an off-line sampling approach,
-independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
+Then, we propose a sequential algorithm to extend a feasible guide path into a sequence of contacts.
+A sampling based-approach is used to generate candidate configurations.
 The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 

### 2016-01-20 body/efort.tex
typos

@@
 Considering: a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
-the equivalent joint torque vector $\mathbf{\tau}$;
+the equivalent joint torque vector $\bm{\tau}$;
 we can define the mechanical work in both spaces:
 \begin{equation*} \label{power}
-\dot{\mathbf{q}}^T \mathbf{\tau} = \dot{\mathbf{p}}^T \mathbf{f}
+\dot{\mathbf{q}}^T \bm{\tau} = \dot{\mathbf{p}}^T \mathbf{f}
 \end{equation*}
@@
 \begin{equation*} \label{ballforce}
-||\mathbf{\tau}||^2 \leq 1
+||\bm{\tau}||^2 \leq 1
 \end{equation*}
@@
 \begin{equation*}
-f_\mathsf{T}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(JJ^{T})\mathbf{m}]^{-\frac{1}{2}}
+f_\mathsf{T}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}	^{T})\mathbf{m}]^{-\frac{1}{2}}
 \end{equation*}

### 2016-01-20 body/glossary.tex
typos

@@
   name=\textit{reachability condition},
-  description={A root configuration $\mathbf{q}^{0}$ that verifies :\\ $W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0_{s^*}(\mathbf{q}^{0}) \cap O = \emptyset$,
+  description={A root configuration $\mathbf{q}^{0}$ that verifies :\\ $W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0_{s^*}(\mathbf{q}^{0}) \cap O = \emptyset$,
   where $s^*$ is a user-defined value}

### 2016-01-20 body/intro.tex
typos

@@
 
-As suggested in \cite{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner, while preserving the completeness of the search.
+As suggested in \cite{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner.
 We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot path.
@@
 use black box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
-at \gls{interactive} frame rates. While this new approach provides an exciting direction of research, currently the resulting motions
+at \gls{interactive} frame rates (We say that a planning method is interactive when the computation time for one step is lesser than the
+time to execute it). While this new approach provides an exciting direction of research, currently the resulting motions
 look unnatural, and do not seem applicable to real robots.
@@
 Regarding robotics applications, it thus appears that optimization based approaches outperform existing planners,
-partly because the interpolation between contact poses ($\mathcal{P}_3$) remains an unsolved problem.
+partly because the real-time interpolation between contact poses ($\mathcal{P}_3$) remains an unsolved problem.
 However recent contributions have been able to bring partial solutions to it \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
@@
 with heuristics for relevant contact selection. These two contributions were required to apply our method to real-robot models, namely HRP-2 and HyQ~\citep{semini11hyqdesignjsce}.
-As opposed to the conference paper, the complete algorithm for the contact planner is provided, and an access to the source code of our implementation is provided.
+In the present paper, a complete algorithm for the contact planner is given, and an access to the source code of our implementation is provided.
 Furthermore, our experiments are supported by a an in-depth analysis of our performance, as well as a discussion on the influence of the parameters to the method.

### 2016-01-20 body/overview.tex
typos

@@
 	\end{overpic}
-\caption{Illustration of several configurations sets used in this paper in a 2D scene.
+\caption{Illustration of several root configurations sets used in this paper in a 2D scene.
 Obstacles are violet, and units are in meters.
@@
 is \gls{quasi-flat}) verifying C2 is not critical for $\mathcal{P}_1$.
-This means that root configurations satisfying C1 and C3 also satisfy C2. For this reason we neglect C2 during the resolution of \Pa, and we worry about it only when solving \Pb.
+This means that root configurations satisfying C1 and C3 also satisfy C2. For this reason we neglect C2 during the resolution of \Pa, and we consider it only when solving \Pb.
 
@@
 We approximate $C_{Contact}^0$ with a set $C_{reach}^0$, defined by the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:dedefeas} -- Red).
-Our objective is to find a root path such that the root (scaled by a factor $s>1$) is not in collision, while the reachable workspace is in collision with the environment.
+Our objective is to find a root path such that the root (scaled by a factor $s \geq 1$) is not in collision, while the reachable workspace is in collision with the environment.
 In Section~\ref{rbprm} we detail how  $C_{reach}^0$ is designed and sampled to compute a guide path with a dedicated planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
@@
 The algorithm is designed so that only one contact can be created or broken between two successive configurations.
-While does not provide guarantees that the interpolation between two configurations is achievable, it appears empirically as a reasonable heuristic.
+While it does not provide guarantees that the interpolation between two configurations is achievable, it appears empirically as a reasonable heuristic.
 Details are presented in Section~\ref{sec:contact}.
@@
 \textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains.
-Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.
+%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.

### 2016-01-20 body/results.tex
typos

@@
    1.1& 12\% & 4\% \\
-   1.15& 07\% & 6\%\\
-   1.2 & 03\% & 7.5\%\\
-   1.25& 02\% & 8.3\%\\
-   1.5 & 01\% & 9.5\%\\
+   1.15& 7\% & 6\%\\
+   1.2 & 3\% & 7.5\%\\
+   1.25& 2\% & 8.3\%\\
+   1.5 & 1\% & 9.5\%\\
  \end{tabular}
@@
 
-\subsubsection{On the choice of heuristics} \label{sec:heuristics}
+\subsubsection{On the choice of heuristics} \label{sec:heuristichoices}
 In our conference paper, the computed motions were only generated using the EFORT heuristic.

### 2016-01-20 cover_letter.tex
cover letter

@@
 
-The extension of our work addresses the main comments of the reviewers:
+Regarding the conference paper, this apper paper presents new contributions that we first highlight:
+
+\begin{itemize}
+\item We introduce a new criterion for asserting the static equilibrium of a robot configuration. It works for arbitrary
+legged robots, and handles any kind of contact configurations and surface orientation. It is formulated as a really efficient Linear Program, and 
+can be used to compute a value of robustness of the configuration. This robustness is useful for real robots applications, where many uncertainties
+in the state estimation of the robot can lead to incorrect evaluations of the equilibrium.
+\item We present two new heuristics that we use to bias our contact planner framework towards configurations relevant for the task being performed.
+\item We present a complete algorithm for contact planning that integrates the robust equilibrium criterion.
+\end{itemize}
+
+The extension of our work also addresses the main comments of the reviewers, and proposes lesser contributions,
+that we detail in the following lines.
 
@@
 comparison to the state of the art (Section 1.1); we clarify our statement regarding
-our computation times (Section 6 and 6.1.6); we make a clear distinction between the
+our computation times and compare our method with the state of the art (Section 6); we make a clear distinction between the
 theoretical concepts we propose and their approximation (Section 2).

### 2016-01-21 body/abstract.tex
Work together with Steve to review up to Section 2

@@
 We address the issue of automatically planning discrete sequences of contacts
-that characterize a multiped motion in a cluttered environment.
+that characterize a multiped motion in a constraining environment.
 To overcome the inherent combinatorial difficulty of this problem, we separate it
 in two sub-problems: first, planning a guide path for the root of the robot and
-then, generating relevant contacts along this path.
+then, generating contacts along this path.
 We propose theoretical contributions for each sub-problem, and implement them
 as a complete contact planner. 
-First, we propose a geometrical characterization of a guide path, the ``reachability condition", which guarantees that a
-guide can be mapped into the contact manifold of the robot. With this condition, asserting the feasibility
-of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
-Then, we propose a sequential algorithm to extend a feasible guide path into a sequence of contacts.
-A sampling based-approach is used to generate candidate configurations.
-The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
-designed for generic robotic applications: a manipulability-based heuristic, EFORT, and a robustness criterion regarding the equilibrium of the robot.
-These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
-variety of robots, including HRP-2, HyQ, and a dexterous hand, with interactive performances.
+First, we generate a guide root path without explicitly computing configurations in static equilibrium, as opposed to previous approaches.
+To do this we propose to approximate the space of equilibrium configurations using the ``reachability condition".
+Verifying this condition boils down to performing simple collision tests and, in the considered environments, this approximation turns out to be accurate.
+Then, we propose an algorithm to extend a root guide path into a sequence of equilibrium configurations.
+A sampling-based approach is used to generate candidate contacts.
+The best candidate is chosen according to user-defined heuristics. We propose four such heuristics, used in all our experiments, and 
+designed for generic robotic applications.
+These contributions are implemented within the open-source software Hpp.
+Our framework allows us to generate complex multi-contact sequences in at most a few seconds for a large variety of robots, including HRP-2, HyQ, and a dexterous hand.

### 2016-01-21 body/intro.tex
Work together with Steve to review up to Section 2

@@
 % !TEX root =  ../main.tex
-We consider the problem of planning an acyclic sequence of contacts describing the motion of a multiped robot in a constrained environment.
-Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to maintain the equilibrium of the system.
+We consider the problem of planning an acyclic sequence of contacts describing the motion of a multiped robot in a constraining environment.
+Acyclic contact planning is a particular class of motion planning in which the robot must be in contact with the environment at every configuration to maintain the static equilibrium.
 
 Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. However, executing
-this behavior on constrained environments is dangerous, if not impossible.
+this behavior on constraining environments is dangerous, if not impossible.
 In an analysis of their participation to the Darpa Robotics Challenge, \citeauthor{atkensondarpa}
@@
 both plan and control[...].  We have seen very few robot planners that are  capable of  generating this  behavior."
-The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of collision-free configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
+The difficulty of addressing such a problem comes both in practice from the proximity to obstacles (which makes tedious the sampling of collision-free configurations) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
 
@@
 
-As suggested in \cite{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner.
-We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot path.
-This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
-This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely rapidly.
-while preserving a high success rate in the demonstrated scenarios.
+As suggested by \cite{Bouyarmane2009}, we believe that these two problems can be treated sequentially within a global planner.
+We go further and claim that this can be done at a much smaller cost, provided we can formulate a computationally-efficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} of the root configuration (i.e.
+there exists a joint configuration such that the robot is in static equilibrium).
+This paper presents a geometrical approximation of this condition, and a concrete implementation of such a decoupled planner.
+This approximation results from a trade-off between a necessary and a sufficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}, allowing us to find solutions extremely rapidly, while preserving a high success rate in the demonstrated scenarios.
 
@@
 in their seminal paper  \citep{Bretl:2006:MPM:1124573.1124585}.
-The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: Planning a discrete sequence of acyclic contact configurations along the path.
+The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of acyclic equilibrium configurations along the path.
 A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence.
-A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential paths.
+A key issue is to avoid combinatorial explosion when considering at the same time the possible contacts and the potential paths.
 This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments.
@@
 To keep reasonable computation times, the method uses a simplified dynamic model for the avatar.
-Still, the computation time is far from interactive  (about 1 minute of computation for a sequence of 20 contacts).
+Still, the computation time is far from real-time  (about 1 minute of computation for a sequence of 20 contacts).
 A similar approach has been considered for manipulation by \cite{gabicciniisrr15}.
-\citeauthor{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed integer problem, but only cyclic, bipedal locomotion is considered.
+\citeauthor{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered.
 \citeauthor{dai2014whole} extend the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
 the contacts manually for more complex interactions.
-In addition to the practical limits of the current implementations, a major drawback of these optimization based approaches is thus that they only offer local convergence when applied to acyclic contact planning.
+In addition to the practical limits of the current implementations, a major drawback of these optimization-based approaches is that they only offer local convergence when applied to acyclic contact planning.
 
 \textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity.
-The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough guide path (i.e.
+The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
 an ad-hoc solution to $\mathcal{P}_1$).
-\Pb  is then addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
-A solution can then be found, efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, but at the cost of prohibitive computation times (several hours) for constrained scenarios for humanoid.
-In the latter case, this approach is suboptimal because the quality of the motion is conditioned by the quality of the guide path,  which is not evaluated \textit{a priori}.
-\citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide path with guarantees of contact feasibility, by extending key frames of the path into whole-body contact configurations in static equilibrium.
-Randomly sampled configurations are projected into the contact sub-manifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide path in the examples presented).
-Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the path.
-\citeauthor{7140082} also propose a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, that characterize the ability to create a contact with
-a discretized environment. This planning phase does not account for collisions, involving that replanning is required in case of failure. This approach is efficient 
-in the scenarios demonstrated. In highly constrained cases such 
-as the car egress scenario we address here, we believe that including collision constraints in the planning is a requirement.
+\Pb is addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
+A solution can then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning},
+but at the cost of prohibitive computation times (up to several hours) for constraining scenarios.
+This approach is suboptimal because the quality of the motion depends on the quality of the guide path.
+\citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide path with guarantees of \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}, by extending key frames of the path into whole-body configurations in static equilibrium.
+Randomly-sampled configurations are projected into the contact sub-manifold using a inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
+Moreover this explicit projection is insufficient to guarantee the feasibility between two key positions in the path.
+\citeauthor{7140082} also propose a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with
+a discretized environment. This planning phase does not account for collisions, implying that replanning is required in case of failure. This approach is efficient 
+in the demonstrated scenarios. In highly-constraining cases such 
+as the car egress scenario that we address here, we believe that including collision constraints in the planning is a requirement.
 
-For completeness, we lastly mention a new kind of approach, recently proposed from the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
-use black box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
-at \gls{interactive} frame rates (We say that a planning method is interactive when the computation time for one step is lesser than the
+For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
+use black-box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
+at \gls{interactive} frame rates (we say that a planning method is interactive when the computation time for one contact switch is less than the
 time to execute it). While this new approach provides an exciting direction of research, currently the resulting motions
@@
 
-Regarding robotics applications, it thus appears that optimization based approaches outperform existing planners,
-partly because the real-time interpolation between contact poses ($\mathcal{P}_3$) remains an unsolved problem.
-However recent contributions have been able to bring partial solutions to it \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
-\citeauthor{Carpentier2016} are able to achieve this with \gls{interactive} performances on the real robot.
-To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly.
-Such a planner holds the promise of a complete real-time multi-contact locomotion system.
+As far as robotics applications are concerned, none of the existing planners is able to solve the problem with \gls{interactive} performances.
+However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
+In particular, \citeauthor{Carpentier2016} are able to achieve this with \gls{interactive} performances on a real robot.
+Therefore, a planner capable of solving \Pa and \Pb efficiently could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
+The main contribution of this paper is exactly this planner.
 
@@
 This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
-we believe that the separation between the generation of the guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
+we believe that the separation between the generation of the root guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
 However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated. \\
 
-\noindent \textbf{Regarding $\mathcal{P}_1$}, the guide path must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces).
-We call this property \textit{equilibrium feasibility}.
+\noindent \textbf{Regarding $\mathcal{P}_1$}, the guide path must guarantee the existence of a contact sequence to actuate it.
+We call this property \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
 This property has not been studied yet; the only way to validate a waypoint in the path is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
@@
 
-\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinite combinations of possible contact sequences for a given root path.
-The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to constrained environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, possibly leading to very tedious transitions).
+\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinitely many combinations of possible contact sequences for a given root path.
+The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to constraining environments (\citeauthor{bouyarmane:lirmm-00777727} and \citeauthor{DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, possibly leading to very tedious transitions).
 \\
 
-\textbf{The key idea of the paper is the \textit{reachability condition}}, an efficient approximation of the space of \gls{equilibrium feasible} configurations.
-We make the strong assumption that the root configurations in this approximation can be efficiently extended into full body configurations in static equilibrium.
-We know that this assumption is not always verified, but we provide empirical evidence that it is often
-the case in the class of problems we consider. We define this class as the set of \textit{\gls{cluttered} contact planning problems}.
-We consider that such problems can be solved with a contact sequence 
-where every configuration contains at least one \gls{quasi-flat} contact support \citep{Prete2016}. A \textit{quasi-flat} contact support denotes a contact
-with a surface where the friction cone contains the direction opposite to the gravity. 
+\textbf{The key idea of the paper is the \textit{reachability condition}}, a computationally-efficient approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
+We assume that most of the times the \textit{reachability condition} implies the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
+%root configurations in this approximation can be efficiently extended into full body configurations in static equilibrium.
+This assumption is not always verified, but we provide empirical evidence that it is in the considered class of problems, the \textit{\gls{cluttered} contact planning problems}.
+Such problems can be solved with a contact sequence 
+where in every configuration at least one contact is \gls{quasi-flat}~\citep{Prete2016}. A contact is \textit{quasi-flat} if the friction cone contains the direction opposite to gravity. 
 %~ This definition includes motions on flat or uneven terrains, but also more complex ones such as stair climbing, car egress, or standing up scenarios, that might require noncoplanar contacts.
-Under this assumption we can take advantage of the decoupling approach to find a solution rapidly.
+%Under this assumption we can take advantage of the decoupling approach to find a solution rapidly.
 
 To address \Pa, we first use the \textit{reachability condition} to plan the guide path. This step is fast because the planning happens
-in a low dimensional space, and does not require explicit contact computation.
+in a low-dimensional space, and does not require explicit contact computations.
 
-Then we address \Pb by extending the path into a sequence of \gls{equilibrium feasible} contact configurations.
-This second step requires computing contact configurations verifying a condition for \textit{static equilibrium feasibility}.
+Then we address \Pb by extending the path into a sequence of whole-body configurations in static equilibrium.
+This second step requires the explicit computation of contact configurations.
 
 This sequential approach is the key to the efficiency of our method, though it
-can result in failures because our hypothesis is not always true. Similarly, in this paper we do not 
-prove that all the contact sequences produced by our planner can be interpolated ($\mathcal{P}_3$), although we have been able to do this
-for some of the sequences demonstrated \citep{Carpentier2016}.
-However, we demonstrate empirically the validity of the approach: the high success rate of our method, combined with the low computation times we obtain, allows us to plan (and re-plan upon failure) multi contact sequences at \gls{interactive} rates.
+can result in failures because our hypothesis is not always true.
+However, we demonstrate empirically the validity of the approach: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
+Similarly, in this paper we do not  prove that all the contact sequences produced by our planner can be interpolated ($\mathcal{P}_3$), although we have been able to do this for some of the demonstrated sequences~\citep{Carpentier2016}.
 
 \subsection{Paper contribution and organization}
-
-In this context, our method proposes contributions to both problems \Pa and \Pb.
+We propose contributions to both problems \Pa and \Pb~.
 \begin{itemize}
-\item First, we propose a low dimensional, efficient sampling based approach to plan guide paths ;
-\item Then we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with \gls{interactive} applications;
-\item We propose three heuristics for contact generation. They bias the planner towards configurations that are robust regarding static equilibrium, or 
-efficient regarding the task being performed.
-\item Finally, we present statistical tests that empirically demonstrate the validity of our approach in the scenarios we consider. 
+\item A low-dimensional, efficient sampling-based approach to plan guide paths.
+\item A very efficient and general implementation of an acyclic contact planner, the first one compatible with \gls{interactive} applications.
+\item Four heuristics for contact generation. They bias the planner towards configurations that are in robust static equilibrium, or 
+more efficient with respect to the task.
+\item Statistical tests that empirically demonstrate the validity of our approach in the considered scenarios. 
 \end{itemize}
 
-Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
-Since the interpolation of a complete path between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
-is the fastest multi-contact planning solution to our knowledge.
+%Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
+%Since the interpolation of a complete path between such configurations can be solved within \gls{interactive} constraints \citep{Carpentier2016}, we claim that our planner is the fastest multi-contact planning solution to our knowledge.
 
-In Section~\ref{overview}, we present the general organization of our method.
+In Section~\ref{overview}, we present an overview of our method.
 Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb.
 In Section~\ref{sec:heuristics}, we present heuristics used for the selection of a contact configuration.
-Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
-that complete those presented in the conference paper.
+Finally, in Section 6 we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios.
+%that complete those presented in the conference paper.
+
 \subsection*{Comparison with our previous work.}
@@
 The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
-This extension completes this work and brings new contributions, resulting from the implementation of our approach real-world robots and problems.
+This extension completes this work and brings new contributions, resulting from the implementation of our approach on real-world robots and problems.
 In particular, a strong contribution is the introduction of a robust equilibrium criterion, designed to ensure the equilibrium of the robot despite bounded errors in the contact forces, coupled
@@
 In the present paper, a complete algorithm for the contact planner is given, and an access to the source code of our implementation is provided.
-Furthermore, our experiments are supported by a an in-depth analysis of our performance, as well as a discussion on the influence of the parameters to the method.
+Furthermore, our experiments are supported by an in-depth analysis of our performances, as well as a discussion on the influence of the parameters to the method.

### 2016-01-21 body/overview.tex
Work together with Steve to review up to Section 2

@@
 Figure~\ref{fig:framework} illustrates our workflow.
-\Pa and \Pb are addressed in a sequential fashion: we first plan a guide path with the \textit{reachability condition}, before
-extending it into a sequence of contact configurations in static equilibrium.
+\Pa and \Pb are addressed in a sequential fashion: we first plan a root guide path with the \textit{reachability condition}, before
+extending it into a sequence of configurations in static equilibrium.
 In this Section we describe this workflow and give some notations used throughout the paper. Some key concepts are introduced,
-the definitions of which can be found in the glossary at the end of this paper.
+whose definitions can be found in the glossary at the end of this paper.
 
@@
 \subsection{Computation of a guide path --- \Pa}
-We first consider the problem of planning a relevant guide path.
-The objective is to compute a path of root configurations that will allow contact creation.
-A desired requirement is to preserve the completeness of the planner: it should be able to explore any valid guide path, but at the same time, any computed guide path must lead to a sequence of contact configurations in static equilibrium.
-The sets we introduce are illustrated in Figure~\ref{fig:dedefeas}.
-
+We first consider the problem of planning a root guide path. 
+%The objective is to compute a path of root configurations that will allow for equilibrium. 
+%A desired requirement is to preserve the completeness of the planner: it should be able to explore any valid guide path, but at the same time, any computed guide path must lead to a sequence of configurations in static equilibrium.
 \begin{figure}[t]
@@
 \end{figure}
-
-More precisely, a root configuration is \gls{equilibrium feasible} if and only if there exists at least one joint configuration such that the robot satisfies two conditions:
+%
+Ideally we would like to sample root configurations that are \gls{equilibrium feasible}.
+A root configuration is \gls{equilibrium feasible} if and only if there exists at least one joint configuration such that:
 \begin{description}
-\item[C1]: is not in collision with the environment, 
+\item[C1]: the robot is not in collision with the environment, 
 %~ \item[C2] has at least one end-effector in contact with the environment, 
-\item[C2]: is in static equilibrium. 
+\item[C2]: the robot is in static equilibrium. 
 \end{description}
-A requirement for C2 is the condition C3: at least one end-effector must be in contact with the environment.
-C1 and C3 define \glslink{contact feasible}{\textit{contact feasibility}}, and can be efficiently approximated even without computing the contact points.
-C2 defines \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}, that we describe with the set $C_{Equil}^0$ (Figure~\ref{fig:dedefeas} -- Green).
-C2 is difficult to approximate and requires the computation of the contact points, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
-
-We make the assumption that for the \textit{\gls{cluttered} contact planning problem} (that we define as admitting a solution where every configuration contains at least one contact
-is \gls{quasi-flat}) verifying C2 is not critical for $\mathcal{P}_1$.
-This means that root configurations satisfying C1 and C3 also satisfy C2. For this reason we neglect C2 during the resolution of \Pa, and we consider it only when solving \Pb.
-
-We define $C_{Contact}^0$ (Figure~\ref{fig:dedefeas} -- Orange) as the set of \gls{contact feasible} root configurations (that can satisfy C1 and C2).
-Planning in $C_{Contact}^0$ boils down to planning in $\mathbb{R}^r$, which has an acceptable practical complexity.
+Verifying C2 is computationally expensive because it requires the projection of the whole-body configuration into a contact posture, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
+For this reason, we are interested in finding an approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} that is computationally efficient.
+A requirement for C2 is the condition:
+\begin{description}
+\item[C3]: at least one end-effector is in contact with the environment.
+\end{description}
+C1 and C3 define \glslink{contact feasible}{\textit{contact feasibility}}, which is easier to verify than \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} because it does not require the computation of contact forces.
+We make the assumption that for the \textit{\gls{cluttered} contact planning problem}, most of the times \gls{contact feasible} root configurations are also \gls{equilibrium feasible}.
+For this reason we neglect C2 during the resolution of \Pa, and we consider it only when solving \Pb.
+However, even verifying \glslink{contact feasible}{\textit{contact feasibility}} is too computationally expensive because it requires the projection of the whole-body configuration into a contact posture.
 
 An intuitive description of \gls{contact feasible} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
-We approximate $C_{Contact}^0$ with a set $C_{reach}^0$, defined by the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:dedefeas} -- Red).
-Our objective is to find a root path such that the root (scaled by a factor $s \geq 1$) is not in collision, while the reachable workspace is in collision with the environment.
-In Section~\ref{rbprm} we detail how  $C_{reach}^0$ is designed and sampled to compute a guide path with a dedicated planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).
+We approximate \glslink{contact feasible}{\textit{contact feasibility}} with the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
+More precisely, a root configuration is \textit{reachable} if the root scaled by a factor $s \geq 1$ is not in collision, while the reachable workspace is in collision with the environment.
+Verifying the \textit{reachability condition} is very efficient in practice.
+In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).
+
+Figure~\ref{fig:dedefeas} gives an insight into the difference between the three conditions, depicting $C_{Equil}^0$, $C_{Contact}^0$ and $C_{Reach}^0$, which are the sets of \gls{equilibrium feasible}, \gls{contact feasible} and \textit{reachable} root configurations, respectively.
 
@@
 contact generation problem is reduced to the number of degrees of freedom of the considered limb. 
-In other words, we consider each limb as a manipulator attached to the root, and select the most relevant contact from a database of precomputed configurations, which is independent from the environment (Figure~\ref{fig:contact_gen}).
+%In other words, we consider each limb as a manipulator attached to the root. 
+We select from a database of precomputed limb configurations (independent from the environment) the ones resulting in the end-effector being close to the environment.
+Limiting our selection to this set of precomputed configurations allow us to drastically reduce the computation times. 
+Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), which we present in Section~\ref{sec:heuristics}.
 
@@
 Details are presented in Section~\ref{sec:contact}.
-
-There is a combinatorial issue in the choice of which contact to create,
-and where to create it.
-This complexity is avoided by selecting the ``most relevant'' contact configurations with user-defined heuristics.
-In Section~\ref{sec:heuristics} we present extensively those heuristics that have been used in our experiments.
 
@@
 \textbf{A robot} is a kinematic chain $R$, comprising \mbox{$n + r$} degrees of freedom (DOFs), with $r \geq 6$. $r$ denotes the world coordinates of the root of the robot,
-as well as extra DOFs not related to the limbs (For instance, HRP-2 has two extra DOFS in the torso, such that we have $r=8$. 
+as well as extra DOFs not related to the limbs (for instance, HRP-2 has two extra DOFS in the torso, such that we have $r=8$). 
 $R$ is composed of $l$ limbs $R^k, 1 \leq k \leq l$, attached to a root.
@@
 We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
-	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ denotes the position and orientation of the root of the robot $R$.
+	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ denotes the world coordinates of the root of the robot $R$.
 \end{itemize}
@@
 \medskip
-The volume encompassing the trunk of the robot is denoted $W^0$
+The volume encompassing the trunk of the robot is denoted $W^0$.
  %~ (Figure~\ref{fig:character}-right: central cylinder). 
- The reachable workspace of a limb $R^k$ is denoted $W^k$ %(Figure~\ref{fig:character}-right: the four ellipses). 
+ The reachable workspace of a limb $R^k$ is denoted $W^k$: %(Figure~\ref{fig:character}-right: the four ellipses). 
 \begin{equation}

### 2016-01-21 body/conclusion.tex
added missing figure on feasibility
small fixes in revised sections

@@
 In this paper we consider acyclic contact planning in cluttered environments, formulated as two sub problems that we address sequentially.
-The first problem is \Pa, that consists in computing a guide path for the root of the robot that can be extended; the second problem is \Pb, that is the computation of a discrete sequence of contacts along the root path.
+The first problem is \Pa, that consists in computing a guide path for the root of the robot that can be extended; the second problem is $\mathcal{P}_2$, that is the computation of a discrete sequence of contacts along the root path.
 Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner.
@@
 %
-\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and \Pb. This results in computation costs close to real-time in
+\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and $\mathcal{P}_2$. This results in computation costs close to real-time in
 known static environments.
@@
 working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
-Lastly, we aim at going beyond static equilibrium and to perform kynodynamic planning in the configuration space. We believe that the most promising direction in this regard is to integrate
+Lastly, we aim at going beyond static equilibrium and to perform kinodynamic planning in the configuration space. We believe that the most promising direction in this regard is to integrate
 the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.

### 2016-01-21 body/intro.tex
small fixes in revised sections

@@
 
-To address \Pa, we first use the \textit{reachability condition} to plan the guide path. This step is fast because the planning happens
+To address $\mathcal{P}_1$, we first use the \textit{reachability condition} to plan the guide path. This step is fast because the planning happens
 in a low-dimensional space, and does not require explicit contact computations.
@@
 \subsection{Paper contribution and organization}
-We propose contributions to both problems \Pa and \Pb~.
+We propose contributions to both problems \Pa and $\mathcal{P}_2$.
 \begin{itemize}
@@
 In Section~\ref{overview}, we present an overview of our method.
-Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb.
+Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and $\mathcal{P}_2$.
 In Section~\ref{sec:heuristics}, we present heuristics used for the selection of a contact configuration.

### 2016-01-21 body/overview.tex
small fixes in revised sections

@@
 is only described by two positional parameters (x and y).
+The root of the robot is indicated with a black cross.
+To compute the reachable workspace, the point on the ankle indicated by a green cross was used.
 $C_{Equil}^0$ is included in $C_{Contact}^0$.
@@
 \textbf{A robot} is a kinematic chain $R$, comprising \mbox{$n + r$} degrees of freedom (DOFs), with $r \geq 6$. $r$ denotes the world coordinates of the root of the robot,
-as well as extra DOFs not related to the limbs (for instance, HRP-2 has two extra DOFS in the torso, such that we have $r=8$). 
-$R$ is composed of $l$ limbs $R^k, 1 \leq k \leq l$, attached to a root.
+as well as extra DOFs not related to the limbs (for instance, HRP-2 has two extra DOFS in the torso, such that we have $r=8$). Abusively, for simplicity,
+we refer to the root as the kinematic chain described by all these DOFS.
+$R$ is composed of $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
 It is described by a configuration $\mathbf{q} \in  \mathbb{R}^r \times \mathbb{R}^n$.
@@
 \medskip
-The volume encompassing the trunk of the robot is denoted $W^0$.
+The volume encompassing the root of the robot is denoted $W^0$.
  %~ (Figure~\ref{fig:character}-right: central cylinder). 
@@
 %~ \adnote{in this definition you don't account for joint limits}
-where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0$ being the null displacement, and  $C^k_{jl}$ is the space
+where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{jl}$ is the space
 of admissible limb joint configurations. We also define $W = \bigcup_{k=1}^{l}W^k$, and
-$W^k(\mathbf{q}^{0})$ (for $1 \leq k \leq l$) as the volume $W^k$ translated and rotated by the rigid displacement~$\mathbf{q}^{0}$.
+$W^k(\mathbf{q}^{0})$ (for $1 \leq k \leq l$) as the volume $W^k$ translated and rotated according to the root configuration $\mathbf{q}^{0}$.
 

### 2016-01-21 body/results.tex
small fixes in revised sections

@@
 For this reason, we now prefer to use our manipulability based heuristic for the legs of the robot, but we still
-use EFORT for the arm, that results in lesser contact repositionning.
+use EFORT for the arm, that results in less contact repositionning.
 

### 2016-01-22 body/overview.tex
3.1

@@
 
+In the remainder of this paper, we use the terms \gls{contact feasible} and \gls{equilibrium feasible} to qualify either a root configuration, a whole-body configuration, or a set
+of such configurations.
+
 \subsection{Generating a discrete sequence of contact configurations --- \Pb}
@@
 \medskip
-\textbf{A robot} is a kinematic chain $R$, comprising \mbox{$n + r$} degrees of freedom (DOFs), with $r \geq 6$. $r$ denotes the world coordinates of the root of the robot,
-as well as extra DOFs not related to the limbs (for instance, HRP-2 has two extra DOFS in the torso, such that we have $r=8$). Abusively, for simplicity,
-we refer to the root as the kinematic chain described by all these DOFS.
-$R$ is composed of $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
-It is described by a configuration $\mathbf{q} \in  \mathbb{R}^r \times \mathbb{R}^n$.
+\textbf{A robot} is a kinematic chain $R$, that we divide into different parts: a root $R^0$, and $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
+The root has $r \geq 6$ DOFS: for instance, HRP-2 has two extra DOFS in the torso, such that we have $r=8$.
+Therefore $R$ is fully described by a configuration $\mathbf{q} \in  \mathbb{R}^{r+n}$.
 We define some relevant projections of $\mathbf{q}$:
@@
 We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
-	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ denotes the world coordinates of the root of the robot $R$.
+	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ denotes the world coordinates of the root $R^0$.
 \end{itemize}
@@
 \medskip
-The volume encompassing the root of the robot is denoted $W^0$.
+The volume encompassing $R^0$ is denoted $W^0$.
  %~ (Figure~\ref{fig:character}-right: central cylinder). 
@@
 \begin{equation}
-  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k \in C^k_{jl}, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\},
+  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k \in C^k_{jl}, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\}
 \end{equation}
@@
 where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{jl}$ is the space
-of admissible limb joint configurations. We also define $W = \bigcup_{k=1}^{l}W^k$, and
-$W^k(\mathbf{q}^{0})$ (for $1 \leq k \leq l$) as the volume $W^k$ translated and rotated according to the root configuration $\mathbf{q}^{0}$.
+of admissible limb joint configurations. We also define \mbox{$W = \bigcup_{k=1}^{l}W^k$}, and
+$W^k(\mathbf{q}^{0})$ (for $1 \leq k \leq l$) as the volume $W^k$ for the root configuration $\mathbf{q}^{0}$.
 

### 2016-01-22 body/rbprm.tex
typos
3.1

@@
 % !TEX root =  ../main.tex
-\section{Computing a guide path in $C_{reach}$ ($\mathcal{P}_1$) }
+\section{Computing a guide path in $C^0_{reach}$ ($\mathcal{P}_1$) }
 \label{rbprm}
 
-We consider the issue of computing a normalized guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow \mathbb{R}^r$ for the root of a multiped robot, connecting user-defined start and goal configurations.
-As said in the previous section, the goal is to enforce that the guide can be discretized into a sequence of \gls{contact feasible} configurations.
-We denote by \gls{ccontact} the contact sub-manifold of the robot.
+We consider the issue of computing an \gls{equilibrium feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{ccequil0} for the root of a multiped robot, connecting user-defined start and goal configurations.
 
-We say that a root configuration $\mathbf{q}^{0}$ is \gls{contact feasible} if it belongs to the set:
-\begin{equation}
-      C_{contact}^0= \{ \mathbf{q}^{0}: \exists \mathbf{q}^{\overline{0}}, \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact} \}
-    \label{eq:pi}
-\end{equation}
+Again, we assume that, for \gls{cluttered} problems, most of the times \contactfeasibility\, implies \equilibriumfeasibility.
+Under this assumption, our goal is to find a \gls{contact feasible} guide path.
+
+To generate such a path efficiently, ideally we need to only sample \gls{contact feasible} root configurations.
+This requires exhibiting a necessary and sufficient condition for \contactfeasibility\,that does not need to compute the whole-body configuration.
+Since we do not know such a condition, we may turn our attention towards either an only-necessary or an only-sufficient condition.
+
+Only-necessary conditions are appealing because they preserve the completeness of the search, while reducing 
+the search space: they provide an outer-approximation of \gls{ccontact0}.
+On the other hand, only-sufficient conditions provide the guarantee that any configuration that satisfies them is indeed \gls{contact feasible}:
+they provide an inner-approximation of \gls{ccontact0}.
+
+In practice, the only-necessary and only-sufficient conditions that we can provide are trivial and provide rather inaccurate approximations of \gls{ccontact0}.
+Therefore, we propose a compromise between them: the \textit{reachability condition}, which is computationally efficient
+and it provides a rather accurate approximation of \gls{ccontact0}. % for \gls{cluttered} problems.
+
+%~ As said in the previous section, the goal is to enforce that the guide can be discretized into a sequence of \gls{equilibrium feasible} configurations.
+We denote by \gls{ccequil} the equilibrium sub-manifold of the robot.
+
+%~ We say that a root configuration $\mathbf{q}^{0}$ is \gls{equilibrium feasible} if it belongs to the set:
+%~ \begin{equation}
+      %~ C_{Equil}^0= \{ \mathbf{q}^{0}: \exists \mathbf{q}^{\overline{0}}, \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{Equil} \}
+    %~ \label{eq:pi}
+%~ \end{equation}
 
 %~ The set of all \textit{geometrically feasible} root configurations is denoted by $C_{reach}$.
-By extension, a path $\mathbf{q}^0(t)$ is \gls{contact feasible} if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{contact}^0$.
+%~ By extension, a path $\mathbf{q}^0(t)$ is \gls{equilibrium feasible} if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{contact}^0$.
 
-For this stage, and for this stage only, we make the assumption that for \gls{cluttered} problems, a \gls{contact feasible} path can be transformed into a \gls{equilibrium feasible} path.
-Under this assumption, for a two-stage acyclic contact planner to be exact and complete, we need the combination of two conditions on a guide path generator: all the generated paths must be \gls{contact feasible} (sufficient condition); the guide planner must be complete, i.e.
-it must be able to discover any \textit{contact feasible} path (necessary condition).
-
-In this Section we characterize \gls{ccontact0} theoretically with a necessary and a sufficient condition for belonging to the set.
-We then approximate \gls{ccontact0} as a space $C_{reach}$, where we can efficiently sample configurations.
-We finally describe our implementation of an efficient RRT planner in $C_{reach}$.
+%~ In this Section we characterize \gls{ccontact0} theoretically with a necessary and a sufficient condition for belonging to the set.
+%~ We then approximate \gls{ccontact0} as a space $C_{reach}$, where we can efficiently sample configurations.
+%~ We finally describe our implementation of an efficient RRT planner in $C_{reach}$.
 
@@
 C_{Nec}^0 = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
- %~ & \wedge & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}
+ %~ & \text{ and } & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}
 \end{equation}
@@
  \begin{equation}
-C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \text{ and } & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}

### 2016-01-25 body/appendix_roms.tex
part 3 with andrea

@@
 
-\subsection{Step 2: computing the trunk scaling workspace $W^0$ of the robot}
+\subsection{Step 2: computing the torso scaling workspace $W^0$ of the robot}
 To define the volume $W^0$ of HRP-2, we proceed in an empirical manner.
-First, we compute the bounding boxes of the robot trunk, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
+First, we compute the bounding boxes of the robot torso, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
 Then, we perform a scaling of these boxes by a factor $s$. 

### 2016-01-25 body/rbprm.tex
part 3 with andrea

@@
 To generate such a path efficiently, ideally we need to only sample \gls{contact feasible} root configurations.
-This requires exhibiting a necessary and sufficient condition for \contactfeasibility\,that does not need to compute the whole-body configuration.
-Since we do not know such a condition, we may turn our attention towards either an only-necessary or an only-sufficient condition.
+This requires exhibiting a necessary and sufficient condition for \contactfeasibility. %,that does not need the computation of the whole-body configuration.
+By default, verifying \contactfeasibility\, implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$. This is the approach chosen by \cite{Bouyarmane2009},
+which is too computationally expensive.
+To formulate a cheaper condition, we may turn our attention towards either an only-necessary or an only-sufficient condition.
 
 Only-necessary conditions are appealing because they preserve the completeness of the search, while reducing 
-the search space: they provide an outer-approximation of \gls{ccontact0}.
+the search space: they provide an outer approximation of \gls{ccontact0}.
 On the other hand, only-sufficient conditions provide the guarantee that any configuration that satisfies them is indeed \gls{contact feasible}:
-they provide an inner-approximation of \gls{ccontact0}.
+they provide an inner approximation of \gls{ccontact0}.
 
-In practice, the only-necessary and only-sufficient conditions that we can provide are trivial and provide rather inaccurate approximations of \gls{ccontact0}.
+In practice, the only-necessary and only-sufficient conditions that we can provide are trivial and give rather inaccurate approximations of \gls{ccontact0}.
 Therefore, we propose a compromise between them: the \textit{reachability condition}, which is computationally efficient
-and it provides a rather accurate approximation of \gls{ccontact0}. % for \gls{cluttered} problems.
+and provides a rather accurate approximation of \gls{ccontact0}. % for \gls{cluttered} problems.
 
@@
 \subsection{Conditions for contact feasibility}
-By default, \glslink{contact feasible}{\textit{contact feasibility}} implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$.
+%~ By default, \glslink{contact feasible}{\textit{contact feasibility}} implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$.
 This is the approach chosen by \cite{Bouyarmane2009}.
@@
 For a contact to be possible, a volume $O_i \in O$ necessarily intersects with the reachable workspace $W(\mathbf{q}^{0})$ (Figure~\ref{fig:contact_gen}--1).
-Furthermore, if $\mathbf{q}^{0}$ is  feasible, then the trunk of the robot $W^0(\mathbf{q}^{0})$ is necessarily not colliding  with the environment $O$.
+Furthermore, if $\mathbf{q}^{0}$ is \gls{contact feasible}, then the torso of the robot $W^0(\mathbf{q}^{0})$ is necessarily not colliding  with the environment $O$.
 
-Therefore we can define an optimistic approximation  $C^0_{Nec} \supset$ \gls{ccontact0} defined as: 
+Therefore we can define an outer approximation  $C^0_{Nec} \supset$ \gls{ccontact0} defined as: 
 \begin{equation}
 C_{Nec}^0 = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
- %~ & \text{ and } & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}
+ %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
 \end{equation}
-The inclusion \gls{ccontact0} $\subset C^0_{Nec}$ is straightforward, but it is very important.
-Planning in $C^0_{Nec}$ allows a strong reduction of the search space, while not discarding 
-any root configuration that could lead to a solution.
+%~ The inclusion \gls{ccontact0} $\subset C^0_{Nec}$ is straightforward, but it is very important.
+%~ Planning in $C^0_{Nec}$ allows a strong reduction of the search space, while not discarding 
+%~ any root configuration that could lead to a solution.
 
-The condition defining $C^0_{Nec}$ is only necessary. This means that it might not be possible to extend a guide path planned 
-in $C^0_{Nec}$ into a sequence of contact configurations.
+%~ The condition defining $C^0_{Nec}$ is only necessary. This means that it might not be possible to extend a guide path planned 
+%~ in $C^0_{Nec}$ into a sequence of contact configurations.
 
@@
 A trivial sufficient condition for \glslink{contact feasible}{\textit{contact feasibility}} can be constructed as a variation of $C^0_{Nec}$, by replacing $W^0$ with a bounding volume $B^{Suf}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
-We denote by \mbox{$C_{Suf}^0 \supset $ \gls{ccontact0}} the set of root configurations corresponding to this sufficient condition.
+We denote by \mbox{$C_{Suf}^0 \subset $ \gls{ccontact0}} the set of root configurations corresponding to this sufficient condition:
 
-In general, the inclusion is strict, which means that we lose the completeness of the two-stage contact planner (i.e.
+\begin{equation}
+C_{Suf}^0 = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } B^{Suf}(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+ %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
+\end{equation}
+
+%~ In general, the inclusion is strict, which means that we lose the completeness of the two-stage contact planner (i.e.
 the planner is not able to discover a path inside \mbox{$C_{reach} \setminus C_{Suf}^0$}).
@@
 The sufficient condition is not interesting in practice since it leads the solver to lose too many interesting paths.
-The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \gls{contact feasible} in practice.
-It might be possible to find an ideal shape $B$ that is necessary and sufficient; however, it seems intuitively very unlikely in general.
+The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \gls{contact feasible}.
+An ideal shape $B$ (with $W^0 \subset B \subset B^{Suf}$) that leads to a necessary and sufficient condition may exist---even if it seems intuitively very unlikely in general. 
+%~ Its construction is out of the scope of this work.
+
+%~ It might be possible to find an ideal shape $B$ that leads to a necessary and sufficient condition; however, it seems intuitively very unlikely in general.
 The construction of a shape $W^0 \subset B \subset B^{Suf}$ leading to a necessary and sufficient condition (or the proof of its nonexistence) is out of the scope of this work.
 
-However, between $W^0$ and $B^{Suf}$, a trade-off can be found between a necessary and a sufficient condition.
+However, using a shape between $W^0$ and $B^{Suf}$ leads to a trade-off between a necessary and a sufficient condition.
 We define $W^0_s$ as the volume $W^0$ subject to a scaling transformation by a factor $s \in \mathbb{R}^+$.
@@
 C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
- %~ & \text{ and } & A_{trunk}(\mathbf{q}^{root}) \cap W = \emptyset \}
+ %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
 \end{equation}
 %
-If $s=1$, then $W^0_s$ = $W^0$, such that $C_1^0$ = $C_{Nec}$. We thus define $s \in \mathbb{R}^+$, since values inferior to $1$ can only introduce more errors.
-The higher $s$, the closer this \gls{reachc} is to being sufficient, and if $s=1$, the search is complete (in the sense that all \gls{contact feasible} configurations can be found).
-The parametrization of $s$ defines a trade-off between these two desirable properties.
-$s$ can be chosen by hand, or automatically as explained in Section~\ref{sec:params}.
-In the rest of the paper, we consider that an appropriate value $s^*$ of $s$ is chosen, and define \gls{creach0} $= C^0_{s^*}$.
+If $s=1$, then $W^0_s$ = $W^0$, such that $C_1^0$ = $C_{Nec}$. We thus consider that $s \geq 1$, since smaller values would only worsen the approximation.
+By increasing $s$, the condition can become sufficient.
+%~ The higher $s$, the closer this \gls{reachc} is to being sufficient, and if $s=1$, the search is complete (in the sense that all \gls{contact feasible} configurations can be found). 
+The parametrization of $s$ then defines a trade-off between these two interesting extremes. 
+We can choose $s$ by hand, or automatically as explained in Section~\ref{sec:params}.
+The chosen value $s=s^*$ defines the \gls{reachc}, therefore we write \gls{creach0} $= C^0_{s^*}$.
+%~ We call this condition the \gls{reachc}, which is defined by the chosen value $s=s^*$, and define \gls{creach0} $= C^0_{s^*}$.
+%~ In the rest of the paper, we consider that an appropriate value $s^*$ of $s$ is chosen, and define \gls{creach0} $= C^0_{s^*}$.
 
-As an example, Figure~\ref{fig:HyQ_roms} presents the volumes $W$ computed for the HyQ robot.
-In Appendix~\ref{app:rom}, we give a generic method to compute these volumes, with the example of HRP-2.
+%~ As an example, Figure~\ref{fig:HyQ_roms} presents the volumes $W$ computed for the HyQ robot.
+In Appendix~\ref{app:rom}, we give a generic method to compute the volumes appearing in the definition of \gls{creach0}, with the example of HRP-2.
 %~ Section~\ref{sec:results} shows that in practice, it is easy to adjust $s$ to keep most of the interesting guides without introducing incorrect guides.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
-  \caption{
-           $W$ volumes for HyQ. The green shapes represent the reachable workspace $W^k$ of each limb. The red shape is $W^0$, a scaling of the convex hull
-           of the torso of the robot.}
-		   \label{fig:HyQ_roms}
-\end{figure}
 
@@
 Any sampling-based motion planner can be used to plan a path in \gls{creach0}. 
-Indeed, contrary to \gls{ccontact}, \gls{creach0} has a non zero measure in the configuration space $C$. Therefore a standard uniform sampling approach
+Indeed, contrary to \gls{ccontact}, \gls{creach0} has a nonzero measure in the configuration space $C$. Therefore a standard uniform sampling approach
 can work, in spite of a high rejection rate. 
-Thus, the only significant change regarding a classical planner is to replace the classical collision checking method with a test of belonging to \gls{creach0} when verifying
-that drawn configurations and associated local paths are valid.
+Thus, the only significant change regarding a classical planner is to replace the collision checking with the \gls{reachc} when verifying
+the drawn configurations and associated local paths.
 
-However, to improve the sampling efficiency while remaining probabilistically complete, we bias the sampling process to generate near obstacle configurations, similarly to~\cite{Amato98choosinggood}.
-First, a configuration is set to a random point on the surface of one randomly chosen obstacle. The root location is then translated and rotated randomly until the \gls{reachc} is satisfied.
+However, to improve the sampling efficiency 
+%~ while remaining probabilistically complete, 
+we bias the sampling process to generate near-obstacle configurations, similarly to~\cite{Amato98choosinggood}.
+%~ First, a configuration is set to a random point on the surface of one randomly chosen obstacle. The root location is then translated and rotated randomly until the \gls{reachc} is satisfied.
 Our current implementation of these modifications is based on the Bi-RRT planner \citep{770022} provided by the HPP software.
 
-Thanks to this set of modifications, the problem of planning an \gls{contact feasible} path in a high dimensional space is reduced to a standard geometry and collision checking problem, in a much lower dimension (6 for HyQ, 8 for HRP-2 that has joints in the torso).
-In so doing we make the problem tractable for sampling-based approaches, with \gls{interactive} computation times.
+Thanks to these modifications, the problem of planning a \gls{contact feasible} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
+By doing this we can solve the problem with a sampling-based approach, in \gls{interactive} computation times.

### 2016-01-26 body/conclusion.tex
finished pass with andrea

@@
 
-In this paper we consider acyclic contact planning in cluttered environments, formulated as two sub problems that we address sequentially.
-The first problem is \Pa, that consists in computing a guide path for the root of the robot that can be extended; the second problem is $\mathcal{P}_2$, that is the computation of a discrete sequence of contacts along the root path.
-Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner.
-We introduced a low-dimensional space $C_{reach}$ that can be mapped
-into the contact sub-manifold of the robot, approximated, and efficiently sampled by our Reachability-Based planner.
-Our contribution to \Pb is a pragmatic contact generation scheme that can take into
-account criteria to enforce interesting properties on the generated contacts (such as robustness, energy efficiency or naturalness).
+In this paper we consider the \gls{cluttered} contact planning problem, formulated as two sub-problems that we address sequentially.
+The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
+ the second problem $\mathcal{P}_2$ is the computation of a discrete sequence of whole-body configurations along the root path.
 
-Aside from the theoretical contributions, our results demonstrate that our method allows a very interesting compromise between three 
+Our contribution to \Pa is the introduction of a low-dimensional space $C_{reach}$, an approximation of the space of \gls{equilibrium feasible} root configurations.
+Thanks to the computationally efficient verification of the \gls{reachc}, we are able to solve \Pa much faster than previous approaches.
+
+%~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
+Our contribution to \Pb is a fast contact generation scheme that can take into
+account criteria to optimize user-defined properties.
+
+Our results demonstrate that our method allows a pragmatic compromise between three 
 criteria that are hard to conciliate: generality, performance, and quality of the solution, making it the first acyclic contact
@@
 \textbf{Regarding generality}, the \gls{reachc}, coupled with an approach based on limb decomposition, 
-allows the method to address arbitrary multiped robots. The only pre-requisite is the specification 
+allows the method to address arbitrary multiped robots in \gls{cluttered} problems. The only pre-requisite is the specification 
 of the volumes $W^0$.
 %
-\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and $\mathcal{P}_2$. This results in computation costs close to real-time in
-known static environments.
+\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
 %
-\textbf{Regarding the quality of the paths}, a parametrization of the \gls{reachc} allows us to compute
-relevant paths in all the presented scenarios, with low rejection rates.
-As for \cite{Bouyarmane2009}, failures can still occur, due to the compromise criterion used in computing the guide path.
+\textbf{Regarding the quality of the paths}, the \gls{reachc} allows us to compute
+\gls{equilibrium feasible} paths in all the presented scenarios, with low rejection rates.
+As for \cite{Bouyarmane2009}, failures can still occur, due to the approximate condition used to compute the guide path.
 The low computational burden of our framework however allows for fast re-planning in case of failure.
 
-One direction for future work is to focus on a more accurate formulation of $C_{reach}$ to address 
+One direction for future work is to focus on a more accurate formulation of $C_{reach}^0$ to address 
 this limitation.
 Our main objective is of course to generate the complete interpolation between the contact sequences.
-We have already obtained some successes for some of the computed sequences~\citep{Carpentier2016}, but we admit that additional
+We have already obtained some success for some of the computed sequences~\citep{Carpentier2016}, but additional
 work is required on the planning side to obtain a seamless workflow. To achieve this, we are currently working on the notion of transition certificate, i.e. formulating
 conditions that guarantee that the interpolation between two contact configurations is dynamically feasible.
-Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
-at the center of mass of the robot.
- We are also
-working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
-Lastly, we aim at going beyond static equilibrium and to perform kinodynamic planning in the configuration space. We believe that the most promising direction in this regard is to integrate
+%~ Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
+%~ at the center of mass of the robot.
+ %~ We are also
+%~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
+Lastly, we aim at going beyond static equilibrium and performing kinodynamic planning. We believe that the most promising direction in this regard is to integrate
 the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
-Addressing these three issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.
+Addressing these two issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.

### 2016-01-26 body/contact.tex
complete, need abstract and section 4
added missing figure on feasibility
finished pass with andrea

@@
 \label{sec:contact}
-Our planner computes guide paths in \gls{creach0}, an over-approximation of the set of \gls{equilibrium feasible} paths, \gls{ccequil0}.
-As an input of this stage, we however assume an \gls{equilibrium feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow \mathbb{R}^r$ for the root of the robot $R$.
-If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section TODO.
-We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.
+Our planner computes guide paths in \gls{creach0}, an approximation of \gls{ccequil0}.
+As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{ccequil0}.
+If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section~\ref{sec:perf}.
+We now consider the second problem of computing a discrete sequence of equilibrium configurations $\mathbf{Q}^{\overline{0}}$ along $\mathbf{q}^0(t)$.
 
-In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a given limb 
-and root location.
+In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a limb, given a
+root location.
 Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.
@@
 \label{sec:single_contact}
-We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$.
-From \glslink{contact feasible}{\textit{contact feasibility}} we know that there exists at least one $\mathbf{q}^{\overline{0}}$ that verifies $\mathbf{q}^k \oplus \mathbf{q}^{\overline{k}} \in$ \gls{ccontact}.
-The issue is then to exhibit locally this $\mathbf{q}^{\overline{0}}$.
-
-While exhibiting analytically a $\mathbf{q}^{\overline{0}}$ does not seem tractable, we can iteratively try to generate one as follows:
+Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
+$R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
+While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
 \begin{enumerate}
@@
 
-We partially trade the completeness property of this approach for a more efficient solution introduced in our previous work~\citep{Tonneau2014}, which we describe
-here for completeness.
+We trade the completeness of this approach for a more efficient solution introduced in our previous work~\citep{Tonneau2014}, which we describe
+here to be exhaustive.
 
-We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimal distance 
+We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimum distance 
 between an effector and a surface of the environment is less than a small value $\epsilon$.
@@
 \item Select the first configuration of $S$. Project the configuration onto contact with inverse kinematics. If $S$ is empty, stop (failure case) (Figure~\ref{fig:contact_gen} (d) and (e));
-\item If a valid solution is found, stop (success case). Otherwise remove the element from $S$ and go back step 5.
+\item If a valid solution is found, stop (success case). Otherwise remove the element from $S$ and go back to step 5.
 \end{enumerate}
@@
 
-It should be noted that a failure case for our contact generator does not necessarily mean that the planner will globally fail. Indeed,
-for a given root configuration, if contact configurations are found for other limbs, resulting in static equilibrium, the configuration will be accepted despite
-the failure to generate a contact.
-
-We thus have defined a real-time contact generator. For a given configuration, the generator can be as complete as required by the context (by increasing the number of samples).
+%~ It should be noted that a failure case for our contact generator does not necessarily mean that the planner will globally fail. Indeed,
+%~ for a given root configuration, if contact configurations are found for other limbs, resulting in static equilibrium, the configuration will be accepted despite
+%~ the failure to generate a contact.
 
@@
 
-As an input, we consider the guide path $\mathbf{q}^0(t)$, automatically discretized into a sequence of $j$ key configurations:  
+As an input, we consider the guide path $\mathbf{q}^0(t)$, discretized into a sequence of $j$ key configurations:  
 \begin{equation*}
@@
 \end{equation*} 
-where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations.
-To ensure continuity in the contact transition phases, we propose a recursive mapping $\pi$, for any $0<i<j$:
-\begin{equation*}
-    \pi\colon\left\{
-    \begin{aligned}		
-        \mathbf{Q}^0 \in C_{equil}^0 & \longrightarrow C_{equil} \\
-        %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
-        \mathbf{q}^{0}_i &  \longrightarrow  \mathbf{q}^{0}_i \oplus g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
-    \end{aligned}
-    \right.
-\end{equation*} 
-$g$ is the method that extends a root configuration into a full-body configuration. At each step, it tries to generate a contact configuration that preserves as much as possible
-the previous contacts while allowing for static equilibrium. The objective is thus to characterize $g$.
-We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
+where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations. %To ensure continuity in the contact transition phases,
+We want to extend the configurations of $\mathbf{Q}^0$ in such a way that continuity is preserved regarding the contact transitions.
+To do so, we define an algorithm that, given the current root configuration, and the previous full body configuration, computes a full body configuration 
+in \gls{ccequil} such that contacts are maintained if possible.
+The first full body configuration of the sequence is given by the initial state of the robot.
+
+ %~ we propose a recursive mapping $\pi$, for any $0<i<j$:
+%~ \begin{equation*}
+    %~ \pi\colon\left\{
+    %~ \begin{aligned}		
+        %~ \mathbf{Q}^0 \in C_{Equil}^0 & \longrightarrow C_{Equil} \\
+        %~ %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
+        %~ \mathbf{q}^{0}_i &  \longrightarrow  g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
+    %~ \end{aligned}
+    %~ \right.
+%~ \end{equation*} 
+%~ $g$ is the method that extends a root configuration into a full-body configuration. At each step, it tries to generate a contact configuration that preserves as much as possible
+%~ the previous contacts while allowing for static equilibrium. The objective is thus to characterize $g$.
+%~ We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
 
@@
 Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
-If possible, a limb in contact at time $i-1$ remains in contact at time $i$.
+If possible, a limb in contact at step $i-1$ remains in contact at step $i$.
 The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits.
-The solver is directly provided by the HPP software.
+The solver is directly provided by the Hpp software.
 
 If the solver fails, the contact is broken and a collision-free configuration is assigned to the limb.
-If two or more contacts are broken in a single time step, intermediate steps are added to reposition the contacts one at a time, before moving.
+If two or more contacts are broken in a single step, one or more intermediate steps are added.
+For these steps the root configuration is the same as for the previous step, with the difference that
+one faulty contact is repositioned, in the hope that it will not be broken at the next step.
 
@@
 %\subsection{Generation of a contact configuration}  
-\subsection{Creating contacts.}
+\subsubsection{Creating contacts.}
 Contacts are created using three rules:
 \begin{enumerate}
-\item Only one contact creation can happen between two consecutive states. This increases the odds that the interpolation between the two states be feasible;
+\item Only one contact creation can happen between two consecutive steps. This increases the odds that the interpolation between the two steps be feasible;
 %\adnote{Before you said that you ensured that transition between two states is feasible...}
-\item A contact is validated if and only if the resulting configuration is in static equilibrium, no matter what user-defined heuristics are used; 
+\item A contact is validated if and only if the resulting configuration is in static equilibrium; 
 \item We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
-was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others have been treated.
+was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.
 \end{enumerate}
 
-These three rules provide a deterministic manner to generate contacts along the discretized guide path.
-For a given \gls{equilibrium feasible} path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
+These three rules provide a deterministic generation of contacts along the discretized guide path.
+For a given \gls{equilibrium feasible} path, our current implementation does not guarantee that the planner will succeed, because
 of the deterministic approach used to break the combinatorial.
-However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}, thanks to 
-the heuristics that we use for contact selection.
+However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.

### 2016-01-26 body/efort.tex
finished pass with andrea

@@
 % !TEX root =  ../main.tex
-\subsection{Manipulability-based heuristics for locally-optimal contact selection}
-The other kind of heuristics that we use are based on the analysis of the motion capabilities
-provided by a candidate configuration regarding the task being achieved.
-These heuristics allow to locally optimize a contact location with respect to a particular task.
-For instance, one can be interested in configurations that allow to efficiently exert a force compatible with the global direction of motion,
+\subsection{Manipulability-based heuristics for contact selection}
+This Section proposes heuristics to select a contact that optimizes desiered capabilities.
+%~ These heuristics allow to locally optimize a contact location with respect to a particular task.
+For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion,
 a high velocity in a given direction, or to stay away from singular configurations.
@@
 For a limb configuration $\mathbf{q}^k$, the Jacobian matrix
- $\mathbf{J}(\mathbf{q}^k)$   defines the relation:
+ $\mathbf{J}^k(\mathbf{q}^k)$   defines the relation:
 
 \begin{equation} \label{jac}
-\mathbf{\dot{p}^k} = \mathbf{J}(\mathbf{q}^k) \dot{\mathbf{q}}^k
+\mathbf{\dot{p}}^k = \mathbf{J}^k(\mathbf{q}^k) \dot{\mathbf{q}}^k
 \end{equation}
 
-For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}(\mathbf{q}^k)$ as $\mathbf{J}$.
+For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
 
-As a linear approximation of the forward-kinematics function $f$ , $\mathbf{J}$ describes how small
+As a linear approximation of a forward-kinematics function, $\mathbf{J}$ describes how small
 variations from the configuration $\mathbf{q}$ affect the position vector $\mathbf{p}$.
-It can be seen as the transformation mapping a velocity in the configuration space into a velocity in
-the Euclidian space.
+%~ It can be seen as the transformation mapping a velocity in the configuration space into a velocity in
+%~ the Euclidian space.
 
@@
 
-We assume that $\mathbf{J}$ is full rank (we are not interested in singular configurations that we discard).
+We assume that $\mathbf{J}$ is full rank (we are not interested in singular configurations, which we discard).
 From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
@@
 \end{equation}
-This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of velocities that can
+This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of end-effector velocities that can
 be reached under the constraint \eqref{ball} for the current configuration.
-The longer an axis of the ellipsoid is, the faster the configuration can move along the direction of the axis.
+The longer the axis of the ellipsoid is, the faster the end-effector can move along the direction of the axis.
 Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.
 
-\begin{figure}[htb]
+\begin{figure}[!tbp]
   \centering
@@
 \begin{equation}
-h_{EFORT}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \mu_0 \mathbf{n}^T \mathbf{m})
+h_{EFORT}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \mathbf{m})
 \end{equation}
-where $\mu_0$ and $\mathbf{n}$ are respectively the friction coefficient and the normal vector of the contact surface.
+where $\mu$ and $\mathbf{n}$ are respectively the friction coefficient and the normal vector of the contact surface.
 
@@
 \begin{equation}
-h_{vel}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{m}]^{-\frac{1}{2}} ( \mu_0 \mathbf{n}^T \mathbf{m})
+h_{vel}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{m}]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \mathbf{m})
 \end{equation}
@@
 In other less demanding cases, manipulability can also be considered to avoid singularities.
-To do so, we can consider the manipulability measure $w$, also given by \citeauthor{Yoshikawa1984}:
+To do so, we can consider the manipulability measure $h_{w}$, also given by \citeauthor{Yoshikawa1984}:
 
-\begin{equation*} \label{ellipsoid}
-w(\mathbf{q}) = \sqrt{det(\mathbf{J}\mathbf{J}^T)}
-\end{equation*}
-$w$ measures the ``distance'' between a given configuration and a singular configuration. When $w$ is equal to 0, the configuration is in a singular state;
-the greater $w$ is, the further away the configuration is from singularity.
-We then derive:
-\begin{equation}
-h_{w}(\mathbf{q}, \mathbf{m}) = w(\mathbf{q})  (\mu_0 \mathbf{n}^T \mathbf{m})
+\begin{equation} \label{ellipsoid}
+h_{w}(\mathbf{q}) = \sqrt{det(\mathbf{J}\mathbf{J}^T)}
 \end{equation}
+$h_{w}$ measures the ``distance'' of a given configuration to singularity. When $h_{w}$ is equal to 0, the configuration is singular;
+the greater $h_{w}$ is, the further away the configuration is from singularity.
+%~ We then derive:
+%~ \begin{equation}
+%~ h_{w}(\mathbf{q}, \mathbf{m}) = w(\mathbf{q})  (\mu \mathbf{n}^T \mathbf{m})
+%~ \end{equation}
 
@@
 can be precomputed off-line), that allow the planner to select the best
-candidates with respect to the local motion task along the guide path.
-This results in faster computation times, and a reduced number of overall contact switches.
+candidates according to user-defined criterion.
+%~ This results in faster computation times, and a reduced number of overall contact switches.

### 2016-01-26 body/equilibrium.tex
typos
finished pass with andrea

@@
 % !TEX root =  ../main.tex
-\section{Heuristics for relevant contact selection}
+\section{Heuristics for contact selection}
 \label{sec:heuristics}
@@
 The planner is designed so that any generated contact configuration is in static equilibrium.
-Equilibrium is of course critical in multipedal locomotion. For this reason we are interested in a robust 
-criterion, that ensures that the robot remains balanced in a real-world application, regardless of perception and control uncertainties.
+%~ Equilibrium is of course critical in multipedal locomotion. For this reason 
+We are interested in a robust 
+criterion, that ensures that the robot remains in equilibrium in a real-world application, regardless of perception and control uncertainties.
 
-We first give a linear program (LP) that asserts that a contact configuration is in static equilibrium.
-From this formulation we derive a new LP that quantifies the robustness of the configuration regarding equilibrium.
-In turn, from this value we can either define a heuristic to choose the most robust candidate, or set a threshold on the required robustness.
+We first give a linear program (LP) that verifies whether a contact configuration allows for static equilibrium.
+From this formulation we derive a new LP that quantifies the robustness of the equilibrium to uncertainties in the contact forces.
+In turn, from this value we can either choose the most robust candidate, or set a threshold on the required robustness.
 While the presented LP is original, it is based on an analysis of the problem that we proposed in \citep{Prete2016}, where the interested reader can find more details.
@@
 \subsubsection{Conditions for static equilibrium.}
-We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:
+We first define the variables of the problem, for $e$ contact points, expressed in world coordinates:
 \begin{itemize}
 \item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
-\item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
+%~ \item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
 \item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
@@
 
-According to Coulomb's law, the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
+According to Coulomb's law, the nonslipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
 As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays of unit length. We choose for instance:
@@
 \begin{equation*}
-\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} : \bm{\beta}_i \ge 0 \wedge \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i,
+\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} : \bm{\beta}_i \ge 0 \text{ and } \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i,
 \end{equation*}
-where $\bm{\beta}_i$ contains the coefficient of the cone generators.
+where $\bm{\beta}_i$ contains the coefficients of the cone generators.
 We can then stack all the constraints to obtain:
 \begin{equation}\label{eq:gen}
-\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \wedge \mathbf{f} = \mathbf{V} \bm{\beta},
+\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \text{ and } \mathbf{f} = \mathbf{V} \bm{\beta},
 \end{equation}
@@
 
-From the Newton-Euler equations, and assuming a null acceleration of the COM, we want the contact forces to compensate the gravity at the center of mass:
+From the Newton-Euler equations, to be in static equilibrium the contact forces have to compensate the gravitational forces:
 
@@
 
-If there exists a $\bm{\beta}^*$ satisfying \eqref{eq:gen} and \eqref{eq:new_eul}, it means that the configuration is statically balanced.
-The problem can then be formulated as a LP:
+If there exists a $\bm{\beta}^*$ satisfying \eqref{eq:gen} and \eqref{eq:new_eul}, it means that the configuration is in static equilibrium.
+The problem can then be formulated as an LP:
 
@@
 
-We observe that if $b_0$ is positive then \eqref{eq:lin_prog} admits a solution, and $b_0$ is proportional to the minimum distance of the equilibrium contact forces to the boundaries of the friction cones.
+We observe that if $b_0$ is positive then \eqref{eq:lin_prog} admits a solution, and $b_0$ is proportional to the minimum distance of the contact forces to the boundaries of the friction cones.
 If $b_0$ is negative, the configuration is not in static equilibrium, and $b_0$ indicates ``how far'' from equilibrium the configuration is. We thus use $b_0$ as a measure of robustness.
@@
 
-In conclusion, the LP~\eqref{eq:dual} provides a method to assert rapidly static equilibrium, and returns a robustness value.
-This robustness criterion can be immediately transformed into a heuristic, and eventually be pondered with others.
-Another option to enforce robustness is to set a minimum acceptable value for $b_0$, but as a result this conservative approach is not complete.
+%~ In conclusion, the LP~\eqref{eq:dual} provides a method to check rapidly static equilibrium, and returns a robustness value.
+%~ This robustness criterion can be immediately transformed into a heuristic, and eventually be weighted with others.
+%~ Another option to enforce robustness is to set a minimum acceptable value for $b_0$, but as a result this conservative approach discards valid solutions.

### 2016-01-26 body/results.tex
finished pass with andrea

@@
 
-We say that the planning is \gls{interactive} when the computation time for one step is lesser than the
-time to execute it. We arbitrarily approximate this time to one second.
+%~ We say that the planning is \gls{interactive} when the computation time for one step is lesser than the
+%~ time to execute it. We arbitrarily approximate this time to one second.
 
@@
 the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 
-The companion video presents the complete contact sequence obtained in all these scenarios.
+%~ The companion video presents the complete contact sequence obtained in all these scenarios.
 
@@
 
-\noindent\textbf{Heuristics:} The extended manipulability $h_w$ is chosen for the feet; $h_{EFORT}$ is chosen for the right arm.
+\noindent\textbf{Heuristics:} The manipulability $h_w$ is chosen for the feet; $h_{EFORT}$ is chosen for the right arm.
 Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
@@
 
-Again, the interpolation motion between the contact stances is out of the scope of this paper.
+Again, the interpolation motion between the contact steps is out of the scope of this paper.
 However it should be noted that the computed plan in this scenario has been executed successfully on the robot~\citep{Carpentier2016}.
@@
 \subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing}):}
-From a bent configuration, a standing-up motion is computed in a cluttered environment.
+From a bent configuration, a standing-up motion is computed in a constraining environment.
 The resulting motion involves using a wall as support, and climbing a 25-cm high step.
@@
 \noindent\textbf{Observations:} The difficulty of this scenario lies in the strong reduction of the reachable workspace induced 
-by the extreme proximity of all obstacles. The planner is able to find a sequence, that consists in many stances.
-The proximity of the obstacles invalidate a large number of contact candidates because of collisions. To respect the condition
-that only one contact must be broken between each phase, the motion has to be decomposed into a large number of stances (61 in average).
-While this scenario is the slower to solve, the planner still computes a solution interactively.
+by the extreme proximity of all obstacles. The planner is able to find a sequence, that consists in many steps.
+The proximity of the obstacles invalidate a large number of contact candidates because of collisions.
+To avoid breaking more than one contact between each step, the motion has to be decomposed into a large number of steps (61 in average).
+While this scenario is the slowest to solve, the planner still computes a solution \glslink{interactive}{interactively}.
  % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
@@
 \noindent\textbf{Observations:} In this context, setting up a really important minimum value for $b_0$ is possible due to the high
-stability of the HyQ robot, and results in more contact switches, in exchange for safety.
-The path-planning time is higher than for the previous HRP-2 robots due to the constraint that the 4 reachable workspace of all legs must
+stability of the HyQ robot, and results in more contact switches, in exchange for safety. The path-planning takes a few seconds due to the constraint that the 4 reachable workspaces of all legs must
 collide with the environment at all times. %\adnote{Why do not relax this constraint and require only 3 legs?} 
-Again, the computed time remains however \gls{interactive}.
+Again, the computation times remain however \gls{interactive}.
 
 \subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs}):}
-In this long scene, HyQ is first required to cross a 55-cm large hole; then, to cross a narrow ``bridge'', large of only 25-cm.
+In this long scene, HyQ is first required to cross a 55-cm large hole; then, to cross a narrow ``bridge'', only 25-cm large.
 
@@
 
-\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $4$.
+\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $10$.
 
@@
 Second, the narrow bridge is hard both for the planner and the contact generator: to make sure that equilibrium is preserved along the traversal,
-the bridge must be approached with the right angle.
+the bridge must be approached with the appropriate angle.
 The difficulty is illustrated in Figure~\ref{fig:HyQ_obs}, where several feet rearrangements are required to cross the hole (although the video shows this best).
@@
 This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
-our contact planner to compute a guide path for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
+our contact planner to compute a contact sequence for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
 Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences in less than 5 seconds.
@@
 \subsection{Influence of the parameters} \label{sec:influence}
-We discuss the several factors can influence the outcome of our planner: the root scaling factor $s$, the choice of heuristics for
-contact generation, the parametrization of the static equilibrium robustness, and lastly, the discretization 
+This subsection discusses the several factors that influence the outcome of our planner: the root scaling factor $s$, the choice of heuristics for
+contact generation, the desired value of the robustness parameter $b_0$, and lastly, the discretization 
 of the guide path.
@@
 For several values of $s$, we generated $10 000$ configurations. 
-We then computed the rates of false positives (ie the resulting configuration satisfies
-the reachability condition, but does not belong to \gls{ccontact0}) and false negatives (ie the resulting configuration does not satisfy the 
+We then computed the rates of false positives (i.e. the configuration satisfies
+the reachability condition, but does not belong to \gls{ccequil0}) and false negatives (i.e. the configuration does not satisfy the 
 reachability condition, but belongs to \gls{ccequil0}).
@@
 statistical tests are not really conclusive since we are only interested in a small area of the environment).
-As it can be expected, the scaling results in a high increase of the false negative, while the false positive decreases.
+As it can be expected, the scaling results in a high increase of the false negatives, while the false positives decrease.
 For HRP-2 we decided to set $s^*=1.2$, that results in less than 3 \% of false positives, and was also effective
@@
 the generation of the plan, the resulting interpolation on the real robot turned out to be harder \citep{Carpentier2016}.
-For this reason, we now prefer to use our manipulability based heuristic for the legs of the robot, but we still
-use EFORT for the arm, that results in less contact repositionning.
+For this reason, we now prefer to use our manipulability-based heuristic for the legs of the robot, but we still
+use EFORT for the arms, that results in less contact repositioning.
 
-\subsubsection{parametrization of the robustness equilibrium criterion} \label{sec:parrob}
+\subsubsection{On the robustness equilibrium criterion} \label{sec:parrob}
 Robustness is really important when considering practical applications on the robot, to account 
-for the various uncertainties that result from environment and state estimator. However,
-maximizing the robustness criterion is not necessarily optimal, because the resulting contacts will stay close to the center
-of mass, thus not favoring the motion. In our experiments, we choose not to maximize the robustness,
+for the various uncertainties that result from environment and state estimation. However,
+maximizing the robustness criterion is not necessarily optimal, because the resulting configurations may be too conservative, thus not favoring the motion.
+In our experiments, we choose not to maximize the robustness,
 but to empirically set a robustness threshold value under which a configuration is not considered to be in
@@
 the planner may fail since we impose the constraint that only one contact change might occur
-between two consecutive stances. For instance, in most scenarios the torso of HRP-2 translates by about 15 cm between two postures, but about only 3 cm
-for the car egress scenario for instance, where we had to lower the step size because of the many contact breaks, induced by the extremely constrained geometry.
+between two consecutive steps. For instance, in most scenarios the torso of HRP-2 moves about 15 cm between two postures, but only 3 cm
+for the car egress scenario, where we had to lower the step size because of the many contact breaks induced by the extremely constraining geometry.
 For future work, we would like to automatically adapt the size of the discretization step to the complexity of the planning, for instance
-by reducing the step size when in the areas where failure occur.
+by reducing the step size when failure occurs.
 
@@
 \subsubsection{Computation times}
-Table~\ref{tab:requestime} summarizes the performance measurement obtained, in terms of computation times.
+Table~\ref{tab:requestime} summarizes the performance measurements obtained, in terms of computation times.
 
@@
 it is also applied every time a new candidate needs to be evaluated. In particular for the car egress scenario,
-the kinematic constraints are really strong. Therefore a large number of projections fail because of joint limits violation.
+the kinematic constraints are really strong. Therefore a large number of projections fail because of joint-limit violations.
 
-For HyQ, there is a more regular repartition of the performances for the rubble scenario. For the obstacle scenario,
+For HyQ, there is a more uniform repartition of the times for the rubble scenario. For the obstacle scenario,
 we observe a large increase of the time spent performing collision checking and inverse kinematics. This is explained
@@
 
-In all scenarios, one can observe that the average computation time for one single step is largely below the second,
+In all scenarios, one can observe that the average computation time for one single step is largely below one second,
 thus allowing to consider \gls{interactive} applications.
 
+\subsubsection{Success rates}
 Table~\ref{tab:requestpercent} summarizes the success rates obtained for each scenario.
-
-\subsubsection{Success rates}
-We observe that, as expected, our planner does not succeed systematically, because of the approximations made in the implementation
-of our theoretical formulation. The extreme situation of the car egress scenario provides the less satisfying results in terms of kinematic failure. This is not
+We observe that, as expected, our planner does not succeed systematically, because of the approximations made in our formulation.
+The extreme situation of the car egress scenario provides the less satisfying results in terms of kinematic failure.
+This is not
 surprising considering the narrowness of the scene.
@@
 Comparing our method with others is not trivial. Indeed, we focus on the contact planning aspect of the problem, and do not address
-the interpolation in this paper, nor provide a strong guarantee that our sequence can be interpolated.
-Papers in the state of the art, when they indicate the computation times, often only provide an overall computation
-time, and we can not afford the important development resources required to reimplement them.
+the interpolation in this paper, nor provide a strong guarantee that our sequences can be interpolated.
+Most papers in the literature only provide the overall computation
+time, without specifying how much is spent in each phase. Due to the complexity
+of the existing approaches,  we cannot afford the development resources required to reimplement them.
 
-Furthermore, the interpolation itself is a problem that is far from being solved.
+Furthermore, the interpolation itself is not completely solved.
 For instance in \cite{DBLP:conf/iser/EscandeKMG08}, the end-effector trajectories are computed automatically with splines, but are manually corrected if they enter in collision.
@@
 
-Regardless, we indicate in Table~\ref{tab:compprev} what is actually computed by the method and in how much time.
-We compare motions obtained in previous works with HRP-2 (or a humanoid avatar) with ours: the two scenarios we selected consist in:  climbing one step, high of 20, 30 or 40 centimeters; 
-moving out of a table, that we consider of a complexity similar to the car egress scenario (or inferior since the feet location remain on the ground in the former case).
-For scenarios that apply, we also indicate the interpolation time that we obtain by feeding the plan to our interpolation method presented in \cite{Carpentier2016}.
+Nonetheless, we indicate in Table~\ref{tab:compprev} what is actually computed by each method and in how much time.
+We compare motions obtained in previous works with HRP-2 (or a humanoid avatar) with ours.
+The two scenarios we selected consist in:  climbing one step, of height 20, 30 or 40 centimeters; 
+getting out of a table, that we consider of a complexity similar to the car egress scenario (or inferior since the feet locations remain on the ground in the former case).
+For scenarios that apply, we also indicate the interpolation time that we obtained by feeding the plan to our interpolation method presented in \cite{Carpentier2016}.
 

### 2016-01-26 main.tex
nico + glossary
complete, need abstract and section 4
typos
finished pass with andrea

@@
 \usepackage{bm}
+\usepackage{booktabs}
+\usepackage{multirow}
 
@@
 \usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+\usepackage{glossaries}
 
@@
 \def\volumeyear{2015}
+
+\makeglossaries
+\input{body/glossary}
 
@@
  %~ Franck Multon\affilnum{3},
-\affiliation{\affilnum{1}LAAS-CNRS, Toulouse, France\\
+\affiliation{\affilnum{1}LAAS-CNRS / Universit\'e Paul Sabatier, Toulouse, France\\
 \affilnum{2}Inria, Rennes, France\\
@@
 This research is supported by Euroc (project under FP7 Grant Agreement  608849),  Entracte (ANR  grant  agreement  13-CORD-002-01),
-the ARO Contract W911NF-14-1-0437,
-the NSF award 1305286, and a grant from Boeing.
+the ARO Contract W911NF-14-1-0437 and the NSF award 1305286. %, and a grant from Boeing.
 \theendnotes
+
+\glsaddall
+\printglossaries
 \bibliographystyle{SageH}

### 2016-01-26 body/appendix_algo.tex
conclusion and appendices as discussed with andrea

@@
 		\While {$i < length(\mathbf{Q}^0)$}
-			\State State $previous = last\_element(states)$
-			\State State $s = \pi(previous, \mathbf{Q}^0[i])$
+			\State State $pState = last\_element(states)$
+			\State State $s =$ \textsc{GenFullBody}$(pState, \mathbf{Q}^0[i])$
 			\If {$s != NULL $}
@@
 				\EndIf				
-				\State $s = $\textsc{RepositionContacts}$(previous)$
+				\State $s = $\textsc{RepositionContacts}$(pState)$
 			\EndIf
@@
 
-At each step, $g$ is called with the previous state as a parameter, as well
-as a new root placement. $g$ returns a new contact configuration, if it succeeded
+At each step, \textsc{GenFullBody} is called with the previous state as a parameter, as well
+as a new root configuration. \textsc{GenFullBody} returns a new contact configuration, if it succeeded
 in computing a configuration with only one contact switch occurring.
-If $g$ failed in achieving this, the method \textsc{RepositionContacts} is called.
+If \textsc{GenFullBody} failed in achieving this, the method \textsc{RepositionContacts} is called.
 It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
@@
 
-The pseudo code for the method $g$ is given by Algorithm~\ref{alg:pi}.
+The pseudo code for the method \textsc{GenFullBody} is given by Algorithm~\ref{alg:pi}.
 
@@
 	%~ \Function{GenerateConfiguration}{}
-	\Function{$g$}{$pState$,$\mathbf{q}^0$}
+	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
 		\State State $newState$
@@
 		\State $nbContactsBroken = 0$
-		\For {\textbf{each} Limb $k$ in $previous.contactLimbs$}
+		\For {\textbf{each} Limb $k$ in $pState.contactLimbs$}
 			\If {$!$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
@@
 		\EndFor
-		\For {\textbf{each} Limb $k$ in $previous.freeLimbs$}
+		\For {\textbf{each} Limb $k$ in $pState.freeLimbs$}
 			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
 				\State $push(newState.contactLimbs,k)$
+				\State $remove(newState.freeLimbs,k)$		
 				\State \textbf{return} $newState$
@@
 		\EndFor
+		\If {\textsc{IsInStaticEquilibrium}$(newState)$}
+			\State \textbf{return} $newState$
+		\Else
 			\State \textbf{return} $NULL$
+		\EndIf
 	\EndFunction
@@
 If it succeeds, the new limb configuration is assigned to $k$. If it fails, a random collision free configuration is assigned to $k$.
+
+The method \textsc{IsInStaticEquilibrium} returns whether a given state is in static equilibrium.
 

### 2016-01-26 body/appendix_roms.tex
conclusion and appendices as discussed with andrea

@@
 Regarding the procedure, we can see that step 2 is conservative (Figure~\ref{fig:hrp2_roms}--right), which 
-is acceptable, especially because the lost set essentially relates to configurations close to singularity (they are close to the boundaries of the reachable workspace).
-We choose again to be less complete
-but more efficient, regarding the number of collision tests to be performed by RB-RRT.
+is acceptable, especially because the lost set essentially relates to configurations close to singularity (they are close to the boundaries of the reachable workspace, and
+often not \gls{contact feasible}, as illustrated in Figure~\ref{fig:dedefeas}, where the exterior boundaries of the reachable workspace appear
+red, thus not belonging to \gls{ccontact0}). We choose again to be less complete but more efficient, regarding the number of collision tests to be performed by RB-RRT.
 In step 1 on the other hand, selecting the convex hull (Figure~\ref{fig:hrp2_roms}--middle) instead of a minimum encompassing shape (Figure~\ref{fig:hrp2_roms}--left) may introduce false positives.

### 2016-01-26 body/conclusion.tex
conclusion and appendices as discussed with andrea

@@
 The low computational burden of our framework however allows for fast re-planning in case of failure.
+Furthermore because of this approximation, the guide search is not complete. The choice is deliberate, because we are convinced
+that it is necessary to trade completeness for efficiency at all stages of the planner.
+However, one direction for future work is to focus on a more accurate formulation of $C_{reach}^0$ to improve
+the approximation.
 
-One direction for future work is to focus on a more accurate formulation of $C_{reach}^0$ to address 
-this limitation.
-Our main objective is of course to generate the complete interpolation between the contact sequences.
-We have already obtained some success for some of the computed sequences~\citep{Carpentier2016}, but additional
+Another limitation of the method is that it currently only applies to \gls{cluttered} problems.
+However, it should first be noted that many problems belong to this class: for instance all the problems
+at the DRC were \gls{cluttered}. Our method is thus already useful for many cases.
+One way of extending its range of application, that we 
+consider for future work, is to include the equilibrium criterion when solving $\mathcal{P}_1$.
+Considering the set of obstacles intersecting with the reachable workspace for a given root configuration as candidates surfaces, we can use them to verify the equilibrium criterion.
+This would give us a necessary condition for \equilibriumfeasibility.
+
+Regarding the interpolation between the contact sequences ($\mathcal{P}_3$), we have already obtained some success for some of the computed sequences~\citep{Carpentier2016}, but additional
 work is required on the planning side to obtain a seamless workflow. To achieve this, we are currently working on the notion of transition certificate, i.e. formulating
@@
 %~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
-Lastly, we aim at going beyond static equilibrium and performing kinodynamic planning. We believe that the most promising direction in this regard is to integrate
+A last limitation of our method is that only static equilibrium configurations are considered for contact planning.
+We aim at performing kinodynamic planning to overcome this limitation. We believe that the most promising direction in this regard is to integrate
 the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
-Addressing these two issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.
+Addressing these two last issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.

### 2016-01-27 body/appendix_algo.tex
nicolas comments

@@
 
-Then, proving that $\mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} = (\mathbf{J}\mathbf{J}^T)^{-1}$ is straight-forward,
-using the SVD decomposition of each term.
+Then, the equality $\mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} = (\mathbf{J}\mathbf{J}^T)^{-1}$ follows from the SVD decomposition of each term
+~\citep{ben2003generalized}.

### 2016-01-27 body/efort.tex
nicolas comments

@@
 
-\subsubsection{The force and velocity ellipsoids.}
+\subsubsection{The force and velocity ellipsoids:}
 For a limb configuration $\mathbf{q}^k$, the Jacobian matrix
@@
 
-\subsubsection{Manipulability-based heuristics.}
+\subsubsection{Manipulability-based heuristics:}
 From these definitions, we can derive three useful heuristics, that all account for the environment and the task being performed.
@@
 \begin{equation}
-h_{EFORT}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \mathbf{m})
+h_{\textrm{\it EFORT}}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \mathbf{m})
 \end{equation}
@@
 
-$h_{EFORT}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
+$h_{\textrm{\it EFORT}}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
 $EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
@@
 
-$h_{EFORT}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
+$h_{\textrm{\it EFORT}}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
 can be precomputed off-line), that allow the planner to select the best

### 2016-01-27 body/equilibrium.tex
nicolas comments

@@
 
-\subsubsection{Conditions for static equilibrium.}
+\subsubsection{Conditions for static equilibrium:}
 We first define the variables of the problem, for $e$ contact points, expressed in world coordinates:

### 2016-01-27 body/intro.tex
nicolas comments

@@
 In particular, \citeauthor{Carpentier2016} are able to achieve this with \gls{interactive} performances on a real robot.
-Therefore, a planner capable of solving \Pa and \Pb efficiently could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
+Therefore, a planner capable of efficiently solving \Pa and \Pb could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
 The main contribution of this paper is exactly this planner.
@@
 We call this property \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
-This property has not been studied yet; the only way to validate a waypoint in the path is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
+This property has not been studied yet; the only rigorous way to validate a waypoint in the path is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
 \\
@@
 
-\subsection*{Comparison with our previous work.}
+\subsection*{Comparison with our previous work}
 The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep{tonneauisrr15}.

### 2016-01-27 body/overview.tex
part 3 with andrea
nicolas comments

@@
 We make the assumption that for the \textit{\gls{cluttered} contact planning problem}, most of the times \gls{contact feasible} root configurations are also \gls{equilibrium feasible}.
-For this reason we neglect C2 during the resolution of \Pa, and we consider it only when solving \Pb.
-However, even verifying \glslink{contact feasible}{\textit{contact feasibility}} is too computationally expensive because it requires the projection of the whole-body configuration into a contact posture.
+For this reason we neglect C2 during the resolution of $\mathcal{P}_1$, and we consider it only when solving \Pb.
+However, even verifying \glslink{contact feasible}{\textit{contact feasibility}} is computationally too expensive because it requires the projection of the whole-body configuration into a contact posture.
 
@@
 The task efficiency is measured using the Extended FORce Transmission ratio (EFORT), proposed by ~\cite{Tonneau:2014:TEC:2619648.2619652}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
+  \caption{
+           Reachable workspace and torso bounding box of HyQ. The green shapes represent the reachable workspace $W^k$ of each limb. The red shape is $W^0$.}
+		   \label{fig:HyQ_roms}
+\end{figure}
 
@@
 \medskip
-The volume encompassing $R^0$ is denoted $W^0$.
+The volume encompassing $R^0$ is denoted $W^0$ (Figure~\ref{fig:HyQ_roms}).
  %~ (Figure~\ref{fig:character}-right: central cylinder). 

### 2016-01-27 body/rbprm.tex
nicolas comments

@@
 
-\subsubsection*{Contact feasibility: necessary condition}
+\subsubsection*{Contact feasibility, a necessary condition:}
 For a contact to be possible, a volume $O_i \in O$ necessarily intersects with the reachable workspace $W(\mathbf{q}^{0})$ (Figure~\ref{fig:contact_gen}--1).
@@
 
-Therefore we can define an outer approximation  $C^0_{Nec} \supset$ \gls{ccontact0} defined as: 
+Therefore we can define an outer approximation  $ \mathcal{C}^0_{\textrm{\it Nec}} \supset$ \gls{ccontact0} defined as: 
 \begin{equation}
-C_{Nec}^0 = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+\mathcal{C}^0_{\textrm{\it Nec}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
 \end{equation}
-%~ The inclusion \gls{ccontact0} $\subset C^0_{Nec}$ is straightforward, but it is very important.
-%~ Planning in $C^0_{Nec}$ allows a strong reduction of the search space, while not discarding 
+%~ The inclusion \gls{ccontact0} $\subset \mathcal{C}^0_{\textrm{\it Nec}}$ is straightforward, but it is very important.
+%~ Planning in $\mathcal{C}^0_{\textrm{\it Nec}}$ allows a strong reduction of the search space, while not discarding 
 %~ any root configuration that could lead to a solution.
 
-%~ The condition defining $C^0_{Nec}$ is only necessary. This means that it might not be possible to extend a guide path planned 
-%~ in $C^0_{Nec}$ into a sequence of contact configurations.
+%~ The condition defining $\mathcal{C}^0_{\textrm{\it Nec}}$ is only necessary. This means that it might not be possible to extend a guide path planned 
+%~ in $\mathcal{C}^0_{\textrm{\it Nec}}$ into a sequence of contact configurations.
 
-\subsubsection*{Contact feasibility: sufficient condition}
+\subsubsection*{Contact feasibility, a sufficient condition:}
 %~ In building $C_{reach}^1$, we directly considered the including hull of the root body $W^{0}$, obtaining a necessary condition. 
-A trivial sufficient condition for \glslink{contact feasible}{\textit{contact feasibility}} can be constructed as a variation of $C^0_{Nec}$, by replacing $W^0$ with a bounding volume $B^{Suf}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
-We denote by \mbox{$C_{Suf}^0 \subset $ \gls{ccontact0}} the set of root configurations corresponding to this sufficient condition:
+A trivial sufficient condition for \glslink{contact feasible}{\textit{contact feasibility}} can be constructed as a variation of $\mathcal{C}^0_{\textrm{\it Nec}}$, by replacing $W^0$ with a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
+We denote by \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ \gls{ccontact0}} the set of root configurations corresponding to this sufficient condition:
 
 \begin{equation}
-C_{Suf}^0 = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } B^{Suf}(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+\mathcal{C}^0_{\textrm{\it Suf}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } B^{\textrm{\it Suf}}(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
@@
 %~ In general, the inclusion is strict, which means that we lose the completeness of the two-stage contact planner (i.e.
-the planner is not able to discover a path inside \mbox{$C_{reach} \setminus C_{Suf}^0$}).
+the planner is not able to discover a path inside \mbox{$C_{reach} \setminus \mathcal{C}^0_{\textrm{\it Suf}}$}).
 However, the sufficient condition guarantees that any such path leads to a valid sequence of contacts.
@@
 The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \gls{contact feasible}.
-An ideal shape $B$ (with $W^0 \subset B \subset B^{Suf}$) that leads to a necessary and sufficient condition may exist---even if it seems intuitively very unlikely in general. 
+An ideal shape $B$ (with $W^0 \subset B \subset B^{\textrm{\it Suf}}$) that leads to a necessary and sufficient condition may exist---even if it seems intuitively very unlikely in general. 
 %~ Its construction is out of the scope of this work.
@@
 %~ It might be possible to find an ideal shape $B$ that leads to a necessary and sufficient condition; however, it seems intuitively very unlikely in general.
-The construction of a shape $W^0 \subset B \subset B^{Suf}$ leading to a necessary and sufficient condition (or the proof of its nonexistence) is out of the scope of this work.
+The construction of a shape $W^0 \subset B \subset B^{\textrm{\it Suf}}$ leading to a necessary and sufficient condition (or the proof of its nonexistence) is out of the scope of this work.
 
-However, using a shape between $W^0$ and $B^{Suf}$ leads to a trade-off between a necessary and a sufficient condition.
+However, using a shape between $W^0$ and $B^{\textrm{\it Suf}}$ leads to a trade-off between a necessary and a sufficient condition.
 We define $W^0_s$ as the volume $W^0$ subject to a scaling transformation by a factor $s \in \mathbb{R}^+$.
@@
 %
-If $s=1$, then $W^0_s$ = $W^0$, such that $C_1^0$ = $C_{Nec}$. We thus consider that $s \geq 1$, since smaller values would only worsen the approximation.
+If $s=1$, then $W^0_s$ = $W^0$, such that $C_1^0$ = $\mathcal{C}^0_{\textrm{\it Nec}}$. We thus consider that $s \geq 1$, since smaller values would only worsen the approximation.
 By increasing $s$, the condition can become sufficient.

### 2016-01-27 body/results.tex
nicolas comments

@@
 In our previous work~\citep{tonneauisrr15} additional results are demonstrated with various virtual avatars (Figure~\ref{fig:robots_old}).
-In this extension we choose to focus on actual robots. We invite the interested reader to watch the ISRR video\endnote{https://www.youtube.com/watch?v=LmLAHgGQJGA}, and 
+In this extension we choose to focus on actual robots. We invite the interested reader to watch the ISRR video (\url{http://youtu.be/LmLAHgGQJGA}), and 
 to refer to the previous paper for a discussion on these results.
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=0.5\linewidth]{figures/stair}
+  \includegraphics[width=1\linewidth]{figures/stair}
   \caption{
@@
 		   \label{fig:stair_robust}
-\end{figure*}
+\end{figure}
 
@@
 
-\noindent\textbf{Heuristics:} The manipulability $h_w$ is chosen for the feet; $h_{EFORT}$ is chosen for the right arm.
+\noindent\textbf{Heuristics:} The manipulability $h_w$ is chosen for the feet; $h_{\textrm{\it EFORT}}$ is chosen for the right arm.
 Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
@@
 Again, the interpolation motion between the contact steps is out of the scope of this paper.
-However it should be noted that the computed plan in this scenario has been executed successfully on the robot~\citep{Carpentier2016}.
+However it should be noted that the computed plan in this scenario has been executed successfully on the robot (\url{http://youtu.be/YjL-DBQgXwk#t=0m28s}).
 
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=0.5\linewidth]{figures/standing}
+  \includegraphics[width=1\linewidth]{figures/standing}
   \caption{
@@
 		   \label{fig:standing}
-\end{figure*}
+\end{figure}
 
@@
 
-\noindent\textbf{Heuristics:} $h_w$ for the feet, $h_{EFORT}$  for the hands.
+\noindent\textbf{Heuristics:} $h_w$ for the feet, $h_{\textrm{\it EFORT}}$  for the hands.
 
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=0.5\linewidth]{figures/darpa}
+  \includegraphics[width=1\linewidth]{figures/darpa}
   \caption{
@@
 		   \label{fig:darpa}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=0.5\linewidth]{figures/HyQ_obs}
+  \includegraphics[width=1\linewidth]{figures/HyQ_obs}
   \caption{
@@
 		   \label{fig:HyQ_obs}
-\end{figure*}
+\end{figure}
 
@@
 
-\noindent\textbf{Heuristics:} $h_w$  for all legs. $h_{EFORT}$  for the right arm. The robustness threshold is set to $2$.
+\noindent\textbf{Heuristics:} $h_w$  for all legs. $h_{\textrm{\it EFORT}}$  for the right arm. The robustness threshold is set to $2$.
 
@@
 
-\noindent\textbf{Heuristics:} $h_{EFORT}$ for all fingers.
+\noindent\textbf{Heuristics:} $h_{\textrm{\it EFORT}}$ for all fingers.
 
@@
 
-\subsubsection{Choosing the scaling factor $s$} \label{sec:params}
+\subsubsection{Choosing the scaling factor $s$:} \label{sec:params}
 To find a convenient value $s^*$ of $s$, we proceeded as follows. 
@@
 
-\subsubsection{On the choice of heuristics} \label{sec:heuristichoices}
+\subsubsection{On the choice of heuristics:} \label{sec:heuristichoices}
 In our conference paper, the computed motions were only generated using the EFORT heuristic.
@@
 the generation of the plan, the resulting interpolation on the real robot turned out to be harder \citep{Carpentier2016}.
-For this reason, we now prefer to use our manipulability-based heuristic for the legs of the robot, but we still
+For this reason, we prefer to use our manipulability-based heuristic for the legs of the robot, but we still
 use EFORT for the arms, that results in less contact repositioning.
 
-\subsubsection{On the robustness equilibrium criterion} \label{sec:parrob}
+\subsubsection{On the robustness equilibrium criterion:} \label{sec:parrob}
 Robustness is really important when considering practical applications on the robot, to account 
 for the various uncertainties that result from environment and state estimation. However,
-maximizing the robustness criterion is not necessarily optimal, because the resulting configurations may be too conservative, thus not favoring the motion.
+maximizing the robustness criterion is often not optimal, because the resulting configurations may be too conservative, thus not favoring the motion.
 In our experiments, we choose not to maximize the robustness,
@@
 
-\subsubsection{Discretization of the guide path} \label{sec:disc}
+\subsubsection{Discretization of the guide path:} \label{sec:disc}
 Currently, to discretize the path, the user defines a fixed step size. The step size
@@
 \subsection{Performance analysis} \label{sec:perf}
-To analyze performance, for each considered scenario, we ran the simulation 1000 times.
+To analyze performance, for each considered scenario, we ran the planner 1000 times.
 We measured the computation time spent in each aspect of the algorithm, and also analyzed the success
@@
 
-\subsubsection{Computation times}
+\subsubsection{Computation times:}
 Table~\ref{tab:requestime} summarizes the performance measurements obtained, in terms of computation times.
@@
 it is also applied every time a new candidate needs to be evaluated. In particular for the car egress scenario,
-the kinematic constraints are really strong. Therefore a large number of projections fail because of joint-limit violations.
+the kinematic constraints are very demanding. Therefore a large number of projections fail because of joint-limit violations.
 
@@
 
-\subsubsection{Success rates}
+\subsubsection{Success rates:}
 Table~\ref{tab:requestpercent} summarizes the success rates obtained for each scenario.
@@
  \hline
-   \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{5.42 min}  \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{2 to 10 min} \\
+   \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}5.42 min}  \\
 							 & \textbf{Ours} + \citeauthor{Carpentier2016}  & $\mathbf{<}$ \textbf{2 ms} & $\mathbf{<}$ \textbf{14 ms} & $ <$ 2s \\
+							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
  \hline
-   \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{4.08 min}  \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{2 to 10 min} \\
+   \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}4.08 min}  \\
 							 & \textbf{Ours}  & $\mathbf{<}$ \textbf{2 ms} & $\mathbf{<}$ \textbf{14 ms} & X \\
+							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
  \hline
-   \multirow{3}{*}{Stair 40 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{10.08 min}  \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{2 to 10 min} \\
+   \multirow{3}{*}{Stair 40 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}10.08 min}  \\
 							 & \textbf{Ours}  & \textbf{5 s} & \textbf{3 s} & X \\
+							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
  \hline
-   \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 10 min & \multicolumn{2}{c}{3.5 hours}  \\
+   \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 10 min & \multicolumn{2}{c}{\cellcolor{blue!25}3.5 hours}  \\
 							 & \textbf{Ours}  & \textbf{529 ms} & \textbf{$<$ 21 s} & X \\
  \end{tabular}
-\caption{TODO VERIFIE STATS A LA MAISON.}
+\caption{Comparison between the computation times obtained by our method and previous ones, for each phase of the planning. Blue cells indicate merged times (because no
+precise time information is available). The X symbol indicates scenarios where we did not address the interpolation of the contact sequence.}
 \label{tab:compprev}
@@
 Most papers in the literature only provide the overall computation
-time, without specifying how much is spent in each phase. Due to the complexity
-of the existing approaches,  we cannot afford the development resources required to reimplement them.
+time, without specifying how much is spent in each phase. Moreover, there exists no off-the-shelf implementation
+of the related methods. Due to their complexity, we cannot afford the development resources required to reimplement them.
 

### 2016-02-03 body/abstract.tex
typos
final version !

@@
-We address the issue of automatically planning discrete sequences of contacts
-that characterize a multiped motion in a constraining environment.
-To overcome the inherent combinatorial difficulty of this problem, we separate it
-in two sub-problems: first, planning a guide path for the root of the robot and
-then, generating contacts along this path.
-We propose theoretical contributions for each sub-problem, and implement them
-as a complete contact planner. 
-First, we generate a guide root path without explicitly computing configurations in static equilibrium, as opposed to previous approaches.
-To do this we propose to approximate the space of equilibrium configurations using the ``reachability condition".
-Verifying this condition boils down to performing simple collision tests and, in the considered environments, this approximation turns out to be accurate.
-Then, we propose an algorithm to extend a root guide path into a sequence of equilibrium configurations.
-A sampling-based approach is used to generate candidate contacts.
-The best candidate is chosen according to user-defined heuristics. We propose four such heuristics, used in all our experiments, and 
-designed for generic robotic applications.
-These contributions are implemented within the open-source software Hpp.
-Our framework allows us to generate complex multi-contact sequences in at most a few seconds for a large variety of robots, including HRP-2, HyQ, and a dexterous hand.
+We present a framework capable of planning 
+a large variety of contact plans, which describe complex multiped
+motions (including humanoid) such as standing up, as climbing stairs using a handrail, crossing rubble and getting out of a car.
+As this choice of scenarios suggests, our framework answers a need demonstrated at the Darpa Robotics Challenge,
+where the lack of an automatic acyclic contact planner was recognized as a major issue.
+
+The key idea of the framework is the introduction of the reachability condition.
+Informally, it verifies that the root configuration of a robot is ``close, but not too close'' from the obstacles of the environment: 
+close enough to allow contact creation, but not too close to avoid collision.
+This approximation of the space of admissible root configurations allows the decomposition of the high-dimensional, combinatorial contact planning problem into two simpler sub-problems:
+first, to plan a guide path for the root of the robot without considering the whole-body configuration	 of the robot; then, to generate a discrete sequence of whole-body configurations in static equilibrium along this path.
+The reachability condition turns the computation of the guide path into a low dimensional collision checking problem, which can be solved in less than a few seconds.
+Then, a deterministic contact selection algorithm tackles the combinatorial aspect of the second problem, the generation of a discrete sequence along the guide path.
+Several innovations make it computationally efficient, including a criterion for verifying the static equilibrium of a configuration, and a set of heuristics used 
+to enforce desirable properties on the configuration.
+
+The design of our planner results from the pragmatic choice of favoring efficiency over exhaustiveness, which we justify empirically:
+in a few seconds, with satisfying success rates, we are able to generate complex contact plans for a large variety of scenarios and robots, including HRP-2, HyQ, and a dexterous hand.

### 2016-02-03 body/results.tex
final version !

@@
 \noindent\textbf{Observations:} In this context, setting up a really important minimum value for $b_0$ is possible due to the high
-stability of the HyQ robot, and results in more contact switches, in exchange for safety. The path-planning takes a few seconds due to the constraint that the 4 reachable workspaces of all legs must
-collide with the environment at all times. %\adnote{Why do not relax this constraint and require only 3 legs?} 
+stability of the HyQ robot, and results in more contact switches, in exchange for safety. The guide path-planning in this scenario takes a few seconds in average, more than
+in any other scenarios.
+This is explained by the necessity of discovering a safe way to ``climb down'' the rubble.
+In this part of the planning, the constraint that the 4 reachable workspaces of all legs must collide with the environment at all times is hard to respect, but enforces the equilibrium of HyQ.
+%\adnote{Why do not relax this constraint and require only 3 legs?}
 Again, the computation times remain however \gls{interactive}.
@@
  \hline
-   Stairs (18) & 6 -- \textbf{13} --  24 & 13 --  \textbf{32} -- 329   & 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{14} \\
-   Standing (24)& 4 -- \textbf{451} --  985 & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 310 --  \textbf{1622} -- 3429 & \textbf{66}  \\
-   Car (61)& 1 -- \textbf{529} --  995 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6775 --  \textbf{20890} -- 52963 & \textbf{342} \\
-   Rubble (71)& 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{3058} -- 13126 & \textbf{43} \\
-   Race (112)& 1 -- \textbf{501} --  995 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2530 --  \textbf{3722} -- 10030 & \textbf{33}
+   Stairs (18) & 5 -- \textbf{6} --  18 & 13 --  \textbf{32} -- 329   & 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{15} \\
+   %~ Standing (24)& 4 -- \textbf{451} --  985 & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 310 --  \textbf{1622} -- 3429 & \textbf{66}  \\
+   Standing (24)& 65 -- \textbf{1086} --  5227 & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 371 --  \textbf{2257} -- 7671 & \textbf{94}  \\
+   %~ Car (61)& 1 -- \textbf{529} --  995 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6775 --  \textbf{20890} -- 52963 & \textbf{342} \\
+   Car (61)& 145 -- \textbf{1055} --  15063 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6919 --  \textbf{21416} -- 67031 & \textbf{351} \\
+   %~ Rubble (71)& 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{3058} -- 13126 & \textbf{43} \\
+   Rubble (71)& 151 -- \textbf{15220} --  125867 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1548 --  \textbf{18058} -- 13126 & \textbf{254} \\
+   %~ Race (112)& 1 -- \textbf{501} --  995 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2530 --  \textbf{3722} -- 10030 & \textbf{33}
+   Race (112)& 267 -- \textbf{966} --  2257 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2797 --  \textbf{4187} -- 11292 & \textbf{37}
    %Standing up & $70$ & 23 & 48\\
  \end{tabular}
-\caption{minimum, \textbf{average} and worst time (in ms) spent in the generation process for each scenario and each critical part of the generation process. The last
-column indicates the average time necessary to compute one contact transition.}
+\caption{minimum, \textbf{average} and worst time (in ms) spent in the generation process for each scenario and each critical part of the generation process (not all parts are timed,
+thus the average total computation time is higher than the sum of each part). The last
+column indicates the average time necessary to compute one contact transition. }
 \label{tab:requestime}
@@
    \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 10 min & \multicolumn{2}{c}{\cellcolor{blue!25}3.5 hours}  \\
-							 & \textbf{Ours}  & \textbf{529 ms} & \textbf{$<$ 21 s} & X \\
+							 & \textbf{Ours}  & \textbf{529 ms} & \textbf{$<$ 22 s} & X \\
  \end{tabular}

### 2016-02-03 cover_letter.tex
final version !

@@
 a support for our argumentation on the strong choices made in the design
-of our planner (Section 1.2 and 2).
+of our planner (Section 1.2 and 2), and a comparison with methods from the literature.
 \end{itemize}

### 2016-02-03 body/abstract.tex
final submission

--- 2016-01-21 body/abstract.tex
@@
-We address the issue of automatically planning discrete sequences of contacts
-that characterize a multiped motion in a constraining environment.
-To overcome the inherent combinatorial difficulty of this problem, we separate it
-in two sub-problems: first, planning a guide path for the root of the robot and
-then, generating contacts along this path.
-We propose theoretical contributions for each sub-problem, and implement them
-as a complete contact planner. 
-First, we generate a guide root path without explicitly computing configurations in static equilibrium, as opposed to previous approaches.
-To do this we propose to approximate the space of equilibrium configurations using the ``reachability condition".
-Verifying this condition boils down to performing simple collision tests and, in the considered environments, this approximation turns out to be accurate.
-Then, we propose an algorithm to extend a root guide path into a sequence of equilibrium configurations.
-A sampling-based approach is used to generate candidate contacts.
-The best candidate is chosen according to user-defined heuristics. We propose four such heuristics, used in all our experiments, and 
-designed for generic robotic applications.
-These contributions are implemented within the open-source software Hpp.
-Our framework allows us to generate complex multi-contact sequences in at most a few seconds for a large variety of robots, including HRP-2, HyQ, and a dexterous hand.
+We present a framework capable of producing contact plans describing complex multiped motions (including humanoid): standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
+Our framework answers a need demonstrated at the Darpa Robotics Challenge,
+where the lack of an automatic acyclic contact planner was recognized a major issue.
+
+Our novel key idea is the reachability condition.
+Informally, it verifies that the root configuration of a robot is ``close, but not too close'' from obstacles: close to allow contact creation, not too close to avoid collision.
+With this approximation of the space of admissible root configurations we decompose the hard contact planning problem into simpler sub-problems: first, to plan a guide path for the root without considering the whole-body configuration; then, to generate a discrete sequence of whole-body configurations in static equilibrium along this path.
+The reachability condition turns the high-dimensional computation of the guide into a collision checking problem, solved in less than a few seconds.
+Then a deterministic contact selection algorithm tackles the combinatorial issue of  generating of a discrete sequence along the guide path.
+Several innovations make it computationally efficient: a criterion for verifying static equilibrium, and a set of heuristics used to enforce desirable properties on the configuration.
+
+Our approach results from the pragmatic choice of favoring efficiency over exhaustiveness, justified empirically: in a few seconds, with satisfying success rates, we generate complex contact plans for various scenarios and robots, namely HRP-2, HyQ, and a dexterous hand.

### 2016-02-03 body/results.tex
final submission

@@
    \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}5.42 min}  \\
-							 & \textbf{Ours} + \citeauthor{Carpentier2016}  & $\mathbf{<}$ \textbf{2 ms} & $\mathbf{<}$ \textbf{14 ms} & $ <$ 2s \\
+							 & \textbf{Ours} + \citeauthor{Carpentier2016}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & $ <$ 2s \\
 							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
@@
    \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}4.08 min}  \\
-							 & \textbf{Ours}  & $\mathbf{<}$ \textbf{2 ms} & $\mathbf{<}$ \textbf{14 ms} & X \\
+							 & \textbf{Ours}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & X \\
 							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\

### 2016-02-03 main.tex
nicolas comments
final version !
final submission

@@
 \usepackage{multirow}
+\usepackage[table]{xcolor}
 
@@
 \section*{Acknowledgements}
-This research is supported by Euroc (project under FP7 Grant Agreement  608849),  Entracte (ANR  grant  agreement  13-CORD-002-01),
-the ARO Contract W911NF-14-1-0437 and the NSF award 1305286. %, and a grant from Boeing.
-\theendnotes
+This research is supported by Euroc (project under FP7 Grant Agreement  608849);  Entracte (ANR  grant  agreement  13-CORD-002-01);
+the ARO Contract W911NF-14-1-0437; and the NSF award 1305286. %, and a grant from Boeing.
+%~ \theendnotes
 

### 2016-09-05 body/abstract.tex
Nouvel abstract Main.tex change pour pb de compilation.

@@
-We present a framework capable of producing contact plans describing complex multiped motions (including humanoid): standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
-Our framework answers a need demonstrated at the Darpa Robotics Challenge,
-where the lack of an automatic acyclic contact planner was recognized a major issue.
+We describe  a method to plan contacts allowing any multiped robot, including a humanoid, to achieve complex motions: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
 
-Our novel key idea is the reachability condition.
-Informally, it verifies that the root configuration of a robot is ``close, but not too close'' from obstacles: close to allow contact creation, not too close to avoid collision.
-With this approximation of the space of admissible root configurations we decompose the hard contact planning problem into simpler sub-problems: first, to plan a guide path for the root without considering the whole-body configuration; then, to generate a discrete sequence of whole-body configurations in static equilibrium along this path.
-The reachability condition turns the high-dimensional computation of the guide into a collision checking problem, solved in less than a few seconds.
+The Darpa Robotics Challenge recognized the lack  of an automatic acyclic contact planner to be a major issue.
+We adress by introducing the reachability condition which, informally, verifies that a robot root remains in a bounded volume above the environment surfaces, close enough to allow contact creation and far enough to allow collision free motion of the limbs.
+We achieve a complex motion by decomposing the hard contact planning into simpler sub-problems: first, to plan a guide path for the root without considering the whole-body configuration; second, to generate a discrete sequence of whole-body configurations in static equilibrium along this path.
+The reachability condition turns the high-dimensional computation of the guide into a collision checking problem.
 Then a deterministic contact selection algorithm tackles the combinatorial issue of  generating of a discrete sequence along the guide path.

### 2016-09-08 body/intro.tex
Introduction remaniée

@@
 % !TEX root =  ../main.tex
-We consider the problem of planning an acyclic sequence of contacts describing the motion of a multiped robot in a constraining environment.
-Acyclic contact planning is a particular class of motion planning in which the robot must be in contact with the environment at every configuration to maintain the static equilibrium.
 
-Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. However, executing
-this behavior on constraining environments is dangerous, if not impossible.
-In an analysis of their participation to the Darpa Robotics Challenge, \citeauthor{atkensondarpa}
-noted: ``Except for egress, no robots in the DRC Finals used
-the  stair  railings  or  any  form  of  bracing.   Even  drunk  people  are  smart  enough  to  use  nearby  supports.
-Full body locomotion (hand holds,  bracing,  leaning against a wall or obstacles) should be easier than our
-current high performance minimum contact locomotion approaches."
+Planning the locomotion of a multipedal robot in a complex environment - made of stairs, constraining obstacles or uneven surfaces - is still an open issue.
+Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}, but executing this behavior on constraining environments is dangerous if not impossible.
+In such a context, the robot has to perform a particular class of motion planning called acyclic contact planning, in which the robot must be in contact with the environment at every configuration to maintain the static equilibrium.
 
-Indeed, the current approach to locomotion planning is to avoid obstacles as much as possible, instead of using them
-to facilitate locomotion. The reason for this, as the authors state, is that ``More contacts make tasks
-mechanically easier, but algorithmically more complicated for planning, and the transitions are difficult to
-both plan and control[...].  We have seen very few robot planners that are  capable of  generating this  behavior."
-The difficulty of addressing such a problem comes both in practice from the proximity to obstacles (which makes tedious the sampling of collision-free configurations) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
+The lack of solution to this planning problem has been highlighted by Atkeson \citeauthor{atkensondarpa}.
+Actually, the current approach to locomotion planning tends to avoid obstacles as much as possible, instead of using them by making contacts to facilitate motion.
+One main reason is that contact planning is algorithmically difficult, both because of the proximity to obstacles which makes tedious the sampling of collision-free configurations, and because of the foliation of the configuration space where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
 
@@
 
-\subsection{Rationale}
+\subsection{Situation and Contributions}
 
@@
 \end{figure*}
-This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
-we believe that the separation between the generation of the root guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
-However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated. \\
 
-\noindent \textbf{Regarding $\mathcal{P}_1$}, the guide path must guarantee the existence of a contact sequence to actuate it.
-We call this property \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
-This property has not been studied yet; the only rigorous way to validate a waypoint in the path is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
-\\
+Our solution belongs the class of decoupled approaches, because we believe that the separation between the generation of the root guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of multi-contact planning.
+Compared to previous approaches, our solution has two main novelties:
 
-\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinitely many combinations of possible contact sequences for a given root path.
-The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to constraining environments (\citeauthor{bouyarmane:lirmm-00777727} and \citeauthor{DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, possibly leading to very tedious transitions).
-\\
-
-\textbf{The key idea of the paper is the \textit{reachability condition}}, a computationally-efficient approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
-We assume that most of the times the \textit{reachability condition} implies the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
-%root configurations in this approximation can be efficiently extended into full body configurations in static equilibrium.
-This assumption is not always verified, but we provide empirical evidence that it is in the considered class of problems, the \textit{\gls{cluttered} contact planning problems}.
-Such problems can be solved with a contact sequence 
-where in every configuration at least one contact is \gls{quasi-flat}~\citep{Prete2016}. A contact is \textit{quasi-flat} if the friction cone contains the direction opposite to gravity. 
-%~ This definition includes motions on flat or uneven terrains, but also more complex ones such as stair climbing, car egress, or standing up scenarios, that might require noncoplanar contacts.
-%Under this assumption we can take advantage of the decoupling approach to find a solution rapidly.
-
-To address $\mathcal{P}_1$, we first use the \textit{reachability condition} to plan the guide path. This step is fast because the planning happens
+\noindent \textbf{Regarding $\mathcal{P}_1$}, we define and use the \textit{reachability condition} to plan the guide path.
+The \textit{reachability condition}, a computationally-efficient approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
+This step is fast because the planning happens
 in a low-dimensional space, and does not require explicit contact computations.
 
-Then we address \Pb by extending the path into a sequence of whole-body configurations in static equilibrium.
+\noindent \textbf{Regarding $\mathcal{P}_2$},  we address \Pb by extending the path into a sequence of whole-body configurations in static equilibrium.
 This second step requires the explicit computation of contact configurations.
+It is guided by dedicated heuristics which fastly synthesize relevant configurations.
 
@@
 However, we demonstrate empirically the validity of the approach: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
-Similarly, in this paper we do not  prove that all the contact sequences produced by our planner can be interpolated ($\mathcal{P}_3$), although we have been able to do this for some of the demonstrated sequences~\citep{Carpentier2016}.
 
-\subsection{Paper contribution and organization}
-We propose contributions to both problems \Pa and $\mathcal{P}_2$.
+As a result, we propose contributions to both problems \Pa and $\mathcal{P}_2$.
 \begin{itemize}
@@
 more efficient with respect to the task.
-\item Statistical tests that empirically demonstrate the validity of our approach in the considered scenarios. 
 \end{itemize}
 
-%Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
-%Since the interpolation of a complete path between such configurations can be solved within \gls{interactive} constraints \citep{Carpentier2016}, we claim that our planner is the fastest multi-contact planning solution to our knowledge.
+%
+%
+%% TO BE MOVED IN DISCUSSION 
+%%
+%%the guide path must guarantee the existence of a contact sequence to actuate it.
+We call this property \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
+This property has not been studied yet; the only rigorous way to validate a waypoint in the path is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
+\\
+%
+%\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinitely many combinations of possible contact sequences for a given root path.
+The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to constraining environments (\citeauthor{bouyarmane:lirmm-00777727} and \citeauthor{DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, possibly leading to very tedious transitions).
+\\
+%
+%
+%\textbf{The key idea of the paper is the \textit{reachability condition}}, a computationally-efficient approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
+%We assume that most of the times the \textit{reachability condition} implies the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
+%%root configurations in this approximation can be efficiently extended into full body configurations in static equilibrium.
+%This assumption is not always verified, but we provide empirical evidence that it is in the considered class of problems, the \textit{\gls{cluttered} contact planning problems}.
+%Such problems can be solved with a contact sequence 
+%where in every configuration at least one contact is \gls{quasi-flat}~\citep{Prete2016}. A contact is \textit{quasi-flat} if the friction cone contains the direction opposite to gravity. 
+%%~ This definition includes motions on flat or uneven terrains, but also more complex ones such as stair climbing, car egress, or standing up scenarios, that might require noncoplanar contacts.
+%%Under this assumption we can take advantage of the decoupling approach to find a solution rapidly.
+%
+%%Similarly, in this paper we do not  prove that all the contact sequences produced by our planner can be interpolated ($\mathcal{P}_3$), although we have been able to do this for some of the demonstrated sequences~\citep{Carpentier2016}.
+%
+%
+%To address $\mathcal{P}_1$, we first use the \textit{reachability condition} to plan the guide path. This step is fast because the planning happens
+%in a low-dimensional space, and does not require explicit contact computations.
+%
+%Then we address \Pb by extending the path into a sequence of whole-body configurations in static equilibrium.
+%This second step requires the explicit computation of contact configurations.
 
-In Section~\ref{overview}, we present an overview of our method.
+%
+%
+%\subsection{Paper contribution and organization}
+%We propose contributions to both problems \Pa and $\mathcal{P}_2$.
+%\begin{itemize}
+%\item A low-dimensional, efficient sampling-based approach to plan guide paths.
+%\item A very efficient and general implementation of an acyclic contact planner, the first one compatible with \gls{interactive} applications.
+%\item Four heuristics for contact generation. They bias the planner towards configurations that are in robust static equilibrium, or 
+%more efficient with respect to the task.
+%\item Statistical tests that empirically demonstrate the validity of our approach in the considered scenarios. 
+%\end{itemize}
+%
+%%Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
+%%Since the interpolation of a complete path between such configurations can be solved within \gls{interactive} constraints \citep{Carpentier2016}, we claim that our planner is the fastest multi-contact planning solution to our knowledge.
+%
+%
+%In Section~\ref{overview}, we present an overview of our method.
 Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and $\mathcal{P}_2$.
@@
 Finally, in Section 6 we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios.
-%that complete those presented in the conference paper.
-
-\subsection*{Comparison with our previous work}
-The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep{tonneauisrr15}.
-The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
-This extension completes this work and brings new contributions, resulting from the implementation of our approach on real-world robots and problems.
-In particular, a strong contribution is the introduction of a robust equilibrium criterion, designed to ensure the equilibrium of the robot despite bounded errors in the contact forces, coupled
-with heuristics for relevant contact selection. These two contributions were required to apply our method to real-robot models, namely HRP-2 and HyQ~\citep{semini11hyqdesignjsce}.
-In the present paper, a complete algorithm for the contact planner is given, and an access to the source code of our implementation is provided.
-Furthermore, our experiments are supported by an in-depth analysis of our performances, as well as a discussion on the influence of the parameters to the method.
+%%that complete those presented in the conference paper.
+%
+%\subsection*{Comparison with our previous work}
+%The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep{tonneauisrr15}.
+%The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
+%This extension completes this work and brings new contributions, resulting from the implementation of our approach on real-world robots and problems.
+%In particular, a strong contribution is the introduction of a robust equilibrium criterion, designed to ensure the equilibrium of the robot despite bounded errors in the contact forces, coupled
+%with heuristics for relevant contact selection. These two contributions were required to apply our method to real-robot models, namely HRP-2 and HyQ~\citep{semini11hyqdesignjsce}.
+%In the present paper, a complete algorithm for the contact planner is given, and an access to the source code of our implementation is provided.
+%Furthermore, our experiments are supported by an in-depth analysis of our performances, as well as a discussion on the influence of the parameters to the method.

### 2016-12-06 cc.sh
compilation script

@@ -1,0 +1,3 @@
+pdflatex main.tex
+bibtex main.aux
+pdflatex main.tex

### 2016-12-08 body/intro.tex
ongoing intro rewriting

@@
 
-Planning the locomotion of a multipedal robot in a complex environment - made of stairs, constraining obstacles or uneven surfaces - is still an open issue.
-Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}, but executing this behavior on constraining environments is dangerous if not impossible.
-In such a context, the robot has to perform a particular class of motion planning called acyclic contact planning, in which the robot must be in contact with the environment at every configuration to maintain the static equilibrium.
+Planning legged robot locomotion in a complex environment (stairs, constraining obstacles, uneven surfaces \dots) remains an open issue.
+To move, a legged robot is required to sequentially create contacts with the environment while avoiding collisions.
+Finding such a contact sequence is an instance of the motion planning problem called acyclic contact planning.
 
-The lack of solution to this planning problem has been highlighted by Atkeson \citeauthor{atkensondarpa}.
-Actually, the current approach to locomotion planning tends to avoid obstacles as much as possible, instead of using them by making contacts to facilitate motion.
-One main reason is that contact planning is algorithmically difficult, both because of the proximity to obstacles which makes tedious the sampling of collision-free configurations, and because of the foliation of the configuration space where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
+Addressing this issue is a pre-requisite to the deployment of robots in human-centered environments, where wheeled robots are not able to navigate.
 
-Previous contributions that embrace the combinatorial provide complete approaches, not applicable in practice because they require hours of computations \citep{Bretl:2006:MPM:1124573.1124585}.
-More recent successes use a local solution, resulting in more reasonable computation times (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
-Neither global nor local methods present satisfying performances, because planning simultaneously the robot trajectory and the contacts that allow
-its execution is a very hard problem.
+To make a legged robot walk on a flat ground, efficient approaches based on the capture point(PRATT) can be used to determine the contact location. Choosing the effector
+to create a contact with is also trivial because walk follows a cyclic pattern.
+Unfortunately, in the general case planning complex contact interactions is extremely challenging.
+At any given time a contact choice must be made between infinitely many possibilities (often a combinatorial discrete choice for the effector, but an infinite choice of locations).
+Furthermore, the contact manifold is foliated~\citep{simeon-manipulation-04}: a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this decision brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles.
+Lastly, the contact manifold has a null measure and is thus impossible to randomly sample because it has a zero measure with respect to the configuration space.
 
-As suggested by \cite{Bouyarmane2009}, we believe that these two problems can be treated sequentially within a global planner.
-We go further and claim that this can be done at a much smaller cost, provided we can formulate a computationally-efficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} of the root configuration (i.e.
+%~ This makes the simultaneous planning of a center of mass trajectory with a contact sequence tedious.
+
+ %~ The contact manifold can be seen as a building with an infinite number of floors and elevators, where each elevator only stops at a subset of the floors.
+We don't have a map of this building.
+In this metaphor, the kinematic and dynamic constraints associated with a contact configuration restrict the motion to lie on
+%~ a floor, until an elevator can be reached, to safely transition to another floor (ie a new contact configuration). Exploring efficiently the building to find 
+%~ the sequence of floor paths and elevator transitions is the scientific question to address.
+ %~ Furthermore, because the contact manifold has a zero measure in the configuration space, we can't sample a configuration that lie on a floor.
+
+%~ This contact manifold has a null measure, which makes it impossible to directly sample contact configurations. Even worse, 
+%~ the contact manifold is foliated~\citep{simeon-manipulation-04}: each contact decision creates kinematic and dynamic constraints that must 
+%~ be respected to transition to another state.
+
+%~ Contact interaction is crucial to maintain equilibrium, though extremely challenging:
+%~ \begin{enumerate}
+%~ \item verifying that a contact allows the motion to be continuously dynamically feasible is computationally expensive; 
+%~ \item the contact manifold has a zero measure (a contact configuration cannot be sampled);
+%~ \item the choice of contacts along the motion (e.g. left or right hand) adds a combinatorial issue, and the set of possible contact locations is infinite; 
+%~ \item kinematic constraints appear upon contact creation, introducing an infinite foliation of the contact manifold~\citep{simeon-manipulation-04};
+%~ \item in theory, maintaining the balance of the robot requires the simultaneous planning of the root trajectory and the contact locations and timing, which increases even more the search space.
+%~ \end{enumerate}
+Most locomotion systems avoid these issues thanks to simplifying assumptions that reduce the complexity (coplanar contacts, cyclic patterns that are not combinatorial)~\citep{Kajita03a}.
+At the DARPA challenge, instead of using the environment for creating contacts that facilitate motion, the chosen strategies to locomotion planning avoided obstacles as much as possible~\citep{atkensondarpa}.
+Previous contributions having demonstrated acyclic contact locomotion have not been used because they are either too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585} or dynamically inconsistent~\citep{Mordatch:2012:DCB:2185520.2185539}.
+
+As suggested by \cite{Bouyarmane2009}, contrary to observation 5 we believe trajectory and contact planning can be decoupled.
+We go further and claim that this can be done at a much smaller cost, provided we can formulate a computationally-efficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} of a root configuration (i.e.
 there exists a joint configuration such that the robot is in static equilibrium).
@@
 
-In the remainder of this introduction, we discuss further the current state of the art. This allows us to situate more precisely our contributions.
+%~ In the remainder of this introduction, we discuss further the current state of the art. This allows us to situate more precisely our contributions.
 

### 2016-12-08 body/intro.tex
ongoin

@@
 
-To make a legged robot walk on a flat ground, efficient approaches based on the capture point(PRATT) can be used to determine the contact location. Choosing the effector
+To make a legged robot walk on a flat ground, efficient approaches based on the capture point~\citep{Pratt2006} can be used to determine the contact location. Choosing the effector
 to create a contact with is also trivial because walk follows a cyclic pattern.
-Unfortunately, in the general case planning complex contact interactions is extremely challenging.
-At any given time a contact choice must be made between infinitely many possibilities (often a combinatorial discrete choice for the effector, but an infinite choice of locations).
+Unfortunately, in the general case planning complex contact interactions is extremely challenging because the determinism disappears.
+At any given time a contact choice must be made between infinitely many possibilities (often a combinatorial discrete choice for the effector, and an infinite choice for the location).
 Furthermore, the contact manifold is foliated~\citep{simeon-manipulation-04}: a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this decision brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles.
-Lastly, the contact manifold has a null measure and is thus impossible to randomly sample because it has a zero measure with respect to the configuration space.
+Lastly, the contact manifold has a null measure and is thus impossible to randomly sample.
 
@@
 %~ \end{enumerate}
-Most locomotion systems avoid these issues thanks to simplifying assumptions that reduce the complexity (coplanar contacts, cyclic patterns that are not combinatorial)~\citep{Kajita03a}.
-At the DARPA challenge, instead of using the environment for creating contacts that facilitate motion, the chosen strategies to locomotion planning avoided obstacles as much as possible~\citep{atkensondarpa}.
+%~ Most locomotion systems avoid these issues thanks to simplifying assumptions that reduce the complexity (coplanar contacts, cyclic patterns that are not combinatorial)~\citep{Kajita03a}.
+For this reason at the DARPA challenge, instead of using the environment for creating contacts that facilitate motion, the chosen strategies fell back to walking patterns, and avoided obstacles as much as possible~\citep{atkensondarpa}, resulting in unsafe behaviors.
 Previous contributions having demonstrated acyclic contact locomotion have not been used because they are either too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585} or dynamically inconsistent~\citep{Mordatch:2012:DCB:2185520.2185539}.
 
-As suggested by \cite{Bouyarmane2009}, contrary to observation 5 we believe trajectory and contact planning can be decoupled.
-We go further and claim that this can be done at a much smaller cost, provided we can formulate a computationally-efficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} of a root configuration (i.e.
+In this paper, we propose a pragmatic approach to tackle the complexity of the acyclic contact planning problem. Rather than planning simultaneously in the configuration space the 
+whole-body trajectory and the contact location and timings, we decouple path and contact planning, as suggested by \citeauthor{Bouyarmane2009}.
+We first plan a path for the geometric root of the robot in a low dimensional space. We formulate this space as a computationally-efficient approximation
+of the \textit{equilibrium feasibile} space, that is the space of root configurations for which there exists a joint configuration such that the robot is in static equilibrium.
+We then use a deterministic algorithm 
+%~ We introduce a computationally-efficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} of a root configuration (i.e.
+there exists a joint configuration such that the robot is in static equilibrium).
+%~ We claim that this can be done at a much smaller cost, provided we can formulate a computationally-efficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} of a root configuration (i.e.
 there exists a joint configuration such that the robot is in static equilibrium).
 This paper presents a geometrical approximation of this condition, and a concrete implementation of such a decoupled planner.
-This approximation results from a trade-off between a necessary and a sufficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}, allowing us to find solutions extremely rapidly, while preserving a high success rate in the demonstrated scenarios.
+This approximation results from a trade-off between a necessary and a sufficient condition for \textit{equilibrium feasibility}, allowing us to find solutions extremely rapidly, while preserving a high success rate in the demonstrated scenarios.
 

### 2016-12-08 code/data/config/conf_hyq_hole.py

@@ -12,18 +12,16 @@ SAVE_DATA                   = True;
 SMOOTH_FILTER_WINDOW_LENGTH = 41;
 MAX_TEST_DURATION           = -1;
 dt                          = 1e-3;
-#INPUT_FILE_NAME             = '../data/2016_ijrr_traj/slide_compressed';
-#INPUT_FILE_NAME             = '../data/2016_ijrr_traj/ijrr/ground_0dl_window1_compressed';
-INPUT_FILE_NAME             = '../data/2016_ijrr_traj/ijrr/test_ground_crouch_conservative_cones/ground_0dl_window1_compressed';
-model_path                  = ["/home/adelpret/repos/20151101_contact_planner_steve/code/data/hyq"];
+INPUT_FILE_NAME             = '/home_local/dev/hpp/src/2016_ijrr_traj/ijrr/test_ground_crouch_conservative_cones/ground_0dl_window1_compressed';
+model_path                  = ["/home_local/dev/hpp/install/share"];
 urdfFileName                = model_path[0] + "/hyq_description/urdf/hyq.urdf";
 freeFlyer                   = True;
 
 ''' CONTROLLER CONFIGURATION '''
 ENABLE_CAPTURE_POINT_LIMITS     = False;
-ENABLE_TORQUE_LIMITS            = True;
-ENABLE_FORCE_LIMITS             = True;
-ENABLE_JOINT_LIMITS             = True;
+ENABLE_TORQUE_LIMITS            = True; # param
+ENABLE_FORCE_LIMITS             = True; # param
+ENABLE_JOINT_LIMITS             = True; # param, modulate with 4 next (position viability false, velocity ok)
 IMPOSE_POSITION_BOUNDS          = True;
 IMPOSE_VELOCITY_BOUNDS          = True;
 IMPOSE_VIABILITY_BOUNDS         = True;

### 2016-12-09 body/intro.tex
ongoing intro

--- 2016-09-08 body/intro.tex
@@
 
-Planning the locomotion of a multipedal robot in a complex environment - made of stairs, constraining obstacles or uneven surfaces - is still an open issue.
-Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}, but executing this behavior on constraining environments is dangerous if not impossible.
-In such a context, the robot has to perform a particular class of motion planning called acyclic contact planning, in which the robot must be in contact with the environment at every configuration to maintain the static equilibrium.
+%~ Planning legged robot locomotion remains an open issue.
+As humans do, a legged robot moves by sequentially creating contacts with the environment while avoiding collisions.
+Finding such a contact sequence is an instance of the motion planning problem called acyclic contact planning.
 
-The lack of solution to this planning problem has been highlighted by Atkeson \citeauthor{atkensondarpa}.
-Actually, the current approach to locomotion planning tends to avoid obstacles as much as possible, instead of using them by making contacts to facilitate motion.
-One main reason is that contact planning is algorithmically difficult, both because of the proximity to obstacles which makes tedious the sampling of collision-free configurations, and because of the foliation of the configuration space where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.
+Once the contact plan is known, efficient approaches can execute the plan and generate a dynamically feasible motion~\citep{Carpentier2016}. However in general
+the contact plan is manually designed after long trials and error.
+Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in human-centered environments, which are cluttered and comprise constraining obstacles such as stairs.
 
-Previous contributions that embrace the combinatorial provide complete approaches, not applicable in practice because they require hours of computations \citep{Bretl:2006:MPM:1124573.1124585}.
-More recent successes use a local solution, resulting in more reasonable computation times (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
-Neither global nor local methods present satisfying performances, because planning simultaneously the robot trajectory and the contacts that allow
-its execution is a very hard problem.
+To make a legged robot walk on a flat ground, efficient approaches based on the capture point~\citep{Pratt2006} can be used to choose the next contact location in a deterministic manner.
+Choosing the effector to create a contact with is trivial because walk follows a cyclic pattern.
+Unfortunately, in the general case planning complex contact interactions is extremely challenging because the determinism disappears.
+At any given time a contact choice must be made between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location).
+Furthermore, the contact manifold is foliated~\citep{simeon-manipulation-04}: a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this decision brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles.
+Lastly, the contact manifold has a null measure and is thus impossible to randomly sample.
+This restriction prevents the use of efficient sampling based planners to solve the problem.
 
-As suggested by \cite{Bouyarmane2009}, we believe that these two problems can be treated sequentially within a global planner.
-We go further and claim that this can be done at a much smaller cost, provided we can formulate a computationally-efficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} of the root configuration (i.e.
-there exists a joint configuration such that the robot is in static equilibrium).
-This paper presents a geometrical approximation of this condition, and a concrete implementation of such a decoupled planner.
-This approximation results from a trade-off between a necessary and a sufficient condition for \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}, allowing us to find solutions extremely rapidly, while preserving a high success rate in the demonstrated scenarios.
+For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
+Aa a result at the DARPA challenge, the robots fell back to unsafe walking patterns to climb stairs, instead of using the provided handrails to faciliate the motion~\citep{atkensondarpa}.
 
-In the remainder of this introduction, we discuss further the current state of the art. This allows us to situate more precisely our contributions.
+Reaching reasonable computation times is critical for a contact planner to be of any practical use.
+Our work thus aims at breaking the complexity of the contact planning problem.
+To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the non-determinism of the contact selection problem.
+First we introduce a low dimensional space, called the \textit{contact-feasible} space, that can be sampled and mapped efficiently to the contact manifold.
+Then, given a path computed in the \textit{contact-feasible} space, we propose a deterministic algorithm to generate a contact sequence along the path.
+This decoupling between the planning of the path and the generation of contacts presents drawbacks and advantages discussed in previous related literature.
+%~ whole-body trajectory and the contact location and timings, we decouple path and contact planning, as suggested by \citeauthor{Bouyarmane2009}.
+%~ We first plan a path for the geometric root of the robot in a low dimensional space. We formulate this space as a computationally-efficient approximation
+%~ of the \textit{equilibrium feasibile} space, that is the space of root configurations for which there exists a joint configuration such that the robot is in static equilibrium.
+%~ We then use a deterministic algorithm
+%~ This paper presents a geometrical approximation of this condition, and a concrete implementation of such a decoupled planner.
+%~ This approximation results from a trade-off between a necessary and a sufficient condition for \textit{equilibrium feasibility}, allowing us to find solutions extremely rapidly, while preserving a high success rate in the demonstrated scenarios.
+
+%~ In the remainder of this introduction, we discuss further the current state of the art. This allows us to situate more precisely our contributions.
 
@@
 This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments.
-Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.
+Following it, seve\-ral papers have applied this approach in specific situations, limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.
 
@@
 \textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence.
-A complete example of the potential offered by such approaches was proposed \citep{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
+A complete example of the potential offered by such approaches was proposed by \citeauthor{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
 To keep reasonable computation times, the method uses a simplified dynamic model for the avatar.
@@
 the contacts manually for more complex interactions.
-In addition to the practical limits of the current implementations, a major drawback of these optimization-based approaches is that they only offer local convergence when applied to acyclic contact planning.
+In addition to the limits of the current implementations, a major drawback of optimization-based approaches is that they only offer local convergence.
 
@@
 This approach is suboptimal because the quality of the motion depends on the quality of the guide path.
-\citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide path with guarantees of \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}, by extending key frames of the path into whole-body configurations in static equilibrium.
+\citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
 Randomly-sampled configurations are projected into the contact sub-manifold using a inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
@@
 \noindent \textbf{Regarding $\mathcal{P}_1$}, we define and use the \textit{reachability condition} to plan the guide path.
-The \textit{reachability condition}, a computationally-efficient approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
+The \textit{reachability condition} defines a low dimensional space that can be mapped to the contact manifoldcomputationally-efficient approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
 This step is fast because the planning happens
@@
 \item A very efficient and general implementation of an acyclic contact planner, the first one compatible with \gls{interactive} applications.
-\item Four heuristics for contact generation. They bias the planner towards configurations that are in robust static equilibrium, or 
-more efficient with respect to the task.
+%~ \item Four heuristics for contact generation. They bias the planner towards configurations that are in robust static equilibrium, or 
+%~ more efficient with respect to the task.
 \end{itemize}

### 2016-12-09 body/intro.tex
cleaning

@@
 %~ Planning legged robot locomotion remains an open issue.
-As humans do, a legged robot moves by sequentially creating contacts with the environment while avoiding collisions.
-Finding such a contact sequence is an instance of the motion planning problem called acyclic contact planning.
+Legged robots move by sequentially creating contacts with the environment while avoiding collisions.
+Finding the contact sequence is an instance of the motion planning problem called acyclic contact planning.
 
@@
 \textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence.
-A complete example of the potential offered by such approaches was proposed by \citeauthor{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
+A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
 To keep reasonable computation times, the method uses a simplified dynamic model for the avatar.

### 2016-12-09 code/data/config/conf_hyq_hole.py

@@ -12,7 +12,7 @@ SAVE_DATA                   = True;
 SMOOTH_FILTER_WINDOW_LENGTH = 41;
 MAX_TEST_DURATION           = -1;
 dt                          = 1e-3;
-INPUT_FILE_NAME             = '/home_local/dev/hpp/src/2016_ijrr_traj/ijrr/test_ground_crouch_conservative_cones/ground_0dl_window1_compressed';
+INPUT_FILE_NAME             = '/home_local/dev/hpp/src/hpp-rbprm-corba/script/scenarios/test35_compressed';
 model_path                  = ["/home_local/dev/hpp/install/share"];
 urdfFileName                = model_path[0] + "/hyq_description/urdf/hyq.urdf";
 freeFlyer                   = True;
@@ -59,7 +59,7 @@ verb=0;             # verbosity level (0, 1, or 2)
 USE_INPUT_CONTACT_POINTS = False;
 DEFAULT_CONTACT_POINTS  = np.matrix([0., 0., 0.]).T    # contact points in local reference frame
 DEFAULT_CONTACT_NORMALS = np.matrix([0., 0., 1.]).T    # contact normals in local reference frame
-mu  = np.array([0.33, 0.1]);          # force and moment friction coefficient
+mu  = np.array([0.6, 0.1]);          # force and moment friction coefficient
 fMin = 0.0;					     # minimum normal force
 
 # SIMULATOR PARAMETERS

### 2016-12-12 body/results.tex
Move Section about heuristics to Appendix

@@
 Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
-For each scenario we indicate the chosen heuristics. We also provide a performance analysis, which shows that the planner is compatible with \gls{interactive} applications,
+%For each scenario we indicate the chosen heuristics. 
+We also provide a performance analysis, which shows that the planner is compatible with \gls{interactive} applications,
 and present the success rates obtained in each scenario. Moreover, we demonstrate the interest of our robustness criterion in the different computed poses.
@@
 The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
-In each scenario we detail the parameters chosen: the heuristics, and the constraints on the reachable workspaces (for instance in all the scenarios,
+In each scenario we detail the parameters chosen: the heuristics (either $h_{\textrm{\it EFORT}}$, $h_{vel}$ or $h_{w}$, and optionally the equilibrium robustness, all of which are defined in the Appendix~\ref{sec:heuristics}), and the constraints on the reachable workspaces (for instance in all the scenarios,
 the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 

### 2016-12-12 body/overview.tex

@@
 Limiting our selection to this set of precomputed configurations allow us to drastically reduce the computation times. 
-Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), which we present in Section~\ref{sec:heuristics}.
+Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), which we present in the Appendix~\ref{sec:heuristics}.
 

### 2016-12-12 body/results.tex

@@
 Table~\ref{tab:requestpercent} summarizes the success rates obtained for each scenario.
-We observe that, as expected, our planner does not succeed systematically, because of the approximations made in our formulation.
+As expected, our planner does not succeed systematically, because of the approximations made in our formulation.
 The extreme situation of the car egress scenario provides the less satisfying results in terms of kinematic failure.
@@
 surprising considering the narrowness of the scene.
-From a pragmatic point of view, regarding the computation times, we claim that our approach provides a satisfying compromise between completeness and efficiency.
-Indeed, the advantage of the framework is that when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan a new contact sequence with a reasonable chance of success.
-The most efficient (and immediate) approach is probably to launch in parallel several instances of the planner for a given problem (our current implementation is single threaded) and to apply any successful result.
+\deladp{From a pragmatic point of view, }regarding \deladp{the} computation times, \deladp{we claim that} our approach provides a satisfying compromise between completeness and efficiency.
+Indeed, \deladp{the advantage of the framework is that} when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan \deladp{a new contact sequence} with a reasonable chance of success.
+The most efficient (and immediate) approach is probably to launch in parallel several instances of the planner \deladp{for a given problem} (our current implementation is single threaded) and to apply any successful result.
 

### 2016-12-12 body/intro.tex
rewrote intro

@@
 
-Reaching reasonable computation times is critical for a contact planner to be of any practical use.
-Our work thus aims at breaking the complexity of the contact planning problem.
+Obtaining reasonable computation times is critical for a contact planner to be of any practical use.
+Our work thus aims at breaking the complexity of the acyclic contact planning problem.
 To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the non-determinism of the contact selection problem.
@@
 The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of acyclic equilibrium configurations along the path.
-A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence.
+A third nontrivial problem, $\mathcal{P}_3$,
+%~ not addressed in this work, 
+then consists in interpolating a complete motion between two postures of the contact sequence.
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contacts and the potential paths.
@@
 
-Our solution belongs the class of decoupled approaches, because we believe that the separation between the generation of the root guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of multi-contact planning.
+Our solution belongs the class of decoupled approaches, because we believe that the separation between the generation of the root guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning.
 Compared to previous approaches, our solution has two main novelties:
 
-\noindent \textbf{Regarding $\mathcal{P}_1$}, we define and use the \textit{reachability condition} to plan the guide path.
-The \textit{reachability condition} defines a low dimensional space that can be mapped to the contact manifoldcomputationally-efficient approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
-This step is fast because the planning happens
-in a low-dimensional space, and does not require explicit contact computations.
+\noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm. The key to its efficiency is that it does not sample
+directly the contact manifold, but an approximation of the \textit{contact feasible} space, for which there exists a mapping to the contact manifold.
+The space is sampled thanks to the \textit{reachability condition} we introduce, a low-dimensional geometric condition for \textit{contact feasibility}.
 
-\noindent \textbf{Regarding $\mathcal{P}_2$},  we address \Pb by extending the path into a sequence of whole-body configurations in static equilibrium.
+\noindent \textbf{Regarding $\mathcal{P}_2$},  we propose a fast method to extend a \textit{contact feasible} path into a sequence of whole-body configurations in static equilibrium.
 This second step requires the explicit computation of contact configurations.
@@
 This sequential approach is the key to the efficiency of our method, though it
-can result in failures because our hypothesis is not always true.
+can result in failures:  we show that the \textit{reachability condition} only approximates the \textit{contact feasible} space.
 However, we demonstrate empirically the validity of the approach: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
@@
 \end{itemize}
+
+Finally, to demonstrate the validity of our approach regarding the complete chain, we show that the contact plans we produce can be successfully executed, either in simulation or on the real HRP-2 robot.
 

### 2016-12-13 body/intro.tex
[review] Reviewing introduction.

@@
 For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
-Aa a result at the DARPA challenge, the robots fell back to unsafe walking patterns to climb stairs, instead of using the provided handrails to faciliate the motion~\citep{atkensondarpa}.
+Aa a result at the DARPA Robotics Challenge, robots relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
 
@@
 Then, given a path computed in the \textit{contact-feasible} space, we propose a deterministic algorithm to generate a contact sequence along the path.
-This decoupling between the planning of the path and the generation of contacts presents drawbacks and advantages discussed in previous related literature.
+This decoupling \deladp{between the planning of the path and the generation of contacts} presents drawbacks and advantages discussed in previous related literature\addadp{, which we summarize in the following}.
 %~ whole-body trajectory and the contact location and timings, we decouple path and contact planning, as suggested by \citeauthor{Bouyarmane2009}.
@@
 
-The issue of planning acyclic contacts was first completely described by Bretl et al.
+\repadp{Acyclic contact planning}{The issue of planning acyclic contacts} was first completely described by Bretl et al.
 in their seminal paper  \citep{Bretl:2006:MPM:1124573.1124585}.
-The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of acyclic equilibrium configurations along the path.
+\repadp{This problem}{The issue} requires the simultaneous handling of two \addadp{sub}problems, $\mathcal{P}_1$: planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of \deladp{acyclic} equilibrium configurations along the path.
 A third nontrivial problem, $\mathcal{P}_3$,
@@
 Randomly-sampled configurations are projected into the contact sub-manifold using a inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
-Moreover this explicit projection is insufficient to guarantee the feasibility between two key positions in the path.
+Moreover this explicit projection is insufficient to guarantee the feasibility between two key \repadp{postures}{positions} in the path.
 \citeauthor{7140082} also propose a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with
@@
 use black-box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
-at \gls{interactive} frame rates (we say that a planning method is interactive when the computation time for one contact switch is less than the
-time to execute it). While this new approach provides an exciting direction of research, currently the resulting motions
+at \gls{interactive} frame rates (we say that a \deladp{planning} method is interactive when the computation time for \repadp{planning}{one contact switch} is less than the
+time \repadp{for execution}{to execute it}). While this new approach provides an exciting direction of research, currently the resulting motions
 look unnatural, and do not seem applicable to real robots.
 
-As far as robotics applications are concerned, none of the existing planners is able to solve the problem with \gls{interactive} performances.
+As far as robotics applications are concerned, none of the existing planners is \deladp{able to solve the problem with} \gls{interactive} \deladp{performances}.
 However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
-In particular, \citeauthor{Carpentier2016} are able to achieve this with \gls{interactive} performances on a real robot.
+In particular, \addadp{the algorithm of} \citeauthor{Carpentier2016} \repadp{is }{are able to achieve this with} \gls{interactive} \deladp{performances on a real robot}.
 Therefore, a planner capable of efficiently solving \Pa and \Pb could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
@@
 
-\subsection{Situation and Contributions}
+\subsection{\deladp{Situation and} Contributions}
 
@@
 
-Our solution belongs the class of decoupled approaches, because we believe that the separation between the generation of the root guide path and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning.
+Our solution belongs to the class of decoupled approaches, because we believe \repadp{this to be}{that the separation between the generation of the root guide path and the contact sequence is} the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning.
 Compared to previous approaches, our solution has two main novelties:
@@
 directly the contact manifold, but an approximation of the \textit{contact feasible} space, for which there exists a mapping to the contact manifold.
-The space is sampled thanks to the \textit{reachability condition} we introduce, a low-dimensional geometric condition for \textit{contact feasibility}.
+\adnote{It is not clear what the difference is between the contact manifold and the contact feasible space}
+The space is sampled thanks to the \textit{reachability condition}\deladp{ we introduce}, a low-dimensional geometric condition for \textit{contact feasibility}.
 
 \noindent \textbf{Regarding $\mathcal{P}_2$},  we propose a fast method to extend a \textit{contact feasible} path into a sequence of whole-body configurations in static equilibrium.
-This second step requires the explicit computation of contact configurations.
-It is guided by dedicated heuristics which fastly synthesize relevant configurations.
+This \deladp{second step} requires the explicit computation of contact configurations\repadp{, which we guide}{.
+It is guided} by dedicated heuristics that quickly synthesize \repadp{feasible}{relevant} configurations.
 
 This sequential approach is the key to the efficiency of our method, though it
-can result in failures:  we show that the \textit{reachability condition} only approximates the \textit{contact feasible} space.
+can result in failures\deladp{: we show that the \textit{reachability condition} only approximates the \textit{contact feasible} space}.
 However, we demonstrate empirically the validity of the approach: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
 
-As a result, we propose contributions to both problems \Pa and $\mathcal{P}_2$.
+\deladp{As a result, we propose contributions to both problems \Pa and $\mathcal{P}_2$.}
 \begin{itemize}
-\item A low-dimensional, efficient sampling-based approach to plan guide paths.
-\item A very efficient and general implementation of an acyclic contact planner, the first one compatible with \gls{interactive} applications.
+\item \deladp{A low-dimensional, efficient sampling-based approach to plan guide paths.}
+\item \deladp{A very efficient and general implementation of an acyclic contact planner, the first one compatible with \gls{interactive} applications.}
 %~ \item Four heuristics for contact generation. They bias the planner towards configurations that are in robust static equilibrium, or 
@@
 
-Finally, to demonstrate the validity of our approach regarding the complete chain, we show that the contact plans we produce can be successfully executed, either in simulation or on the real HRP-2 robot.
+Finally, to demonstrate the validity of our approach\deladp{ regarding the complete chain}, we show that the generated contact plans can be successfully executed, either in simulation or on the real HRP-2 robot.
 

### 2017-01-05 body/intro.tex
merged andreas comment + updating validation code

@@
 
-Finally, to demonstrate the validity of our approach\deladp{ regarding the complete chain}, we show that the generated contact plans can be successfully executed, either in simulation or on the real HRP-2 robot.
+Finally, to demonstrate the validity of our approach \deladp{ regarding the complete chain}, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
 

### 2017-01-05 code/data/config/conf_hyq_bridge.py

@@ -12,7 +12,7 @@ SAVE_DATA                   = True;
 SMOOTH_FILTER_WINDOW_LENGTH = 41;
 MAX_TEST_DURATION           = -1;
 dt                          = 1e-3;
-INPUT_FILE_NAME             = '/home_local/dev/hpp/src/2016_ijrr_traj/ground_crouch_ok/test34_compressed';
+INPUT_FILE_NAME             = '/home_local/dev/hpp/src/2016_ijrr_traj/bridge_ok/test88_compressed';
 model_path                  = ["/home_local/dev/hpp/install/share"];
 urdfFileName                = model_path[0] + "/hyq_description/urdf/hyq.urdf";
 sceneFileName               = model_path[0] + "/hpp-rbprm-corba/meshes/groundcrouch.stl";

### 2017-01-05 code/data/config/conf_hyq_darpa.py

@@ -12,10 +12,10 @@ SAVE_DATA                   = True;
 SMOOTH_FILTER_WINDOW_LENGTH = 41;
 MAX_TEST_DURATION           = -1;
 dt                          = 1e-3;
-INPUT_FILE_NAME             = '/home_local/dev/hpp/src/2016_ijrr_traj/ground_crouch_ok/test34_compressed';
+INPUT_FILE_NAME             = '/home_local/dev/hpp/src/hpp-rbprm-corba/script/scenarios/test55_compressed';
 model_path                  = ["/home_local/dev/hpp/install/share"];
 urdfFileName                = model_path[0] + "/hyq_description/urdf/hyq.urdf";
-sceneFileName               = model_path[0] + "/hpp-rbprm-corba/meshes/groundcrouch.stl";
+sceneFileName               = model_path[0] + "/hpp-rbprm-corba/meshes/darpareduced.stl";
 freeFlyer                   = True;
 
 ''' CONTROLLER CONFIGURATION '''
@@ -35,7 +35,7 @@ USE_JOINT_VELOCITY_ESTIMATOR    = False;
 ACCOUNT_FOR_ROTOR_INERTIAS      = True;
 
 # CONTROLLER GAINS
-kp_posture  = 15 #30.0; #1.0;   # proportional gain of postural task
+kp_posture  = 30 #15.0; #1.0;   # proportional gain of postural task
 kd_posture  = 2*sqrt(kp_posture);
 kp_constr   = 100.0;   # constraint proportional feedback gain
 kd_constr   = 2*sqrt(kp_constr);   # constraint derivative feedback gain

### 2017-01-05 code/python/track_locomotion_traj.py

@@ -10,6 +10,7 @@ if cwd+'/data/config' not in sys.path:
 
 #import test1_planar as conf
 import conf_hyq_hole as conf
+#~ import conf_hyq_bridge as conf
 
 from pinocchio_inv_dyn.standard_qp_solver import StandardQpSolver
 from pinocchio_inv_dyn.simulator import Simulator
@@ -414,9 +415,27 @@ final_time           = np.zeros(N_SOLVERS);
 final_time_step      = np.zeros(N_SOLVERS, np.int);
 controller_balance   = N_SOLVERS*[False,];
 
+''' Cosmetics '''
+gepgui = simulator.viewer.robot.viewer.gui
+simulator.viewer.addMesh("ground",conf.sceneFileName)
+gepgui.setColor('world/ground', [0.8,0.4,0.,1])
+gepgui.applyConfiguration("world/ground",[0,0,0.01,1,0,0,0]); 
+gepgui.setVisibility('world/floor', 'OFF')
+gepgui.refresh()
+
+
 for s in conf.SOLVER_TO_INTEGRATE:
     controller_balance[s] = startSimulation(q0, v0, s);
 #    cProfile.run('startSimulation(q0, v0, s);');
+
+
+gepgui.setVisibility('world/com', 'OFF')
+gepgui.setVisibility('world/cp', 'OFF')
+gepgui.setVisibility('world/lf_foot_joint', 'OFF')
+gepgui.setVisibility('world/rf_foot_joint', 'OFF')
+gepgui.setVisibility('world/lh_foot_joint', 'OFF')
+gepgui.setVisibility('world/rh_foot_joint', 'OFF')
+gepgui.refresh()
 
 if(conf.PLAY_MOTION_AT_THE_END):
     print "Gonna play computed motion";
@@ -477,3 +496,5 @@ if(conf.SHOW_FIGURES and conf.PLOT_JOINT_TRAJ):
         ax[0].plot(q_ref[7+j,:].A.squeeze(), 'r--');
         ax[0].set_title("Joint "+str(j));
         plt.show();
+        
+

### 2017-01-05 body/intro.tex
refreshing introduction

--- 2016-12-13 body/intro.tex
@@
 Legged robots move by sequentially creating contacts with the environment while avoiding collisions.
-Finding the contact sequence is an instance of the motion planning problem called acyclic contact planning.
+After years of research, such robots autonomously walk on flat ground, but struggle to achieve more complex motions.
+To climb stairs using a handrail for instance, a humanoid robot must plan the contacts to create with its feet and hands to perform the motion safely.
+The issue of computing this contact sequence is a motion planning problem called acyclic contact planning.
 
@@
 
-To make a legged robot walk on a flat ground, efficient approaches based on the capture point~\citep{Pratt2006} can be used to choose the next contact location in a deterministic manner.
-Choosing the effector to create a contact with is trivial because walk follows a cyclic pattern.
-Unfortunately, in the general case planning complex contact interactions is extremely challenging because the determinism disappears.
+To make a legged robot walk on a flat ground, efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
+Furthermore, choosing the effector to create a contact with is trivial because walk follows a cyclic pattern (For a humanoid the left foot always follows the right foot).
+Unfortunately, in the general case planning complex contact interactions is extremely challenging:
 At any given time a contact choice must be made between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location).
-Furthermore, the contact manifold is foliated~\citep{simeon-manipulation-04}: a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this decision brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles.
-Lastly, the contact manifold has a null measure and is thus impossible to randomly sample.
+Furthermore, a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles; we say that the contact manifold is foliated~\citep{simeon-manipulation-04}.
+Lastly, the contact manifold has a null measure and is thus impossible to sample.
 This restriction prevents the use of efficient sampling based planners to solve the problem.
@@
 For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
-Aa a result at the DARPA Robotics Challenge, robots relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
+As a result at the DARPA Robotics Challenge, robots relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
 
@@
 Our work thus aims at breaking the complexity of the acyclic contact planning problem.
-To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the non-determinism of the contact selection problem.
+To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorial of the contact selection problem.
 First we introduce a low dimensional space, called the \textit{contact-feasible} space, that can be sampled and mapped efficiently to the contact manifold.
 Then, given a path computed in the \textit{contact-feasible} space, we propose a deterministic algorithm to generate a contact sequence along the path.
-This decoupling \deladp{between the planning of the path and the generation of contacts} presents drawbacks and advantages discussed in previous related literature\addadp{, which we summarize in the following}.
+This decoupling presents drawbacks and advantages discussed in previous related literature, which we summarize in the following.
 %~ whole-body trajectory and the contact location and timings, we decouple path and contact planning, as suggested by \citeauthor{Bouyarmane2009}.
@@
 Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}.
-These approaches are intrinsically not able to adapt to new situations or discover complex behaviors in unforeseen contexts.
+These approaches are by definition not able to discover complex behaviors in unforeseen contexts.
 
-\repadp{Acyclic contact planning}{The issue of planning acyclic contacts} was first completely described by Bretl et al.
-in their seminal paper  \citep{Bretl:2006:MPM:1124573.1124585}.
-\repadp{This problem}{The issue} requires the simultaneous handling of two \addadp{sub}problems, $\mathcal{P}_1$: planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of \deladp{acyclic} equilibrium configurations along the path.
+The issue of planning acyclic contacts was first completely described by \citeauthor{Bretl:2006:MPM:1124573.1124585}.
+The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path.
 A third nontrivial problem, $\mathcal{P}_3$,
@@
 Two main directions have been explored.
-\textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence.
+\textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}.
 A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
 To keep reasonable computation times, the method uses a simplified dynamic model for the avatar.
-Still, the computation time is far from real-time  (about 1 minute of computation for a sequence of 20 contacts).
+Still, the computation time is far from real-time  (about 1 minute of computation for 20 contacts).
 A similar approach has been considered for manipulation by \cite{gabicciniisrr15}.
@@
 \citeauthor{dai2014whole} extend the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
-the contacts manually for more complex interactions.
-In addition to the limits of the current implementations, a major drawback of optimization-based approaches is that they only offer local convergence.
+the contacts manually for complex interactions.
+In addition to the limits of the current implementations, optimization-based approaches only converge locally.
 
@@
 \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
-Randomly-sampled configurations are projected into the contact sub-manifold using a inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
-Moreover this explicit projection is insufficient to guarantee the feasibility between two key \repadp{postures}{positions} in the path.
+Randomly-sampled configurations are projected to the contact manifold using a inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
+Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path.
 \citeauthor{7140082} also propose a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with
@@
 use black-box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
-at \gls{interactive} frame rates (we say that a \deladp{planning} method is interactive when the computation time for \repadp{planning}{one contact switch} is less than the
-time \repadp{for execution}{to execute it}). While this new approach provides an exciting direction of research, currently the resulting motions
+at \gls{interactive} frame rates (we say that a method is interactive when the computation time for planning is less than the
+time for execution). While this new approach provides an exciting direction of research, currently the resulting motions
 look unnatural, and do not seem applicable to real robots.
 
-As far as robotics applications are concerned, none of the existing planners is \deladp{able to solve the problem with} \gls{interactive} \deladp{performances}.
+As far as robotics applications are concerned, none of the existing planners is \gls{interactive}.
 However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
-In particular, \addadp{the algorithm of} \citeauthor{Carpentier2016} \repadp{is }{are able to achieve this with} \gls{interactive} \deladp{performances on a real robot}.
+In particular, our algorithm proposed in \citeauthor{Carpentier2016} is \gls{interactive}.
 Therefore, a planner capable of efficiently solving \Pa and \Pb could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
@@
 
-\subsection{\deladp{Situation and} Contributions}
+\subsection{Contributions}
 
@@
 
-Our solution belongs to the class of decoupled approaches, because we believe \repadp{this to be}{that the separation between the generation of the root guide path and the contact sequence is} the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning.
+Our solution belongs to the class of decoupled approaches, because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning.
 Compared to previous approaches, our solution has two main novelties:
@@
 \noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm. The key to its efficiency is that it does not sample
-directly the contact manifold, but an approximation of the \textit{contact feasible} space, for which there exists a mapping to the contact manifold.
-\adnote{It is not clear what the difference is between the contact manifold and the contact feasible space}
-The space is sampled thanks to the \textit{reachability condition}\deladp{ we introduce}, a low-dimensional geometric condition for \textit{contact feasibility}.
+directly the contact manifold, but an approximation of the \textit{contact feasible} space.
+The \textit{contact feasible} space is a low dimensional space for which there exists a mapping to the contact manifold.
+The space is sampled thanks to the \textit{reachability condition}, a low-dimensional geometric condition for \textit{contact feasibility}.
 
 \noindent \textbf{Regarding $\mathcal{P}_2$},  we propose a fast method to extend a \textit{contact feasible} path into a sequence of whole-body configurations in static equilibrium.
-This \deladp{second step} requires the explicit computation of contact configurations\repadp{, which we guide}{.
-It is guided} by dedicated heuristics that quickly synthesize \repadp{feasible}{relevant} configurations.
+This  requires the explicit computation of contact configurations.
+It is guided by dedicated heuristics that quickly synthesize feasible configurations.
 
 This sequential approach is the key to the efficiency of our method, though it
-can result in failures\deladp{: we show that the \textit{reachability condition} only approximates the \textit{contact feasible} space}.
+can result in failures.
 However, we demonstrate empirically the validity of the approach: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
@@
 
-Finally, to demonstrate the validity of our approach\deladp{ regarding the complete chain}, we show that the generated contact plans can be successfully executed, either in simulation or on the real HRP-2 robot.
+Finally, to demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
 

### 2017-01-09 body/appendix_roms.tex
final version !
replaced contact feasible with contact reachable to match with reachability confition

--- 2016-01-25 body/appendix_roms.tex
@@
+\clearpage
 \section{Generating the $W$ volumes for HRP-2}
@@
 Regarding the procedure, we can see that step 2 is conservative (Figure~\ref{fig:hrp2_roms}--right), which 
-is acceptable, especially because the lost set essentially relates to configurations close to singularity (they are close to the boundaries of the reachable workspace).
-We choose again to be less complete
-but more efficient, regarding the number of collision tests to be performed by RB-RRT.
+is acceptable, especially because the lost set essentially relates to configurations close to singularity (they are close to the boundaries of the reachable workspace, and
+often not \gls{contact reachable}, as illustrated in Figure~\ref{fig:dedefeas}, where the exterior boundaries of the reachable workspace appear
+red, thus not belonging to \gls{ccontact0}). We choose again to be less complete but more efficient, regarding the number of collision tests to be performed by RB-RRT.
 In step 1 on the other hand, selecting the convex hull (Figure~\ref{fig:hrp2_roms}--middle) instead of a minimum encompassing shape (Figure~\ref{fig:hrp2_roms}--left) may introduce false positives.

### 2017-01-09 body/glossary.tex
replaced contact feasible with contact reachable to match with reachability confition

@@
   name=$C_{Contact}^0$,
-  description={Set of \textit{contact feasible} root configurations $\mathbf{q}^{0}$, that verify $\exists \mathbf{q}^{\overline{0}},  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact}$}
+  description={Set of \textit{contact reachable} root configurations $\mathbf{q}^{0}$, that verify $\exists \mathbf{q}^{\overline{0}},  \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{contact}$}
 }
 
-\newglossaryentry{contact feasible}
+\newglossaryentry{contact reachable}
 {
-  name=\textit{contact feasible},
+  name=\textit{contact reachable},
   description={Refers to a root configuration that can be extended into a collision free contact configuration}

### 2017-01-09 body/intro.tex
replaced contact feasible with contact reachable to match with reachability confition

@@
 \noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm. The key to its efficiency is that it does not sample
-directly the contact manifold, but an approximation of the \textit{contact feasible} space.
-The \textit{contact feasible} space is a low dimensional space for which there exists a mapping to the contact manifold.
-The space is sampled thanks to the \textit{reachability condition}, a low-dimensional geometric condition for \textit{contact feasibility}.
+directly the contact manifold, but an approximation of the \textit{contact reachable} space.
+The \textit{contact reachable} space is a low dimensional space for which there exists a mapping to the contact manifold.
+The space is sampled thanks to the \textit{reachability condition}, a low-dimensional geometric condition for \textit{contact reachability}.
 
-\noindent \textbf{Regarding $\mathcal{P}_2$},  we propose a fast method to extend a \textit{contact feasible} path into a sequence of whole-body configurations in static equilibrium.
+\noindent \textbf{Regarding $\mathcal{P}_2$},  we propose a fast method to extend a \textit{contact reachable} path into a sequence of whole-body configurations in static equilibrium.
 This  requires the explicit computation of contact configurations.
@@
 
-\deladp{As a result, we propose contributions to both problems \Pa and $\mathcal{P}_2$.}
-\begin{itemize}
-\item \deladp{A low-dimensional, efficient sampling-based approach to plan guide paths.}
-\item \deladp{A very efficient and general implementation of an acyclic contact planner, the first one compatible with \gls{interactive} applications.}
-%~ \item Four heuristics for contact generation. They bias the planner towards configurations that are in robust static equilibrium, or 
-%~ more efficient with respect to the task.
-\end{itemize}
-
 Finally, to demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
+For HRP-2, we detail the complete computation times we obtained to address sequentially the three problems, and compare them with respect to related work, thus clearly demonstrating that our method is order of magnitude faster than any existing method.
 

### 2017-01-09 body/overview.tex
replaced contact feasible with contact reachable to match with reachability confition

@@
 
-%~ \begin{table*}
-%~ \centering
-%~ \begin{tabular}{ c | >{\arraybackslash}m{200pt} | c}
-   %~ Robot configuration set &  \centering Definition: $\mathbf{q} \in$ set $\Rightarrow$?  & Root configuration set\\
- %~ \hline
- %~ \hline
-   %~ $C_{contact}$ & $\mathbf{q}$ collision free; at least one end effector in contact. &  $\mathbf{q}^{0} \in C_{contact}^0$ is ``contact feasible''.  \\
- %~ \hline
-   %~ $C_{Equil} \subset C_{contact} $ & $\mathbf{q}$  in static equilibrium &  $\mathbf{q}^{0} \in C_{Equil}^0$ is ``equilibrium feasible''. \\
- %~ \end{tabular}
-%~ \caption{TODO.}
-%~ \label{tab:glossary}
-%~ \quad
-%~ \end{table*}
-
 Figure~\ref{fig:framework} illustrates our workflow.
@@
 extending it into a sequence of configurations in static equilibrium.
-In this Section we describe this workflow and give some notations used throughout the paper. Some key concepts are introduced,
-whose definitions can be found in the glossary at the end of this paper.
+In this Section we describe this workflow and give some notations used throughout the paper. 
+\delst{Some key concepts are introduced,
+whose definitions can be found in the glossary at the end of this paper.}
 
@@
 \subsection{Computation of a guide path --- \Pa}
-We first consider the problem of planning a root guide path. 
+We first consider the problem of planning a root guide path. All root configurations 
+As done by \citeauthor{Bouyarmane2009}, the aim is that 
 %The objective is to compute a path of root configurations that will allow for equilibrium. 
@@
 \end{description}
-C1 and C3 define \glslink{contact feasible}{\textit{contact feasibility}}, which is easier to verify than \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} because it does not require the computation of contact forces.
-We make the assumption that for the \textit{\gls{cluttered} contact planning problem}, most of the times \gls{contact feasible} root configurations are also \gls{equilibrium feasible}.
+C1 and C3 define \glslink{contact reachable}{\textit{contact reachability}}, which is easier to verify than \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} because it does not require the computation of contact forces.
+We make the assumption that for the \textit{\gls{cluttered} contact planning problem}, most of the times \gls{contact reachable} root configurations are also \gls{equilibrium feasible}.
 For this reason we neglect C2 during the resolution of $\mathcal{P}_1$, and we consider it only when solving \Pb.
-However, even verifying \glslink{contact feasible}{\textit{contact feasibility}} is computationally too expensive because it requires the projection of the whole-body configuration into a contact posture.
+However, even verifying \glslink{contact reachable}{\textit{contact reachability}} is computationally too expensive because it requires the projection of the whole-body configuration into a contact posture.
 
-An intuitive description of \gls{contact feasible} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
-We approximate \glslink{contact feasible}{\textit{contact feasibility}} with the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
+An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
+We approximate \glslink{contact reachable}{\textit{contact reachability}} with the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
 More precisely, a root configuration is \textit{reachable} if the root scaled by a factor $s \geq 1$ is not in collision, while the reachable workspace is in collision with the environment.
@@
 
-Figure~\ref{fig:dedefeas} gives an insight into the difference between the three conditions, depicting $C_{Equil}^0$, $C_{Contact}^0$ and $C_{Reach}^0$, which are the sets of \gls{equilibrium feasible}, \gls{contact feasible} and \textit{reachable} root configurations, respectively.
+Figure~\ref{fig:dedefeas} gives an insight into the difference between the three conditions, depicting $C_{Equil}^0$, $C_{Contact}^0$ and $C_{Reach}^0$, which are the sets of \gls{equilibrium feasible}, \gls{contact reachable} and \textit{reachable} root configurations, respectively.
 
-In the remainder of this paper, we use the terms \gls{contact feasible} and \gls{equilibrium feasible} to qualify either a root configuration, a whole-body configuration, or a set
+In the remainder of this paper, we use the terms \gls{contact reachable} and \gls{equilibrium feasible} to qualify either a root configuration, a whole-body configuration, or a set
 of such configurations.

### 2017-01-09 body/rbprm.tex
replaced contact feasible with contact reachable to match with reachability confition

@@
 
-Again, we assume that, for \gls{cluttered} problems, most of the times \contactfeasibility\, implies \equilibriumfeasibility.
-Under this assumption, our goal is to find a \gls{contact feasible} guide path.
+Again, we assume that, for \gls{cluttered} problems, most of the times \contactreachability\, implies \equilibriumfeasibility.
+Under this assumption, our goal is to find a \gls{contact reachable} guide path.
 
-To generate such a path efficiently, ideally we need to only sample \gls{contact feasible} root configurations.
-This requires exhibiting a necessary and sufficient condition for \contactfeasibility. %,that does not need the computation of the whole-body configuration.
-By default, verifying \contactfeasibility\, implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$. This is the approach chosen by \cite{Bouyarmane2009},
+To generate such a path efficiently, ideally we need to only sample \gls{contact reachable} root configurations.
+This requires exhibiting a necessary and sufficient condition for \contactreachability. %,that does not need the computation of the whole-body configuration.
+By default, verifying \contactreachability\, implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$. This is the approach chosen by \cite{Bouyarmane2009},
 which is too computationally expensive.
@@
 the search space: they provide an outer approximation of \gls{ccontact0}.
-On the other hand, only-sufficient conditions provide the guarantee that any configuration that satisfies them is indeed \gls{contact feasible}:
+On the other hand, only-sufficient conditions provide the guarantee that any configuration that satisfies them is indeed \gls{contact reachable}:
 they provide an inner approximation of \gls{ccontact0}.
@@
 
-\subsection{Conditions for contact feasibility}
-%~ By default, \glslink{contact feasible}{\textit{contact feasibility}} implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$.
+\subsection{Conditions for contact reachability}
+%~ By default, \glslink{contact reachable}{\textit{contact reachability}} implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$.
 This is the approach chosen by \cite{Bouyarmane2009}.
 However, computing $\mathbf{q}^{\overline{0}}$ is expensive in terms of computation time.
-In this section we rather define a necessary condition and a sufficient condition for \glslink{contact feasible}{\textit{contact feasibility}} that do not require this explicit computation.
+In this section we rather define a necessary condition and a sufficient condition for \glslink{contact reachable}{\textit{contact reachability}} that do not require this explicit computation.
 
-\subsubsection*{Contact feasibility, a necessary condition:}
+\subsubsection*{contact reachability, a necessary condition:}
 For a contact to be possible, a volume $O_i \in O$ necessarily intersects with the reachable workspace $W(\mathbf{q}^{0})$ (Figure~\ref{fig:contact_gen}--1).
-Furthermore, if $\mathbf{q}^{0}$ is \gls{contact feasible}, then the torso of the robot $W^0(\mathbf{q}^{0})$ is necessarily not colliding  with the environment $O$.
+Furthermore, if $\mathbf{q}^{0}$ is \gls{contact reachable}, then the torso of the robot $W^0(\mathbf{q}^{0})$ is necessarily not colliding  with the environment $O$.
 
@@
 
-\subsubsection*{Contact feasibility, a sufficient condition:}
+\subsubsection*{contact reachability, a sufficient condition:}
 %~ In building $C_{reach}^1$, we directly considered the including hull of the root body $W^{0}$, obtaining a necessary condition. 
-A trivial sufficient condition for \glslink{contact feasible}{\textit{contact feasibility}} can be constructed as a variation of $\mathcal{C}^0_{\textrm{\it Nec}}$, by replacing $W^0$ with a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
+A trivial sufficient condition for \glslink{contact reachable}{\textit{contact reachability}} can be constructed as a variation of $\mathcal{C}^0_{\textrm{\it Nec}}$, by replacing $W^0$ with a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
 We denote by \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ \gls{ccontact0}} the set of root configurations corresponding to this sufficient condition:
@@
 
-\subsection{Contact feasibility: a compromise \gls{reachc}}
+\subsection{contact reachability: a compromise \gls{reachc}}
 The sufficient condition is not interesting in practice since it leads the solver to lose too many interesting paths.
-The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \gls{contact feasible}.
+The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \gls{contact reachable}.
 An ideal shape $B$ (with $W^0 \subset B \subset B^{\textrm{\it Suf}}$) that leads to a necessary and sufficient condition may exist---even if it seems intuitively very unlikely in general. 
@@
 By increasing $s$, the condition can become sufficient.
-%~ The higher $s$, the closer this \gls{reachc} is to being sufficient, and if $s=1$, the search is complete (in the sense that all \gls{contact feasible} configurations can be found). 
+%~ The higher $s$, the closer this \gls{reachc} is to being sufficient, and if $s=1$, the search is complete (in the sense that all \gls{contact reachable} configurations can be found). 
 The parametrization of $s$ then defines a trade-off between these two interesting extremes. 
@@
 
-Thanks to these modifications, the problem of planning a \gls{contact feasible} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
+Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
 By doing this we can solve the problem with a sampling-based approach, in \gls{interactive} computation times.

### 2017-01-09 body/conclusion.tex
updated overview

@@
- \section{Discussion and future work} 
+ \section{Discussion} 
+\label{sec:discussion}
+
+\subsection{validity and interest of the approximations}
+
+TODO equilibirum as contact reachability
+
+\begin{figure}[t]
+\centering
+  \begin{overpic}[width=1\linewidth]{figures/2D_feas}
+		\put (15,) {$C_{Equil}^0$      $\subset$} 
+		\put (47,) {$C_{Contact}^0$ $\approx$ } 
+		\put (76,) {$C_{Reach}^0$} 
+	\end{overpic}
+\caption{Illustration of several root configurations sets used in this paper in a 2D scene.
+Obstacles are violet, and units are in meters.
+To show the sets in a 2D representation, all the rotational joints of HRP-2 are locked in the shown configuration, such that a torso configuration
+is only described by two positional parameters (x and y).
+The root of the robot is indicated with a black cross.
+To compute the reachable workspace, the point on the ankle indicated by a green cross was used.
+$C_{Equil}^0$ is included in $C_{Contact}^0$.
+$C_{Reach}^0$ approximates $C_{Contact}^0$.
+Depending on a parametrization, we can obtain $C_{Contact}^0 \subset C_{Reach}^0$.
+Considering the configurations around the top obstacle, we can observe a dramatic
+divergence between  $C_{Equil}^0$  and $C_{Contact}^0$ when the problem is not \textit{\gls{cluttered}}.}
+		   \label{fig:dedefeas}
+\end{figure}
+
+Figure~\ref{fig:dedefeas} gives an insight into the difference between the three conditions, depicting $C_{Equil}^0$, $C_{Contact}^0$ and $C_{Reach}^0$, which are the sets of \gls{equilibrium feasible}, \gls{contact reachable} and \textit{reachable} root configurations, respectively.
+
+TODO blah regarding contact plan generation, and ignoring the mid contact conf
+The first constraint is less restrictive than the one proposed by \citeauthor{DBLP:conf/iser/EscandeKMG08}, because our method
+allows to break and create one contact
+
+%~ \subsection{feasibility }
+
+\section{Conclusion} 
 \label{sec:conclusion}

### 2017-01-09 body/intro.tex
updated overview

@@
 can result in failures.
-However, we demonstrate empirically the validity of the approach: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
+However, we demonstrate empirically its interest: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
 
-Finally, to demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
+Finally, to further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
 For HRP-2, we detail the complete computation times we obtained to address sequentially the three problems, and compare them with respect to related work, thus clearly demonstrating that our method is order of magnitude faster than any existing method.

### 2017-01-09 body/overview.tex
updated overview

@@
 In this Section we describe this workflow and give some notations used throughout the paper.
-\delst{Some key concepts are introduced,
-whose definitions can be found in the glossary at the end of this paper.}
 
@@
 \subsection{Computation of a guide path --- \Pa}
-We first consider the problem of planning a root guide path. All root configurations 
-As done by \citeauthor{Bouyarmane2009}, the aim is that 
-%The objective is to compute a path of root configurations that will allow for equilibrium. 
-%A desired requirement is to preserve the completeness of the planner: it should be able to explore any valid guide path, but at the same time, any computed guide path must lead to a sequence of configurations in static equilibrium.
+We first consider the problem of planning a root guide path (Figure~\ref{fig:framework}--$\mathcal{P}_1$).
+The dimension of the path is equal to the number of degrees of freedom of the root of the robot.
+As for previous works~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: the root configurations of the path must be extended into full-body static equilibrium configurations.
+Previous work verify \textit{equilibrium feasibility} by explicitly computing such a joint configuration.
+We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact rechability}, illustrated in the following.
+
+An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a) to allow contact creation; not too close, because the robot must avoid collision.
+\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).
+More precisely, a root configuration is \textit{contact reachable} if the root scaled by a factor $s \geq 1$ is not in collision, while the reachable workspace is in collision.
+%~ Verifying the \textit{reachability condition} is very efficient in practice.
+To plan a root path, we then use a RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
+This is detailed in Section~\ref{rbprm}.
+%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).
+
 \begin{figure}[t]
 \centering
-  \begin{overpic}[width=1\linewidth]{figures/2D_feas}
-		\put (15,) {$C_{Equil}^0$      $\subset$} 
-		\put (47,) {$C_{Contact}^0$ $\approx$ } 
-		\put (76,) {$C_{Reach}^0$} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
+  \begin{overpic}[width=0.8\linewidth]{figures/reachability_cond}
 	\end{overpic}
-\caption{Illustration of several root configurations sets used in this paper in a 2D scene.
-Obstacles are violet, and units are in meters.
-To show the sets in a 2D representation, all the rotational joints of HRP-2 are locked in the shown configuration, such that a torso configuration
-is only described by two positional parameters (x and y).
-The root of the robot is indicated with a black cross.
-To compute the reachable workspace, the point on the ankle indicated by a green cross was used.
-$C_{Equil}^0$ is included in $C_{Contact}^0$.
-$C_{Reach}^0$ approximates $C_{Contact}^0$.
-Depending on a parametrization, we can obtain $C_{Contact}^0 \subset C_{Reach}^0$.
-Considering the configurations around the top obstacle, we can observe a dramatic
-divergence between  $C_{Equil}^0$  and $C_{Contact}^0$ when the problem is not \textit{\gls{cluttered}}.}
-		   \label{fig:dedefeas}
+\caption{The reachability condition is verified by the right configuration: the trunk (red) is free of collisions while the limbs reachable workspace (green) intersect the environment.}
+		   \label{fig:reach_cond}
 \end{figure}
+
 %
-Ideally we would like to sample root configurations that are \gls{equilibrium feasible}.
+%~ Ideally we would like to sample root configurations that are \gls{equilibrium feasible}.
 A root configuration is \gls{equilibrium feasible} if and only if there exists at least one joint configuration such that:
-\begin{description}
-\item[C1]: the robot is not in collision with the environment, 
-%~ \item[C2] has at least one end-effector in contact with the environment, 
-\item[C2]: the robot is in static equilibrium. 
-\end{description}
-Verifying C2 is computationally expensive because it requires the projection of the whole-body configuration into a contact posture, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
-For this reason, we are interested in finding an approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} that is computationally efficient.
-A requirement for C2 is the condition:
-\begin{description}
-\item[C3]: at least one end-effector is in contact with the environment.
-\end{description}
-C1 and C3 define \glslink{contact reachable}{\textit{contact reachability}}, which is easier to verify than \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} because it does not require the computation of contact forces.
-We make the assumption that for the \textit{\gls{cluttered} contact planning problem}, most of the times \gls{contact reachable} root configurations are also \gls{equilibrium feasible}.
-For this reason we neglect C2 during the resolution of $\mathcal{P}_1$, and we consider it only when solving \Pb.
-However, even verifying \glslink{contact reachable}{\textit{contact reachability}} is computationally too expensive because it requires the projection of the whole-body configuration into a contact posture.
-
-An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
-We approximate \glslink{contact reachable}{\textit{contact reachability}} with the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
-More precisely, a root configuration is \textit{reachable} if the root scaled by a factor $s \geq 1$ is not in collision, while the reachable workspace is in collision with the environment.
-Verifying the \textit{reachability condition} is very efficient in practice.
-In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).
-
-Figure~\ref{fig:dedefeas} gives an insight into the difference between the three conditions, depicting $C_{Equil}^0$, $C_{Contact}^0$ and $C_{Reach}^0$, which are the sets of \gls{equilibrium feasible}, \gls{contact reachable} and \textit{reachable} root configurations, respectively.
+%~ \begin{description}
+%~ \item[C1]: the robot is not in collision with the environment, 
+%~ \item[C2]: the robot is in static equilibrium. 
+%~ \end{description}
+%~ Verifying C2 is computationally expensive because it requires the projection of the whole-body configuration into a contact posture, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
+%~ For this reason, we are interested in finding an approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} that is computationally efficient.
+%~ A requirement for C2 is the condition:
+%~ \begin{description}
+%~ \item[C3]: at least one end-effector is in contact with the environment.
+%~ \end{description}
+%~ C1 and C3 define \glslink{contact reachable}{\textit{contact reachability}}, which is easier to verify than \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} because it does not require the computation of contact forces.
+%~ We make the assumption that for the \textit{\gls{cluttered} contact planning problem}, most of the times \gls{contact reachable} root configurations are also \gls{equilibrium feasible}.
+%~ For this reason we neglect C2 during the resolution of $\mathcal{P}_1$, and we consider it only when solving \Pb.
+%~ However, even verifying \glslink{contact reachable}{\textit{contact reachability}} is computationally too expensive because it requires the projection of the whole-body configuration into a contact posture.
+%~ 
+%~ An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
+%~ We approximate \glslink{contact reachable}{\textit{contact reachability}} with the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
+%~ More precisely, a root configuration is \textit{reachable} if the root scaled by a factor $s \geq 1$ is not in collision, while the reachable workspace is in collision with the environment.
+%~ Verifying the \textit{reachability condition} is very efficient in practice.
+%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).
 
@@
 
-The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--\Pb).
-With a fixed root position, the dimensionality of the 
-contact generation problem is reduced to the number of degrees of freedom of the considered limb. 
-%In other words, we consider each limb as a manipulator attached to the root. 
-We select from a database of precomputed limb configurations (independent from the environment) the ones resulting in the end-effector being close to the environment.
-Limiting our selection to this set of precomputed configurations allow us to drastically reduce the computation times. 
-Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), which we present in the Appendix~\ref{sec:heuristics}.
+The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--$\mathcal{P}_2$).
+To achieve this the root path is first decomposed into a sequence of discrete root configurations, according to a user-defined discretization step.
+Each root configuration is then extended into a whole-body configuration, subject to three constraints:
+\begin{itemize} 
+\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
+\item Each configuration is in static equilibrium.
+\item Each configuration is collision-free.
+\end{itemize}
 
-From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root configuration, an inverse-kinematics solver 
-is used to maintain the contacts that were previously existing. Possibly, some of these contacts cannot be maintained (because of joint limits or collisions).
-Then the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
-The algorithm is designed so that only one contact can be created or broken between two successive configurations.
+The algorithm thus proceeds iteratively, starting from the whole-body initial configuration of the robot.
+
+The algorithm takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually.
+Candidate limb configurations are selected from a database (independent from the environment) those resulting in the end-effector being close to the environment.
+Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.
+
+The details of the contact sequence generation are given in Section~\ref{sec:contact}.
+
+%~ From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root configuration, an inverse-kinematics solver 
+%~ is used to maintain the contacts that were previously existing. Possibly, some of these contacts cannot be maintained (because of joint limits or collisions).
+%~ Then the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
+%~ The algorithm is designed so that only one contact can be created or broken between two successive configurations.
 While it does not provide guarantees that the interpolation between two configurations is achievable, it appears empirically as a reasonable heuristic.
-Details are presented in Section~\ref{sec:contact}.
 

### 2017-01-10 body/conclusion.tex
overview + part 4

@@
 \subsection{validity and interest of the approximations}
+
+Again, we assume that, for \gls{cluttered} problems, most of the times \contactreachability\, implies \equilibriumfeasibility.
+Under this assumption, our goal is to find a \gls{contact reachable} guide path.
+
+To generate such a path efficiently, ideally we need to only sample \gls{contact reachable} root configurations.
+This requires exhibiting a necessary and sufficient condition for \contactreachability. %,that does not need the computation of the whole-body configuration.
+By default, verifying \contactreachability\, implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$. This is the approach chosen by \cite{Bouyarmane2009},
+which is too computationally expensive.
+To formulate a cheaper condition, we may turn our attention towards either an only-necessary or an only-sufficient condition.
+
+Only-necessary conditions are appealing because they preserve the completeness of the search, while reducing 
+the search space: they provide an outer approximation of $C_{Contact}^0$.
+On the other hand, only-sufficient conditions provide the guarantee that any configuration that satisfies them is indeed \gls{contact reachable}:
+they provide an inner approximation of $C_{Contact}^0$.
+
+In practice, the only-necessary and only-sufficient conditions that we can provide are trivial and give rather inaccurate approximations of $C_{Contact}^0$.
+Therefore, we propose a compromise between them: the \textit{reachability condition}, which is computationally efficient
+and provides a rather accurate approximation of $C_{Contact}^0$.
 

### 2017-01-10 body/contact.tex
typos
nicolas comments
overview + part 4

@@
 \label{sec:contact}
-Our planner computes guide paths in \gls{creach0}, an approximation of \gls{ccequil0}.
-As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{ccequil0}.
+Our planner computes guide paths in $C_{Reach}^0$ , an approximation of \gls{$C_{Equil}^0$}.
+As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Equil}^0$}.
 If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section~\ref{sec:perf}.
@@
 
-\subsubsection{Maintaining a contact in the sequence.}
+\subsubsection{Maintaining a contact in the sequence:}
 
@@
 The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits.
-The solver is directly provided by the Hpp software.
+The solver is directly provided by the HPP software.
 
@@
 %\subsection{Generation of a contact configuration}  
-\subsubsection{Creating contacts.}
+\subsubsection{Creating contacts:}
 Contacts are created using three rules:

### 2017-01-10 body/overview.tex
overview + part 4

@@
 Figure~\ref{fig:framework} illustrates our workflow.
-\Pa and \Pb are addressed in a sequential fashion: we first plan a root guide path with the \textit{reachability condition}, before
+\Pa and \Pb are addressed sequentially: we first plan a root guide path with the \textit{reachability condition}, before
 extending it into a sequence of configurations in static equilibrium.
-In this Section we describe this workflow and give some notations used throughout the paper.
 
@@
 
-\subsection{Notation conventions and definitions} \label{notations}
+\section{Notation conventions and definitions} \label{notations}
 
@@
 \medskip
-\textbf{A robot} is a kinematic chain $R$, that we divide into different parts: a root $R^0$, and $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
+\textbf{A robot} is a kinematic chain $R$ composed of: a root $R^0$, and $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
 The root has $r \geq 6$ DOFS: for instance, HRP-2 has two extra DOFS in the torso, such that we have $r=8$.
 Therefore $R$ is fully described by a configuration $\mathbf{q} \in  \mathbb{R}^{r+n}$.
-We define some relevant projections of $\mathbf{q}$:
+$\mathbf{q}$ is decomposed as follows:
 \begin{itemize}
-	\item $\mathbf{q}^k$ denotes the configuration (a vector of joint values) of the limb $R^k$; % (Figure~\ref{fig:character});
-	\item $\mathbf{q}^{\overline{k}}$ denotes the vector of joint values of R \textbf{not} related to $R^k$.
+	\item $\mathbf{q}^k$ is a configuration (a vector of joint values) of the limb $R^k$; % (Figure~\ref{fig:character});
+	\item $\mathbf{q}^{\overline{k}}$ is a vector of joint values of R \textbf{not} related to $R^k$.
 We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
-	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ denotes the world coordinates of the root $R^0$.
+	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ is the world coordinates vector of $R^0$.
 \end{itemize}
@@
 \medskip
-The volume encompassing $R^0$ is denoted $W^0$ (Figure~\ref{fig:HyQ_roms}).
+\noindent $W^0$ is a volume encompassing $R^0$ (Figure~\ref{fig:HyQ_roms}). \\
  %~ (Figure~\ref{fig:character}-right: central cylinder). 
- The reachable workspace of a limb $R^k$ is denoted $W^k$: %(Figure~\ref{fig:character}-right: the four ellipses). 
+ $W^k$ is the reachable workspace of a limb $R^k$: %(Figure~\ref{fig:character}-right: the four ellipses). 
 \begin{equation}
@@
 \textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains.
+
+\medskip
+$C_{Contact}$ is the set of whole body-configurations in contact and collision-free.
+
+\medskip
+$C_{Equil} \subset C_{Contact}$ is the set of whole body-configurations in static equilibrium and collision-free.
+
+\medskip
+For any set $C_{X}$, we define $C_{X}^0$:
+\begin{equation*}
+  C_{X}^0 = \left\{ \mathbf{q}^{0},  \exists \mathbf{q}^{\overline{0}}: \mathbf{q}^0  \oplus \mathbf{q}^{\overline{0}} \in C_{X} \right\}
+\end{equation*}
+
 %~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.

### 2017-01-10 body/rbprm.tex
overview + part 4

@@
 % !TEX root =  ../main.tex
-\section{Computing a guide path in $C^0_{reach}$ ($\mathcal{P}_1$) }
+%~ \section{Computing a guide path in $C^0_{reach}$ ($\mathcal{P}_1$) }
+\section{Root path planning in the \textit{contact reachable space}}
 \label{rbprm}
 
-We consider the issue of computing an \gls{equilibrium feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{ccequil0} for the root of a multiped robot, connecting user-defined start and goal configurations.
+In the root path planning phase of our framework, we only consider the root configuration $\mathbf{q}^0$ defined in the previous Section,
+as well as the environment $O$.
 
-Again, we assume that, for \gls{cluttered} problems, most of the times \contactreachability\, implies \equilibriumfeasibility.
-Under this assumption, our goal is to find a \gls{contact reachable} guide path.
+Given a start and a goal configurations, we aim at computing a guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ $\mathbb{R}^r$ verifying:
+\begin{equation*} \label{eq:path}
+\forall t \in [0,1], \mathbf{q}^0(t)  \in C_{Equil}^0
+\end{equation*}
+This means that any root configuration must be extended into a whole-body, static equilibrium configuration.
+$C_{Equil}^0$  can't be described analytically, because the contact manifold has a zero measure.
 
-To generate such a path efficiently, ideally we need to only sample \gls{contact reachable} root configurations.
-This requires exhibiting a necessary and sufficient condition for \contactreachability. %,that does not need the computation of the whole-body configuration.
-By default, verifying \contactreachability\, implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$. This is the approach chosen by \cite{Bouyarmane2009},
-which is too computationally expensive.
-To formulate a cheaper condition, we may turn our attention towards either an only-necessary or an only-sufficient condition.
-
-Only-necessary conditions are appealing because they preserve the completeness of the search, while reducing 
-the search space: they provide an outer approximation of \gls{ccontact0}.
-On the other hand, only-sufficient conditions provide the guarantee that any configuration that satisfies them is indeed \gls{contact reachable}:
-they provide an inner approximation of \gls{ccontact0}.
-
-In practice, the only-necessary and only-sufficient conditions that we can provide are trivial and give rather inaccurate approximations of \gls{ccontact0}.
-Therefore, we propose a compromise between them: the \textit{reachability condition}, which is computationally efficient
-and provides a rather accurate approximation of \gls{ccontact0}. % for \gls{cluttered} problems.
-
-%~ As said in the previous section, the goal is to enforce that the guide can be discretized into a sequence of \gls{equilibrium feasible} configurations.
-We denote by \gls{ccequil} the equilibrium sub-manifold of the robot.
-
-%~ We say that a root configuration $\mathbf{q}^{0}$ is \gls{equilibrium feasible} if it belongs to the set:
-%~ \begin{equation}
-      %~ C_{Equil}^0= \{ \mathbf{q}^{0}: \exists \mathbf{q}^{\overline{0}}, \mathbf{q}^{0} \oplus \mathbf{q}^{\overline{0}} \in C_{Equil} \}
-    %~ \label{eq:pi}
-%~ \end{equation}
-
-%~ The set of all \textit{geometrically feasible} root configurations is denoted by $C_{reach}$.
-%~ By extension, a path $\mathbf{q}^0(t)$ is \gls{equilibrium feasible} if $\forall t \in [0,1], \mathbf{q}^0(t) \in C_{contact}^0$.
-
-%~ In this Section we characterize \gls{ccontact0} theoretically with a necessary and a sufficient condition for belonging to the set.
-%~ We then approximate \gls{ccontact0} as a space $C_{reach}$, where we can efficiently sample configurations.
-%~ We finally describe our implementation of an efficient RRT planner in $C_{reach}$.
-
-%~ We want all the configurations of $q^0(t)$ to belong to a set $C_{reach}$ defined as:
-%~ 
-%~ \begin{eqnarray}
-%~ C_{reach}= \{ \mathbf{q}^{0} & : & \exists f, f(\mathbf{q}^{0}) \in C_{contact} \}
-%~ \end{eqnarray}
+The strong hypothesis of this work is that for a large variety of locomotion tasks, we can define a space  $C_{Reach}^0 \simeq C_{Contact}^0$, such that 
+\begin{equation} \label{eq:creach}
+\forall t \in [0,1], \mathbf{q}^0(t) \in C_{Reach}^0 \Rightarrow \mathbf{q}^0(t)  \in C_{Equil}^0
+\end{equation}
+% for \gls{cluttered} problems.
+We call  $C_{Reach}^0$ the \textit{contact reachable workspace}, and detail its construction in the following.
+The validity of this hypothesis is discussed in depth in Section~\ref{sec:discussion}.
+%~ We describe the construction of $C_{Reach}^0$ in the following.
 
 \subsection{Conditions for contact reachability}
-%~ By default, \glslink{contact reachable}{\textit{contact reachability}} implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$.
-This is the approach chosen by \cite{Bouyarmane2009}.
-However, computing $\mathbf{q}^{\overline{0}}$ is expensive in terms of computation time.
-In this section we rather define a necessary condition and a sufficient condition for \glslink{contact reachable}{\textit{contact reachability}} that do not require this explicit computation.
+The contact reachable workspace is defined as a compromise between two necessary and a sufficient condition for contact creation.
 
-\subsubsection*{contact reachability, a necessary condition:}
-For a contact to be possible, a volume $O_i \in O$ necessarily intersects with the reachable workspace $W(\mathbf{q}^{0})$ (Figure~\ref{fig:contact_gen}--1).
-Furthermore, if $\mathbf{q}^{0}$ is \gls{contact reachable}, then the torso of the robot $W^0(\mathbf{q}^{0})$ is necessarily not colliding  with the environment $O$.
-
-Therefore we can define an outer approximation  $ \mathcal{C}^0_{\textrm{\it Nec}} \supset$ \gls{ccontact0} defined as: 
+\textbf{necessary conditions:}
+For a contact to be possible, a volume $O_i \in O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot (Figure~\ref{fig:contact_gen}--1).
+Also the torso of the robot $W^0(\mathbf{q}^{0})$ is necessarily collision-free.
+Therefore we can define an outer approximation  $ \mathcal{C}^0_{\textrm{\it Nec}} \supset$ $C_{Contact}^0$ defined as: 
 \begin{equation}
-\mathcal{C}^0_{\textrm{\it Nec}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+\mathcal{C}^0_{\textrm{\it Nec}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
 \end{equation}
-%~ The inclusion \gls{ccontact0} $\subset \mathcal{C}^0_{\textrm{\it Nec}}$ is straightforward, but it is very important.
+%~ The inclusion $C_{Contact}^0$ $\subset \mathcal{C}^0_{\textrm{\it Nec}}$ is straightforward, but it is very important.
 %~ Planning in $\mathcal{C}^0_{\textrm{\it Nec}}$ allows a strong reduction of the search space, while not discarding 
@@
 
-\subsubsection*{contact reachability, a sufficient condition:}
-%~ In building $C_{reach}^1$, we directly considered the including hull of the root body $W^{0}$, obtaining a necessary condition. 
-A trivial sufficient condition for \glslink{contact reachable}{\textit{contact reachability}} can be constructed as a variation of $\mathcal{C}^0_{\textrm{\it Nec}}$, by replacing $W^0$ with a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces to be in contact.
-We denote by \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ \gls{ccontact0}} the set of root configurations corresponding to this sufficient condition:
-
+\textbf{a sufficient condition:}
+Similarly we can define an inner approximation \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ $C_{Contact}^0$} by considering a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces in contact.
 \begin{equation}
-\mathcal{C}^0_{\textrm{\it Suf}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } B^{\textrm{\it Suf}}(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+\mathcal{C}^0_{\textrm{\it Suf}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge B^{\textrm{\it Suf}}(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
@@
 
-\subsection{contact reachability: a compromise \gls{reachc}}
-The sufficient condition is not interesting in practice since it leads the solver to lose too many interesting paths.
-The necessary condition is not perfect either, since the first stage of the planner would stop on a guide that is not \gls{contact reachable}.
-An ideal shape $B$ (with $W^0 \subset B \subset B^{\textrm{\it Suf}}$) that leads to a necessary and sufficient condition may exist---even if it seems intuitively very unlikely in general. 
-%~ Its construction is out of the scope of this work.
+\subsection{The compromising reachability condition}
+The ideal shape $B^*, W^0 \subset B* \subset B^{\textrm{\it Suf}}$ that defines a necessary \textbf{and} sufficient condition for contact creation has no analytical definition.
+Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.
 
-%~ It might be possible to find an ideal shape $B$ that leads to a necessary and sufficient condition; however, it seems intuitively very unlikely in general.
-The construction of a shape $W^0 \subset B \subset B^{\textrm{\it Suf}}$ leading to a necessary and sufficient condition (or the proof of its nonexistence) is out of the scope of this work.
-
-However, using a shape between $W^0$ and $B^{\textrm{\it Suf}}$ leads to a trade-off between a necessary and a sufficient condition.
+%~ However, using a shape between $W^0$ and $B^{\textrm{\it Suf}}$ leads to a trade-off between a necessary and a sufficient condition. 
 We define $W^0_s$ as the volume $W^0$ subject to a scaling transformation by a factor $s \in \mathbb{R}^+$.
@@
  \begin{equation}
-C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+ \label{eq:reachc}
+C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
@@
 %
-If $s=1$, then $W^0_s$ = $W^0$, such that $C_1^0$ = $\mathcal{C}^0_{\textrm{\it Nec}}$. We thus consider that $s \geq 1$, since smaller values would only worsen the approximation.
-By increasing $s$, the condition can become sufficient.
-%~ The higher $s$, the closer this \gls{reachc} is to being sufficient, and if $s=1$, the search is complete (in the sense that all \gls{contact reachable} configurations can be found). 
-The parametrization of $s$ then defines a trade-off between these two interesting extremes. 
-We can choose $s$ by hand, or automatically as explained in Section~\ref{sec:params}.
-The chosen value $s=s^*$ defines the \gls{reachc}, therefore we write \gls{creach0} $= C^0_{s^*}$.
-%~ We call this condition the \gls{reachc}, which is defined by the chosen value $s=s^*$, and define \gls{creach0} $= C^0_{s^*}$.
-%~ In the rest of the paper, we consider that an appropriate value $s^*$ of $s$ is chosen, and define \gls{creach0} $= C^0_{s^*}$.
-
-%~ As an example, Figure~\ref{fig:HyQ_roms} presents the volumes $W$ computed for the HyQ robot.
-In Appendix~\ref{app:rom}, we give a generic method to compute the volumes appearing in the definition of \gls{creach0}, with the example of HRP-2.
-%~ Section~\ref{sec:results} shows that in practice, it is easy to adjust $s$ to keep most of the interesting guides without introducing incorrect guides.
+The parametrization of $s$ defines a trade-off:
+If $s=1$, then $W^0_s$ = $W^0$, such that $C_1^0$ = $\mathcal{C}^0_{\textrm{\it Nec}}$.
+ %~ We thus consider that $s \geq 1$, since smaller values would only worsen the approximation.
+By increasing $s$, the condition can become sufficient, but less and less necessary.  
+Eq.~\ref{eq:reachc} thus defines the \textit{reachability condition}. We fix a value $s^*$ for $s$ and define  $C_{Reach}^0 = C^0_{s^*}$.
+The computation of $s^*$ is detailed in Section~\ref{sec:params}. 
+In Appendix~\ref{app:rom}, we give a generic method to compute the $W$ volumes appearing in the definition of $C_{Reach}^0$.
 
 \subsection{Computing the guide path in $C_{reach}^0$}
-Any sampling-based motion planner can be used to plan a path in \gls{creach0}. 
-Indeed, contrary to \gls{ccontact}, \gls{creach0} has a nonzero measure in the configuration space $C$. Therefore a standard uniform sampling approach
-can work, in spite of a high rejection rate. 
-Thus, the only significant change regarding a classical planner is to replace the collision checking with the \gls{reachc} when verifying
+$C_{Reach}^0$ can be sampled efficiently thanks to Eq.~\ref{eq:reachc}, and can thus be used with a standard motion planner.
+Thus, the only significant change regarding a classical planner is to replace the collision checking with the \textit{rechability condition} when verifying
 the drawn configurations and associated local paths.
@@
 Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
-By doing this we can solve the problem with a sampling-based approach, in \gls{interactive} computation times.
+%~ By doing this we can solve the problem with a sampling-based approach, in \gls{interactive} computation times.

### 2017-01-10 body/results.tex
overview + part 4

@@
 We then computed the rates of false positives (i.e. the configuration satisfies
-the reachability condition, but does not belong to \gls{ccequil0}) and false negatives (i.e. the configuration does not satisfy the 
-reachability condition, but belongs to \gls{ccequil0}).
+the reachability condition, but does not belong to \gls{$C_{Equil}^0$}) and false negatives (i.e. the configuration does not satisfy the 
+reachability condition, but belongs to \gls{$C_{Equil}^0$}).
 

### 2017-01-10 body/intro.tex

@@
 %~ Planning legged robot locomotion remains an open issue.
-Legged robots move by sequentially creating contacts with the environment while avoiding collisions.
-After years of research, such robots autonomously walk on flat ground, but struggle to achieve more complex motions.
-To climb stairs using a handrail for instance, a humanoid robot must plan the contacts to create with its feet and hands to perform the motion safely.
-The issue of computing this contact sequence is a motion planning problem called acyclic contact planning.
+Legged robots move by sequentially creating contacts with the environment\deladp{ while avoiding collisions}.
+After years of research, such robots \addadp{can} autonomously walk on flat ground, but struggle to \repadp{navigate more complex environments}{achieve more complex motions}.
+To climb stairs using a handrail for instance, a humanoid robot must plan the contacts to create with its feet and hands \deladp{to perform the motion safely}.
+\deladp{The issue of} computing \repadp{such a}{this} contact sequence is a\addadp{n open} motion planning problem called acyclic contact planning.
 
-Once the contact plan is known, efficient approaches can execute the plan and generate a dynamically feasible motion~\citep{Carpentier2016}. However in general
-the contact plan is manually designed after long trials and error.
+Once the contact plan is known, efficient approaches \repadp{exist to}{can execute the plan and} generate a dynamically feasible motion~\citep{Carpentier2016}. However in general
+the contact plan is manually designed \repadp{by a time consuming trial-and-error procedure}{after long trials and error}.
 Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in human-centered environments, which are cluttered and comprise constraining obstacles such as stairs.
 
-To make a legged robot walk on a flat ground, efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
-Furthermore, choosing the effector to create a contact with is trivial because walk follows a cyclic pattern (For a humanoid the left foot always follows the right foot).
+To make a legged robot walk on \deladp{a} flat ground, efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
+Furthermore, choosing the effector with which to create a contact is trivial because walk\addadp{ing} follows a cyclic pattern (For a humanoid the left foot always follows the right foot).
 Unfortunately, in the general case planning complex contact interactions is extremely challenging:
-At any given time a contact choice must be made between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location).
+At any given time a contact must be chosen between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location).
 Furthermore, a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles; we say that the contact manifold is foliated~\citep{simeon-manipulation-04}.
 Lastly, the contact manifold has a null measure and is thus impossible to sample.
-This restriction prevents the use of efficient sampling based planners to solve the problem.
+This restriction prevents the use of efficient sampling-based planners\deladp{ to solve the problem}.
 
@@
 To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorial of the contact selection problem.
-First we introduce a low dimensional space, called the \textit{contact-feasible} space, that can be sampled and mapped efficiently to the contact manifold.
+First we introduce a low-dimensional space, called the \textit{contact-feasible} space, that can be sampled and mapped efficiently to the contact manifold.
 Then, given a path computed in the \textit{contact-feasible} space, we propose a deterministic algorithm to generate a contact sequence along the path.
@@
 The issue of planning acyclic contacts was first completely described by \citeauthor{Bretl:2006:MPM:1124573.1124585}.
-The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path.
+The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a \deladp{relevant} guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path.
 A third nontrivial problem, $\mathcal{P}_3$,
@@
 A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
-To keep reasonable computation times, the method uses a simplified dynamic model for the avatar.
-Still, the computation time is far from real-time  (about 1 minute of computation for 20 contacts).
+To get reasonable computation times, the method uses a simplified dynamic model for the avatar.
+Still, the \repadp{method}{computation time} is far from real-time  (about 1 minute of computation for 20 contacts).
 A similar approach has been considered for manipulation by \cite{gabicciniisrr15}.
-\citeauthor{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered.
-\citeauthor{dai2014whole} extend the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
+\citeauthor{DBLP:conf/humanoids/DeitsT14} proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered.
+\citeauthor{dai2014whole} extended the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
 the contacts manually for complex interactions.
@@
 \textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity.
-The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
-an ad-hoc solution to $\mathcal{P}_1$).
-\Pb is addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
-A solution can then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning},
+The feasibility and interest of the decoupling has been shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
+an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
+A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning},
 but at the cost of prohibitive computation times (up to several hours) for constraining scenarios.
 This approach is suboptimal because the quality of the motion depends on the quality of the guide path.
-\citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
-Randomly-sampled configurations are projected to the contact manifold using a inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
+\citeauthor{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
+Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
 Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path.
-\citeauthor{7140082} also propose a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with
+\citeauthor{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with
 a discretized environment. This planning phase does not account for collisions, implying that replanning is required in case of failure. This approach is efficient 
@@
 For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
-use black-box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
+used black-box physics simulators to perform Model Predictive Control for \deladp{the motion of} a humanoid character, and managed to obtain dynamically-consistent motions
 at \gls{interactive} frame rates (we say that a method is interactive when the computation time for planning is less than the
@@
 directly the contact manifold, but an approximation of the \textit{contact reachable} space.
-The \textit{contact reachable} space is a low dimensional space for which there exists a mapping to the contact manifold.
-The space is sampled thanks to the \textit{reachability condition}, a low-dimensional geometric condition for \textit{contact reachability}.
+The \textit{contact reachable} space is a low-dimensional space for which there exists a mapping to the contact manifold.
+\deladp{The space is sampled thanks to the \textit{reachability condition}, a low-dimensional geometric condition for \textit{contact reachability}.}
 
@@
 Finally, to further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
-For HRP-2, we detail the complete computation times we obtained to address sequentially the three problems, and compare them with respect to related work, thus clearly demonstrating that our method is order of magnitude faster than any existing method.
+For HRP-2, we detail the complete computation times \deladp{we obtained} to address sequentially the three problems, and compare them \deladp{with respect} to related work\deladp{, thus clearly} demonstrating that our method is order of magnitude faster\deladp{ than any existing method}.
 

### 2017-01-10 body/intro.tex
accepted andrea s update on introduction

@@
 %~ Planning legged robot locomotion remains an open issue.
-Legged robots move by sequentially creating contacts with the environment\deladp{ while avoiding collisions}.
-After years of research, such robots \addadp{can} autonomously walk on flat ground, but struggle to \repadp{navigate more complex environments}{achieve more complex motions}.
-To climb stairs using a handrail for instance, a humanoid robot must plan the contacts to create with its feet and hands \deladp{to perform the motion safely}.
-\deladp{The issue of} computing \repadp{such a}{this} contact sequence is a\addadp{n open} motion planning problem called acyclic contact planning.
+Legged robots move by sequentially creating contacts with the environment.
+After years of research, such robots can autonomously walk on flat ground, but struggle to navigate more complex environments.
+To climb stairs using a handrail, a humanoid robot must plan the contacts to create with its feet and hands.
+Computing such a contact sequence is an open motion planning problem called acyclic contact planning.
 
-Once the contact plan is known, efficient approaches \repadp{exist to}{can execute the plan and} generate a dynamically feasible motion~\citep{Carpentier2016}. However in general
-the contact plan is manually designed \repadp{by a time consuming trial-and-error procedure}{after long trials and error}.
+Once the contact plan is known, efficient approaches exist to generate a dynamically feasible motion~\citep{Carpentier2016}. However in general
+the contact plan is manually designed by a time consuming trial-and-error procedure.
 Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in human-centered environments, which are cluttered and comprise constraining obstacles such as stairs.
 
-To make a legged robot walk on \deladp{a} flat ground, efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
-Furthermore, choosing the effector with which to create a contact is trivial because walk\addadp{ing} follows a cyclic pattern (For a humanoid the left foot always follows the right foot).
+To make a legged robot walk on flat ground, efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
+Furthermore, choosing the effector with which to create a contact is trivial because walking follows a cyclic pattern (For a humanoid the left foot always follows the right foot).
 Unfortunately, in the general case planning complex contact interactions is extremely challenging:
@@
 Lastly, the contact manifold has a null measure and is thus impossible to sample.
-This restriction prevents the use of efficient sampling-based planners\deladp{ to solve the problem}.
+This restriction prevents the use of efficient sampling-based planners.
 
@@
 The issue of planning acyclic contacts was first completely described by \citeauthor{Bretl:2006:MPM:1124573.1124585}.
-The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a \deladp{relevant} guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path.
+The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path.
 A third nontrivial problem, $\mathcal{P}_3$,
@@
 To get reasonable computation times, the method uses a simplified dynamic model for the avatar.
-Still, the \repadp{method}{computation time} is far from real-time  (about 1 minute of computation for 20 contacts).
+Still, the method is far from real-time  (about 1 minute of computation for 20 contacts).
 A similar approach has been considered for manipulation by \cite{gabicciniisrr15}.
@@
 For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
-used black-box physics simulators to perform Model Predictive Control for \deladp{the motion of} a humanoid character, and managed to obtain dynamically-consistent motions
+used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
 at \gls{interactive} frame rates (we say that a method is interactive when the computation time for planning is less than the
@@
 The \textit{contact reachable} space is a low-dimensional space for which there exists a mapping to the contact manifold.
-\deladp{The space is sampled thanks to the \textit{reachability condition}, a low-dimensional geometric condition for \textit{contact reachability}.}
 
@@
 Finally, to further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
-For HRP-2, we detail the complete computation times \deladp{we obtained} to address sequentially the three problems, and compare them \deladp{with respect} to related work\deladp{, thus clearly} demonstrating that our method is order of magnitude faster\deladp{ than any existing method}.
+For HRP-2, we detail the complete computation times to address sequentially the three problems, and compare them to related work, demonstrating that our method is order of magnitude faster.
 

### 2017-01-11 body/rbprm.tex
finished first revision of rbprm part

@@
 $C_{Reach}^0$ can be sampled efficiently thanks to Eq.~\ref{eq:reachc}, and can thus be used with a standard motion planner.
-Thus, the only significant change regarding a classical planner is to replace the collision checking with the \textit{rechability condition} when verifying
+Thus, the only significant change regarding a classical planner is to replace the collision checking with the \textit{reachability condition} when verifying
 the drawn configurations and associated local paths.
 
-However, to improve the sampling efficiency 
-%~ while remaining probabilistically complete, 
-we bias the sampling process to generate near-obstacle configurations, similarly to~\cite{Amato98choosinggood}.
-%~ First, a configuration is set to a random point on the surface of one randomly chosen obstacle. The root location is then translated and rotated randomly until the \gls{reachc} is satisfied.
+\delst{However, to improve the sampling efficiency 
+we bias the sampling process to generate near-obstacle configurations, similarly to cite{Amato98choosinggood}.}
 Our current implementation of these modifications is based on the Bi-RRT planner \citep{770022} provided by the HPP software.
 
-Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
+In this Section we have thus presented a guide path planner for the geometric root of a robot. This planner is implemented as low a dimensional sampling based 
+algorithm. Given start and goal configurations, it outputs a continuous path for the root of the robot, such that for any configuration in the path there exists a joint
+configuration that results in static equilibrium.
+%~ Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
 %~ By doing this we can solve the problem with a sampling-based approach, in \gls{interactive} computation times.

### 2017-01-11 body/rbprm.tex
[rbprm]

@@
 In this Section we have thus presented a guide path planner for the geometric root of a robot. This planner is implemented as low a dimensional sampling based 
-algorithm. Given start and goal configurations, it outputs a continuous path for the root of the robot, such that for any configuration in the path there exists a joint
-configuration that results in static equilibrium.
+algorithm.
+Given start and goal configurations, it outputs a continuous path for the root of the robot.
+Thanks to the \textit{reachability condition}, we assume that any for any configuration in the path there exists a joint configuration that results in static equilibrium.
 %~ Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).

### 2017-01-11 body/overview.tex
Review Overview

@@
 Figure~\ref{fig:framework} illustrates our workflow.
-\Pa and \Pb are addressed sequentially: we first plan a root guide path with the \textit{reachability condition}, before
+\Pa and \Pb are addressed sequentially: we first plan a root guide path \deladp{with the \textit{reachability condition}}, before
 extending it into a sequence of configurations in static equilibrium.
@@
 The dimension of the path is equal to the number of degrees of freedom of the root of the robot.
-As for previous works~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: the root configurations of the path must be extended into full-body static equilibrium configurations.
-Previous work verify \textit{equilibrium feasibility} by explicitly computing such a joint configuration.
+\repadp{Similarly to}{As for} previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: the root configurations \deladp{of the path} must be \repadp{extensible}{extended} into full-body static equilibrium configurations.
+Previous work verifies \textit{equilibrium feasibility} by explicitly computing such a \deladp{joint} configuration.
 We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact rechability}, illustrated in the following.
 
-An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a) to allow contact creation; not too close, because the robot must avoid collision.
-\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).
-More precisely, a root configuration is \textit{contact reachable} if the root scaled by a factor $s \geq 1$ is not in collision, while the reachable workspace is in collision.
+An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' \addadp{to the environment}: close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a) to allow contact creation; not too close, because the robot must avoid collision.
+\deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
+More precisely, a root configuration is \textit{contact reachable} if the root scaled by a \addadp{user-defined} factor $s \geq 1$ is not in collision, while the reachable workspace is in collision.
 %~ Verifying the \textit{reachability condition} is very efficient in practice.
-To plan a root path, we then use a RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
+To plan a root path, we then use an RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
 This is detailed in Section~\ref{rbprm}.
@@
 The algorithm takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually.
-Candidate limb configurations are selected from a database (independent from the environment) those resulting in the end-effector being close to the environment.
+Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database (\repadp{computed offline independently}{independent} from the environment).
 Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.
@@
 
-\section{Notation conventions and definitions} \label{notations}
+\section{Notation \deladp{conventions} and definitions} \label{notations}
 

### 2017-01-11 body/rbprm.tex
Review Rbprm

@@
 
-In the root path planning phase of our framework, we only consider the root configuration $\mathbf{q}^0$ defined in the previous Section,
+\repadp{During the root path planning}{In the root path planning phase of our framework,} we only consider the root configuration $\mathbf{q}^0$ defined in the previous Section,
 as well as the environment $O$.
@@
 This means that any root configuration must be extended into a whole-body, static equilibrium configuration.
-$C_{Equil}^0$  can't be described analytically, because the contact manifold has a zero measure.
+$C_{Equil}^0$  cannot be described analytically\deladp{, because the contact manifold has a zero measure}.
 
@@
 \textbf{necessary conditions:}
-For a contact to be possible, a volume $O_i \in O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot (Figure~\ref{fig:contact_gen}--1).
-Also the torso of the robot $W^0(\mathbf{q}^{0})$ is necessarily collision-free.
-Therefore we can define an outer approximation  $ \mathcal{C}^0_{\textrm{\it Nec}} \supset$ $C_{Contact}^0$ defined as: 
+For a contact to be possible, \repadp{an obstacle}{a volume} $O_i \in O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot (Figure~\ref{fig:contact_gen}--1).
+Also the torso of the robot $W^0(\mathbf{q}^{0})$ \repadp{must}{is} necessarily \addadp{be} collision-free.
+Therefore we can define an outer approximation  $ \mathcal{C}^0_{\textrm{\it Nec}} \supset$ $C_{Contact}^0$ as: 
 \begin{equation}
@@
 
-\textbf{a sufficient condition:}
+\textbf{sufficient condition:}
 Similarly we can define an inner approximation \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ $C_{Contact}^0$} by considering a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces in contact.
@@
 $C_{Reach}^0$ can be sampled efficiently thanks to Eq.~\ref{eq:reachc}, and can thus be used with a standard motion planner.
-Thus, the only significant change regarding a classical planner is to replace the collision checking with the \textit{reachability condition} when verifying
-the drawn configurations and associated local paths.
+\repadp{The}{Thus, the} only significant change \deladp{regarding a classical planner} is to replace the collision checking with the \textit{reachability condition} \deladp{when verifying
+the drawn configurations and associated local paths}.
 
@@
 
-In this Section we have thus presented a guide path planner for the geometric root of a robot. This planner is implemented as low a dimensional sampling based 
+\repadp{This Section has}{In this Section we have thus} presented a guide path planner for the geometric root of a robot. This planner is implemented as a low-dimensional sampling-based 
 algorithm.
 Given start and goal configurations, it outputs a continuous path for the root of the robot.
-Thanks to the \textit{reachability condition}, we assume that any for any configuration in the path there exists a joint configuration that results in static equilibrium.
+\deladp{Thanks to the \textit{reachability condition}, we assume that for any configuration in the path there exists a joint configuration that results in static equilibrium.}
 %~ Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).

### 2017-01-11 body/overview.tex
[Overview - Andrea_s comment]

@@
 Figure~\ref{fig:framework} illustrates our workflow.
-\Pa and \Pb are addressed sequentially: we first plan a root guide path \deladp{with the \textit{reachability condition}}, before
+\Pa and \Pb are addressed sequentially: we first plan a root guide path, before
 extending it into a sequence of configurations in static equilibrium.
@@
 The dimension of the path is equal to the number of degrees of freedom of the root of the robot.
-\repadp{Similarly to}{As for} previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: the root configurations \deladp{of the path} must be \repadp{extensible}{extended} into full-body static equilibrium configurations.
-Previous work verifies \textit{equilibrium feasibility} by explicitly computing such a \deladp{joint} configuration.
+Similarly to previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: the root configurations must be extensible into full-body static equilibrium configurations.
+Previous work verifies \textit{equilibrium feasibility} by explicitly computing such a configuration.
 We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact rechability}, illustrated in the following.
 
-An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' \addadp{to the environment}: close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a) to allow contact creation; not too close, because the robot must avoid collision.
-\deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
-More precisely, a root configuration is \textit{contact reachable} if the root scaled by a \addadp{user-defined} factor $s \geq 1$ is not in collision, while the reachable workspace is in collision.
+An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' to the environment: close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a) to allow contact creation; not too close, because the robot must avoid collision.
+%~ \deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
+More precisely, a root configuration is \textit{contact reachable} if the root scaled by a user-defined factor $s \geq 1$ is not in collision, while the reachable workspace is in collision.
 %~ Verifying the \textit{reachability condition} is very efficient in practice.
@@
 The algorithm takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually.
-Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database (\repadp{computed offline independently}{independent} from the environment).
+Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database computed offline independently from the environment).
 Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.
@@
 
-\section{Notation \deladp{conventions} and definitions} \label{notations}
+\section{Notation and definitions} \label{notations}
 

### 2017-01-11 body/rbprm.tex
[RBPRM] andrea s comment integrated

@@
 
-\repadp{During the root path planning}{In the root path planning phase of our framework,} we only consider the root configuration $\mathbf{q}^0$ defined in the previous Section,
+During the root path planning we only consider the root configuration $\mathbf{q}^0$ defined in the previous Section,
 as well as the environment $O$.
@@
 This means that any root configuration must be extended into a whole-body, static equilibrium configuration.
-$C_{Equil}^0$  cannot be described analytically\deladp{, because the contact manifold has a zero measure}.
+$C_{Equil}^0$  cannot be described analytically.
 
@@
 \textbf{necessary conditions:}
-For a contact to be possible, \repadp{an obstacle}{a volume} $O_i \in O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot (Figure~\ref{fig:contact_gen}--1).
-Also the torso of the robot $W^0(\mathbf{q}^{0})$ \repadp{must}{is} necessarily \addadp{be} collision-free.
+For a contact to be possible, an obstacle $O_i \in O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot (Figure~\ref{fig:contact_gen}--1).
+Also the torso of the robot $W^0(\mathbf{q}^{0})$ must necessarily be collision-free.
 Therefore we can define an outer approximation  $ \mathcal{C}^0_{\textrm{\it Nec}} \supset$ $C_{Contact}^0$ as: 
@@
 $C_{Reach}^0$ can be sampled efficiently thanks to Eq.~\ref{eq:reachc}, and can thus be used with a standard motion planner.
-\repadp{The}{Thus, the} only significant change \deladp{regarding a classical planner} is to replace the collision checking with the \textit{reachability condition} \deladp{when verifying
-the drawn configurations and associated local paths}.
+The only significant change is to replace the collision checking with the \textit{reachability condition}.
 
@@
 
-\repadp{This Section has}{In this Section we have thus} presented a guide path planner for the geometric root of a robot. This planner is implemented as a low-dimensional sampling-based 
-algorithm.
-Given start and goal configurations, it outputs a continuous path for the root of the robot.
-\deladp{Thanks to the \textit{reachability condition}, we assume that for any configuration in the path there exists a joint configuration that results in static equilibrium.}
+This Section has presented a guide path planner for the geometric root of a robot. This planner is implemented as a low-dimensional sampling-based 
+algorithm. Given start and goal configurations, it outputs a continuous path for the root of the robot. 
+%~ \deladp{Thanks to the \textit{reachability condition}, we assume that for any configuration in the path there exists a joint configuration that results in static equilibrium.}
 %~ Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).

### 2017-01-11 body/contact.tex
[Contact planning] + cosmetics in overview intro and rbprm

@@
 \label{sec:contact}
-Our planner computes guide paths in $C_{Reach}^0$ , an approximation of \gls{$C_{Equil}^0$}.
-As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Equil}^0$}.
-If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section~\ref{sec:perf}.
-We now consider the second problem of computing a discrete sequence of equilibrium configurations $\mathbf{Q}^{\overline{0}}$ along $\mathbf{q}^0(t)$.
+In the second phase, we compute a discrete sequence of static equilibrium configurations $\mathbf{Q}^{\overline{0}}$ given a root path
+$\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Reach}^0 \simeq C_{Equil}^0$}. This contact planner is built on top of
+a contact generator, used to generate static equilibrium configurations. We first describe the generator, before detailing the contact planning algorithm.
+%~ Our planner computes guide paths in $C_{Reach}^0$ , an approximation of \gls{$C_{Equil}^0$}.
+%~ As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Equil}^0$}.
+%~ If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section~\ref{sec:perf}.
+%~ We now consider the second problem of computing a discrete sequence of equilibrium configurations $\mathbf{Q}^{\overline{0}}$ along $\mathbf{q}^0(t)$.
 
-In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a limb, given a
-root location.
-Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.
+%~ In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a limb, given a
+%~ root location.
+%~ Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.
 
-Our criterion to assert efficiently the static equilibrium of the system
-is described in Section~\ref{sec:heuristics}.
+%~ Our criterion to assert efficiently the static equilibrium of the system
+%~ is described in Appendix~\ref{sec:heuristics}.
 
@@
 \label{sec:single_contact}
-Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
-$R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
-While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
-\begin{enumerate}
-\item Generate randomly a collision-free limb configuration;
-\item Project the end-effector onto the closest surface with inverse kinematics;
-\item If a valid solution is found, stop. Otherwise repeat from step 1.
-\end{enumerate}
 
-We trade the completeness of this approach for a more efficient solution introduced in our previous work~\citep{Tonneau2014}, which we describe
-here to be exhaustive.
+\begin{figure*}
+  \centering
+  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
+		\put (1,1) {a} 
+		\put (22,1) {b} 
+		\put (42,1) {c} 
+		\put (62,1) {d} 
+		\put (83,1) {e} 
+	\end{overpic}
+  \caption{Generation of a contact configuration for the right leg of HRP-2.
+(a): Selection of reachable obstacles.
+(b): Entries of the limb samples database (with $N = 4$).
+(c): With a proximity query on the octree database, configurations too far from obstacles are eliminated.
+(d): The best candidate according to a user-defined heuristic $h$ is chosen.
+(e): The final contact is achieved using inverse kinematics.}
+  \label{fig:contact_gen}
+\end{figure*}
 
-We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimum distance 
-between an effector and a surface of the environment is less than a small value $\epsilon$.
+A contact generator is defined for a given limb $R^k$, and only affects the dofs of $R^k$.
+Given a robot configuration, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
+\begin{equation}
+\label{eq:contact_gen}
+	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \wedge \mathbf{q}^k \in  C_{Contact}^k 
+\end{equation}
+
+Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the considered effector onto the closest surface with an inverse kinematics solver.
+This process is repeated until either a solution for Eq.~\ref{eq:contact_gen} is found or the generator fails, according to a user-defined maximum number of trials.
+We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
+
+%~ Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
+%~ $R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
+%~ While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
+%~ \begin{enumerate}
+%~ \item Generate randomly a collision-free limb configuration;
+%~ \item Project the end-effector onto the closest surface with inverse kinematics;
+%~ \item If a valid solution is found, stop. Otherwise repeat from step 1.
+%~ \end{enumerate}
+
+%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
+
+We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum distance 
+between an effector and an obstacle is lesser than $\epsilon \in \mathbb{R}$.
 We then apply the following steps:
@@
 \item At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
-\item According to a user-defined heuristic $h$, sort the elements of $S$;
-\item Select the first configuration of $S$. Project the configuration onto contact with inverse kinematics. If $S$ is empty, stop (failure case) (Figure~\ref{fig:contact_gen} (d) and (e));
-\item If a valid solution is found, stop (success case). Otherwise remove the element from $S$ and go back to step 5.
+\item Use a user-defined heuristic $h$ to sort $S$;
+\item If $S$ is empty, stop (failure). Else select the first configuration of $S$. Project it onto contact using inverse kinematics. (Figure~\ref{fig:contact_gen} (d) and (e));
+\item If Eq.~\ref{eq:contact_gen} is verified, stop (success). Otherwise remove the element from $S$ and go to step 5.
 \end{enumerate}
 
-This approach has two main advantages.
-First, this allows us to select a large number of candidates in a single proximity request.
-Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
-a locally-optimal candidate.
-Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
-Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
-$N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
-The reader is invited to refer to our previous work for an extensive discussion on the optimal value of $N$ \citep{Tonneau2014}.
+%~ This approach has two main advantages.
+%~ First, this allows us to select a large number of candidates in a single proximity request.
+%~ Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
+%~ a locally-optimal candidate.
+%~ Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
+%~ Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
+%~ $N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
+The reader is invited to refer to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
+of the $N$ parameter \citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ used are detailed in Appendix~\ref{sec:heuristics}.
 
@@
 
-\subsection{Extension of the guide path}
-Using the contact generator, we define an iterative algorithm to generate the contact
-sequence along the guide path.
-Although we address acyclic contact sequences, the algorithm is deterministic in the order in which 
-the contacts are created, allowing it to break the combinatorial.
-The complete algorithm can be found in Appendix~\ref{app:contact}.
-In the remainder of the section we provide an intuition of it.
+\subsection{Contact planning algorithm}
+Starting from an initial whole-body configuration, we use the contact generator to compute a sequence
+of whole-body configurations  $\mathbf{Q}^{\overline{0}}$ along the root path $\mathbf{q}^0(t)$.
+The algorithm details can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
+%~ Although we address acyclic contact sequences, the algorithm is deterministic in the order in which 
+%~ the contacts are created, allowing it to break the combinatorial.
 
-As an input, we consider the guide path $\mathbf{q}^0(t)$, discretized into a sequence of $j$ key configurations:  
+First, the root path $\mathbf{q}^0(t)$ is discretized into a sequence of $j$ key configurations:  
 \begin{equation*}
@@
 \end{equation*} 
-where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations. %To ensure continuity in the contact transition phases,
-We want to extend the configurations of $\mathbf{Q}^0$ in such a way that continuity is preserved regarding the contact transitions.
-To do so, we define an algorithm that, given the current root configuration, and the previous full body configuration, computes a full body configuration 
-in \gls{ccequil} such that contacts are maintained if possible.
-The first full body configuration of the sequence is given by the initial state of the robot.
+where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
+Each root configuration of $\mathbf{Q}^0$ is then extended into a whole-body configuration such that:
+\begin{itemize} 
+\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
+\item Each configuration is in static equilibrium.
+\item Each configuration is collision-free.
+\end{itemize}
+
+%~ We want to extend the configurations of $\mathbf{Q}^0$ in such a way that continuity is preserved regarding the contact transitions.
+%~ To do so, we define an algorithm that, given the current root configuration, and the previous full body configuration, computes a full body configuration 
+%~ in $C_{Equil}$ such that contacts are maintained if possible.
+%~ The first full body configuration of the sequence is given by the initial state of the robot.
 
@@
 
-Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
-If possible, a limb in contact at step $i-1$ remains in contact at step $i$.
-The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits.
-The solver is directly provided by the HPP software.
-
-If the solver fails, the contact is broken and a collision-free configuration is assigned to the limb.
+%~ Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
+If kinematically possible, a limb in contact at step $i-1$ remains in contact at step $i$ (Figure~\ref{fig:break_contact}). 
+%~ The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits. 
+%~ The solver is directly provided by the HPP software.
+Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
 If two or more contacts are broken in a single step, one or more intermediate steps are added.
-For these steps the root configuration is the same as for the previous step, with the difference that
-one faulty contact is repositioned, in the hope that it will not be broken at the next step.
+%~ For these steps the root configuration is the same as for the previous step, with the difference that
+%~ one faulty contact is repositioned, in the hope that it will not be broken at the next step.
 
@@
 \begin{enumerate}
-\item Only one contact creation can happen between two consecutive steps. This increases the odds that the interpolation between the two steps be feasible;
+\item At most one contact creation happens between two consecutive steps. 
+%~ This increases the odds that the interpolation between the two steps be feasible;
 %\adnote{Before you said that you ensured that transition between two states is feasible...}
@@
 
-These three rules provide a deterministic generation of contacts along the discretized guide path.
-For a given \gls{equilibrium feasible} path, our current implementation does not guarantee that the planner will succeed, because
-of the deterministic approach used to break the combinatorial.
+The set of rules defined in this Section define a deterministic contact planning algorithm.
+The determinism of this algorithm makes it efficient, at the cost of not being probabilistically complete.
 However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.

### 2017-01-11 body/overview.tex
[Contact planning] + cosmetics in overview intro and rbprm

@@
 %
-\begin{figure*}
-  \centering
-  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
-		\put (1,1) {a} 
-		\put (22,1) {b} 
-		\put (42,1) {c} 
-		\put (62,1) {d} 
-		\put (83,1) {e} 
-	\end{overpic}
-  \caption{Generation of a contact configuration for the right leg of HRP-2.
-(a): Selection of reachable obstacles.
-(b): Entries of the limb samples database (with $N = 4$).
-(c): With a proximity query on the octree database, configurations too far from obstacles are eliminated.
-(d): The best candidate according to a user-defined heuristic $h$ is chosen.
-(e): The final contact is achieved using inverse kinematics.}
-  \label{fig:contact_gen}
-\end{figure*}
 
@@
 
-An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' to the environment: close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a) to allow contact creation; not too close, because the robot must avoid collision.
+An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' to the environment: close, because a contact surface must be partially included in the reachable workspace of the robot to allow contact creation; not too close, because the robot must avoid collision.
 %~ \deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
-More precisely, a root configuration is \textit{contact reachable} if the root scaled by a user-defined factor $s \geq 1$ is not in collision, while the reachable workspace is in collision.
+More precisely, a root configuration is \textit{contact reachable} if the root scaled by a user-defined factor $s \geq 1$ is not in collision (Figure~\ref{fig:reach_cond} - red shape), while the reachable workspace is in collision (Figure~\ref{fig:reach_cond} - green shapes).
 %~ Verifying the \textit{reachability condition} is very efficient in practice.
@@
 To achieve this the root path is first decomposed into a sequence of discrete root configurations, according to a user-defined discretization step.
-Each root configuration is then extended into a whole-body configuration, subject to three constraints:
-\begin{itemize} 
-\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
-\item Each configuration is in static equilibrium.
-\item Each configuration is collision-free.
-\end{itemize}
-
+Each root configuration is then extended into a whole-body configuration in static equilibrium.
 The algorithm thus proceeds iteratively, starting from the whole-body initial configuration of the robot.
-
-The algorithm takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually.
-Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database computed offline independently from the environment).
-Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.
+It takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually. 
+%~ Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database computed offline independently from the environment).
+%~ Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.
 
@@
 The task efficiency is measured using the Extended FORce Transmission ratio (EFORT), proposed by ~\cite{Tonneau:2014:TEC:2619648.2619652}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
-  \caption{
-           Reachable workspace and torso bounding box of HyQ. The green shapes represent the reachable workspace $W^k$ of each limb. The red shape is $W^0$.}
-		   \label{fig:HyQ_roms}
-\end{figure}
 
@@
 $C_{Contact}$ is the set of whole body-configurations in contact and collision-free.
+$C_{Contact}^k \subset C_{Contact}$ is the set of whole body-configurations where at least $R^k$ is in contact.
 
@@
 %~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.
+\begin{figure}
+  \centering
+  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
+  \caption{
+           Reachable workspace and torso bounding box of HyQ. The green shapes represent the reachable workspace $W^k$ of each limb. The red shape is $W^0$.}
+		   \label{fig:HyQ_roms}
+\end{figure}

### 2017-01-12 body/appendix_algo.tex
[Appendix] + starting results

--- 2016-01-26 body/appendix_algo.tex
@@
 
-\section{Derivation of the manipulability ellipsoid}
-\label{app:manipulability}
-
-Again, we assume that $\mathbf{J}$ is full rank. We discard the $k$ indices, and write the pseudo-inverse of $\mathbf{J}$ as $\mathbf{J}^{\dagger}$.
-\begin{eqnarray*}
-\mathbf{\dot{p}} & =  & \mathbf{J} \dot{\mathbf{q}} \\ 
-\mathbf{\dot{q}} & =  & \mathbf{J}^{\dagger} \dot{\mathbf{p}} \\ 
-\mathbf{\dot{q}}^T & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \\ 
-\mathbf{\dot{q}}^T\mathbf{\dot{q}} & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} \dot{\mathbf{p}}\\ 
-\end{eqnarray*}
-
-Then, proving that $\mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} = (\mathbf{J}\mathbf{J}^T)^{-1}$ is straight-forward,
-using the SVD decomposition of each term.
+%~ \section{Derivation of the manipulability ellipsoid}
+%~ \label{app:manipulability}
+%~ 
+%~ Again, we assume that $\mathbf{J}$ is full rank. We discard the $k$ indices, and write the pseudo-inverse of $\mathbf{J}$ as $\mathbf{J}^{\dagger}$.
+%~ \begin{eqnarray*}
+%~ \mathbf{\dot{p}} & =  & \mathbf{J} \dot{\mathbf{q}} \\ 
+%~ \mathbf{\dot{q}} & =  & \mathbf{J}^{\dagger} \dot{\mathbf{p}} \\ 
+%~ \mathbf{\dot{q}}^T & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \\ 
+%~ \mathbf{\dot{q}}^T\mathbf{\dot{q}} & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} \dot{\mathbf{p}}\\ 
+%~ \end{eqnarray*}
+%~ 
+%~ Then, the equality $\mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} = (\mathbf{J}\mathbf{J}^T)^{-1}$ follows from the SVD decomposition of each term
+%~ ~\citep{ben2003generalized}.

### 2017-01-12 body/appendix_roms.tex
merging appendices
[Appendix] + starting results

--- 2016-01-25 body/appendix_roms.tex
@@
+%~ \clearpage
 \section{Generating the $W$ volumes for HRP-2}
@@
 
-In this Appendix,
-we detail our method to generate the volumes $W$ used
+We detail our method to generate the volumes $W$ used
 in RB-RRT, with the example of HRP-2.
-
-We consider the robot HRP-2, and proceed to the decomposition 
-of its kinematic chain into four limbs $R^k$.
+The kinematic chain is split into four limbs $R^k$.
 The arms are connected to the shoulders, and the legs to the root.
@@
 	\end{overpic}
-\caption{Different approximations of the range of motion of the right arm HRP-2. Left: non convex-hull, computed with the powercrust algorithm~\citep{Amenta:2001:PC:376957.376986}. Middle:
+\caption{Different approximations of the range of motion of the right arm of HRP-2. Left: non convex-hull, computed with the powercrust algorithm~\citep{Amenta:2001:PC:376957.376986}. Middle:
 convex hull of the reachable workspace. Right: Simplified hull used in our experiments.}
@@
 Regarding the procedure, we can see that step 2 is conservative (Figure~\ref{fig:hrp2_roms}--right), which 
-is acceptable, especially because the lost set essentially relates to configurations close to singularity (they are close to the boundaries of the reachable workspace).
-We choose again to be less complete
-but more efficient, regarding the number of collision tests to be performed by RB-RRT.
+is acceptable, especially because the lost set essentially relates to configurations close to singularity (they are close to the boundaries of the reachable workspace, and
+often not \gls{contact reachable}, as illustrated in Figure~\ref{fig:dedefeas}, where the exterior boundaries of the reachable workspace appear
+red, thus not belonging to $C_{Contact}^0$). We choose again to be less complete but more efficient, regarding the number of collision tests to be performed by RB-RRT.
 In step 1 on the other hand, selecting the convex hull (Figure~\ref{fig:hrp2_roms}--middle) instead of a minimum encompassing shape (Figure~\ref{fig:hrp2_roms}--left) may introduce false positives.

### 2017-01-12 body/efort.tex
[Appendix] + starting results

@@
 \subsection{Manipulability-based heuristics for contact selection}
-This Section proposes heuristics to select a contact that optimizes desiered capabilities.
+This Section proposes heuristics to select a contact that optimizes desired capabilities.
 %~ These heuristics allow to locally optimize a contact location with respect to a particular task.
@@
 a high velocity in a given direction, or to stay away from singular configurations.
-In this section, we derive three such heuristics from the work on manipulability by \cite{Yoshikawa1984}, that we present first.
+We derive three such heuristics from the work by \cite{Yoshikawa1984}, recalled here. %, that we present first.
 
 \subsubsection{The force and velocity ellipsoids:}
-For a limb configuration $\mathbf{q}^k$, the Jacobian matrix
- $\mathbf{J}^k(\mathbf{q}^k)$   defines the relation:
-
-\begin{equation} \label{jac}
-\mathbf{\dot{p}}^k = \mathbf{J}^k(\mathbf{q}^k) \dot{\mathbf{q}}^k
-\end{equation}
-
-For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
-
-As a linear approximation of a forward-kinematics function, $\mathbf{J}$ describes how small
-variations from the configuration $\mathbf{q}$ affect the position vector $\mathbf{p}$.
-%~ It can be seen as the transformation mapping a velocity in the configuration space into a velocity in
-%~ the Euclidian space.
-
-Now we consider the unit ball in the configuration space $C$ defined by the set of joint velocities
-for which the norm is at most 1:
-\begin{equation} \label{ball}
-||\dot{\mathbf{q}}||^2 \leq 1
-\end{equation}
-
-We assume that $\mathbf{J}$ is full rank (we are not interested in singular configurations, which we discard).
-From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
-\begin{equation} \label{jaceq}
-\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} = \dot{\mathbf{q}}^T \dot{\mathbf{q}}
-\end{equation}
-
-We can use \eqref{jaceq} to map the ball into an ellipsoid in the Euclidian space $\mathbb{R}^m$:
-
-\begin{equation} \label{ellipsoid}
-\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} \leq 1
-\end{equation}
-This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of end-effector velocities that can
-be reached under the constraint \eqref{ball} for the current configuration.
-The longer the axis of the ellipsoid is, the faster the end-effector can move along the direction of the axis.
-Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.
 
@@
 	\end{overpic}
-  % \includegraphics[width=1\linewidth]{ressources/screens/sampling/canap/bd2_cmyk}
   \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 dofs and 2 segments.
-Only the horizontal and vertical speeds are shown (not the rotation speeds), since it would require being able to draw in four dimensions.}
+Only the horizontal and vertical speeds are shown (not the rotation speeds).}
+ %~ since it would require being able to draw in four dimensions.}
 		   \label{sec:efort_ellipsoid}
@@
 
-Similarly to the velocity ellipsoid, \citeauthor{Yoshikawa1984} also defines the force ellipsoid.
-Considering: a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
-the equivalent joint torque vector $\bm{\tau}$;
-we can define the mechanical work in both spaces:
-\begin{equation*} \label{power}
-\dot{\mathbf{q}}^T \bm{\tau} = \dot{\mathbf{p}}^T \mathbf{f}
-\end{equation*}
+We consider: a limb configuration $\mathbf{q}^k$; its end effector position $\mathbf{p}^k$; its Jacobian matrix
+$\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
+\citeauthor{Yoshikawa1984} defines the velocity~\ref{eq:vel} and force~\ref{eq:vel} ellipsoids:
 
-Exploiting \eqref{jac}, we can easily see that the set of achievable forces in $\mathbb{R}^m$ subject to the constraint:
-\begin{equation*} \label{ballforce}
-||\bm{\tau}||^2 \leq 1
-\end{equation*}
-is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right):
-\begin{equation} \label{ellipsoidforce}
+ \begin{equation} 
+ \label{eq:vel}
+\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} \leq 1 
+\end{equation}
+
+ \begin{equation} 
+ \label{eq:for}
 \mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
 \end{equation}
+
+They describe the set of end-effector velocities (respectively forces) that can
+be reached under the constraint $||\dot{\mathbf{q}}||^2 \leq 1$ for the current configuration.
+The longer the axis of the ellipsoid is, the more important the velocity (resp. force) of the end-effector the direction of the axis can be (Figure~\ref{sec:efort_ellipsoid}).
+
+%~ Similarly, \citeauthor{Yoshikawa1984} also defines the force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right):
+
+%~ \begin{equation} \label{ellipsoidforce}
+%~ \mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
+%~ \end{equation}
+%~ with $\mathbf{f}$ is a force vector expressed in the task space $\mathbb{R}^m$;
 
@@
 In our problem, to compare candidate configurations, we include the quality of the contact surface, and choose $\mathbf{m}$ as the direction
-opposite to the local motion (thus given by the difference between two consecutive root positions):
+opposite to the local motion (given by the difference between two consecutive root positions):
 
@@
 the greater $h_{w}$ is, the further away the configuration is from singularity.
-%~ We then derive:
-%~ \begin{equation}
-%~ h_{w}(\mathbf{q}, \mathbf{m}) = w(\mathbf{q})  (\mu \mathbf{n}^T \mathbf{m})
+
+%~ $h_{\textrm{\it EFORT}}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
+%~ can be precomputed off-line), that allow the planner to select the best
+%~ candidates according to user-defined criterion.
+
+%~ This results in faster computation times, and a reduced number of overall contact switches.
+
+%~ where the $\dot$ symbol represents the time derivative.  It describes the set of end-effector velocities that can
+%~ be reached under the constraint \eqref{ball} for the current configuration.
+%~ The longer the axis of the ellipsoid is, the faster the end-effector can move along the direction of the axis (Figure~\ref{sec:efort_ellipsoid}).
+ %~ 
+%~ Similarly, \citeauthor{Yoshikawa1984} also defines the force ellipsoid(Figure~\ref{sec:efort_ellipsoid} - right):
+ %~ 
+%~ \begin{equation} \label{ellipsoidforce}
+%~ \mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
+%~ \end{equation}
+%~ where $\mathbf{f}$ is a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
+%~ the equivalent joint torque vector $\bm{\tau}$;
+ %~ 
+%~ \begin{equation} \label{jac}
+%~ \mathbf{\dot{p}}^k = \mathbf{J}^k(\mathbf{q}^k) \dot{\mathbf{q}}^k
+%~ \end{equation}
+%~ 
+%~ For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
+ %~ 
+%~ As a linear approximation of a forward-kinematics function, $\mathbf{J}$ describes how small
+%~ variations from the configuration $\mathbf{q}$ affect the position vector $\mathbf{p}$.
+
+%~ Now we consider the unit ball in the configuration space $C$ defined by the set of joint velocities
+%~ for which the norm is at most 1:
+%~ \begin{equation} \label{ball}
+%~ ||\dot{\mathbf{q}}||^2 \leq 1
+%~ \end{equation}
+%~ 
+%~ We assume that $\mathbf{J}$ is full rank (we are not interested in singular configurations, which we discard).
+%~ From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
+%~ \begin{equation} \label{jaceq}
+%~ \mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} = \dot{\mathbf{q}}^T \dot{\mathbf{q}}
 %~ \end{equation}
 
-$h_{\textrm{\it EFORT}}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
-can be precomputed off-line), that allow the planner to select the best
-candidates according to user-defined criterion.
-%~ This results in faster computation times, and a reduced number of overall contact switches.
+%~ We can use \eqref{jaceq} to map the ball into an ellipsoid in the Euclidian space $\mathbb{R}^m$:
+
+%~ \begin{equation} \label{ellipsoid}
+%~ \mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} \leq 1
+%~ \end{equation}
+%~ This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of end-effector velocities that can
+%~ be reached under the constraint \eqref{ball} for the current configuration.
+%~ The longer the axis of the ellipsoid is, the faster the end-effector can move along the direction of the axis.
+%~ Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.
+
+%~ \begin{figure}[!tbp]
+  %~ \centering
+	%~ \begin{overpic}[width=1\linewidth]{figures/EFORT/ellipsoid}
+		%~ \put (8.5,1.2) {\small{Velocity ellipsoid}}
+		%~ \put (62.5,1.2) {\small{Force ellipsoid} \tiny{(scale 0.5)}}
+	%~ \end{overpic}
+  %~ \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 dofs and 2 segments.
+%~ Only the horizontal and vertical speeds are shown (not the rotation speeds), since it would require being able to draw in four dimensions.}
+		   %~ \label{sec:efort_ellipsoid}
+%~ \end{figure}
+
+%~ Similarly to the velocity ellipsoid, \citeauthor{Yoshikawa1984} also defines the force ellipsoid.
+%~ Considering: a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
+%~ the equivalent joint torque vector $\bm{\tau}$;
+%~ we can define the mechanical work in both spaces:
+%~ \begin{equation*} \label{power}
+%~ \dot{\mathbf{q}}^T \bm{\tau} = \dot{\mathbf{p}}^T \mathbf{f}
+%~ \end{equation*}
+%~ 
+%~ 
+%~ Exploiting \eqref{jac}, we can easily see that the set of achievable forces in $\mathbb{R}^m$ subject to the constraint:
+%~ \begin{equation*} \label{ballforce}
+%~ ||\bm{\tau}||^2 \leq 1
+%~ \end{equation*}
+%~ is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right):
+%~ \begin{equation} \label{ellipsoidforce}
+%~ \mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
+%~ \end{equation}

### 2017-01-12 body/equilibrium.tex
ongoing intro rewriting
Move Section about heuristics to Appendix
[Appendix] + starting results

@@
 % !TEX root =  ../main.tex
-\section{Heuristics for contact selection}
+\section{Heuristics and criteria for contact selection}
 \label{sec:heuristics}
-\subsection{A heuristic for robust static equilibrium}
+\subsection{A criterion for robust static equilibrium}
 The planner is designed so that any generated contact configuration is in static equilibrium.
-%~ Equilibrium is of course critical in multipedal locomotion. For this reason 
+%~ Equilibrium is of course critical in legged locomotion. For this reason 
 We are interested in a robust 
@@
 %~ \item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
-\item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
+\item $m \in \mathbb{R}$ is the robot mass;
 \item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;

### 2017-01-12 body/results.tex
[Appendix] + starting results

@@
 Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
-%For each scenario we indicate the chosen heuristics. 
-We also provide a performance analysis, which shows that the planner is compatible with \gls{interactive} applications,
-and present the success rates obtained in each scenario. Moreover, we demonstrate the interest of our robustness criterion in the different computed poses.
 Finally, a last example suggests possible applications to dexterous manipulation.
+At the end of the video, we validate our contact plans by exhibiting a solution to the interpolation problem $\mathcal{P}_3$ for these scenarios.
 
-In our previous work~\citep{tonneauisrr15} additional results are demonstrated with various virtual avatars (Figure~\ref{fig:robots_old}).
-In this extension we choose to focus on actual robots. We invite the interested reader to watch the ISRR video (\url{http://youtu.be/LmLAHgGQJGA}), and 
-to refer to the previous paper for a discussion on these results.
+At the end of this Section, we provide the \textit{interactive} computation times obtained in each case.
+We also compare the times obtained with HRP-2 with respect to previous demonstrations.
+
+These scenarios complete the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
+%~ (Figure~\ref{fig:robots_old}).
+ %~ We invite the interested reader to watch the ISRR video (\url{http://youtu.be/LmLAHgGQJGA}), and 
+%~ to refer to the previous paper for a discussion on these results.
 

### 2017-01-12 body/results.tex
[Results] darpa scenario

@@
 Finally, a last example suggests possible applications to dexterous manipulation.
+
 At the end of the video, we validate our contact plans by exhibiting a solution to the interpolation problem $\mathcal{P}_3$ for these scenarios.
 
-At the end of this Section, we provide the \textit{interactive} computation times obtained in each case.
-We also compare the times obtained with HRP-2 with respect to previous demonstrations.
+At the end of this Section, we analyze the influence of the parameters of our framework on the output, and provide the \textit{interactive} computation times obtained in each case.
+We also compare the times obtained with HRP-2 with respect to previous works.
 
@@
 The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
-In each scenario we detail the parameters chosen: the heuristics (either $h_{\textrm{\it EFORT}}$, $h_{vel}$ or $h_{w}$, and optionally the equilibrium robustness, all of which are defined in the Appendix~\ref{sec:heuristics}), and the constraints on the reachable workspaces (for instance in all the scenarios,
-the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 
+In each scenario we detail the contacts involved and the heuristics chosen (either $h_{\textrm{\it EFORT}}$, $h_{vel}$ or $h_{w}$, all of which are defined in the Appendix~\ref{sec:heuristics}).
+ %~ and the constraints on the reachable workspaces (for instance in all the scenarios, the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 
 %~ The companion video presents the complete contact sequence obtained in all these scenarios.
@@
 \noindent\textbf{Heuristics:} The manipulability $h_w$ is chosen for the feet; $h_{\textrm{\it EFORT}}$ is chosen for the right arm.
-Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
+%~ Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
 
-\noindent\textbf{Observations:}
-This scenario illustrates best the importance of the equilibrium-robustness criterion.
-With a robust approach, more states are required to reach the last step (15 rather than 13 in average).
-However, when the last step is reached by both feet, in the nonrobust case the contacts are extremely close to 
-the cone limits (Figure~\ref{fig:stair_comp}).
-
-The geometry of the environment is easily addressed by our planner, and the contact planning is several times faster than real time in this scenario.
-
-Again, the interpolation motion between the contact steps is out of the scope of this paper.
+%~ \noindent\textbf{Observations:}
+%~ This scenario illustrates best the importance of the equilibrium-robustness criterion.
+%~ With a robust approach, more states are required to reach the last step (15 rather than 13 in average).
+%~ However, when the last step is reached by both feet, in the nonrobust case the contacts are extremely close to 
+%~ the cone limits (Figure~\ref{fig:stair_comp}).
+%~ 
+%~ 
+%~ The geometry of the environment is easily addressed by our planner, and the contact planning is several times faster than real time in this scenario.
+%~ 
+%~ Again, the interpolation motion between the contact steps is out of the scope of this paper.
 However it should be noted that the computed plan in this scenario has been executed successfully on the robot (\url{http://youtu.be/YjL-DBQgXwk#t=0m28s}).
-
-\begin{figure}
-  \centering
-  %~ \includegraphics[width=0.6\linewidth]{figures/stair_robust}
-  \begin{overpic}[width=0.5\linewidth]{figures/stair_robust}
-		\put (17,5) {\small{\color{red}$b_0 = 0.23$}} 
-		\put (79,5) {\small{\color{green}$b_0 = 6.16$}} 
-		%~ \put (68,) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
-	\end{overpic}
-  \caption{
-           Evaluation of the robustness $b_0$ of two contact configurations. Although in equilibrium, the left configuration is on the verge of slipping.}
-		   \label{fig:stair_comp}
-\end{figure}
+%~ 
+%~ \begin{figure}
+  %~ \centering
+  %~ \begin{overpic}[width=0.5\linewidth]{figures/stair_robust}
+		%~ \put (17,5) {\small{\color{red}$b_0 = 0.23$}} 
+		%~ \put (79,5) {\small{\color{green}$b_0 = 6.16$}} 
+	%~ \end{overpic}
+  %~ \caption{
+           %~ Evaluation of the robustness $b_0$ of two contact configurations. Although in equilibrium, the left configuration is on the verge of slipping.}
+		   %~ \label{fig:stair_comp}
+%~ \end{figure}
 
@@
 
-\noindent\textbf{Observations:} The scenario illustrates well the acyclic aspect of the planning. For instance, in the four first frames of Figure~\ref{fig:standing}, we can see that the right foot
-is moved twice, with the left foot in between, before the configuration allows HRP-2 to move its hand.
-Because the contacts are tried in a FIFO manner, the fact that the output contact sequence is acyclic shows that a cyclic approach (with a finite state machine for instance) is not sufficient
-for the computed path. The reason for this is not reachability, but equilibrium. The planning is slower than for the stair scenario (because the contact generation fails more),
-though it remains compatible with \gls{interactive} performances. % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
+%~ \noindent\textbf{Observations:} The scenario illustrates well the acyclic aspect of the planning. For instance, in the four first frames of Figure~\ref{fig:standing}, we can see that the right foot
+%~ is moved twice, with the left foot in between, before the configuration allows HRP-2 to move its hand.
+%~ Because the contacts are tried in a FIFO manner, the fact that the output contact sequence is acyclic shows that a cyclic approach (with a finite state machine for instance) is not sufficient
+%~ for the computed path. The reason for this is not reachability, but equilibrium. The planning is slower than for the stair scenario (because the contact generation fails more),
+%~ though it remains compatible with \gls{interactive} performances. % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
 
-\subsubsection{HRP-2 -- Car egress (Figure~\ref{fig:car}):}
+\subsubsection{HRP-2 -- Polaris car egress (Figure~\ref{fig:car}):}
 This scenario is inspired from the Darpa challenge car egress scenario (\url{http://cpc.cx/edH}). HRP-2 has 
-to find the contact sequence that allows it to step out of a car.
+to find the contact sequence that allows it to step out of the Polaris car.
 
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{figures/car}
+  \includegraphics[width=1\linewidth]{figures/polaris}
   \caption{
-           Selected frames from the car egress scenario. }
+           Selected frames from the polaris egress scenario. }
 		   \label{fig:car}
@@
 
-\noindent\textbf{Observations:} The difficulty of this scenario lies in the strong reduction of the reachable workspace induced 
-by the extreme proximity of all obstacles. The planner is able to find a sequence, that consists in many steps.
-The proximity of the obstacles invalidate a large number of contact candidates because of collisions.
+%~ \noindent\textbf{Observations:} The difficulty of this scenario lies in the strong reduction of the reachable workspace induced 
+%~ by the extreme proximity of all obstacles. The planner is able to find a sequence, that consists in many steps.
+%~ The proximity of the obstacles invalidate a large number of contact candidates because of collisions.
 To avoid breaking more than one contact between each step, the motion has to be decomposed into a large number of steps (61 in average).
-While this scenario is the slowest to solve, the planner still computes a solution \glslink{interactive}{interactively}.
- % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
+%~ While this scenario is the slowest to solve, the planner still computes a solution \glslink{interactive}{interactively}.
 
@@
 
-\noindent\textbf{Observations:} In this context, setting up a really important minimum value for $b_0$ is possible due to the high
-stability of the HyQ robot, and results in more contact switches, in exchange for safety. The guide path-planning in this scenario takes a few seconds in average, more than
-in any other scenarios.
+%~ \noindent\textbf{Observations:} In this context, setting up a really important minimum value for $b_0$ is possible due to the high
+%~ stability of the HyQ robot, and results in more contact switches, in exchange for safety. The guide path-planning in this scenario takes a few seconds in average, more than
+%~ in any other scenarios.
 This is explained by the necessity of discovering a safe way to ``climb down'' the rubble.
@@
 %\adnote{Why do not relax this constraint and require only 3 legs?}
-Again, the computation times remain however \gls{interactive}.
+%~ Again, the computation times remain however \gls{interactive}.
 
@@
 
-\noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is a hard case for a contact planner.
-While finding a guide path above the hole is easy for the guide planner, finding a sequence of contacts that allows for equilibrium is not trivial.
-Second, the narrow bridge is hard both for the planner and the contact generator: to make sure that equilibrium is preserved along the traversal,
-the bridge must be approached with the appropriate angle.
-The difficulty is illustrated in Figure~\ref{fig:HyQ_obs}, where several feet rearrangements are required to cross the hole (although the video shows this best).
-The planner however succeeds in finding a feasible sequence in the end, again with \gls{interactive} computation times.
+%~ \noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is a hard case for a contact planner.
+%~ While finding a guide path above the hole is easy for the guide planner, finding a sequence of contacts that allows for equilibrium is not trivial.
+%~ Second, the narrow bridge is hard both for the planner and the contact generator: to make sure that equilibrium is preserved along the traversal,
+%~ the bridge must be approached with the appropriate angle.
+%~ The difficulty is illustrated in Figure~\ref{fig:HyQ_obs}, where several feet rearrangements are required to cross the hole (although the video shows this best).
+%~ The planner however succeeds in finding a feasible sequence in the end, again with \gls{interactive} computation times.
 
@@
 
-\noindent\textbf{Observations:} This scenario is designed to illustrate concretely the computation times of the planner.
-In the video, the footsteps indicating the contact sequence appear at the average speed of their computation (including the guide-path planning).
+%~ \noindent\textbf{Observations:} This scenario is designed to illustrate concretely the computation times of the planner.
+%~ In the video, the footsteps indicating the contact sequence appear at the average speed of their computation (including the guide-path planning).
 
@@
 This subsection discusses the several factors that influence the outcome of our planner: the root scaling factor $s$, the choice of heuristics for
-contact generation, the desired value of the robustness parameter $b_0$, and lastly, the discretization 
+contact generation, 
+%~ the desired value of the robustness parameter $b_0$, 
+and lastly, the discretization 
 of the guide path.
@@
 
-The obtained results for HRP-2 are summarized in Table~\ref{tab:scale}, averaged over all scenes (except for the car egress: in this scenario, 
+The obtained results for HRP-2 are shown in Table~\ref{tab:scale}, averaged over all scenes (except for the car egress: in this scenario, 
 statistical tests are not really conclusive since we are only interested in a small area of the environment).
 As it can be expected, the scaling results in a high increase of the false negatives, while the false positives decrease.
-For HRP-2 we decided to set $s^*=1.2$, that results in less than 3 \% of false positives, and was also effective
-for the car egress scenario.
+For HRP-2 we decided to set $s^*=1.2$, which results in less than 3 \% of false positives.
+%~ , and was also effective for the car egress scenario.
 
@@
 
-\subsubsection{On the robustness equilibrium criterion:} \label{sec:parrob}
-Robustness is really important when considering practical applications on the robot, to account 
-for the various uncertainties that result from environment and state estimation. However,
-maximizing the robustness criterion is often not optimal, because the resulting configurations may be too conservative, thus not favoring the motion.
+%~ \subsubsection{On the robustness equilibrium criterion:} \label{sec:parrob}
+%~ Robustness is really important when considering practical applications on the robot, to account 
+%~ for the various uncertainties that result from environment and state estimation. However,
+%~ maximizing the robustness criterion is often not optimal, because the resulting configurations may be too conservative, thus not favoring the motion.
 In our experiments, we choose not to maximize the robustness,
-but to empirically set a robustness threshold value under which a configuration is not considered to be in
-static equilibrium (If no candidate reaches the threshold, instead of failing, the algorithm can eventually return the ``more robust'' configuration
-found). Currently for HRP-2, a threshold value of 2 subjectively gives the best results in the considered scenarios, while 10 seems to be a good choice for HyQ. Both values do not
-significantly slow down the planning times.
+%~ but to empirically set a robustness threshold value under which a configuration is not considered to be in
+%~ static equilibrium (If no candidate reaches the threshold, instead of failing, the algorithm can eventually return the ``more robust'' configuration
+%~ found). Currently for HRP-2, a threshold value of 2 subjectively gives the best results in the considered scenarios, while 10 seems to be a good choice for HyQ. Both values do not
+%~ significantly slow down the planning times.
 
@@
 To analyze performance, for each considered scenario, we ran the planner 1000 times.
-We measured the computation time spent in each aspect of the algorithm, and also analyzed the success
-rate obtained for each scenario.
+We measured the computation time spent in each aspect of the algorithm, and analyzed their success rate.
 

### 2017-01-12 body/contact.tex
Review section 5

@@
 
-\subsection{Definition of a contact generator}
+\subsection{\repadp{Contact}{Definition of a contact} generator}
 \label{sec:single_contact}
@@
 
-A contact generator is defined for a given limb $R^k$, and only affects the dofs of $R^k$.
-Given a robot configuration, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
+\repadp{The contact generation treats each limb $R^k$ independently.}{A contact generator is defined for a given limb $R^k$, and only affects the DoFs of $R^k$.}
+Given a \repadp{configuration of the root and the other limbs}{robot configuration}, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
 \begin{equation}
@@
 
-Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the considered effector onto the closest surface with an inverse kinematics solver.
-This process is repeated until either a solution for Eq.~\ref{eq:contact_gen} is found or the generator fails, according to a user-defined maximum number of trials.
+Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the \deladp{considered} effector onto the closest surface with an inverse kinematics solver.
+This process is repeated until either a solution for Eq.~\ref{eq:contact_gen} is found, or the generator fails, according to a user-defined maximum number of trials.
 We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
@@
 We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum distance 
-between an effector and an obstacle is lesser than $\epsilon \in \mathbb{R}$.
+between an effector and an obstacle is less than $\epsilon \in \mathbb{R}$.
 We then apply the following steps:
@@
 The reader is invited to refer to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
-of the $N$ parameter \citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ used are detailed in Appendix~\ref{sec:heuristics}.
+of the \repadp{parameter $N$}{$N$ parameter}~\citep{Tonneau2014}.
+Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ \deladp{used} are detailed in Appendix~\ref{sec:heuristics}.
 
@@
 \item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
+\item \addadp{At most one contact is added between two consecutive configurations.} 
 \item Each configuration is in static equilibrium.
@@
 Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
-If two or more contacts are broken in a single step, one or more intermediate steps are added.
+If two or more contacts \repadp{cannot be maintained between two consecutive configurations}{are broken in a single step}, one or more intermediate \repadp{configurations}{steps} are added.
 %~ For these steps the root configuration is the same as for the previous step, with the difference that
@@
 \subsubsection{Creating contacts:}
-Contacts are created using three rules:
+Contacts are created using \deladp{three rules:} \addadp{a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest.
+If the contact creation was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
 \begin{enumerate}
-\item At most one contact creation happens between two consecutive steps. 
+\item \deladp{At most one contact creation happens between two consecutive steps. }
 %~ This increases the odds that the interpolation between the two steps be feasible;
 %\adnote{Before you said that you ensured that transition between two states is feasible...}
-\item A contact is validated if and only if the resulting configuration is in static equilibrium; 
-\item We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
-was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.
+\item \deladp{A contact is validated if and only if the resulting configuration is in static equilibrium;}
+\item \deladp{We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
+was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
 \end{enumerate}
 
-The set of rules defined in this Section define a deterministic contact planning algorithm.
+\repadp{This set of rules}{The set of rules defined in this Section} defines a deterministic contact planning algorithm.
 The determinism of this algorithm makes it efficient, at the cost of not being probabilistically complete.
-However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.
+\deladp{However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.}

### 2017-01-12 body/overview.tex
Review section 5

@@
 We first consider the problem of planning a root guide path (Figure~\ref{fig:framework}--$\mathcal{P}_1$).
-The dimension of the path is equal to the number of degrees of freedom of the root of the robot.
+The dimension of the path is equal to the number of degrees of freedom (DoFs) of the root of the robot.
 Similarly to previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: the root configurations must be extensible into full-body static equilibrium configurations.
@@
 \textbf{A robot} is a kinematic chain $R$ composed of: a root $R^0$, and $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
-The root has $r \geq 6$ DOFS: for instance, HRP-2 has two extra DOFS in the torso, such that we have $r=8$.
+The root has $r \geq 6$ DoFs: for instance, HRP-2 has two extra DoFs in the torso, such that we have $r=8$.
 Therefore $R$ is fully described by a configuration $\mathbf{q} \in  \mathbb{R}^{r+n}$.

### 2017-01-12 body/rbprm.tex
Review section 5

@@
 
-\textbf{sufficient condition:}
+\textbf{sufficient condition\addadp{s}:}
 Similarly we can define an inner approximation \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ $C_{Contact}^0$} by considering a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces in contact.
@@
 The only significant change is to replace the collision checking with the \textit{reachability condition}.
-
-\delst{However, to improve the sampling efficiency 
-we bias the sampling process to generate near-obstacle configurations, similarly to cite{Amato98choosinggood}.}
 Our current implementation of these modifications is based on the Bi-RRT planner \citep{770022} provided by the HPP software.
 
+%~ \delst{However, to improve the sampling efficiency 
+%~ we bias the sampling process to generate near-obstacle configurations, similarly to cite{Amato98choosinggood}.}
+
+\subsubsection*{}
 This Section has presented a guide path planner for the geometric root of a robot. This planner is implemented as a low-dimensional sampling-based 

### 2017-01-12 body/contact.tex
updated figure on julien s advice, integrated [CONTACT] modifs by Andrea

@@
 
-\subsection{\repadp{Contact}{Definition of a contact} generator}
+\subsection{Contact generator}
 \label{sec:single_contact}
@@
 
-\repadp{The contact generation treats each limb $R^k$ independently.}{A contact generator is defined for a given limb $R^k$, and only affects the DoFs of $R^k$.}
-Given a \repadp{configuration of the root and the other limbs}{robot configuration}, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
+The contact generation treats each limb $R^k$ independently.
+Given a configuration of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
 \begin{equation}
@@
 
-Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the \deladp{considered} effector onto the closest surface with an inverse kinematics solver.
+Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
 This process is repeated until either a solution for Eq.~\ref{eq:contact_gen} is found, or the generator fails, according to a user-defined maximum number of trials.
@@
 The reader is invited to refer to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
-of the \repadp{parameter $N$}{$N$ parameter}~\citep{Tonneau2014}.
-Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ \deladp{used} are detailed in Appendix~\ref{sec:heuristics}.
+of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
 
@@
 \item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
-\item \addadp{At most one contact is added between two consecutive configurations.} 
+\item At most one contact is added between two consecutive configurations.
 \item Each configuration is in static equilibrium.
@@
 Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
-If two or more contacts \repadp{cannot be maintained between two consecutive configurations}{are broken in a single step}, one or more intermediate \repadp{configurations}{steps} are added.
+If two or more contacts can't be maintained between two consecutive configurations, one or more intermediate configurations are added.
 %~ For these steps the root configuration is the same as for the previous step, with the difference that
@@
 	\end{overpic}
-\caption{Contacts are maintained if joint limits and collisions constraints are respected (2). They are broken otherwise(3,4).}
+\caption{Contacts are maintained if joint limits and collisions constraints are respected (2). They are broken otherwise(3,4). The green line represents the root path.}
 		   \label{fig:break_contact}
@@
 \subsubsection{Creating contacts:}
-Contacts are created using \deladp{three rules:} \addadp{a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest.
-If the contact creation was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
-\begin{enumerate}
-\item \deladp{At most one contact creation happens between two consecutive steps. }
-%~ This increases the odds that the interpolation between the two steps be feasible;
-%\adnote{Before you said that you ensured that transition between two states is feasible...}
-\item \deladp{A contact is validated if and only if the resulting configuration is in static equilibrium;}
-\item \deladp{We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
-was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
-\end{enumerate}
+Contacts are created using a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest.
+If the contact creation was not successful, the limb is pushed on top of the queue, and will only be tried again after the others.
+%~ \begin{enumerate}
+%~ \item \deladp{At most one contact creation happens between two consecutive steps. }
+%~ \item \deladp{A contact is validated if and only if the resulting configuration is in static equilibrium;}
+%~ \item \deladp{We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
+%~ was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
+%~ \end{enumerate}
 
-\repadp{This set of rules}{The set of rules defined in this Section} defines a deterministic contact planning algorithm.
+This set of rules defines a deterministic contact planning algorithm.
 The determinism of this algorithm makes it efficient, at the cost of not being probabilistically complete.
-\deladp{However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.}
+%~ \deladp{However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.}

### 2017-01-12 body/intro.tex
[Contact planning] + cosmetics in overview intro and rbprm
updated figure on julien s advice, integrated [CONTACT] modifs by Andrea

@@
 % !TEX root =  ../main.tex
+\newcommand{\Pa}{$\mathcal{P}_1$ }
+\newcommand{\Pb}{$\mathcal{P}_2$ }
 
@@
 
+\begin{figure*}
+  \centering
+  \begin{overpic}[width=0.8\linewidth]{figures/workflow}
+    %~ \put (6.4,1.8) {\normalsize{Request}} 
+    \put (30,10.8) {\large{\color{white}\Pa} }
+    %~ \put (31.6,2) {\scriptsize{\color{white}RB-RRT}} 
+    \put (66.4,10.8) {\large{\color{white}\Pb} }
+    %~ \put (72.5,2) {\tiny{\color{white}Contact}} 
+	\put (15,1) {a} 
+	\put (50,1) {b} 
+	\put (85,1) {c} 
+  \end{overpic}
+  \vspace{-1em}
+  \caption{
+    Overview of our two-stage framework. Given a path request between start and goal positions (left image), \Pa is the problem of computing a guide path in the space
+    of \textit{equilibrium feasible} root configurations.
+We achieve this by defining a geometric condition, the \textit{reachability condition} (abstracted with the transparent cylinders on the middle image).
+\Pb is then the problem of extending the path into a discrete sequence of contact configurations using an iterative algorithm (right image).}
+  \label{fig:framework}
+\end{figure*}
+
 \subsection{State of the art}
-
-\newcommand{\Pa}{$\mathcal{P}_1$ }
-\newcommand{\Pb}{$\mathcal{P}_2$ }
 
@@
 
-\begin{figure*}
-  \centering
-  \begin{overpic}[width=0.8\linewidth]{figures/workflow}
-    %~ \put (6.4,1.8) {\normalsize{Request}} 
-    \put (30,10.8) {\large{\color{white}\Pa} }
-    %~ \put (31.6,2) {\scriptsize{\color{white}RB-RRT}} 
-    \put (66.4,10.8) {\large{\color{white}\Pb} }
-    %~ \put (72.5,2) {\tiny{\color{white}Contact}} 
-  \end{overpic}
-  \vspace{-1em}
-  \caption{
-    Overview of our two-stage framework. Given a path request between start and goal positions (left image), \Pa is the problem of computing a guide path in the space
-    of \textit{equilibrium feasible} root configurations.
-We achieve this by defining a geometric condition, the \textit{reachability condition} (abstracted with the transparent cylinders on the middle image).
-\Pb is then the problem of extending the path into a discrete sequence of contact configurations using an iterative algorithm (right image).}
-  \label{fig:framework}
-\end{figure*}
-
 Our solution belongs to the class of decoupled approaches, because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning.
@@
 
-Finally, to further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
-For HRP-2, we detail the complete computation times to address sequentially the three problems, and compare them to related work, demonstrating that our method is order of magnitude faster.
+To further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
+For HRP-2, we detail the complete computation times to address sequentially the three problems, and compare them to related work, demonstrating that our method is orders of magnitude faster.
+
+Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem.
 

### 2017-01-12 body/overview.tex
updated figure on julien s advice, integrated [CONTACT] modifs by Andrea

@@
 Figure~\ref{fig:framework} illustrates our workflow.
-\Pa and \Pb are addressed sequentially: we first plan a root guide path, before
-extending it into a sequence of configurations in static equilibrium.
+\Pa and \Pb are addressed sequentially: From a given problem (a) we first plan a root guide path (b), before
+extending it into a sequence of static equilibrium configurations(c). In the case where step c fails,
+our framework invalidates the computed guide path, and restarts the planning from b.
 
@@
 
-\subsection{Generating a discrete sequence of contact configurations --- \Pb}
-
 The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--$\mathcal{P}_2$).
@@
 The task efficiency is measured using the Extended FORce Transmission ratio (EFORT), proposed by ~\cite{Tonneau:2014:TEC:2619648.2619652}.
+
+\subsection{Generating a discrete sequence of contact configurations --- \Pb}
 
@@
 \medskip
+Finally we define some relevant of the configuration space $C$.
 $C_{Contact}$ is the set of whole body-configurations in contact and collision-free.

### 2017-01-12 body/rbprm.tex
updated figure on julien s advice, integrated [CONTACT] modifs by Andrea

@@
 
-\textbf{sufficient condition\addadp{s}:}
+\textbf{sufficient condition:}
 Similarly we can define an inner approximation \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ $C_{Contact}^0$} by considering a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces in contact.

### 2017-01-12 body/results.tex
updated figure on julien s advice, integrated [CONTACT] modifs by Andrea

@@
 
-At the end of this Section, we analyze the influence of the parameters of our framework on the output, and provide the \textit{interactive} computation times obtained in each case.
+At the end of this Section, we discuss the role of the parameters of our framework on the output. We then provide the \textit{interactive} computation times obtained in each case.
 We also compare the times obtained with HRP-2 with respect to previous works.
@@
 configurations close to singularities (and joint limits). While this does not significantly impact
-the generation of the plan, the resulting interpolation on the real robot turned out to be harder \citep{Carpentier2016}.
+the generation of the plan, the resulting interpolation on the real robot turned out to be harder.
 For this reason, we prefer to use our manipulability-based heuristic for the legs of the robot, but we still
@@
    %~ Car (61)& 1 -- \textbf{529} --  995 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6775 --  \textbf{20890} -- 52963 & \textbf{342} \\
-   Car (61)& 145 -- \textbf{1055} --  15063 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6919 --  \textbf{21416} -- 67031 & \textbf{351} \\
+   Car TODO (61)& 145 -- \textbf{1055} --  15063 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6919 --  \textbf{21416} -- 67031 & \textbf{351} \\
    %~ Rubble (71)& 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{3058} -- 13126 & \textbf{43} \\
@@
 \begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | c}
-  &  Path planning & Equilibrium feasibility & Kinematic failure & Equilibrium failure \\
+  &  Path planning success rate & Success rate & Kinematic failure & Equilibrium failure \\
  \hline
@@
    Standing up & 68\% & 87.8\% & 6.1\% & 6.1\% \\
-   Car egress & 39\% & 77.0\% & 21.0\% &  2.0\% \\
+   Car egress TODO & 39\% & 77.0\% & 21.0\% &  2.0\% \\
    Rubble & 74\% & 97.9\% & 0.1\% & 2.0\% \\
-   Obstacle race & 58.0\% & 95.7\% & 1.8\% & 2.5\% \\
+   Obstacle race TODO & 58.0\% & 95.7\% & 1.8\% & 2.5\% \\
  \end{tabular}
-\caption{Success rates for each scenario, rounded to the first decimal. The first value indicates the percentage of successful complete contact plannings for 1000 tests; The second value
+\caption{ TODO AMBIGUITE POURCENTAGE Success rates for each scenario, rounded to the first decimal.
+The first value indicates the percentage of successful complete contact plannings for 1000 tests; The second value
 indicates the percentage of equilibrium feasible root configurations: considering each limb individually, indicates the percentage of root configurations of the guide path that
@@
 it is also applied every time a new candidate needs to be evaluated. In particular for the car egress scenario,
-the kinematic constraints are very demanding. Therefore a large number of projections fail because of joint-limit violations.
+the kinematic constraints are very demanding to avoid collisions.
 
@@
 In all scenarios, one can observe that the average computation time for one single step is largely below one second,
-thus allowing to consider \gls{interactive} applications.
+thus allowing to consider \gls{interactive} applications. This would allow to consider online autonomous planning of the robot motion.
 
@@
 Table~\ref{tab:requestpercent} summarizes the success rates obtained for each scenario.
+Despite the complexity of the scenarios adressed, as well as the approximations made in our formulation, our planner succeeds in a large majority of cases.
 As expected, our planner does not succeed systematically, because of the approximations made in our formulation.
@@
 surprising considering the narrowness of the scene.
-\deladp{From a pragmatic point of view, }regarding \deladp{the} computation times, \deladp{we claim that} our approach provides a satisfying compromise between completeness and efficiency.
-Indeed, \deladp{the advantage of the framework is that} when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan \deladp{a new contact sequence} with a reasonable chance of success.
-The most efficient (and immediate) approach is probably to launch in parallel several instances of the planner \deladp{for a given problem} (our current implementation is single threaded) and to apply any successful result.
+Regarding computation times, our approach provides a satisfying compromise between completeness and efficiency.
+Indeed, when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan with a reasonable chance of success.
+The most efficient (and immediate) approach is probably to launch in parallel several instances of the planner (our current implementation is single threaded) and to apply any successful result.
+
+%~ \begin{table*}
+%~ \centering
+%~ \begin{tabular}{ c | c | c | c | c }
+ %~ Scenario & Method  & Guide Path & Contact sequence & Interpolation \\
+ %~ \hline
+   %~ \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}5.42 min}  \\
+							 %~ & \textbf{Ours} + \citeauthor{Carpentier2016}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & $ <$ 2s \\
+							 %~ & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
+ %~ \hline
+   %~ \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}4.08 min}  \\
+							 %~ & \textbf{Ours}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & X \\
+							 %~ & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
+ %~ \hline
+   %~ \multirow{3}{*}{Stair 40 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}10.08 min}  \\
+							 %~ & \textbf{Ours}  & \textbf{5 s} & \textbf{3 s} & X \\
+							 %~ & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
+ %~ \hline
+   %~ \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 10 min & \multicolumn{2}{c}{\cellcolor{blue!25}3.5 hours}  \\
+							 %~ & \textbf{Ours}  & \textbf{529 ms} & \textbf{$<$ 22 s} & X \\
+ %~ \end{tabular}
+%~ \caption{Comparison between the computation times obtained by our method and previous ones, for each phase of the planning. Blue cells indicate merged times (because no
+%~ precise time information is available). The X symbol indicates scenarios where we did not address the interpolation of the contact sequence.}
+%~ \label{tab:compprev}
+%~ \quad
+%~ \end{table*}
 
@@
 							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
- \hline
-   \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 10 min & \multicolumn{2}{c}{\cellcolor{blue!25}3.5 hours}  \\
-							 & \textbf{Ours}  & \textbf{529 ms} & \textbf{$<$ 22 s} & X \\
+ %~ \hline
+   %~ \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 10 min & \multicolumn{2}{c}{\cellcolor{blue!25}3.5 hours}  \\
+							 %~ & \textbf{Ours}  & \textbf{529 ms} & \textbf{$<$ 22 s} & X \\
  \end{tabular}
@@
 \end{table*}
-\subsection{Comparison with previous work} \label{sec:compa}
-Comparing our method with others is not trivial. Indeed, we focus on the contact planning aspect of the problem, and do not address
-the interpolation in this paper, nor provide a strong guarantee that our sequences can be interpolated.
-Most papers in the literature only provide the overall computation
-time, without specifying how much is spent in each phase. Moreover, there exists no off-the-shelf implementation
-of the related methods. Due to their complexity, we cannot afford the development resources required to reimplement them.
 
-Furthermore, the interpolation itself is not completely solved.
+\subsection{Validation of the contact plans, and comparison with previous work} \label{sec:compa}
+To validate our plans, we had to implement our own interpolation method to address $\mathcal{P}_3$.
+We refer the curious reader to \citep{Carpentier2016} for the implementation details of our method. 
+This motion is in turn validated in our dynamic simulator~\citep{DelPrete2015b}. The stair climbing scenario was validated on the real HRP-2 robot \citep{Carpentier2016}.
+The obtained motions are available in the video.
+
+Few contributions provided computation times on the complete problem, necessary to provide comparison between our approaches. 
+Furthermore, $\mathcal{P}_3$ is an open problem in the presence of obstacles.
 For instance in \cite{DBLP:conf/iser/EscandeKMG08}, the end-effector trajectories are computed automatically with splines, but are manually corrected if they enter in collision.
@@
 Recent methods suffer from similar issues \citep{Carpentier2016}.
+The only valid scenarios addressed completely and automatically in previous works are thus the stair climbing scenarios with HRP-2 
+To compare our approach with previous contributions, we considered those that addressed the complete problem, and selected scenarios similar to ours, thus focusing on the HRP-2 robot.
+
+Table~\ref{tab:compprev} presents the results we obtained
 

### 2017-01-12 body/rbprm.tex
[RESULTS] Julien s comments

@@
 \subsection{The compromising reachability condition}
+\label{sec:scaling}
 The ideal shape $B^*, W^0 \subset B* \subset B^{\textrm{\it Suf}}$ that defines a necessary \textbf{and} sufficient condition for contact creation has no analytical definition.

### 2017-01-12 body/results.tex
[RESULTS] Julien s comments

@@
 
-\subsection{Influence of the parameters} \label{sec:influence}
-This subsection discusses the several factors that influence the outcome of our planner: the root scaling factor $s$, the choice of heuristics for
-contact generation, 
-%~ the desired value of the robustness parameter $b_0$, 
-and lastly, the discretization 
-of the guide path.
+\subsection{Role of the main parameters} \label{sec:influence}
+We discuss the several factors that influence the outcome of our planner: the root scaling factor $s$ (Section~\ref{sec:scaling}), the choice of heuristics for
+contact generation (Appendix~\ref{sec:heuristics}), and lastly, the discretization step for the guide path. The appropriate value for these parameters
+is computed empirically based on statistical analysis or trials and errors.
 
@@
 For several values of $s$, we generated $10 000$ configurations. 
-We then computed the rates of false positives (i.e. the configuration satisfies
-the reachability condition, but does not belong to \gls{$C_{Equil}^0$}) and false negatives (i.e. the configuration does not satisfy the 
-reachability condition, but belongs to \gls{$C_{Equil}^0$}).
-
+We then computed the sensitivity of the reachability condition (proportion of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}).
+Similarly we comuputed the specificity of the condition (proportion of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}).
 The obtained results for HRP-2 are shown in Table~\ref{tab:scale}, averaged over all scenes (except for the car egress: in this scenario, 
 statistical tests are not really conclusive since we are only interested in a small area of the environment).
-As it can be expected, the scaling results in a high increase of the false negatives, while the false positives decrease.
-For HRP-2 we decided to set $s^*=1.2$, which results in less than 3 \% of false positives.
+
+As it can be expected, the scaling results in a high decrease of the sensitivity, with an decrease of the specificity.
+For HRP-2 we decided to set $s^*=1.2$.
 %~ , and was also effective for the car egress scenario.
@@
 \begin{tabular}{c | c | c}
-   Value of $s$ &  False positive & False negative\\
+   Value of $s$ &  Sensitivity & Specificity\\
  \hline
-   1   & 24\% & 0 \%\\
-   1.1& 12\% & 4\% \\
-   1.15& 7\% & 6\%\\
-   1.2 & 3\% & 7.5\%\\
-   1.25& 2\% & 8.3\%\\
-   1.5 & 1\% & 9.5\%\\
+   1   & 76\% & 100 \%\\
+   1.1& 88\% & 96\% \\
+   1.15& 93\% & 94\%\\
+   1.2 & 97\% & 92.5\%\\
+   1.25& 98\% & 91.7\%\\
+   1.5 & 99\% & 90.5\%\\
  \end{tabular}
-\caption{Percentage of true and false positive in the reachability condition, depending on the scaling value of $W^0$.}
+\caption{Sensitivity and specificity values the reachability condition, depending on the scaling value of $W^0$.}
 \label{tab:scale}
@@
 \subsubsection{On the choice of heuristics:} \label{sec:heuristichoices}
-In our conference paper, the computed motions were only generated using the EFORT heuristic.
+In our conference paper~\citep{tonneauisrr15}, the computed motions were only generated using the EFORT heuristic.
 EFORT is designed for tasks requiring to exert important forces (such as pushing / pulling / climbing). 
@@
 For this reason, we prefer to use our manipulability-based heuristic for the legs of the robot, but we still
-use EFORT for the arms, that results in less contact repositioning.
+use EFORT for the arms, that results in less contact repositionings.
 
@@
 \subsubsection{Discretization of the guide path:} \label{sec:disc}
-Currently, to discretize the path, the user defines a fixed step size. The step size
+The discretization step is a user-defined, fixed parameter. The step
 has an influence on the output of the planner: if too large steps are taken,
 the planner may fail since we impose the constraint that only one contact change might occur
-between two consecutive steps. For instance, in most scenarios the torso of HRP-2 moves about 15 cm between two postures, but only 3 cm
-for the car egress scenario, where we had to lower the step size because of the many contact breaks induced by the extremely constraining geometry.
-For future work, we would like to automatically adapt the size of the discretization step to the complexity of the planning, for instance
-by reducing the step size when failure occurs.
+between two consecutive steps. On the other hand, a small step will not impact the success rate of the planner, 
+but may generate unnecessary states. In most scenarios the torso of HRP-2 moves about 15 cm between two postures, but only 3 cm
+for the car egress scenario to handle the geometry of the polaris.
+For future work, we would like to automatically adapt the size of the discretization step to the complexity of the environment.
 

### 2017-01-12 body/results.tex
updated results

@@
    Stairs (18) 	& 5 -- \textbf{6} --  18 		 & 13 --  \textbf{32} -- 329   	& 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{15} \\
-   %~ Standing (24)& 4 -- \textbf{451} --  985 & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 310 --  \textbf{1622} -- 3429 & \textbf{66}  \\
    Standing (24)& 65 -- \textbf{1086} --  5227   & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 371 --  \textbf{2257} -- 7671 & \textbf{94}  \\
-   %~ Car (61)& 1 -- \textbf{529} --  995 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6775 --  \textbf{20890} -- 52963 & \textbf{342} \\
-   Car TODO (61)& 145 -- \textbf{1055} --  15063 & 64 --  \textbf{85} -- 144   & 394 --  \textbf{735} -- 1422 & 3947 --  \textbf{13069} -- 36262 & 6919 --  \textbf{21416} -- 67031 & \textbf{351} \\
-   %~ Rubble (71)& 3 -- \textbf{524} --  993 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1400 --  \textbf{3058} -- 13126 & \textbf{43} \\
-   Rubble (71)& 151 -- \textbf{15220} --  125867 & 242 --  \textbf{511} -- 3480   & 233 --  \textbf{505} -- 4564 & 180 --  \textbf{414} -- 3518 & 1548 --  \textbf{18058} -- 13126 & \textbf{254} \\
-   %~ Race (112)& 1 -- \textbf{501} --  995 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2530 --  \textbf{3722} -- 10030 & \textbf{33}
-   Race (112)& 267 -- \textbf{966} --  2257 & 266 --  \textbf{449} -- 3956   & 824 --  \textbf{1061} -- 2130 & 666 --  \textbf{874} -- 1613 & 2797 --  \textbf{4187} -- 11292 & \textbf{37}
-   %Standing up & $70$ & 23 & 48\\
+   Car (86)& 320 -- \textbf{6971} --  44002 & 409 --  \textbf{1766} -- 14752   	& 297 -- \textbf{1187} -- 8483 & 3154 --  \textbf{15323} -- 165541 & 5834 --  \textbf{31391} -- 281000 & \textbf{365}\\
+   Rubble (82)& 37 -- \textbf{573} --  1685 & 583 --  \textbf{2714} -- 9459 & 491 --  \textbf{1971} -- 6273 & 269 --  \textbf{706} -- 3118 & 1811 --  \textbf{7195} -- 23241 & \textbf{86} \\
+   Race (134)& 14 -- \textbf{51} --  125 & 455 --  \textbf{1359} -- 21045   & 397 --  \textbf{923} -- 9924 & 228 --  \textbf{471} -- 5415 & 1436 --  \textbf{3343} -- 41446 & \textbf{25}
  \end{tabular}
@@
 \centering
-\begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | c}
-  &  Path planning success rate & Success rate & Kinematic failure & Equilibrium failure \\
+\begin{tabular}{ l | c}
+  &  Path planning success rate \\
  \hline
-   Steep stairs & 100\%  & 99.5\% & 0.1\% & 0.4\% \\
-   Standing up & 68\% & 87.8\% & 6.1\% & 6.1\% \\
-   Car egress TODO & 39\% & 77.0\% & 21.0\% &  2.0\% \\
-   Rubble & 74\% & 97.9\% & 0.1\% & 2.0\% \\
-   Obstacle race TODO & 58.0\% & 95.7\% & 1.8\% & 2.5\% \\
+   Steep stairs     	& 100\% \\
+   Standing up 			& 68\% 	\\
+   Car egress 			& 77\% 	\\
+   Rubble 				& 97\% 	\\
+   Obstacle race        & 88.0\% 	\\
  \end{tabular}
-\caption{ TODO AMBIGUITE POURCENTAGE Success rates for each scenario, rounded to the first decimal.
-The first value indicates the percentage of successful complete contact plannings for 1000 tests; The second value
-indicates the percentage of equilibrium feasible root configurations: considering each limb individually, indicates the percentage of root configurations of the guide path that
-led to a feasible contact. Kinematic failure is the percentage of contact generations that failed because no collision-free candidate was found. Equilibrium failure is the percentage of contact
-generations that failed because no candidate respected the static equilibrium condition.}
+\caption{ Percentage of successful complete contact planning rates for each scenario, rounded to the first decimal.}
+\label{tab:sucess_planning}
+\quad
+\end{table*}
+
+\begin{table*}
+\centering
+\begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | c}
+  &  Equilibrium success rate & Kinematic failure & Equilibrium failure \\
+ \hline
+   Steep stairs 	& 99.5\%  & 0.1\% 	& 0.4\% \\
+   Standing up 		& 87.8\%  & 6.1\% 	& 6.1\% \\
+   Car egress 		& 66.2\%  & 15.9\% 	& 17.9\% \\
+   Rubble 			& 97.54\% & 0.16\% 	& 2.3\% \\
+   Obstacle race 	& 92.4\%  & 0.15\% 	& 7.45\% \\
+ \end{tabular}
+\caption{Success rates obtained for the generation of static equilibrium contact configurations for each scenario, rounded to the first decimal. Column 1 indicates 
+indicates the rate of contact generation that suceeeded. In the cases where the generation fails, it can be
+either a kinematic issue (column 2), or because no contact configuration led to a static equilibrium configuration (column 3). Note that a failure in the contact generation
+is not equivalent to a failure of the contact planning algorithm.}
 \label{tab:requestpercent}
@@
 \subsubsection{Success rates:}
-Table~\ref{tab:requestpercent} summarizes the success rates obtained for each scenario.
+Table~\ref{tab:sucess_planning} summarizes the successful planning rates obtained for each scenario.
 Despite the complexity of the scenarios adressed, as well as the approximations made in our formulation, our planner succeeds in a large majority of cases.
-As expected, our planner does not succeed systematically, because of the approximations made in our formulation.
-The extreme situation of the car egress scenario provides the less satisfying results in terms of kinematic failure.
-This is not
-surprising considering the narrowness of the scene.
-Regarding computation times, our approach provides a satisfying compromise between completeness and efficiency.
+
+Table~\ref{tab:requestpercent} presents, for each scenario, the rate of contact generation that succeeded.
+Note that a failure in contact generation for a root configuration is not equivalent to a failure in the contact plan.
+It simply means that another limb was tested for contact generation for the same root configuration.
+As expected, the more constrained scenario, the car egress one, provides the less satisfying results, despite the high success rate of the planner.
+
+These results confirm that our approach provides a satisfying compromise between completeness and efficiency, thus allowing to consider online planning
+while controlling the robot.
 Indeed, when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan with a reasonable chance of success.
-The most efficient (and immediate) approach is probably to launch in parallel several instances of the planner (our current implementation is single threaded) and to apply any successful result.
+The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible is probably to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.
 

### 2017-01-13 body/conclusion.tex
[Discussion]

--- 2017-01-09 body/conclusion.tex
@@
- \section{Discussion} 
+ \section{Discussion: validity and interest of our contact planner} 
 \label{sec:discussion}
 
-\subsection{validity and interest of the approximations}
+\subsection{Accuracy and performance gain of the reachability condition}
 
-TODO equilibirum as contact reachability
+As demonstrated in the results section, the main interest of our method is time performance. Time performance is critical to consider using a multi-contact planner for 
+online applications with the robot, a goal not satisfied by any of the previous contributions.
+Our method addresses highly constrained environments while improving the search time by orders of magnitude.
+This high performance is reached at the cost of some approximations we discuss here.
+%~ In previous approaches, we identified 2 sources of high computational cost.
+
+The first approximation is the verification of \contactreachability\ ($\mathbf{q}^0 \in C_{Contact}^0$).
+Our \textit{reachability condition} ($\mathbf{q}^0 \in C_{Reach}^0$) is computationally efficient and provides an accurate approximation of $C_{Contact}^0$ (Section~\ref{sec:scaling}).
+This is demonstrated by the second column of Table~\ref{tab:requestpercent}, and illustrated by Figure~\ref{fig:dedefeas}.
+Indeed, in the large majority of cases, (84\% in the worst case), we are able to find a contact configuration for any configuration in $C_{Reach}^0$.
+
+%~ These results clearly demonstrate that the reachability condition, presented in Section~\ref{sec:scaling}, provides an accurate representation of the 
+%~ contact we combine only-necessary and only-sufficient conditions which are trivial to estimate.
+Only-necessary conditions are appealing because they preserve the completeness of the search, while reducing
+%~ the search space: they provide an outer approximation of $C_{Contact}^0$.
+On the other hand, only-sufficient conditions provide the guarantee that any configuration that satisfies them is indeed \gls{contact reachable}: they provide an inner approximation of $C_{Contact}^0$.
+
+The main source of computational cost identified in previous works is the verification of \equilibriumfeasibility. 
+The strong assumption of our work is that for the class of problems we consider \contactreachability\ implies \equilibriumfeasibility.
+Our scenarios show that the  assumption is verified in the majority of cases when at least one contact surface is \textit{quasi flat}, that is when
+the friction cone of the contact surface contains the direction opposite to the gravity.
+Figure~\ref{fig:dedefeas} illustrate this observation, demonstrated empirically by the third column of Table~\ref{tab:requestpercent}. In the worst case, in our experiment
+the assumption was verified for 82\% of the total amount of trials.
+
+In the example of \cite{Bouyarmane2009}, the verification of \equilibriumfeasibility implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$, requiring
+several minutes of planning. Our method, in comparison, takes from a few milliseconds to several seconds.
+
+These results clearly justify the pragmatic decisions of this paper. What is lost in completeness is largely compensated by the computation times we obtain.
+
+\subsection{Addressing $\mathcal{P}_3$}
+This paper focuses on the contact planning aspect of multi-contact locomotion. While it does not address directly the continuous motion generation,
+we had to prove that the plans produced by our planner are useful and can be executed by the considered robots. The problem $\mathcal{P}_3$ remains open,
+and most existing solutions do not consider collisions with the environment during the planning, making it hard to validate constrained problems such as the polaris scenario
+or the table egress in \cite{DBLP:conf/iser/EscandeKMG08}.
+The companion video shows the results we obtained in simulation and on the real HRP-2 robot. Despite the fact that our method does not consider joint torque limits in the planning
+phase, the robot is able to execute our plans, achieving to demonstrate the interest of our framework.
 
@@
 \end{figure}
-
-Figure~\ref{fig:dedefeas} gives an insight into the difference between the three conditions, depicting $C_{Equil}^0$, $C_{Contact}^0$ and $C_{Reach}^0$, which are the sets of \gls{equilibrium feasible}, \gls{contact reachable} and \textit{reachable} root configurations, respectively.
-
-TODO blah regarding contact plan generation, and ignoring the mid contact conf
-The first constraint is less restrictive than the one proposed by \citeauthor{DBLP:conf/iser/EscandeKMG08}, because our method
-allows to break and create one contact
 

### 2017-01-13 body/results.tex
[Discussion]

@@
 Table~\ref{tab:sucess_planning} summarizes the successful planning rates obtained for each scenario.
-Despite the complexity of the scenarios adressed, as well as the approximations made in our formulation, our planner succeeds in a large majority of cases.
+Despite the complexity of the scenarios adressed, as well as the approximations made in our formulation, our planner succeeds in the large majority of cases.
 
@@
 These results confirm that our approach provides a satisfying compromise between completeness and efficiency, thus allowing to consider online planning
-while controlling the robot.
-Indeed, when the contact generation fails, it does so rapidly, which allows us to rapidly re-plan with a reasonable chance of success.
+while controlling the robot. Indeed, when the contact planning fails, it does so rapidly, which allows us to rapidly re-plan with a reasonable chance of success.
 The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible is probably to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.

### 2017-01-13 body/conclusion.tex
conclusion

@@
 
-In the example of \cite{Bouyarmane2009}, the verification of \equilibriumfeasibility implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$, requiring
+In the example of \cite{Bouyarmane2009}, the verification of \equilibriumfeasibility\ implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$, requiring
 several minutes of planning. Our method, in comparison, takes from a few milliseconds to several seconds.
@@
 Depending on a parametrization, we can obtain $C_{Contact}^0 \subset C_{Reach}^0$.
-Considering the configurations around the top obstacle, we can observe a dramatic
-divergence between  $C_{Equil}^0$  and $C_{Contact}^0$ when the problem is not \textit{\gls{cluttered}}.}
+Considering the configurations around the top obstacle, we can observe a similarity between  $C_{Equil}^0$  and $C_{Contact}^0$ when the reachable workspace of the legs includes \textit{quasi-flat} surfaces.}
 		   \label{fig:dedefeas}
@@
 
-In this paper we consider the \gls{cluttered} contact planning problem, formulated as two sub-problems that we address sequentially.
+In this paper we consider the multi contact planning problem, formulated as two sub-problems that we address sequentially.
 The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
@@
 
-Our contribution to \Pa is the introduction of a low-dimensional space $C_{reach}$, an approximation of the space of \gls{equilibrium feasible} root configurations.
-Thanks to the computationally efficient verification of the \gls{reachc}, we are able to solve \Pa much faster than previous approaches.
+Our contribution to \Pa is the introduction of a low-dimensional space $C_{reach}^0$, an approximation of the space of \gls{equilibrium feasible} root configurations.
+$C_{reach}^0$ can be efficiently sampled and has a low-dimension. For these reasons we are able to solve \Pa much faster than previous approaches.
 
@@
 %
-\textbf{Regarding generality}, the \gls{reachc}, coupled with an approach based on limb decomposition, 
-allows the method to address arbitrary multiped robots in \gls{cluttered} problems. The only pre-requisite is the specification 
-of the volumes $W^0$.
+\textbf{Regarding generality}, the \textit{reachability condition}, coupled with an approach based on limb decomposition, 
+allows the method to address automatically arbitrary multiped robots.
 %
@@
 %
-\textbf{Regarding the quality of the paths}, the \gls{reachc} allows us to compute
-\gls{equilibrium feasible} paths in all the presented scenarios, with low rejection rates.
+\textbf{Regarding the quality of the paths}, we are able to compute
+\gls{equilibrium feasible} paths in all the presented scenarios, with a high success rates.
 As for \cite{Bouyarmane2009}, failures can still occur, due to the approximate condition used to compute the guide path.
@@
 
-Another limitation of the method is that it currently only applies to \gls{cluttered} problems.
-However, it should first be noted that many problems belong to this class: for instance all the problems
-at the DRC were \gls{cluttered}. Our method is thus already useful for many cases.
-One way of extending its range of application, that we 
-consider for future work, is to include the equilibrium criterion when solving $\mathcal{P}_1$.
+Our method applies to any \textit{quasi-flat} scenario, where at least one contact friction cone contains
+the direction opposed to the gravity. This class of scenarios include all the problems proposed at the DARPA Robotics challenge.
+One way of further extending its range of application, that we consider for future work, is to include the equilibrium criterion when solving $\mathcal{P}_1$.
 Considering the set of obstacles intersecting with the reachable workspace for a given root configuration as candidates surfaces, we can use them to verify the equilibrium criterion.
@@
 
-Regarding the interpolation between the contact sequences ($\mathcal{P}_3$), we have already obtained some success for some of the computed sequences~\citep{Carpentier2016}, but additional
-work is required on the planning side to obtain a seamless workflow. To achieve this, we are currently working on the notion of transition certificate, i.e. formulating
-conditions that guarantee that the interpolation between two contact configurations is dynamically feasible.
+While we have exhibited complete multi contact locomotion obtained with our contact planner, our main concern for future work is to address the interpolation
+between contact sequences ($\mathcal{P}_3$), not addressed in this work, which remains an open issue in really constrained scenarios.
+Adressing $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem during the interpolation,	but also improving
+our plans with transition certificates, i.e. guarantees that the interpolation between two contact configurations is feasible.
 %~ Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
@@
 %~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
-A last limitation of our method is that only static equilibrium configurations are considered for contact planning.
-We aim at performing kinodynamic planning to overcome this limitation. We believe that the most promising direction in this regard is to integrate
-the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
-Addressing these two last issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.
+Finally, we aim at performing kinodynamic planning to remove the constraint that configurations in the contact plan should be in static equilibrium.
+We believe that the most promising direction in this regard is to integrate the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
+Addressing these two issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.

### 2017-01-14 body/abstract.tex
abstract

@@
-We describe  a method to plan contacts allowing any multiped robot, including a humanoid, to achieve complex motions: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
+We present a framework capable of producing contact plans describing complex multiped motions: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
+The need for such a planner was highlighted at the Darpa Robotics Challenge, where required multi contact behaviors
+could not be demonstrated.
 
-The Darpa Robotics Challenge recognized the lack  of an automatic acyclic contact planner to be a major issue.
-We adress by introducing the reachability condition which, informally, verifies that a robot root remains in a bounded volume above the environment surfaces, close enough to allow contact creation and far enough to allow collision free motion of the limbs.
-We achieve a complex motion by decomposing the hard contact planning into simpler sub-problems: first, to plan a guide path for the root without considering the whole-body configuration; second, to generate a discrete sequence of whole-body configurations in static equilibrium along this path.
-The reachability condition turns the high-dimensional computation of the guide into a collision checking problem.
-Then a deterministic contact selection algorithm tackles the combinatorial issue of  generating of a discrete sequence along the guide path.
-Several innovations make it computationally efficient: a criterion for verifying static equilibrium, and a set of heuristics used to enforce desirable properties on the configuration.
+Indeed, current planners suffer from prohibitive computation times, explained by the difficulty
+of generating contact postures: the feasible space (the contact manifold) is of high dimension,
+and impossible to sample.
 
-Our approach results from the pragmatic choice of favoring efficiency over exhaustiveness, justified empirically: in a few seconds, with satisfying success rates, we generate complex contact plans for various scenarios and robots, namely HRP-2, HyQ, and a dexterous hand.
+We tackle this issue with our novel key idea, the reachability condition. This condition allows us to define
+a geometric approximation of the contact manifold, which is of low dimension and can be sampled efficiently.
+Informally, the condition verifies that the root configuration of a robot is “close, but not too close” from obstacles: close to allow contact creation, not too close to avoid collision.
+Thanks to this condition we decompose the hard contact planning problem into
+simpler sub-problems: first, to plan a guide path for the root without considering the whole-body configuration, using a sampling
+based algorithm; then, to generate a discrete sequence of whole-body configurations in static equilibrium along this path, using a new deterministic
+contact selection algorithm.
+
+Our approach results from the pragmatic choice of favoring efficiency over completeness, justified empirically: in a
+few seconds, with satisfying success rates, we generate complex contact plans for various scenarios and robots, namely
+HRP-2, HyQ, and a dexterous hand. These plans are then validated in dynamic simulations, or on the real HRP-2 robot, thus demonstrating 
+the first interactive multi-contact planner.

### 2017-01-16 body/intro.tex
updates

@@
 To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorial of the contact selection problem.
-First we introduce a low-dimensional space, called the \textit{contact-feasible} space, that can be sampled and mapped efficiently to the contact manifold.
-Then, given a path computed in the \textit{contact-feasible} space, we propose a deterministic algorithm to generate a contact sequence along the path.
+First we introduce a low-dimensional space, called the \textit{contact reachable} space, that can be sampled and mapped efficiently to the contact manifold.
+Then, given a path computed in the \textit{contact reachable} space, we propose a deterministic algorithm to generate a contact sequence along the path.
 This decoupling presents drawbacks and advantages discussed in previous related literature, which we summarize in the following.
@@
 This approach is suboptimal because the quality of the motion depends on the quality of the guide path.
-\citeauthor{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
-Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
+\citeauthor{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, that is that for each root configuration in the guide path, there exists a joint configuration that results in static equilibrium.
+They did it by extending key frames of the path into whole-body configurations using a generalized inverse kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
 Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path.
@@
 For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
-used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
+used black-box physics simulators to perform Model Predictive Control for an avatar, and obtained dynamically-consistent motions
 at \gls{interactive} frame rates (we say that a method is interactive when the computation time for planning is less than the

### 2017-01-16 body/overview.tex
updates

@@
 
+\subsection{Generating a discrete sequence of contact configurations --- \Pb}
 The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--$\mathcal{P}_2$).
@@
 The task efficiency is measured using the Extended FORce Transmission ratio (EFORT), proposed by ~\cite{Tonneau:2014:TEC:2619648.2619652}.
-
-\subsection{Generating a discrete sequence of contact configurations --- \Pb}
 

### 2017-01-16 body/rbprm.tex
updates

@@
 \textbf{sufficient condition:}
-Similarly we can define an inner approximation \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ $C_{Contact}^0$} by considering a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces in contact.
+Similarly we can define an inner approximation \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ $C_{Contact}^0$} by considering a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces.
 \begin{equation}

### 2017-01-16 body/results.tex
updates

@@
 
-As it can be expected, the scaling results in a high decrease of the sensitivity, with an decrease of the specificity.
+As it can be expected, the scaling results in a high decrease of the sensitivity, with a decrease of the specificity.
 For HRP-2 we decided to set $s^*=1.2$.
@@
 
-For HyQ, there is a more uniform repartition of the times for the rubble scenario. For the obstacle scenario,
-we observe a large increase of the time spent performing collision checking and inverse kinematics. This is explained
-by the complexity of crossing the bridge: the robot has to stand high and create contacts close to each other because
-the bridge is narrow. In many cases this results in collisions, thus a large number of contacts have to be evaluated.
+On the other hand for HyQ most of the time is spent testing the static equilibrium of the candidates configurations..
 

### 2017-01-16 body/contact.tex
updates
Jospeh s comments

@@
 In the second phase, we compute a discrete sequence of static equilibrium configurations $\mathbf{Q}^{\overline{0}}$ given a root path
-$\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Reach}^0 \simeq C_{Equil}^0$}. This contact planner is built on top of
+$\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Reach}^0$}. This contact planner is built on top of
 a contact generator, used to generate static equilibrium configurations. We first describe the generator, before detailing the contact planning algorithm.
@@
 (b): Entries of the limb samples database (with $N = 4$).
-(c): With a proximity query on the octree database, configurations too far from obstacles are eliminated.
+(c): With a proximity query between the octree database and the obstacles, configurations too far from obstacles are discarded.
 (d): The best candidate according to a user-defined heuristic $h$ is chosen.
@@
 Contacts are created using a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest.
-If the contact creation was not successful, the limb is pushed on top of the queue, and will only be tried again after the others.
+If the contact creation does not succeeds, the limb is pushed on top of the queue, and will only be tried again after the others.
 %~ \begin{enumerate}

### 2017-01-16 body/overview.tex
Jospeh s comments

@@
 \begin{equation}
-  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k \in C^k_{jl}, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\}
+  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k \in C^k_{j.lim}, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\}
 \end{equation}
 %~ \adnote{in this definition you don't account for joint limits}
-where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{jl}$ is the space
+where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{j.lim}$ is the space
 of admissible limb joint configurations. We also define \mbox{$W = \bigcup_{k=1}^{l}W^k$}, and

### 2017-01-16 body/rbprm.tex
Jospeh s comments

@@
 \label{sec:scaling}
-The ideal shape $B^*, W^0 \subset B* \subset B^{\textrm{\it Suf}}$ that defines a necessary \textbf{and} sufficient condition for contact creation has no analytical definition.
+The ideal shape $B^*, W^0 \subset B* \subset B^{\textrm{\it Suf}}$ that defines a necessary \textbf{and} sufficient condition for contact creation has no explicit definition.
 Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.

### 2017-01-16 body/appendix_hpp.tex
florent s comment

--- 2015-12-08 body/appendix_hpp.tex
@@
 \label{app:hpp}
-Our planner is implemented using the Humanoid Path Planner (HPP) software.
+Our planner is implemented using the Humanoid Path Planner (HPP) software, introduced in~\cite{7759083}.
 HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS, that can be found at \url{http://projects.laas.fr/gepetto/index.php/Software/Main}.
@@
 Our implementation of the planner is also open source, and can be found at \url{https://github.com/stonneau/hpp-rbprm}.
-Contrary to the mature HPP, at the time of submitting this article,
-our planner is not officialy released, mostly for lack of a complete documentation and a tutorial.
-
-However a reader familiar with HPP should be able to use the library.
-Our goal is to make our code useful for the community, and we are working on releasing a stable 
-version of our planner as soon as possible.

### 2017-01-16 body/conclusion.tex
florent s comment

@@
 This paper focuses on the contact planning aspect of multi-contact locomotion. While it does not address directly the continuous motion generation,
-we had to prove that the plans produced by our planner are useful and can be executed by the considered robots. The problem $\mathcal{P}_3$ remains open,
+we had to prove that the plans produced by our planner are useful and can be executed by the considered robots. Problem $\mathcal{P}_3$ remains open,
 and most existing solutions do not consider collisions with the environment during the planning, making it hard to validate constrained problems such as the polaris scenario

### 2017-01-16 body/contact.tex
florent s comment

@@
 Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
-This process is repeated until either a solution for Eq.~\ref{eq:contact_gen} is found, or the generator fails, according to a user-defined maximum number of trials.
+This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.
 We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
@@
 \item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
-\item At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
+\item At runtime, when contact creation is required, intersect the octree and the environment to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
 \item Use a user-defined heuristic $h$ to sort $S$;

### 2017-01-16 body/intro.tex
florent s comment

--- 2017-01-12 body/intro.tex
@@
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contacts and the potential paths.
-This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments.
+\citeauthor{Bretl:2006:MPM:1124573.1124585}'s seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments.
 Following it, seve\-ral papers have applied this approach in specific situations, limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.
@@
 Two main directions have been explored.
-\textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}.
+\textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}.
 A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
@@
 can result in failures.
-However, we demonstrate empirically its interest: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
+However, we demonstrate empirically its interest: the high success rate and low computation times allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
 

### 2017-01-16 body/overview.tex
florent s comment

@@
 The dimension of the path is equal to the number of degrees of freedom (DoFs) of the root of the robot.
-Similarly to previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: the root configurations must be extensible into full-body static equilibrium configurations.
-Previous work verifies \textit{equilibrium feasibility} by explicitly computing such a configuration.
+Similarly to previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: there must exist a joint configuration that results in static equilibrium for each root configuration.
+Previous work verifiy \textit{equilibrium feasibility} by explicitly computing such a configuration.
 We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact rechability}, illustrated in the following.
@@
 The root has $r \geq 6$ DoFs: for instance, HRP-2 has two extra DoFs in the torso, such that we have $r=8$.
-Therefore $R$ is fully described by a configuration $\mathbf{q} \in  \mathbb{R}^{r+n}$.
+Therefore $R$ is fully described by a configuration $\mathbf{q} \in  \mathbb{R}^{r+n}$, with $n$ the number of joint DoFs.
 $\mathbf{q}$ is decomposed as follows:
@@
 %~ \adnote{in this definition you don't account for joint limits}
-where $\mathbf{p}^k$ denotes the end-effector position of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{j.lim}$ is the space
+where $\mathbf{p}^k$ denotes the end-effector position (in the root frame) of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{j.lim}$ is the space
 of admissible limb joint configurations. We also define \mbox{$W = \bigcup_{k=1}^{l}W^k$}, and
@@
 \medskip
-Finally we define some relevant of the configuration space $C$.
+Finally we define some relevant subsets of the configuration space $C$.
 $C_{Contact}$ is the set of whole body-configurations in contact and collision-free.

### 2017-01-16 body/rbprm.tex
florent s comment

@@
 \textbf{necessary conditions:}
-For a contact to be possible, an obstacle $O_i \in O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot (Figure~\ref{fig:contact_gen}--1).
+For a contact to be possible, an obstacle $O_i \subset O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot (Figure~\ref{fig:contact_gen}--1).
 Also the torso of the robot $W^0(\mathbf{q}^{0})$ must necessarily be collision-free.
@@
 The only significant change is to replace the collision checking with the \textit{reachability condition}.
-Our current implementation of these modifications is based on the Bi-RRT planner \citep{770022} provided by the HPP software.
+Our current implementation of these modifications is based on the Bi-RRT planner \citep{770022} provided by the HPP software~\citep{7759083}.
 

### 2017-01-16 body/results.tex
florent s comment

@@
 We then computed the sensitivity of the reachability condition (proportion of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}).
-Similarly we comuputed the specificity of the condition (proportion of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}).
+Similarly we computed the specificity of the condition (proportion of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}).
 The obtained results for HRP-2 are shown in Table~\ref{tab:scale}, averaged over all scenes (except for the car egress: in this scenario, 
@@
 These results confirm that our approach provides a satisfying compromise between completeness and efficiency, thus allowing to consider online planning
-while controlling the robot. Indeed, when the contact planning fails, it does so rapidly, which allows us to rapidly re-plan with a reasonable chance of success.
+while controlling the robot. Indeed, when the contact planning fails, it fails rapidly. This allows us to rapidly re-plan with a reasonable chance of success.
 The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible is probably to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.
@@
 To validate our plans, we had to implement our own interpolation method to address $\mathcal{P}_3$.
-We refer the curious reader to \citep{Carpentier2016} for the implementation details of our method. 
+We refer the reader to \citep{Carpentier2016} for the implementation details of our method. 
 This motion is in turn validated in our dynamic simulator~\citep{DelPrete2015b}. The stair climbing scenario was validated on the real HRP-2 robot \citep{Carpentier2016}.

### 2017-01-16 body/abstract.tex
Review abstract and results

@@
-We present a framework capable of producing contact plans describing complex multiped motions: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
-The need for such a planner was highlighted at the Darpa Robotics Challenge, where required multi contact behaviors
+% !TEX root =  ../main.tex
+\adnote{What about replacing 'multiped' with 'legged' everywhere?}
+\repadp{We present a contact planning framework for complex legged locomotion tasks:}{We present a framework capable of producing contact plans describing complex multiped motions}: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
+The need for such a planner was highlighted at the Darpa Robotics Challenge, where \deladp{required} multi-contact behaviors
 could not be demonstrated.
 
-Indeed, current planners suffer from prohibitive computation times, explained by the difficulty
+Indeed, current planners suffer from prohibitive computation times, \repadp{due to}{explained by} the difficulty
 of generating contact postures: the feasible space (the contact manifold) is of high dimension,
@@
 
-We tackle this issue with our novel key idea, the reachability condition. This condition allows us to define
+We tackle this issue with our novel \deladp{key} idea, the reachability condition. This condition allows us to define
 a geometric approximation of the contact manifold, which is of low dimension and can be sampled efficiently.
-Informally, the condition verifies that the root configuration of a robot is “close, but not too close” from obstacles: close to allow contact creation, not too close to avoid collision.
-Thanks to this condition we decompose the hard contact planning problem into
-simpler sub-problems: first, to plan a guide path for the root without considering the whole-body configuration, using a sampling
-based algorithm; then, to generate a discrete sequence of whole-body configurations in static equilibrium along this path, using a new deterministic
-contact selection algorithm.
+Informally, the condition verifies that the root configuration of a robot is €œclose, but not too close to obstacles: close to allow contact creation, not too close to avoid collision.
+Thanks to this condition we decompose the hard contact planning problem into two
+simpler sub-problems: first, planning a guide path for the robot root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a new deterministic
+contact-selection algorithm.
 
-Our approach results from the pragmatic choice of favoring efficiency over completeness, justified empirically: in a
+Our approach results from the pragmatic choice of favoring efficiency over completeness, \repadp{which we justify empirically}{justified empirically}: in a
 few seconds, with satisfying success rates, we generate complex contact plans for various scenarios and robots, namely
-HRP-2, HyQ, and a dexterous hand. These plans are then validated in dynamic simulations, or on the real HRP-2 robot, thus demonstrating 
+HRP-2, HyQ, and a dexterous hand. These plans are then validated either in dynamic simulations, or on the real HRP-2 robot, thus demonstrating 
 the first interactive multi-contact planner.

### 2017-01-16 body/results.tex
Review abstract and results

@@
 \label{sec:results}
-In this Section we present some of the results obtained with our planner. The complete sequences computed are shown in the companion video.
+In this \repadp{section}{Section} we present some of the results obtained with our planner. The complete sequences \deladp{computed} are shown in the companion video.
 Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
@@
 
-At the end of this Section, we discuss the role of the parameters of our framework on the output. We then provide the \textit{interactive} computation times obtained in each case.
+At the end of this \repadp{section}{Section}, we discuss the role of the parameters of our framework\deladp{ on the output}.
+We then provide the \textit{interactive} computation times obtained in each case.
 We also compare the times obtained with HRP-2 with respect to previous works.
 
-These scenarios complete the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
+These scenarios \repadp{complement}{complete} the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
 %~ (Figure~\ref{fig:robots_old}).
@@
 In all the scenarios considered, the formulation of the problem is always the same:
-a start and goal root configurations are provided as an input of the scenario.
+a start and goal root configurations are provided as \deladp{an} input \deladp{of the scenario}.
 The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
@@
 
-The goal is to climb three 15-cm high steps. This height requires HRP-2 to use a ramp to perform the task.
+The goal is to climb three 15-cm high steps. This height requires HRP-2 to use a ramp to perform the task. \adnote{It is not clear why, so either explain why, or remove the comment.}
 
@@
 \subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing}):}
-From a bent configuration, a standing-up motion is computed in a constraining environment.
-The resulting motion involves using a wall as support, and climbing a 25-cm high step.
+From a bent configuration, \repadp{the robot has to stand up}{a standing-up motion is computed in a constraining environment.
+The resulting motion involves} using a wall as support, and climbing a 25-cm high step.
 
@@
 \subsubsection{HRP-2 -- Polaris car egress (Figure~\ref{fig:car}):}
-This scenario is inspired from the Darpa challenge car egress scenario (\url{http://cpc.cx/edH}). HRP-2 has 
-to find the contact sequence that allows it to step out of the Polaris car.
+\repadp{In this scenario---inspired from the DRC car egress (\url{http://cpc.cx/edH})---HRP-2 has}{This scenario is inspired from the DARPA-Robotics-Challenge car egress scenario (\url{http://cpc.cx/edH}).
+HRP-2 has to find the contact sequence that allows it} to step out of the Polaris car.
 
@@
 
-\subsubsection{HyQ -- Darpa-style rubble (Figure~\ref{fig:darpa})}
-The quadruped robot is given the task to cross a rubble composed of bricks rotated at different angles and directions.
+\subsubsection{HyQ -- DRC-style rubble (Figure~\ref{fig:darpa})}
+The quadruped robot \repadp{must}{is given the task to} cross a rubble composed of bricks rotated at different angles and directions.
 
@@
   \caption{
-           Robust crossing of rubbles by HyQ ($b_0 > 20$). }
+           Robust crossing of rubbles by HyQ \deladp{($b_0 > 20$)}. }
 		   \label{fig:darpa}
@@
 \subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs}):}
-In this long scene, HyQ is first required to cross a 55-cm large hole; then, to cross a narrow ``bridge'', only 25-cm large.
+In this long scene, HyQ \repadp{has to cross a 55-cm large hole, followed by}{is first required to cross a 55-cm large hole; then, to cross} a narrow ``bridge'', only 25-cm large.
 
@@
   \caption{
-           Crossing a hole contact sequence for HyQ ($b_0 > 4$). }
+           Crossing a hole contact sequence for HyQ \deladp{($b_0 > 4$)}. }
 		   \label{fig:HyQ_obs}
@@
 \subsubsection{HRP-2 -- Path re-planning (Figure~\ref{fig:re-planning}):}
-In this long scene, HRP-2 plans a path through several obstacles. the scene is edited during the execution of the motion: a stair is added,
+In this long scene, HRP-2 plans a path through several obstacles. The scene is edited during the execution of the motion: a stair is added,
 some stepping stones are removed, and part of the final staircase is deleted. All these modifications require re-planning.
@@
 \subsection{Role of the main parameters} \label{sec:influence}
-We discuss the several factors that influence the outcome of our planner: the root scaling factor $s$ (Section~\ref{sec:scaling}), the choice of heuristics for
+We discuss the several factors \adnote{Either use 'the' or use 'several'} that influence the outcome of our planner: the root scaling factor $s$ (Section~\ref{sec:scaling}), the \deladp{choice of} heuristics for
 contact generation (Appendix~\ref{sec:heuristics}), and lastly, the discretization step for the guide path. The appropriate value for these parameters
-is computed empirically based on statistical analysis or trials and errors.
+is computed empirically based on statistical analysis \adnote{Are you sure you wanna use the word 'statistical'?} or trials and errors.
 
 \subsubsection{Choosing the scaling factor $s$:} \label{sec:params}
-To find a convenient value $s^*$ of $s$, we proceeded as follows. 
+\deladp{To find a convenient value $s^*$ of $s$, we proceeded as follows. }
 For several values of $s$, we generated $10 000$ configurations. 
-We then computed the sensitivity of the reachability condition (proportion of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}).
-Similarly we computed the specificity of the condition (proportion of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}).
+We then computed the sensitivity of the reachability condition (\repadp{percentage}{proportion} of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}).
+Similarly we computed the specificity of the condition (\repadp{percentage}{proportion} of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}).
 The obtained results for HRP-2 are shown in Table~\ref{tab:scale}, averaged over all scenes (except for the car egress: in this scenario, 
@@
 
-As it can be expected, the scaling results in a high decrease of the sensitivity, with a decrease of the specificity.
+As it can be expected, the scaling results in a high \repadp{increase}{decrease} of the sensitivity, with a decrease of the specificity.
 For HRP-2 we decided to set $s^*=1.2$.
@@
  \end{tabular}
-\caption{Sensitivity and specificity values the reachability condition, depending on the scaling value of $W^0$.}
+\caption{Sensitivity and specificity values \addadp{of} the reachability condition, depending on the scaling value \addadp{$s$} of $W^0$.}
 \label{tab:scale}
@@
 
-\subsubsection{On the choice of heuristics:} \label{sec:heuristichoices}
-In our conference paper~\citep{tonneauisrr15}, the computed motions were only generated using the EFORT heuristic.
+\subsubsection{\repadp{Choosing the heuristics}{On the choice of heuristics}:} \label{sec:heuristichoices}
+In our conference paper~\citep{tonneauisrr15}, the computed motions were \deladp{only} generated using the EFORT heuristic.
 EFORT is designed for tasks requiring to exert important forces (such as pushing / pulling / climbing). 
-Regarding locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
+\repadp{In}{Regarding} locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
 configurations close to singularities (and joint limits). While this does not significantly impact
-the generation of the plan, the resulting interpolation on the real robot turned out to be harder.
+the generation of the plan, the resulting interpolation \deladp{on the real robot} turned out to be harder.
 For this reason, we prefer to use our manipulability-based heuristic for the legs of the robot, but we still
-use EFORT for the arms, that results in less contact repositionings.
+use EFORT for the arms, which results in less contact repositionings.
 
@@
 For future work, we would like to automatically adapt the size of the discretization step to the complexity of the environment.
+\adnote{Maybe you should say which discretization step you used.}
 
 \subsection{Performance analysis} \label{sec:perf}
-To analyze performance, for each considered scenario, we ran the planner 1000 times.
-We measured the computation time spent in each aspect of the algorithm, and analyzed their success rate.
+To analyze performance, we ran the planner 1000 times for each considered scenario.
+We measured the computation time spent in each \repadp{part}{aspect} of the algorithm, and analyzed their success rate.
 
@@
 \subsubsection{Computation times:}
-Table~\ref{tab:requestime} summarizes the performance measurements obtained, in terms of computation times.
+Table~\ref{tab:requestime} summarizes the \deladp{performance measurements obtained, in terms of} computation times.
 
-Considering the repartition of the computation time, for HRP-2, most of the time is spent performing inverse kinematics.
+\repadp{F}{Considering the repartition of the computation time, f}or HRP-2, most of the time \repadp{was}{is} spent performing inverse kinematics.
 This is not surprising considering the number of calls to the methods: IK projection is used intensively to maintain contact continuity between two postures; 
@@
 
-On the other hand for HyQ most of the time is spent testing the static equilibrium of the candidates configurations..
+On the other hand for HyQ most of the time is spent testing the static equilibrium of the candidate configurations.
 
 In all scenarios, one can observe that the average computation time for one single step is largely below one second,
-thus allowing to consider \gls{interactive} applications. This would allow to consider online autonomous planning of the robot motion.
+thus allowing to consider \gls{interactive} applications\repadp{ and}{. This would allow to consider} online autonomous planning of the robot motion.
 
 \subsubsection{Success rates:}
-Table~\ref{tab:sucess_planning} summarizes the successful planning rates obtained for each scenario.
-Despite the complexity of the scenarios adressed, as well as the approximations made in our formulation, our planner succeeds in the large majority of cases.
+Table~\ref{tab:sucess_planning} summarizes the successful planning rates\deladp{ obtained for each scenario}.
+Despite the complexity of the scenarios addressed, as well as the approximations made in our formulation, our planner \repadp{succeeded}{succeeds} in the large majority of cases.
 
-Table~\ref{tab:requestpercent} presents, for each scenario, the rate of contact generation that succeeded.
+Table~\ref{tab:requestpercent} presents\deladp{, for each scenario,} the rate of \addadp{successful} contact generation\deladp{ that succeeded}.
 Note that a failure in contact generation for a root configuration is not equivalent to a failure in the contact plan.
 It simply means that another limb was tested for contact generation for the same root configuration.
-As expected, the more constrained scenario, the car egress one, provides the less satisfying results, despite the high success rate of the planner.
+As expected, the more constrained scenario, the car egress \deladp{one}, provides the \repadp{worst}{less satisfying} results, despite the high success rate of the planner.
 
@@
 while controlling the robot. Indeed, when the contact planning fails, it fails rapidly. This allows us to rapidly re-plan with a reasonable chance of success.
-The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible is probably to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.
+The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible \repadp{would be}{is probably} to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.
 

### 2017-01-16 body/conclusion.tex
Review discussion and conclusion

@@
+% !TEX root =  ../main.tex
  \section{Discussion: validity and interest of our contact planner} 
@@
 
-As demonstrated in the results section, the main interest of our method is time performance. Time performance is critical to consider using a multi-contact planner for 
+As demonstrated in the results section, the main interest of our method is time performance. Time performance is critical \deladp{to consider using a multi-contact planner} for 
 online applications with the robot, a goal not satisfied by any of the previous contributions.
 Our method addresses highly constrained environments while improving the search time by orders of magnitude.
-This high performance is reached at the cost of some approximations we discuss here.
+This high performance is reached at the cost of some approximations that we discuss here.
 %~ In previous approaches, we identified 2 sources of high computational cost.
@@
 This is demonstrated by the second column of Table~\ref{tab:requestpercent}, and illustrated by Figure~\ref{fig:dedefeas}.
-Indeed, in the large majority of cases, (84\% in the worst case), we are able to find a contact configuration for any configuration in $C_{Reach}^0$.
+Indeed, in the large majority of cases, (84\% in the worst case \adnote{I do not get where this 84 comes from}), we are able to find a contact configuration for any configuration in $C_{Reach}^0$.
 
@@
 The strong assumption of our work is that for the class of problems we consider \contactreachability\ implies \equilibriumfeasibility.
-Our scenarios show that the  assumption is verified in the majority of cases when at least one contact surface is \textit{quasi flat}, that is when
+Our scenarios show that the  assumption is verified in the majority of cases when at least one contact surface is \textit{quasi flat}~\cite{Prete2016}, that is when
 the friction cone of the contact surface contains the direction opposite to the gravity.
-Figure~\ref{fig:dedefeas} illustrate this observation, demonstrated empirically by the third column of Table~\ref{tab:requestpercent}. In the worst case, in our experiment
+Figure~\ref{fig:dedefeas} illustrates this observation, demonstrated empirically by the third column of Table~\ref{tab:requestpercent}. In the worst case, in our experiment
 the assumption was verified for 82\% of the total amount of trials.
@@
 \subsection{Addressing $\mathcal{P}_3$}
-This paper focuses on the contact planning aspect of multi-contact locomotion. While it does not address directly the continuous motion generation,
-we had to prove that the plans produced by our planner are useful and can be executed by the considered robots. Problem $\mathcal{P}_3$ remains open,
+This paper focuses on \deladp{the} contact planning \deladp{aspect} of multi-contact locomotion. While it does not address directly the continuous motion generation,
+we had to prove that the plans produced by our planner \deladp{are useful and} can be executed by the considered robots.
+Problem $\mathcal{P}_3$ remains \repadp{an active research topic~\cite{Carpentier2016, herzog2015trajectory}}{open},
 and most existing solutions do not consider collisions with the environment during the planning, making it hard to validate constrained problems such as the polaris scenario
 or the table egress in \cite{DBLP:conf/iser/EscandeKMG08}.
-The companion video shows the results we obtained in simulation and on the real HRP-2 robot. Despite the fact that our method does not consider joint torque limits in the planning
+The companion video shows the results that we obtained in simulation and on the real HRP-2 robot. Despite the fact that our method does not consider joint torque limits in the planning
 phase, the robot is able to execute our plans, achieving to demonstrate the interest of our framework.
+\adnote{Probably you should add in appendix an explanation of how you solve $P_3$ and refer to it here.}
 
@@
 
-In this paper we consider the multi contact planning problem, formulated as two sub-problems that we address sequentially.
+In this paper we consider the multi-contact planning problem, formulated as two sub-problems that we address sequentially.
 The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
@@
 %~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
-Our contribution to \Pb is a fast contact generation scheme that can take into
-account criteria to optimize user-defined properties.
+Our contribution to \Pb is a fast contact generation scheme that can \repadp{optimize user-defined criteria}{take into
+account criteria to optimize user-defined properties}.
 
@@
 %
-\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
+\textbf{Regarding performance}, our framework is \deladp{really} efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
 %
 \textbf{Regarding the quality of the paths}, we are able to compute
-\gls{equilibrium feasible} paths in all the presented scenarios, with a high success rates.
+\gls{equilibrium feasible} paths in all the presented scenarios, with high success rates.
 As for \cite{Bouyarmane2009}, failures can still occur, due to the approximate condition used to compute the guide path.
 The low computational burden of our framework however allows for fast re-planning in case of failure.
-Furthermore because of this approximation, the guide search is not complete. The choice is deliberate, because we are convinced
+Furthermore, because of this approximation, the guide search is not complete. The choice is deliberate, because we \repadp{believe}{are convinced}
 that it is necessary to trade completeness for efficiency at all stages of the planner.
@@
 
-Our method applies to any \textit{quasi-flat} scenario, where at least one contact friction cone contains
-the direction opposed to the gravity. This class of scenarios include all the problems proposed at the DARPA Robotics challenge.
-One way of further extending its range of application, that we consider for future work, is to include the equilibrium criterion when solving $\mathcal{P}_1$.
-Considering the set of obstacles intersecting with the reachable workspace for a given root configuration as candidates surfaces, we can use them to verify the equilibrium criterion.
+Our method applies to any scenario where at least one contact friction cone contains
+the direction opposed to the gravity (i.e. \textit{quasi-flat}). This class of scenarios include all the problems proposed at the DARPA Robotics Challenge.
+One way to further extend its range of application, which we consider for future work, is to include the equilibrium criterion when solving $\mathcal{P}_1$.
+Considering the set of obstacles intersecting with the reachable workspace for a given root configuration as candidate surfaces, we can use them to verify the equilibrium criterion.
 This would give us a necessary condition for \equilibriumfeasibility.
 
-While we have exhibited complete multi contact locomotion obtained with our contact planner, our main concern for future work is to address the interpolation
-between contact sequences ($\mathcal{P}_3$), not addressed in this work, which remains an open issue in really constrained scenarios.
-Adressing $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem during the interpolation,	but also improving
+While we have exhibited complete multi-contact locomotion obtained with our contact planner, our main concern for future work is to address the interpolation
+between contact sequences ($\mathcal{P}_3$), not addressed here, and which remains an open issue in highly-constrained scenarios.
+Solving $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem \deladp{during the interpolation},	but also improving
 our plans with transition certificates, i.e. guarantees that the interpolation between two contact configurations is feasible.
@@
 %~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
-Finally, we aim at performing kinodynamic planning to remove the constraint that configurations in the contact plan should be in static equilibrium.
+Finally, we aim at performing kinodynamic planning to remove the constraint that configurations in the contact plan \deladp{should} be in static equilibrium.
 We believe that the most promising direction in this regard is to integrate the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.

### 2017-01-18 body/abstract.tex
andrea s comment + reviewers

@@
 % !TEX root =  ../main.tex
-\adnote{What about replacing 'multiped' with 'legged' everywhere?}
-\repadp{We present a contact planning framework for complex legged locomotion tasks:}{We present a framework capable of producing contact plans describing complex multiped motions}: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
-The need for such a planner was highlighted at the Darpa Robotics Challenge, where \deladp{required} multi-contact behaviors
+We present a contact planning framework for complex legged locomotion tasks: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
+The need for such a planner was highlighted at the Darpa Robotics Challenge, where multi-contact behaviors
 could not be demonstrated.
 
-Indeed, current planners suffer from prohibitive computation times, \repadp{due to}{explained by} the difficulty
+Indeed, current planners suffer from prohibitive computation times, due to the difficulty
 of generating contact postures: the feasible space (the contact manifold) is of high dimension,
@@
 
-We tackle this issue with our novel \deladp{key} idea, the reachability condition. This condition allows us to define
+We tackle this issue with our novel  idea, the reachability condition. This condition allows us to define
 a geometric approximation of the contact manifold, which is of low dimension and can be sampled efficiently.
-Informally, the condition verifies that the root configuration of a robot is €œclose, but not too close to obstacles: close to allow contact creation, not too close to avoid collision.
+Informally, the condition verifies that the root configuration of a robot is close, but not too close to obstacles: close to allow contact creation, not too close to avoid collision.
 Thanks to this condition we decompose the hard contact planning problem into two
@@
 
-Our approach results from the pragmatic choice of favoring efficiency over completeness, \repadp{which we justify empirically}{justified empirically}: in a
+Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically: in a
 few seconds, with satisfying success rates, we generate complex contact plans for various scenarios and robots, namely

### 2017-01-18 body/conclusion.tex
andrea s comment + reviewers

@@
 
-As demonstrated in the results section, the main interest of our method is time performance. Time performance is critical \deladp{to consider using a multi-contact planner} for 
+As demonstrated in the results section, the main interest of our method is time performance. Time performance is critical for 
 online applications with the robot, a goal not satisfied by any of the previous contributions.
@@
 This is demonstrated by the second column of Table~\ref{tab:requestpercent}, and illustrated by Figure~\ref{fig:dedefeas}.
-Indeed, in the large majority of cases, (84\% in the worst case \adnote{I do not get where this 84 comes from}), we are able to find a contact configuration for any configuration in $C_{Reach}^0$.
+Indeed, in the large majority of cases, (84\% in the worst car egress case), we are able to find a contact configuration for any configuration in $C_{Reach}^0$.
 
@@
 \subsection{Addressing $\mathcal{P}_3$}
-This paper focuses on \deladp{the} contact planning \deladp{aspect} of multi-contact locomotion. While it does not address directly the continuous motion generation,
-we had to prove that the plans produced by our planner \deladp{are useful and} can be executed by the considered robots.
-Problem $\mathcal{P}_3$ remains \repadp{an active research topic~\cite{Carpentier2016, herzog2015trajectory}}{open},
+This paper focuses on contact planning of multi-contact locomotion. While it does not address directly the continuous motion generation,
+we had to prove that the plans produced by our planner can be executed by the considered robots. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
 and most existing solutions do not consider collisions with the environment during the planning, making it hard to validate constrained problems such as the polaris scenario
@@
 phase, the robot is able to execute our plans, achieving to demonstrate the interest of our framework.
-\adnote{Probably you should add in appendix an explanation of how you solve $P_3$ and refer to it here.}
+%~ \adnote{Probably you should add in appendix an explanation of how you solve $P_3$ and refer to it here.}
 
@@
 %~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
-Our contribution to \Pb is a fast contact generation scheme that can \repadp{optimize user-defined criteria}{take into
-account criteria to optimize user-defined properties}.
+Our contribution to \Pb is a fast contact generation scheme that canoptimize user-defined criteria.
 
@@
 %
-\textbf{Regarding performance}, our framework is \deladp{really} efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
+\textbf{Regarding performance}, our framework is efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
 %
@@
 The low computational burden of our framework however allows for fast re-planning in case of failure.
-Furthermore, because of this approximation, the guide search is not complete. The choice is deliberate, because we \repadp{believe}{are convinced}
+Furthermore, because of this approximation, the guide search is not complete. The choice is deliberate, because we believe
 that it is necessary to trade completeness for efficiency at all stages of the planner.
@@
 between contact sequences ($\mathcal{P}_3$), not addressed here, and which remains an open issue in highly-constrained scenarios.
-Solving $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem \deladp{during the interpolation},	but also improving
+Solving $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem,	but also improving
 our plans with transition certificates, i.e. guarantees that the interpolation between two contact configurations is feasible.
@@
 %~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
-Finally, we aim at performing kinodynamic planning to remove the constraint that configurations in the contact plan \deladp{should} be in static equilibrium.
+Finally, we aim at performing kinodynamic planning to remove the constraint that configurations in the contact plan be in static equilibrium.
 We believe that the most promising direction in this regard is to integrate the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.

### 2017-01-18 body/contact.tex
andrea s comment + reviewers

@@
 %~ is described in Appendix~\ref{sec:heuristics}.
+
+\subsection{Definition of a contact sequence}
+In previous contributions, a contact plan is defined as a sequence of configurations that describe a quasi-static equilibrium configuration
+for each existing phase of the motion. For instance, a walk cycle would be described by three key configurations: a double support configuration,
+a simple configuration (the contact is broken), and another double support configuration (a contact is created).
+Our definition of the contact plan differs: in this example it would simply consists in the two double support configurations.
+However, in the case where the contact broken and the contact created sequentially are not related to the same effector, our planner outputs two distinct states.
+This representation is sufficient to describe all the contact constraints of the problem. Furthermore it removes the unnecessary constraint of having a simple support,
+quasi static configuration in the path as in the example. As shown in the validation video, this allows our framework to produce motions that are not quasi static, but dynamic.
 
@@
 Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
-If two or more contacts can't be maintained between two consecutive configurations, one or more intermediate configurations are added.
+If two or more contacts can't be maintained between two consecutive configurations, one or more intermediate configurations are added, to ensure
+that at most one contact is broken between two sequential configurations.
 %~ For these steps the root configuration is the same as for the previous step, with the difference that

### 2017-01-18 body/results.tex
andrea s comment + reviewers

@@
 \label{sec:results}
-In this \repadp{section}{Section} we present some of the results obtained with our planner. The complete sequences \deladp{computed} are shown in the companion video.
+In this section we present some of the results obtained with our planner. The complete sequences are shown in the companion video.
 Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
@@
 
-At the end of this \repadp{section}{Section}, we discuss the role of the parameters of our framework\deladp{ on the output}.
-We then provide the \textit{interactive} computation times obtained in each case.
+At the end of this section, we discuss the role of the parameters of our framework. We then provide the \textit{interactive} computation times obtained in each case.
 We also compare the times obtained with HRP-2 with respect to previous works.
 
-These scenarios \repadp{complement}{complete} the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
+These scenarios complement the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
 %~ (Figure~\ref{fig:robots_old}).
@@
 In all the scenarios considered, the formulation of the problem is always the same:
-a start and goal root configurations are provided as \deladp{an} input \deladp{of the scenario}.
+a start and goal root configurations are provided as input.
 The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
@@
 
-The goal is to climb three 15-cm high steps. This height requires HRP-2 to use a ramp to perform the task. \adnote{It is not clear why, so either explain why, or remove the comment.}
+The goal is to climb three 15-cm high steps.
 
@@
 \subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing}):}
-From a bent configuration, \repadp{the robot has to stand up}{a standing-up motion is computed in a constraining environment.
-The resulting motion involves} using a wall as support, and climbing a 25-cm high step.
+From a bent configuration, the robot has to stand up using a wall as support, and climbing a 25-cm high step.
 
@@
 \subsubsection{HRP-2 -- Polaris car egress (Figure~\ref{fig:car}):}
-\repadp{In this scenario---inspired from the DRC car egress (\url{http://cpc.cx/edH})---HRP-2 has}{This scenario is inspired from the DARPA-Robotics-Challenge car egress scenario (\url{http://cpc.cx/edH}).
-HRP-2 has to find the contact sequence that allows it} to step out of the Polaris car.
+In this scenario inspired from the DRC car egress HRP-2 has to step out of the Polaris car.
 
@@
 \subsubsection{HyQ -- DRC-style rubble (Figure~\ref{fig:darpa})}
-The quadruped robot \repadp{must}{is given the task to} cross a rubble composed of bricks rotated at different angles and directions.
+The quadruped robot must cross a rubble composed of bricks rotated at different angles and directions.
 
@@
   \caption{
-           Robust crossing of rubbles by HyQ \deladp{($b_0 > 20$)}. }
+           Robust crossing of rubbles by HyQ. }
 		   \label{fig:darpa}
@@
 \subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs}):}
-In this long scene, HyQ \repadp{has to cross a 55-cm large hole, followed by}{is first required to cross a 55-cm large hole; then, to cross} a narrow ``bridge'', only 25-cm large.
+In this long scene, HyQ has to cross a 55-cm large hole, followed by a narrow ``bridge'', only 25-cm large.
 
@@
   \caption{
-           Crossing a hole contact sequence for HyQ \deladp{($b_0 > 4$)}. }
+           Crossing a hole contact sequence for HyQ. }
 		   \label{fig:HyQ_obs}
@@
 \subsection{Role of the main parameters} \label{sec:influence}
-We discuss the several factors \adnote{Either use 'the' or use 'several'} that influence the outcome of our planner: the root scaling factor $s$ (Section~\ref{sec:scaling}), the \deladp{choice of} heuristics for
+We discuss the factors that influence the outcome of our planner: the root scaling factor $s$ (Section~\ref{sec:scaling}), the heuristics for
 contact generation (Appendix~\ref{sec:heuristics}), and lastly, the discretization step for the guide path. The appropriate value for these parameters
-is computed empirically based on statistical analysis \adnote{Are you sure you wanna use the word 'statistical'?} or trials and errors.
+is computed empirically based on use-case analysis or trials and errors.
 
 \subsubsection{Choosing the scaling factor $s$:} \label{sec:params}
-\deladp{To find a convenient value $s^*$ of $s$, we proceeded as follows. }
+%~ \deladp{To find a convenient value $s^*$ of $s$, we proceeded as follows. }
 For several values of $s$, we generated $10 000$ configurations. 
-We then computed the sensitivity of the reachability condition (\repadp{percentage}{proportion} of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}).
-Similarly we computed the specificity of the condition (\repadp{percentage}{proportion} of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}).
+We then computed the sensitivity of the reachability condition (percentage of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}).
+Similarly we computed the specificity of the condition (percentage of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}).
 The obtained results for HRP-2 are shown in Table~\ref{tab:scale}, averaged over all scenes (except for the car egress: in this scenario, 
@@
 
-As it can be expected, the scaling results in a high \repadp{increase}{decrease} of the sensitivity, with a decrease of the specificity.
+As it can be expected, the scaling results in a high increasepercentage of the sensitivity, with a decrease of the specificity.
 For HRP-2 we decided to set $s^*=1.2$.
@@
  \end{tabular}
-\caption{Sensitivity and specificity values \addadp{of} the reachability condition, depending on the scaling value \addadp{$s$} of $W^0$.}
+\caption{Sensitivity and specificity values of the reachability condition, depending on the scaling value $s$ of $W^0$.}
 \label{tab:scale}
@@
 
-\subsubsection{\repadp{Choosing the heuristics}{On the choice of heuristics}:} \label{sec:heuristichoices}
-In our conference paper~\citep{tonneauisrr15}, the computed motions were \deladp{only} generated using the EFORT heuristic.
+\subsubsection{Choosing the heuristics:} \label{sec:heuristichoices}
+In our conference paper~\citep{tonneauisrr15}, the computed motions were generated using the EFORT heuristic.
 EFORT is designed for tasks requiring to exert important forces (such as pushing / pulling / climbing). 
-\repadp{In}{Regarding} locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
+In locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
 configurations close to singularities (and joint limits). While this does not significantly impact
-the generation of the plan, the resulting interpolation \deladp{on the real robot} turned out to be harder.
+the generation of the plan, the resulting interpolation turned out to be harder.
 For this reason, we prefer to use our manipulability-based heuristic for the legs of the robot, but we still
@@
 For future work, we would like to automatically adapt the size of the discretization step to the complexity of the environment.
-\adnote{Maybe you should say which discretization step you used.}
 
@@
 To analyze performance, we ran the planner 1000 times for each considered scenario.
-We measured the computation time spent in each \repadp{part}{aspect} of the algorithm, and analyzed their success rate.
+We measured the computation time spent in each part of the algorithm, and analyzed their success rate.
 
@@
  \hline
-   Steep stairs     	& 100\% \\
-   Standing up 			& 68\% 	\\
-   Car egress 			& 77\% 	\\
+   Stairs     	& 100\% \\
+   Standing			& 68\% 	\\
+   Car 			& 77\% 	\\
    Rubble 				& 97\% 	\\
-   Obstacle race        & 88.0\% 	\\
+   Race        & 88.0\% 	\\
  \end{tabular}
@@
 \subsubsection{Computation times:}
-Table~\ref{tab:requestime} summarizes the \deladp{performance measurements obtained, in terms of} computation times.
+Table~\ref{tab:requestime} summarizes the computation times.
 
-\repadp{F}{Considering the repartition of the computation time, f}or HRP-2, most of the time \repadp{was}{is} spent performing inverse kinematics.
+For HRP-2, most of the time was spent performing inverse kinematics.
 This is not surprising considering the number of calls to the methods: IK projection is used intensively to maintain contact continuity between two postures; 
@@
 In all scenarios, one can observe that the average computation time for one single step is largely below one second,
-thus allowing to consider \gls{interactive} applications\repadp{ and}{. This would allow to consider} online autonomous planning of the robot motion.
+thus allowing to consider \gls{interactive} applications and online autonomous planning of the robot motion.
 
 \subsubsection{Success rates:}
-Table~\ref{tab:sucess_planning} summarizes the successful planning rates\deladp{ obtained for each scenario}.
-Despite the complexity of the scenarios addressed, as well as the approximations made in our formulation, our planner \repadp{succeeded}{succeeds} in the large majority of cases.
+Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
+Despite the complexity of the scenarios addressed, as well as the approximations made in our formulation, our planner succeeded in the large majority of cases.
 
-Table~\ref{tab:requestpercent} presents\deladp{, for each scenario,} the rate of \addadp{successful} contact generation\deladp{ that succeeded}.
+Table~\ref{tab:requestpercent} presents the rate of successful contact generation.
 Note that a failure in contact generation for a root configuration is not equivalent to a failure in the contact plan.
 It simply means that another limb was tested for contact generation for the same root configuration.
-As expected, the more constrained scenario, the car egress \deladp{one}, provides the \repadp{worst}{less satisfying} results, despite the high success rate of the planner.
+As expected, the more constrained scenario, the car egress, provides the less satisfying results, despite the high success rate of the planner.
 
@@
 while controlling the robot. Indeed, when the contact planning fails, it fails rapidly. This allows us to rapidly re-plan with a reasonable chance of success.
-The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible \repadp{would be}{is probably} to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.
+The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible would be to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.
 

### 2017-01-18 main.tex
Nouvel abstract Main.tex change pour pb de compilation.
Move Section about heuristics to Appendix
replaced contact feasible with contact reachable to match with reachability confition
updated overview
merging appendices
florent s comment
andrea s comment + reviewers

@@
 \definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
+\newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
+\newcommand{\addadp}[1]{\added[id=adp]{#1}}
+\newcommand{\repadp}[2]{\replaced[id=adp]{#1}{#2}}
+\definechangesauthor[name={Steve Tonneau}, color=blue]{st}
+\newcommand{\delst}[1]{\deleted[id=st]{#1}}
+\newcommand{\addst}[1]{\added[id=st]{#1}}
+\newcommand{\repst}[2]{\replaced[id=st]{#1}{#2}}
+\newcommand{\gls}[1]{\textit{#1}}
+\newcommand{\glslink}[2]{{#2}}
 
+%~ \usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red,draft]{hyperref}
 \usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
-\usepackage{glossaries}
+%~ \usepackage{glossaries}
 
@@
 
-\makeglossaries
-\input{body/glossary}
+%~ \makeglossaries
+%~ \input{body/glossary}
 
@@
 
-\title{An efficient acyclic contact planner for multiped robots}
+\title{An efficient acyclic contact planner for legged robots}
 
@@
  %~ Franck Multon\affilnum{3},
-\affiliation{\affilnum{1}LAAS-CNRS / Universit\'e Paul Sabatier, Toulouse, France\\
+\affiliation{\affilnum{1}LAAS-CNRS / Universit\'e de Toulouse, France\\
 \affilnum{2}Inria, Rennes, France\\
@@
 \input{body/contact}
-\input{body/equilibrium}
-\input{body/efort}
 \input{body/results}
@@
 
-\glsaddall
-\printglossaries
+%~ \appendix
+%~ \section*{Appendix}
+
+%~ \glsaddall
+%~ \printglossaries
 \bibliographystyle{SageH}
@@
 \appendix
+\input{body/equilibrium}
+\input{body/efort}
 \input{body/appendix_roms}

### 2017-01-18 cover2.tex
cover  letter

\title{Cover letter}

\documentclass[12pt]{article}

\begin{document}
\maketitle

This cover letter for the manuscript ``An efficient acyclic contact planner for
legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity. We left the original cover
letter at the end of this document.

We would first like to thank the reviewers for their previous feedback on the initial manuscript. We believe we have answered
all of their concerns, the main ones being that our method had not been tested on a real robot or a simulation, and that the paper was too verbose.
In the following lines we are going to describe how we improved our paper in this regard, and answer to the other remarks of the reviewers.

\section{Validation of the method (addressing  $\mathcal{P}_3$) }
Because we had already validated some of our plans, we were really surprised by the remark of reviewer 2 stating that our plans are not executable, even in simulation.
The stair climbing scenario has been demonstrated
on the real HRP-2 robot, while the standing up had been demonstrated in simulation. While it was mentioned in the paper,
we did not insist on these demonstrations because our main contribution focuses on the other problems.

We believe that, as reviewer 1 notices, this observation was due to the fact that we do not clearly define what we mean by a contact plan.
This is now detailed in section 5.1.. In any case, the validation videos joined to the submission should now establish that our plans are in indeed feasible, even without considering
the torque constraints.

In this submission, we have thus joined the simulation and real life videos of all the scenarios we demonstrate.
These videos show that the plans we compute correspond to feasible motions.
We thank the reviewers for this comment,
since we now believe that our paper is stronger with the complete demonstration. However, while we briefly describe how we solve $\mathcal{P}_3$, the focus of the 
paper remains $\mathcal{P}_1$ and $\mathcal{P}_2$.

\section{Success rates of the method}
At this point, we believe it is important to clarify our position regarding both reviewers concerns of the success rates of our paper.
We did something that no other contact planning paper does, that is providing the success rates of our contact planner. Our method is not always successful, we assume it and argue that it is mostly successful, and that it can always find a feasible solution fast enough for online applications by replanning upon failure.

In previous works, it is in fact unknown whether the methods are always successful or not. We simply have no idea. For instance, the CIO of Mordatch,
because of the local convergence of the method, \textbf{necessarily} fails in some scenarios. We just do not know how often. Similarly, Escande et al. had to manually edit 
the end effector trajectories to avoid collisions with the environment, but we do not know how often. Considering this,
we do not understand why we are criticized for providing our success rates, which actually demonstrate the interet of our approach.

\section{Originality and interest of the work}
We strongly disagree with reviewer 1 regarding the originality of the work. Reviewer 1 states that we simply experiment with minor heuristic adjustments.
Bretl formalized well the problem without providing an algorithm applicable to the general case. Escande focused on the contact planning $\mathcal{P}_2$. 
Our work tackles $\mathcal{P}_1$ as well, in a much more efficient manner than Bouyarmane et al.. 
As for Hauser, these works are limited by the computational complexity of considering the full dimension of the problem.

The reachability condition provides a dimensionality reduction of the problem. This is clearly an originality regarding the previous work who always consider the whole-body problem.
We believe that this reduction is the only way to make the contact planning problem tractable, in the same way that many researchers (such as ourselves, but also Wieber, Tedrake, Righetti and others) use dimensionality reduction successfully to address
the trajectory optimization problem $\mathcal{P}_3$.

Furthermore, the computational time provided by our method is a contribution in itself: it makes the difference
between a method that can be used at the Darpa challenge or not, as we show in our introduction.

\section{A clearer paper}
As requested, we have deeply shortened the paper. More importantly, we have done an important work of defining in simpler terms
the different concepts that we use in the paper, and reduced their number.
We believe that the paper is much more clearer in its current form.

\section{Justification of the simplifying assumptions}
Regarding several comments from both reviewers, stating that our paper does not justify the fact that contact feasibility implies equilibrium feasibility,
we now have added a section that discusses in depth the assumption (Section 7). We clearly describe which hypothesis (we call it the quasi-flat hypothesis, referring
to the contacts in the configurations, rather than the vague cluttered term) are required to validate our approach, and analyze
the success rates we obtained in each scenario to justify that the assumption works.

\section{Comparison with other methods}
As noted by reviewer 2, in our previous paper it was indeed complex to compare our work with previous contributions,
since they were adressing the complete problem. We now give the complete computation times and compare them with previous works,
clearly demonstrating that our method outperforms them by orders of magnitude (Section 6.4).

\section{On the static equilibrium of each configuration}
Reviewer 1 stated that ``the planned contact path is only discrete with static equilibrium considered for each
configuration, which is not very realistic''. First, we recall that this assumption is always made in the previous works that address cluttered scenarios,
such as the work from Bretl and Escande. Work that do not make this assumption such as Mordatch and al. only offer local convergence and can't address these scenarios. Furthermore, this limitation is in fact less restrictive in our approach that the one of Escande, since while our plan is quasi static, our resulting motion is dynamic, as now explained in the paper (Section 5.1).

\section{Actual car model of the DARPA Challenge}
As asked by reviewer 2, we now use the polaris car model for demonstrating our egress scenario.

\end{document}

### 2017-01-19 code/data/config/conf_hrp2_polaris.py
updated code for polaris scenario

@@ -12,22 +12,17 @@ SAVE_DATA                   = True;
 SMOOTH_FILTER_WINDOW_LENGTH = 41;
 MAX_TEST_DURATION           = -1;
 dt                          = 1e-3;
-FOLDER = '/home_local/dev/hpp/src/2016_ijrr_traj/darpa_ok/'
-INPUT_FILE_NAME             = [FOLDER+'seq_1',FOLDER+'seq_2',FOLDER+'seq_3',FOLDER+'seq_4',FOLDER+'seq_5',FOLDER+'seq_6'];
-#INPUT_FILE_NAME             = [FOLDER+'seq_5', FOLDER+'seq_6'];
-RECORD_FILE_NAME             = 'seq_all';
-RECORD                      = True
-CAMERA_TRANSFORM            = [-1.9623876810073853, -5.876553058624268, 3.1662111282348633, 0.8184708952903748, 0.5468565821647644, -0.09847398102283478, -0.14613758027553558];
+INPUT_FILE_NAME             = '/home_local/dev/hpp/src/2016_ijrr_traj/polaris_2_compressed';
 model_path                  = ["/home_local/dev/hpp/install/share"];
-urdfFileName                = model_path[0] + "/hyq_description/urdf/hyq.urdf";
-sceneFileName               = model_path[0] + "/hpp-rbprm-corba/meshes/darpareduced.stl";
+urdfFileName                = model_path[0] + "/hrp2_14_description/urdf/hrp2_14_reduced.urdf";
+sceneFileName               = model_path[0] + "/hpp-rbprm-corba/meshes/polaris.stl";
 freeFlyer                   = True;
 
 ''' CONTROLLER CONFIGURATION '''
 ENABLE_CAPTURE_POINT_LIMITS     = False;
 ENABLE_TORQUE_LIMITS            = True; # param
 ENABLE_FORCE_LIMITS             = True; # param
-ENABLE_JOINT_LIMITS             = True; # param, modulate with 4 next (position viability false, velocity ok)
+ENABLE_JOINT_LIMITS             = False; # param, modulate with 4 next (position viability false, velocity ok)
 IMPOSE_POSITION_BOUNDS          = True;
 IMPOSE_VELOCITY_BOUNDS          = True;
 IMPOSE_VIABILITY_BOUNDS         = True;
@@ -40,7 +35,7 @@ USE_JOINT_VELOCITY_ESTIMATOR    = False;
[…]

### 2017-01-19 code/python/track_locomotion_traj.py

@@ -9,8 +9,10 @@ if cwd+'/data/config' not in sys.path:
     sys.path += [cwd+'/data/config',];
 
 #import test1_planar as conf
-import conf_hyq_hole as conf
+#~ import conf_hyq_hole as conf
 #~ import conf_hyq_bridge as conf
+import conf_hyq_darpa as conf
+#import conf_hrp2_polaris as conf
 
 from pinocchio_inv_dyn.standard_qp_solver import StandardQpSolver
 from pinocchio_inv_dyn.simulator import Simulator
@@ -111,7 +113,7 @@ def updateConstraints(t, i, q, v, invDynForm, contacts, ee_tasks):
             Ni = np.matrix(contacts[name]['N']).T;
             for j in range(Pi.shape[1]):
 #                print "    contact point %d in local frame:"%j, Pi[:,j].T, Ni[:,j].T;
-                print "    contact point %d in world frame:"%j, oMi.act_point(Pi[:,j]).T, (oMi.rotation * Ni[:,j]).T;
+                print "    contact point %d in world frame:"%j, oMi.act(Pi[:,j]).T, (oMi.rotation * Ni[:,j]).T;
         else:
             Pi = conf.DEFAULT_CONTACT_POINTS;   # contact points is expressed in local frame
             Ni = oMi.rotation.T * conf.DEFAULT_CONTACT_NORMALS; # contact normal is in world frame
@@ -255,6 +257,27 @@ def startSimulation(q0, v0, solverId):
         
     return True;
 
+def fusion_data(files):
+    all_data = {'Q':[],'fly':[],'C':[],'frameswitches':[] }
+    current_offset=0
+    for fi in files:
+        f = open(fi, 'rb');
+        print(len(all_data['C']))
+        data = pickle.load(f);
+        for k, el in enumerate(data['frameswitches']):
+            #if(k != 0 or len(all_data['C']) == 0):
+            all_data['frameswitches'] = all_data['frameswitches'] + [[current_offset + el[0], el[1]]]
+        current_offset += len(data['Q'])
+        all_data['Q']   += data['Q'][:];
+        all_data['fly'] += data['fly'][:];
[…]

### 2017-01-19 cover2.tex
[COVER] modif of the first part of the letter.

@@
 
-This cover letter for the manuscript ``An efficient acyclic contact planner for
-legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity. We left the original cover
-letter at the end of this document.
+%This cover letter for the manuscript ``An efficient acyclic contact planner for legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity.
+This cover letter introduces the first revision of the manuscript ``An efficient acyclic contact planner for legged robots'', resubmited for the ISSR'15 sepcial issue.
+As discussed with the Editor, we hoped to have reviewer continuity for the revision.
+The cover letter then plays more the role of a rebutal.
+We left the original cover letter at the end of this document.
 
@@
 
+\section{Recall of main paper contributions}
+The paper is concerned with the problem of planning acyclic contact planning. 
+More precisely, we propose a very efficient method to address what we believe is the core of the difficulty of the contact planning problem: discovering the sequence of key contact posture that would lead to a safe and stable contact trajectory of the robot.
+Using the terminlogy of the paper, our contact planner addressed the two sub-problems $\mathcal{P}_1$ (finding a guide path for the robot root) and $\mathcal{P}_2$ (finding a sequence of stable contact postures and corresponding contact locations).
+Our claim is that the solution of our planner leads to a straight-forward resolution of the last sub-problem $\mathcal{P}_3$ (finding the complete trajectory connecting the contact postures).
+In particular, our method is able to interactive performances, i.e. the planner needs less time to compute the solution than what is needed for the robot to execute it.
+This practical result corresponds to a dramatic improve of the state of the art, for example comparing to the capabilities of planning acyclic contact motions demonstrated by the humanoid teams during the Darpa Robotics Challenge last year.
+
+Regarding this contribution, the reviews mostly criticized the structure of the paper (too verbose) and the unjustified claim that the contact sequence output by the planner leads to easy resolution of $\mathcal{P}_3$ hence complete resolution of the contact planning problem.
+
+In the revision, we implemented a solver inspired by the state of the art to solve $\mathcal{P}_3$ and empiricially proved the claim that our $\mathcal{P}_1$-$\mathcal{P}_2$ planner leads to easily solving $\mathcal{P}_3$.
+All the robot contact sequences reported in the paper have been converted to full contact trajectories using this $\mathcal{P}_3$ solution.
+We also executed one of them on the real robot.
+This extension, along with the correction of the other remarks by the reviewers, have been integregated in the revision paper, that we also reworked to make it clearer, shorter and sharper.
+
+Details of these modifications follow.
+
 \section{Validation of the method (addressing  $\mathcal{P}_3$) }
-Because we had already validated some of our plans, we were really surprised by the remark of reviewer 2 stating that our plans are not executable, even in simulation.
-The stair climbing scenario has been demonstrated
-on the real HRP-2 robot, while the standing up had been demonstrated in simulation. While it was mentioned in the paper,
-we did not insist on these demonstrations because our main contribution focuses on the other problems.
+In his review, reviewer 2 stated that our plans are likely to be not executable, even in simulation.
 
-We believe that, as reviewer 1 notices, this observation was due to the fact that we do not clearly define what we mean by a contact plan.
-This is now detailed in section 5.1.. In any case, the validation videos joined to the submission should now establish that our plans are in indeed feasible, even without considering
-the torque constraints.
+\begin{quote}
+  \textit{A quick glance at the sequences displayed in the video reveals that the "motion patterns" (and I understand these are not "motions" per se, but "contact plans") are not physically consistent (even reagrding them as mere contact plans and not as the final motions).
+The robot moves a contact without needing to readjust the posture in order to release that contact (or "set it free", by nullifying the contact forces on that contact).
+This results in sequences in which the contacts seem to be "slided" rather than removed and repositioned.
+This denotes a significant flaw in the algorithm.
+}
+\end{quote}
 
-In this submission, we have thus joined the simulation and real life videos of all the scenarios we demonstrate.
+We were really surprised by the remark, in particular because we had already validated some of our plans.
+The stair climbing scenario has been demonstrated on the real HRP-2 robot, while the standing up had been demonstrated in simulation. 
+While it was mentioned in the paper, we did not insist on these demonstrations because our main contribution focuses on the other problems. 
+We believe that, as reviewer 1 notices, this observation was due to the fact that we do not clearly define what we mean by a contact plan (i.e.
+a discrete sequence of key contact posture with the corresponding contact locations).
+We thank the reviewers for these comments, since it pointed out that what we believed to be an evidence must be carefully explained and justified.
+
+A rigorous definition of what we mean by contat plan is now detailed in section 5.1.
+
+%In any case, the validation videos joined to the submission should now establish that our plans are in indeed feasible, even without considering the torque constraints.
+
+We also implemented a solver inspired by the state of the art to dynamically interpolate the contact plan and produce a complete trajectory of the robot, that we validated in a full-dynamics simulator.
+This implementation completely validates that the contact plan is feasible by the real robot.
+We also indeed executed one of the computed trajectory (climbing the stairs with HRP-2).
+
+We have joined the full-dynamics simulation and real life videos of all the scenarios we demonstrate.
 These videos show that the plans we compute correspond to feasible motions.
-We thank the reviewers for this comment,
-since we now believe that our paper is stronger with the complete demonstration. However, while we briefly describe how we solve $\mathcal{P}_3$, the focus of the 
-paper remains $\mathcal{P}_1$ and $\mathcal{P}_2$.
+While we briefly describe how we solve $\mathcal{P}_3$, the focus of the paper remains $\mathcal{P}_1$ and $\mathcal{P}_2$. 
+The solution of $\mathcal{P}_3$ is reported in the appendix to emphasize the main focus.
+
+\newcommand{\mP}[1]{$\mathcal{P}_#1$}
 
 \section{Success rates of the method}
-At this point, we believe it is important to clarify our position regarding both reviewers concerns of the success rates of our paper.
-We did something that no other contact planning paper does, that is providing the success rates of our contact planner.
-Our method is not always successful, we assume it and argue that it is mostly successful, and that it can always find a feasible solution fast enough for online applications by replanning upon failure.
+At this point, we believe it is important to clarify our position regarding the concerns of both reviewers about the success rates of our paper.
+In the paper, we are providing the success rates of our contact planner. 
+Such an empirical study has yet never been provided for other contact planners (arguabily because previous works did not reached the maturity level that would have permit such an empirical study).
+Our method is not always successful, as very often with probabilistic searches.
+More precisely, the hypothesis taken by \mP1 may lead to infeasible \mP2 problem, and similarly \mP2 hypotheses may lead to infeasible to infeasible \mP3.
+We accept this fact and argue that it is mostly successful (i.e. more than 90\% of success in reasonable ``DRC-like'' scenarios).
+Even in case of failure, the planner is based on random sampling that approximate probabilistic completeness, and can be restarted to obtain another candidate trajectory.
+In practice, it can always find a feasible solution fast enough for online applications by replanning upon failure.
 

### 2017-01-19 body/intro.tex
citing liu

@@
 Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path.
-\citeauthor{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with
-a discretized environment. This planning phase does not account for collisions, implying that replanning is required in case of failure. This approach is efficient 
-in the demonstrated scenarios. In highly-constraining cases such 
-as the car egress scenario that we address here, we believe that including collision constraints in the planning is a requirement.
+\citeauthor{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment.
+This planning phase does not account for collisions, implying that replanning is required in case of failure.
+In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}.
 
@@
 used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
-at \gls{interactive} frame rates (we say that a method is interactive when the computation time for planning is less than the
-time for execution). While this new approach provides an exciting direction of research, currently the resulting motions
-look unnatural, and do not seem applicable to real robots.
+at \gls{interactive} frame rates (a method is interactive when the computation time for planning is less than the
+execution time). While this new approach provides an exciting direction of research, currently the resulting motions
+ do not seem applicable to real robots.
 
@@
 
-\noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm. The key to its efficiency is that it does not sample
-directly the contact manifold, but an approximation of the \textit{contact reachable} space.
+\noindent \textbf{Regarding $\mathcal{P}_1$}.
+We propose a fast guide path planning algorithm.
+The key to its efficiency is that it does not sample directly the contact manifold, but an approximation of the \textit{contact reachable} space.
 The \textit{contact reachable} space is a low-dimensional space for which there exists a mapping to the contact manifold.

### 2017-01-19 body/appendix_optim.tex
complete paper

\section{Addressing $\mathcal{P}_3$}
\label{app:optim}
The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
To address the others scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox}), which can be integrated directly in our motion planner software.
This implementation is slower, but handles collision avoidance in the trajectory generation phase.
It uses a modified formulation of the optimization problem, that we present here.

\subsection{The centroidal cone}
We rewrite Eq.~\ref{eq:new_eul} in the general case:

%~ \begin{align} \label{eq:new_eul}
%~ \underbrace{
%~ \mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
%~ \hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
%~ }_\mathbf{G} \bm{\beta}, = 
%~ \underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
%~ \underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
%~ \end{align}

\begin{align} \label{eq:new_eul_acc}
\mathbf{G} \bm{\beta} = 
\underbrace{\mat{m (\mathbf{\ddot{c}} - {g}) \\m (\mathbf{\ddot{c}} - {g}) + \dot{L}}}_{\mathbf{w}}
%~ \underbrace{m (\mathbf{\ddot{c}} - \mathbf{g})  \\ m (\mathbf{\ddot{dc}} - \mathbf{g}) + \mathbf{\dot{L}} }_{\mathbf{w}}
\end{align}

where $\dot{L}$ is the angular momentum expressed at the com.
Eq.~\ref{eq:new_eul} defines a 6 dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
this cone determines all the admissible wrenches $\mathbf{w}$ such that the robot won't fall. The objective 
is thus to plan a trajectory for $c$ and $\dot{L}$ that remains inside $\mathcal{K}$. As a cone,
$\mathcal{K}$ can be described as a linear inequality, obtained with the double description method~\citep{Fukuda1996}:

\begin{equation}
\label{eq:cone_k}
	\mathcal{K} =  \left\{ \mathbf{w}, \mathbf{A}\mathbf{w} \leq \mathbf{b	} \right\}
\end{equation}

We now consider a two contact configurations  $\mathbf{q}_0$ and $\mathbf{q}_1$ given by our planner: in the general case one contact is broken and one created to get from
$\mathbf{q}_0$ to $\mathbf{q}_1$. The motion is thus constrained by three cones $\mathcal{K}_s, s =0 \dots 2$. We manually define the duration of each contact phase, of total
time $T$, and formulate the following optimization problem:

\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
%~ \find \quad & \mx{R} \in \Rv{3\times T}, \mxdd{C} \in  \Rv{6\times T} \\
\minimize_{\mathbf{c}(t), \mathbf{w}(t)}  \quad & \sum_{s=0}^{2} \int_{t_s}^{t_s + \Delta t_s}  h(\mathbf{c}, \mathbf{w})  \\
\st \quad & \mathbf{\dot{c}} = f(\mathbf{c}, \mathbf{w}) \\
	\quad & \mathbf{A}_s \mathbf{w}(t)  \leq \mathbf{b}_s , \forall t \in  \left[t_s, t_s + \Delta t_s \right[ \\
	\quad & \mathbf{c}(0)  = \mathbf{c}_{\mathbf{q}_0} \\
	\quad & \mathbf{c}(T)  = \mathbf{c}_{\mathbf{q}_1} \\
	\quad & \mathbf{c}(0)  = \mathbf{\dot{c}}(0)  = \mathbf{\ddot{c}}(0) = \mathbf{0}\\
	\quad & \mathbf{c}(T)  = \mathbf{\dot{c}}(T)  = \mathbf{\ddot{c}}(T) = \mathbf{0}\\
    %~ \quad & ({\vct{r}}\tra\vct{r}=1,  \forall t  ?) \\
    %~ \quad & \delta t \sum\limits_1^{T-1}{\vc{r}^1}\tra {\vc{r}^1} \ge t_{transit} \\
    %~ \quad & \mx{A}_{\int} \mxdd{C} \leq \vc{a}_{\int} \\
    %~ \quad & \mx{V}_{\int} \mxd{C} \leq \vc{v}_{\int} \\
    %~ \quad & \mx{B}_{\int}^{1} \mx{C} \leq \vc{b}_{\int} \\
    %~ \quad & [\vc{c}_0\tra, \vcd{c}_0\tra]\tra = \vc{x}_0 \\
    %~ \quad & \vc{c}_T = \vc{c}_T^{input} \\
\end{aligned} \end{equation}

where $h$ is a cost function (We typically minimize the angular momentum and acceleration over the whole trajectory, and use
a log barrier to constrain kinematically the center of mass position with respect to the contact locations).
This formulation can trivially be extended over the whole contact sequence.
In our implementation, the problem is discretized using time steps of 100 ms.

As an output, the solution of this optimization problem gives us a admissible center of mass trajectory.
To compute the whole body motion, we use a two step approach.

First, we plan a kinematic motion for the robot, subject to the contact constraints and center of mass constraints.
This is achieved using a RRT planner. Using a sampling based planner guarantees that the effector will find a whole-body motion that is collision free.

Planning the whole-body takes approximatively 1.5 seconds for a complete contact transition.

This motion is then validated in our dynamic simulator~\citep{DelPrete2015b}, using the whole-body trajectory as a postural task to an inverse dynamics
solver, that accounts for the torque constraints of the robot. This final result is shown in the videos.

### 2017-01-19 body/results.tex
complete paper

@@
 \centering
-\begin{tabular}{ c | c | c | c | c }
- Scenario & Method  & Guide Path & Contact sequence & Interpolation \\
+\begin{tabular}{ c | c | c }
+ Scenario & Method  & Complete computation time \\
  \hline
-   \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}5.42 min}  \\
-							 & \textbf{Ours} + \citeauthor{Carpentier2016}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & $ <$ 2s \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
+   \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  5.42 min  \\							 
+							  & \citeauthor{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
+							 & \textbf{Ours} + \citeauthor{Carpentier2016}  & \textbf{$ <$ 2s} \\
  \hline
-   \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}4.08 min}  \\
-							 & \textbf{Ours}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & X \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
+   \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  4.08 min  \\
+							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
+							 & \textbf{Ours}  & \textbf{$ <$ 2s}   \\
  \hline
-   \multirow{3}{*}{Stair 40 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}10.08 min}  \\
-							 & \textbf{Ours}  & \textbf{5 s} & \textbf{3 s} & X \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
- %~ \hline
-   %~ \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 10 min & \multicolumn{2}{c}{\cellcolor{blue!25}3.5 hours}  \\
-							 %~ & \textbf{Ours}  & \textbf{529 ms} & \textbf{$<$ 22 s} & X \\
+   \multirow{3}{*}{Stair 40 cm} & \citeauthor{Hauser06usingmotion} &  10.08 min  \\
+							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
+							 & \textbf{Ours}   & \textbf{$ <$ 5s}   \\
+ \hline
+   \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 3.5 hours  \\
+							 & \textbf{Ours}  & \textbf{$<$ 60 s} \\
+
  \end{tabular}
-\caption{Comparison between the computation times obtained by our method and previous ones, for each phase of the planning. Blue cells indicate merged times (because no
-precise time information is available). The X symbol indicates scenarios where we did not address the interpolation of the contact sequence.}
+\caption{Comparison between the computation times obtained by our method and previous ones for addressing the whole problem.}
 \label{tab:compprev}
@@
 \subsection{Validation of the contact plans, and comparison with previous work} \label{sec:compa}
-To validate our plans, we had to implement our own interpolation method to address $\mathcal{P}_3$.
-We refer the reader to \citep{Carpentier2016} for the implementation details of our method. 
-This motion is in turn validated in our dynamic simulator~\citep{DelPrete2015b}. The stair climbing scenario was validated on the real HRP-2 robot \citep{Carpentier2016}.
+To validate our plans, we either use the framework proposed by \citeauthor{Carpentier2016} or our own implementation of a $\mathcal{P}_3$ solver, detailed in Appendix~\ref{app:optim}.
 The obtained motions are available in the video.
@@
 Few contributions provided computation times on the complete problem, necessary to provide comparison between our approaches. 
-Furthermore, $\mathcal{P}_3$ is an open problem in the presence of obstacles.
-For instance in \cite{DBLP:conf/iser/EscandeKMG08}, the end-effector trajectories are computed automatically with splines, but are manually corrected if they enter in collision.
-In \citep{Mordatch:2012:DCB:2185520.2185539}, internal joint collisions can occur in the resulting motion.
-Recent methods suffer from similar issues \citep{Carpentier2016}.
-The only valid scenarios addressed completely and automatically in previous works are thus the stair climbing scenarios with HRP-2 
-To compare our approach with previous contributions, we considered those that addressed the complete problem, and selected scenarios similar to ours, thus focusing on the HRP-2 robot.
+Furthermore, $\mathcal{P}_3$ remains an open problem in the presence of obstacles.
+The only valid scenarios addressed completely in previous works are thus the stair climbing scenarios of different heights proposed by \citeauthor{Hauser06usingmotion}, and the table egress scenario by ~\citeauthor{DBLP:conf/iser/EscandeKMG08}, which we consider to be of similar complexity with respect to the car egress scenario (we did not consider the stairs in the scene).
+Both scenarios use HRP-2.
 
-Table~\ref{tab:compprev} presents the results we obtained
-
-Nonetheless, we indicate in Table~\ref{tab:compprev} what is actually computed by each method and in how much time.
-We compare motions obtained in previous works with HRP-2 (or a humanoid avatar) with ours.
-The two scenarios we selected consist in:  climbing one step, of height 20, 30 or 40 centimeters; 
-getting out of a table, that we consider of a complexity similar to the car egress scenario (or inferior since the feet locations remain on the ground in the former case).
-For scenarios that apply, we also indicate the interpolation time that we obtained by feeding the plan to our interpolation method presented in \cite{Carpentier2016}.
-
-Despite the difficulty of comparing the methods, we believe the results presented in Table~\ref{tab:compprev} indicate that our planner holds the promise of planning complete trajectories significantly faster than previous works.
+Table~\ref{tab:compprev} presents the computation times we obtained for these scenarios, clearly demonstrating that our approach is order of magnitude faster that previous works.

### 2017-01-19 body/appendix_optim.tex
cosmetics

@@
 
-\subsection{The centroidal cone}
 We rewrite Eq.~\ref{eq:new_eul} in the general case:
@@
 where $\dot{L}$ is the angular momentum expressed at the com.
-Eq.~\ref{eq:new_eul} defines a 6 dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
+Eq.~\ref{eq:new_eul_acc} defines a 6 dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
 this cone determines all the admissible wrenches $\mathbf{w}$ such that the robot won't fall. The objective 
@@
 	\quad & \mathbf{A}_s \mathbf{w}(t)  \leq \mathbf{b}_s , \forall t \in  \left[t_s, t_s + \Delta t_s \right[ \\
+	\quad & \mathbf{D}_s \mathbf{c}(t)  \leq \mathbf{d}_s , \forall t \in  \left[t_s, t_s + \Delta t_s \right[ \\
 	\quad & \mathbf{c}(0)  = \mathbf{c}_{\mathbf{q}_0} \\
@@
 
-where $h$ is a cost function (We typically minimize the angular momentum and acceleration over the whole trajectory, and use
-a log barrier to constrain kinematically the center of mass position with respect to the contact locations).
+$h$ is a cost function (We typically minimize the angular momentum and acceleration over the whole trajectory, and use
+a log barrier to constrain kinematically the center of mass position with respect to the contact locations). $f$ is the colocation method
+bounding the position, velocity and acceleration of the center of mass. 
+$\mathbf{D}_s$ and $\mathbf{d}_s$ denote stacked kinematic constraints on the center of mass position, determined by the active contact locations.
+The inequalities for each contact are determined in the same way that the reachable workspace is computed in Appendix~\ref{app:rom}, with
+the effector serving as root (a triangulated object is a polytope, and as such can be described as a set of linear inequalities).
+
 This formulation can trivially be extended over the whole contact sequence.

### 2017-01-20 body/appendix_optim.tex
updated appendix optim

@@
 The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
+The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.
 To address the others scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox}), which can be integrated directly in our motion planner software.
-This implementation is slower, but handles collision avoidance in the trajectory generation phase.
-It uses a modified formulation of the optimization problem, that we present here.
+This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the cotnact forces were used.
+The center of mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.
 
@@
 \mathbf{G} \bm{\beta} = 
-\underbrace{\mat{m (\mathbf{\ddot{c}} - {g}) \\m (\mathbf{\ddot{c}} - {g}) + \dot{L}}}_{\mathbf{w}}
+\underbrace{\mat{m (\mathbf{\ddot{c}} - \mathbf{g}) \\m (\mathbf{\ddot{c}} - \mathbf{g}) + \dot{L}}}_{\mathbf{w}}
 %~ \underbrace{m (\mathbf{\ddot{c}} - \mathbf{g})  \\ m (\mathbf{\ddot{dc}} - \mathbf{g}) + \mathbf{\dot{L}} }_{\mathbf{w}}
@@
 Eq.~\ref{eq:new_eul_acc} defines a 6 dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
-this cone determines all the admissible wrenches $\mathbf{w}$ such that the robot won't fall. The objective 
-is thus to plan a trajectory for $c$ and $\dot{L}$ that remains inside $\mathcal{K}$. As a cone,
-$\mathcal{K}$ can be described as a linear inequality, obtained with the double description method~\citep{Fukuda1996}:
+this cone determines all the contact forces that generate an admissible wrench $\mathbf{w}$ such that the robot won't fall.
+The face form of $\mathcal{K}$ can be computed using the double description method~\citep{Fukuda1996}, resulting in the following linear inequalities:
 
@@
 
-We now consider a two contact configurations  $\mathbf{q}_0$ and $\mathbf{q}_1$ given by our planner: in the general case one contact is broken and one created to get from
-$\mathbf{q}_0$ to $\mathbf{q}_1$. The motion is thus constrained by three cones $\mathcal{K}_s, s =0 \dots 2$. We manually define the duration of each contact phase, of total
-time $T$, and formulate the following optimization problem:
+The objective 
+is then to plan a trajectory for the center of mass such that the generated $\mathbf{w}$ always verifies Eq.~\ref{eq:cone_k}. 
+We now consider two contact configurations  $\mathbf{q}_0$ and $\mathbf{q}_1$ computed by our planner: in the general case one contact is broken and one created to get from
+$\mathbf{q}_0$ to $\mathbf{q}_1$. The motion is thus constrained by three cones $\mathcal{K}_s, s =0 \dots 2$. We manually define the duration of each of the three contact phases.
+We call the total duration of the motion $T$, and formulate the following optimization problem:
 
@@
 
-$h$ is a cost function (We typically minimize the angular momentum and acceleration over the whole trajectory, and use
-a log barrier to constrain kinematically the center of mass position with respect to the contact locations). $f$ is the colocation method
-bounding the position, velocity and acceleration of the center of mass. 
+As a cost function, $h$ typically minimizes the angular momentum and acceleration variation over the whole trajectory. $f$ is the integration method
+that ties together the position, velocity and acceleration of the center of mass. 
 $\mathbf{D}_s$ and $\mathbf{d}_s$ denote stacked kinematic constraints on the center of mass position, determined by the active contact locations.
@@
 
-First, we plan a kinematic motion for the robot, subject to the contact constraints and center of mass constraints.
-This is achieved using a RRT planner. Using a sampling based planner guarantees that the effector will find a whole-body motion that is collision free.
+First, we plan a kinematic motion for the robot, subject to the contact constraints. We also constrain the center of mass to follow
+the computed trajectory. This is achieved using a constraint-based RRT planner~\citep{7759083}. As a result we obtain a whole-body motion that is collision free.
 
-Planning the whole-body takes approximatively 1.5 seconds for a complete contact transition.
+The entire resolution takes approximatively 1.5 seconds for a complete contact transition.
 
-This motion is then validated in our dynamic simulator~\citep{DelPrete2015b}, using the whole-body trajectory as a postural task to an inverse dynamics
+This motion is then validated in our dynamic simulator~\citep{DelPrete2015b}, using the trajectory as a reference postural task to an inverse dynamics
 solver, that accounts for the torque constraints of the robot. This final result is shown in the videos.

### 2017-01-20 cover2.tex
cover letter

@@
-\title{Cover letter}
+\documentclass[a4paper]{article}
+ \usepackage{fullpage}
+\usepackage{amsfonts} % pour les lettres maths creuses \mathbb
+\usepackage{amsmath}
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\usepackage[frenchb]{babel}
+\usepackage{aeguill}
+\usepackage{graphicx}
+\usepackage{hyperref}
+\usepackage{color}
+\usepackage{listings}
+\usepackage{pdfpages}
 
-\documentclass[12pt]{article}
+\newcommand\dx{\dot{x}}
+\newcommand\dy{\dot{y}}
+\newcommand\ddx{\ddot{x}}
+\newcommand\ddy{\ddot{y}}
+\def\real{{\mathbb R}}
+\frenchbsetup%
+{%
+StandardItemLabels=true,%
+ItemLabels=\ding{43},%
+}%
 
+\author {}
+\title {Cover letter for the IJRR submission \\ ``An efficient acyclic contact planner for multiped robots''}
+\date {}
 \begin{document}
@@
 
+Dear Editor, \\
+first of all we thank you for the time spent on considering our orignal manuscript. Indeed,
 %This cover letter for the manuscript ``An efficient acyclic contact planner for legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity.
-This cover letter introduces the first revision of the manuscript ``An efficient acyclic contact planner for legged robots'', resubmited for the ISSR'15 sepcial issue.
-As discussed with the Editor, we hoped to have reviewer continuity for the revision.
-The cover letter then plays more the role of a rebutal.
-We left the original cover letter at the end of this document.
+this cover letter introduces the first revision of the manuscript ``An efficient acyclic contact planner for legged robots'', resubmitted for the ISSR'15 sepcial issue.
+As discussed with prof. Hollerbach, we hoped to have reviewer continuity for the revision.
+The cover letter then plays more the role of a rebutal. We left the original cover letter at the end of this document, as well as the original ISRR reviews.
+Therefore, we kindly ask that this cover letter be forwarded to the reviewers of the new manuscript.
 
-We would first like to thank the reviewers for their previous feedback on the initial manuscript. We believe we have answered
+We would then like to thank the reviewers for their previous feedback on the initial manuscript. We believe we have answered
 all of their concerns, the main ones being that our method had not been tested on a real robot or a simulation, and that the paper was too verbose.
@@
 The paper is concerned with the problem of planning acyclic contact planning. 
-More precisely, we propose a very efficient method to address what we believe is the core of the difficulty of the contact planning problem: discovering the sequence of key contact posture that would lead to a safe and stable contact trajectory of the robot.
-Using the terminlogy of the paper, our contact planner addressed the two sub-problems $\mathcal{P}_1$ (finding a guide path for the robot root) and $\mathcal{P}_2$ (finding a sequence of stable contact postures and corresponding contact locations).
+More precisely, we propose a very efficient method to address what we believe is the core of the difficulty of the contact planning problem: discovering the sequence of key contact postures that would lead to a safe and stable contact trajectory of the robot.
+Using the terminology of the paper, our contact planner addressed the two sub-problems $\mathcal{P}_1$ (finding a guide path for the robot root) and $\mathcal{P}_2$ (finding a sequence of stable contact postures and corresponding contact locations).
 Our claim is that the solution of our planner leads to a straight-forward resolution of the last sub-problem $\mathcal{P}_3$ (finding the complete trajectory connecting the contact postures).
-In particular, our method is able to interactive performances, i.e. the planner needs less time to compute the solution than what is needed for the robot to execute it.
-This practical result corresponds to a dramatic improve of the state of the art, for example comparing to the capabilities of planning acyclic contact motions demonstrated by the humanoid teams during the Darpa Robotics Challenge last year.
+In particular, our method proposes interactive performances, i.e. the planner needs less time to compute the solution than what is needed for the robot to execute it.
+This practical result corresponds to a significant improvement of the state of the art, for example regarding the capabilities of acyclic contact planners demonstrated by the humanoid teams during the Darpa Robotics Challenge last year.
 
-Regarding this contribution, the reviews mostly criticized the structure of the paper (too verbose) and the unjustified claim that the contact sequence output by the planner leads to easy resolution of $\mathcal{P}_3$ hence complete resolution of the contact planning problem.
+Regarding this contribution, the reviews mostly criticized the structure of the paper (too verbose) and the unjustified claim that the contact sequence output by the planner leads to a feasible resolution of $\mathcal{P}_3$.
 
@@
 We also executed one of them on the real robot.
-This extension, along with the correction of the other remarks by the reviewers, have been integregated in the revision paper, that we also reworked to make it clearer, shorter and sharper.
+This extension, along with the correction of the other remarks by the reviewers, have been integregated in the revision paper, that we also reworked to make it clearer, shorter and sharper (the core of the paper is now 13 pages long, as opposed to 17 before) than the original manuscript.
 
-Details of these modifications follow.
+The details of these modifications follow.
 
@@
 While it was mentioned in the paper, we did not insist on these demonstrations because our main contribution focuses on the other problems. 
-We believe that, as reviewer 1 notices, this observation was due to the fact that we do not clearly define what we mean by a contact plan (i.e.
+We believe that, as reviewer 1 notices, this observation was due to the fact that we did not clearly define what we mean by a contact plan (i.e.
 a discrete sequence of key contact posture with the corresponding contact locations).
@@
 
-A rigorous definition of what we mean by contat plan is now detailed in section 5.1.
+A rigorous definition of a contact plan is now detailed in section 5.1.
 
@@
 In the paper, we are providing the success rates of our contact planner. 
-Such an empirical study has yet never been provided for other contact planners (arguabily because previous works did not reached the maturity level that would have permit such an empirical study).
+Such an empirical study has yet never been provided for other contact planners.
 Our method is not always successful, as very often with probabilistic searches.
-More precisely, the hypothesis taken by \mP1 may lead to infeasible \mP2 problem, and similarly \mP2 hypotheses may lead to infeasible to infeasible \mP3.
-We accept this fact and argue that it is mostly successful (i.e. more than 90\% of success in reasonable ``DRC-like'' scenarios).
+More precisely, the hypothesis taken by \mP1 may lead to an infeasible \mP2 problem, and similarly \mP2 hypotheses may lead to an infeasible \mP3.
+We accept this fact and argue that it is mostly successful (i.e. more than 77\% of success in reasonable ``DRC-like'' scenarios).
 Even in case of failure, the planner is based on random sampling that approximate probabilistic completeness, and can be restarted to obtain another candidate trajectory.
@@
 the end effector trajectories to avoid collisions with the environment, but we do not know how often. Considering this,
-we do not understand why we are criticized for providing our success rates, which actually demonstrate the interet of our approach.
+we do not understand why we are criticized for providing our success rates, which actually demonstrate the interest of our approach.
 
 \section{Originality and interest of the work}
-We strongly disagree with reviewer 1 regarding the originality of the work. Reviewer 1 states that we simply experiment with minor heuristic adjustments.
+We strongly disagree with reviewer 2 regarding the originality of the work. Reviewer 1 states that we simply experiment with minor heuristic adjustments.
 Bretl formalized well the problem without providing an algorithm applicable to the general case. Escande focused on the contact planning $\mathcal{P}_2$. 
 Our work tackles $\mathcal{P}_1$ as well, in a much more efficient manner than Bouyarmane et al.. 
-As for Hauser, these works are limited by the computational complexity of considering the full dimension of the problem.
+These works, including the one of Hauser, are limited by the computational complexity of considering the full dimension of the problem.
 
@@
 such as the work from Bretl and Escande.
-Work that do not make this assumption such as Mordatch and al.
+Works that do not make this assumption such as Mordatch and al.
 only offer local convergence and can't address these scenarios.
@@
 \end{document}
+
+\author {}
+\title {Original Cover letter for the IJRR submission \\ ``An efficient acyclic contact planner for multiped robots''}
+\date {}
+\begin{document}
+\maketitle
+
+Dear editor and reviewers, \\ \\
+
+The current submission is an extension to our conference paper ``A reachability-based planner for sequences of acyclic contacts in
+cluttered environments'', accepted to ISRR 15 (\url{http://stevetonneau.fr/files/publications/isrr15/isrr15.pdf}).
+
+In these two papers, our objective is to propose automatic methods
+to compute the contact sequence that characterizes a robot motion in
+cluttered environment.
+
+We joined to the present submission the reviews that were made on the
+conference paper. They follow the cover letter.
+
+Regarding the conference paper, this apper paper presents new contributions that we first highlight:
+
+\begin{itemize}
+\item We introduce a new criterion for asserting the static equilibrium of a robot configuration. It works for arbitrary
+legged robots, and handles any kind of contact configurations and surface orientation. It is formulated as a really efficient Linear Program, and 
+can be used to compute a value of robustness of the configuration. This robustness is useful for real robots applications, where many uncertainties
+in the state estimation of the robot can lead to incorrect evaluations of the equilibrium.
+\item We present two new heuristics that we use to bias our contact planner framework towards configurations relevant for the task being performed.
+\item We present a complete algorithm for contact planning that integrates the robust equilibrium criterion.
+\end{itemize}
+
+The extension of our work also addresses the main comments of the reviewers, and proposes lesser contributions,
+that we detail in the following lines.
+
+\begin{itemize}
+\item They were concerned by the fact that our demonstrations were
+considering virtual avatars and not actual robots. In the present
+submission, all our examples are applied to the HRP-2 and HyQ robots (Section 6).
+To achieve this, a new contribution is proposed in our paper:
+We introduce a robustness criterion for asserting the static equilibrium
+of a configuration (Section 5.1). We use it in combination with new heuristics that we
+propose to bias the contact generation process towards configurations
+relevant for the motion being performed (Section 5.2).
+With the exception of the dextrous hand, all the demonstrations shown in
+the paper and the video are entirely new.
+
+\item They also judged that the conference paper provided a theoretical
+description of our approach, but lacked the details on how to apply
+the method on an actual robot. We now explain the complete process
+for the HRP-2 in the present submission (Appendix A).
+
+\item Concerning remarks made on the reproducibility of the method,
+we now provide the complete pseudo code of the algorithm, as well as the
+source code of our application (Appendix B).
+
+\item The paper has been largely re-written: we provide a clearer
+comparison to the state of the art (Section 1.1); we clarify our statement regarding
+our computation times and compare our method with the state of the art (Section 6); we make a clear distinction between the
+theoretical concepts we propose and their approximation (Section 2).
+
+\item Regarding the computation time performances, compared to our conference paper, the planner is much more efficient.
+A fundamental change is the reimplementation of the algorithm as a Bi-RRT planner instead of the PRM
+approach. The code is also more efficient thanks to its complete rewriting into the HPP platform.
+
+\item Finally, we propose an analysis of the performances of our method and
+the success rates we obtained for each scenario (Section 6.2). This analysis provides
+a support for our argumentation on the strong choices made in the design
+of our planner (Section 1.2 and 2), and a comparison with methods from the literature.
+\end{itemize}
+
+We are convinced that these points not only provide a complete answer to
+the reviews of the conference paper, but also bring new contributions to
+the community, that justify a journal publication.
+
+%~ On a last note, as required by the IJRR submission form,
+%~ we proposed a list of editors and reviewers for the submission.
+%~ We affirm that we don't have conflicts of interest with the people we propose.
+%~ For complete clarity however, we want to mention that some of the authors
+%~ are proposing two workshops for the ICRA conference this year.
+%~ Among the names we proposed for reviews, Karen Liu, kris Hauser, and Quang-Cuong Pham have accepted to be invited speakers.
+
+We thank the editor and the reviewers for their time.
+
+Sincerely, \\
+
+The authors
+
+\includepdf[pages=-]{reviewisrr.pdf}
+
+\end{document}

### 2017-01-20 cover2.tex
cosmetics

@@
 \section{Originality and interest of the work}
-We strongly disagree with reviewer 2 regarding the originality of the work. Reviewer 1 states that we simply experiment with minor heuristic adjustments.
+We strongly disagree with reviewer 2 regarding the originality of the work when he states that we simply ``\textit{experiment with minor heuristic adjustments}''.
 Bretl formalized well the problem without providing an algorithm applicable to the general case. Escande focused on the contact planning $\mathcal{P}_2$. 
@@
 \section{On the static equilibrium of each configuration}
-Reviewer 1 stated that ``the planned contact path is only discrete with static equilibrium considered for each
-configuration, which is not very realistic''. First, we recall that this assumption is always made in the previous works that address cluttered scenarios,
-such as the work from Bretl and Escande.
+Reviewer 1 stated that ``\textit{the planned contact path is only discrete with static equilibrium considered for each
+configuration, which is not very realistic''}.
+While we agree with the observation, we first recall that this assumption is made in the previous works that address cluttered scenarios, such as the work from Bretl and Escande.
 Works that do not make this assumption such as Mordatch and al.

### 2017-01-20 body/contact.tex
Review 5.1 and 6.4

--- 2017-01-16 body/contact.tex
@@
 %~ is described in Appendix~\ref{sec:heuristics}.
+
+\subsection{Definition of a contact sequence}
+In previous contributions\adnote{Add reference here}, a contact plan is defined as a sequence of \deladp{configurations that describe a} quasi-static equilibrium configurations
+for each \repadp{contact}{existing} phase \deladp{of the motion}.
+For instance, a walk cycle would be described by three key configurations: a double-support configuration, a single-support configuration (a contact is broken), and another double-support configuration (a contact is created).
+Our definition of contact plan differs: between two consecutive configurations we allow both a contact break and a contact creation---if they are on the same effector. 
+In the previous example, our contact plan would simply consist of the two double-support configurations. 
+%However, if the contact broken and the contact created sequentially are not on the same effector, our planner outputs two distinct states.
+This representation is sufficient to describe all the contact constraints of the problem. Furthermore it removes the need to have single-support quasi-static configurations as in the example. 
+As shown in the companion video, this allows our framework to produce dynamic motions.
 

### 2017-01-20 body/results.tex
Review 5.1 and 6.4

@@
 To validate our plans, we either use the framework proposed by \citeauthor{Carpentier2016} or our own implementation of a $\mathcal{P}_3$ solver, detailed in Appendix~\ref{app:optim}.
-The obtained motions are available in the video.
+The companion video shows the obtained motions.
 
-Few contributions provided computation times on the complete problem, necessary to provide comparison between our approaches. 
+Few contributions provided computation times on the complete problem. %, necessary to provide comparison between our approaches. 
 Furthermore, $\mathcal{P}_3$ remains an open problem in the presence of obstacles.
-The only valid scenarios addressed completely in previous works are thus the stair climbing scenarios of different heights proposed by \citeauthor{Hauser06usingmotion}, and the table egress scenario by ~\citeauthor{DBLP:conf/iser/EscandeKMG08}, which we consider to be of similar complexity with respect to the car egress scenario (we did not consider the stairs in the scene).
-Both scenarios use HRP-2.
+The only valid scenarios addressed completely in previous works are thus the stair-climbing scenarios of different heights proposed by \citeauthor{Hauser06usingmotion}, and the table-egress scenario by~\citeauthor{DBLP:conf/iser/EscandeKMG08}, which we consider to be of similar complexity with respect to the car-egress scenario (we did not consider the stairs in the scene).
+Both scenarios are tested with HRP-2.
 
-Table~\ref{tab:compprev} presents the computation times we obtained for these scenarios, clearly demonstrating that our approach is order of magnitude faster that previous works.
+Table~\ref{tab:compprev} presents the computation times for these scenarios, clearly demonstrating that our approach is order of magnitude faster than previous works.

### 2017-01-20 body/appendix_optim.tex
Review appendix C

@@
+% !TEX root =  ../main.tex
 \section{Addressing $\mathcal{P}_3$}
@@
 The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
-The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.
-To address the others scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox}), which can be integrated directly in our motion planner software.
-This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the cotnact forces were used.
-The center of mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.
+\deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
+To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox}), which can be integrated directly in our motion planner software.
+This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
+The center-of-mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.
 
@@
 \mathbf{G} \bm{\beta} = 
-\underbrace{\mat{m (\mathbf{\ddot{c}} - \mathbf{g}) \\m (\mathbf{\ddot{c}} - \mathbf{g}) + \dot{L}}}_{\mathbf{w}}
+\underbrace{\mat{m (\mathbf{\ddot{c}} - \mathbf{g}) \\m \mathbf{c} \times (\mathbf{\ddot{c}} - \mathbf{g}) + \dot{L}}}_{\mathbf{w}}
 %~ \underbrace{m (\mathbf{\ddot{c}} - \mathbf{g})  \\ m (\mathbf{\ddot{dc}} - \mathbf{g}) + \mathbf{\dot{L}} }_{\mathbf{w}}
@@
 where $\dot{L}$ is the angular momentum expressed at the com.
-Eq.~\ref{eq:new_eul_acc} defines a 6 dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
-this cone determines all the contact forces that generate an admissible wrench $\mathbf{w}$ such that the robot won't fall.
+Eq.~\ref{eq:new_eul_acc} defines a 6-dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
+this cone determines all the admissible wrenches $\mathbf{w}$ that can be generated by contact forces inside their friction cones.
 The face form of $\mathcal{K}$ can be computed using the double description method~\citep{Fukuda1996}, resulting in the following linear inequalities:
@@
 We now consider two contact configurations  $\mathbf{q}_0$ and $\mathbf{q}_1$ computed by our planner: in the general case one contact is broken and one created to get from
-$\mathbf{q}_0$ to $\mathbf{q}_1$. The motion is thus constrained by three cones $\mathcal{K}_s, s =0 \dots 2$. We manually define the duration of each of the three contact phases.
+$\mathbf{q}_0$ to $\mathbf{q}_1$. We manually define the duration of each of the three contact phases.
+In each phase $s$ the centroidal wrench $\mathbf{w}$ is constrained to lie inside a cone $\mathcal{K}_s, s =0 \dots 2$.
 We call the total duration of the motion $T$, and formulate the following optimization problem:
@@
 %~ \find \quad & \mx{R} \in \Rv{3\times T}, \mxdd{C} \in  \Rv{6\times T} \\
-\minimize_{\mathbf{c}(t), \mathbf{w}(t)}  \quad & \sum_{s=0}^{2} \int_{t_s}^{t_s + \Delta t_s}  h(\mathbf{c}, \mathbf{w})  \\
-\st \quad & \mathbf{\dot{c}} = f(\mathbf{c}, \mathbf{w}) \\
-	\quad & \mathbf{A}_s \mathbf{w}(t)  \leq \mathbf{b}_s , \forall t \in  \left[t_s, t_s + \Delta t_s \right[ \\
-	\quad & \mathbf{D}_s \mathbf{c}(t)  \leq \mathbf{d}_s , \forall t \in  \left[t_s, t_s + \Delta t_s \right[ \\
+\minimize_{\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)}  \quad & \sum_{s=0}^{2} \int_{t_s}^{t_s + \Delta t_s}  h(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)) \text{dt}  \\
+\st \quad & \mathbf{\ddot{c}}(t) = f(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{w}(t)) \\
+	\quad & \mathbf{A}_s \mathbf{w}(t)  \leq \mathbf{b}_s , \forall t \in  \left[t_s, t_s + \Delta t_s \right[, \forall s \\
+	\quad & \mathbf{D}_s \mathbf{c}(t)  \leq \mathbf{d}_s , \forall t \in  \left[t_s, t_s + \Delta t_s \right[, \forall s \\
 	\quad & \mathbf{c}(0)  = \mathbf{c}_{\mathbf{q}_0} \\
@@
 
-As a cost function, $h$ typically minimizes the angular momentum and acceleration variation over the whole trajectory. $f$ is the integration method
-that ties together the position, velocity and acceleration of the center of mass. 
+The cost function $h$ is a weighted sum of the angular momentum and center-of-mass acceleration variation over the whole trajectory. 
+The system dynamics $f$ ties together the position, velocity and acceleration of the center of mass. 
 $\mathbf{D}_s$ and $\mathbf{d}_s$ denote stacked kinematic constraints on the center of mass position, determined by the active contact locations.
 The inequalities for each contact are determined in the same way that the reachable workspace is computed in Appendix~\ref{app:rom}, with
-the effector serving as root (a triangulated object is a polytope, and as such can be described as a set of linear inequalities).
+the effector serving as root. % (a triangulated object is a polytope, and as such can be described as a set of linear inequalities).
 
@@
 
-As an output, the solution of this optimization problem gives us a admissible center of mass trajectory.
-To compute the whole body motion, we use a two step approach.
+The output of this optimization problem is an admissible center-of-mass trajectory.
+To compute the whole body motion, we use a two-step approach.
 
 First, we plan a kinematic motion for the robot, subject to the contact constraints. We also constrain the center of mass to follow
-the computed trajectory. This is achieved using a constraint-based RRT planner~\citep{7759083}. As a result we obtain a whole-body motion that is collision free.
+the computed trajectory. This is achieved using a constraint-based RRT planner~\citep{7759083}. As a result we obtain a collision-free whole-body motion.
 
@@
 
-This motion is then validated in our dynamic simulator~\citep{DelPrete2015b}, using the trajectory as a reference postural task to an inverse dynamics
-solver, that accounts for the torque constraints of the robot. This final result is shown in the videos.
+This motion is then validated in our dynamic simulator~\citep{DelPrete2015b}, using the trajectory as a reference postural task for an inverse-dynamics
+controller that accounts for all the robot constraints. This final result is shown in the companion video.

### 2017-01-20 body/contact.tex
Review appendix C

@@
 \subsection{Definition of a contact sequence}
-In previous contributions\adnote{Add reference here}, a contact plan is defined as a sequence of \deladp{configurations that describe a} quasi-static equilibrium configurations
-for each \repadp{contact}{existing} phase \deladp{of the motion}.
+In previous contributions\adnote{Add reference here}, a contact plan is defined as a sequence of quasi-static equilibrium configurations
+for each contact phase.
 For instance, a walk cycle would be described by three key configurations: a double-support configuration, a single-support configuration (a contact is broken), and another double-support configuration (a contact is created).
@@
 %However, if the contact broken and the contact created sequentially are not on the same effector, our planner outputs two distinct states.
-This representation is sufficient to describe all the contact constraints of the problem. Furthermore it removes the need to have single-support quasi-static configurations as in the example. 
+This representation is sufficient to describe all the contact phases. Furthermore it removes the need to have single-support quasi-static configurations as in the example. 
 As shown in the companion video, this allows our framework to produce dynamic motions.

### 2017-01-23 body/abstract.tex
andrea s last modif plus 250 words abstract

@@
 
-We tackle this issue with our novel  idea, the reachability condition. This condition allows us to define
+We tackle this issue by introducing the reachability condition. This condition allows us to define
 a geometric approximation of the contact manifold, which is of low dimension and can be sampled efficiently.
@@
 Thanks to this condition we decompose the hard contact planning problem into two
-simpler sub-problems: first, planning a guide path for the robot root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a new deterministic
-contact-selection algorithm.
+sub-problems: first, planning a guide path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm.
 
 Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically: in a
-few seconds, with satisfying success rates, we generate complex contact plans for various scenarios and robots, namely
-HRP-2, HyQ, and a dexterous hand. These plans are then validated either in dynamic simulations, or on the real HRP-2 robot, thus demonstrating 
+few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ, and a dexterous hand.
+These plans are then validated either in dynamic simulations, or on the real HRP-2 robot, thus demonstrating
 the first interactive multi-contact planner.

### 2017-01-23 body/appendix_optim.tex
andrea s last modif plus 250 words abstract

@@
 The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
-\deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
+%~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
 To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox}), which can be integrated directly in our motion planner software.

### 2017-01-23 body/conclusion.tex
andrea s last modif plus 250 words abstract

@@
 
-The main source of computational cost identified in previous works is the verification of \equilibriumfeasibility. 
+Another source of computational cost identified in previous works is the verification of \equilibriumfeasibility. 
 The strong assumption of our work is that for the class of problems we consider \contactreachability\ implies \equilibriumfeasibility.
@@
 Figure~\ref{fig:dedefeas} illustrates this observation, demonstrated empirically by the third column of Table~\ref{tab:requestpercent}. In the worst case, in our experiment
-the assumption was verified for 82\% of the total amount of trials.
-
+the assumption was verified for 82\% of the total amount of trials that verified \contactreachability\.
 In the example of \cite{Bouyarmane2009}, the verification of \equilibriumfeasibility\ implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$, requiring
@@
 
-These results clearly justify the pragmatic decisions of this paper. What is lost in completeness is largely compensated by the computation times we obtain.
+These results clearly justify the pragmatic decisions of this paper.
 

### 2017-01-23 body/contact.tex
andrea s last modif plus 250 words abstract

@@
 \subsection{Definition of a contact sequence}
-In previous contributions\adnote{Add reference here}, a contact plan is defined as a sequence of quasi-static equilibrium configurations
+In previous contributions~\citep{DBLP:conf/iser/EscandeKMG08}, a contact plan is defined as a sequence of quasi-static equilibrium configurations
 for each contact phase.

### 2017-01-23 cover2.tex
[COVER] Added tabular to comment.

@@
 \usepackage{pdfpages}
+\usepackage{longtable}
 
@@
 
-Dear Editor, \\
-first of all we thank you for the time spent on considering our orignal manuscript. Indeed,
+Dear Editor, \medskip
+
+First of all we thank you for the time spent on considering our original manuscript. Indeed,
 %This cover letter for the manuscript ``An efficient acyclic contact planner for legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity.
-this cover letter introduces the first revision of the manuscript ``An efficient acyclic contact planner for legged robots'', resubmitted for the ISSR'15 sepcial issue.
-As discussed with prof. Hollerbach, we hoped to have reviewer continuity for the revision.
-The cover letter then plays more the role of a rebutal. We left the original cover letter at the end of this document, as well as the original ISRR reviews.
+this cover letter introduces the first revision of the manuscript ``An efficient acyclic contact planner for legged robots'', resubmitted for the ISSR'15 special issue.
+As discussed with Prof. Hollerbach, we hope to have reviewer continuity for the revision.
+The cover letter then plays more the role of a rebuttal. We left the original cover letter at the end of this document, as well as the original ISRR reviews.
 Therefore, we kindly ask that this cover letter be forwarded to the reviewers of the new manuscript.
@@
 \section{Recall of main paper contributions}
-The paper is concerned with the problem of planning acyclic contact planning. 
+The paper is concerned with the problem of planning acyclic contact movements. 
 More precisely, we propose a very efficient method to address what we believe is the core of the difficulty of the contact planning problem: discovering the sequence of key contact postures that would lead to a safe and stable contact trajectory of the robot.
@@
 
-In the revision, we implemented a solver inspired by the state of the art to solve $\mathcal{P}_3$ and empiricially proved the claim that our $\mathcal{P}_1$-$\mathcal{P}_2$ planner leads to easily solving $\mathcal{P}_3$.
-All the robot contact sequences reported in the paper have been converted to full contact trajectories using this $\mathcal{P}_3$ solution.
+In the revision, we implemented a solver inspired by the state of the art to solve $\mathcal{P}_3$ and empirically proved the claim that our $\mathcal{P}_1$-$\mathcal{P}_2$ planner leads to easily solving $\mathcal{P}_3$.
+All the robot contact sequences reported in the paper have been converted to full contact trajectories using this $\mathcal{P}_3$ solution, that were validated in an contact simulator.
 We also executed one of them on the real robot.
-This extension, along with the correction of the other remarks by the reviewers, have been integregated in the revision paper, that we also reworked to make it clearer, shorter and sharper (the core of the paper is now 13 pages long, as opposed to 17 before) than the original manuscript.
+This extension, along with the correction of the other remarks by the reviewers, have been integrated in the revision paper, that we also reworked to make it clearer, shorter and sharper (the core of the paper is now 13 pages long, as opposed to 17 before) than the original manuscript.
 
@@
 \section{Validation of the method (addressing  $\mathcal{P}_3$) }
-In his review, reviewer 2 stated that our plans are likely to be not executable, even in simulation.
+In his review, Reviewer 2 stated that our plans are likely to be not executable, even in simulation.
 
@@
 While it was mentioned in the paper, we did not insist on these demonstrations because our main contribution focuses on the other problems. 
-We believe that, as reviewer 1 notices, this observation was due to the fact that we did not clearly define what we mean by a contact plan (i.e.
+We believe that, as Reviewer 1 notices, this observation was due to the fact that we did not clearly define what we mean by a contact plan (i.e.
 a discrete sequence of key contact posture with the corresponding contact locations).
@@
 Our method is not always successful, as very often with probabilistic searches.
-More precisely, the hypothesis taken by \mP1 may lead to an infeasible \mP2 problem, and similarly \mP2 hypotheses may lead to an infeasible \mP3.
+More precisely, the hypothesis taken by \mP1 may lead to an unfeasible \mP2 problem, and similarly \mP2 hypotheses may lead to an unfeasible \mP3.
 We accept this fact and argue that it is mostly successful (i.e. more than 77\% of success in reasonable ``DRC-like'' scenarios).
 Even in case of failure, the planner is based on random sampling that approximate probabilistic completeness, and can be restarted to obtain another candidate trajectory.
-In practice, it can always find a feasible solution fast enough for online applications by replanning upon failure.
+In practice, it can always find a feasible solution fast enough for online applications by re-planning upon failure.
 
-In previous works, it is in fact unknown whether the methods are always successful or not. We simply have no idea. For instance, the CIO of Mordatch,
+Reading the papers reporting previous works in contact planning, it is in fact not discussed whether the methods are always successful or not.
+We simply have no idea.
+For instance, the CIO of Mordatch,
 because of the local convergence of the method, \textbf{necessarily} fails in some scenarios. We just do not know how often. Similarly, Escande et al. had to manually edit 
@@
 \section{Originality and interest of the work}
-We strongly disagree with reviewer 2 regarding the originality of the work when he states that we simply ``\textit{experiment with minor heuristic adjustments}''.
+We strongly disagree with Reviewer 2 regarding the originality of the work when he states that we simply ``\textit{experiment with minor heuristic adjustments}''.
 Bretl formalized well the problem without providing an algorithm applicable to the general case. Escande focused on the contact planning $\mathcal{P}_2$. 
-Our work tackles $\mathcal{P}_1$ as well, in a much more efficient manner than Bouyarmane et al.. 
+Our work tackles $\mathcal{P}_1$ as well, in a much more efficient manner than Bouyarmane et al.
 These works, including the one of Hauser, are limited by the computational complexity of considering the full dimension of the problem.
+In practice, the underlying complexity imply that they cannot lead to implementation that can be used in practical scenarios, as emphasized by the lack of such contact solvers during the DRC (as discussed by Atkinson).
 
-The reachability condition provides a dimensionality reduction of the problem. This is clearly an originality regarding the previous work who always consider the whole-body problem.
-We believe that this reduction is the only way to make the contact planning problem tractable, in the same way that many researchers (such as ourselves, but also Wieber, Tedrake, Righetti and others) use dimensionality reduction successfully to address
-the trajectory optimization problem $\mathcal{P}_3$.
+We claim that the reachability condition provides a dimensionality reduction of the problem. This is clearly an originality regarding the previous work who always consider the whole-body problem.
+We believe that this reduction is the only way to make the contact planning problem tractable, in the same way that many researchers (such as ourselves, but also Wieber, Tedrake, Righetti and others) use dimensionality reduction successfully to address the trajectory optimization problem $\mathcal{P}_3$.
 
@@
 the different concepts that we use in the paper, and reduced their number.
-We believe that the paper is much more clearer in its current form.
+We believe that the paper is much clearer in its current form.
 
@@
 \section{Comparison with other methods}
-As noted by reviewer 2, in our previous paper it was indeed complex to compare our work with previous contributions,
-since they were adressing the complete problem. We now give the complete computation times and compare them with previous works,
+As noted by Reviewer 2, in our previous paper it was indeed complex to compare our work with previous contributions,
+since they were addressing the complete problem. We now give the complete computation times and compare them with previous works,
 clearly demonstrating that our method outperforms them by orders of magnitude (Section 6.4).
@@
 Works that do not make this assumption such as Mordatch and al.
-only offer local convergence and can't address these scenarios.
+only offer local convergence and cannot address these scenarios (e.g.
+on the ``Polaris car egress'' scenario reported in our paper).
 Furthermore, this limitation is in fact less restrictive in our approach that the one of Escande, since while our plan is quasi static, our resulting motion is dynamic, as now explained in the paper (Section 5.1).
@@
 \section{Actual car model of the DARPA Challenge}
-As asked by reviewer 2, we now use the polaris car model for demonstrating our egress scenario.
+As asked by Reviewer 2, we now use the Polaris car model for demonstrating our egress scenario.
+
+\newpage
+\section{Recap of Reviewer comments}
+
+\subsection{Reviewer 1}
+\noindent
+\begin{longtable}{|p{21em}|p{21em}|}
+\hline
+The paper addresses an important problem of planning robot motion involving contacts.
+It extends the idea of planning acyclic contacts by decoupling the issue into (1) planning a guide path for the root of the robot, and (2) planning a discrete sequence of acyclic and static equilibrium contact configurations along the path, with an interpolation of motion between two consecutive contact postures in the sequence.
+The introduced planner considers both collision avoidance and maintenance of desired contacts and aims at real-time or interactive performance.
+&Thanks for the comment.
+\\ \hline % ----------------------------------------------------------------------------------
+For the decoupled approach to work, the planned guide path must guarantee that a sequence of stable contact configurations is associated with it to execute the path with desired contacts and yet without undesired collisions; such a guide path is called “equilibrium feasible”.
+The issue is how to find such a guide path efficiently.
+One key assumption of the paper is that for cluttered environments, “most of the times” a guide path that is “contact feasible” (referring to that at least one end-effector is in contact with the environment and there is no undesired collision between the robot and the environment) is also “equilibrium feasible”.
+It further assumes that if the reachable space of the robot along the guide path is in collision with the environment but the robot root configuration is not in collision, called the “reachability condition”, then the guide path is “contact feasible”.
+With those assumptions, the paper treated guide path planning as a geometric problem and used an RRT planner to solve the problem.
+& We agree with this summary.
+\\ \hline % ----------------------------------------------------------------------------------
+However, the question is whether the above key assumption (of 
+“contact feasible” implies “equilibrium feasible”) is reasonable.
+The paper did not provide justification for the assumption, and the assumption does not seem intuitive.
+The paper used terms such as “most of the times” and “cluttered contact planning” without clear and quantifiable definitions.
+With the above assumptions, planning becomes quite straightforward, and the planning algorithm used is rather conventional.
+&
+We modified the paper to more rigorously discuss these hypotheses. 
+We also believe that the main justification of their validity is through statistical tests, as reported in the result section.\newline
+See Section 6 for details. 
+\\ \hline % ----------------------------------------------------------------------------------
+For the next problem of planning a sequence of contact configurations given a “contact feasible” root path, the paper further decomposed the problem by planning the configuration of one limb at a time, assuming the root configuration and the other limb configurations were fixed.
+For each limb, N sampled configurations were considered (sampled off-line), and they were indexed by the corresponding end-effector position and stored in an octree beforehand.
+Because all that was done off-line, it is not clear how N was decided. 
+The on-line planning step would search through the stored limb configurations (sorted according to some heuristic), one limb at a time (it seems), to find an almost-contact configuration and create a feasible limb contact configuration out of it.
+Different limbs were considered in some deterministic order: the limb that was contact-free the longest was considered first, and contact configurations were generated as long as the limb could maintain contact.
+It is a bit confusing when interpolated contact configurations would be added between two adjacent contact configurations.
+Was it “in the hope” to keep a limb in contact or change limbs in contact?
+& 
+--- I THINK WE MUST DISCUSS THAT HERE ---
+\\ \hline % ----------------------------------------------------------------------------------
+The manipulator Jacobian for each limb was assumed full rank, and no singular configurations were considered. 
+&
+--- I THINK WE MUST DISCUSS THAT HERE ---
+\\ \hline % ----------------------------------------------------------------------------------
+The main issue with the proposed planner is that assumptions (commented earlier) and decisions were made in the planner conveniently in order to simplify the problem but without much justification, and there is no guarantee of a solution.
+The entire approach is heuristic and somewhat ad hoc.
+&
+We agree that our approach is mostly heuristic.
+Heuristics are somehow underestimated in robotics, where optimality principles are often preferred (this is not the case in other fields, like discrete optimization, data routing or multi-agent diagnosis in web applications).
+
+The main hypotheses of the papers are: the division of contact planning in three problems; the reachability condition approximating the equilibrium condition; and the heuristics for building the contact sequence (\mP2) along the guide path (\mP1).
+We believe that these main hypotheses are properly discussed, some being theoretically justified (reachability condition), the other being statistically asserted.
+
+We agree that the method also requires some other minor choices, like the ones discussed by the reviewers. 
+We do not believe that these minor technical choices are important. 
+If other choices are made, similar results would likely be obtained as well.
+To keep the paper as short as possible, we did not spent time explaining these choices.
+Any reader that would try to re-implement our approach would find enough details to reproduce our choices.
+The source code of the planner is also available (url in the paper) in case the exact implementation is needed.
+
+Providing more details about any single technical choice would require 10 more pages, which is in contraction to the request of the reviewer to have the paper less verbose.
+
+In the revision, we emphasized what are the key assumptions, and spent more time to justify them properly.
+In the appositive, we more clearly pointed what assumptions are the results of technical choices and do not impact the solver performances.
+
+\\ \hline % ----------------------------------------------------------------------------------
+Another major limitation is that the planned contact path is only discrete with static equilibrium considered for each configuration, which is not very realistic.
+The paper also pointed out this limitation in the Conclusions.
+&
+Our contribution is on solving \mP1-\mP2 efficiently. 
+We believe that this is the core of the contact planning problem, as argued in introduction.
+Indeed, solving \mP3 in all the cases studied in the paper was quite straight-forward.
+
+Clearly, dividing the complete problem in 3 subproblems will ease the solver and possibly obstruct some solutions. The question is whether this choice is a sane trade-off or not. 
+We insist that this division is quite classical in the state of the art.
+In particular, the only planner able to cope with this division (by Mordatch et al) only leads to local convergence.
+
+In the new version, we asserted the realism of the contact sequences by implementing a solution to \mP3, outputing full trajectories that were tested in a contact simulator and on the real HRP-2 robot.
+
+See Sections 2 and 6 for details.
+
+\\ \hline % ----------------------------------------------------------------------------------
+The important problem of transitioning motion from one contact configuration to the next was left out as beyond the scope of the paper. 
+&
+--- I THINK WE MUST DISCUSS THAT HERE ---
+\\ \hline % ----------------------------------------------------------------------------------
+The planner was tested in a number of diverse simulation examples, and there were related comparisons with existing work.
+This is commendable.
+However, there is no experimental result on a real robot.
+&
+The contribution of the paper is on subproblems \mP1 and \mP2. 
+We initially did not put robot results to emphasize this aspect and avoid confusing the reader.
+Following the reviewer recommendations, we added a solver of subproblem \mP3, leading to complete movement, one of them being executed on the real robot. 
+\newline See Section 2 for details.
+\\ \hline % ----------------------------------------------------------------------------------
+The writing of the paper can be much improved and made much more concise.
+The description of contact path planning in Section 4 is broken into different sub-sections, but not entirely in the order of how the sequence is generated, which is confusing.
+In Section 4.1, steps 3—6 talk about on-line generation, but the related decisions were introduced in Section 4.2.1 and 4.2.2, and none of those sub-sections has a title that clearly indicates the planning procedure.
+The paper has used unnecessary notations and symbols on the one hand, but vague languages such as “most of the time”, “cluttered problems”, “in the hope”, etc., on the other hand.
+Some symbols were not used consistently, for instance, m is used both to indicate mass and a direction.
+&
+Thanks for carefully listing these problems. 
+The paper was completely rewritten to make it shorter and more rigorous.
+\\ \hline % ----------------------------------------------------------------------------------
+\end{longtable}
+
+\newpage
+\subsection{Reviewer 2}
+\noindent
+\begin{longtable}{|p{21em}|p{21em}|}
+\hline
+The paper presents a method for producing contact plans for the locomotion of multiped robots.
+It decomposes the problem into two sub-problems : the computation of a guide path is which the robot is floating close enough to the environment, and subsequently the positioning of contacts on the environment along the guide.
+& Thanks for this summary, that we agree with.
+\\ \hline % ----------------------------------------------------------------------------------
+My major concern about this piece of work is its lack of originality.
+The problem seems to have been well formalized and studied in numerous previous works (Hauser et al, Escande et al, etc.).
+The authors experiment with minor heuristic adjustments.
+I do not consider them as major contributions that warrant a publication in IJRR by the standards of the journal.
+The authors end up with the main claim of the paper being a reduced computation time in comparison with these previous works.
+As admitted by the authors themselves, even this gain in computation time is not really comparable to the previous works since the outputs are different (previous works go all the way to the motion generation, and account for different constraints).
+&
+We do not agree with this comment.
+The main contribution of our method are the reduction hypothesis that we are using to break the complexity of contact planning.
+The problem reduction, leading to pragmatic implementation, is a step-change in how the problem of contact planning should be considered.
+
+Although we clearly acknowledge the contribution of previous works, in particular by Bretl, Hauser, Escande, Bouyarmane and their co-authors, they are all mostly nonpractical contributions.
+The  planners resulting of their works are interesting to study, but can barely be used in practice, in large part due to their computational complexity. 
+Even the work of Mordatch and colleagues, which is solving the problem only locally and partially (no collision and self-collision checking, partial dynamics, etc) requires minutes of computations for seconds of execution.
+
+We can clearly compare the performances of our planner with previous methods (even if not as accurately than what we would like): our method outperforms them by orders of magnitude.
+Even if considering our contribution only on the practical aspect of computation time,  it makes the difference between a method that can be used at the Darpa challenge or not, as we show in our introduction.
+
+\\ \hline % ----------------------------------------------------------------------------------
+The work is largely incomplete, and far from being self-sufficient.
+A quick glance at the sequences displayed in the video reveals that the "motion patterns" (and I understand these are not "motions" per se, but "contact plans") are not physically consistent (even regarding them as mere contact plans and not as the final motions).
+The robot moves a contact without needing to readjust the posture in order to release that contact (or "set it free", by nullifying the contact forces on that contact).
+This results in sequences in which the contacts seem to be "slided" rather than removed and repositioned.
+This denotes a significant flaw in the algorithm.
+At no point in the paper do the authors explicitly define what a "contact plan" is.
+Hence we do not know what is the output plan of their framework they are trying to obtain (obviously that output is not the same as that of previous works which feature the above-mentioned "releasing of contacts" property in their own definitions of "contact plans").
+The authors should clearly define what a valid output plan of their framework is.
+&
+Once more, we cannot agree with this comment.
+We believe that the comment is partly due to a problem of definition of what we mean by ``contact sequence''.
+The solution to subproblem \mP2 is a discrete sequence of static posture in stable contact (as detailed in Section 2).
+All the sequences reported in the paper are realistic in that sense.
+
+In the original paper, we claimed that any of these contact sequences would directly lead to a full contact trajectory (\mP3).
+However, we did not asserted this claim. 
+In the new version, we implemented a solver \mP3 and validated that all the contact sequences reported in the paper are leading to full contact trajectories, whose dynamic plausibility has been checked by execution on the real robot for one of them, and in a dynamic contact simulator for the others.
+
+See Sections 2 and 6 for details.
+
+\\ \hline % ----------------------------------------------------------------------------------
+There is no way to assess the validity of the work, as opposed to the previous works.
+From what I can judge from the video, the motions will not be able to be performed with a real robot nor even in physics simulation.
+The only way to legitimately argue with the latter statement is by providing a demonstration of the execution of the plans in simulation or with the real robots.
+I think non-trivial work still needs to be accomplished to "plug" the output of the presented framework into any existing motion generator or controller.
+If the only envisioned application of the framework is the coming future work of the authors to turn it into a valid motion controller, then I think the work would be complete only by then and its validity assessable from a scientific standpoint (i.e.
+not speculation) only at that stage.
+&
+We agree that this remark is justified by the lack of facts in our original paper.
+Our claim was that any contact sequence output by our planner easily lead to full contact trajectory.
+We wrongly assumed that this was an evidence when looking at the contact sequences, and neglected to prove it.
+
+In the revision, we have implement a solve \mP3 and converted all contact sequences in full contact trajectory, validated in simulation or on the real robot.
+We must insist that this validation was straight-forward, on a scientific point of view. 
+As initially stated, the produced contact sequences indeed lead directly to full trajectories.
+
+See Section 2 for details.
+
+\\ \hline % ----------------------------------------------------------------------------------
+There are many algorithm weakening assumptions.
+The most obvious one being the replacement of the equilibrium feasibility condition with the contact feasibility condition, then replacing the latter with the reachability condition.
+I understand that the authors assume that these approximations are valid, especially in "cluttered" environment.
+However I believe it should not be too difficult to think of counter-example problems in which each of this approximations creates a degenerate plan.
+&
+We agree with this remark. 
+The reduction models that we propose are valid approximations in a reasonable domain of the robot environment.
+Experimentally, we demonstrated that this validity domain is quite large.
+It is yet possible to find counter-examples, and we cite some of them in the paper.
+Our planner is only a contribution to the problem, not the ultimate solution to the contact-planning problem.
+
+In terms of genericity, two key aspects must be considered.
+First, our planner is good enough for 90\% of the situations that nowadays legged robot can meet.
+Second, the general idea of reducing the contact planning problem is valid in any situations that we can imagine, and the heuristics and approximate models that we proposed can be adapted to extend the planner capabilities.
+Automatic synthesis of these models is indeed a research direction that we are currently considering.
+
+In the revision, we now justified these assumptions by theoretical results, explanations about the rationale behind the assumption, and statistical validation reported in the experimental results.
+\\ \hline % ----------------------------------------------------------------------------------
+The joint torque limit constraint is absent from the static equilibrium condition.
+This is a major approximation that does indeed speed up computation with respect to this constraint being accounted for in some of the compared works, but results in weaker, less physically valid plans (again the only way to argue with that is to show that the motions are performed in real or in simulation)
+&
+We agree that the torque limits are not considered by our planner.
+However, although this is a hard limit on the robot, we do not agree that it plays a dramatic role in contact planning.
+Indeed, the torque limits are considered in our \mP3 validation, and does not negatively impact the execution of our contact sequences.
+We state and experimentally verified that considering the torque limits at the planning level is not necessary.
+\\ \hline % ----------------------------------------------------------------------------------
+"These two contributions were required to apply our method to real robot models", the use of "real robot" is somewhat misleading there.
+&
+By ``real robot'', we meant a realistic modeling of the robot, with true joint limits, collision constraint, etc. 
+During rewriting, this particular sentence was removed.
+We now used ``real robot'' only to discuss the experiments we did on the physical robot device.
+\\ \hline % ----------------------------------------------------------------------------------
+A minor remark : since the DARPA challenge is one of the motivations of the work then I do not understand why the authors did not use the actual model of the car used in the DARPA challenge and used the toy model instead ?
+A great part of the difficulty is removed there, numerous collisions occur with detail parts of the car (steering wheel, hand break, outside frame, etc) and contribute to the challenging aspect of the problems.
+&
+We used the simplified car model because the rendering seemed clearer to read.
+As requested by the reviewer, we implemented the same movement on the Polaris model taken from the DRC.
+Our HRP-2 robot being to small to go down the car (compared to DRC HRP-2 ``Kai'' robot), we added two steps.
+\\ \hline % ----------------------------------------------------------------------------------
+Finally, in the presentation department, I found the paper to be excessively long, I do not know what are the pages recommendations of the journal, but he paper is too verbose, the algorithms too partitioned into different sections and subsections.
+I think a major rewriting and restructuring is possible and the ideas can be conveyed in a more synthetic manner.
+&
+The paper was rewritten, see Section 5.
+\\ \hline % ----------------------------------------------------------------------------------
+Overall, I think the work is very promising, and I hope that my detailed comments and questions show my enthusiasm and interest of the subject.
+I understand that all the approximations and heuristics are driven by a sense of  pragmatism and practicality rather than pure theoretical soundness, which are probably legitimate motivations in their own right.
+However I also think that the work is still not mature enough, and will largely benefit from a validation of these approximations with a real robot or simulation experiment after the framework is complete.
+I encourage the authors to continue along the completion of the work.
+&
+It was indeed difficult to fully feel the enthusiasm of the reviewer, although we definitely appreciated his detailed comment and effort to improve the quality of the paper.
+In conclusion, we truly believe that our solution answer to the main difficulty of contact planning, by building a realistic contact sequence which easily leads to the construction of a complete contact trajectory, and behind, the execution of it on the real robot.
+The addition of a \mP3 solver in the revised paper experimentally justifies this claim.
+The results that we demonstrated have never been reached so far by any previous works, neither in term of qualitative results, number of reported scenario, statistical analysis of the algorithms or combinatorial efficiency.
+The contribution that we are reporting is not a marginal increment or a practical implementation, but a step-change in how the contact planning problem should be considered, leading to the first realistic implementation.
+We hope that the revision of the paper will help the reviewer to meet our opinion.
+
+\\ \hline % ----------------------------------------------------------------------------------
+\end{longtable}
 
@@
 
-Regarding the conference paper, this apper paper presents new contributions that we first highlight:
+Regarding the conference paper, this appear paper presents new contributions that we first highlight:
 
@@
 relevant for the motion being performed (Section 5.2).
-With the exception of the dextrous hand, all the demonstrations shown in
+With the exception of the dexterous hand, all the demonstrations shown in
 the paper and the video are entirely new.
@@
 \item Regarding the computation time performances, compared to our conference paper, the planner is much more efficient.
-A fundamental change is the reimplementation of the algorithm as a Bi-RRT planner instead of the PRM
+A fundamental change is the re-implementation of the algorithm as a Bi-RRT planner instead of the PRM
 approach. The code is also more efficient thanks to its complete rewriting into the HPP platform.

### 2017-01-23 body/efort.tex
Review section 5
complete paper
no symbol redefinition

@@
+\section{Heuristics and criteria for contact selection}
+\label{sec:heuristics}
 % !TEX root =  ../main.tex
@@
 	\end{overpic}
-  \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 dofs and 2 segments.
+  \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 DoFs and 2 segments.
 Only the horizontal and vertical speeds are shown (not the rotation speeds).}
@@
 We thus consider the force ellipsoid as a basis for our heuristic.
-In a given direction $\mathbf{m}$, the length of the ellipsoid is given by the force-transmission ratio \citep{1087795}:
+In a given direction $\uprho$, the length of the ellipsoid is given by the force-transmission ratio \citep{1087795}:
 \begin{equation*}
-f_\mathsf{T}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}	^{T})\mathbf{m}]^{-\frac{1}{2}}
+f_\mathsf{T}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}	^{T})\uprho]^{-\frac{1}{2}}
 \end{equation*}
 
-In our problem, to compare candidate configurations, we include the quality of the contact surface, and choose $\mathbf{m}$ as the direction
+In our problem, to compare candidate configurations, we include the quality of the contact surface, and choose $\uprho$ as the direction
 opposite to the local motion (given by the difference between two consecutive root positions):
@@
 \begin{equation}
-h_{\textrm{\it EFORT}}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \mathbf{m})
+h_{\textrm{\it EFORT}}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
 \end{equation}
@@
 \begin{equation}
-h_{vel}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{m}]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \mathbf{m})
+h_{vel}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
 \end{equation}

### 2017-01-23 body/equilibrium.tex
complete paper
no symbol redefinition

@@
 % !TEX root =  ../main.tex
-\section{Heuristics and criteria for contact selection}
-\label{sec:heuristics}
+
 \subsection{A criterion for robust static equilibrium}
@@
 	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
-	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local Cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
-	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
+	\item $\mathbf{n}_i,\mathbf{\boldsymbol\upgamma}_{i1},\mathbf{\boldsymbol\upgamma}_{i2}$ form a local Cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
+	with the contact surface normal, and the $\mathbf{\boldsymbol\upgamma}_i$s are tangent vectors.
 	\end{itemize}
@@
 \begin{equation*}
-\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{t}_{i1} & \mathbf{n}_{i} -\mu \mathbf{t}_{i1} & \mathbf{n}_{i} + \mu \mathbf{t}_{i2} & \mathbf{n}_{i} - \mu \mathbf{t}_{i2}}^T
+\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{\boldsymbol\upgamma}_{i1} & \mathbf{n}_{i} -\mu \mathbf{\boldsymbol\upgamma}_{i1} & \mathbf{n}_{i} + \mu \mathbf{\boldsymbol\upgamma}_{i2} & \mathbf{n}_{i} - \mu \mathbf{\boldsymbol\upgamma}_{i2}}^T
 \end{equation*}

### 2017-01-23 cover2.tex
finished cover letter

@@
 
-First of all we thank you for the time spent on considering our original manuscript. Indeed,
+First of all we thank you for the time spent considering our original manuscript. Indeed,
 %This cover letter for the manuscript ``An efficient acyclic contact planner for legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity.
@@
 
-\section{Recall of main paper contributions}
+Furthermore, we included the integrality of the previous reviews, decomposed in an array. Each row contains a paragraph of the review, as well as our answer to the
+specific paragraph.
+
+\section{Rebutal}
+\subsection{Main contributions of the paper}
+More precisely, we propose a very efficient method to address what we believe is the core of the difficulty of the contact planning problem: discovering the sequence of key contact postures that would lead to a safe and stable contact trajectory of the robot.
 The paper is concerned with the problem of planning acyclic contact movements. 
-More precisely, we propose a very efficient method to address what we believe is the core of the difficulty of the contact planning problem: discovering the sequence of key contact postures that would lead to a safe and stable contact trajectory of the robot.
 Using the terminology of the paper, our contact planner addressed the two sub-problems $\mathcal{P}_1$ (finding a guide path for the robot root) and $\mathcal{P}_2$ (finding a sequence of stable contact postures and corresponding contact locations).
-Our claim is that the solution of our planner leads to a straight-forward resolution of the last sub-problem $\mathcal{P}_3$ (finding the complete trajectory connecting the contact postures).
-In particular, our method proposes interactive performances, i.e. the planner needs less time to compute the solution than what is needed for the robot to execute it.
+Our claim is that the solution produced by our planner leads to a feasible sub-problem $\mathcal{P}_3$ (finding the complete trajectory connecting the contact postures).
+We demonstrate that our method provides interactive performances, i.e. the planner needs less time to compute the solution than what is needed for the robot to execute it.
 This practical result corresponds to a significant improvement of the state of the art, for example regarding the capabilities of acyclic contact planners demonstrated by the humanoid teams during the Darpa Robotics Challenge last year.
 
-Regarding this contribution, the reviews mostly criticized the structure of the paper (too verbose) and the unjustified claim that the contact sequence output by the planner leads to a feasible resolution of $\mathcal{P}_3$.
+Regarding this contribution, the reviews essentially criticized the structure of the paper (too verbose) and the undemonstrated claim that the contact sequence output by the planner leads to a feasible resolution of $\mathcal{P}_3$.
 
-In the revision, we implemented a solver inspired by the state of the art to solve $\mathcal{P}_3$ and empirically proved the claim that our $\mathcal{P}_1$-$\mathcal{P}_2$ planner leads to easily solving $\mathcal{P}_3$.
-All the robot contact sequences reported in the paper have been converted to full contact trajectories using this $\mathcal{P}_3$ solution, that were validated in an contact simulator.
+In the revision, we implemented a solver inspired by the state of the art to solve $\mathcal{P}_3$ and empirically proved the claim that our $\mathcal{P}_1$-$\mathcal{P}_2$ planner leads to a feasible $\mathcal{P}_3$.
+All the robot contact sequences reported in the paper have been converted to full contact trajectories using this $\mathcal{P}_3$ solution, that were validated in an inverse dynamics simulator.
 We also executed one of them on the real robot.
@@
 
-\section{Validation of the method (addressing  $\mathcal{P}_3$) }
+\subsection{Validation of the method (addressing  $\mathcal{P}_3$) }
 In his review, Reviewer 2 stated that our plans are likely to be not executable, even in simulation.
@@
 While it was mentioned in the paper, we did not insist on these demonstrations because our main contribution focuses on the other problems. 
-We believe that, as Reviewer 1 notices, this observation was due to the fact that we did not clearly define what we mean by a contact plan (i.e.
+We believe that, as Reviewer 2 notices, his observation is due to the fact that we did not clearly define what we mean by a contact plan (i.e.
 a discrete sequence of key contact posture with the corresponding contact locations).
-We thank the reviewers for these comments, since it pointed out that what we believed to be an evidence must be carefully explained and justified.
-
+We thank the reviewers for these comments, which pointed out that we had to carefully explain and justify this aspect of the paper.
 A rigorous definition of a contact plan is now detailed in section 5.1.
@@
 
-\section{Success rates of the method}
+\subsection{Success rates of the method}
 At this point, we believe it is important to clarify our position regarding the concerns of both reviewers about the success rates of our paper.
@@
 We accept this fact and argue that it is mostly successful (i.e. more than 77\% of success in reasonable ``DRC-like'' scenarios).
-Even in case of failure, the planner is based on random sampling that approximate probabilistic completeness, and can be restarted to obtain another candidate trajectory.
+Even in case of failure, since the planner is based on random sampling it can be restarted to obtain another candidate trajectory.
 In practice, it can always find a feasible solution fast enough for online applications by re-planning upon failure.
@@
 
-\section{Originality and interest of the work}
-We strongly disagree with Reviewer 2 regarding the originality of the work when he states that we simply ``\textit{experiment with minor heuristic adjustments}''.
+\subsection{Originality and interest of the work}
+\label{orig}
+We strongly disagree with Reviewer 1 and 2 regarding the originality of the work as ``\textit{heuristic adjustments}''.
 Bretl formalized well the problem without providing an algorithm applicable to the general case. Escande focused on the contact planning $\mathcal{P}_2$. 
@@
 
-\section{A clearer paper}
-As requested, we have deeply shortened the paper. More importantly, we have done an important work of defining in simpler terms
+\subsection{A clearer paper}
+As requested by both reviewers, we have deeply shortened the paper. More importantly, we have done an important work of defining in simpler terms
 the different concepts that we use in the paper, and reduced their number.
@@
 
-\section{Justification of the simplifying assumptions}
+More specifically, Reviewer 1 stated that the section related to the contact generation, in particular, was confusing. We rewrote Section 2.2 and Section 5 to give a clearer
+description of the method, especially regarding the addition of intermediate contact configurations.
+
+\subsection{Justification of the simplifying assumptions}
 Regarding several comments from both reviewers, stating that our paper does not justify the fact that contact feasibility implies equilibrium feasibility,
@@
 
-\section{Comparison with other methods}
+\subsection{Comparison with other methods}
 As noted by Reviewer 2, in our previous paper it was indeed complex to compare our work with previous contributions,
@@
 
-\section{On the static equilibrium of each configuration}
-Reviewer 1 stated that ``\textit{the planned contact path is only discrete with static equilibrium considered for each
-configuration, which is not very realistic''}.
-While we agree with the observation, we first recall that this assumption is made in the previous works that address cluttered scenarios, such as the work from Bretl and Escande.
-Works that do not make this assumption such as Mordatch and al.
-only offer local convergence and cannot address these scenarios (e.g.
-on the ``Polaris car egress'' scenario reported in our paper).
-Furthermore, this limitation is in fact less restrictive in our approach that the one of Escande, since while our plan is quasi static, our resulting motion is dynamic, as now explained in the paper (Section 5.1).
-
-\section{Actual car model of the DARPA Challenge}
+\subsection{Actual car model of the DARPA Challenge}
 As asked by Reviewer 2, we now use the Polaris car model for demonstrating our egress scenario.
@@
 We modified the paper to more rigorously discuss these hypotheses. 
-We also believe that the main justification of their validity is through statistical tests, as reported in the result section.\newline
+We also believe that the main justification of their validity is shown by the statistical tests we provide in the result section.\newline
 See Section 6 for details. 
@@
 Because all that was done off-line, it is not clear how N was decided. 
+&
+As explained in the paper, in our previous work (Tonneau et al.
+2014), we discussed in details  how to choose N.
+We chose to refer to the paper rather than giving details  because of the space constraints of
+the paper. In our experiments,
+we always set $N=10000$.
+
+\\ \hline % ----
 The on-line planning step would search through the stored limb configurations (sorted according to some heuristic), one limb at a time (it seems), to find an almost-contact configuration and create a feasible limb contact configuration out of it.
@@
 & 
---- I THINK WE MUST DISCUSS THAT HERE ---
+The main reason to add intermediate contact configuration
+is to respect the conditions imposed by our planner: at most one contact must be broken between two configurations. Sometimes the kinematic constraints are such
+that for a given discretization step, several contacts have to be broken at once. Adding intermediate states allows to reposition one contact before the other.
+Concretely, when an intermediate contact is added, one contact is broken and relocated.
+This is the role of \textsc{RepositionContacts}, as explained in Appendix D.
+We clarified this aspect in section 5.1.
+
 \\ \hline % ----------------------------------------------------------------------------------
@@
 &
---- I THINK WE MUST DISCUSS THAT HERE ---
+For locomotion tasks, we are interested in avoiding the singular configurations that are not desirable.
+Indeed, in our approach we discard these singular configurations, at the sampling phase, or by favoring non singular configurations
+using our manipulability based heuristics.
+
+However, this choice is not a limitation of our method: if singular configurations were required for some reason, the method would be able
+to handle them.
+
 \\ \hline % ----------------------------------------------------------------------------------
@@
 &
-We agree that our approach is mostly heuristic.
-Heuristics are somehow underestimated in robotics, where optimality principles are often preferred (this is not the case in other fields, like discrete optimization, data routing or multi-agent diagnosis in web applications).
+As detailed in Section~\ref{orig} of this cover letter, we disagree regarding the fact that our approach is purely heuristic.
+We claim that the dimensionality reduction proposed to address \mP1 is an important contribution to the field.
 
 The main hypotheses of the papers are: the division of contact planning in three problems; the reachability condition approximating the equilibrium condition; and the heuristics for building the contact sequence (\mP2) along the guide path (\mP1).
-We believe that these main hypotheses are properly discussed, some being theoretically justified (reachability condition), the other being statistically asserted.
+We believe that these main hypotheses are properly discussed, some being theoretically justified (reachability condition), the others being statistically asserted.
 
-We agree that the method also requires some other minor choices, like the ones discussed by the reviewers. 
+%~ Then, regarding the heuristics we used to generate contacts and solve \mP2, we believe that heuristics are somehow underestimated in robotics, where optimality principles are often preferred.
+This is not the case in other fields, like discrete optimization, data routing or multi-agent diagnosis in web applications.
+We agree that the method also requires some other minor choices, especially regarding \mP2, like the ones discussed by the reviewers. 
 We do not believe that these minor technical choices are important. 
@@
 To keep the paper as short as possible, we did not spent time explaining these choices.
-Any reader that would try to re-implement our approach would find enough details to reproduce our choices.
-The source code of the planner is also available (url in the paper) in case the exact implementation is needed.
-
-Providing more details about any single technical choice would require 10 more pages, which is in contraction to the request of the reviewer to have the paper less verbose.
+%~ Any reader that would try to re-implement our approach would find enough details to reproduce our choices.
+%~ The source code of the planner is also available (url in the paper) in case the exact implementation is needed.
+%~ 
+%~ Providing more details about any single technical choice would require 10 more pages, which is in contraction to the request of the reviewer to have the paper less verbose.
 
@@
 &
-Our contribution is on solving \mP1-\mP2 efficiently. 
-We believe that this is the core of the contact planning problem, as argued in introduction.
-Indeed, solving \mP3 in all the cases studied in the paper was quite straight-forward.
-
-Clearly, dividing the complete problem in 3 subproblems will ease the solver and possibly obstruct some solutions. The question is whether this choice is a sane trade-off or not. 
-We insist that this division is quite classical in the state of the art.
-In particular, the only planner able to cope with this division (by Mordatch et al) only leads to local convergence.
+We first recall that this assumption is made in the previous works that address cluttered scenarios, such as the work from Bretl and Escande.
+Works that do not make this assumption such as Mordatch and al.
+only offer local convergence and cannot address these scenarios (e.g.
+on the ``Polaris car egress'' scenario reported in our paper).
+Furthermore, this limitation is in fact less restrictive in our approach that the one of Escande, since while our plan is quasi static, our resulting motion is dynamic, as now explained in the paper (Section 5.1).
 
 In the new version, we asserted the realism of the contact sequences by implementing a solution to \mP3, outputing full trajectories that were tested in a contact simulator and on the real HRP-2 robot.
-
 See Sections 2 and 6 for details.
@@
 The important problem of transitioning motion from one contact configuration to the next was left out as beyond the scope of the paper. 
-&
---- I THINK WE MUST DISCUSS THAT HERE ---
-\\ \hline % ----------------------------------------------------------------------------------
 The planner was tested in a number of diverse simulation examples, and there were related comparisons with existing work.
@@
 
-In the original paper, we claimed that any of these contact sequences would directly lead to a full contact trajectory (\mP3).
-However, we did not asserted this claim. 
+In the original paper, we claimed that our contact sequences would directly lead feasible problem (\mP3), and thus a feasible trajectory.
+However, we did not assert this claim. 
 In the new version, we implemented a solver \mP3 and validated that all the contact sequences reported in the paper are leading to full contact trajectories, whose dynamic plausibility has been checked by execution on the real robot for one of them, and in a dynamic contact simulator for the others.
@@
 
-In the revision, we have implement a solve \mP3 and converted all contact sequences in full contact trajectory, validated in simulation or on the real robot.
-We must insist that this validation was straight-forward, on a scientific point of view. 
-As initially stated, the produced contact sequences indeed lead directly to full trajectories.
+In the revision, we have implement a solver for \mP3 and converted all contact sequences to whole-body trajectories, validated in simulation or on the real robot.
+%~ We must insist that this validation was straight-forward, on a scientific point of view. 
+As initially stated, the produced contact sequences indeed lead directly to complete trajectories.
 
@@
 In terms of genericity, two key aspects must be considered.
-First, our planner is good enough for 90\% of the situations that nowadays legged robot can meet.
+First, our planner is good enough for the majority of situations that nowadays legged robot can meet.
 Second, the general idea of reducing the contact planning problem is valid in any situations that we can imagine, and the heuristics and approximate models that we proposed can be adapted to extend the planner capabilities.
@@
 &
-It was indeed difficult to fully feel the enthusiasm of the reviewer, although we definitely appreciated his detailed comment and effort to improve the quality of the paper.
-In conclusion, we truly believe that our solution answer to the main difficulty of contact planning, by building a realistic contact sequence which easily leads to the construction of a complete contact trajectory, and behind, the execution of it on the real robot.
+%~ It was indeed difficult to fully feel the enthusiasm of the reviewer, although
+We definitely appreciated the detailed comments and effort to improve the quality of the paper.
+In conclusion, we truly believe that our solution answers to the main difficulty of contact planning, by building a realistic contact sequence which easily leads to the construction of a complete contact trajectory, and behind, the execution of it on the real robot.
 The addition of a \mP3 solver in the revised paper experimentally justifies this claim.
@@
 The contribution that we are reporting is not a marginal increment or a practical implementation, but a step-change in how the contact planning problem should be considered, leading to the first realistic implementation.
-We hope that the revision of the paper will help the reviewer to meet our opinion.
+We hope that the revision of the paper will help the reviewers to meet our opinion.
 

### 2017-01-23 cover2.tex
[COVER] Minor final remark.

@@
 
+In this section, we answer to each point of the reviews, while referring to the first section of the letter.
+
 \subsection{Reviewer 1}
@@
 \begin{longtable}{|p{21em}|p{21em}|}
+\hline \bf Reviewer comment & \bf Author answer \\
 \hline
@@
 &
-As detailed in Section~\ref{orig} of this cover letter, we disagree regarding the fact that our approach is purely heuristic.
-We claim that the dimensionality reduction proposed to address \mP1 is an important contribution to the field.
+%As detailed in Section~\ref{orig} of this cover letter, we disagree regarding the fact that our approach is purely heuristic.
+%We claim that the dimensionality reduction proposed to address \mP1 is an important contribution to the field.
+The main contribution of the paper are the dimensionality reduction breaking the complexity of the contact planning problem.
+The main reduction comes from the reachability condition used to solve subproblem \mP1, which is not a heuristic but rather an approximate condition.
+Some rigorous heuristics are introduced to solve \mP2.
+We therefore disagree regarding the fact that our approach is purely heuristic.
 

### 2017-04-19 cover2.tex
commit before switchin to tro

--- 2017-01-23 cover2.tex
@@
-\documentclass[a4paper]{article}
- \usepackage{fullpage}
-\usepackage{amsfonts} % pour les lettres maths creuses \mathbb
-\usepackage{amsmath}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[frenchb]{babel}
-\usepackage{aeguill}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{color}
-\usepackage{listings}
-\usepackage{pdfpages}
-\usepackage{longtable}
+\title{Cover letter}
 
-\newcommand\dx{\dot{x}}
-\newcommand\dy{\dot{y}}
-\newcommand\ddx{\ddot{x}}
-\newcommand\ddy{\ddot{y}}
-\def\real{{\mathbb R}}
-\frenchbsetup%
-{%
-StandardItemLabels=true,%
-ItemLabels=\ding{43},%
-}%
+\documentclass[12pt]{article}
 
-\author {}
-\title {Cover letter for the IJRR submission \\ ``An efficient acyclic contact planner for multiped robots''}
-\date {}
 \begin{document}
@@
 
-Dear Editor, \medskip
+This cover letter for the manuscript ``An efficient acyclic contact planner for
+legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity. We left the original cover
+letter at the end of this document.
 
-First of all we thank you for the time spent on considering our original manuscript. Indeed,
-%This cover letter for the manuscript ``An efficient acyclic contact planner for legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity.
-this cover letter introduces the first revision of the manuscript ``An efficient acyclic contact planner for legged robots'', resubmitted for the ISSR'15 special issue.
-As discussed with Prof. Hollerbach, we hope to have reviewer continuity for the revision.
-The cover letter then plays more the role of a rebuttal. We left the original cover letter at the end of this document, as well as the original ISRR reviews.
-Therefore, we kindly ask that this cover letter be forwarded to the reviewers of the new manuscript.
-
-We would then like to thank the reviewers for their previous feedback on the initial manuscript. We believe we have answered
-all of their concerns, the main ones being that our method had not been tested on a real robot or a simulation, and that the paper was too verbose.
+We would first like to thank the reviewers for their previous feedback on the initial manuscript. We believe we have answered
+all of their concerns, the main ones being that our method had not been tested on a real robot nor in simulation, and that the paper was too verbose.
 In the following lines we are going to describe how we improved our paper in this regard, and answer to the other remarks of the reviewers.
 
-\section{Recall of main paper contributions}
-The paper is concerned with the problem of planning acyclic contact movements. 
-More precisely, we propose a very efficient method to address what we believe is the core of the difficulty of the contact planning problem: discovering the sequence of key contact postures that would lead to a safe and stable contact trajectory of the robot.
-Using the terminology of the paper, our contact planner addressed the two sub-problems $\mathcal{P}_1$ (finding a guide path for the robot root) and $\mathcal{P}_2$ (finding a sequence of stable contact postures and corresponding contact locations).
-Our claim is that the solution of our planner leads to a straight-forward resolution of the last sub-problem $\mathcal{P}_3$ (finding the complete trajectory connecting the contact postures).
-In particular, our method proposes interactive performances, i.e. the planner needs less time to compute the solution than what is needed for the robot to execute it.
-This practical result corresponds to a significant improvement of the state of the art, for example regarding the capabilities of acyclic contact planners demonstrated by the humanoid teams during the Darpa Robotics Challenge last year.
+\section{Validation of the method (addressing  $\mathcal{P}_3$) }
+Because we had already validated some of our plans, we were really surprised by the remark of reviewer 2 stating that our plans are not executable, even in simulation.
+The stair climbing scenario has been demonstrated
+on the real HRP-2 robot, while the standing up had been demonstrated in simulation in (Carpentier et al., ICRA 2015). While it was mentioned in the paper,
+we did not insist on these demonstrations because our main contribution focuses on the other problems $\mathcal{P}_1$ and $\mathcal{P}_2$.
 
-Regarding this contribution, the reviews mostly criticized the structure of the paper (too verbose) and the unjustified claim that the contact sequence output by the planner leads to a feasible resolution of $\mathcal{P}_3$.
+We believe that, as reviewer 1 notices, this observation was due to the fact that we do not clearly define what we mean by a contact plan.
+This is now detailed in section 5.1.. In any case, the validation videos joined to the submission should now establish that our plans are in indeed feasible, even without considering
+the torque constraints.
 
-In the revision, we implemented a solver inspired by the state of the art to solve $\mathcal{P}_3$ and empirically proved the claim that our $\mathcal{P}_1$-$\mathcal{P}_2$ planner leads to easily solving $\mathcal{P}_3$.
-All the robot contact sequences reported in the paper have been converted to full contact trajectories using this $\mathcal{P}_3$ solution, that were validated in an contact simulator.
-We also executed one of them on the real robot.
-This extension, along with the correction of the other remarks by the reviewers, have been integrated in the revision paper, that we also reworked to make it clearer, shorter and sharper (the core of the paper is now 13 pages long, as opposed to 17 before) than the original manuscript.
-
-The details of these modifications follow.
-
-\section{Validation of the method (addressing  $\mathcal{P}_3$) }
-In his review, Reviewer 2 stated that our plans are likely to be not executable, even in simulation.
-
-\begin{quote}
-  \textit{A quick glance at the sequences displayed in the video reveals that the "motion patterns" (and I understand these are not "motions" per se, but "contact plans") are not physically consistent (even reagrding them as mere contact plans and not as the final motions).
-The robot moves a contact without needing to readjust the posture in order to release that contact (or "set it free", by nullifying the contact forces on that contact).
-This results in sequences in which the contacts seem to be "slided" rather than removed and repositioned.
-This denotes a significant flaw in the algorithm.
-}
-\end{quote}
-
-We were really surprised by the remark, in particular because we had already validated some of our plans.
-The stair climbing scenario has been demonstrated on the real HRP-2 robot, while the standing up had been demonstrated in simulation. 
-While it was mentioned in the paper, we did not insist on these demonstrations because our main contribution focuses on the other problems. 
-We believe that, as Reviewer 1 notices, this observation was due to the fact that we did not clearly define what we mean by a contact plan (i.e.
-a discrete sequence of key contact posture with the corresponding contact locations).
-We thank the reviewers for these comments, since it pointed out that what we believed to be an evidence must be carefully explained and justified.
-
-A rigorous definition of a contact plan is now detailed in section 5.1.
-
-%In any case, the validation videos joined to the submission should now establish that our plans are in indeed feasible, even without considering the torque constraints.
-
-We also implemented a solver inspired by the state of the art to dynamically interpolate the contact plan and produce a complete trajectory of the robot, that we validated in a full-dynamics simulator.
-This implementation completely validates that the contact plan is feasible by the real robot.
-We also indeed executed one of the computed trajectory (climbing the stairs with HRP-2).
-
-We have joined the full-dynamics simulation and real life videos of all the scenarios we demonstrate.
+In this submission, we have thus joined the simulation and real life videos of all the scenarios we demonstrate.
 These videos show that the plans we compute correspond to feasible motions.
-While we briefly describe how we solve $\mathcal{P}_3$, the focus of the paper remains $\mathcal{P}_1$ and $\mathcal{P}_2$. 
-The solution of $\mathcal{P}_3$ is reported in the appendix to emphasize the main focus.
-
-\newcommand{\mP}[1]{$\mathcal{P}_#1$}
+We thank the reviewers for this comment,
+since we now believe that our paper is stronger with the complete demonstration. However, while we briefly describe how we solve $\mathcal{P}_3$, the focus of the 
+paper remains $\mathcal{P}_1$ and $\mathcal{P}_2$.
 
 \section{Success rates of the method}
-At this point, we believe it is important to clarify our position regarding the concerns of both reviewers about the success rates of our paper.
-In the paper, we are providing the success rates of our contact planner. 
-Such an empirical study has yet never been provided for other contact planners.
-Our method is not always successful, as very often with probabilistic searches.
-More precisely, the hypothesis taken by \mP1 may lead to an unfeasible \mP2 problem, and similarly \mP2 hypotheses may lead to an unfeasible \mP3.
-We accept this fact and argue that it is mostly successful (i.e. more than 77\% of success in reasonable ``DRC-like'' scenarios).
-Even in case of failure, the planner is based on random sampling that approximate probabilistic completeness, and can be restarted to obtain another candidate trajectory.
-In practice, it can always find a feasible solution fast enough for online applications by re-planning upon failure.
+At this point, we believe it is important to clarify our position regarding both reviewers concerns of the success rates of our paper.
+We did something that no other contact planning paper does, that is providing the success rates of our contact planner.
+Our method is not always successful, we assume it and argue that it is mostly successful, and that it can always find a feasible solution fast enough for online applications by replanning upon failure.
 
-Reading the papers reporting previous works in contact planning, it is in fact not discussed whether the methods are always successful or not.
-We simply have no idea.
-For instance, the CIO of Mordatch,
+In previous works, it is in fact unknown whether the methods are always successful or not. We simply have no idea. For instance, the CIO of Mordatch,
 because of the local convergence of the method, \textbf{necessarily} fails in some scenarios. We just do not know how often. Similarly, Escande et al. had to manually edit 
 the end effector trajectories to avoid collisions with the environment, but we do not know how often. Considering this,
-we do not understand why we are criticized for providing our success rates, which actually demonstrate the interest of our approach.
+we do not understand why we are criticized for providing our success rates, which actually demonstrate the interet of our approach.
 
 \section{Originality and interest of the work}
-We strongly disagree with Reviewer 2 regarding the originality of the work when he states that we simply ``\textit{experiment with minor heuristic adjustments}''.
+We strongly disagree with reviewer 1 regarding the originality of the work. Reviewer 1 states that we simply experiment with minor heuristic adjustments.
 Bretl formalized well the problem without providing an algorithm applicable to the general case. Escande focused on the contact planning $\mathcal{P}_2$. 
-Our work tackles $\mathcal{P}_1$ as well, in a much more efficient manner than Bouyarmane et al.
-These works, including the one of Hauser, are limited by the computational complexity of considering the full dimension of the problem.
-In practice, the underlying complexity imply that they cannot lead to implementation that can be used in practical scenarios, as emphasized by the lack of such contact solvers during the DRC (as discussed by Atkinson).
+Our work tackles $\mathcal{P}_1$ as well, in a much more efficient manner than Bouyarmane et al.. 
+As for Hauser, these works are limited by the computational complexity of considering the full dimension of the problem.
 
-We claim that the reachability condition provides a dimensionality reduction of the problem. This is clearly an originality regarding the previous work who always consider the whole-body problem.
-We believe that this reduction is the only way to make the contact planning problem tractable, in the same way that many researchers (such as ourselves, but also Wieber, Tedrake, Righetti and others) use dimensionality reduction successfully to address the trajectory optimization problem $\mathcal{P}_3$.
+The reachability condition provides a dimensionality reduction of the problem. This is clearly an originality regarding the previous work who always consider the whole-body problem.
+We believe that this reduction is the only way to make the contact planning problem tractable, in the same way that many researchers (such as ourselves, but also Wieber, Tedrake, Righetti and others) use dimensionality reduction successfully to address
+the trajectory optimization problem $\mathcal{P}_3$.
 
@@
 the different concepts that we use in the paper, and reduced their number.
-We believe that the paper is much clearer in its current form.
+We believe that the paper is much more clearer in its current form.
 
@@
 \section{Comparison with other methods}
-As noted by Reviewer 2, in our previous paper it was indeed complex to compare our work with previous contributions,
-since they were addressing the complete problem. We now give the complete computation times and compare them with previous works,
+As noted by reviewer 2, in our previous paper it was indeed complex to compare our work with previous contributions,
+since they were adressing the complete problem. We now give the complete computation times and compare them with previous works,
 clearly demonstrating that our method outperforms them by orders of magnitude (Section 6.4).
@@
 \section{On the static equilibrium of each configuration}
-Reviewer 1 stated that ``\textit{the planned contact path is only discrete with static equilibrium considered for each
-configuration, which is not very realistic''}.
-While we agree with the observation, we first recall that this assumption is made in the previous works that address cluttered scenarios, such as the work from Bretl and Escande.
-Works that do not make this assumption such as Mordatch and al.
-only offer local convergence and cannot address these scenarios (e.g.
-on the ``Polaris car egress'' scenario reported in our paper).
+Reviewer 1 stated that ``the planned contact path is only discrete with static equilibrium considered for each
+configuration, which is not very realistic''. First, we recall that this assumption is always made in the previous works that address cluttered scenarios,
+such as the work from Bretl and Escande.
+Work that do not make this assumption such as Mordatch and al.
+only offer local convergence and can't address these scenarios.
 Furthermore, this limitation is in fact less restrictive in our approach that the one of Escande, since while our plan is quasi static, our resulting motion is dynamic, as now explained in the paper (Section 5.1).
@@
 \section{Actual car model of the DARPA Challenge}
-As asked by Reviewer 2, we now use the Polaris car model for demonstrating our egress scenario.
-
-\newpage
-\section{Recap of Reviewer comments}
-
-\subsection{Reviewer 1}
-\noindent
-\begin{longtable}{|p{21em}|p{21em}|}
-\hline
-The paper addresses an important problem of planning robot motion involving contacts.
-It extends the idea of planning acyclic contacts by decoupling the issue into (1) planning a guide path for the root of the robot, and (2) planning a discrete sequence of acyclic and static equilibrium contact configurations along the path, with an interpolation of motion between two consecutive contact postures in the sequence.
-The introduced planner considers both collision avoidance and maintenance of desired contacts and aims at real-time or interactive performance.
-&Thanks for the comment.
-\\ \hline % ----------------------------------------------------------------------------------
-For the decoupled approach to work, the planned guide path must guarantee that a sequence of stable contact configurations is associated with it to execute the path with desired contacts and yet without undesired collisions; such a guide path is called “equilibrium feasible”.
-The issue is how to find such a guide path efficiently.
-One key assumption of the paper is that for cluttered environments, “most of the times” a guide path that is “contact feasible” (referring to that at least one end-effector is in contact with the environment and there is no undesired collision between the robot and the environment) is also “equilibrium feasible”.
-It further assumes that if the reachable space of the robot along the guide path is in collision with the environment but the robot root configuration is not in collision, called the “reachability condition”, then the guide path is “contact feasible”.
-With those assumptions, the paper treated guide path planning as a geometric problem and used an RRT planner to solve the problem.
-& We agree with this summary.
-\\ \hline % ----------------------------------------------------------------------------------
-However, the question is whether the above key assumption (of 
-“contact feasible” implies “equilibrium feasible”) is reasonable.
-The paper did not provide justification for the assumption, and the assumption does not seem intuitive.
-The paper used terms such as “most of the times” and “cluttered contact planning” without clear and quantifiable definitions.
-With the above assumptions, planning becomes quite straightforward, and the planning algorithm used is rather conventional.
-&
-We modified the paper to more rigorously discuss these hypotheses. 
-We also believe that the main justification of their validity is through statistical tests, as reported in the result section.\newline
-See Section 6 for details. 
-\\ \hline % ----------------------------------------------------------------------------------
-For the next problem of planning a sequence of contact configurations given a “contact feasible” root path, the paper further decomposed the problem by planning the configuration of one limb at a time, assuming the root configuration and the other limb configurations were fixed.
-For each limb, N sampled configurations were considered (sampled off-line), and they were indexed by the corresponding end-effector position and stored in an octree beforehand.
-Because all that was done off-line, it is not clear how N was decided. 
-The on-line planning step would search through the stored limb configurations (sorted according to some heuristic), one limb at a time (it seems), to find an almost-contact configuration and create a feasible limb contact configuration out of it.
-Different limbs were considered in some deterministic order: the limb that was contact-free the longest was considered first, and contact configurations were generated as long as the limb could maintain contact.
-It is a bit confusing when interpolated contact configurations would be added between two adjacent contact configurations.
-Was it “in the hope” to keep a limb in contact or change limbs in contact?
-& 
---- I THINK WE MUST DISCUSS THAT HERE ---
-\\ \hline % ----------------------------------------------------------------------------------
-The manipulator Jacobian for each limb was assumed full rank, and no singular configurations were considered. 
-&
---- I THINK WE MUST DISCUSS THAT HERE ---
-\\ \hline % ----------------------------------------------------------------------------------
-The main issue with the proposed planner is that assumptions (commented earlier) and decisions were made in the planner conveniently in order to simplify the problem but without much justification, and there is no guarantee of a solution.
-The entire approach is heuristic and somewhat ad hoc.
-&
-We agree that our approach is mostly heuristic.
-Heuristics are somehow underestimated in robotics, where optimality principles are often preferred (this is not the case in other fields, like discrete optimization, data routing or multi-agent diagnosis in web applications).
-
-The main hypotheses of the papers are: the division of contact planning in three problems; the reachability condition approximating the equilibrium condition; and the heuristics for building the contact sequence (\mP2) along the guide path (\mP1).
-We believe that these main hypotheses are properly discussed, some being theoretically justified (reachability condition), the other being statistically asserted.
-
-We agree that the method also requires some other minor choices, like the ones discussed by the reviewers. 
-We do not believe that these minor technical choices are important. 
-If other choices are made, similar results would likely be obtained as well.
-To keep the paper as short as possible, we did not spent time explaining these choices.
-Any reader that would try to re-implement our approach would find enough details to reproduce our choices.
-The source code of the planner is also available (url in the paper) in case the exact implementation is needed.
-
-Providing more details about any single technical choice would require 10 more pages, which is in contraction to the request of the reviewer to have the paper less verbose.
-
-In the revision, we emphasized what are the key assumptions, and spent more time to justify them properly.
-In the appositive, we more clearly pointed what assumptions are the results of technical choices and do not impact the solver performances.
-
-\\ \hline % ----------------------------------------------------------------------------------
-Another major limitation is that the planned contact path is only discrete with static equilibrium considered for each configuration, which is not very realistic.
-The paper also pointed out this limitation in the Conclusions.
-&
-Our contribution is on solving \mP1-\mP2 efficiently. 
-We believe that this is the core of the contact planning problem, as argued in introduction.
-Indeed, solving \mP3 in all the cases studied in the paper was quite straight-forward.
-
-Clearly, dividing the complete problem in 3 subproblems will ease the solver and possibly obstruct some solutions. The question is whether this choice is a sane trade-off or not. 
-We insist that this division is quite classical in the state of the art.
-In particular, the only planner able to cope with this division (by Mordatch et al) only leads to local convergence.
-
-In the new version, we asserted the realism of the contact sequences by implementing a solution to \mP3, outputing full trajectories that were tested in a contact simulator and on the real HRP-2 robot.
-
-See Sections 2 and 6 for details.
-
-\\ \hline % ----------------------------------------------------------------------------------
-The important problem of transitioning motion from one contact configuration to the next was left out as beyond the scope of the paper. 
-&
---- I THINK WE MUST DISCUSS THAT HERE ---
-\\ \hline % ----------------------------------------------------------------------------------
-The planner was tested in a number of diverse simulation examples, and there were related comparisons with existing work.
-This is commendable.
-However, there is no experimental result on a real robot.
-&
-The contribution of the paper is on subproblems \mP1 and \mP2. 
-We initially did not put robot results to emphasize this aspect and avoid confusing the reader.
-Following the reviewer recommendations, we added a solver of subproblem \mP3, leading to complete movement, one of them being executed on the real robot. 
-\newline See Section 2 for details.
-\\ \hline % ----------------------------------------------------------------------------------
-The writing of the paper can be much improved and made much more concise.
-The description of contact path planning in Section 4 is broken into different sub-sections, but not entirely in the order of how the sequence is generated, which is confusing.
-In Section 4.1, steps 3—6 talk about on-line generation, but the related decisions were introduced in Section 4.2.1 and 4.2.2, and none of those sub-sections has a title that clearly indicates the planning procedure.
-The paper has used unnecessary notations and symbols on the one hand, but vague languages such as “most of the time”, “cluttered problems”, “in the hope”, etc., on the other hand.
-Some symbols were not used consistently, for instance, m is used both to indicate mass and a direction.
-&
-Thanks for carefully listing these problems. 
-The paper was completely rewritten to make it shorter and more rigorous.
-\\ \hline % ----------------------------------------------------------------------------------
-\end{longtable}
-
-\newpage
-\subsection{Reviewer 2}
-\noindent
-\begin{longtable}{|p{21em}|p{21em}|}
-\hline
-The paper presents a method for producing contact plans for the locomotion of multiped robots.
-It decomposes the problem into two sub-problems : the computation of a guide path is which the robot is floating close enough to the environment, and subsequently the positioning of contacts on the environment along the guide.
-& Thanks for this summary, that we agree with.
-\\ \hline % ----------------------------------------------------------------------------------
-My major concern about this piece of work is its lack of originality.
-The problem seems to have been well formalized and studied in numerous previous works (Hauser et al, Escande et al, etc.).
-The authors experiment with minor heuristic adjustments.
-I do not consider them as major contributions that warrant a publication in IJRR by the standards of the journal.
-The authors end up with the main claim of the paper being a reduced computation time in comparison with these previous works.
-As admitted by the authors themselves, even this gain in computation time is not really comparable to the previous works since the outputs are different (previous works go all the way to the motion generation, and account for different constraints).
-&
-We do not agree with this comment.
-The main contribution of our method are the reduction hypothesis that we are using to break the complexity of contact planning.
-The problem reduction, leading to pragmatic implementation, is a step-change in how the problem of contact planning should be considered.
-
-Although we clearly acknowledge the contribution of previous works, in particular by Bretl, Hauser, Escande, Bouyarmane and their co-authors, they are all mostly nonpractical contributions.
-The  planners resulting of their works are interesting to study, but can barely be used in practice, in large part due to their computational complexity. 
-Even the work of Mordatch and colleagues, which is solving the problem only locally and partially (no collision and self-collision checking, partial dynamics, etc) requires minutes of computations for seconds of execution.
-
-We can clearly compare the performances of our planner with previous methods (even if not as accurately than what we would like): our method outperforms them by orders of magnitude.
-Even if considering our contribution only on the practical aspect of computation time,  it makes the difference between a method that can be used at the Darpa challenge or not, as we show in our introduction.
-
-\\ \hline % ----------------------------------------------------------------------------------
-The work is largely incomplete, and far from being self-sufficient.
-A quick glance at the sequences displayed in the video reveals that the "motion patterns" (and I understand these are not "motions" per se, but "contact plans") are not physically consistent (even regarding them as mere contact plans and not as the final motions).
-The robot moves a contact without needing to readjust the posture in order to release that contact (or "set it free", by nullifying the contact forces on that contact).
-This results in sequences in which the contacts seem to be "slided" rather than removed and repositioned.
-This denotes a significant flaw in the algorithm.
-At no point in the paper do the authors explicitly define what a "contact plan" is.
-Hence we do not know what is the output plan of their framework they are trying to obtain (obviously that output is not the same as that of previous works which feature the above-mentioned "releasing of contacts" property in their own definitions of "contact plans").
-The authors should clearly define what a valid output plan of their framework is.
-&
-Once more, we cannot agree with this comment.
-We believe that the comment is partly due to a problem of definition of what we mean by ``contact sequence''.
-The solution to subproblem \mP2 is a discrete sequence of static posture in stable contact (as detailed in Section 2).
-All the sequences reported in the paper are realistic in that sense.
-
-In the original paper, we claimed that any of these contact sequences would directly lead to a full contact trajectory (\mP3).
-However, we did not asserted this claim. 
-In the new version, we implemented a solver \mP3 and validated that all the contact sequences reported in the paper are leading to full contact trajectories, whose dynamic plausibility has been checked by execution on the real robot for one of them, and in a dynamic contact simulator for the others.
-
-See Sections 2 and 6 for details.
-
-\\ \hline % ----------------------------------------------------------------------------------
-There is no way to assess the validity of the work, as opposed to the previous works.
-From what I can judge from the video, the motions will not be able to be performed with a real robot nor even in physics simulation.
-The only way to legitimately argue with the latter statement is by providing a demonstration of the execution of the plans in simulation or with the real robots.
-I think non-trivial work still needs to be accomplished to "plug" the output of the presented framework into any existing motion generator or controller.
-If the only envisioned application of the framework is the coming future work of the authors to turn it into a valid motion controller, then I think the work would be complete only by then and its validity assessable from a scientific standpoint (i.e.
-not speculation) only at that stage.
-&
-We agree that this remark is justified by the lack of facts in our original paper.
-Our claim was that any contact sequence output by our planner easily lead to full contact trajectory.
-We wrongly assumed that this was an evidence when looking at the contact sequences, and neglected to prove it.
-
-In the revision, we have implement a solve \mP3 and converted all contact sequences in full contact trajectory, validated in simulation or on the real robot.
-We must insist that this validation was straight-forward, on a scientific point of view. 
-As initially stated, the produced contact sequences indeed lead directly to full trajectories.
-
-See Section 2 for details.
-
-\\ \hline % ----------------------------------------------------------------------------------
-There are many algorithm weakening assumptions.
-The most obvious one being the replacement of the equilibrium feasibility condition with the contact feasibility condition, then replacing the latter with the reachability condition.
-I understand that the authors assume that these approximations are valid, especially in "cluttered" environment.
-However I believe it should not be too difficult to think of counter-example problems in which each of this approximations creates a degenerate plan.
-&
-We agree with this remark. 
-The reduction models that we propose are valid approximations in a reasonable domain of the robot environment.
-Experimentally, we demonstrated that this validity domain is quite large.
-It is yet possible to find counter-examples, and we cite some of them in the paper.
-Our planner is only a contribution to the problem, not the ultimate solution to the contact-planning problem.
-
-In terms of genericity, two key aspects must be considered.
-First, our planner is good enough for 90\% of the situations that nowadays legged robot can meet.
-Second, the general idea of reducing the contact planning problem is valid in any situations that we can imagine, and the heuristics and approximate models that we proposed can be adapted to extend the planner capabilities.
-Automatic synthesis of these models is indeed a research direction that we are currently considering.
-
-In the revision, we now justified these assumptions by theoretical results, explanations about the rationale behind the assumption, and statistical validation reported in the experimental results.
-\\ \hline % ----------------------------------------------------------------------------------
-The joint torque limit constraint is absent from the static equilibrium condition.
-This is a major approximation that does indeed speed up computation with respect to this constraint being accounted for in some of the compared works, but results in weaker, less physically valid plans (again the only way to argue with that is to show that the motions are performed in real or in simulation)
-&
-We agree that the torque limits are not considered by our planner.
-However, although this is a hard limit on the robot, we do not agree that it plays a dramatic role in contact planning.
-Indeed, the torque limits are considered in our \mP3 validation, and does not negatively impact the execution of our contact sequences.
-We state and experimentally verified that considering the torque limits at the planning level is not necessary.
-\\ \hline % ----------------------------------------------------------------------------------
-"These two contributions were required to apply our method to real robot models", the use of "real robot" is somewhat misleading there.
-&
-By ``real robot'', we meant a realistic modeling of the robot, with true joint limits, collision constraint, etc. 
-During rewriting, this particular sentence was removed.
-We now used ``real robot'' only to discuss the experiments we did on the physical robot device.
-\\ \hline % ----------------------------------------------------------------------------------
-A minor remark : since the DARPA challenge is one of the motivations of the work then I do not understand why the authors did not use the actual model of the car used in the DARPA challenge and used the toy model instead ?
-A great part of the difficulty is removed there, numerous collisions occur with detail parts of the car (steering wheel, hand break, outside frame, etc) and contribute to the challenging aspect of the problems.
-&
-We used the simplified car model because the rendering seemed clearer to read.
-As requested by the reviewer, we implemented the same movement on the Polaris model taken from the DRC.
-Our HRP-2 robot being to small to go down the car (compared to DRC HRP-2 ``Kai'' robot), we added two steps.
-\\ \hline % ----------------------------------------------------------------------------------
-Finally, in the presentation department, I found the paper to be excessively long, I do not know what are the pages recommendations of the journal, but he paper is too verbose, the algorithms too partitioned into different sections and subsections.
-I think a major rewriting and restructuring is possible and the ideas can be conveyed in a more synthetic manner.
-&
-The paper was rewritten, see Section 5.
-\\ \hline % ----------------------------------------------------------------------------------
-Overall, I think the work is very promising, and I hope that my detailed comments and questions show my enthusiasm and interest of the subject.
-I understand that all the approximations and heuristics are driven by a sense of  pragmatism and practicality rather than pure theoretical soundness, which are probably legitimate motivations in their own right.
-However I also think that the work is still not mature enough, and will largely benefit from a validation of these approximations with a real robot or simulation experiment after the framework is complete.
-I encourage the authors to continue along the completion of the work.
-&
-It was indeed difficult to fully feel the enthusiasm of the reviewer, although we definitely appreciated his detailed comment and effort to improve the quality of the paper.
-In conclusion, we truly believe that our solution answer to the main difficulty of contact planning, by building a realistic contact sequence which easily leads to the construction of a complete contact trajectory, and behind, the execution of it on the real robot.
-The addition of a \mP3 solver in the revised paper experimentally justifies this claim.
-The results that we demonstrated have never been reached so far by any previous works, neither in term of qualitative results, number of reported scenario, statistical analysis of the algorithms or combinatorial efficiency.
-The contribution that we are reporting is not a marginal increment or a practical implementation, but a step-change in how the contact planning problem should be considered, leading to the first realistic implementation.
-We hope that the revision of the paper will help the reviewer to meet our opinion.
-
-\\ \hline % ----------------------------------------------------------------------------------
-\end{longtable}
+As asked by reviewer 2, we now use the polaris car model for demonstrating our egress scenario.
 
 \end{document}
-
-\author {}
-\title {Original Cover letter for the IJRR submission \\ ``An efficient acyclic contact planner for multiped robots''}
-\date {}
-\begin{document}
-\maketitle
-
-Dear editor and reviewers, \\ \\
-
-The current submission is an extension to our conference paper ``A reachability-based planner for sequences of acyclic contacts in
-cluttered environments'', accepted to ISRR 15 (\url{http://stevetonneau.fr/files/publications/isrr15/isrr15.pdf}).
-
-In these two papers, our objective is to propose automatic methods
-to compute the contact sequence that characterizes a robot motion in
-cluttered environment.
-
-We joined to the present submission the reviews that were made on the
-conference paper. They follow the cover letter.
-
-Regarding the conference paper, this appear paper presents new contributions that we first highlight:
-
-\begin{itemize}
-\item We introduce a new criterion for asserting the static equilibrium of a robot configuration. It works for arbitrary
-legged robots, and handles any kind of contact configurations and surface orientation. It is formulated as a really efficient Linear Program, and 
-can be used to compute a value of robustness of the configuration. This robustness is useful for real robots applications, where many uncertainties
-in the state estimation of the robot can lead to incorrect evaluations of the equilibrium.
-\item We present two new heuristics that we use to bias our contact planner framework towards configurations relevant for the task being performed.
-\item We present a complete algorithm for contact planning that integrates the robust equilibrium criterion.
-\end{itemize}
-
-The extension of our work also addresses the main comments of the reviewers, and proposes lesser contributions,
-that we detail in the following lines.
-
-\begin{itemize}
-\item They were concerned by the fact that our demonstrations were
-considering virtual avatars and not actual robots. In the present
-submission, all our examples are applied to the HRP-2 and HyQ robots (Section 6).
-To achieve this, a new contribution is proposed in our paper:
-We introduce a robustness criterion for asserting the static equilibrium
-of a configuration (Section 5.1). We use it in combination with new heuristics that we
-propose to bias the contact generation process towards configurations
-relevant for the motion being performed (Section 5.2).
-With the exception of the dexterous hand, all the demonstrations shown in
-the paper and the video are entirely new.
-
-\item They also judged that the conference paper provided a theoretical
-description of our approach, but lacked the details on how to apply
-the method on an actual robot. We now explain the complete process
-for the HRP-2 in the present submission (Appendix A).
-
-\item Concerning remarks made on the reproducibility of the method,
-we now provide the complete pseudo code of the algorithm, as well as the
-source code of our application (Appendix B).
-
-\item The paper has been largely re-written: we provide a clearer
-comparison to the state of the art (Section 1.1); we clarify our statement regarding
-our computation times and compare our method with the state of the art (Section 6); we make a clear distinction between the
-theoretical concepts we propose and their approximation (Section 2).
-
-\item Regarding the computation time performances, compared to our conference paper, the planner is much more efficient.
-A fundamental change is the re-implementation of the algorithm as a Bi-RRT planner instead of the PRM
-approach. The code is also more efficient thanks to its complete rewriting into the HPP platform.
-
-\item Finally, we propose an analysis of the performances of our method and
-the success rates we obtained for each scenario (Section 6.2). This analysis provides
-a support for our argumentation on the strong choices made in the design
-of our planner (Section 1.2 and 2), and a comparison with methods from the literature.
-\end{itemize}
-
-We are convinced that these points not only provide a complete answer to
-the reviews of the conference paper, but also bring new contributions to
-the community, that justify a journal publication.
-
-%~ On a last note, as required by the IJRR submission form,
-%~ we proposed a list of editors and reviewers for the submission.
-%~ We affirm that we don't have conflicts of interest with the people we propose.
-%~ For complete clarity however, we want to mention that some of the authors
-%~ are proposing two workshops for the ICRA conference this year.
-%~ Among the names we proposed for reviews, Karen Liu, kris Hauser, and Quang-Cuong Pham have accepted to be invited speakers.
-
-We thank the editor and the reviewers for their time.
-
-Sincerely, \\
-
-The authors
-
-\includepdf[pages=-]{reviewisrr.pdf}
-
-\end{document}

### 2017-04-19 cc.sh
updated compilation script

@@ -1,3 +1,4 @@
-pdflatex main.tex
+pdflatex main_tro.tex
 bibtex main.aux
-pdflatex main.tex
+pdflatex main_tro.tex
+mv main_tro.pdf root.pdf

### 2017-04-19 body/efort.tex
updated references

--- 2017-01-12 body/efort.tex
@@
 $\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
-\citeauthor{Yoshikawa1984} defines the velocity~\ref{eq:vel} and force~\ref{eq:vel} ellipsoids:
+Yoshikawa \citeauthor{Yoshikawa1984} defines the velocity~\ref{eq:vel} and force~\ref{eq:vel} ellipsoids:
 
@@
 In other less demanding cases, manipulability can also be considered to avoid singularities.
-To do so, we can consider the manipulability measure $h_{w}$, also given by \citeauthor{Yoshikawa1984}:
+To do so, we can consider the manipulability measure $h_{w}$, also given by Yoshikawa \citeauthor{Yoshikawa1984}:
 

### 2017-04-19 body/intro.tex
first compiling tro version
updated references

--- 2017-01-16 body/intro.tex
@@
 %~ Planning legged robot locomotion remains an open issue.
-Legged robots move by sequentially creating contacts with the environment.
+\IEEEPARstart{L}{egged} robots move by sequentially creating contacts with the environment.
 After years of research, such robots can autonomously walk on flat ground, but struggle to navigate more complex environments.
@@
 
-The issue of planning acyclic contacts was first completely described by \citeauthor{Bretl:2006:MPM:1124573.1124585}.
+The issue of planning acyclic contacts was first completely described by Bretl\citeauthor{Bretl:2006:MPM:1124573.1124585}.
 The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path.
@@
 A key issue is to avoid combinatorial explosion when considering at the same time the possible contacts and the potential paths.
-\citeauthor{Bretl:2006:MPM:1124573.1124585}'s seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments.
+Bretl's seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments.
 Following it, seve\-ral papers have applied this approach in specific situations, limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.
 
-Most of the papers that followed the work of \citeauthor{Bretl:2006:MPM:1124573.1124585} have explored alternative formulations to handle the combinatorial issue.
+Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorial issue.
 Two main directions have been explored.
@@
 A similar approach has been considered for manipulation by \cite{gabicciniisrr15}.
-\citeauthor{DBLP:conf/humanoids/DeitsT14} proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered.
-\citeauthor{dai2014whole} extended the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
+In \citeauthor{DBLP:conf/humanoids/DeitsT14}, Deits et Tedrake proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered.
+Dai et al. \citeauthor{dai2014whole} extended the work of Posa et al. \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
 the contacts manually for complex interactions.
@@
 \textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity.
-The feasibility and interest of the decoupling has been shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
-an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
+The feasibility and interest of the decoupling has been shown by Escande et al.
+who manually set up a rough root guide path (i.e.
+an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide \citeauthor{DBLP:conf/iser/EscandeKMG08}.
 A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning},
@@
 This approach is suboptimal because the quality of the motion depends on the quality of the guide path.
-\citeauthor{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
+Bouyarmane et al.
+precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium \citeauthor{Bouyarmane2009}.
 Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
 Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path.
-\citeauthor{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with
+Chung and Khatib also proposed a decoupled approach \citeauthor{7140082}, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with
 a discretized environment. This planning phase does not account for collisions, implying that replanning is required in case of failure. This approach is efficient 
@@
 However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
-In particular, our algorithm proposed in \citeauthor{Carpentier2016} is \gls{interactive}.
+In particular, our algorithm proposed in \cite{Carpentier2016} is \gls{interactive}.
 Therefore, a planner capable of efficiently solving \Pa and \Pb could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.

### 2017-04-19 body/results.tex
updated references

--- 2017-01-18 body/results.tex
@@
  \hline
-   \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}5.42 min}  \\
+   \multirow{3}{*}{Stair 20 cm} & Hauser et al. \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}5.42 min}  \\
 							 & \textbf{Ours} + \citeauthor{Carpentier2016}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & $ <$ 2s \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
+							 & Mordatch et al. \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
  \hline
-   \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}4.08 min}  \\
+   \multirow{3}{*}{Stair 30 cm} & Hauser et al. \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}4.08 min}  \\
 							 & \textbf{Ours}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & X \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
+							 &  Mordatch et al. \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
  \hline
-   \multirow{3}{*}{Stair 40 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}10.08 min}  \\
+   \multirow{3}{*}{Stair 40 cm} & Hauser et al. \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}10.08 min}  \\
 							 & \textbf{Ours}  & \textbf{5 s} & \textbf{3 s} & X \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
+							 &  Mordatch et al. \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
  %~ \hline

### 2017-04-19 cc.sh

@@ -1,4 +1,4 @@
 pdflatex main_tro.tex
-bibtex main.aux
+bibtex main_tro.aux
 pdflatex main_tro.tex
 mv main_tro.pdf root.pdf

### 2017-04-19 body/appendix_optim.tex

@@
 $\mathbf{q}_0$ to $\mathbf{q}_1$. We manually define the duration of each of the three contact phases.
-In each phase $s$ the centroidal wrench $\mathbf{w}$ is constrained to lie inside a cone $\mathcal{K}_s, s =0 \dots 2$.
+In each phase $s$ the centroidal wrench $\mathbf{w}$ is constrained to lie inside a cone $\mathcal{K}_u, u =0 \dots 2$.
 We call the total duration of the motion $T$, and formulate the following optimization problem:
@@
 %~ \find \quad & \mx{R} \in \Rv{3\times T}, \mxdd{C} \in  \Rv{6\times T} \\
-\minimize_{\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)}  \quad & \sum_{s=0}^{2} \int_{t_s}^{t_s + \Delta t_s}  h(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)) \text{dt}  \\
-\st \quad & \mathbf{\ddot{c}}(t) = f(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{w}(t)) \\
-	\quad & \mathbf{A}_s \mathbf{w}(t)  \leq \mathbf{b}_s , \forall t \in  \left[t_s, t_s + \Delta t_s \right[, \forall s \\
-	\quad & \mathbf{D}_s \mathbf{c}(t)  \leq \mathbf{d}_s , \forall t \in  \left[t_s, t_s + \Delta t_s \right[, \forall s \\
+\minimize_{\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)}  \quad & \sum_{u=0}^{2} \int_{t_u}^{t_u + \Delta t_u}  \ell(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)) \text{dt}  \\
+\st \quad & \mathbf{\ddot{c}}(t) = dyn(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{w}(t)) \\
+	\quad & \mathbf{A}_u \mathbf{w}(t)  \leq \mathbf{b}_u , \forall t \in  \left[t_u, t_u + \Delta t_u \right[, \forall u \\
+	\quad & \mathbf{Y}_u \mathbf{c}(t)  \leq \mathbf{y}_u , \forall t \in  \left[t_u, t_u + \Delta t_u \right[, \forall u \\
 	\quad & \mathbf{c}(0)  = \mathbf{c}_{\mathbf{q}_0} \\
@@
 
-The cost function $h$ is a weighted sum of the angular momentum and center-of-mass acceleration variation over the whole trajectory. 
-The system dynamics $f$ ties together the position, velocity and acceleration of the center of mass. 
-$\mathbf{D}_s$ and $\mathbf{d}_s$ denote stacked kinematic constraints on the center of mass position, determined by the active contact locations.
+The cost function $\ell$ is a weighted sum of the angular momentum and center-of-mass acceleration variation over the whole trajectory. 
+The system dynamics $dyn$ tie together the position, velocity and acceleration of the center of mass. 
+$\mathbf{Y}_u$ and $\mathbf{y}_u$ denote stacked kinematic constraints on the center of mass position, determined by the active contact locations.
 The inequalities for each contact are determined in the same way that the reachable workspace is computed in Appendix~\ref{app:rom}, with

### 2017-04-19 body/intro.tex

@@
 For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
-As a result at the DARPA Robotics Challenge, robots relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
+As a result at the DARPA Robotics Challenge, robots relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion.
+This lack was clearly identified by the challengers themselves: ``Except for egress, no robots in the DRC Finals used
+the  stair  railings  or  any  form  of  bracing''~\citep{atkensondarpa}.
 
@@
 Our work thus aims at breaking the complexity of the acyclic contact planning problem.
-To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorial of the contact selection problem.
+To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorics of the contact selection problem.
 First we introduce a low-dimensional space, called the \textit{contact-feasible} space, that can be sampled and mapped efficiently to the contact manifold.
@@
   \caption{
-    Overview of our two-stage framework. Given a path request between start and goal positions (left image), \Pa is the problem of computing a guide path in the space
+    Overview of our two-stage framework. Given a path query between start and goal positions (left image), \Pa is the problem of computing a guide path in the space
     of \textit{equilibrium feasible} root configurations.
@@
 
-Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorial issue.
+Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorics issue.
 Two main directions have been explored.
-\textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}.
+\textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorics of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}.
 A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
@@
 Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path.
-Chung and Khatib also proposed a decoupled approach \citeauthor{7140082}, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with
-a discretized environment. This planning phase does not account for collisions, implying that replanning is required in case of failure. This approach is efficient 
-in the demonstrated scenarios. In highly-constraining cases such 
-as the car egress scenario that we address here, we believe that including collision constraints in the planning is a requirement.
+Chung and Khatib also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment \citeauthor{7140082} .
+This planning phase does not account for collisions, implying that replanning is required in case of failure.
+In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}.
 

### 2017-04-19 body/results.tex

--- 2017-01-20 body/results.tex
@@
 
+TODO
+
 Table~\ref{tab:compprev} presents the computation times for these scenarios, clearly demonstrating that our approach is order of magnitude faster than previous works.
+
+Table~\ref{tab:compprev} presents the results we obtained
+
+Nonetheless, we indicate in Table~\ref{tab:compprev} what is actually computed by each method and in how much time.
+We compare motions obtained in previous works with HRP-2 (or a humanoid avatar) with ours.
+The two scenarios we selected consist in:  climbing one step, of height 20, 30 or 40 centimeters; 
+getting out of a table, that we consider of a complexity similar to the car egress scenario (or inferior since the feet locations remain on the ground in the former case).
+For scenarios that apply, we also indicate the interpolation time that we obtained by feeding the plan to our interpolation method presented in \cite{Carpentier2016}.
+
+Despite the difficulty of comparing the methods, we believe the results presented in Table~\ref{tab:compprev} indicate that our planner holds the promise of planning complete trajectories significantly faster than previous works.

### 2017-04-19 body/conclusion.tex

@@
 %~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
-Our contribution to \Pb is a fast contact generation scheme that canoptimize user-defined criteria.
+Our contribution to \Pb is a fast contact generation scheme that can optimize user-defined criteria.
 
@@
 \textbf{Regarding generality}, the \textit{reachability condition}, coupled with an approach based on limb decomposition, 
-allows the method to address automatically arbitrary multiped robots.
+allows the method to address automatically arbitrary legged robots.
 %
@@
 While we have exhibited complete multi-contact locomotion obtained with our contact planner, our main concern for future work is to address the interpolation
-between contact sequences ($\mathcal{P}_3$), not addressed here, and which remains an open issue in highly-constrained scenarios.
-Solving $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem,	but also improving
-our plans with transition certificates, i.e. guarantees that the interpolation between two contact configurations is feasible.
+between contact sequences ($\mathcal{P}_3$), which remains an open issue in highly-constrained scenarios.
+Solving $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem in the interpolation phase, an issue 
+not addressed by existing frameworks. We aim at providing our plans with transition certificates, that would define constraints on $\mathcal{P}_3$, under which
+the transition between two contact configurations is feasible and collision-free.
 %~ Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
@@
 We believe that the most promising direction in this regard is to integrate the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
-Addressing these two issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.
+Addressing these two issues is essential to bridge the gap between the planning and control aspects of legged locomotion.

### 2017-04-19 body/contact.tex
no symbol redefinition

@@
 In the second phase, we compute a discrete sequence of static equilibrium configurations $\mathbf{Q}^{\overline{0}}$ given a root path
-$\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Reach}^0$}. This contact planner is built on top of
-a contact generator, used to generate static equilibrium configurations. We first describe the generator, before detailing the contact planning algorithm.
+$\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Reach}^0$}.
+This contact planner uses a contact generator, used to generate static equilibrium configurations.
+We first describe the contact planning algorithm, before describing
+the contact generator.
 %~ Our planner computes guide paths in $C_{Reach}^0$ , an approximation of \gls{$C_{Equil}^0$}.
@@
 
-\subsection{Contact generator}
-\label{sec:single_contact}
-
-\begin{figure*}
-  \centering
-  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
-		\put (1,1) {a} 
-		\put (22,1) {b} 
-		\put (42,1) {c} 
-		\put (62,1) {d} 
-		\put (83,1) {e} 
-	\end{overpic}
-  \caption{Generation of a contact configuration for the right leg of HRP-2.
-(a): Selection of reachable obstacles.
-(b): Entries of the limb samples database (with $N = 4$).
-(c): With a proximity query between the octree database and the obstacles, configurations too far from obstacles are discarded.
-(d): The best candidate according to a user-defined heuristic $h$ is chosen.
-(e): The final contact is achieved using inverse kinematics.}
-  \label{fig:contact_gen}
-\end{figure*}
-
-The contact generation treats each limb $R^k$ independently.
-Given a configuration of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
-\begin{equation}
-\label{eq:contact_gen}
-	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \wedge \mathbf{q}^k \in  C_{Contact}^k 
-\end{equation}
-
-Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
-This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.
-We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
-
-%~ Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
-%~ $R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
-%~ While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
-%~ \begin{enumerate}
-%~ \item Generate randomly a collision-free limb configuration;
-%~ \item Project the end-effector onto the closest surface with inverse kinematics;
-%~ \item If a valid solution is found, stop. Otherwise repeat from step 1.
-%~ \end{enumerate}
-
-%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
-
-We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum distance 
-between an effector and an obstacle is less than $\epsilon \in \mathbb{R}$.
-We then apply the following steps:
-\begin{enumerate}
-\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
-\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
-\item At runtime, when contact creation is required, intersect the octree and the environment to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
-\item Use a user-defined heuristic $h$ to sort $S$;
-\item If $S$ is empty, stop (failure). Else select the first configuration of $S$. Project it onto contact using inverse kinematics. (Figure~\ref{fig:contact_gen} (d) and (e));
-\item If Eq.~\ref{eq:contact_gen} is verified, stop (success). Otherwise remove the element from $S$ and go to step 5.
-\end{enumerate}
-
-%~ This approach has two main advantages.
-%~ First, this allows us to select a large number of candidates in a single proximity request.
-%~ Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
-%~ a locally-optimal candidate.
-%~ Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
-%~ Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
-%~ $N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
-The reader is invited to refer to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
-of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
-
-%~ It should be noted that a failure case for our contact generator does not necessarily mean that the planner will globally fail. Indeed,
-%~ for a given root configuration, if contact configurations are found for other limbs, resulting in static equilibrium, the configuration will be accepted despite
-%~ the failure to generate a contact.
-
 \subsection{Contact planning algorithm}
-Starting from an initial whole-body configuration, we use the contact generator to compute a sequence
+Starting from an initial whole-body configuration, we compute a sequence
 of whole-body configurations  $\mathbf{Q}^{\overline{0}}$ along the root path $\mathbf{q}^0(t)$.
-The algorithm details can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
+The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
 %~ Although we address acyclic contact sequences, the algorithm is deterministic in the order in which 
@@
 
-This set of rules defines a deterministic contact planning algorithm.
-The determinism of this algorithm makes it efficient, at the cost of not being probabilistically complete.
 %~ \deladp{However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.}
+
+\subsection{Contact generator}
+\label{sec:single_contact}
+
+\begin{figure*}
+  \centering
+  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
+		\put (1,1) {a} 
+		\put (22,1) {b} 
+		\put (42,1) {c} 
+		\put (62,1) {d} 
+		\put (83,1) {e} 
+	\end{overpic}
+  \caption{Generation of a contact configuration for the right leg of HRP-2.
+(a): Selection of reachable obstacles.
+(b): Entries of the limb samples database (with $N = 4$).
+(c): With a proximity query between the octree database and the obstacles, configurations too far from obstacles are discarded.
+(d): The best candidate according to a user-defined heuristic $h$ is chosen.
+(e): The final contact is achieved using inverse kinematics.}
+  \label{fig:contact_gen}
+\end{figure*}
+
+To generate contacts, we proceed as follows.
+The contact generator treats each limb $R^k$ independently.
+Given a configuration of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
+\begin{equation}
+\label{eq:contact_gen}
+	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \wedge \mathbf{q}^k \in  C_{Contact}^k 
+\end{equation}
+
+Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
+This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.
+We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
+
+%~ Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
+%~ $R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
+%~ While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
+%~ \begin{enumerate}
+%~ \item Generate randomly a collision-free limb configuration;
+%~ \item Project the end-effector onto the closest surface with inverse kinematics;
+%~ \item If a valid solution is found, stop. Otherwise repeat from step 1.
+%~ \end{enumerate}
+
+%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
+
+We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum distance 
+between an effector and an obstacle is less than $\epsilon \in \mathbb{R}$.
+We then apply the following steps:
+\begin{enumerate}
+\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$ (We choose $N=10^3$);
+\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
+\item At runtime, when contact creation is required, intersect the octree and the environment to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
+\item Use a user-defined heuristic $h$ to sort $S$;
+\item If $S$ is empty, stop (failure). Else select the first configuration of $S$. Project it onto contact using inverse kinematics. (Figure~\ref{fig:contact_gen} (d) and (e));
+\item If Eq.~\ref{eq:contact_gen} is verified, stop (success). Otherwise remove the element from $S$ and go to step 5.
+\end{enumerate}
+
+%~ This approach has two main advantages.
+%~ First, this allows us to select a large number of candidates in a single proximity request.
+%~ Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
+%~ a locally-optimal candidate.
+%~ Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
+%~ Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
+%~ $N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
+The reader is invited to refer to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
+of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
+
+%~ It should be noted that a failure case for our contact generator does not necessarily mean that the planner will globally fail. Indeed,
+%~ for a given root configuration, if contact configurations are found for other limbs, resulting in static equilibrium, the configuration will be accepted despite
+%~ the failure to generate a contact.
+
+%~ In this Section, we have thus defined a deterministic contact planning algorithm.
+%~ The determinism of this algorithm makes it efficient, at the cost of not being probabilistically complete.

### 2017-04-19 cover_letter.tex

@@
 \author {}
-\title {Cover letter for the IJRR submission \\ ``An efficient acyclic contact planner for multiped robots''}
+\title {Cover letter for the original IJRR submission}
 \date {}
@@
 The current submission is an extension to our conference paper ``A reachability-based planner for sequences of acyclic contacts in
-cluttered environments'', accepted to ISRR 15 (\url{http://stevetonneau.fr/files/publications/isrr15/isrr15.pdf}).
+cluttered environments'', accepted to ISRR 15.
 
@@
 
-Sincerely, \\
-
-The authors
-
 \includepdf[pages=-]{reviewisrr.pdf}

### 2017-04-19 body/efort.tex
updated citations

--- 2017-01-23 body/efort.tex
@@
 $\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
-\citeauthor{Yoshikawa1984} defines the velocity~\ref{eq:vel} and force~\ref{eq:vel} ellipsoids:
+Yoshikawa~\cite{Yoshikawa1984} defines the velocity~\ref{eq:vel} and force~\ref{eq:vel} ellipsoids:
 
@@
 The longer the axis of the ellipsoid is, the more important the velocity (resp. force) of the end-effector the direction of the axis can be (Figure~\ref{sec:efort_ellipsoid}).
-
-%~ Similarly, \citeauthor{Yoshikawa1984} also defines the force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right):
-
-%~ \begin{equation} \label{ellipsoidforce}
-%~ \mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
-%~ \end{equation}
-%~ with $\mathbf{f}$ is a force vector expressed in the task space $\mathbb{R}^m$;
 
@@
 In other less demanding cases, manipulability can also be considered to avoid singularities.
-To do so, we can consider the manipulability measure $h_{w}$, also given by \citeauthor{Yoshikawa1984}:
+To do so, we can consider the manipulability measure $h_{w}$, also given by Yoshikawa:
 
@@
 the greater $h_{w}$ is, the further away the configuration is from singularity.
-
-%~ $h_{\textrm{\it EFORT}}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
-%~ can be precomputed off-line), that allow the planner to select the best
-%~ candidates according to user-defined criterion.
-
-%~ This results in faster computation times, and a reduced number of overall contact switches.
-
-%~ where the $\dot$ symbol represents the time derivative.  It describes the set of end-effector velocities that can
-%~ be reached under the constraint \eqref{ball} for the current configuration.
-%~ The longer the axis of the ellipsoid is, the faster the end-effector can move along the direction of the axis (Figure~\ref{sec:efort_ellipsoid}).
- %~ 
-%~ Similarly, \citeauthor{Yoshikawa1984} also defines the force ellipsoid(Figure~\ref{sec:efort_ellipsoid} - right):
- %~ 
-%~ \begin{equation} \label{ellipsoidforce}
-%~ \mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
-%~ \end{equation}
-%~ where $\mathbf{f}$ is a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
-%~ the equivalent joint torque vector $\bm{\tau}$;
- %~ 
-%~ \begin{equation} \label{jac}
-%~ \mathbf{\dot{p}}^k = \mathbf{J}^k(\mathbf{q}^k) \dot{\mathbf{q}}^k
-%~ \end{equation}
-%~ 
-%~ For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
- %~ 
-%~ As a linear approximation of a forward-kinematics function, $\mathbf{J}$ describes how small
-%~ variations from the configuration $\mathbf{q}$ affect the position vector $\mathbf{p}$.
-
-%~ Now we consider the unit ball in the configuration space $C$ defined by the set of joint velocities
-%~ for which the norm is at most 1:
-%~ \begin{equation} \label{ball}
-%~ ||\dot{\mathbf{q}}||^2 \leq 1
-%~ \end{equation}
-%~ 
-%~ We assume that $\mathbf{J}$ is full rank (we are not interested in singular configurations, which we discard).
-%~ From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
-%~ \begin{equation} \label{jaceq}
-%~ \mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} = \dot{\mathbf{q}}^T \dot{\mathbf{q}}
-%~ \end{equation}
-
-%~ We can use \eqref{jaceq} to map the ball into an ellipsoid in the Euclidian space $\mathbb{R}^m$:
-
-%~ \begin{equation} \label{ellipsoid}
-%~ \mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} \leq 1
-%~ \end{equation}
-%~ This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of end-effector velocities that can
-%~ be reached under the constraint \eqref{ball} for the current configuration.
-%~ The longer the axis of the ellipsoid is, the faster the end-effector can move along the direction of the axis.
-%~ Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.
-
-%~ \begin{figure}[!tbp]
-  %~ \centering
-	%~ \begin{overpic}[width=1\linewidth]{figures/EFORT/ellipsoid}
-		%~ \put (8.5,1.2) {\small{Velocity ellipsoid}}
-		%~ \put (62.5,1.2) {\small{Force ellipsoid} \tiny{(scale 0.5)}}
-	%~ \end{overpic}
-  %~ \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 dofs and 2 segments.
-%~ Only the horizontal and vertical speeds are shown (not the rotation speeds), since it would require being able to draw in four dimensions.}
-		   %~ \label{sec:efort_ellipsoid}
-%~ \end{figure}
-
-%~ Similarly to the velocity ellipsoid, \citeauthor{Yoshikawa1984} also defines the force ellipsoid.
-%~ Considering: a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
-%~ the equivalent joint torque vector $\bm{\tau}$;
-%~ we can define the mechanical work in both spaces:
-%~ \begin{equation*} \label{power}
-%~ \dot{\mathbf{q}}^T \bm{\tau} = \dot{\mathbf{p}}^T \mathbf{f}
-%~ \end{equation*}
-%~ 
-%~ 
-%~ Exploiting \eqref{jac}, we can easily see that the set of achievable forces in $\mathbb{R}^m$ subject to the constraint:
-%~ \begin{equation*} \label{ballforce}
-%~ ||\bm{\tau}||^2 \leq 1
-%~ \end{equation*}
-%~ is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right):
-%~ \begin{equation} \label{ellipsoidforce}
-%~ \mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
-%~ \end{equation}

### 2017-04-19 body/intro.tex
updated citations

--- 2017-01-23 body/intro.tex
@@
 
-The issue of planning acyclic contacts was first completely described by \citeauthor{Bretl:2006:MPM:1124573.1124585}.
+The issue of planning acyclic contacts was first completely described by Bretl~\cite{Bretl:2006:MPM:1124573.1124585}.
 The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path.
@@
 A similar approach has been considered for manipulation by \cite{gabicciniisrr15}.
-\citeauthor{DBLP:conf/humanoids/DeitsT14} proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered.
-\citeauthor{dai2014whole} extended the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
+Deits and Tedrake proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered, and equilibrium is not considered~\cite{DBLP:conf/humanoids/DeitsT14}.
+Dai et al.\cite{dai2014whole} extended the work of Posa et al.\cite{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
 the contacts manually for complex interactions.
@@
 \textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity.
-The feasibility and interest of the decoupling has been shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
+The feasibility and interest of the decoupling has been shown by Escande et al.\cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
 an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
@@
 This approach is suboptimal because the quality of the motion depends on the quality of the guide path.
-\citeauthor{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
+Bouyarmane et al.~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
 Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
 Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path.
-\citeauthor{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment.
+Chung and Khatib~\cite{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment.
 This planning phase does not account for collisions, implying that replanning is required in case of failure.
@@
 Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem.
-
-%
-%
-%% TO BE MOVED IN DISCUSSION 
-%%
-%%the guide path must guarantee the existence of a contact sequence to actuate it.
-We call this property \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
-This property has not been studied yet; the only rigorous way to validate a waypoint in the path is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
-\\
-%
-%\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinitely many combinations of possible contact sequences for a given root path.
-The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to constraining environments (\citeauthor{bouyarmane:lirmm-00777727} and \citeauthor{DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, possibly leading to very tedious transitions).
-\\
-%
-%
-%\textbf{The key idea of the paper is the \textit{reachability condition}}, a computationally-efficient approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
-%We assume that most of the times the \textit{reachability condition} implies the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}}.
-%%root configurations in this approximation can be efficiently extended into full body configurations in static equilibrium.
-%This assumption is not always verified, but we provide empirical evidence that it is in the considered class of problems, the \textit{\gls{cluttered} contact planning problems}.
-%Such problems can be solved with a contact sequence 
-%where in every configuration at least one contact is \gls{quasi-flat}~\citep{Prete2016}. A contact is \textit{quasi-flat} if the friction cone contains the direction opposite to gravity. 
-%%~ This definition includes motions on flat or uneven terrains, but also more complex ones such as stair climbing, car egress, or standing up scenarios, that might require noncoplanar contacts.
-%%Under this assumption we can take advantage of the decoupling approach to find a solution rapidly.
-%
-%%Similarly, in this paper we do not  prove that all the contact sequences produced by our planner can be interpolated ($\mathcal{P}_3$), although we have been able to do this for some of the demonstrated sequences~\citep{Carpentier2016}.
-%
-%
-%To address $\mathcal{P}_1$, we first use the \textit{reachability condition} to plan the guide path. This step is fast because the planning happens
-%in a low-dimensional space, and does not require explicit contact computations.
-%
-%Then we address \Pb by extending the path into a sequence of whole-body configurations in static equilibrium.
-%This second step requires the explicit computation of contact configurations.
-
-%
-%
-%\subsection{Paper contribution and organization}
-%We propose contributions to both problems \Pa and $\mathcal{P}_2$.
-%\begin{itemize}
-%\item A low-dimensional, efficient sampling-based approach to plan guide paths.
-%\item A very efficient and general implementation of an acyclic contact planner, the first one compatible with \gls{interactive} applications.
-%\item Four heuristics for contact generation. They bias the planner towards configurations that are in robust static equilibrium, or 
-%more efficient with respect to the task.
-%\item Statistical tests that empirically demonstrate the validity of our approach in the considered scenarios. 
-%\end{itemize}
-%
-%%Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
-%%Since the interpolation of a complete path between such configurations can be solved within \gls{interactive} constraints \citep{Carpentier2016}, we claim that our planner is the fastest multi-contact planning solution to our knowledge.
-%
-%
-%In Section~\ref{overview}, we present an overview of our method.
-Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and $\mathcal{P}_2$.
-In Section~\ref{sec:heuristics}, we present heuristics used for the selection of a contact configuration.
-Finally, in Section 6 we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios.
-%%that complete those presented in the conference paper.
-%
-%\subsection*{Comparison with our previous work}
-%The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep{tonneauisrr15}.
-%The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
-%This extension completes this work and brings new contributions, resulting from the implementation of our approach on real-world robots and problems.
-%In particular, a strong contribution is the introduction of a robust equilibrium criterion, designed to ensure the equilibrium of the robot despite bounded errors in the contact forces, coupled
-%with heuristics for relevant contact selection. These two contributions were required to apply our method to real-robot models, namely HRP-2 and HyQ~\citep{semini11hyqdesignjsce}.
-%In the present paper, a complete algorithm for the contact planner is given, and an access to the source code of our implementation is provided.
-%Furthermore, our experiments are supported by an in-depth analysis of our performances, as well as a discussion on the influence of the parameters to the method.

### 2017-04-19 body/results.tex
updated citations

--- 2017-01-20 body/results.tex
@@
 
-%~ \begin{table*}
-%~ \centering
-%~ \begin{tabular}{ c | c | c | c | c }
- %~ Scenario & Method  & Guide Path & Contact sequence & Interpolation \\
- %~ \hline
-   %~ \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}5.42 min}  \\
-							 %~ & \textbf{Ours} + \citeauthor{Carpentier2016}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & $ <$ 2s \\
-							 %~ & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
- %~ \hline
-   %~ \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}4.08 min}  \\
-							 %~ & \textbf{Ours}  & \textbf{5 ms} & $\mathbf{<}$ \textbf{18 ms} & X \\
-							 %~ & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
- %~ \hline
-   %~ \multirow{3}{*}{Stair 40 cm} & \citeauthor{Hauser06usingmotion} &  \multicolumn{3}{c}{\cellcolor{blue!25}10.08 min}  \\
-							 %~ & \textbf{Ours}  & \textbf{5 s} & \textbf{3 s} & X \\
-							 %~ & \citeauthor{Mordatch:2012:DCB:2185520.2185539} &  \multicolumn{3}{c}{\cellcolor{blue!25}2 to 10 min} \\
- %~ \hline
-   %~ \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 10 min & \multicolumn{2}{c}{\cellcolor{blue!25}3.5 hours}  \\
-							 %~ & \textbf{Ours}  & \textbf{529 ms} & \textbf{$<$ 22 s} & X \\
- %~ \end{tabular}
-%~ \caption{Comparison between the computation times obtained by our method and previous ones, for each phase of the planning. Blue cells indicate merged times (because no
-%~ precise time information is available). The X symbol indicates scenarios where we did not address the interpolation of the contact sequence.}
-%~ \label{tab:compprev}
-%~ \quad
-%~ \end{table*}
-
 \begin{table*}
@@
  \hline
-   \multirow{3}{*}{Stair 20 cm} & \citeauthor{Hauser06usingmotion} &  5.42 min  \\							 
-							  & \citeauthor{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
-							 & \textbf{Ours} + \citeauthor{Carpentier2016}  & \textbf{$ <$ 2s} \\
+   \multirow{3}{*}{Stair 20 cm} & Hauser~\cite{Hauser06usingmotion} &  5.42 min  \\							 
+							  & Mordatch et al.\cite{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
+							 & \textbf{Ours} + \cite{Carpentier2016}  & \textbf{$ <$ 2s} \\
  \hline
-   \multirow{3}{*}{Stair 30 cm} & \citeauthor{Hauser06usingmotion} &  4.08 min  \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
+   \multirow{3}{*}{Stair 30 cm} & Hauser~\cite{Hauser06usingmotion} &  4.08 min  \\
+							 & Mordatch et al.\cite{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
 							 & \textbf{Ours}  & \textbf{$ <$ 2s}   \\
  \hline
-   \multirow{3}{*}{Stair 40 cm} & \citeauthor{Hauser06usingmotion} &  10.08 min  \\
-							 & \citeauthor{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
+   \multirow{3}{*}{Stair 40 cm} & Hauser \cite{Hauser06usingmotion} &  10.08 min  \\
+							 & Mordatch et al.\cite{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
 							 & \textbf{Ours}   & \textbf{$ <$ 5s}   \\
  \hline
-   \multirow{2}{*}{Table (car) egress} & \citeauthor{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 3.5 hours  \\
+   \multirow{2}{*}{Table (car) egress} & Bouyarmane et al.~\cite{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 3.5 hours  \\
 							 & \textbf{Ours}  & \textbf{$<$ 60 s} \\
@@
 \subsection{Validation of the contact plans, and comparison with previous work} \label{sec:compa}
-To validate our plans, we either use the framework proposed by \citeauthor{Carpentier2016} or our own implementation of a $\mathcal{P}_3$ solver, detailed in Appendix~\ref{app:optim}.
+To validate our plans, we either use the framework proposed in \cite{Carpentier2016} or our own implementation of a $\mathcal{P}_3$ solver, detailed in Appendix~\ref{app:optim}.
 The companion video shows the obtained motions.
@@
 Furthermore, $\mathcal{P}_3$ remains an open problem in the presence of obstacles.
-The only valid scenarios addressed completely in previous works are thus the stair-climbing scenarios of different heights proposed by \citeauthor{Hauser06usingmotion}, and the table-egress scenario by~\citeauthor{DBLP:conf/iser/EscandeKMG08}, which we consider to be of similar complexity with respect to the car-egress scenario (we did not consider the stairs in the scene).
+The only valid scenarios addressed completely in previous works are thus the stair-climbing scenarios of different heights proposed by Hauser in \cite{Hauser06usingmotion}, and the table-egress scenario by Escande et al.
+in~\cite{DBLP:conf/iser/EscandeKMG08}, which we consider to be of similar complexity with respect to the car-egress scenario (we did not consider the stairs in the scene).
 Both scenarios are tested with HRP-2.

### 2017-04-19 body/abstract.tex
corrections fro mrev 1

@@
 % !TEX root =  ../main.tex
-We present a contact planning framework for complex legged locomotion tasks: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
-The need for such a planner was highlighted at the Darpa Robotics Challenge, where multi-contact behaviors
-could not be demonstrated.
+We present a contact planner for complex legged locomotion tasks: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
+The need for such a planner was shown at the Darpa Robotics Challenge, where such behaviors
+could not be demonstrated (except for egress).
 

### 2017-04-19 body/conclusion.tex
corrections fro mrev 1

@@
 This paper focuses on contact planning of multi-contact locomotion. While it does not address directly the continuous motion generation,
-we had to prove that the plans produced by our planner can be executed by the considered robots. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
-and most existing solutions do not consider collisions with the environment during the planning, making it hard to validate constrained problems such as the polaris scenario
-or the table egress in \cite{DBLP:conf/iser/EscandeKMG08}.
-The companion video shows the results that we obtained in simulation and on the real HRP-2 robot. Despite the fact that our method does not consider joint torque limits in the planning
-phase, the robot is able to execute our plans, achieving to demonstrate the interest of our framework.
+we had to prove that the plans produced by our planner can be executed by the considered robots.
+Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory}, for which the implemented a solution in this paper, described in appendix
+%~ and most existing solutions do not consider collisions with the environment during the planning, making it hard to validate constrained problems such as the polaris scenario
+%~ or the table egress in \cite{DBLP:conf/iser/EscandeKMG08}.
+The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.
+In these scenarios the robot is able to execute our plans, while respecting the constraints inherent to the problem: joint kinematics and torque limits and
+collision avoidance.
 %~ \adnote{Probably you should add in appendix an explanation of how you solve $P_3$ and refer to it here.}

### 2017-04-19 body/overview.tex
corrections fro mrev 1

@@
 \medskip
-\textbf{A robot} is a kinematic chain $R$ composed of: a root $R^0$, and $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
+\textbf{A robot} is a kinematic tree $R$ composed of: a root $R^0$, and $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
 The root has $r \geq 6$ DoFs: for instance, HRP-2 has two extra DoFs in the torso, such that we have $r=8$.
@@
 \medskip
-\noindent $W^0$ is a volume encompassing $R^0$ (Figure~\ref{fig:HyQ_roms}). \\
+\noindent $W^0$ is a volume encompassing $R^0$ (Figure~\ref{fig:HyQ_roms}), or equal to it. $W^0$ is typically a low-polygonal bounding shape
+of $R^0$ for performance reasons.\\
  %~ (Figure~\ref{fig:character}-right: central cylinder). 
@@
 where $\mathbf{p}^k$ denotes the end-effector position (in the root frame) of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{j.lim}$ is the space
-of admissible limb joint configurations. We also define \mbox{$W = \bigcup_{k=1}^{l}W^k$}, and
-$W^k(\mathbf{q}^{0})$ (for $1 \leq k \leq l$) as the volume $W^k$ for the root configuration $\mathbf{q}^{0}$.
+of admissible limb joint configurations. We also define \mbox{$W = \bigcup_{k=1}^{l}W^k$}.
+
+Each $W^i, 0 \leq i \leq l$ is  attached to one joint of the root, such that 
+$W^i(\mathbf{q}^{0})$ is the volume $W^i$ transformed the by root configuration $\mathbf{q}^{0}$.
 

### 2017-04-19 body/rbprm.tex
cosmetics
corrections fro mrev 1

@@
 %~ \section{Computing a guide path in $C^0_{reach}$ ($\mathcal{P}_1$) }
-\section{Root path planning in the \textit{contact reachable space}}
+\section{Root path planning in the contact reachable space}
 \label{rbprm}
@@
 \begin{equation}
-\mathcal{C}^0_{\textrm{\it Nec}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+\mathcal{C}^0_{\textrm{\it Nec}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
@@
 \begin{equation}
-\mathcal{C}^0_{\textrm{\it Suf}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge B^{\textrm{\it Suf}}(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+\mathcal{C}^0_{\textrm{\it Suf}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ B^{\textrm{\it Suf}}(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
@@
  \label{eq:reachc}
-C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
+C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
  %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
@@
 The only significant change is to replace the collision checking with the \textit{reachability condition}.
-Our current implementation of these modifications is based on the Bi-RRT planner \citep{770022} provided by the HPP software~\citep{7759083}.
+Our current implementation of these modifications is based on the Bi-RRT planner \citep{770022} provided by the HPP software~\citep{7759083}. \\
 
@@
 
-\subsubsection*{}
 This Section has presented a guide path planner for the geometric root of a robot. This planner is implemented as a low-dimensional sampling-based 

### 2017-04-19 body/conclusion.tex
cosmetics

@@
 \subsection{Addressing $\mathcal{P}_3$}
-This paper focuses on contact planning of multi-contact locomotion. While it does not address directly the continuous motion generation,
-we had to prove that the plans produced by our planner can be executed by the considered robots.
-Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory}, for which the implemented a solution in this paper, described in appendix
-%~ and most existing solutions do not consider collisions with the environment during the planning, making it hard to validate constrained problems such as the polaris scenario
-%~ or the table egress in \cite{DBLP:conf/iser/EscandeKMG08}.
+This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation,
+. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
+for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
 The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.

### 2017-04-19 body/intro.tex
cosmetics

@@
 Our work thus aims at breaking the complexity of the acyclic contact planning problem.
-To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorial of the contact selection problem.
+To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorics of the contact selection problem.
 First we introduce a low-dimensional space, called the \textit{contact reachable} space, that can be sampled and mapped efficiently to the contact manifold.
@@
 
-Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorial issue.
+Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorics.
 Two main directions have been explored.

### 2017-04-19 body/results.tex
cosmetics

@@
 
-\begin{table*}
+\begin{table}
 \centering
@@
 \quad
-\end{table*}
+\end{table}
 
-\begin{table*}
+\begin{table}
 \centering
-\begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{65pt} | c}
+\begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{35pt} | >{\centering\arraybackslash}m{35pt} | c}
   &  Equilibrium success rate & Kinematic failure & Equilibrium failure \\
@@
 \quad
-\end{table*}
+\end{table}
 
@@
 
-\begin{table*}
+\begin{table}
 \centering
 \begin{tabular}{ c | c | c }
- Scenario & Method  & Complete computation time \\
+ Scenario & Method  & Computation time \\
  \hline

### 2017-04-19 body/conclusion.tex
reviewer 2

@@
 Figure~\ref{fig:dedefeas} illustrates this observation, demonstrated empirically by the third column of Table~\ref{tab:requestpercent}. In the worst case, in our experiment
-the assumption was verified for 82\% of the total amount of trials that verified \contactreachability\.
+the assumption was verified for 82\% of the total amount of trials that verified \textit{contact reachability}.
 In the example of \cite{Bouyarmane2009}, the verification of \equilibriumfeasibility\ implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$, requiring
@@
 Our results demonstrate that our method allows a pragmatic compromise between three 
-criteria that are hard to conciliate: generality, performance, and quality of the solution, making it the first acyclic contact
+criteria that are hard to reconcile: generality, performance, and quality of the solution, making it the first acyclic contact
 planner compatible with \gls{interactive} applications.

### 2017-04-19 body/overview.tex
reviewer 2

@@
 The dimension of the path is equal to the number of degrees of freedom (DoFs) of the root of the robot.
-Similarly to previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: there must exist a joint configuration that results in static equilibrium for each root configuration.
+Similarly to previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: there must exist a joint configuration that results in static equilibrium for each root configuration\footnote{Enforcing static equilibrium is a classical, conservative approach to reduce the search space and considering states  with non-zero accelerations and velocities, which can't be connected trivially.
+This does not mean that the final motion will necessary be quasi-static.}.
 Previous work verifiy \textit{equilibrium feasibility} by explicitly computing such a configuration.

### 2017-04-19 body/results.tex
reviewer 2

@@
 In this section we present some of the results obtained with our planner. The complete sequences are shown in the companion video.
-Specifically, we demonstrate the planner for two really different robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
+Specifically, we demonstrate the planner for two legged robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
 Finally, a last example suggests possible applications to dexterous manipulation.
@@
 
-\subsubsection{Choosing the scaling factor $s$:} \label{sec:params}
+\subsubsection{Choosing the scaling factor $s$} \label{sec:params}
 %~ \deladp{To find a convenient value $s^*$ of $s$, we proceeded as follows. }
@@
 
-As it can be expected, the scaling results in a high increasepercentage of the sensitivity, with a decrease of the specificity.
+As it can be expected, the scaling results in a high increase of the sensitivity, with a decrease of the specificity.
 For HRP-2 we decided to set $s^*=1.2$.
@@
 
-\subsubsection{Choosing the heuristics:} \label{sec:heuristichoices}
+\subsubsection{Choosing the heuristics} \label{sec:heuristichoices}
 In our conference paper~\citep{tonneauisrr15}, the computed motions were generated using the EFORT heuristic.
@@
 In locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
-configurations close to singularities (and joint limits). While this does not significantly impact
+configurations close to singularities (and joint limits). While this did not significantly impact
 the generation of the plan, the resulting interpolation turned out to be harder.
 For this reason, we prefer to use our manipulability-based heuristic for the legs of the robot, but we still
-use EFORT for the arms, which results in less contact repositionings.
+use EFORT for the arms, which results in fewer contact repositionings.
 
@@
 In our experiments, we choose not to maximize the robustness,
-%~ but to empirically set a robustness threshold value under which a configuration is not considered to be in
+%~ but to empirically se	t a robustness threshold value under which a configuration is not considered to be in
 %~ static equilibrium (If no candidate reaches the threshold, instead of failing, the algorithm can eventually return the ``more robust'' configuration
@@
 
-\subsubsection{Discretization of the guide path:} \label{sec:disc}
+\subsubsection{Discretization of the guide path} \label{sec:disc}
 The discretization step is a user-defined, fixed parameter. The step

### 2017-04-20 body/abstract.tex
mise en page

@@
 
-Indeed, current planners suffer from prohibitive computation times, due to the difficulty
-of generating contact postures: the feasible space (the contact manifold) is of high dimension,
-and impossible to sample.
+Current planners suffer from prohibitive computation times, due to the difficulty
+of generating contact postures in a high-dimensional solution space that is impossible to sample.
 

### 2017-04-20 body/intro.tex
mise en page

@@
 the contact plan is manually designed by a time consuming trial-and-error procedure.
-Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in human-centered environments, which are cluttered and comprise constraining obstacles such as stairs.
+Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in cluttered, human-centered environments, which comprise constraining obstacles (stairs...).
 
@@
 For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
-As a result at the DARPA Robotics Challenge, robots relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
+As a result at the DARPA Robotics Challenge, the participants stated that except for egress, the robots did not use multi contact strategies: they relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
 
@@
 Then, given a path computed in the \textit{contact reachable} space, we propose a deterministic algorithm to generate a contact sequence along the path.
-This decoupling presents drawbacks and advantages discussed in previous related literature, which we summarize in the following.
+This decoupling presents pros and cons discussed in previous related literature, summarized in the following.
 %~ whole-body trajectory and the contact location and timings, we decouple path and contact planning, as suggested by \citeauthor{Bouyarmane2009}.
@@
 Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem.
+
+This paper extends our previous conference paper~\citep{tonneauisrr15} in several ways.
+Our approach is now demonstrated on actual robots and in simulation, contrary to virtual avatars before, in a large variety of contexts.
+To achieve this we introduce a new criterion to efficiently compute the equilibrium of the robot while accounting for uncertainties of the real world.
+We provide the complete pseudo code of algorithm.
+We give the complete procedure to use our planner with any legged robot and provide our source code, allowing any user to directly use our method.
+We also propose a solution to $\mathcal{P}_3$, making our framework complete.
+Finally we provide an in-depth analysis of the results we obtain, and a theoretical discussion on the capabilities of the planner and its current limitations.

### 2017-04-20 body/contact.tex
cosmetics
reviewer 2
cosmetics + performance

@@
 The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
-%~ Although we address acyclic contact sequences, the algorithm is deterministic in the order in which 
-%~ the contacts are created, allowing it to break the combinatorial.
 
@@
 
-\subsubsection{Maintaining a contact in the sequence:}
+\subsubsection{Maintaining a contact in the sequence}
 
@@
 %\subsection{Generation of a contact configuration}  
-\subsubsection{Creating contacts:}
+\subsubsection{Creating contacts}
 Contacts are created using a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest.
@@
 \label{eq:contact_gen}
-	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \wedge \mathbf{q}^k \in  C_{Contact}^k 
+	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \textrm{ \textbf{and}}\ \mathbf{q}^k \in  C_{Contact}^k 
 \end{equation}

### 2017-04-20 body/overview.tex
cosmetics + performance

@@
 
-\subsection{Computation of a guide path --- \Pa}
+\subsection{Computation of a guide path --- \Pa (Section~\ref{rbprm})}
 We first consider the problem of planning a root guide path (Figure~\ref{fig:framework}--$\mathcal{P}_1$).
@@
 This does not mean that the final motion will necessary be quasi-static.}.
-Previous work verifiy \textit{equilibrium feasibility} by explicitly computing such a configuration.
+Previous works verifiy \textit{equilibrium feasibility} by explicitly computing such a configuration.
 We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact rechability}, illustrated in the following.
@@
 To plan a root path, we then use an RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
-This is detailed in Section~\ref{rbprm}.
 %~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).
@@
 
-In the remainder of this paper, we use the terms \gls{contact reachable} and \gls{equilibrium feasible} to qualify either a root configuration, a whole-body configuration, or a set
+In the remainder of this paper, we use the terms \gls{contact reachable} and \gls{equilibrium feasible} to qualify either a root or a whole-body configuration, or a set
 of such configurations.
 
-\subsection{Generating a discrete sequence of contact configurations --- \Pb}
+\subsection{Generating a discrete sequence of contact configurations --- \Pb (Section~\ref{sec:contact})}
 The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--$\mathcal{P}_2$).
@@
 
-The details of the contact sequence generation are given in Section~\ref{sec:contact}.
+%~ The details of the contact sequence generation are given in Section~\ref{sec:contact}.
 
@@
 The root has $r \geq 6$ DoFs: for instance, HRP-2 has two extra DoFs in the torso, such that we have $r=8$.
-Therefore $R$ is fully described by a configuration $\mathbf{q} \in  \mathbb{R}^{r+n}$, with $n$ the number of joint DoFs.
+Thus $R$ is fully described by a configuration (a vector of joint values) $\mathbf{q} \in  \mathbb{R}^{r+n}$, with $n$ the number of joint DoFs.
 $\mathbf{q}$ is decomposed as follows:
 \begin{itemize}
-	\item $\mathbf{q}^k$ is a configuration (a vector of joint values) of the limb $R^k$; % (Figure~\ref{fig:character});
+	\item $\mathbf{q}^k$ is a configuration  of the limb $R^k$; % (Figure~\ref{fig:character});
 	\item $\mathbf{q}^{\overline{k}}$ is a vector of joint values of R \textbf{not} related to $R^k$.
@@
 \medskip
-\noindent $W^0$ is a volume encompassing $R^0$ (Figure~\ref{fig:HyQ_roms}), or equal to it. $W^0$ is typically a low-polygonal bounding shape
-of $R^0$ for performance reasons.\\
+We then define a set of 3d volumes  $W^i, 0 \leq i \leq l$, each  attached to one joint of the root, such that 
+$W^i(\mathbf{q}^{0})$ describes the world position of $W^i$ for the root configuration $\mathbf{q}^{0}$.
+
+\noindent $W^0$ is a volume encompassing $R^0$ (Figure~\ref{fig:HyQ_roms}), or equal to it\footnote{$W^0$ is typically a low-polygonal bounding shape
+of $R^0$ for performance.}.\\
  %~ (Figure~\ref{fig:character}-right: central cylinder). 
@@
 
-Each $W^i, 0 \leq i \leq l$ is  attached to one joint of the root, such that 
-$W^i(\mathbf{q}^{0})$ is the volume $W^i$ transformed the by root configuration $\mathbf{q}^{0}$.
+%~ Each $W^i, 0 \leq i \leq l$ is  attached to one joint of the root, such that 
+%~ $W^i(\mathbf{q}^{0})$ is the volume $W^i$ transformed the by root configuration $\mathbf{q}^{0}$.
 
@@
   \caption{
-           Reachable workspace and torso bounding box of HyQ. The green shapes represent the reachable workspace $W^k$ of each limb. The red shape is $W^0$.}
+           Reachable workspace and torso bounding box of HyQ. Each green shape represent a reachable workspace $W^k$ of a limb. The red shape is $W^0$.}
 		   \label{fig:HyQ_roms}

### 2017-04-20 body/results.tex
cosmetics + performance

@@
 
-At the end of the video, we validate our contact plans by exhibiting a solution to the interpolation problem $\mathcal{P}_3$ for these scenarios.
+At the end of the video, we validate our contact plans by exhibiting a solution to the interpolation problem $\mathcal{P}_3$.
 
@@
 In all the scenarios considered, the formulation of the problem is always the same:
-a start and goal root configurations are provided as input.
+a start and goal root configurations are provided as input (except in the stair climbing scenario where the start whole body configuration is given).
 The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
@@
 %\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}
-\subsubsection{HRP-2 -- Steep staircase (Figure~\ref{fig:stair_robust}):}
+\subsubsection{HRP-2 -- Steep staircase (Figure~\ref{fig:stair_robust})}
 
@@
 
-\subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing}):}
+\subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing})}
 From a bent configuration, the robot has to stand up using a wall as support, and climbing a 25-cm high step.
@@
 
-\subsubsection{HRP-2 -- Polaris car egress (Figure~\ref{fig:car}):}
+\subsubsection{HRP-2 -- Polaris car egress (Figure~\ref{fig:car})}
 In this scenario inspired from the DRC car egress HRP-2 has to step out of the Polaris car.
@@
 
-\subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs}):}
+\subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs})}
 In this long scene, HyQ has to cross a 55-cm large hole, followed by a narrow ``bridge'', only 25-cm large.
@@
 
-\subsubsection{HRP-2 -- Path re-planning (Figure~\ref{fig:re-planning}):}
+\subsubsection{HRP-2 -- Path re-planning (Figure~\ref{fig:re-planning})}
 In this long scene, HRP-2 plans a path through several obstacles. The scene is edited during the execution of the motion: a stair is added,
@@
 
-\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot}):}
+\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot})}
 This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
@@
 \subsection{Performance analysis} \label{sec:perf}
-To analyze performance, we ran the planner 1000 times for each considered scenario.
+To analyze performance, we ran the planner 1000 times for each scenario.
 We measured the computation time spent in each part of the algorithm, and analyzed their success rate.
@@
 
-\subsubsection{Computation times:}
-Table~\ref{tab:requestime} summarizes the computation times.
+\subsubsection{Computation times (Table~\ref{tab:requestime})}
+ %~ summarizes the computation times.
 
@@
 
-\subsubsection{Success rates:}
-Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
-Despite the complexity of the scenarios addressed, as well as the approximations made in our formulation, our planner succeeded in the large majority of cases.
+\subsubsection{Success rates (Table~\ref{tab:sucess_planning})}
+%~ Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
+Despite the complexity of the scenarios and the approximations made in our formulation, our planner succeeded in the large majority of cases.
 
@@
 \subsection{Validation of the contact plans, and comparison with previous work} \label{sec:compa}
-To validate our plans, we either use the framework proposed in \cite{Carpentier2016} or our own implementation of a $\mathcal{P}_3$ solver, detailed in Appendix~\ref{app:optim}.
+To validate our plans, we either use the framework of \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
 The companion video shows the obtained motions.
@@
 Table~\ref{tab:compprev} presents the computation times for these scenarios, clearly demonstrating that our approach is order of magnitude faster than previous works.
+
+To provide a completely fair comparison, it would be required to compare the methods on the same computers and exact same scenarios, which is not possible
+because the source code of the other methods is often not available. Despite this observation and the fact that some results are old, we believe that the results show that our method
+performs better. Indeed, a difference in technology does not justify a difference as important as the one demonstrated in Table~\ref{tab:sucess_planning}.

### 2017-04-20 body/appendix_algo.tex
final version ?

@@
 in which contacts are tested: we try to replace older contacts first when necessary.
-Thus the algorithm is deterministic even though it can handle acyclic motions.
+Thus the algorithm is deterministic even though it can handle acyclic motions. \\
 
@@
 in computing a configuration with only one contact switch occurring.
-If \textsc{GenFullBody} failed in achieving this, the method \textsc{RepositionContacts} is called.
+Otherwise, the method \textsc{RepositionContacts} is called.
 It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
 This repositioning allows to increase the odds that the contact can be maintained at the next step.
-
-The pseudo code for the method \textsc{GenFullBody} is given by Algorithm~\ref{alg:pi}.
+Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.
 

### 2017-04-20 body/appendix_hpp.tex

@@
 HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS, that can be found at \url{http://projects.laas.fr/gepetto/index.php/Software/Main}.
-The HPP modules implement the standard tools and algorithms used in motion planning,
+HPP implements the standard tools and algorithms used in motion planning,
 such as the Bi-RRT planner from which RB-RRT is derived.
 
-The robot models used in our experiments are described using the standard urdf file format, that is 
-compatible with HPP.
+The robot models used in our experiments are described using the standard urdf file format, compatible with HPP.
 

### 2017-04-20 body/conclusion.tex

@@
 
-These results clearly justify the pragmatic decisions of this paper.
+These results clearly justify our pragmatic approach.
 
@@
 %~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
-Finally, we aim at performing kinodynamic planning to remove the constraint that configurations in the contact plan be in static equilibrium.
-We believe that the most promising direction in this regard is to integrate the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
+Finally, we aim at performing kinodynamic planning to remove the constraint that configurations be in static equilibrium.
+We believe that the most promising direction in this regard is to integrate the notion of Admissible Velocity Propagation \citep{DBLP:conf/rss/PhamCN13}.
 Addressing these two issues is essential to bridge the gap between the planning and control aspects of legged locomotion.

### 2017-04-20 body/efort.tex
reviewer 2

@@
 
-\subsubsection{The force and velocity ellipsoids:}
+\subsubsection{The force and velocity ellipsoids}
 
@@
 $\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
-Yoshikawa~\cite{Yoshikawa1984} defines the velocity~\ref{eq:vel} and force~\ref{eq:vel} ellipsoids:
+Yoshikawa~\cite{Yoshikawa1984} defines the velocity(~\ref{eq:vel}) and force~(\ref{eq:for}) ellipsoids:
 
@@
 
-\subsubsection{Manipulability-based heuristics:}
+\subsubsection{Manipulability-based heuristics}
 From these definitions, we can derive three useful heuristics, that all account for the environment and the task being performed.

### 2017-04-20 body/equilibrium.tex

@@
 
-\subsubsection{Conditions for static equilibrium:}
+\subsubsection{Conditions for static equilibrium}
 We first define the variables of the problem, for $e$ contact points, expressed in world coordinates:
@@
 
-\subsubsection{Formulation of a robust LP:}
+\subsubsection{Formulation of a robust LP}
 Let $b_0 \in \mathbb{R}$ be a scalar value. We now define the following LP:

### 2017-04-20 body/intro.tex

@@
 Deits and Tedrake proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered, and equilibrium is not considered~\cite{DBLP:conf/humanoids/DeitsT14}.
-Dai et al.\cite{dai2014whole} extended the work of Posa et al.\cite{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
+Dai et al.~\cite{dai2014whole} extended the work of Posa et al.~\cite{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
 the contacts manually for complex interactions.
@@
 \textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity.
-The feasibility and interest of the decoupling has been shown by Escande et al.\cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
+The feasibility and interest of the decoupling has been shown by Escande et al.
+\cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
 an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
@@
 This approach is suboptimal because the quality of the motion depends on the quality of the guide path.
-Bouyarmane et al.~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
+Bouyarmane et al.
+~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
 Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).

### 2017-04-20 body/rbprm.tex

@@
 
-This Section has presented a guide path planner for the geometric root of a robot. This planner is implemented as a low-dimensional sampling-based 
-algorithm. Given start and goal configurations, it outputs a continuous path for the root of the robot. 
+This Section has presented a guide path planner for the geometric root of a robot, implemented as a low-dimensional sampling-based 
+algorithm. Given start and goal configurations, it outputs a continuous path for the robot's root. 
 %~ \deladp{Thanks to the \textit{reachability condition}, we assume that for any configuration in the path there exists a joint configuration that results in static equilibrium.}

### 2017-05-02 body/abstract.tex
Abstract nmsd.

@@
 could not be demonstrated (except for egress).
-
-Current planners suffer from prohibitive computation times, due to the difficulty
-of generating contact postures in a high-dimensional solution space that is impossible to sample.
-
-We tackle this issue by introducing the reachability condition. This condition allows us to define
-a geometric approximation of the contact manifold, which is of low dimension and can be sampled efficiently.
-Informally, the condition verifies that the root configuration of a robot is close, but not too close to obstacles: close to allow contact creation, not too close to avoid collision.
+%
+%This planning problem is theoretically known to be a specific instance of a larger class of complex planning problems (``manipulation'' or ``rearrangement'' planning).
+Current planners suffer from prohibitive algorithmic complexity, because they deploy a tree of robot configurations projected in contact with the environment.
+Contact configurations are costly to compute (because sampled and projected in high dimension) and have no equivalence property (i.e.
+two neighbor configurations cannot be directly connected in the planning graph).
+%
+We tackle this issue by introducing a reduction property: the reachability condition. 
+This condition defines a geometric approximation of the contact manifold, which is of low dimension, has Cartesian topology and can be efficiently sampled and explored.
+%
+The hard contact planning problem can then be separated into two sub-problems: 
+first, we plan a path for the root without considering the whole-body configuration, using a sampling-based algorithm; 
+second, we generate a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm. 
+%
+%We tackle this issue by introducing the reachability condition. This condition allows us to define
+%a geometric approximation of the contact manifold, which is of low dimension and can be sampled efficiently.
+%Informally, the condition verifies that the root configuration of a robot is close, but not too close to obstacles: close to allow contact creation, not too close to avoid collision.
 Thanks to this condition we decompose the hard contact planning problem into two
-sub-problems: first, planning a guide path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm.
-
-Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically: in a
-few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ, and a dexterous hand.
-These plans are then validated either in dynamic simulations, or on the real HRP-2 robot, thus demonstrating
-the first interactive multi-contact planner.
+%sub-problems: first, planning a guide path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm.
+%
+The reduction breaks the algorithm complexity encountered in other contact planners, logically leading to the open-source implementation of the first interactive multi-contact planner.
+%
+While no contact planner has yet been proposed with theoretical completeness, we empirically show by a large number of trials ($>$10,000) that it is not an issue for our method:
+in a few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ and a dexterous hand. 
+These plans are then validated either in dynamic simulations and on the real HRP-2 robot.
+%, thus demonstrating the first interactive multi-contact planner.
+%Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically

### 2017-05-02 body/intro.tex
First modif of intro.

@@
 \IEEEPARstart{L}{egged} robots move by sequentially creating contacts with the environment.
-After years of research, such robots can autonomously walk on flat ground, but struggle to navigate more complex environments.
-To climb stairs using a handrail, a humanoid robot must plan the contacts to create with its feet and hands.
-Computing such a contact sequence is an open motion planning problem called acyclic contact planning.
+After years of research, such robots can autonomously walk on flat ground, but struggle to navigate in more complex environments, e.g. to climb stairs using a handrail.
+In such a case a humanoid robot must plan the contacts it musts create with its feet and hands. Computing such a movement is an open motion-planning problem called acyclic contact planning.
+Most of the complexity of this problem lies in the choice of the sequence of contacts, i.e.
+the underlying decomposition of the trajectory into contact phases where specific points of the robot body are exerting forces on specific locations of the environment.
+%
+Once the contact plan is known, efficient approaches exist to generate a dynamically feasible motion~\citep{Carpentier2016}.
 
-Once the contact plan is known, efficient approaches exist to generate a dynamically feasible motion~\citep{Carpentier2016}. However in general
-the contact plan is manually designed by a time consuming trial-and-error procedure.
+%% However in general
+%% the contact plan is manually designed by a time consuming trial-and-error procedure.
 Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in cluttered, human-centered environments, which comprise constraining obstacles (stairs...).
 
-To make a legged robot walk on flat ground, efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
-Furthermore, choosing the effector with which to create a contact is trivial because walking follows a cyclic pattern (For a humanoid the left foot always follows the right foot).
-Unfortunately, in the general case planning complex contact interactions is extremely challenging:
-At any given time a contact must be chosen between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location).
-Furthermore, a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles; we say that the contact manifold is foliated~\citep{simeon-manipulation-04}.
-Lastly, the contact manifold has a null measure and is thus impossible to sample.
-This restriction prevents the use of efficient sampling-based planners.
+In the specific (and simple) case of gaited biped locomotion on flat ground, choosing the effector with which to create a contact is trivial because walking follows a cyclic pattern (left foot always follows right foot).
+Efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location. 
+%
+The general case of planning contact sequences is more challenging.
+At any given time a contact results of a combinatorial discrete choice for both the effector and contact surface, and a continuous choice for the contact location.
+Any choice of contact then kinematically and dynamically constrains the possible motion of the robot.
+Furthermore there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles.
+From this property, we say that the contact manifold is foliated~\citep{simeon-manipulation-04}. 
+The foliation prevents the use of efficient sampling-based planners for two reasons.
+(i) First, each submanifold of the foliation has a zero measure and cannot be directly sample.
+A sample is rather obtained by sampling a ``free flying'' configuration and explicitly projecting it in contact (which is a costly numerical operation).
+(ii) Second, the foliated topology turns the exploration by spreading a graph of configurations (probability roadmap, rapidely exploring random trees) into an unefficient random process, where many useless nodes are sampled on parallel submanifolds.
+The total algorithm complexity of classical contact planners comes both from the number of nodes of the graph that are sampled during the exploration (ii) and to the cost of the projection when sampling a new configuration (i).
+%
+For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
+As a result at the DARPA Robotics Challenge, the participants stated that, except for egress, the robots did not use multi contact strategies: they relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
 
-For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
-As a result at the DARPA Robotics Challenge, the participants stated that except for egress, the robots did not use multi contact strategies: they relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
-
-Obtaining reasonable computation times is critical for a contact planner to be of any practical use.
-Our work thus aims at breaking the complexity of the acyclic contact planning problem.
-To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorics of the contact selection problem.
-First we introduce a low-dimensional space, called the \textit{contact reachable} space, that can be sampled and mapped efficiently to the contact manifold.
-Then, given a path computed in the \textit{contact reachable} space, we propose a deterministic algorithm to generate a contact sequence along the path.
+%Obtaining reasonable computation times is critical for a contact planner to be of any practical use. 
+Our work aims at breaking the complexity of the acyclic contact planning problem.  
+To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold (i), and the combinatorics of the contact selection problem (ii). 
+To that end, we introduce a reduction property that we name the reachability condition.
+This condition relaxes the strict contact condition.
+It induces a low-dimensional space, called the \textit{contact reachable} space, that has a Cartesian topology, i.e. it can be efficiently sampled.
+Under some hypothesis, a path in the \textit{contact reachable} space can be easily mapped to the (high-dimensionnal) contact manifold. 
+This approach of the problem enables us to separate the contact planning problem in two tractable subproblems.
 This decoupling presents pros and cons discussed in previous related literature, summarized in the following.

### 2017-05-03 body/intro.tex
Candidate final I.0

@@
 \IEEEPARstart{L}{egged} robots move by sequentially creating contacts with the environment.
-After years of research, such robots can autonomously walk on flat ground, but struggle to navigate in more complex environments, e.g. to climb stairs using a handrail.
-In such a case a humanoid robot must plan the contacts it musts create with its feet and hands. Computing such a movement is an open motion-planning problem called acyclic contact planning.
+After years of research, such robots can autonomously walk on flat ground, but struggle to navigate in more complex environments, where deciding where to create a contact with its feet and possibly its hands is nontrivial, e.g.
+to climb stairs using a handrail.
+%In such a case a humanoid robot must plan the contacts it musts create with its feet and hands.
+Computing such a movement is an open motion-planning problem called acyclic contact planning.
 Most of the complexity of this problem lies in the choice of the sequence of contacts, i.e.
 the underlying decomposition of the trajectory into contact phases where specific points of the robot body are exerting forces on specific locations of the environment.
+Tackling this complexity is the main objective of this paper.
 %
@@
 For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
-As a result at the DARPA Robotics Challenge, the participants stated that, except for egress, the robots did not use multi contact strategies: they relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
+As a result at the DARPA Robotics Challenge, the participants stated that the robots did not use multi contact strategies (except for egress): they relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
 
@@
 To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold (i), and the combinatorics of the contact selection problem (ii). 
-To that end, we introduce a reduction property that we name the reachability condition.
+To that end, we introduce a reduction property: the \textit{reachability condition}.
 This condition relaxes the strict contact condition.
-It induces a low-dimensional space, called the \textit{contact reachable} space, that has a Cartesian topology, i.e. it can be efficiently sampled.
+It induces a low-dimensional space, called the \textit{contact reachable} space, that has Cartesian topology, i.e. it can be efficiently sampled.
 Under some hypothesis, a path in the \textit{contact reachable} space can be easily mapped to the (high-dimensionnal) contact manifold. 
-This approach of the problem enables us to separate the contact planning problem in two tractable subproblems.
+This approach enables us to separate the contact planning problem in two tractable subproblems.
 This decoupling presents pros and cons discussed in previous related literature, summarized in the following.
@@
 used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
-at \gls{interactive} frame rates (a method is interactive when the computation time for planning is less than the
-execution time). While this new approach provides an exciting direction of research, currently the resulting motions
+at \gls{interactive} frame rates\footnote{We say that a method is \underline{interactive} when the computation time for planning is less than the
+execution time.}. While this new approach provides an exciting direction of research, currently the resulting motions
  do not seem applicable to real robots.

### 2017-05-03 body/abstract.tex
Tentative complete paper.

@@
 % !TEX root =  ../main.tex
-We present a contact planner for complex legged locomotion tasks: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
+We present a contact planner for complex legged locomotion tasks: standing up, climbing stairs using a handrail, crossing rubble, getting out of a car.
 The need for such a planner was shown at the Darpa Robotics Challenge, where such behaviors
@@
 Current planners suffer from prohibitive algorithmic complexity, because they deploy a tree of robot configurations projected in contact with the environment.
-Contact configurations are costly to compute (because sampled and projected in high dimension) and have no equivalence property (i.e.
+%Contact configurations are costly to compute (because sampled and projected in high dimension) and have no equivalence property (i.e.
 two neighbor configurations cannot be directly connected in the planning graph).
@@
 %
-The reduction breaks the algorithm complexity encountered in other contact planners, logically leading to the open-source implementation of the first interactive multi-contact planner.
+The reduction breaks the algorithm complexity encountered in previous works, logically leading to the  first interactive implementation of a contact planner (open source).
 %
@@
 in a few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ and a dexterous hand. 
-These plans are then validated either in dynamic simulations and on the real HRP-2 robot.
+These plans are validated either in dynamic simulations and on the real HRP-2 robot.
 %, thus demonstrating the first interactive multi-contact planner.

### 2017-05-03 body/conclusion.tex
Tentative complete paper.

@@
 
-As demonstrated in the results section, the main interest of our method is time performance. Time performance is critical for 
-online applications with the robot, a goal not satisfied by any of the previous contributions.
+As demonstrated in the results section, the main interest of our method is reduced algorithmic complexity, leading to interactive implementation. This property is critical for 
+online applications with the robot and was not reached by any of the previous contributions.
 Our method addresses highly constrained environments while improving the search time by orders of magnitude.
@@
 
-The first approximation is the verification of \contactreachability\ ($\mathbf{q}^0 \in C_{Contact}^0$).
+The first approximation occurs when verifying the \contactreachability\ ($\mathbf{q}^0 \in C_{Contact}^0$).
 Our \textit{reachability condition} ($\mathbf{q}^0 \in C_{Reach}^0$) is computationally efficient and provides an accurate approximation of $C_{Contact}^0$ (Section~\ref{sec:scaling}).
@@
 These results clearly justify our pragmatic approach.
+Moreover, refining the reachibility condition to makes it more accurate while keeping the separated approach seems a fruitful research direction.
 
 \subsection{Addressing $\mathcal{P}_3$}
-This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation,
-. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
+This paper focuses on contact planning of multi-contact locomotion. Following the spirit of previous works~\cite{Bretl:2006:MPM:1124573.1124585,DBLP:conf/iser/EscandeKMG08,7140082},
+it does not address directly the continuous motion generation. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
 for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
@@
  the second problem $\mathcal{P}_2$ is the computation of a discrete sequence of whole-body configurations along the root path.
-
+%
 Our contribution to \Pa is the introduction of a low-dimensional space $C_{reach}^0$, an approximation of the space of \gls{equilibrium feasible} root configurations.
 $C_{reach}^0$ can be efficiently sampled and has a low-dimension. For these reasons we are able to solve \Pa much faster than previous approaches.
-
+%
 %~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
@@
 
-Our results demonstrate that our method allows a pragmatic compromise between three 
+Our results demonstrate that our method allows a pragmatic trade-off between three 
 criteria that are hard to reconcile: generality, performance, and quality of the solution, making it the first acyclic contact

### 2017-05-03 body/contact.tex
Tentative complete paper.

@@
 
-First, the root path $\mathbf{q}^0(t)$ is discretized into a sequence of $j$ key configurations:  
+An initial discretization of the root path $\mathbf{q}^0(t)$ into a sequence of $j$ key configurations is first chosen:  
 \begin{equation*}
-	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
+	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; ...; \mathbf{q}^0_{i}; ...; \mathbf{q}^0_{j-1}]
 \end{equation*} 
 where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
-Each root configuration of $\mathbf{Q}^0$ is then extended into a whole-body configuration such that:
+Each root configuration $\mathbf{q}^0_i$ of $\mathbf{Q}^0$ is then extended into a whole-body configuration, while possibly additional discrete steps are added, such that:
 \begin{itemize} 
@@
 Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
-If two or more contacts can't be maintained between two consecutive configurations, one or more intermediate configurations are added, to ensure
+If two or more contacts cannot be maintained between two consecutive configurations, one or more intermediate configurations are added, to ensure
 that at most one contact is broken between two sequential configurations.
@@
 
-To generate contacts, we proceed as follows.
-The contact generator treats each limb $R^k$ independently.
-Given a configuration of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
-\begin{equation}
-\label{eq:contact_gen}
-	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \textrm{ \textbf{and}}\ \mathbf{q}^k \in  C_{Contact}^k 
-\end{equation}
+Given a configuration of the root and the list of effectors that should be in contact, the contact generator computes the configuration of the limbs such that contacts are properly satisfied and the robot is in static equilibrium.
+In previous works~\cite{DBLP:conf/iser/EscandeKMG08,Bouyarmane2009}, the generation of contact is typically implemented by randomly sampling limbs configurations and projecting the whole robot configuration onto the closest surfaces with an inverse kinematics solver.
+In case of failure of the projection, the process would randomly iterate.
 
-Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
-This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.
-We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
-
+We propose two modifications of this general algorithm principle.
+First our contact generator handle each limb $R^k$ independently.
+By handling each limb separatly, we reduce the complexity of the generation of contact configurations.
+This is made possible thanks to the reachibility condition in $\mathcal{P}_1$ that produces a root path that we can afford not to modify in $\mathcal{P}_2$, and because we allow both a contact break and a contact creation between two consecutive configurations of the contact sequence.
+%
+%% Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that limb $R^k$ is in contact and the robot $R$ is in static equilibrium:
+%% \begin{equation}
+%% \label{eq:contact_gen}
+%% 	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k,  s.t.\ \mathbf{q}^k \in  C_{Contact}^k  \textrm{ \textbf{and}}\  (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil}
+%% \end{equation}
+%% The limb configuration $\mathbf{q}^{\overline{k}}$ also defines the location of the contact points.
+%
+%% In previous work, the generation of contact configuration is typically achieved by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver, and iterating until
+%
+%% The contact generator is implemented by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
+%% This process is repeated until either a solution for \eqref{eq:contact_gen} is found, or the generator fails, according to a user-defined maximum number of trials.
+%% We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
+%
 %~ Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
@@
 %~ \end{enumerate}
-
+%
 %~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
-
+Second, we rely on off-line generation of configuration candidates.
 We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum distance 
@@
 \begin{enumerate}
-\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$ (We choose $N=10^3$);
+\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$ (we choose $N=10^3$);
 \item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
@@
 \item Use a user-defined heuristic $h$ to sort $S$;
-\item If $S$ is empty, stop (failure). Else select the first configuration of $S$. Project it onto contact using inverse kinematics. (Figure~\ref{fig:contact_gen} (d) and (e));
-\item If Eq.~\ref{eq:contact_gen} is verified, stop (success). Otherwise remove the element from $S$ and go to step 5.
+\item If $S$ is empty, stop (failure). Else select the first configuration of $S$. Project it onto contact using inverse kinematics (Figure~\ref{fig:contact_gen} (d) and (e));
+\item If projections succeed and equilibrium is verified, stop (success). Otherwise remove the element from $S$ and go to step 5.
 \end{enumerate}
@@
 %~ $N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
-The reader is invited to refer to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
+The reader is refered to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
 of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.

### 2017-05-03 body/intro.tex
Tentative complete paper.

@@
 (ii) Second, the foliated topology turns the exploration by spreading a graph of configurations (probability roadmap, rapidely exploring random trees) into an unefficient random process, where many useless nodes are sampled on parallel submanifolds.
-The total algorithm complexity of classical contact planners comes both from the number of nodes of the graph that are sampled during the exploration (ii) and to the cost of the projection when sampling a new configuration (i).
+The total algorithmic complexity of classical contact planners comes from both the number of graph nodes sampled during exploration (ii) and the cost of the projection when sampling new configurations (i).
 %
@@
 Our work aims at breaking the complexity of the acyclic contact planning problem.  
-To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold (i), and the combinatorics of the contact selection problem (ii). 
+We deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold (i), and the combinatorics of the contact selection problem (ii). 
 To that end, we introduce a reduction property: the \textit{reachability condition}.
@@
 Under some hypothesis, a path in the \textit{contact reachable} space can be easily mapped to the (high-dimensionnal) contact manifold. 
-This approach enables us to separate the contact planning problem in two tractable subproblems.
+This approach enables us to separate the contact planning in two tractable subproblems.
 This decoupling presents pros and cons discussed in previous related literature, summarized in the following.
@@
 %~ In the remainder of this introduction, we discuss further the current state of the art. This allows us to situate more precisely our contributions.
+
+\subsection{State of the art}
+
+Additionally to robotics, acyclic contact planning is a problem of interest in neurosciences, biomechanics and computer animation.
+Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163} or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}.
+These approaches are by definition not able to discover complex behaviors in unforeseen contexts.
+
+The issue of planning acyclic contacts was first completely described by Bretl~\cite{Bretl:2006:MPM:1124573.1124585}.
+It requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path.
+A third nontrivial problem, $\mathcal{P}_3$,
+%~ not addressed in this work, 
+then consists in interpolating a complete motion between two configurations of the contact sequence.
+A key issue is to avoid combinatorial explosion when considering at the same time the possible contacts and the potential paths.
+Bretl's seminal paper proposes a first effective algorithm, able to handle simple situations, but not applicable to arbitrary environments.
+Following it, seve\-ral papers have applied this approach in specific situations, limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.
+
+Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorics.
+Two main directions have been explored.
+\textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}.
+A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
+To get reasonable computation times, the method uses a simplified dynamic model for the avatar.
+Still, the method is far from real-time  (about 1 minute of computation for 20 contacts).
+A similar approach has been considered for manipulation by \cite{gabicciniisrr15}.
+Deits and Tedrake proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered, and equilibrium is not considered~\cite{DBLP:conf/humanoids/DeitsT14}.
+Dai et al.~\cite{dai2014whole} extended the work of Posa et al.~\cite{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
+the contacts manually for complex interactions.
+Model predictive control on short horizon of the whole contact dynamcis~\cite{tassa_icra14,hamalainen_cpbp_2015} leads to dynamically-consistent motions at \gls{interactive} frame rates\footnote{We say that a method is \underline{interactive} when the computation time for planning is less than the execution time.}, although with very few look-ahead capabilities and yet not realistically applicable to physical robots.
+In addition to the limits of the current implementations, optimization-based approaches only converge locally.
+
+\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity.
+The feasibility and interest of the decoupling has been shown by Escande et al.
+\cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
+an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
+A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning},
+but at the cost of prohibitive computation times (up to several hours) for constraining scenarios.
+This approach is suboptimal because the quality of the motion depends on the quality of the guide path.
+Bouyarmane et al.
+~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
+Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
+Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path.
+Chung and Khatib~\cite{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment.
+This planning phase does not account for collisions, implying that replanning is required in case of failure.
+In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}.
+
+%% For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
+%% used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
+%% at \gls{interactive} frame rates\footnote{We say that a method is \underline{interactive} when the computation time for planning is less than the
+%% execution time.}. While this new approach provides an exciting direction of research, currently the resulting motions
+%%  do not seem applicable to real robots.
+
+As far as robotics applications are concerned, none of the existing planners is \gls{interactive}.
+However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
+In particular, our algorithm proposed in \cite{Carpentier2016} is \gls{interactive}.
+Therefore, a planner capable of efficiently solving \Pa and \Pb could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
+The main contribution of this paper is exactly this planner.
 
@@
 
-\subsection{State of the art}
-
-Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences, biomechanics, and virtual character animation.
-Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}.
-These approaches are by definition not able to discover complex behaviors in unforeseen contexts.
-
-The issue of planning acyclic contacts was first completely described by Bretl~\cite{Bretl:2006:MPM:1124573.1124585}.
-The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path.
-A third nontrivial problem, $\mathcal{P}_3$,
-%~ not addressed in this work, 
-then consists in interpolating a complete motion between two postures of the contact sequence.
-A key issue is to avoid combinatorial explosion when considering at the same time the possible contacts and the potential paths.
-Bretl's seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments.
-Following it, seve\-ral papers have applied this approach in specific situations, limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.
-
-Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorics.
-Two main directions have been explored.
-\textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}.
-A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.
-To get reasonable computation times, the method uses a simplified dynamic model for the avatar.
-Still, the method is far from real-time  (about 1 minute of computation for 20 contacts).
-A similar approach has been considered for manipulation by \cite{gabicciniisrr15}.
-Deits and Tedrake proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered, and equilibrium is not considered~\cite{DBLP:conf/humanoids/DeitsT14}.
-Dai et al.~\cite{dai2014whole} extended the work of Posa et al.~\cite{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
-the contacts manually for complex interactions.
-In addition to the limits of the current implementations, optimization-based approaches only converge locally.
-
-\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity.
-The feasibility and interest of the decoupling has been shown by Escande et al.
-\cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e.
-an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
-A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning},
-but at the cost of prohibitive computation times (up to several hours) for constraining scenarios.
-This approach is suboptimal because the quality of the motion depends on the quality of the guide path.
-Bouyarmane et al.
-~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium.
-Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented).
-Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path.
-Chung and Khatib~\cite{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment.
-This planning phase does not account for collisions, implying that replanning is required in case of failure.
-In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}.
-
-For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
-used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
-at \gls{interactive} frame rates\footnote{We say that a method is \underline{interactive} when the computation time for planning is less than the
-execution time.}. While this new approach provides an exciting direction of research, currently the resulting motions
- do not seem applicable to real robots.
-
-As far as robotics applications are concerned, none of the existing planners is \gls{interactive}.
-However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
-In particular, our algorithm proposed in \cite{Carpentier2016} is \gls{interactive}.
-Therefore, a planner capable of efficiently solving \Pa and \Pb could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
-The main contribution of this paper is exactly this planner.
-
 \subsection{Contributions}
 
-Our solution belongs to the class of decoupled approaches, because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning.
+Our solution belongs to the class of decoupled approaches, i.e.
+we propose specific algorithms to efficiently solve both $\mathcal{P}_1$ and $\mathcal{P}_2$, while relying on state-of-the-art solution to $\mathcal{P}_3$ to obtain the whole movement.
+Our main contribution is the definition of a reduction property, the \textit{reachibility condition}.%,  that leads to a large decrease of the overall algorithmic complexity.
 Compared to previous approaches, our solution has two main novelties:
 
-\noindent \textbf{Regarding $\mathcal{P}_1$}.
-We propose a fast guide path planning algorithm.
+\noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm.
 The key to its efficiency is that it does not sample directly the contact manifold, but an approximation of the \textit{contact reachable} space.
-The \textit{contact reachable} space is a low-dimensional space for which there exists a mapping to the contact manifold.
+The \textit{contact reachable} space is a low-dimensional space and can be explored using standard sample-based planning.
+Under some conditions, a mapping exists from a trajectory in the \textit{contact reachable} space to the contact manifold.
 
 \noindent \textbf{Regarding $\mathcal{P}_2$}, we propose a fast method to extend a \textit{contact reachable} path into a sequence of whole-body configurations in static equilibrium. 
-This  requires the explicit computation of contact configurations.
-It is guided by dedicated heuristics that quickly synthesize feasible configurations.
+Thanks to the reachibility condition, the sequence can be extended exactly on the trajectory computed in $\mathcal{P}_1$ (and not along it, like in previous works~\cite{DBLP:conf/iser/EscandeKMG08,Bouyarmane2009}), which makes $\mathcal{P}_2$ a simple 1-D algorithm.
+It requires the explicit computation of contact configurations, guided by dedicated heuristics that quickly synthesize feasible configurations.
 
-This sequential approach is the key to the efficiency of our method, though it
-can result in failures.
-However, we demonstrate empirically its interest: the high success rate and low computation times allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
+The reachibility condition is the key to the strict separation between $\mathcal{P}_1$ and $\mathcal{P}_2$, hence to the low complexity of our planner.
+It  may also result in failure (i.e. some guide paths $\mathcal{P}_1$ are not extendable in a full sequence $\mathcal{P}_2$).
+However, we empirically show by a large number of trials on various scenarios of practical interest that the rate of success is very high, while the algorithm efficiency allows the planner to iterate in case of failure.
+All scenarios are validated by planning a whole dynamic trajectory using a state-of-the-art implementation of $\mathcal{P}_3$ which is executed either in simulation or on the real robot.
 
-To further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
+The paper extends our previous papers presenting the reachibility condition~\cite{tonneauisrr15} and the connection of our contact planner with a solver~$\mathcal{P}_3$~\cite{Carpentier2016}.
+In this paper, we introduce a new efficient criterion of the robot balance, provide the complete pseudo-code an release our source code that can be used to directly reproduce our results.
+We also provide a solution to $\mathcal{P}_3$.
+The planner is validated on robot models of HyQ and HRP-2 (while \cite{tonneauisrr15} used unrealistic avatar models) with complete performance analysis.
+
+%This sequential approach is the key to the efficiency of our method, though it can result in failures. 
+%
+%However, we demonstrate empirically its interest: the high success rate and low computation times allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
+
+%% To further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
 For HRP-2, we detail the complete computation times to address sequentially the three problems, and compare them to related work, demonstrating that our method is orders of magnitude faster.
 
-Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem.
+%% Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem.
 
-This paper extends our previous conference paper~\citep{tonneauisrr15} in several ways.
+%% This paper extends our previous conference paper~\citep{tonneauisrr15} in several ways.
 Our approach is now demonstrated on actual robots and in simulation, contrary to virtual avatars before, in a large variety of contexts.

### 2017-05-03 body/overview.tex
Tentative complete paper.

@@
 Figure~\ref{fig:framework} illustrates our workflow.
-\Pa and \Pb are addressed sequentially: From a given problem (a) we first plan a root guide path (b), before
-extending it into a sequence of static equilibrium configurations(c). In the case where step c fails,
-our framework invalidates the computed guide path, and restarts the planning from b.
+\Pa and \Pb are addressed sequentially: from a given problem (a) we first plan a root guide path (b), before
+extending it into a sequence of static equilibrium configurations (c). In the case where step (c) fails,
+our framework invalidates the computed guide path, and restarts the planning from (b).
 

### 2017-05-03 body/rbprm.tex
Tentative complete paper.

@@
 %~ \section{Computing a guide path in $C^0_{reach}$ ($\mathcal{P}_1$) }
-\section{Root path planning in the contact reachable space}
+\section{Root path planning in the contact reachable space ($\mathcal{P}_1$)}
 \label{rbprm}
@@
 as well as the environment $O$.
-
+%
 Given a start and a goal configurations, we aim at computing a guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ $\mathbb{R}^r$ verifying:
@@
 $C_{Equil}^0$  cannot be described analytically.
-
+%
 The strong hypothesis of this work is that for a large variety of locomotion tasks, we can define a space  $C_{Reach}^0 \simeq C_{Contact}^0$, such that 
@@
 
-\textbf{necessary conditions:}
+\noindent\textbf{Necessary conditions:}
 For a contact to be possible, an obstacle $O_i \subset O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot (Figure~\ref{fig:contact_gen}--1).
@@
 
-\textbf{sufficient condition:}
+\noindent\textbf{Sufficient condition:}
 Similarly we can define an inner approximation \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ $C_{Contact}^0$} by considering a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces.
@@
 Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.
-
+%
 %~ However, using a shape between $W^0$ and $B^{\textrm{\it Suf}}$ leads to a trade-off between a necessary and a sufficient condition. 
@@
 By increasing $s$, the condition can become sufficient, but less and less necessary.  
-Eq.~\ref{eq:reachc} thus defines the \textit{reachability condition}. We fix a value $s^*$ for $s$ and define  $C_{Reach}^0 = C^0_{s^*}$.
+Eq.~\eqref{eq:reachc} thus defines the \textit{reachability condition}. We fix a value $s^*$ for $s$ and define  $C_{Reach}^0 = C^0_{s^*}$.
 The computation of $s^*$ is detailed in Section~\ref{sec:params}. 
@@
 \subsection{Computing the guide path in $C_{reach}^0$}
-$C_{Reach}^0$ can be sampled efficiently thanks to Eq.~\ref{eq:reachc}, and can thus be used with a standard motion planner.
+$C_{Reach}^0$ can be sampled efficiently thanks to \eqref{eq:reachc}, and can thus be used with a standard motion planner.
 The only significant change is to replace the collision checking with the \textit{reachability condition}.
@@
 
-This Section has presented a guide path planner for the geometric root of a robot, implemented as a low-dimensional sampling-based 
-algorithm. Given start and goal configurations, it outputs a continuous path for the robot's root. 
+%This Section has presented a guide path planner for the geometric root of a robot, implemented as a low-dimensional sampling-based 
+%algorithm. Given start and goal configurations, it outputs a continuous path for the root of the robot.
+
+The result of this section is a guide path for the root of the robot (problem $\mathcal{P}_1$), computed with simple modification over a bi-RRT planner using the reachability condition.
+We now explain how this path of the root can be extended to a sequence of whole-body contact configurations.
+
 %~ \deladp{Thanks to the \textit{reachability condition}, we assume that for any configuration in the path there exists a joint configuration that results in static equilibrium.}

### 2017-05-03 body/results.tex
Tentative complete paper.

@@
 Specifically, we demonstrate the planner for two legged robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
+For all scenarios, the video displays a valid contact sequence outputs by our planner (i.e. $\mathcal{P}_2$ result).
+They are then validated by exhibiting a dynamically-consistant solution to the interpolation problem $\mathcal{P}_3$, which is displayed at the video end.
 Finally, a last example suggests possible applications to dexterous manipulation.
+These scenarios complement our previous results with virtual avatars (Figure~\ref{fig:robots_old})~\citep{tonneauisrr15}.
+%
+%At the end of the video, we validate our contact plans by exhibiting a solution to the interpolation problem $\mathcal{P}_3$.
 
-At the end of the video, we validate our contact plans by exhibiting a solution to the interpolation problem $\mathcal{P}_3$.
+In this section, we first reports the scenarios that were solved by our planner (Sec.~\ref{sec:scenarios}) 
+before discussing  the role of the parameters of our planner (Sec.~\ref{sec:influence}) and its performances (\textit{interactive} computation times, success rates -- Sec.~\ref{sec:perf}). 
+%We then provide the \textit{interactive} computation times obtained in each case.
+%We also compare the times obtained with HRP-2 with respect to previous works.
 
-At the end of this section, we discuss the role of the parameters of our framework. We then provide the \textit{interactive} computation times obtained in each case.
-We also compare the times obtained with HRP-2 with respect to previous works.
-
-These scenarios complement the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
 %~ (Figure~\ref{fig:robots_old}).
@@
 	\end{overpic}
-\caption{Virtual avatars in various scenarios demonstrated in our conference paper.}
+\caption{Virtual avatars in various scenarios demonstrated in our conference paper~\cite{tonneauisrr15} (see video on YouTube \url{http://youtu.be/LmLAHgGQJGA}).}
 		   \label{fig:robots_old}
@@
 
-\subsection{Description of the scenarios}
+\subsection{Description of the scenarios}\label{sec:scenarios}
 In all the scenarios considered, the formulation of the problem is always the same:
@@
 contact generation (Appendix~\ref{sec:heuristics}), and lastly, the discretization step for the guide path. The appropriate value for these parameters
-is computed empirically based on use-case analysis or trials and errors.
+is computed empirically based on use-case analysis or trials and errors (the same value is used in all the scenarios).
 
@@
 As it can be expected, the scaling results in a high increase of the sensitivity, with a decrease of the specificity.
-For HRP-2 we decided to set $s^*=1.2$.
+For HRP-2 and HyQ we decided to set $s^*=1.2$.
 %~ , and was also effective for the car egress scenario.
@@
 In locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
-configurations close to singularities (and joint limits). While this did not significantly impact
+configurations close to singularities and joint limits. While this did not significantly impact
 the generation of the plan, the resulting interpolation turned out to be harder.
@@
 for the car egress scenario to handle the geometry of the polaris.
+This parameter is similar to the step duration used e.g. in~\cite{Mordatch:2012:DCB:2185520.2185539}.
 For future work, we would like to automatically adapt the size of the discretization step to the complexity of the environment.
@@
 To analyze performance, we ran the planner 1000 times for each scenario.
-We measured the computation time spent in each part of the algorithm, and analyzed their success rate.
+We measured the computation time spent in each part of the algorithm and analyzed their success rate.
 
@@
  \hline
-   Stairs     	& 100\% \\
-   Standing			& 68\% 	\\
-   Car 			& 77\% 	\\
+   Steep stairs     	& 100\% \\
+   Standing up			& 68\% 	\\
+   Car egress			& 77\% 	\\
    Rubble 				& 97\% 	\\
-   Race        & 88.0\% 	\\
+   Obstacle race        & 88\% 	\\
  \end{tabular}
-\caption{ Percentage of successful complete contact planning rates for each scenario, rounded to the first decimal.}
+\caption{ Rates of successful complete contact planning for each scenario.}
 \label{tab:sucess_planning}
@@
 \begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{35pt} | >{\centering\arraybackslash}m{35pt} | c}
-  &  Equilibrium success rate & Kinematic failure & Equilibrium failure \\
+  &  Success rate of the contact generator & Kinematic failure & Equilibrium failure \\
  \hline
@@
  \end{tabular}
-\caption{Success rates obtained for the generation of static equilibrium contact configurations for each scenario, rounded to the first decimal. Column 1 indicates 
-indicates the rate of contact generation that suceeeded. In the cases where the generation fails, it can be
-either a kinematic issue (column 2), or because no contact configuration led to a static equilibrium configuration (column 3). Note that a failure in the contact generation
-is not equivalent to a failure of the contact planning algorithm.}
+\caption{Success rate of the contact generator , whose failures may be due to kinematic issue or equilibrium issue. 
+%% Success rates obtained for the generation of static equilibrium contact configurations for each scenario, rounded to the first decimal. Column 1 indicates 
+%% indicates the rate of contact generation that suceeeded. In the cases where the generation fails, it can be
+%% either a kinematic issue (column 2), or because no contact configuration led to a static equilibrium configuration (column 3). 
+%Note that a failure in the contact generation is not equivalent to a failure of the contact planning algorithm.
+}
 \label{tab:requestpercent}
@@
 %~ Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
-Despite the complexity of the scenarios and the approximations made in our formulation, our planner succeeded in the large majority of cases.
+Despite the complexity of the scenarios, our planner succeeded in the large majority of cases.
 
-Table~\ref{tab:requestpercent} presents the rate of successful contact generation.
-Note that a failure in contact generation for a root configuration is not equivalent to a failure in the contact plan.
-It simply means that another limb was tested for contact generation for the same root configuration.
+It is also interesting to observe the causes of failure of the contact generator (Table~\ref{tab:requestpercent}).
+As it iterates until sucess, one failure of the contact generator is not equivalent to a failure in the contact plan (but corresponds to increased computation time).
 As expected, the more constrained scenario, the car egress, provides the less satisfying results, despite the high success rate of the planner.
 
-These results confirm that our approach provides a satisfying compromise between completeness and efficiency, thus allowing to consider online planning
-while controlling the robot. Indeed, when the contact planning fails, it fails rapidly. This allows us to rapidly re-plan with a reasonable chance of success.
-The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible would be to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.
+These results correspond to an empirical validation of the \textit{reachibility condition}.
+As previously explained, this condition is not sufficient to guarantee that a path $\mathcal{P}_1$ is expendable in a sequence~$\mathcal{P}_2$.
+However we see in practice that, even in complex scenario, this condition is a fair approximation.
+Moreover, when the contact planning fails, it fails rapidly. 
+It is then immediate to simply invalidate the corresponding edge of the planning graph and to rapidly re-plan it with a reasonable chance of success.
+
+%% These results confirm that our approach provides a satisfying trade-off between completeness and efficiency, thus allowing to consider online planning
+%% while controlling the robot. Indeed, when the contact planning fails, it fails rapidly. This allows us to rapidly re-plan with a reasonable chance of success.
+%% The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible would be to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.
 
@@
 
-\subsection{Validation of the contact plans, and comparison with previous work} \label{sec:compa}
-To validate our plans, we either use the framework of \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
+\subsection{Validation of the contact plans}
+The contact sequences computed by our planner are finally used to generate a whole dynamics trajectory using~$\mathcal{P}_3$.
+For this last stage, we relied on existing formulation, either directly using existing solvers~\cite{Carpentier2016} or implementing our own version a $\mathcal{P}_3$ solver, described in Appendix~\ref{app:optim}.
+Because $\mathcal{P}_3$ is not a contribution of this paper, we do not provide here computation scores (see~\cite{Carpentier2016} for details).
+The resulting whole trajectories are dynamically consistant: they are smooth; contact forces corresponds to the robot motion; joint motion and torques are within the acceptable ranges specified by the robot manufacturers.
+All the resulting movements have been executed in a complete dynamic simulator.
+We executed one of them using the real HRP-2 robot
+%
+%To validate our plans, we either use the framework of \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
 The companion video shows the obtained motions.
 
-Few contributions provided computation times on the complete problem. %, necessary to provide comparison between our approaches. 
+\subsection{Comparison with previous work} \label{sec:compa}
+
+We did our best to provide a fair comparison of the computation complexity of our method with the state of the art. 
+%Few contributions provided computation times on the complete problem. 
+However existing benchmarks for motion planning algorithms~\cite{moll2014extensible} do not yet encompass contact planning.
+Moreover, the source code of the previous methods of the state of the art is often not available.
+Providing a fair comparison with the algorithms performing on the same computer and on the same scenarios is yet out of reach.
+A step in this direction is the open-source release of our source code (see Appendix~\ref{app:hpp}) that allows any reader to reproduce our results.
+%
 Furthermore, $\mathcal{P}_3$ remains an open problem in the presence of obstacles. 
-The only valid scenarios addressed completely in previous works are thus the stair-climbing scenarios of different heights proposed by Hauser in \cite{Hauser06usingmotion}, and the table-egress scenario by Escande et al.
-in~\cite{DBLP:conf/iser/EscandeKMG08}, which we consider to be of similar complexity with respect to the car-egress scenario (we did not consider the stairs in the scene).
-Both scenarios are tested with HRP-2.
+The only valid scenarios completely addressed  in previous works are thus the stair-climbing scenarios of different heights  in \cite{Hauser06usingmotion}, and the table-egress scenario in~\cite{DBLP:conf/iser/EscandeKMG08}(whose  complexity is similar to our car-egress scenario).
+Both our scenarios are tested with HRP-2.
 
-Table~\ref{tab:compprev} presents the computation times for these scenarios, clearly demonstrating that our approach is order of magnitude faster than previous works.
+Table~\ref{tab:compprev} presents the computation times for these scenarios.
+The scores clearly demonstrate that our approach is order of magnitude faster than previous works.
 
-To provide a completely fair comparison, it would be required to compare the methods on the same computers and exact same scenarios, which is not possible
-because the source code of the other methods is often not available. Despite this observation and the fact that some results are old, we believe that the results show that our method
-performs better. Indeed, a difference in technology does not justify a difference as important as the one demonstrated in Table~\ref{tab:sucess_planning}.
+ %~ Despite this observation and the fact that some results are old, we believe that the results show that our method
+%~ performs better. Indeed, a difference in technology does not justify a difference as important as the one demonstrated in Table~\ref{tab:sucess_planning}.

### 2017-05-09 body/abstract.tex
abstract

--- 2017-04-20 body/abstract.tex
@@
 
-Current planners suffer from prohibitive computation times, due to the difficulty
-of generating contact postures in a high-dimensional solution space that is impossible to sample.
+Current planners suffer from their prohibitive algorithmic complexity, because they deploy a tree of robot
+configurations projected in contact with the environment.
+%~ computation times, due to the difficulty
+%~ of generating contact postures in a high-dimensional solution space that is impossible to sample.
 
-We tackle this issue by introducing the reachability condition. This condition allows us to define
-a geometric approximation of the contact manifold, which is of low dimension and can be sampled efficiently.
-Informally, the condition verifies that the root configuration of a robot is close, but not too close to obstacles: close to allow contact creation, not too close to avoid collision.
-Thanks to this condition we decompose the hard contact planning problem into two
-sub-problems: first, planning a guide path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm.
+We tackle this issue by introducing a reduction property: the reachability condition. This condition defines
+a geometric approximation of the contact manifold, which is of low dimension, presents a Cartesian topology, and can be efficiently sampled and explored.
+%~ Informally, the condition verifies that the root configuration of a robot is close, but not too close to obstacles: close to allow contact creation, not too close to avoid collision.
+Thanks to this condition we decompose
+The hard contact planning problem can then be decomposed into two
+%~ sub-problems: first, planning a guide path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm.
+sub-problems: first, we plan a path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, we generate a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm.
 
-Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically: in a
-few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ, and a dexterous hand.
+The reduction breaks the algorithm complexity encountered
+in previous works, resulting in the first interactive
+implementation of a contact planner (open source). While
+no contact planner has yet been proposed with theoretical
+completeness, we empirically show the interest of our framework:
+in a
+few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ and a dexterous hand.
+These plans are validated either in dynamic simulations, or on the real HRP-2 robot.
+
+%~ Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically: in a
+%~ few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ, and a dexterous hand.
 These plans are then validated either in dynamic simulations, or on the real HRP-2 robot, thus demonstrating
-the first interactive multi-contact planner.
+%~ the first interactive multi-contact planner.

### 2017-05-09 body/intro.tex
nicolas modifs

--- 2017-05-02 body/intro.tex
@@
 \IEEEPARstart{L}{egged} robots move by sequentially creating contacts with the environment.
-After years of research, such robots can autonomously walk on flat ground, but struggle to navigate in more complex environments, e.g. to climb stairs using a handrail.
-In such a case a humanoid robot must plan the contacts it musts create with its feet and hands. Computing such a movement is an open motion-planning problem called acyclic contact planning.
-Most of the complexity of this problem lies in the choice of the sequence of contacts, i.e.
-the underlying decomposition of the trajectory into contact phases where specific points of the robot body are exerting forces on specific locations of the environment.
-%
+After years of research, such robots can autonomously walk on flat ground, but struggle to navigate more complex environments.
+%~ where
+Deciding
+where
+to
+create a 
+contact with its feet and
+possibly
+its
+hands
+is
+nontrivial,
+e.g. to climb stairs using a
+handrail.
+%~ To climb stairs using a handrail, a humanoid robot must plan the contacts to create with its feet and hands. 
+%~ Computing such motion is an open motion planning problem called acyclic contact planning.
+
+Most of the complexity of this problem lies in the contact planning, i.e. the underlying decomposition
+of the trajectory into contact phases where specific points
+of
+the
+robot body are exerting forces on specific locations
+of
+the
+environment. Tackling this complexity is the main
+objective
+of this paper.
+
 Once the contact plan is known, efficient approaches exist to generate a dynamically feasible motion~\citep{Carpentier2016}.
+ %~ However in general
+%~ the contact plan is manually designed by a time consuming trial-and-error procedure.
+Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in cluttered, human-centered environments, which comprise constraining obstacles (stairs...).
+In the specific (and simple) case of gaited biped locomotion on flat ground, 
+%~ efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location. Furthermore, 
+choosing the effector with which to create a contact is trivial because walking follows a cyclic pattern (the left foot always follows the right foot).
+Efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
+In the general case planning complex contact interactions is extremely challenging:
+At any given time a contact must be chosen between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location).
+Furthermore, a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles; we say that the contact manifold is foliated~\citep{simeon-manipulation-04}.
+%~ Lastly, the contact manifold has a null measure and is thus impossible to sample. This restriction prevents the use of efficient sampling-based planners.
+The
+foliation prevents the use of efficient sampling-based
+planners
+for two reasons. (i) First, each submanifold of the         
+foliation
+has
+a zero measure and cannot be directly sampled.         
+A
+sample
+is
+rather obtained by sampling a “free flying”         
+configuration
+and
+explicitly projecting it in contact (which is a         
+costly
+numerical
+operation).
+(ii) Second, the foliated topology       
+turns
+the
+exploration
+by
+spreading
+a graph of configurations        
+(probabilistic
+roadmap,
+rapidely
+exploring
+random trees) into      
+an
+unefficient
+random
+process,
+where
+many
+useless nodes       
+are
+sampled
+on
+parallel
+submanifolds.
+The
+total
+algorithmic       
+complexity
+of
+classical
+contact
+planners
+comes
+from
+both the        
+number
+of
+graph
+nodes
+sampled
+during
+exploration
+(ii) and
+the cost of the projection when sampling new configurations
+(i).
 
-%% However in general
-%% the contact plan is manually designed by a time consuming trial-and-error procedure.
-Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in cluttered, human-centered environments, which comprise constraining obstacles (stairs...).
+For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
+As a result at the DARPA Robotics Challenge, the participants stated that except for egress, the robots did not use multi contact strategies: they relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
 
-In the specific (and simple) case of gaited biped locomotion on flat ground, choosing the effector with which to create a contact is trivial because walking follows a cyclic pattern (left foot always follows right foot).
-Efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location. 
-%
-The general case of planning contact sequences is more challenging.
-At any given time a contact results of a combinatorial discrete choice for both the effector and contact surface, and a continuous choice for the contact location.
-Any choice of contact then kinematically and dynamically constrains the possible motion of the robot.
-Furthermore there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles.
-From this property, we say that the contact manifold is foliated~\citep{simeon-manipulation-04}. 
-The foliation prevents the use of efficient sampling-based planners for two reasons.
-(i) First, each submanifold of the foliation has a zero measure and cannot be directly sample.
-A sample is rather obtained by sampling a ``free flying'' configuration and explicitly projecting it in contact (which is a costly numerical operation).
-(ii) Second, the foliated topology turns the exploration by spreading a graph of configurations (probability roadmap, rapidely exploring random trees) into an unefficient random process, where many useless nodes are sampled on parallel submanifolds.
-The total algorithm complexity of classical contact planners comes both from the number of nodes of the graph that are sampled during the exploration (ii) and to the cost of the projection when sampling a new configuration (i).
-%
-For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}.
-As a result at the DARPA Robotics Challenge, the participants stated that, except for egress, the robots did not use multi contact strategies: they relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.
-
-%Obtaining reasonable computation times is critical for a contact planner to be of any practical use. 
+%~ Obtaining reasonable computation times is critical for a contact planner to be of any practical use.
  Our work aims at breaking the complexity of the acyclic contact planning problem.
-To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold (i), and the combinatorics of the contact selection problem (ii). 
-To that end, we introduce a reduction property that we name the reachability condition.
-This condition relaxes the strict contact condition.
-It induces a low-dimensional space, called the \textit{contact reachable} space, that has a Cartesian topology, i.e. it can be efficiently sampled.
-Under some hypothesis, a path in the \textit{contact reachable} space can be easily mapped to the (high-dimensionnal) contact manifold. 
-This approach of the problem enables us to separate the contact planning problem in two tractable subproblems.
+To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorics of the contact selection problem.
+First we introduce a low-dimensional space, called the \textit{contact reachable} space, that can be sampled and mapped efficiently to the contact manifold.
+Then, given a path computed in the \textit{contact reachable} space, we propose a deterministic algorithm to generate a contact sequence along the path.
 This decoupling presents pros and cons discussed in previous related literature, summarized in the following.
@@
 
-For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
-used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
-at \gls{interactive} frame rates (a method is interactive when the computation time for planning is less than the
-execution time). While this new approach provides an exciting direction of research, currently the resulting motions
- do not seem applicable to real robots.
+%~ For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
+%~ used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
+%~ at \gls{interactive} frame rates (a method is interactive when the computation time for planning is less than the
+%~ execution time). While this new approach provides an exciting direction of research, currently the resulting motions
+ %~ do not seem applicable to real robots.
 
@@
 
-Our solution belongs to the class of decoupled approaches, because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning.
+Our solution belongs to the class of decoupled approaches, 
+i.e. 
+we
+propose
+specific
+algorithms
+to
+efficiently
+solve
+both
+\Pa
+and \Pb
+while 
+relying 
+on 
+state-of-the-art 
+solution 
+to 
+$\mathcal{P}_3$
+to 
+obtain
+the
+whole
+movement.
+Our
+main
+contribution
+is
+the
+definition        
+of
+a reduction property, the reachability condition.
+
+%~ because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. 
 Compared to previous approaches, our solution has two main novelties:
@@
 
-This sequential approach is the key to the efficiency of our method, though it
+The reachability condition is the key to the strict separation between $\mathcal{P}_1$ and $\mathcal{P}_2$, hence to the low complexity of our planner, though it
 can result in failures.

### 2017-05-09 body/results.tex
nicolas modifs

--- 2017-04-20 body/results.tex
@@
 
-\subsection{Validation of the contact plans, and comparison with previous work} \label{sec:compa}
+\subsection{Validation of the contact plans} \label{sec:valid}
 To validate our plans, we either use the framework of \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
@@
 
-Few contributions provided computation times on the complete problem. %, necessary to provide comparison between our approaches. 
-Furthermore, $\mathcal{P}_3$ remains an open problem in the presence of obstacles.
+\subsection{Comparison with previous work} \label{sec:compa}
+%~ Few contributions provided computation times on the complete problem.
+To provide a completely fair comparison, it would be required to compare the methods on the same computers and exact same scenarios, which is not possible because
+%~ the source code of the other methods is often not available. %, necessary to provide comparison between our approaches. 
+We did our best to provide a fair comparison of the computation complexity of our method with the state of the art. 
+%Few contributions provided computation times on the complete problem. 
+However existing benchmarks for motion planning algorithms~\cite{moll2014extensible} do not yet encompass contact planning.
+Moreover, the source code of the previous methods of the state of the art is often not available.
+Providing a fair comparison with the algorithms performing on the same computer and on the same scenarios is yet out of reach.
+A step in this direction is the open-source release of our source code (see Appendix~\ref{app:hpp}) that allows any reader to reproduce our results.
+Furthermore, $\mathcal{P}_3$ remains challenging in the presence of obstacles.
 The only valid scenarios addressed completely in previous works are thus the stair-climbing scenarios of different heights proposed by Hauser in \cite{Hauser06usingmotion}, and the table-egress scenario by Escande et al.
@@
 
-To provide a completely fair comparison, it would be required to compare the methods on the same computers and exact same scenarios, which is not possible
-because the source code of the other methods is often not available. Despite this observation and the fact that some results are old, we believe that the results show that our method
-performs better. Indeed, a difference in technology does not justify a difference as important as the one demonstrated in Table~\ref{tab:sucess_planning}.
+ %~ Despite this observation and the fact that some results are old, we believe that the results show that our method
+%~ performs better. Indeed, a difference in technology does not justify a difference as important as the one demonstrated in Table~\ref{tab:sucess_planning}.

### 2017-05-09 body/contact.tex
typos

--- 2017-04-20 body/contact.tex
@@
 
-Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
+Contact generation is typically addressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
 This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.

### 2017-05-09 body/intro.tex
typos

@@
 planners
-for two reasons. (i) First, each submanifold of the         
+for two reasons. (i) First, each sub-manifold of the         
 foliation
@@
 (probabilistic
-roadmap,
-rapidely
+road-map,
+rapidly
 exploring
@@
 an
-unefficient
+inefficient
 random
@@
 parallel
-submanifolds.
+sub-manifolds.
 The
@@
 Chung and Khatib~\cite{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment.
-This planning phase does not account for collisions, implying that replanning is required in case of failure.
+This planning phase does not account for collisions, implying that re-planning is required in case of failure.
 In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}.

### 2017-05-09 body/overview.tex
typos

--- 2017-04-20 body/overview.tex
@@
 
-Figure~\ref{fig:framework} illustrates our workflow.
+Figure~\ref{fig:framework} illustrates our work flow.
 \Pa and \Pb are addressed sequentially: From a given problem (a) we first plan a root guide path (b), before
@@
 This does not mean that the final motion will necessary be quasi-static.}.
-Previous works verifiy \textit{equilibrium feasibility} by explicitly computing such a configuration.
-We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact rechability}, illustrated in the following.
+Previous works verify \textit{equilibrium feasibility} by explicitly computing such a configuration.
+We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact reachability}, illustrated in the following.
 

### 2017-05-09 body/results.tex
typos
car model is not polaris

@@
 
-\subsubsection{HRP-2 -- Polaris car egress (Figure~\ref{fig:car})}
-In this scenario inspired from the DRC car egress HRP-2 has to step out of the Polaris car.
+\subsubsection{HRP-2 -- Car egress (Figure~\ref{fig:car})}
+In this scenario inspired from the DRC car egress HRP-2 has to step out of a car.
 
@@
   \caption{
-           Selected frames from the polaris egress scenario. }
+           Selected frames from the car egress scenario. }
 		   \label{fig:car}
@@
 but may generate unnecessary states. In most scenarios the torso of HRP-2 moves about 15 cm between two postures, but only 3 cm
-for the car egress scenario to handle the geometry of the polaris.
+for the car egress scenario to handle the geometry of the car.
 For future work, we would like to automatically adapt the size of the discretization step to the complexity of the environment.
@@
 \caption{Success rates obtained for the generation of static equilibrium contact configurations for each scenario, rounded to the first decimal. Column 1 indicates 
-indicates the rate of contact generation that suceeeded. In the cases where the generation fails, it can be
+indicates the rate of contact generation that succeeded. In the cases where the generation fails, it can be
 either a kinematic issue (column 2), or because no contact configuration led to a static equilibrium configuration (column 3). Note that a failure in the contact generation

### 2017-05-09 body/conclusion.tex
final version

--- 2017-04-20 body/conclusion.tex
@@
 
-As demonstrated in the results section, the main interest of our method is time performance. Time performance is critical for 
-online applications with the robot, a goal not satisfied by any of the previous contributions.
+As demonstrated in the results section, the main interest of our method is the reduction of the algorithmic complexity of the problem, which leads to an interactive 
+application. This property is critical for 
+online applications with the robot and was not proposed by any of the previous contributions.
 Our method addresses highly constrained environments while improving the search time by orders of magnitude.

### 2017-05-09 body/overview.tex
final version

@@
 \Pa and \Pb are addressed sequentially: From a given problem (a) we first plan a root guide path (b), before
-extending it into a sequence of static equilibrium configurations(c). In the case where step c fails,
-our framework invalidates the computed guide path, and restarts the planning from b.
+extending it into a sequence of static equilibrium configurations (c). In the case where step (c) fails,
+our framework invalidates the computed guide path, and restarts the planning from (b).
 

### 2017-05-09 body/contact.tex
final final

--- 2017-04-20 body/contact.tex
@@
 
-To generate contacts, we proceed as follows.
-The contact generator treats each limb $R^k$ independently.
-Given a configuration of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
+Given a configuration of the root and the list of effectors that should be in contact, the contact generator computes the configuration of the limbs such that contacts are properly satisfied and the robot is in static equilibrium:
+
 \begin{equation}
@@
 
-Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
-This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.
-We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
+In previous works~\cite{DBLP:conf/iser/EscandeKMG08,Bouyarmane2009}, the generation of contact is typically implemented by randomly sampling configurations and projecting the whole robot configuration onto the closest surfaces with an inverse kinematics solver.
+In case of failure of the projection, the process would randomly iterate.
+
+We propose two modifications of this general algorithm principle.
+First our contact generator handles each limb $R^k$ independently.
+By handling each limb separately, we reduce the complexity of the generation of contact configurations.
+This is made possible thanks to the reachability condition in $\mathcal{P}_1$ that produces a root path that we can afford not to modify in $\mathcal{P}_2$, and because we allow both a contact break and a contact creation between two consecutive configurations of the contact sequence.
+Second, we rely on off-line generation of configuration candidates.
+%~ Contact generation is typically addressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
+%~ This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.
+%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
 
@@
 \begin{enumerate}
-\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$ (We choose $N=10^3$);
+\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$ (We choose $N=10^4$);
 \item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
@@
 %~ $N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
-The reader is invited to refer to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
+The reader is referred to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
 of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.

### 2017-05-09 body/intro.tex
final final

@@
 
-This paper extends our previous conference paper~\citep{tonneauisrr15} in several ways.
+The paper extends our previous papers presenting the reachability condition~\cite{tonneauisrr15} and the connection of our contact planner with a solver~$\mathcal{P}_3$~\cite{Carpentier2016}.
+In this paper, we introduce a new efficient criterion of the robot balance, provide the complete pseudo-code and release our source code that can be used to directly reproduce our results.
+We also provide a solution to $\mathcal{P}_3$.
+The planner is validated on robot models of HyQ and HRP-2 (while \cite{tonneauisrr15} used unrealistic avatar models) with complete performance analysis.
+
+%~ This paper extends our previous conference paper~\citep{tonneauisrr15} in several ways.
 Our approach is now demonstrated on actual robots and in simulation, contrary to virtual avatars before, in a large variety of contexts.

### 2017-09-25 reviews_tro/reviewer_response.tex
start reviewer response

\documentclass[a4paper]{article}
 \usepackage{fullpage}
\usepackage{amsfonts} % pour les lettres maths creuses \mathbb
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[frenchb]{babel}
\usepackage{aeguill}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\usepackage{listings}
\usepackage{pdfpages}

\newcommand\dx{\dot{x}}
\newcommand\dy{\dot{y}}
\newcommand\ddx{\ddot{x}}
\newcommand\ddy{\ddot{y}}
\def\real{{\mathbb R}}
\frenchbsetup%
{%
StandardItemLabels=true,%
ItemLabels=\ding{43},%
}%

\author {}
\title {Cover letter for the resubmission of our paper ``An efficient acyclic contact planner for multiped robots''}
\date {}
\begin{document}
\maketitle

\begin{quote}
  \textit{As for novel contributions, this did not seem to be addressed/explained
until page 3 (in the last paragraph of Sec. I). Much of the methodology
has already been presented in earlier work, and the particular aspects
that set THIS paper apart seem (as claimed by the authors) to be: 
}
\end{quote}

We agree that our original submission is not clear on the novelty between this paper and previous contributions.
This paper must be understood as an extension of the ISRR paper [20], while the other cited contributions from our team 
are exploited to validate our approach in a complete pipeline.
In this extension, the algorithm used to compute a sequence of contacts is the same as in [20] (although this
algorithm is not explicitely given in [20]), with the introduction of a fastest test for verifying the robust quasi static equilibrium of a contact posture.
We reformulated the end of Sec I to state clearly that the novelty of this paper essentially lies in the experimental setup and validation we achieved with real robot models, in particular
in the empirical justification of the method, and its validation in a dynamic simulation. This effort led to the proposal of the first open source, complete framework for multi contact motion synthesis, since 
the validation required to write a solver for the problem P3.

\begin{quote}  \textit{ (1) A criterion for robust static equilibrium, but it isn't
described until page 12, in Appendix B, Section B. [\dots]  It would really help to describe this MUCH earlier in
the paper (than page 13), if it is a significant novelty of your paper.}
\end{quote}

We agree with the remark. To better distinguish between what is novel and not, we modified the paper to move the content of Appendix B.B, as well as D  (the pseudo code of our algorithm) and E (reference to our source code) to the text itself. This results in the criterion being presented much earlier.

\begin{quote}  \textit{It is a pretty intuitive/typical check for quantifying a
margin for friction conesque violations (using an approximation for
the friction cone, with x and y direction forces considered
independently). }
\end{quote}
Thanks for the comment. To our knowledge, what is common in the literature is to reduce the friction coefficient to obtain more conservative bounds on the admissible forces, and thus a ``safety margin'' to the real cone boundary. An issue with this formulation is that the resulting margin is really small near the origin of the cone, and really important as we go further away from it. With our formulation, the obtained margin is constant everywhere on the cone, and thus guaranteed to be greater than the value $b_0$ we compute. We believe this formulation is new, but would be happy to cite any relevant reference if we happen to be wrong. This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.

\begin{quote}  \textit{
Provides complete pseudocode.  This is great! However,it's
not fully clear how much of this is novel vs just re-presenting the
same algorithms previously publilshed.	If only code in App. B and C
are new, that should be clear. Conversely, if most of this presentation
was NOT outlined in such detail before, that should also be clear (to
the authors' credit). Either way, a brief clarification in the intro is
recommended.}
\end{quote}

We agree with the comment. Indeed, the pseudocode was not present in [20]. We included it in the core of the paper and clarified that it is novel.

\begin{quote}  \textit{(3)	The contribution to provide a solution to P3 doesn't happen
until page 13 (Appendix C) and as with contribution (1), it needs a
better, intuitive summary (briefly) when first mentioned, as well as a
roadmap for the reader as to where it will be described. It is also
ambiguous in the paper whether to what extent (if any) this solution is
novel, versus a presentation of work from [1]. "Provide a solution"
seemed to imply novelty, in the context of this paragraph -- yet later,
the authors (near the end of page 2) state they use a "state-of-the-art
solution to P3", which implies the P3 solution is not a claim of
novelty.}
\end{quote}

Indeed, we were not clear on P3. In the new version of the paper, we do not consider that our solution
is a new contribution, although our formulation differs from [1] (our control variable is the acceleration
of the center of mass, where in [1] the control variable is the contact forces). We provide it here for completeness, because it corresponds to our implementation, and 
was required to validate our approach. The main contribution of the paper is the proposal and validation of an algorithm for P1 and P2.
We rephrased the text to clearly indicate that the algorithm is not a contribution.

\begin{quote}  \textit{
(4)	The last contribution is validation using an in-house
simulator, as opposed to more approximate (unrealistic avatar) models
used in previous work.	As with the previous contributions claimed, you
should flesh out just a bit more detail, along with a roadmap to where
a particular contribution appears in the paper, to prepare and guide
the reader a bit more. More precisely, what are the differences between
these two simulation approaches (previous and current)? You need a
better description of what your in-house simulator does and does not
do; a vague citation to another paper based on that simulator [38] is
not sufficient.}
\end{quote}

Thanks for the comment. Regarding the absence of details on the dynamic simulator, first of all we apologize because [38] is not the appropriate reference.
We added a section where we describe the simulation in more details.

Regarding the difference between our previous and current work, the postures computed in [20] were never validated in a simulator, nor the complete motions since 
we did not have a solution to P3 at the time. Then the main
difference between the avatars of [20] and the real robots we now address are the joint limitations and auto-collisions probabilities.
The robots are much more constrained in this regard, which makes the planning harder. We thus had to demonstrate that our algorithm is also working
for these models, while maintaining decent performances.
We added this missing piece of information to the paper, at the end of Section I.

\begin{quote}  \textit{[\dots] the authors need to better
distinguish what the novel contributions are, and to highlight where
these contributions will be presented in the paper.  Also, many
important details are in the appendices. The paper would be improved
with more intuitive comments/summaries on these details, rather than
just writing (for example), we use a bunch of heuristics, and they are
described in Appendix X.}
\end{quote}

We agree. The novel details of the paper have been moved from the appendix to the core of the paper, and are explicitely detailed at the end of Section I.

\begin{quote}  \textit{
Also, where you describe "sensitivity" vs "specificity", it would be
useful to give a more intuitive explanation, example [\dots]}
\end{quote}
We updated Section TODO to illustrate the meaning of those terms.

\begin{quote}  \textit{
However, I suggest the paper will be improved if
the presentation focuses more on highlighting (a) what is novel, and
(b)
what take-away messages you have, about why the whole framework is
designed the way it is [within the main text; not hidden in
appendices].}
\end{quote}

TODO

%~ 
%~ \begin{itemize}
%~ \item \textbf{from avatars to real robots.} [20] presents a proof of concept applied to plan contact sequences for virtual avatars. The main
%~ difference between these avatars and the real robots we now address are the joint limitations and auto-collisions probabilities.
%~ The robots are much more constrained in this regard, which makes the planning harder. We thus had to demonstrate that our algortihm is also working
%~ for these models, while maintaining decent performances.
%~ \item \textbf{a more efficient verification for quasi-static equilibrium.} The criteria for robust equilibrium we use replaces the one proposed in [20]. The formulation is new,
%~ but also much faster than before. Also, it provides a good way to differentiate between different contact candidates and select the most relevant one.
%~ \item \textbf{statistical validation.} Compared to [20], our paper provides an extensive analysis of the efficiency of the algorithm. This empirical
%~ study is required to justify that the simplifications made by our approach result in successful planning in general.
%~ \item \textbf{validation in simulation}: Finally, using state of the art approaches, we validated our contact plans by first generating a continuous whole body motion to interpolate them (P3),
%~ before validating them in a state of the art simulator or on the real robot. Although our formulation of P3 is novel regarding [1] (our control variable is the acceleration
%~ of the center of mass, where in [1] the control variables are the contact forces), we do not claim a strong contribution.
%~ \end{itemize}
%~ 
%~ It seems to us that this new material answers the T-RO guidelines regarding the novelty with respect to a conference paper.
%~ This is not clearly stated in the original submission, and we thus rewrote the end of Section I to reflect this.

\end{document}

### 2017-09-26 reviews_tro/reviewer_response.tex
r1 answers

@@
 \usepackage{pdfpages}
+% Another method to track changes
+% Examples of usage:
+% - This is \added[id=per,remark={we need this}]{new} text.
+% - This is \deleted[id=per,remark=obsolete]{unnecessary}text.
+% - This is \replaced[id=per]{nice}{bad} text.
+% To print the list of change use: \listofchanges
+\usepackage{changes}	% use this for the working version
+%\usepackage[final]{changes} % use this for the final version
+\definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
+\newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
+\newcommand{\addadp}[1]{\added[id=adp]{#1}}
+\newcommand{\repadp}[2]{\replaced[id=adp]{#1}{#2}}
+\definechangesauthor[name={Steve Tonneau}, color=blue]{st}
+\newcommand{\delst}[1]{\deleted[id=st]{#1}}
+\newcommand{\addst}[1]{\added[id=st]{#1}}
+\newcommand{\repst}[2]{\replaced[id=st]{#1}{#2}}
+\newcommand{\gls}[1]{\textit{#1}}
+\newcommand{\glslink}[2]{{#2}}
 
@@
 \newcommand\ddy{\ddot{y}}
-\def\real{{\mathbb R}}
+\def\real{{\mathbb R} }
+\newcommand\quot[1]{\begin{quote} \textit{quote}: \textbf{#1}\end{quote}}
+\newcommand\as[1]{\begin{quote} \textit{answer}: {#1}\end{quote} \leavevmode \\ }
 \frenchbsetup%
@@
 }%
-
+\DeclareUnicodeCharacter{00A0}{ }
 \author {}
@@
 
-\begin{quote}
-  \textit{As for novel contributions, this did not seem to be addressed/explained
+\section{Answers to reviewer 1}
+
+\quot {As for novel contributions, this did not seem to be addressed/explained
 until page 3 (in the last paragraph of Sec. I). Much of the methodology
@@
 }
-\end{quote}
 
-We agree that our original submission is not clear on the novelty between this paper and previous contributions.
+\as{We agree that our original submission is not clear on the novelty between this paper and previous contributions.
 This paper must be understood as an extension of the ISRR paper [20], while the other cited contributions from our team 
@@
 This effort led to the proposal of the first open source, complete framework for multi contact motion synthesis, since
-the validation required to write a solver for the problem P3.
+the validation required to write a solver for the problem P3. }
 
-\begin{quote}  \textit{ (1) A criterion for robust static equilibrium, but it isn't
+\quot { (1) A criterion for robust static equilibrium, but it isn't
 described until page 12, in Appendix B, Section B. [\dots]  It would really help to describe this MUCH earlier in
 the paper (than page 13), if it is a significant novelty of your paper.}
-\end{quote}
 
-We agree with the remark.
+\as{We agree with the remark.
 To better distinguish between what is novel and not, we modified the paper to move the content of Appendix B.B, as well as D  (the pseudo code of our algorithm) and E (reference to our source code) to the text itself.
-This results in the criterion being presented much earlier.
+This results in the criterion being presented much earlier.}
 
-\begin{quote}  \textit{It is a pretty intuitive/typical check for quantifying a
+\quot{It is a pretty intuitive/typical check for quantifying a
 margin for friction conesque violations (using an approximation for
@@
 independently). }
-\end{quote}
-Thanks for the comment.
+
+\as{Thanks for the comment.
 To our knowledge, what is common in the literature is to reduce the friction coefficient to obtain more conservative bounds on the admissible forces, and thus a ``safety margin'' to the real cone boundary.
@@
 We believe this formulation is new, but would be happy to cite any relevant reference if we happen to be wrong.
-This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.
+This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.}
 
-\begin{quote}  \textit{
+\quot{
 Provides complete pseudocode.  This is great! However,it's
@@
 recommended.}
-\end{quote}
 
-We agree with the comment. Indeed, the pseudocode was not present in [20]. We included it in the core of the paper and clarified that it is novel.
+\as{We agree with the comment. Indeed, the pseudocode was not present in [20]. We included it in the core of the paper and clarified that it is novel.}
 
-\begin{quote}  \textit{(3)	The contribution to provide a solution to P3 doesn't happen
+\quot{(3)	The contribution to provide a solution to P3 doesn't happen
 until page 13 (Appendix C) and as with contribution (1), it needs a
@@
 novelty.}
-\end{quote}
 
-Indeed, we were not clear on P3. In the new version of the paper, we do not consider that our solution
+\as{Indeed, we were not clear on P3. In the new version of the paper, we do not consider that our solution
 is a new contribution, although our formulation differs from [1] (our control variable is the acceleration
@@
 was required to validate our approach. The main contribution of the paper is the proposal and validation of an algorithm for P1 and P2.
-We rephrased the text to clearly indicate that the algorithm is not a contribution.
+We rephrased the text to clearly indicate that the algorithm is not a contribution. }
 
-\begin{quote}  \textit{
+\quot{
 (4)	The last contribution is validation using an in-house
@@
 not sufficient.}
-\end{quote}
 
-Thanks for the comment. Regarding the absence of details on the dynamic simulator, first of all we apologize because [38] is not the appropriate reference.
+\as{Thanks for the comment. Regarding the absence of details on the dynamic simulator, first of all we apologize because [38] is not the appropriate reference.
 We added a section where we describe the simulation in more details.
@@
 for these models, while maintaining decent performances.
-We added this missing piece of information to the paper, at the end of Section I.
+We added this missing piece of information to the paper, at the end of Section I.}
 
-\begin{quote}  \textit{[\dots] the authors need to better
+\quot{[\dots] the authors need to better
 distinguish what the novel contributions are, and to highlight where
@@
 described in Appendix X.}
-\end{quote}
 
-We agree. The novel details of the paper have been moved from the appendix to the core of the paper, and are explicitely detailed at the end of Section I.
+\as{We agree. The novel details of the paper have been moved from the appendix to the core of the paper, and are explicitely detailed at the end of Section I.}
 
-\begin{quote}  \textit{
+\quot{
 Also, where you describe "sensitivity" vs "specificity", it would be
 useful to give a more intuitive explanation, example [\dots]}
-\end{quote}
-We updated Section TODO to illustrate the meaning of those terms.
 
-\begin{quote}  \textit{
+\as{We updated Section TODO to illustrate the meaning of those terms.}
+
+\quot{
 However, I suggest the paper will be improved if
@@
 appendices].}
-\end{quote}
 
-TODO
+\as{TODO}
 
+\quot{
+p.6 – this would be a great place to give more ``intuition'' for the heuristics.
+Hw seems to be used in essentially all foot contact, with $h_{efort}$ for hand/arm contacts, and ALSO $h_{vel}$ not seeming to be used at all.}
+\as{Thanks for the suggestion. The text was updated accordingly, and the heuristic $h_{vel}$ was removed since indeed it is not used.}
+
+\quot{
+p.6 six fingered hand is slightly misleading, since no rolling (as fully admitted in the text). I'd just cut this? \\
+p.9 Not sure if Fig. 14 is worth including. It looks like its just Fig. 9 from the ISRR paper (?) [20], and there is still no ``full solution'' here -
+ just an illustration of potential future work. 
+ (At a minimum, the figure/solution should probably be referenced as having already appeared in the previous ISSR paper?)
+}
+
+\as{As also suggested by reviewer TODO, we removed the references to the results of the previous work [20], including those.}
+
+\quot{
+p.8– somehow, “Table II” doesn’t appear until after “Table IV” – which is a frustrating choice for latex to make.
+Not sure what you can do to change that, but it would be preferable to have tables (and figures) appear in the correct order.
+}
+
+\as{We reorganized the paper such that the Table appear in the right order.}
+
+\quot{
+p.8. Need to define ``discretization step'' more precisely, I think(?)… Min distance? Max distance? Nominal distance? In time? In Distance? In all 6 DOF??
+}
+
+\as{We updated the text to indicate that the discretization step refers to the number of postures generated along a path. For instance, a discretization step of 0.1 means
+that 10 postures will be generated along a path of length 1.
+Here the length is the distance traversed by the root along the path, computed as a ponderation between the 3D position and joint variations of the root.
+
+Finally, we also took into account all of the remaining suggestions in the pdf file attached. We do not detail them here since they refer essentially to the style 
+and the writing of the paper.}
+
+\quot{
+p.9 ‘addresses highly constrained environment’ – actually, not very well, correct?
+This was the “exception” case, for which results were rather poor, right?
+Also a case where “heuristics” are not so good for getting true (reliable) holds on grasp points, etc.?
+}
+
+\as{\textbf{This quote annoys me, and I kind of want to maintain our statement. Addressing the car scenario, even with a success rate of 68 \%, seems like a nice achievement to me.
+Also, in my opinion all our scenarios are highly constrained. What do you guys think?}}
+
+\quot{
+The first part of App.
+B is referred to within this Appendix as “new minor contributions derived from previous work”.
+These aren’t so novel, perhaps, and what is more useful to the reader is a better intuition for when to use each within the MAIN BODY of the paper, where they are originally described/associated to particular simulations.
+}
+
+\as{Agreed. We removed the incriminated text and provided an intuition of the heuristic in the text.}
+
+\quot{github link seems to be dead? (The link on p. 15 is OK… should they be the same?)}
+
+\as{We are suprised by this comment because after testing it, the link seems to be valid.
+In any case, in accordance with reviewer TODO, we changed all the links source codes to provide a SHA link that will bring the user to the accurate revision of the projects considered.}
+
+\quot{remaining remarks in the pdf file provided by reviewer 1}
+\as{The remaining comments from reviewer 1 were all addressed. Because they did not call for a discussion, as they are essentially related to style or typos, we do not list them here.}
+
+\quot{p.14 – not a good description of the simulation software used \dots Ref [38] focuses on robustness of simulations, but presumably, the testing in the present work is deterministic.
+ just using the same simulator? }
+\as{Thanks for the comment. As mentionned before in this text, we added a paragraph (TODO location) where we describe more acccurately the simulator used to perform the validation.}
 %~ 

### 2017-09-26 reviews_tro/reviewer_response.tex
finished r1

@@
 
-\quot{p.14 – not a good description of the simulation software used \dots Ref [38] focuses on robustness of simulations, but presumably, the testing in the present work is deterministic.
- just using the same simulator? }
-\as{Thanks for the comment. As mentionned before in this text, we added a paragraph (TODO location) where we describe more acccurately the simulator used to perform the validation.}
+%~ \quot{p.14 – not a good description of the simulation software used \dots Ref [38] focuses on robustness of simulations, but presumably, the testing in the present work is deterministic.
+ %~ just using the same simulator? Not much description is given of how/whether contacts/slipping are modeled\dots why isn’t a 3rd­‐part software used, or at least software with better 
+ %~ documentation? Better info is needed to understand what the simulation truly represents.}
+%~ \as{Thanks for the comment. As mentionned before in this text, we added a paragraph (TODO location) where we describe more acccurately the simulator used to perform the validation.}
 %~ 
@@
 
+\section{Answers to reviewer 4}
+\quot{I am a bit skeptical about the readability of the figures.
+Some of them seem to bee too small (e.g.
+Fig.
+3, Fig.
+4), some are of quite poor quality (why the robot silhouettes in Fig.
+4 are blurred ?), and the purpose and meaning of Fig.
+6 is rather unclear.}
+\as{Thanks for the comment, with which we agree in general.
+We increased the size of Fig.
+3 and 4.
+The robot silhouettes in Fig.
+4 are blurred when they describe a previous state of the robots.
+We added this precision in the text.
+\\Regarding the quality of our Figures, we are quite suprised by the comment.
+All of our simulation Figures are taken from high resolution rendering in blender.
+The other Figures are drawn from manually designed powerpoint vectorial graphics (converted to png).
+If the reviewer would be so kind as to point out specifically which Figures are of poor quality, we will gadly replace them.\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so the Fig.
+6 has been removed.}
+
 \end{document}

### 2017-09-26 reviews_tro/reviewer_response.tex
reviewer 4 response

@@
 
-\as{\textbf{This quote annoys me, and I kind of want to maintain our statement. Addressing the car scenario, even with a success rate of 68 \%, seems like a nice achievement to me.
-Also, in my opinion all our scenarios are highly constrained. What do you guys think?}}
+\as{\textbf{\textcolor{red}{This quote annoys me, and I kind of want to maintain our statement. Addressing the car scenario, even with a success rate of 68 \%, seems like a nice achievement to me.
+Also, in my opinion all our scenarios are highly constrained. What do you guys think?}}}
 
@@
 \section{Answers to reviewer 4}
+
+\quot{ I suggest that a few papers related to more general motion and contact planning for multi-legged robots can be added.
+Particularly, the problem of ``foothold selection'' for 4 or 6-legged robots travelling over uneven terrains is a problem related to the one tackled in the paper.
+It would be nice to select a few representative works (e.g.
+from the DARPA Learning Locomotion project or
+papers related to the ETHZ StarlETH and ANYmal robots) and explain the differences between those approaches and the problem solved in the paper.}
+\as{Thanks for the relevant comment. As observed by the reviewer, our state of the art voluntarily focuses on multi-contact planning related work, but it is relevant
+to mention the mentioned papers. We added a paragraph at the end of the state of the art to discuss them. As a matter of fact, we believe these approaches could be considered within
+our framework to replace specific parts.}
+
+\quot{in my opinion a T-RO regular paper should be more self-containing.
+In several points the authors heavily rely on previous publications, e.g.
+when presenting the RRT-based planner (Section IV-C).
+This part should be expanded, describing in more detail what exactly has been changed with respect to the reference paper.
+}
+\as{Thanks for the comment. As it is also suggested by reviewer 1 and the AE, we modified the paper for it to be self contained. The references to previous work now solely indicate
+that a detailed method is not a contribution of the present paper. \\
+
+In the particular case of Section IV-C, in fact the code of our implementation of the Bi-RRT with respect to the original one is exactly the same, with a single difference: when
+traditionally a collision detection predicate is called to validate a configuration, we replace this predicate with a call to the reachability condition.
+We modified the text to explain this more clearly.
+As the Bi-RRT is a famous standard algorithm in the literature, we do not think it is necessary to include it in this paper, but will gladly do so if required.
+}
+
 \quot{I am a bit skeptical about the readability of the figures.
-Some of them seem to bee too small (e.g.
+Some of them seem to be too small (e.g.
 Fig.
@@
 \as{Thanks for the comment, with which we agree in general.
-We increased the size of Fig.
+\\We increased the size of Fig.
 3 and 4.
@@
 
+\quot{In general, the simulation results presented qualitatively in the paper are a bit hard to interpret without the accompanying video, particularly if someone looks at a b/w print of the paper (as it may be looking in T-RO).}
+\as{\textbf{\textcolor{red}{TODO This is true, although I think he has a crappy printer because I don't think it's that bad. I don't know what to do about it though. I can't simply remove the Figures,
+because he tends to say he would like the paper to be more independent from the video...}}}
+
+\quot{ I also found the description of the environment model being
+used,
+and the way it is used in planning quite superficial. It is said, that
+``an octree data structure'' is used to keep the sampled the
+end-effector positions, and it is ``intersect'' with the environment.
+The idea is OK, I understand that the octree is a sample-based
+representation of the effector's workspace, and using some environment
+model it is possible to decide if the given configuration can provide a
+contact point or not. but how is the environment described ? How
+efficient is this intersection operation ? Does this method require the
+environment to be described by some specific objects (e.g. polygons) ?
+What about the (unavoidable) uncertainty if a world model acquired via
+on-line sensing is considered ? Perhaps some of these issues (e.g.
+uncertainty) are not in the scope of the paper, but I can imagine that
+a robotic researcher who reads the paper and is about to use the open
+source software it advertises is very interested in the answers.}
+
+\as{Thanks for the comment. The choice of not developing too much on the environment description and octree goes along with 
+our effort to maintain the paper as readable as possible, since it already contains various concepts. However,
+we agree that additional information must be added. We added a Section TODO where we indicate that the environment is represented as a polygon soup (or 3D mesh), and
+that we assume that it is fully known. Although the integration of on-line sensing and state estimation in the planner is an issue currently considered by our team, it is indeed
+out of the scope of this paper. We thus also indicate that we assume the environment and state of the robot are fully known. We also indicated the algorithmic complexity of the intersection
+method to describe how efficient the intersection operation is.}
+
+\quot{I appreciate the thorough presentation of results in the paper, however
+it seems that this presentation is somewhat biased towards simulation
+results. Simulations are perfectly OK for planning algorithms, but if
+you decided to present the experimental results (HRP-2), then at least some
+illustrations should be included in the paper. now they are only in the
+accompanying video.  also a comparison between the results of the same
+task(s) executed in dynamic simulation and on the real robot should
+convince the reader that the planning method's good performance does
+not depend too much on the idealized simulation environment. }
+\as{We agree with the remark, and included some illustrations of the experimental results on HRP-2 in the paper. \\
+We now elaborate on the validity of the simulation with respect to an experiment on the real robot. As also suggested by reviewer 1, we added a paragraph
+in the paper (TODO where) to provide more information on the simulation we use to validate our motions.
+This simulation is an implementation of the state of the art approaches used by the community.
+Of course, it does not guarantee that any motion it validates will be accurately played on the robot, but it is used extensively by the community as a reasonable guarantee of feasibility.
+The suggestion of the reviewer to compare the simulation results with those on the real robot is interesting.
+However, our robot HRP-2 is not torque controlled, but position controlled, therefore as of today a comparison is not possible.}
 \end{document}

### 2017-09-27 reviews_tro/reviewer_response.tex
all reviews

@@
 However, our robot HRP-2 is not torque controlled, but position controlled, therefore as of today a comparison is not possible.}
+
+\section{Answers to reviewer 6}
+\quot{ My primary concern, however,
+is that the elements described in this paper have been described in
+previous published work, which makes the particular contributions of
+this work unclear. To be specific, the reachability-guided planner
+appears identical to the planner in [20] (as stated by the authors in
+the end of Sect. I.B.), and the optimal control formulation is from [1]
+(again as stated in I.B.).}
+\as{Thanks for the comment.
+Indeed, all reviewers agreed that the contribution with respect to previous work was not clearly stated.
+We updated the paper to reflect the following.
+Firstly, our planner is almost identical to the one in [20], with important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
+for checking static equilibrium (see next question). The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation. The main
+novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].
+We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}
+
+\quot{The authors highlight a particular
+contribution of this work to be the efficient robust balance criterion,
+but I had trouble understanding this particular contribution. Does this
+refer to the approximation of equilibrium feasibility by contact
+reachability? In that case, how does this differ from just applying the
+planner in [20] directly? Or does this refer to the robustness
+parameter $b_0$ in Appendix B Eq. (15)?}
+\as{Indeed, this part was also not clear. The robust balance criterion (renamed equilibrium criterion) indeed refers to the parameter $b_0$ in Appendix B. The solved
+LP replaces the static equilibrium test performed in [20], which was not computationally efficient, and took up to 4ms per candidate to compute. We updated the paper to explain
+this in details.}
+
+\quot{In addition, sections IV and V are very similar to the description of
+the planner in [20]. That's not surprising, of course, but there is no
+indication in those sections that the work being presented is a
+rephrasing of an earlier publication.[\dots] }
+\as{We agree with the comment, and indeed made this clearer in the new submission.}
+
+\quot{I do, however,
+believe that the fact that the technical approaches are rephrasings of
+prior work should be made more clear (or that sections which can be
+replaced by citations of [20] and [1] be removed).}
+\as{Thanks for the advice. This suggestion was in contradiction with the other reviewers, who thought that the paper should be self-contained.
+The associate editor decided that the paper must be self contained, so we maintained the indicated sections, but made clearer the fact that they are rephrasing of previous work. }
+
+\quot{*I think I understand what the authors mean by solving at
+"interactive" rates, but this is worth defining. In my mind, it means
+something like "before the human operator gets bored", or roughly > 0.5
+Hz. But since the authors make the claim that this is the "first
+interactive implementation of a contact planner", then that statement
+should be made quantitative and verifiable}
+\as{We agree with the relevant comment, and updated the paper accordingly.}
+
+\quot{* In Section I.A. the authors refer to Bretl's work on multi-contact
+planning as the inspiration for the P1,P2,P3 decomposition of the
+problem. But it is not, as the authors state, always necessary to
+decompose motion planning into choosing a series of quasi-static
+configurations. There are a variety of feasible motions with no static
+equilibrium configurations, so it is worth mentioning that this
+decomposition only solves a subset of motion planning problem.}
+\as{We apologize if this was not made clear enough in our first submission. In the conclusion, we indicate that the planner is limited to specific scenarios (where quasi-static motion
+is always an option). We updated the state of the art to indicate that the decomposition in general is restricitve.}
+
+\quot{* Also in Sect. I.A, the authors state that "optimization-based
+approaches only converge locally". This is not necessarily true, as
+convex or mixed-integer convex optimizations can (sometimes) be solved
+to a verifiable global optimum. I would suggest removing this sentence
+entirely.}
+\as{Thanks for the suggestion. Our phrasing was indeed misleading and inaccurate. Our intention was to explain that the geometry of the scene make the problem highly non-convex, and
+that approaches such as the CIO, if not initialized with an appropriate initial guess, might never get out of the local minima. We followed the reviewer's advice and simply removed the sentence. }
+
+\quot{* I am not sure why the $C_{reach}$ hypothesis is described as the ``strong''
+hypothesis in, e.g., Sect. IV. Is there a corresponding weak
+hypothesis?}
+\as{We apologize for the confusion, that comes from a inaccurate use of English by the main author. What we meant was that the hypothesis is TODO}
+
+\quot{* In Sect. IV.B, the authors state that the ideal shape $B^*$ ``has no
+explicit definition''. For a particular configuration  $q_0$, $B^*$ is simply
+the surface of the robot, right? So I assume that this claim refers to
+there being no explicit definition of B* which is the volume around
+$B^*(q_0)$ for every $q_0$? I think I agree with the authors, but I found
+this claim hard to follow and would appreciate some clarification}
+\as{We updated the Section IV B.
+to better explain our claim.
+Indeed, as the reviewer states $B^*$  is the volume for every $q_0$.
+If such a volume existed, we would know exactly whether or not a given root configuration can result in a valid contact configuration.
+We of course doubt that $B^*$ actually exists. }
+
+\quot{* In Sect. V.A., the authors mention that the ability to plan a contact
+break and creation with the same limb allows the creation of dynamic
+motions. This is quite an important ability for an otherwise
+quasi-static planner, but this ability isn't described again until the
+appendix. It would be worth explaining why this is true in the paper
+itself. }
+\as{Thanks for the suggestion. We added a paragraph to better explain this in Section V.A.}
+
+\quot{In Sect. V.B, the authors mention inserting an additional state
+between contact configurations if two or more contacts cannot be
+maintained. Shouldn't this step appear somewhere in Alg. 1?}
+\as{We apologize for the lack of clarity. This is actually handled by the method \textsc{RepositionContacts}. We now realize that the name is misleading. Indeed,
+\textsc{RepositionContacts} either adds a new contact, or tries to reposition one already existing. In Appendix D, we state:
+\textit{Otherwise, the method \textsc{RepositionContacts} is called. It repositions one end effector (either a free
+limb, or the oldest active contact) towards a new contact
+position if possible. }
+
+We changed the name to \textsc{TODO}, with the hope that it makes things clearer.}
+
+\quot{* In Appendix A.A, it is mentioned that step 2 in the workspace
+computation is conservative, with reference to figure 17-right. But
+Fig. 17-right is the simplification of the convex hull, corresponding
+to step 3, and it's actually hard to tell from the figure that it is
+conservative. Not knowing anything about Blender's decimate tool, I
+assume it is operating by removing vertices, which would indeed be
+conservative, but some clarification would be useful here.}
+\as{Indeed, the decimate tools removes vertices, and this was not explained clearly in the paper. We updated the text accordingly}
+
+\quot{* Finally, I would suggest providing explicit git SHAs or tags for all
+of the github links. I've personally made the mistake of linking to a
+github project in a paper, only to have that project change and
+eventually delete the specific code I was citing. }
+\as{The reviewer is absolutely right. We indeed encounter the issue with other researchers that are currently using our planner. 
+To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project.
+Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the reviewer.}
+
 \end{document}

### 2017-09-27 body/contact.tex
starting modifications

@@
 
-%~ This approach has two main advantages.
-%~ First, this allows us to select a large number of candidates in a single proximity request.
-%~ Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
-%~ a locally-optimal candidate.
-%~ Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
-%~ Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
-%~ $N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
-The reader is referred to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
-of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
+%~ The reader is referred to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
+%~ of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
+In all our experiments, the heuristic $h$ is implemented as a variation of a manipulability-based heuristic~\cite{Yoshikawa1984}. The manipulability is a real number that quantifies how 
+``good'' a configuration is to perform a given task, based on the analysis of the Jacobian matrix.
+With such heuristics, a configuration can be chosen because it is far from singularities, and thus allows mobility in all directions.
+On the contrary, it can be chosen because it is particularly efficient to exert a force in a desired direction.
+In our experiments, the former solution is usually chosen for computing leg contacts, while the latter is used for computing hand contacts.
+We recall the manipulability measure and its derivatives in Appendix~\ref{sec:heuristics}.
 
-%~ It should be noted that a failure case for our contact generator does not necessarily mean that the planner will globally fail. Indeed,
-%~ for a given root configuration, if contact configurations are found for other limbs, resulting in static equilibrium, the configuration will be accepted despite
-%~ the failure to generate a contact.
+Finally, to verify that a configuration is in static equilibrium, we use a new robust LP formulation. It replaces the computationally inefficient double description
+approach used in our previous work~\cite{tonneauisrr15}.
 
-%~ In this Section, we have thus defined a deterministic contact planning algorithm.
-%~ The determinism of this algorithm makes it efficient, at the cost of not being probabilistically complete.
+%~ TODO: $h$ in appendix, the static equilibrium is here ~\ref{sec:Equil}
+
+\input{body/equilibrium}

### 2017-09-27 body/efort.tex
starting modifications

@@
-\section{Heuristics and criteria for contact selection}
+%~ \section{Heuristics and criteria for contact selection}
+%~ \label{sec:heuristics}
+% !TEX root =  ../main.tex
+%~ \subsection{Manipulability-based heuristics for contact selection}
+\section{Manipulability-based heuristics for contact selection}
 \label{sec:heuristics}
-% !TEX root =  ../main.tex
-\subsection{Manipulability-based heuristics for contact selection}
 This Section proposes heuristics to select a contact that optimizes desired capabilities.
@@
 In other less demanding cases, manipulability can also be considered to avoid singularities.
-To do so, we can consider the manipulability measure $h_{w}$, also given by Yoshikawa:
+To do so, we can consider the manipulability measure $h_{w}$, also given by Yoshikawa~\cite{Yoshikawa1984}:
 

### 2017-09-27 body/equilibrium.tex
typos
starting modifications

@@
 
-\subsection{A criterion for robust static equilibrium}
-The planner is designed so that any generated contact configuration is in static equilibrium.
-%~ Equilibrium is of course critical in legged locomotion. For this reason 
-We are interested in a robust 
-criterion, that ensures that the robot remains in equilibrium in a real-world application, regardless of perception and control uncertainties.
+\section{A criterion for robust static equilibrium}
+\label{sec:Equil}
+%~ The planner is designed so that any generated contact configuration is in static equilibrium.
+%~ We are interested in a robust 
+%~ criterion, that ensures that the robot remains in equilibrium in a real-world application, regardless of perception and control uncertainties.
 
-We first give a linear program (LP) that verifies whether a contact configuration allows for static equilibrium.
+We first give a linear program (LP) that verifies whether a contact configuration allows for static equilibrium. This LP is the same that was proposed in  \citep{Prete2016}.
 From this formulation we derive a new LP that quantifies the robustness of the equilibrium to uncertainties in the contact forces.
 In turn, from this value we can either choose the most robust candidate, or set a threshold on the required robustness. 
-While the presented LP is original, it is based on an analysis of the problem that we proposed in \citep{Prete2016}, where the interested reader can find more details.
+%~ While the presented LP is original, it is based on an analysis of the problem that we proposed in, where the interested reader can find more details.
 
@@
 \item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
-%~ \item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
 \item $m \in \mathbb{R}$ is the robot mass;
@@
 
-According to Coulomb's law, the nonslipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
+According to Coulomb's law, the non-slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
 As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays of unit length. We choose for instance:
@@
 can thus be expressed as a positive combination of its four generating rays.
-%~ \deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}
 
@@
 
-%~ \begin{align}
-%~ \underbrace{\mat{m (\ddot{\mathbf{c}} - \mathbf{g})  \\ m \mathbf{c} \times (\ddot{\mathbf{c}} - \mathbf{g}) + \dot{\mathbf{L}}}}_\mathbf{w}
-%~ = 
-%~ \underbrace{
-%~ \mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
-%~ \hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
-%~ }_\mathbf{G}
-%~ \bm{\beta},
-%~ \end{align}
-%~ where $\mathbf{w}\in \Rv{6}$ is the so-called gravito-inertial wrench (GIW) \citep{qiu:dhm:2011, Caron2015} and $\hat{\mathbf{p}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{p}$.
-
-%~ Static balance assumes zero acceleration thus we can write:
 \begin{align} \label{eq:new_eul}
@@
 where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.
-%~ The right hand side of \eqref{eq:new_eul} is independent of $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as: \adnote{It is not really necessary to show that equilibrium is independent of the com altitude here, anyway the com is given, not a variable, so it does not change the size of the problem}
-%~ \begin{align} \label{eq:new_eul2d}
-%~ \mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
-%~ \end{align}
 
@@
 Thus the optimal value of the LP~\eqref{eq:dual} is indeed the optimal $b_0$.
-
-%~ In conclusion, the LP~\eqref{eq:dual} provides a method to check rapidly static equilibrium, and returns a robustness value.
-%~ This robustness criterion can be immediately transformed into a heuristic, and eventually be weighted with others.
-%~ Another option to enforce robustness is to set a minimum acceptable value for $b_0$, but as a result this conservative approach discards valid solutions.

### 2017-09-27 body/intro.tex
starting modifications

@@
 
-Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem.
+Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem. \\
 
-The paper extends our previous papers presenting the reachability condition~\cite{tonneauisrr15} and the connection of our contact planner with a solver~$\mathcal{P}_3$~\cite{Carpentier2016}.
-In this paper, we introduce a new efficient criterion of the robot balance, provide the complete pseudo-code and release our source code that can be used to directly reproduce our results.
-We also provide a solution to $\mathcal{P}_3$.
-The planner is validated on robot models of HyQ and HRP-2 (while \cite{tonneauisrr15} used unrealistic avatar models) with complete performance analysis.
+\noindent \textbf{Comparison with our previous work:}
+The present paper is an extension of our conference paper~\cite{tonneauisrr15}. As such, the motion planning algorithm presented here is the same as the one presented in the
+previous work. Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper. However, we provide the pseudo-code of the algorithm,
+absent from the original paper. We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation.
+Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
+all the technical information required to use the planner.
+
+To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
+We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work~\cite{Carpentier2016}.
+The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
+Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of the state of the art simulator proposed in TODO.
+We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand the experimental contribution of this paper.
+
+Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi contact motions from scratch.
+The novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
+As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are much more constrained kinematically than the virtual avatars used before,
+thus making the planning problem even harder. Through an extensive empirical analysis, we demonstrate that the method remains interesting in this context, qualitatively and efficiently.
 

### 2017-09-28 reviews_tro/reviewer_response.tex
starting modifications
ongoing, added TODO / DONE macros in letter

@@
 \newcommand{\glslink}[2]{{#2}}
+\newcommand{\done}[0]{\textcolor{green}{DONE}}
+\newcommand{\ndone}[0]{\textcolor{red}{TODO}}
 
@@
 algorithm is not explicitely given in [20]), with the introduction of a fastest test for verifying the robust quasi static equilibrium of a contact posture.
-We reformulated the end of Sec I to state clearly that the novelty of this paper essentially lies in the experimental setup and validation we achieved with real robot models, in particular
+We reformulated the end of Section I.B to state clearly that the novelty of this paper essentially lies in the experimental setup and validation we achieved with real robot models, in particular
 in the empirical justification of the method, and its validation in a dynamic simulation.
 This effort led to the proposal of the first open source, complete framework for multi contact motion synthesis, since
-the validation required to write a solver for the problem P3. }
+the validation required to write a solver for the problem P3. }\done
 
@@
 \as{We agree with the remark.
-To better distinguish between what is novel and not, we modified the paper to move the content of Appendix B.B, as well as D  (the pseudo code of our algorithm) and E (reference to our source code) to the text itself.
-This results in the criterion being presented much earlier.}
+To better distinguish between what is novel and not, we modified the paper to move the content of Appendix B.B (now Section VI), as well as D  (the pseudo code of our algorithm, now Section V.B.2) and E (reference to our source code, now Section VII) to the text itself.
+This results in the criterion being presented much earlier.}\done
 
@@
 We believe this formulation is new, but would be happy to cite any relevant reference if we happen to be wrong.
-This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.}
+This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.}\done
 
@@
 
-\as{We agree with the comment. Indeed, the pseudocode was not present in [20]. We included it in the core of the paper and clarified that it is novel.}
+\as{We agree with the comment. Indeed, the pseudocode was not present in [20]. We included it in the core of the paper and clarified that it is novel at the end
+of Section I.B.}\done
 
@@
 was required to validate our approach. The main contribution of the paper is the proposal and validation of an algorithm for P1 and P2.
-We rephrased the text to clearly indicate that the algorithm is not a contribution. }
+We rephrased the text to clearly indicate that the algorithm is not a contribution. We made the choice not to give an intuition nor describe the solver in the core
+of the paper because it is not required to understand the contributions to P1 and P2.}\done
 
@@
 for these models, while maintaining decent performances.
-We added this missing piece of information to the paper, at the end of Section I.}
+We added this missing piece of information to the paper, at the end of Section I.B.}
+\textcolor{red}{TODO ANDREA}.
 
@@
 
-\as{We agree. The novel details of the paper have been moved from the appendix to the core of the paper, and are explicitely detailed at the end of Section I.}
+\as{We agree. The novel details of the paper have been moved from the appendix to the core of the paper, and are explicitely detailed at the end of Section I.B.} \done
 
@@
 
-\as{We updated Section TODO to illustrate the meaning of those terms.}
+\as{We updated Section VIII.B.1) to illustrate the meaning of those terms.} \done
 
@@
 
-\as{TODO}
+\as{TODO}\ndone
 
@@
 Hw seems to be used in essentially all foot contact, with $h_{efort}$ for hand/arm contacts, and ALSO $h_{vel}$ not seeming to be used at all.}
-\as{Thanks for the suggestion. The text was updated accordingly, and the heuristic $h_{vel}$ was removed since indeed it is not used.}
+\as{Thanks for the suggestion. The text was updated accordingly, and the heuristic $h_{vel}$ was removed since indeed it is not used.} \done
 
@@
 
-\as{As also suggested by reviewer TODO, we removed the references to the results of the previous work [20], including those.}
+\as{As also suggested by reviewer TODO, we removed the references to the results of the previous work [20], including those.}\done
 
@@
 
-\as{We reorganized the paper such that the Table appear in the right order.}
+\as{We reorganized the paper such that the Table appear in the right order.}\ndone
 
@@
 Finally, we also took into account all of the remaining suggestions in the pdf file attached. We do not detail them here since they refer essentially to the style 
-and the writing of the paper.}
+and the writing of the paper.}\ndone
 
@@
 \as{\textbf{\textcolor{red}{This quote annoys me, and I kind of want to maintain our statement. Addressing the car scenario, even with a success rate of 68 \%, seems like a nice achievement to me.
-Also, in my opinion all our scenarios are highly constrained. What do you guys think?}}}
+Also, in my opinion all our scenarios are highly constrained. What do you guys think?}}}\ndone
 
@@
 
-\as{Agreed. We removed the incriminated text and provided an intuition of the heuristic in the text.}
+\as{Agreed. We removed the incriminated text and provided an intuition of the heuristic in the text.}\ndone
 
@@
 \as{We are suprised by this comment because after testing it, the link seems to be valid.
-In any case, in accordance with reviewer TODO, we changed all the links source codes to provide a SHA link that will bring the user to the accurate revision of the projects considered.}
+In any case, in accordance with reviewer TODO, we changed all the links source codes to provide a SHA link that will bring the user to the accurate revision of the projects considered.}\ndone
 
 \quot{remaining remarks in the pdf file provided by reviewer 1}
-\as{The remaining comments from reviewer 1 were all addressed. Because they did not call for a discussion, as they are essentially related to style or typos, we do not list them here.}
+\as{The remaining comments from reviewer 1 were all addressed. Because they did not call for a discussion, as they are essentially related to style or typos, we do not list them here.}\ndone
 
@@
 to mention the mentioned papers. We added a paragraph at the end of the state of the art to discuss them. As a matter of fact, we believe these approaches could be considered within
-our framework to replace specific parts.}
+our framework to replace specific parts.}\ndone
 
@@
 If the reviewer would be so kind as to point out specifically which Figures are of poor quality, we will gadly replace them.\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so the Fig.
-6 has been removed.}
+6 has been removed.}\ndone
 
@@
 \as{\textbf{\textcolor{red}{TODO This is true, although I think he has a crappy printer because I don't think it's that bad. I don't know what to do about it though. I can't simply remove the Figures,
-because he tends to say he would like the paper to be more independent from the video...}}}
+because he tends to say he would like the paper to be more independent from the video...}}}\ndone
 
@@
 out of the scope of this paper. We thus also indicate that we assume the environment and state of the robot are fully known. We also indicated the algorithmic complexity of the intersection
-method to describe how efficient the intersection operation is.}
+method to describe how efficient the intersection operation is.} \ndone
 
@@
 The suggestion of the reviewer to compare the simulation results with those on the real robot is interesting.
-However, our robot HRP-2 is not torque controlled, but position controlled, therefore as of today a comparison is not possible.}
+However, our robot HRP-2 is not torque controlled, but position controlled, therefore as of today a comparison is not possible.}\ndone
 
@@
 novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].
-We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}
+We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}\ndone
 
@@
 LP replaces the static equilibrium test performed in [20], which was not computationally efficient, and took up to 4ms per candidate to compute. We updated the paper to explain
-this in details.}
+this in details.}\ndone
 
@@
 rephrasing of an earlier publication.[\dots] }
-\as{We agree with the comment, and indeed made this clearer in the new submission.}
+\as{We agree with the comment, and indeed made this clearer in the new submission.}\ndone
 
@@
 \as{Thanks for the advice. This suggestion was in contradiction with the other reviewers, who thought that the paper should be self-contained.
-The associate editor decided that the paper must be self contained, so we maintained the indicated sections, but made clearer the fact that they are rephrasing of previous work. }
+The associate editor decided that the paper must be self contained, so we maintained the indicated sections, but made clearer the fact that they are rephrasing of previous work. }\ndone
 
@@
 should be made quantitative and verifiable}
-\as{We agree with the relevant comment, and updated the paper accordingly.}
+\as{We agree with the relevant comment, and updated the paper accordingly.}\ndone
 
@@
 \as{We apologize if this was not made clear enough in our first submission. In the conclusion, we indicate that the planner is limited to specific scenarios (where quasi-static motion
-is always an option). We updated the state of the art to indicate that the decomposition in general is restricitve.}
+is always an option). We updated the state of the art to indicate that the decomposition in general is restricitve.}\ndone
 
@@
 \as{Thanks for the suggestion. Our phrasing was indeed misleading and inaccurate. Our intention was to explain that the geometry of the scene make the problem highly non-convex, and
-that approaches such as the CIO, if not initialized with an appropriate initial guess, might never get out of the local minima. We followed the reviewer's advice and simply removed the sentence. }
+that approaches such as the CIO, if not initialized with an appropriate initial guess, might never get out of the local minima.
+We followed the reviewer's advice and simply removed the sentence.
+}\ndone
 
@@
 hypothesis?}
-\as{We apologize for the confusion, that comes from a inaccurate use of English by the main author. What we meant was that the hypothesis is TODO}
+\as{We apologize for the confusion, that comes from a inaccurate use of English by the main author. What we meant was that the hypothesis is TODO}\ndone
 
@@
 If such a volume existed, we would know exactly whether or not a given root configuration can result in a valid contact configuration.
-We of course doubt that $B^*$ actually exists. }
+We of course doubt that $B^*$ actually exists. }\ndone
 
@@
 itself. }
-\as{Thanks for the suggestion. We added a paragraph to better explain this in Section V.A.}
+\as{Thanks for the suggestion. We added a paragraph to better explain this in Section V.A.}\ndone
 
@@
 
-We changed the name to \textsc{TODO}, with the hope that it makes things clearer.}
+We changed the name to \textsc{TODO}, with the hope that it makes things clearer.}\ndone
 
@@
 conservative, but some clarification would be useful here.}
-\as{Indeed, the decimate tools removes vertices, and this was not explained clearly in the paper. We updated the text accordingly}
+\as{Indeed, the decimate tools removes vertices, and this was not explained clearly in the paper. We updated the text accordingly}\ndone
 
@@
 To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project.
-Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the reviewer.}
+Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the reviewer.}\ndone
 

### 2017-09-28 body/abstract.tex
ongoing, added TODO / DONE macros in letter

@@
 in a
-few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ and a dexterous hand.
+few seconds, with high success rates, we generate complex contact plans for various scenarios and two robots, HRP-2 and HyQ.
 These plans are validated either in dynamic simulations, or on the real HRP-2 robot.

### 2017-09-28 body/appendix_algo.tex
ongoing, added TODO / DONE macros in letter

@@
-\section{Algorithms for the discretization of a path}
+\subsubsection{Pseudo-code of the Algorithm}
 \label{app:contact}
@@
 
-\textsc{GenerateContact}$(\mathbf{q}^0,k)$ is a call to the contact generator presented in Section~\ref{sec:single_contact}.
+The pseudo code for the method \textsc{RepositionContacts} is given by Algorithm~\ref{alg:repo}.
+
+\textsc{GenerateContact}$(\mathbf{q}^0,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
  It generates a contact configuration in static equilibrium, and assigns the corresponding configuration to $k$.
 If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.
-
-The pseudo code for the method \textsc{RepositionContacts} is given by Algorithm~\ref{alg:repo}.
 

### 2017-09-28 body/conclusion.tex
ongoing, added TODO / DONE macros in letter

@@
 \subsection{Addressing $\mathcal{P}_3$}
-This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation,
+This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation
 . Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},

### 2017-09-28 body/contact.tex
ongoing, added TODO / DONE macros in letter

@@
 of whole-body configurations  $\mathbf{Q}^{\overline{0}}$ along the root path $\mathbf{q}^0(t)$.
-The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
-
+We first give an intuition of the algorithm, before providing its complete pseudo-code.
+%~ The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
+\subsubsection{Algorithm overview}
 First, the root path $\mathbf{q}^0(t)$ is discretized into a sequence of $j$ key configurations:  
@@
 
-\subsubsection{Maintaining a contact in the sequence}
+\paragraph{Maintaining a contact in the sequence}
 
@@
 %\subsection{Generation of a contact configuration}  
-\subsubsection{Creating contacts}
+\paragraph{Creating contacts}
 Contacts are created using a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest.
 If the contact creation does not succeeds, the limb is pushed on top of the queue, and will only be tried again after the others.
+\\ \\
 %~ \begin{enumerate}
@@
 %~ \deladp{However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.}
+
+\input{body/appendix_algo}
 
@@
 Finally, to verify that a configuration is in static equilibrium, we use a new robust LP formulation. It replaces the computationally inefficient double description
-approach used in our previous work~\cite{tonneauisrr15}.
+approach used in our previous work~\cite{tonneauisrr15}, and presented in the following Section~\ref{sec:Equil}.
 

### 2017-09-28 body/efort.tex
ongoing, added TODO / DONE macros in letter

@@
 \label{sec:heuristics}
-This Section proposes heuristics to select a contact that optimizes desired capabilities.
+This Section proposes two heuristics to select a contact that optimizes desired capabilities.
 %~ These heuristics allow to locally optimize a contact location with respect to a particular task.
-For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion,
-a high velocity in a given direction, or to stay away from singular configurations.
-We derive three such heuristics from the work by \cite{Yoshikawa1984}, recalled here. %, that we present first.
+For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion, or to stay away from singular configurations.
+We derive these heuristics from the work by \cite{Yoshikawa1984}, recalled here. %, that we present first.
 
@@
 \subsubsection{Manipulability-based heuristics}
-From these definitions, we can derive three useful heuristics, that all account for the environment and the task being performed.
-The first one, EFORT, was introduced by \cite{Tonneau2014}; the other two are new minor contributions, derived from these previous works.
+From these definitions, we can derive two useful heuristics, that all account for the environment and the task being performed.
+The first one, EFORT, was introduced by \cite{Tonneau2014}; the second one derives the manipulability measure proposed by Yoshikawa~\cite{Yoshikawa1984}.
 
@@
 
-If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
-\begin{equation}
-h_{vel}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
-\end{equation}
+%~ If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
+%~ \begin{equation}
+%~ h_{vel}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
+%~ \end{equation}
 
-$h_{\textrm{\it EFORT}}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
+%~ $h_{\textrm{\it EFORT}}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
+$h_{\textrm{\it EFORT}}$ will favor contacts that allow large efforts.
 $EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.

### 2017-09-28 body/intro.tex
ongoing, added TODO / DONE macros in letter

@@
 The present paper is an extension of our conference paper~\cite{tonneauisrr15}. As such, the motion planning algorithm presented here is the same as the one presented in the
-previous work. Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper. However, we provide the pseudo-code of the algorithm,
-absent from the original paper. We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation.
+previous work.
+Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper, with the exception of Section~\ref{app:contact}, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
+We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section~\ref{sec:Equil}..
 Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide

### 2017-09-28 body/results.tex
ongoing, added TODO / DONE macros in letter

@@
 
-These scenarios complement the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
+%~ These scenarios complement the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
 %~ (Figure~\ref{fig:robots_old}).
@@
 
-\begin{figure}[t]
-\centering
-  \begin{overpic}[width=1\linewidth]{figures/robots_old}
-		%~ \put (5,) {1)} 
-		%~ \put (37,) {2)} 
-		%~ \put (68,) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
-	\end{overpic}
-\caption{Virtual avatars in various scenarios demonstrated in our conference paper.}
-		   \label{fig:robots_old}
-\end{figure}
+%~ \begin{figure}[t]
+%~ \centering
+  %~ \begin{overpic}[width=1\linewidth]{figures/robots_old}
+	%~ \end{overpic}
+%~ \caption{Virtual avatars in various scenarios demonstrated in our conference paper.}
+		   %~ \label{fig:robots_old}
+%~ \end{figure}
 
@@
 
-\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot})}
-This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
-our contact planner to compute a contact sequence for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
-Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences in less than 5 seconds.
-
-\begin{figure*}
-\centering
-  \begin{overpic}[width=1\linewidth]{figures/penrot}
-	\end{overpic}
-\caption{Contact sequence found for a pen manipulation in a zero gravity environment.}
-		   \label{fig:penrot}
-\end{figure*}
-
-\noindent\textbf{Contacts involved:} Three finger-tips.
-
-\noindent\textbf{Heuristics:} $h_{\textrm{\it EFORT}}$ for all fingers.
+%~ \subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot})}
+%~ This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
+%~ our contact planner to compute a contact sequence for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
+%~ Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences in less than 5 seconds.
+%~ 
+%~ \begin{figure*}
+%~ \centering
+  %~ \begin{overpic}[width=1\linewidth]{figures/penrot}
+	%~ \end{overpic}
+%~ \caption{Contact sequence found for a pen manipulation in a zero gravity environment.}
+		   %~ \label{fig:penrot}
+%~ \end{figure*}
+%~ 
+ %~ 
+%~ \noindent\textbf{Contacts involved:} Three finger-tips.
+%~ 
+%~ \noindent\textbf{Heuristics:} $h_{\textrm{\it EFORT}}$ for all fingers.
 
@@
 For several values of $s$, we generated $10 000$ configurations. 
-We then computed the sensitivity of the reachability condition (percentage of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}).
-Similarly we computed the specificity of the condition (percentage of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}).
+%~ We then computed the sensitivity of the reachability condition (percentage of configurations in $C_{Reach}^0$ belonging to \gls{$C_{Contact}^0$}).
+%~ Similarly we computed the specificity of the condition (percentage of configurations \textbf{not} in $C_{Reach}^0$ \textbf{not} belonging to \gls{$C_{Contact}^0$}).
+We then computed to sensitivity and specificity of the reachability condition.
+In this context, the sensitivity refers to the percentage of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}.
+If a sampled configuration is in $C_{Reach}^0$, but our method is unable to generate a contact configuration from it, as a result the sensitivity
+decreases. The sensitivity thus illustrates the confidence we have that any configuration in $C_{Reach}^0$ will effectively lead to a contact configuration.
+Conversely, the specificity refers to the percentage of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}.
+If a sampled configuration is \textbf{not} in $C_{Reach}^0$, but our method is able to generate a contact configuration from it, as a result the specificity
+decreases.
+The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to in $C_{Reach}^0$ (or informally, the confidence that we are not missing too many valid solutions).
+We thus look for a compromise between sensitivity and specificity.
+
 The obtained results for HRP-2 are shown in Table~\ref{tab:scale}, averaged over all scenes (except for the car egress: in this scenario, 

### 2017-09-29 reviews_tro/reviewer_response.tex
removed change package dep
adding added text in answer

@@
 % To print the list of change use: \listofchanges
-\usepackage{changes}	% use this for the working version
+%~ \usepackage{changes}	% use this for the working version
 %\usepackage[final]{changes} % use this for the final version
-\definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
-\newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
-\newcommand{\addadp}[1]{\added[id=adp]{#1}}
-\newcommand{\repadp}[2]{\replaced[id=adp]{#1}{#2}}
-\definechangesauthor[name={Steve Tonneau}, color=blue]{st}
-\newcommand{\delst}[1]{\deleted[id=st]{#1}}
-\newcommand{\addst}[1]{\added[id=st]{#1}}
-\newcommand{\repst}[2]{\replaced[id=st]{#1}{#2}}
-\newcommand{\gls}[1]{\textit{#1}}
-\newcommand{\glslink}[2]{{#2}}
+%~ \definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
+%~ \newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
+%~ \newcommand{\addadp}[1]{\added[id=adp]{#1}}
+%~ \newcommand{\repadp}[2]{\replaced[id=adp]{#1}{#2}}
+%~ \definechangesauthor[name={Steve Tonneau}, color=blue]{st}
+%~ \newcommand{\delst}[1]{\deleted[id=st]{#1}}
+%~ \newcommand{\addst}[1]{\added[id=st]{#1}}
+%~ \newcommand{\repst}[2]{\replaced[id=st]{#1}{#2}}
+%~ \newcommand{\gls}[1]{\textit{#1}}
+%~ \newcommand{\glslink}[2]{{#2}}
 \newcommand{\done}[0]{\textcolor{green}{DONE}}
@@
 \def\real{{\mathbb R} }
-\newcommand\quot[1]{\begin{quote} \textit{quote}: \textbf{#1}\end{quote}}
-\newcommand\as[1]{\begin{quote} \textit{answer}: {#1}\end{quote} \leavevmode \\ }
+\newcommand\quot[1]{\begin{quote} \underline{quote}: \textbf{#1}\end{quote}}
+\newcommand\as[1]{\begin{quote} \underline{answer}: {#1}\end{quote} \leavevmode \\ }
+\newcommand\qt[1]{\begin{quote} \underline{added text in paper}: \textit{#1}\end{quote} \leavevmode \\ }
 \frenchbsetup%
@@
 
-\as{We agree that our original submission is not clear on the novelty between this paper and previous contributions.
-This paper must be understood as an extension of the ISRR paper [20], while the other cited contributions from our team 
-are exploited to validate our approach in a complete pipeline.
-In this extension, the algorithm used to compute a sequence of contacts is the same as in [20] (although this
-algorithm is not explicitely given in [20]), with the introduction of a fastest test for verifying the robust quasi static equilibrium of a contact posture.
-We reformulated the end of Section I.B to state clearly that the novelty of this paper essentially lies in the experimental setup and validation we achieved with real robot models, in particular
-in the empirical justification of the method, and its validation in a dynamic simulation.
+\as{Thanks for the comment.
+Indeed, all reviewers agreed that the contribution with respect to previous work was not clearly stated.
+We updated the paper to reflect the following.
+Firstly, our planner is almost identical to the one in [20], with important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
+for checking static equilibrium (see next question). The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation. The main
+novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].
+We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}
+\qt{The present paper is an extension of our conference paper[20]. As such, the motion planning algorithm presented here is the same as the one presented in the
+previous work.
+Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
+We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section 6.
+\\
+Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
+all the technical information required to use the planner. \\
+To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
+We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work[1].
+The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
+Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of the state of the art simulator proposed in \textcolor{red}{TODO}.
+We do not consider these additions as contributions per se, but describing them here is required to understand the experimental contribution of this paper.
+Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi contact motions from scratch.
+The novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
+As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are much more constrained kinematically than the virtual avatars used before,
+thus making the planning problem even harder. Through an extensive empirical analysis, we demonstrate that the method remains interesting in this context, qualitatively and efficiently.}
+\done
+
+%~ \as{We agree that our original submission is not clear on the novelty between this paper and previous contributions.
+%~ This paper must be understood as an extension of the ISRR paper [20], while the other cited contributions from our team 
+%~ are exploited to validate our approach in a complete pipeline.
+%~ In this extension, the algorithm used to compute a sequence of contacts is the same as in [20] (although this
+%~ algorithm is not explicitely given in [20]), with the introduction of a fastest test for verifying the robust quasi static equilibrium of a contact posture.
+%~ We reformulated the end of Section I.B to state clearly that the novelty of this paper essentially lies in the experimental setup and validation we achieved with real robot models, in particular
+%~ in the empirical justification of the method, and its validation in a dynamic simulation.
 This effort led to the proposal of the first open source, complete framework for multi contact motion synthesis, since
-the validation required to write a solver for the problem P3. }\done
+%~ the validation required to write a solver for the problem P3. }\done
 
@@
 entirely.}
-\as{Thanks for the suggestion. Our phrasing was indeed misleading and inaccurate. Our intention was to explain that the geometry of the scene make the problem highly non-convex, and
+\as{Thanks for the suggestion. Our phrasing was indeed misleading and inaccurate. Our intention was to explain that the geometry of the scene makes the problem highly non-convex, and
 that approaches such as the CIO, if not initialized with an appropriate initial guess, might never get out of the local minima.

### 2017-09-29 reviews_tro/reviewer_response.tex
ongoing

@@
 %~ \newcommand{\repst}[2]{\replaced[id=st]{#1}{#2}}
-%~ \newcommand{\gls}[1]{\textit{#1}}
-%~ \newcommand{\glslink}[2]{{#2}}
+\newcommand{\gls}[1]{\textit{#1}}
+\newcommand{\glslink}[2]{{#2}}
 \newcommand{\done}[0]{\textcolor{green}{DONE}}
@@
 \newcommand\quot[1]{\begin{quote} \underline{quote}: \textbf{#1}\end{quote}}
-\newcommand\as[1]{\begin{quote} \underline{answer}: {#1}\end{quote} \leavevmode \\ }
+\newcommand\as[1]{\begin{quote} \underline{answer}: {#1}\end{quote} }
 \newcommand\qt[1]{\begin{quote} \underline{added text in paper}: \textit{#1}\end{quote} \leavevmode \\ }
+\newcommand\jp{ \leavevmode \\}
 \frenchbsetup%
@@
 To better distinguish between what is novel and not, we modified the paper to move the content of Appendix B.B (now Section VI), as well as D  (the pseudo code of our algorithm, now Section V.B.2) and E (reference to our source code, now Section VII) to the text itself.
-This results in the criterion being presented much earlier.}\done
+This results in the criterion being presented much earlier.}\done \jp
 
@@
 \as{We agree with the comment. Indeed, the pseudocode was not present in [20]. We included it in the core of the paper and clarified that it is novel at the end
-of Section I.B.}\done
+of Section I.B.}
+\qt{Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.}
+\done
 
@@
 
-\as{Indeed, we were not clear on P3. In the new version of the paper, we do not consider that our solution
+\as{The main contribution of the paper is the proposal and validation of an algorithm for P1 and P2. 
+In the new version of the paper, we do not consider that our solution to P3
 is a new contribution, although our formulation differs from [1] (our control variable is the acceleration
 of the center of mass, where in [1] the control variable is the contact forces). We provide it here for completeness, because it corresponds to our implementation, and 
-was required to validate our approach. The main contribution of the paper is the proposal and validation of an algorithm for P1 and P2.
+was required to validate our approach. 
 We rephrased the text to clearly indicate that the algorithm is not a contribution. We made the choice not to give an intuition nor describe the solver in the core
-of the paper because it is not required to understand the contributions to P1 and P2.}\done
+of the paper because it is not required to understand the contributions to P1 and P2.}
+\qt{ We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work[1].
+The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
+Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of the state of the art simulator proposed in \textcolor{red}{TODO}.
+We do not consider these additions as contributions per se, but describing them here is required to understand the experimental contribution of this paper.}\done
 
@@
 
-\as{We updated Section VIII.B.1) to illustrate the meaning of those terms.} \done
+\as{We updated Section VIII.B.1) to illustrate the meaning of those terms.}
+\qt{We then computed to sensitivity and specificity of the reachability condition.
+In this context, the sensitivity refers to the percentage of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}.
+If a sampled configuration is in $C_{Reach}^0$, but our method is unable to generate a contact configuration from it, as a result the sensitivity
+decreases. The sensitivity thus illustrates the confidence we have that any configuration in $C_{Reach}^0$ will effectively lead to a contact configuration.
+Conversely, the specificity refers to the percentage of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}.
+If a sampled configuration is \textbf{not} in $C_{Reach}^0$, but our method is able to generate a contact configuration from it, as a result the specificity
+decreases.
+The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to in $C_{Reach}^0$ (or informally, the confidence that we are not missing too many valid solutions).
+We thus look for a compromise between sensitivity and specificity.} \done
 

### 2017-09-29 body/intro.tex
typo

@@
 Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper, with the exception of Section~\ref{app:contact}, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
-We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section~\ref{sec:Equil}..
+We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section~\ref{sec:Equil}.
 Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide

### 2017-09-29 reviews_tro/reviewer_response.tex
finished quoting papers in problems questions already answered

@@
 Hw seems to be used in essentially all foot contact, with $h_{efort}$ for hand/arm contacts, and ALSO $h_{vel}$ not seeming to be used at all.}
-\as{Thanks for the suggestion. The text was updated accordingly, and the heuristic $h_{vel}$ was removed since indeed it is not used.} \done
+\as{Thanks for the suggestion. The text was updated accordingly, and the heuristic $h_{vel}$ was removed since indeed it is not used.} 
+\qt{In all our experiments, the heuristic $h$ is implemented as a variation of a manipulability-based heuristic [27]. The manipulability is a real number that quantifies how 
+``good'' a configuration is to perform a given task, based on the analysis of the Jacobian matrix.
+With such heuristics, a configuration can be chosen because it is far from singularities, and thus allows mobility in all directions.
+On the contrary, it can be chosen because it is particularly efficient to exert a force in a desired direction.
+In our experiments, the former solution is usually chosen for computing leg contacts, while the latter is used for computing hand contacts.
+We recall the manipulability measure and its derivatives in Appendix B.}\done
 

### 2017-09-29 reviews_tro/reviewer_response.tex
finished r1, added footnote on disctance

@@
 
-\as{We updated the text to indicate that the discretization step refers to the number of postures generated along a path. For instance, a discretization step of 0.1 means
-that 10 postures will be generated along a path of length 1.
-Here the length is the distance traversed by the root along the path, computed as a ponderation between the 3D position and joint variations of the root.
-
-Finally, we also took into account all of the remaining suggestions in the pdf file attached. We do not detail them here since they refer essentially to the style 
-and the writing of the paper.}\ndone
+\as{The distance considered is the distance travelled along the path, computed as a weighted 6D Euclidian variation along it. We updated the text to clarify this}.
+\qt{(Section V.B.1) $j$ depends on a user-defined variable, called the discretization step. It corresponds to the ratio between the length of the path $\mathbf{q}^0(t)$, and the number
+of configurations selected along it to create the contact configurations. The length of the path is computed as the weighted 6D Euclidian distance
+travelled along the it, with a weight of 0.7 for the translation part, and 0.3
+for the orientation part.}
 
@@
 
-\as{\textbf{\textcolor{red}{This quote annoys me, and I kind of want to maintain our statement. Addressing the car scenario, even with a success rate of 68 \%, seems like a nice achievement to me.
-Also, in my opinion all our scenarios are highly constrained. What do you guys think?}}}\ndone
+\as{Thanks for the comment. On this particular point however, we wish to maintain our statement. For scenarios such as the car egress, a success rate of $68 \%$ means that the planner
+will fail at some point, but more importantly it means that if we restart the planning a few times with the same problem, there is a reasonable probability that it will find a feasible solution.
+In this regard, the car egress scenario is effectively addressed, because we found a solution for it. \\
+Furthermore, we believe that this result is to be compared with the current state of the art : we are not aware of any other method who was able
+to compute a valid motion from scratch, even in simulation. The closest work to this achievement is the one from Escande et al., but the trajectories were manually interpolated
+after the contact plan was designed, and even so we do not have any idea on the success rate of the method. \\
+Now regarding the heuristics, we guess that the reviewer is referring to the large number of repositionnings that occured in the contact plan. We believe that this issue
+is more related to the kinematic and environmental constraints than the heuristic for the choice of contacts. Developing heuristics that would take into account the environment to limit 
+contact repositioning is indeed out of the scope of this work.}
+
+\done
 
@@
 
-\as{Agreed. We removed the incriminated text and provided an intuition of the heuristic in the text.}\ndone
+\as{Agreed. We removed the incriminated text and provided an intuition of the heuristic in the text, as mentioned above.}\done
 
@@
 
-\quot{remaining remarks in the pdf file provided by reviewer 1}
+\quot{remaining remarks by reviewer 1}
 \as{The remaining comments from reviewer 1 were all addressed. Because they did not call for a discussion, as they are essentially related to style or typos, we do not list them here.}\ndone

### 2017-09-29 body/contact.tex
finished r1, added footnote on disctance

@@
 \end{equation*}
+
 where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
+$j$ depends on a user-defined variable, called the discretization step.
+It corresponds to the ratio between the length of the path $\mathbf{q}^0(t)$\footnote{The length of the path is computed as the weighted 6D Euclidian distance
+travelled along the it, with a weight of $0.7$ for the translation part, and $0.3$ for the orientation part.}, and the number
+of configurations selected along it to create the contact configurations. 
 Each root configuration of $\mathbf{Q}^0$ is then extended into a whole-body configuration such that:

### 2017-09-29 reviews_tro/reviewer_response.tex
updated bi rrt description

@@
 that a detailed method is not a contribution of the present paper. \\ 
-
 In the particular case of Section IV-C, in fact the code of our implementation of the Bi-RRT with respect to the original one is exactly the same, with a single difference: when
@@
 }
+\qt{$C_{Reach}^0$ can be sampled efficiently thanks to Eq.~5, and can thus be used with any standard motion planner.
+Our current implementation uses the Bi-RRT planner [25] provided by the HPP software [26].
+Our implementation is exactly the same as the pseudo-code of the original planner (which does not detail the configuration validation method).
+With respect to a ``classic'' implementation, the only difference is that instead of validating a configuration using collision detection, we validate it with the \textit{reachability condition}.} \done
 

### 2017-09-29 body/rbprm.tex
updated bi rrt description

--- 2017-04-20 body/rbprm.tex
@@
 \subsection{Computing the guide path in $C_{reach}^0$}
-$C_{Reach}^0$ can be sampled efficiently thanks to Eq.~\ref{eq:reachc}, and can thus be used with a standard motion planner.
-The only significant change is to replace the collision checking with the \textit{reachability condition}.
-Our current implementation of these modifications is based on the Bi-RRT planner \citep{770022} provided by the HPP software~\citep{7759083}. \\
+$C_{Reach}^0$ can be sampled efficiently thanks to Eq.~\ref{eq:reachc}, and can thus be used with any standard motion planner.
+%~ The only significant change is to replace the collision checking with the \textit{reachability condition}.
+Our current implementation uses the Bi-RRT planner \citep{770022} provided by the HPP software~\citep{7759083}.
+Our implementation is exactly the same as the pseudo-code of the original planner (which does not detail the configuration validation method).
+With respect to a ``classic'' implementation, the only difference is that instead of validating a configuration using collision detection, we validate it with the \textit{reachability condition}.\\
 

### 2017-09-29 reviews_tro/reviewer_response.tex
octree environment

@@
 our effort to maintain the paper as readable as possible, since it already contains various concepts. However,
-we agree that additional information must be added. We added a Section TODO where we indicate that the environment is represented as a polygon soup (or 3D mesh), and
+we agree that additional information must be added. In Section III we now indicate that the environment is represented as a polygon soup (or 3D mesh), and
 that we assume that it is fully known. Although the integration of on-line sensing and state estimation in the planner is an issue currently considered by our team, it is indeed
-out of the scope of this paper. We thus also indicate that we assume the environment and state of the robot are fully known. We also indicated the algorithmic complexity of the intersection
-method to describe how efficient the intersection operation is.} \ndone
+out of the scope of this paper. We thus also indicate that we assume the environment and state of the robot are fully known. To perform the intersection computation, we use
+the fcl collision detection library, which natively proposes the intersection method. We added a reference to the library in the paper. The performance of the method is included
+in the ``collision'' column of Table II. We updated the legend of Table II to indicate that this time is negligeable in comparison to the other collision detections that occur.
+} 
+\qt{(Section III) \textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains. $O$ is represented
+as a polygon soup (or mesh), where the normal of each surface is known. No further requirement is needed
+by our approach. In this work, we assume the environment is fully known. State uncertainty is out of the scope of the paper. \\
+(Legend of Table II)  The Collision column times includes the (negligeable) octree intersection operation necessary to retrieve the candidate samples.}
+\done
 

### 2017-09-29 body/contact.tex
octree environment

@@
 \item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
-\item At runtime, when contact creation is required, intersect the octree and the environment to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
+\item At runtime, when contact creation is required, intersect the octree and the environment\footnote{this operation is achieved natively by the fcl library https://flexible-collision-library.github.io/} to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
 \item Use a user-defined heuristic $h$ to sort $S$;

### 2017-09-29 reviews_tro/reviewer_response.tex
Review response to reviewers

@@
 To our knowledge, what is common in the literature is to reduce the friction coefficient to obtain more conservative bounds on the admissible forces, and thus a ``safety margin'' to the real cone boundary.
-An issue with this formulation is that the resulting margin is really small near the origin of the cone, and really important as we go further away from it.
-With our formulation, the obtained margin is constant everywhere on the cone, and thus guaranteed to be greater than the value $b_0$ we compute.
+An issue with this formulation is that the resulting margin may be too small for small forces, and too large for large contact forces.
+With our formulation, the obtained margin is constant, and thus guaranteed to be greater than the value $b_0$ we compute.
 We believe this formulation is new, but would be happy to cite any relevant reference if we happen to be wrong.
@@
 is a new contribution, although our formulation differs from [1] (our control variable is the acceleration
-of the center of mass, where in [1] the control variable is the contact forces). We provide it here for completeness, because it corresponds to our implementation, and 
+of the center of mass, whereas in [1] the control variable is the contact forces). We provide it here for completeness, because it corresponds to our implementation, and 
 was required to validate our approach. 
@@
 The robots are much more constrained in this regard, which makes the planning harder. We thus had to demonstrate that our algorithm is also working
-for these models, while maintaining decent performances.
+for these models, while maintaining good performance.
 We added this missing piece of information to the paper, at the end of Section I.B.}
@@
 \as{We updated Section VIII.B.1) to illustrate the meaning of those terms.}
-\qt{We then computed to sensitivity and specificity of the reachability condition.
+\qt{We then computed the sensitivity and specificity of the reachability condition.
 In this context, the sensitivity refers to the percentage of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}.
@@
 decreases.
-The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to in $C_{Reach}^0$ (or informally, the confidence that we are not missing too many valid solutions).
+The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to $C_{Reach}^0$ (or informally, the confidence that we are not missing valid solutions).
 We thus look for a compromise between sensitivity and specificity.} \done
@@
 In this regard, the car egress scenario is effectively addressed, because we found a solution for it. \\
-Furthermore, we believe that this result is to be compared with the current state of the art : we are not aware of any other method who was able
+Furthermore, we believe that this result is to be compared with the current state of the art: we are not aware of any other method that was able
 to compute a valid motion from scratch, even in simulation. The closest work to this achievement is the one from Escande et al., but the trajectories were manually interpolated
 after the contact plan was designed, and even so we do not have any idea on the success rate of the method. \\
-Now regarding the heuristics, we guess that the reviewer is referring to the large number of repositionnings that occured in the contact plan. We believe that this issue
+Now regarding the heuristics, we guess that the reviewer is referring to the large number of repositionings that occurred in the contact plan. We believe that this issue
 is more related to the kinematic and environmental constraints than the heuristic for the choice of contacts. Developing heuristics that would take into account the environment to limit 
@@
 papers related to the ETHZ StarlETH and ANYmal robots) and explain the differences between those approaches and the problem solved in the paper.}
-\as{Thanks for the relevant comment. As observed by the reviewer, our state of the art voluntarily focuses on multi-contact planning related work, but it is relevant
-to mention the mentioned papers. We added a paragraph at the end of the state of the art to discuss them. As a matter of fact, we believe these approaches could be considered within
+\as{Thanks for the relevant comment.
+As observed by the reviewer, our state of the art voluntarily focuses on multi-contact planning related work, but it makes sense to cite the suggested papers.
+We added a paragraph at the end of the state of the art to discuss them.
+As a matter of fact, we believe these approaches could be considered within
 our framework to replace specific parts.}\ndone
@@
 }
-\as{Thanks for the comment. As it is also suggested by reviewer 1 and the AE, we modified the paper for it to be self contained. The references to previous work now solely indicate
-that a detailed method is not a contribution of the present paper. \\ 
+\as{Thanks for the comment. As it is also suggested by reviewer 1 and the AE, we modified the paper to make it self contained. The references to previous work now solely indicate
+that a certain method is not a contribution of the present paper. \\ 
 In the particular case of Section IV-C, in fact the code of our implementation of the Bi-RRT with respect to the original one is exactly the same, with a single difference: when
@@
 We modified the text to explain this more clearly.
-As the Bi-RRT is a famous standard algorithm in the literature, we do not think it is necessary to include it in this paper, but will gladly do so if required.
+As the Bi-RRT is a famous standard algorithm in the literature, we do not think it is necessary to include it in this paper, but we will gladly do so if required.
 }
@@
 We added this precision in the text.
-\\Regarding the quality of our Figures, we are quite suprised by the comment.
+\\Regarding the quality of our Figures, we are quite surprised by the comment.
 All of our simulation Figures are taken from high resolution rendering in blender.
@@
 \quot{In general, the simulation results presented qualitatively in the paper are a bit hard to interpret without the accompanying video, particularly if someone looks at a b/w print of the paper (as it may be looking in T-RO).}
-\as{\textbf{\textcolor{red}{TODO This is true, although I think he has a crappy printer because I don't think it's that bad. I don't know what to do about it though. I can't simply remove the Figures,
-because he tends to say he would like the paper to be more independent from the video...}}}\ndone
+\as{All of our simulation figures are taken from high resolution rendering in blender, so we wouldn't know how to improve their quality.
+If the reviewers think the paper would be better without these figures, we are willing to remove them.
+However, we believe that they would make the paper more dependent on the video, so we would prefer to leave them.}
+%\as{\textbf{\textcolor{red}{TODO This is true, although I think he has a crappy printer because I don't think it's that bad.
+I don't know what to do about it though.
+I can't simply remove the Figures, because he tends to say he would like the paper to be more independent from the video...}}}\ndone
 
@@
 
-\as{Thanks for the comment. The choice of not developing too much on the environment description and octree goes along with 
+\as{Thanks for the comment. The choice of not detailing too much the environment and octree description goes along with 
 our effort to maintain the paper as readable as possible, since it already contains various concepts. However,
@@
 the fcl collision detection library, which natively proposes the intersection method. We added a reference to the library in the paper. The performance of the method is included
-in the ``collision'' column of Table II. We updated the legend of Table II to indicate that this time is negligeable in comparison to the other collision detections that occur.
+in the ``collision'' column of Table II. We updated the legend of Table II to indicate that this time is negligible in comparison to the other collision detections that occur.
 } 
@@
 by our approach. In this work, we assume the environment is fully known. State uncertainty is out of the scope of the paper. \\
-(Legend of Table II)  The Collision column times includes the (negligeable) octree intersection operation necessary to retrieve the candidate samples.}
+(Legend of Table II)  The Collision column times includes the (negligible) octree intersection operation necessary to retrieve the candidate samples.}
 \done
@@
 This simulation is an implementation of the state of the art approaches used by the community.
-Of course, it does not guarantee that any motion it validates will be accurately played on the robot, but it is used extensively by the community as a reasonable guarantee of feasibility.
+Of course, it does not guarantee that the motion could be executed on the robot, but it is used extensively by the community as a reasonable feasibility test.
 The suggestion of the reviewer to compare the simulation results with those on the real robot is interesting.
@@
 hypothesis?}
-\as{We apologize for the confusion, that comes from a inaccurate use of English by the main author. What we meant was that the hypothesis is TODO}\ndone
+\as{We apologize for the confusion, that comes from an inaccurate use of English. What we meant was that the hypothesis is TODO}\ndone
 

### 2017-09-29 body/appendix_optim.tex
ongoing, added TODO / DONE macros in letter
Review paper according to reviewers suggestions

@@
 
-\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
+\begin{equation} \label{eq:qp_rob} \begin{aligned}
 %~ \find \quad & \mx{R} \in \Rv{3\times T}, \mxdd{C} \in  \Rv{6\times T} \\
@@
 \end{aligned} \end{equation}
-
+\adnote{Did you really use w as a variable of your problem?
+Or do you compute it inside your constraint function as a function of the CoM pos and acc?
+And what about dL?
+It appears in the definition of w but then it is no longer there in the problem definition!}
 The cost function $\ell$ is a weighted sum of the angular momentum and center-of-mass acceleration variation over the whole trajectory. 
@@
 
-This motion is then validated in our dynamic simulator~\citep{DelPrete2015b}, using the trajectory as a reference postural task for an inverse-dynamics
-controller that accounts for all the robot constraints. This final result is shown in the companion video.
+%This motion is then validated in our dynamic simulator~\citep{DelPrete2015b}, using the trajectory as a reference postural task for an inverse-dynamics
+%controller that accounts for all the robot constraints. This final result is shown in the companion video.

### 2017-09-29 body/conclusion.tex
Review paper according to reviewers suggestions

@@
-% !TEX root =  ../main.tex
+% !TEX root =  ../main_tro.tex
  \section{Discussion: validity and interest of our contact planner} 
@@
 \subsection{Addressing $\mathcal{P}_3$}
+\adnote{This subsection seems redundant to me}
 This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation

### 2017-09-29 body/intro.tex
Review paper according to reviewers suggestions

@@
-% !TEX root =  ../main.tex
+% !TEX root =  ../main_tro.tex
 \newcommand{\Pa}{$\mathcal{P}_1$ }
@@
 Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper, with the exception of Section~\ref{app:contact}, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
-We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section~\ref{sec:Equil}.
+We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section~\ref{sec:Equil}.
 Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
@@
 The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
-Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of the state of the art simulator proposed in TODO.
+Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
 We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand the experimental contribution of this paper.

### 2017-09-29 body/results.tex
octree environment
Review paper according to reviewers suggestions

@@
-% !TEX root =  ../main.tex
+% !TEX root =  ../main_tro.tex
 \section{Results}
@@
 thus the average total computation time is higher than the sum of each part). The last
-column indicates the average time necessary to compute one contact transition. }
+column indicates the average time necessary to compute one contact transition.
+The Collision column times includes the (negligeable) octree intersection operation necessary to retrieve the candidate samples.}
 \label{tab:requestime}
@@
 \subsection{Validation of the contact plans} \label{sec:valid}
-To validate our plans, we either use the framework of \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
+\adnote{I added the description of the simulations here, but I wonder whether this part should not be moved to the beginning of the Results section.}
+To generate continuous movements from our contact plans we used either the framework proposed in \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
+The resulting movements have been validated either on the real HRP-2 robot (details can be found in~\cite{Carpentier2016}), or with our dynamic simulator, based on a state-of-the-art algorithm~\cite{Kaufman2008}.
+In the simulations we controlled the robot with a standard inverse-dynamics controller~\cite{DelPrete2015b}.
+This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
+The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
 The companion video shows the obtained motions.
@@
 Providing a fair comparison with the algorithms performing on the same computer and on the same scenarios is yet out of reach.
-A step in this direction is the open-source release of our source code (see Appendix~\ref{app:hpp}) that allows any reader to reproduce our results.
+A step in this direction is the open-source release of our source code (see Section~\ref{app:hpp}) that allows any reader to reproduce our results.
 Furthermore, $\mathcal{P}_3$ remains challenging in the presence of obstacles.

### 2017-10-02 body/appendix_optim.tex
andrea s comments

@@
 %~ \find \quad & \mx{R} \in \Rv{3\times T}, \mxdd{C} \in  \Rv{6\times T} \\
-\minimize_{\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)}  \quad & \sum_{u=0}^{2} \int_{t_u}^{t_u + \Delta t_u}  \ell(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)) \text{dt}  \\
-\st \quad & \mathbf{\ddot{c}}(t) = dyn(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{w}(t)) \\
-	\quad & \mathbf{A}_u \mathbf{w}(t)  \leq \mathbf{b}_u , \forall t \in  \left[t_u, t_u + \Delta t_u \right[, \forall u \\
+%~ \minimize_{\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)}  \quad & \sum_{u=0}^{2} \int_{t_u}^{t_u + \Delta t_u}  \ell(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)) \text{dt}  \\
+\minimize_{\mathbf{\ddot{c}}(t), \dot{L}(t)}  \quad & \sum_{u=0}^{2} \int_{t_u}^{t_u + \Delta t_u}  \ell(\mathbf{\ddot{c}}(t), \dot{L}(t)) \text{dt}  \\
+%~ \st \quad & \mathbf{\ddot{c}}(t) = dyn(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{w}(t)) \\
+%~ \st \quad & \mathbf{c}(t) =  dyn(\mathbf{\ddot{c}}, \mathbf{c}_{\mathbf{q}_0} )  \\
+\st \quad & \mathbf{A}_u \mathbf{w}(t)  \leq \mathbf{b}_u , \forall t \in  \left[t_u, t_u + \Delta t_u \right[, \forall u \\
 	\quad & \mathbf{Y}_u \mathbf{c}(t)  \leq \mathbf{y}_u , \forall t \in  \left[t_u, t_u + \Delta t_u \right[, \forall u \\
@@
 \end{aligned} \end{equation}
-\adnote{Did you really use w as a variable of your problem?
+%~ \adnote{Did you really use w as a variable of your problem?
 Or do you compute it inside your constraint function as a function of the CoM pos and acc?
@@
 The cost function $\ell$ is a weighted sum of the angular momentum and center-of-mass acceleration variation over the whole trajectory.
-The system dynamics $dyn$ tie together the position, velocity and acceleration of the center of mass. 
+The center-of-mass positions and velocities $\mathbf{c}(t)$, $\mathbf{\dot{c}}(t)$ are interval variables, obtained through the double integration of $\mathbf{\ddot{c}}(t)$.
+Then $\mathbf{w}(t)$ is obtained directly from these variables. $\mathbf{c}_{\mathbf{q}_0}$ and $\mathbf{c}_{\mathbf{q}_1}$ are the center-of-mass positions for configurations
+ $\mathbf{q}_0$ and $\mathbf{q}_1$ respectively.
+%~ The system dynamics $dyn$ tie together the position, velocity and acceleration of the center of mass. 
 $\mathbf{Y}_u$ and $\mathbf{y}_u$ denote stacked kinematic constraints on the center of mass position, determined by the active contact locations.

### 2017-10-02 body/results.tex
andrea s comments

@@
 Specifically, we demonstrate the planner for two legged robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
-Finally, a last example suggests possible applications to dexterous manipulation.
+%~ Finally, a last example suggests possible applications to dexterous manipulation.
 
-At the end of the video, we validate our contact plans by exhibiting a solution to the interpolation problem $\mathcal{P}_3$.
+Our contact plans are then interpolated with a dedicated solution to the interpolation problem $\mathcal{P}_3$. This allows us to validate the obtained
+motions in a dynamic simulator. This validation is an important contribution as it increases the confidence that the contact plans we compute can effectively
+result in feasible motions on the real robot. One motion is demonstrated on the real HRP-2 robot.
+%~ The obtained motions can be seen at at the end of the video.
 
@@
 We also compare the times obtained with HRP-2 with respect to previous works.
+
+\subsection{Experimental validation of the contact plans} \label{sec:valid}
+%~ \adnote{I added the description of the simulations here, but I wonder whether this part should not be moved to the beginning of the Results section.}
+To generate continuous movements from our contact plans we used either the framework proposed in \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
+The resulting movements have been validated either on the real HRP-2 robot (details can be found in~\cite{Carpentier2016}), or with our dynamic simulator, based on a state-of-the-art algorithm~\cite{Kaufman2008}.
+In the simulations we controlled the robot with a standard inverse-dynamics controller~\cite{DelPrete2015b}.
+This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
+The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
+The companion video shows the obtained motions.
 
@@
 %~ Similarly we computed the specificity of the condition (percentage of configurations \textbf{not} in $C_{Reach}^0$ \textbf{not} belonging to \gls{$C_{Contact}^0$}).
-We then computed to sensitivity and specificity of the reachability condition.
+We then computed the sensitivity and specificity of the reachability condition.
 In this context, the sensitivity refers to the percentage of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}.
@@
 decreases.
-The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to in $C_{Reach}^0$ (or informally, the confidence that we are not missing too many valid solutions).
+The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to $C_{Reach}^0$ (or informally, the confidence that we are not missing valid solutions).
 We thus look for a compromise between sensitivity and specificity.
@@
 
-\subsection{Validation of the contact plans} \label{sec:valid}
-\adnote{I added the description of the simulations here, but I wonder whether this part should not be moved to the beginning of the Results section.}
-To generate continuous movements from our contact plans we used either the framework proposed in \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
-The resulting movements have been validated either on the real HRP-2 robot (details can be found in~\cite{Carpentier2016}), or with our dynamic simulator, based on a state-of-the-art algorithm~\cite{Kaufman2008}.
-In the simulations we controlled the robot with a standard inverse-dynamics controller~\cite{DelPrete2015b}.
-This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
-The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
-The companion video shows the obtained motions.
-
 \subsection{Comparison with previous work} \label{sec:compa}

### 2017-10-02 reviews_tro/reviewer_response.tex
andrea s comments
sota update on bretls work

@@
 \section{Answers to reviewer 1}
+\ndone reecriture p3 + enlever main video
 
@@
 We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}
-\qt{The present paper is an extension of our conference paper[20]. As such, the motion planning algorithm presented here is the same as the one presented in the
+\qt{The present paper is an extension of our conference paper [20]. As such, the motion planning algorithm presented here is the same as the one presented in the
 previous work.
 Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
-We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section 6.
+We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section VI.
 \\
@@
 \as{Thanks for the comment. Regarding the absence of details on the dynamic simulator, first of all we apologize because [38] is not the appropriate reference.
-We added a section where we describe the simulation in more details.
+We added a section VIII.A where we describe the simulation in more details.
 
@@
 We added this missing piece of information to the paper, at the end of Section I.B.}
-\textcolor{red}{TODO ANDREA}.
+\qt{(Section VIII.A) To generate continuous movements from our contact plans we used either the framework proposed in [1], or our own implementation of a $\mathcal{P}_3$ solver (Appendix C).
+The resulting movements have been validated either on the real HRP-2 robot (details can be found in [1]), or with our dynamic simulator, based on a state-of-the-art algorithm [25].
+In the simulations we controlled the robot with a standard inverse-dynamics controller[31].
+This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
+The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
+The companion video shows the obtained motions.}
+\done
+%~ \textcolor{red}{TODO ANDREA}.
 
@@
 We now elaborate on the validity of the simulation with respect to an experiment on the real robot. As also suggested by reviewer 1, we added a paragraph
-in the paper (TODO where) to provide more information on the simulation we use to validate our motions.
+in the paper (Section VIII.A)  to provide more information on the simulation we use to validate our motions.
 This simulation is an implementation of the state of the art approaches used by the community.
@@
 The suggestion of the reviewer to compare the simulation results with those on the real robot is interesting.
-However, our robot HRP-2 is not torque controlled, but position controlled, therefore as of today a comparison is not possible.}\ndone
+However, our robot HRP-2 is not torque controlled, but position controlled, therefore as of today a comparison is not possible.}
+\qt{(Section VIII.A) To generate continuous movements from our contact plans we used either the framework proposed in [1], or our own implementation of a $\mathcal{P}_3$ solver (Appendix C).
+The resulting movements have been validated either on the real HRP-2 robot (details can be found in [1]), or with our dynamic simulator, based on a state-of-the-art algorithm [25].
+In the simulations we controlled the robot with a standard inverse-dynamics controller[31].
+This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
+The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
+The companion video shows the obtained motions.}
+\done
 
@@
 appears identical to the planner in [20] (as stated by the authors in
-the end of Sect. I.B.), and the optimal control formulation is from [1]
+the end of Section I.B.), and the optimal control formulation is from [1]
 (again as stated in I.B.).}
@@
 novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].
-We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}\ndone
+We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}
+\qt{(Section I.B) The present paper is an extension of our conference paper [20]. As such, the motion planning algorithm presented here is the same as the one presented in the
+previous work.
+Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
+We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section VI.
+\\
+Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
+all the technical information required to use the planner. \\
+To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
+We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work[1].
+The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
+Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of the state of the art simulator proposed in \textcolor{red}{TODO}.
+We do not consider these additions as contributions per se, but describing them here is required to understand the experimental contribution of this paper.
+Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi contact motions from scratch.
+The novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
+As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are much more constrained kinematically than the virtual avatars used before,
+thus making the planning problem even harder. Through an extensive empirical analysis, we demonstrate that the method remains interesting in this context, qualitatively and efficiently.}\done
 
@@
 parameter $b_0$ in Appendix B Eq. (15)?}
-\as{Indeed, this part was also not clear. The robust balance criterion (renamed equilibrium criterion) indeed refers to the parameter $b_0$ in Appendix B. The solved
+\as{Indeed, this part was also not clear. The robust balance criterion (renamed equilibrium criterion) indeed refers to the parameter $b_0$ in Appendix B (now Section VI). The solved
 LP replaces the static equilibrium test performed in [20], which was not computationally efficient, and took up to 4ms per candidate to compute. We updated the paper to explain
-this in details.}\ndone
+this in details.}
+\qt{(Section I.B) We also improve the computational performance
+of the method by replacing the static equilibrium test with a more efficient test based on an
+new LP formulation presented in Section VI.}
+\done
 
@@
 rephrasing of an earlier publication.[\dots] }
-\as{We agree with the comment, and indeed made this clearer in the new submission.}\ndone
+\as{We agree with the comment, and indeed made this clearer in the new submission.}
+\qt{(Section I.B) Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.}\done
 
@@
 \as{Thanks for the advice. This suggestion was in contradiction with the other reviewers, who thought that the paper should be self-contained.
-The associate editor decided that the paper must be self contained, so we maintained the indicated sections, but made clearer the fact that they are rephrasing of previous work. }\ndone
+The associate editor decided that the paper must be self contained, so we maintained the indicated sections, but made clearer the fact that they are rephrasing of previous work. }\done
 
@@
 should be made quantitative and verifiable}
-\as{We agree with the relevant comment, and updated the paper accordingly.}\ndone
+\as{We agree with the relevant comment, and updated the paper accordingly.}
+\qt{(End of Section I.A.) We define an interactive planner as one that outputs a solution in a few seconds or less.}\done
 
@@
 \as{We apologize if this was not made clear enough in our first submission. In the conclusion, we indicate that the planner is limited to specific scenarios (where quasi-static motion
-is always an option). We updated the state of the art to indicate that the decomposition in general is restricitve.}\ndone
+is always an option) ``\textit{Our method applies to any scenario where at least one
+contact friction cone contains the direction opposed to the
+gravity (i.e. quasi-flat)[\dots]}''. We updated the state of the art to indicate that the decomposition in general is restricitve.}
+\qt{(End of before last paragraph of Section I.A) A limitation with
+this kind of approach, shared by our own method, is that most of the existing planners only address a subset of the problem, because their ability to find a solution depends
+on the existence of quasi-static equilibrium configurations along a feasible path, which is too restrictive in the general case.}\done
 

### 2017-10-02 body/conclusion.tex
sota update on bretls work

@@
 
-\subsection{Accuracy and performance gain of the reachability condition}
+%~ \subsection{Accuracy and performance gain of the reachability condition}
 
@@
 
-\subsection{Addressing $\mathcal{P}_3$}
-\adnote{This subsection seems redundant to me}
-This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation
-. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
-for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
-The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.
-In these scenarios the robot is able to execute our plans, while respecting the constraints inherent to the problem: joint kinematics and torque limits and
-collision avoidance.
-%~ \adnote{Probably you should add in appendix an explanation of how you solve $P_3$ and refer to it here.}
+%~ \subsection{Addressing $\mathcal{P}_3$}
+%~ \adnote{This subsection seems redundant to me}
+%~ This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation
+%~ . Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
+%~ for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
+%~ The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.
+%~ In these scenarios the robot is able to execute our plans, while respecting the constraints inherent to the problem: joint kinematics and torque limits and
+%~ collision avoidance.
 
@@
 \end{figure}
-
-%~ \subsection{feasibility }
 
@@
 Considering the set of obstacles intersecting with the reachable workspace for a given root configuration as candidate surfaces, we can use them to verify the equilibrium criterion.
-This would give us a necessary condition for \equilibriumfeasibility.
+This would give us a necessary condition for \equilibriumfeasibility. %, which we investigate in~\cite{fernbach:hal-01486933}.
 

### 2017-10-02 body/intro.tex
sota update on bretls work

@@
 In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}.
+A limitation with
+these approaches (including our method) is that the existing planners only address a subset of the problem, because their ability to find a solution depends
+on the existence of quasi-static equilibrium configurations along a feasible path, which is too restrictive in the general case.
 
@@
 
-As far as robotics applications are concerned, none of the existing planners is \gls{interactive}.
+As far as robotics applications are concerned, none of the existing planners is \gls{interactive}(We define an interactive planner as one that outputs a solution in a few seconds or less) . That is 
 However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.

### 2017-10-02 reviews_tro/reviewer_response.tex
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

@@
 We followed the reviewer's advice and simply removed the sentence.
-}\ndone
+}\done
 
@@
 hypothesis?}
-\as{We apologize for the confusion, that comes from an inaccurate use of English. What we meant was that the hypothesis is TODO}\ndone
+\as{We apologize for the confusion, that comes from an inaccurate use of English. What we meant by ``strong'' is that the hypothesis is constraining, because
+it defines a solution space different than the real one, and that we need to demonstrate that it makes sense.  We removed ``strong'' and replaced it by ``main'' to clarify. }\done
 
@@
 If such a volume existed, we would know exactly whether or not a given root configuration can result in a valid contact configuration.
-We of course doubt that $B^*$ actually exists. }\ndone
+We of course doubt that $B^*$ actually exists. }
+\qt{The ideal shape $B^*, W^0 \subset B^* \subset B^{\textrm{\it Suf}}$ would define a necessary \textbf{and} sufficient condition for contact creation. 
+It would guarantee that any root configuration $\mathbf{q}^{0} \in B^*$ would result in a contact configuration, while any $\mathbf{q}^{0} \notin B^*$ could not.
+To our knowledge  $B^*$ has no explicit definition.
+Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.}\done
 
@@
 
-We changed the name to \textsc{TODO}, with the hope that it makes things clearer.}\ndone
+We changed the name to \textsc{IntermediateContactState}, with the hope that it makes things clearer.}\done
 
@@
 conservative, but some clarification would be useful here.}
-\as{Indeed, the decimate tools removes vertices, and this was not explained clearly in the paper. We updated the text accordingly}\ndone
+\as{Indeed, the decimate tools removes vertices, and this was not explained clearly in the paper. We updated the text accordingly}
+\qt{The resulting polytope can contain a very large number of faces.
+A last step is thus to simplify it with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}).
+This tool removes a user-defined amount of vertices (and faces) of the polytope, thus resulting
+in a convervative approximation of the original shape. For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.}\done
 
@@
 \as{The reviewer is absolutely right. We indeed encounter the issue with other researchers that are currently using our planner. 
-To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project.
-Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the reviewer.}\ndone
+To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project, on a separate branch
+dedicated to the installation of the planner.
+Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the reviewer.}
+\qt{(Section VII) Both HPP and our planner can be simply downloaded and compiled by following the instructions on
+\url{https://humanoid-path-planner.github.io/hpp-doc/download.html?branch=rbprm}. \\
+(Appendix C) To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software.
+}
+\done
 

### 2017-10-02 body/appendix_algo.tex
Review paper according to reviewers suggestions
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

@@
+% !TEX root =  ../main_tro.tex
 \subsubsection{Pseudo-code of the Algorithm}
@@
 
-\begin{algorithm}
+\begin{algorithm}[!tbp]
 \caption{Discretization of a path} \label{interpolate}
@@
 				\EndIf				
-				\State $s = $\textsc{RepositionContacts}$(pState)$
+				\State $s = $\textsc{IntermediateContactState}$(pState)$
 			\EndIf
@@
 in computing a configuration with only one contact switch occurring.
-Otherwise, the method \textsc{RepositionContacts} is called.
+Otherwise, the method \textsc{IntermediateContactState} is called.
 It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
@@
 
-\begin{algorithm}
+\begin{algorithm}[!tbp]
 \caption{Full body contact generation method} \label{interpolate}
@@
 
-The pseudo code for the method \textsc{RepositionContacts} is given by Algorithm~\ref{alg:repo}.
+The pseudo code for the method \textsc{IntermediateContactState} is given by Algorithm~\ref{alg:repo}.
 
@@
 
-\begin{algorithm}
-\caption{Performs contact repositioning for one limb} \label{interpolate}
+\begin{algorithm}[!tbp]
+\caption{Adds or repositions a contact for one limb} \label{interpolate}
 	\begin{algorithmic}[1]
-	\Function{RepositionContacts}{$state$}
+	\Function{IntermediateContactState}{$state$}
 		\State $i=0$

### 2017-10-02 body/appendix_hpp.tex
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

@@
 Our planner is implemented using the Humanoid Path Planner (HPP) software, introduced in~\cite{7759083}.
-HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS, that can be found at \url{http://projects.laas.fr/gepetto/index.php/Software/Main}.
+HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS.
 HPP implements the standard tools and algorithms used in motion planning,
@@
 
-Our implementation of the planner is also open source, and can be found at \url{https://github.com/stonneau/hpp-rbprm}.
+Our implementation of the planner is also open source.
+Both HPP and our planner can be simply downloaded and compiled by following the instructions on
+\url{https://humanoid-path-planner.github.io/hpp-doc/download.html?branch=rbprm}.

### 2017-10-02 body/appendix_optim.tex
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

@@
 %~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
-To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox}), which can be integrated directly in our motion planner software.
+To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software.
 This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.

### 2017-10-02 body/appendix_roms.tex
corrections fro mrev 1
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

@@
 in RB-RRT, with the example of HRP-2.
-The kinematic chain is split into four limbs $R^k$.
+The kinematic tree is split into four limbs $R^k$.
 The arms are connected to the shoulders, and the legs to the root.
@@
 \item The resulting polytope can contain a very large number of faces.
-A last step is thus to simplify it in a conservative way with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}).
-For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
+A last step is thus to simplify it with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}).
+This tool removes a user-defined amount of vertices (and faces) of the polytope, thus resulting
+in a convervative approximation of the original shape. For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
 \end{enumerate}

### 2017-10-02 body/rbprm.tex
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

@@
 
-The strong hypothesis of this work is that for a large variety of locomotion tasks, we can define a space  $C_{Reach}^0 \simeq C_{Contact}^0$, such that 
+The main hypothesis of this work is that for a large variety of locomotion tasks, we can define a space  $C_{Reach}^0 \simeq C_{Contact}^0$, such that 
 \begin{equation} \label{eq:creach}
@@
 \label{sec:scaling}
-The ideal shape $B^*, W^0 \subset B* \subset B^{\textrm{\it Suf}}$ that defines a necessary \textbf{and} sufficient condition for contact creation has no explicit definition.
+The ideal shape $B^*, W^0 \subset B^* \subset B^{\textrm{\it Suf}}$ would define a necessary \textbf{and} sufficient condition for contact creation. 
+It would guarantee that any root configuration $\mathbf{q}^{0} \in B^*$ would result in a contact configuration, while any $\mathbf{q}^{0} \notin B^*$ could not.
+To our knowledge  $B^*$ has no explicit definition.
 Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.

### 2017-10-02 reviews_tro/reviewer_response.tex
finished pdf

@@
 
-\quot{github link seems to be dead? (The link on p. 15 is OK… should they be the same?)}
+\quot{github link seems to be dead? (The link on p. 15 is OK\dots should they be the same?)}
 
 \as{We are suprised by this comment because after testing it, the link seems to be valid.
-In any case, in accordance with reviewer TODO, we changed all the links source codes to provide a SHA link that will bring the user to the accurate revision of the projects considered.}\ndone
+In any case, in accordance with reviewer 6, we changed all the links source codes to provide a SHA link that will bring the user to the accurate revision of the projects considered.}
+\qt{(Section VII) Both HPP and our planner can be simply downloaded and compiled by following the instructions on
+\url{https://humanoid-path-planner.github.io/hpp-doc/download.html?branch=rbprm}. \\
+(Appendix C) To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software.
+}\done
+
+\quot{p.5, on right, ``less than epsilon'' between end effector and world - but doesn’t the orientation of the end effector matter tremendously here, too?}
+\as{
+The orientation of the end effector matters, but is not problematic regarding the way we generate contacts.
+Using an epsilon is an effective way to remove many unrelevant configurations, and
+we believe that more complex solutions aiming at taking into account the orientation would end up being counter productive.
+ Indeed, using only 3D distance, we will end up with a few configurations
+that can't be projected correctly, and will end up being discarded. But the majority of irrelevant configurations would already have been rejected with
+the efficient 3D intersection query. Conversely, introducing a 6D octree rather than the current 3D one would make the problem much more complex: instead
+of using well known 3D operations to retrieve our candidate configurations, we would have to implement our own, new solutions. The added cost of doing this is probably 
+much higher than the computational cost of discarding a few configurations that could not be projected.  } \done
 

### 2017-10-02 body/intro.tex
finished pdf

@@
 Two main directions have been explored.
-\textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}.
+\textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial aspect of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}.
 A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}.

### 2017-10-02 body/results.tex
finished pdf

@@
 In all the scenarios considered, the formulation of the problem is always the same:
-a start and goal root configurations are provided as input (except in the stair climbing scenario where the start whole body configuration is given).
+a start and goal root configuration are provided as input (except in the stair climbing scenario where the start whole body configuration is given).
 The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
@@
 %\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}
-\subsubsection{HRP-2 -- Steep staircase (Figure~\ref{fig:stair_robust})}
+\subsubsection{HRP-2 -- Staircase with high steps (Figure~\ref{fig:stair_robust})}
 
@@
   \caption{
-           HRP-2 in the steep stair climbing scenario. }
+           HRP-2 in the stair climbing scenario. }
 		   \label{fig:stair_robust}
@@
 In our conference paper~\citep{tonneauisrr15}, the computed motions were generated using the EFORT heuristic.
-EFORT is designed for tasks requiring to exert important forces (such as pushing / pulling / climbing). 
+EFORT is designed for tasks requiring large magnitude contact forces (such as pushing / pulling / climbing). 
 In locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
@@
 To analyze performance, we ran the planner 1000 times for each scenario.
-We measured the computation time spent in each part of the algorithm, and analyzed their success rate.
+We measured the computation time spent in each part of the algorithm, and analyzed success rate.
 
@@
  \hline
-   Steep stairs 	& 99.5\%  & 0.1\% 	& 0.4\% \\
+   High stairs 	& 99.5\%  & 0.1\% 	& 0.4\% \\
    Standing up 		& 87.8\%  & 6.1\% 	& 6.1\% \\
@@
 It simply means that another limb was tested for contact generation for the same root configuration.
-As expected, the more constrained scenario, the car egress, provides the less satisfying results, despite the high success rate of the planner.
+As expected, a more constrained scenario such as the car egress provides less satisfying results, despite the high success rate of the planner.
 
-These results confirm that our approach provides a satisfying compromise between completeness and efficiency, thus allowing to consider online planning
+These results confirm that our approach provides a satisfying compromise between completeness and efficiency, thus enabling online planning
 while controlling the robot. Indeed, when the contact planning fails, it fails rapidly. This allows us to rapidly re-plan with a reasonable chance of success.

### 2017-10-03 reviews_tro/reviewer_response.tex
update on contact break / creation

@@
 
-\as{TODO}\ndone
+\as{Thanks for the comments. We believe our new version of the papers answers to a). In particular, as mentionned above the novel content have been moved from the Appendix
+to the core of the paper. Regarding b), the take-away message is that to address the hard problem of multi contact planning, it is necessary to use 
+abstraction and simplification, through the decomposition of the problem into three sub-problems, and through dimensionality reduction.
+We believe that the paper develops these aspects to explain the design of this planner. In the introduction, it is done
+with an extensive state of the art that explains how the community came up with the P1, P2, P3 decomposition to tackle a problem that is too hard. In the conclusion, it is explained 
+with an analysis of the results we obtain, that justify our pragmatic choice of going further in the abstraction to obtain interactive planning performances.
+To follow the reviewer's advice, we updated the conclusion
+to include a synthesized give away message.}
+\qt{In this paper we consider the multi-contact planning problem, formulated as three sub-problems  $\mathcal{P}_1$,  $\mathcal{P}_2$,  $\mathcal{P}_3$, addressed sequentially.
+While we propose a global framework that handles all these problems, our contribution focuses on  $\mathcal{P}_1$ and  $\mathcal{P}_2$.
+The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
+the second problem $\mathcal{P}_2$ is the computation of a discrete sequence of whole-body configurations along the root path.
+We believe that this decomposition is currently the most promising approach towards
+a global resolution of the problem. We also claim to have achieved a significant step towards this objective thanks to the dimensionality reduction provided by
+the reachability condition. With our results and the release of our source code, we hope to inspire further research in this direction.}\done
 
@@
 
-\as{As also suggested by reviewer TODO, we removed the references to the results of the previous work [20], including those.}\done
+\as{As also suggested by reviewer 4, we removed the references to the results of the previous work [20], including those.}\done
 
@@
 \quot{remaining remarks by reviewer 1}
-\as{The remaining comments from reviewer 1 were all addressed. Because they did not call for a discussion, as they are essentially related to style or typos, we do not list them here.}\ndone
+\as{The remaining comments from reviewer 1 were all addressed. Because they did not call for a discussion, as they are essentially related to style or typos, we do not list them here.}\done
 
@@
 itself. }
-\as{Thanks for the suggestion. We added a paragraph to better explain this in Section V.A.}\ndone
+\as{Thanks for the suggestion. We added a paragraph to better explain this in Section V.A.}
+\qt{This representation is sufficient to describe all the contact phases, because the single support phase is implicitely described.
+Furthermore it removes the need to compute a single-support quasi-static configuration as in the example.
+Indeed, there might be a case where no quasi-static solution exists for the single support phase (because of the environment), but
+there exists a dynamic motion connecting the two double support states.
+Such motion will be computed by our framework, because the quasi-static constraint is only required at the contact planning phase; as shown in the companion video, and explained in Appendix C, our framework is able to produce dynamic motions.
+}
+\done
 

### 2017-10-03 body/conclusion.tex
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others
update on contact break / creation

@@
 Another source of computational cost identified in previous works is the verification of \equilibriumfeasibility. 
-The strong assumption of our work is that for the class of problems we consider \contactreachability\ implies \equilibriumfeasibility.
+The main assumption of our work is that for the class of problems we consider \contactreachability\ implies \equilibriumfeasibility.
 Our scenarios show that the  assumption is verified in the majority of cases when at least one contact surface is \textit{quasi flat}~\cite{Prete2016}, that is when
@@
 
-In this paper we consider the multi-contact planning problem, formulated as two sub-problems that we address sequentially.
+In this paper we consider the multi-contact planning problem, formulated as three sub-problems  $\mathcal{P}_1$,  $\mathcal{P}_2$,  $\mathcal{P}_3$, addressed sequentially.
+While we propose a global framework that handles all these problems, our contribution focuses on  $\mathcal{P}_1$ and  $\mathcal{P}_2$.
 The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
 the second problem $\mathcal{P}_2$ is the computation of a discrete sequence of whole-body configurations along the root path.
+We believe that this decomposition is currently the most promising approach towards
+a global resolution of the problem. We also claim to have achieved a significant step towards this objective thanks to the dimensionality reduction provided by
+the reachability condition. With our results and the release of our source code, we hope to inspire further research in this direction.
 
@@
 planner compatible with \gls{interactive} applications.
+
 %

### 2017-10-03 body/contact.tex
update on contact break / creation

@@
 %However, if the contact broken and the contact created sequentially are not on the same effector, our planner outputs two distinct states.
-This representation is sufficient to describe all the contact phases. Furthermore it removes the need to have single-support quasi-static configurations as in the example. 
-As shown in the companion video, this allows our framework to produce dynamic motions.
+This representation is sufficient to describe all the contact phases, because the single support phase is implicitely described.
+Furthermore it removes the need to compute a single-support quasi-static configuration as in the example.
+Indeed, there might be a case where no quasi-static solution exists for the single support phase (because of the environment), but
+there exists a dynamic motion connecting the two double support states.
+Such motion will be computed by our framework, because the quasi-static constraint is only required at the contact planning phase; as shown in the companion video, and explained in Appendix~\ref{app:optim}, our framework is able to produce dynamic motions.
 

### 2017-10-04 reviews_tro/reviewer_response.tex
rescaled figures 3 and 4

@@
 4 are blurred ?), and the purpose and meaning of Fig.
-6 is rather unclear.}
+6 is rather unclear.
+\\ In general, the simulation results presented qualitatively in the paper are a bit hard to interpret without the accompanying video, particularly if someone looks at a b/w print of the paper (as it may be looking in T-RO).}
 \as{Thanks for the comment, with which we agree in general.
@@
 We added this precision in the text.
-\\Regarding the quality of our Figures, we are quite surprised by the comment.
-All of our simulation Figures are taken from high resolution rendering in blender.
+\\Regarding the quality of our Figures,  our simulation Figures are taken from high resolution rendering in blender.
 The other Figures are drawn from manually designed powerpoint vectorial graphics (converted to png).
-If the reviewer would be so kind as to point out specifically which Figures are of poor quality, we will gadly replace them.\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so the Fig.
+We assume that the issue encountered comes from the compression of the pdf file to fit the size constraints of the
+submission site.
+We tried to improve this compression, but if the paper is accepted, we believe that the uncompressed pdf obtained from the source of the paper will result in good quality pictures.
+If the reviewers think the paper would be better without these figures, we are willing to remove them.
+However, we believe that they would make the paper more dependent on the video, so we would prefer to leave them.
+\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so the Fig.
 6 has been removed.}\ndone
 
-\quot{In general, the simulation results presented qualitatively in the paper are a bit hard to interpret without the accompanying video, particularly if someone looks at a b/w print of the paper (as it may be looking in T-RO).}
-\as{All of our simulation figures are taken from high resolution rendering in blender, so we wouldn't know how to improve their quality.
+%~ \quot{In general, the simulation results presented qualitatively in the paper are a bit hard to interpret without the accompanying video, particularly if someone looks at a b/w print of the paper (as it may be looking in T-RO).}
+%~ \as{ThAll of our simulation figures are taken from high resolution rendering in blender, so we wouldn't know how to improve their quality.
 If the reviewers think the paper would be better without these figures, we are willing to remove them.

### 2017-10-04 body/contact.tex
rescaled figures 3 and 4

@@
 % !TEX root =  ../main.tex
+\begin{figure*}[!hbtp]
+\centering
+  \begin{overpic}[width=0.6\linewidth]{figures/break_contact}
+		\put (0,4) {1} 
+		\put (25,4) {2} 
+		\put (50,4) {3} 
+		\put (76,4) {4} 
+		%~ \put (68,58) {3.a)} 
+		%~ \put (5,27) {3.b)} 
+		%~ \put (37,27) {4.a)} 
+		%~ \put (68,27) {4.b)} 
+	\end{overpic}
+\caption{Contacts are maintained if joint limits and collisions constraints are respected (2). They are broken otherwise(3,4). The green line represents the root path. The blurred character
+represents the previous contact configuration.}
+		   \label{fig:break_contact}
+\end{figure*}
+
 \section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
 \label{sec:contact}
+
 In the second phase, we compute a discrete sequence of static equilibrium configurations $\mathbf{Q}^{\overline{0}}$ given a root path
@@
 %~ one faulty contact is repositioned, in the hope that it will not be broken at the next step.
-
-\begin{figure}[t]
-\centering
-  \begin{overpic}[width=0.9\linewidth]{figures/break_contact}
-		\put (0,4) {1} 
-		\put (25,4) {2} 
-		\put (50,4) {3} 
-		\put (76,4) {4} 
-		%~ \put (68,58) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
-	\end{overpic}
-\caption{Contacts are maintained if joint limits and collisions constraints are respected (2). They are broken otherwise(3,4). The green line represents the root path.}
-		   \label{fig:break_contact}
-\end{figure}
 

### 2017-10-04 body/overview.tex
octree environment
rescaled figures 3 and 4

@@
 
+\begin{figure}[t]
+\centering
+  \begin{overpic}[width=0.8\linewidth]{figures/reachability_cond}
+	\end{overpic}
+\caption{The reachability condition is verified by the right configuration: the trunk (red) is free of collisions while the limbs reachable workspace (green) intersect the environment.}
+		   \label{fig:reach_cond}
+\end{figure}
+
+\begin{figure*}[!htbp]
+  \centering
+  \includegraphics[width=0.7\linewidth]{figures/HyQ_roms}
+  \caption{
+           Reachable workspace and torso bounding box of HyQ. Each green shape represent a reachable workspace $W^k$ of a limb. The red shape is $W^0$.}
+		   \label{fig:HyQ_roms}
+\end{figure*}
 Figure~\ref{fig:framework} illustrates our work flow.
@@
 %~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).
-
-\begin{figure}[t]
-\centering
-  \begin{overpic}[width=0.8\linewidth]{figures/reachability_cond}
-	\end{overpic}
-\caption{The reachability condition is verified by the right configuration: the trunk (red) is free of collisions while the limbs reachable workspace (green) intersect the environment.}
-		   \label{fig:reach_cond}
-\end{figure}
 
@@
 The task efficiency is measured using the Extended FORce Transmission ratio (EFORT), proposed by ~\cite{Tonneau:2014:TEC:2619648.2619652}.
+
+%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.
 
@@
 \medskip
-\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains.
+\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains. $O$ is represented
+as a polygon soup (or mesh), where the normal of each surface is known. No further requirement is needed
+by our approach. In this work, we assume the environment is fully known. State uncertainty is out of the scope of the paper.
 
@@
 \end{equation*}
-
-%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.
-\begin{figure}
-  \centering
-  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
-  \caption{
-           Reachable workspace and torso bounding box of HyQ. Each green shape represent a reachable workspace $W^k$ of a limb. The red shape is $W^0$.}
-		   \label{fig:HyQ_roms}
-\end{figure}

### 2017-10-04 body/rbprm.tex
finished pdf
rescaled figures 3 and 4

@@
 
-Given a start and a goal configurations, we aim at computing a guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ $\mathbb{R}^r$ verifying:
+Given start and goal configurations, we aim at computing a guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ $\mathbb{R}^r$ verifying:
 \begin{equation*} \label{eq:path}
@@
 \textbf{necessary conditions:}
-For a contact to be possible, an obstacle $O_i \subset O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot (Figure~\ref{fig:contact_gen}--1).
+For a contact to be possible, an obstacle $O_i \subset O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot. % (Figure~\ref{fig:contact_gen}--a).
  Also the torso of the robot $W^0(\mathbf{q}^{0})$ must necessarily be collision-free.

### 2017-10-04 reviews_tro/reviewer_response.tex
finished modifs on paper, now onto quest for typos

@@
 We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}
-\qt{The present paper is an extension of our conference paper [20]. As such, the motion planning algorithm presented here is the same as the one presented in the
+\qt{(Section I.B) The present paper is an extension of our conference paper[20]. As such, the motion planning algorithm presented here is the same as the one presented in the
 previous work.
 Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
-We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section VI.
-\\
+We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section VI.
 Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
-all the technical information required to use the planner. \\
+all the technical information required to use the planner.\\
 To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
@@
 The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
-Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of the state of the art simulator proposed in \textcolor{red}{TODO}.
-We do not consider these additions as contributions per se, but describing them here is required to understand the experimental contribution of this paper.
-Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi contact motions from scratch.
+Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
+We do not consider these additions as contributions per se, but describing them here is required to understand the experimental contribution of this paper.\\
+Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
 The novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
@@
 of the paper because it is not required to understand the contributions to P1 and P2.}
-\qt{ We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work[1].
+\qt{(Section I.B) We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work[1].
 The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
-Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of the state of the art simulator proposed in \textcolor{red}{TODO}.
+Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
 We do not consider these additions as contributions per se, but describing them here is required to understand the experimental contribution of this paper.}\done
@@
 
-\as{We updated Section VIII.B.1) to illustrate the meaning of those terms.}
-\qt{We then computed the sensitivity and specificity of the reachability condition.
+\as{We updated Section VIII.B.1 to illustrate the meaning of those terms.}
+\qt{(Section VIII.B.1) We then computed the sensitivity and specificity of the reachability condition.
 In this context, the sensitivity refers to the percentage of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}.
@@
 \as{Thanks for the comments. We believe our new version of the papers answers to a). In particular, as mentionned above the novel content have been moved from the Appendix
-to the core of the paper. Regarding b), the take-away message is that to address the hard problem of multi contact planning, it is necessary to use 
+to the core of the paper. Regarding b), the take-away message is that to address the hard problem of multi-contact planning, it is necessary to use 
 abstraction and simplification, through the decomposition of the problem into three sub-problems, and through dimensionality reduction.
@@
 to include a synthesized give away message.}
-\qt{In this paper we consider the multi-contact planning problem, formulated as three sub-problems  $\mathcal{P}_1$,  $\mathcal{P}_2$,  $\mathcal{P}_3$, addressed sequentially.
+\qt{(Section X) In this paper we consider the multi-contact planning problem, formulated as three sub-problems  $\mathcal{P}_1$,  $\mathcal{P}_2$,  $\mathcal{P}_3$, addressed sequentially.
 While we propose a global framework that handles all these problems, our contribution focuses on  $\mathcal{P}_1$ and  $\mathcal{P}_2$.
@@
 \as{Thanks for the suggestion. The text was updated accordingly, and the heuristic $h_{vel}$ was removed since indeed it is not used.} 
-\qt{In all our experiments, the heuristic $h$ is implemented as a variation of a manipulability-based heuristic [27]. The manipulability is a real number that quantifies how 
+\qt{(Section V.C) In all our experiments, the heuristic $h$ is implemented as a variation of a manipulability-based heuristic [27]. The manipulability is a real number that quantifies how 
 ``good'' a configuration is to perform a given task, based on the analysis of the Jacobian matrix.
@@
 We added a paragraph at the end of the state of the art to discuss them.
-As a matter of fact, we believe these approaches could be considered within
-our framework to replace specific parts.}\ndone
+As a matter of fact, we believe that the learning approaches
+proposed for computing relevant effector locations could be considered within
+our framework to provide good contact generation heuristics.}
+\qt{(Section I.A) Other contributions to legged locomotion, not directly related to multi-contact motion, are worth mentioning, as they rely on a similar decomposition of the problem.
+First a discrete sequence
+of contact sets is planned (at the so-called footstep planning phase), using a low-dimensional abstraction of the robot to account for its kinematic constraints [22,23].
+In [23], a ``pose certificate'' is obtained by generating a whole-body configuration for each set through inverse kinematics, as done in $\mathcal{P}_2$.
+Then a
+motion is generated along the sequence through the use of optimization techniques. 
+The solutions proposed are designed for cyclic locomotion in quasi-flat scenarios, where the support polygon is a relevant method for equilibrium checks.
+They thus cannot be generalized to to multi contact locomotion.
+However, some contributions on specific parts of the problem could be applied directly in our case. 
+For instance, learning ``terrain costs'', based on expert knowledge as proposed in [24], could define good heuristics to compute the next contact location for an effector.
+Although we did not try to include such formulation in this paper, it would be straightforward to integrate such heuristic in our planner.}\done
 
@@
 }
-\qt{$C_{Reach}^0$ can be sampled efficiently thanks to Eq.~5, and can thus be used with any standard motion planner.
+\qt{(Section IV-C) $C_{Reach}^0$ can be sampled efficiently thanks to Eq.~5, and can thus be used with any standard motion planner.
 Our current implementation uses the Bi-RRT planner [25] provided by the HPP software [26].
@@
 \\We increased the size of Fig.
-3 and 4.
+3 and 4 as much as possible to hold on one column, because we were not satisfied with the result obtained after trying to put them on a double column.
 The robot silhouettes in Fig.
@@
 (Legend of Table II)  The Collision column times includes the (negligible) octree intersection operation necessary to retrieve the candidate samples.}
-\done
+\ndone
 
@@
 We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}
-\qt{(Section I.B) The present paper is an extension of our conference paper [20]. As such, the motion planning algorithm presented here is the same as the one presented in the
+\qt{(Section I.B)  The present paper is an extension of our conference paper[20]. As such, the motion planning algorithm presented here is the same as the one presented in the
 previous work.
 Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
-We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on an new LP formulation presented in Section VI.
-\\
+We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section VI.
 Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
-all the technical information required to use the planner. \\
+all the technical information required to use the planner.\\
 To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
@@
 \as{We agree with the relevant comment, and updated the paper accordingly.}
-\qt{(End of Section I.A.) We define an interactive planner as one that outputs a solution in a few seconds or less.}\done
+\qt{(End of Section I.A.) We define an interactive planner as one for which the time to plan a motion is in the same order of magnitude as the time to execute it (For instance, computing one contact change should take about one or two seconds).}\done
 
@@
 }
-\done
+\ndone
 

### 2017-10-04 body/intro.tex
finished modifs on paper, now onto quest for typos

@@
 on the existence of quasi-static equilibrium configurations along a feasible path, which is too restrictive in the general case.
+Other contributions to legged locomotion, not directly related to multi-contact motion, are worth mentioning, as they rely on a similar decomposition of the problem.
+First a discrete sequence
+of contact sets is planned (at the so-called footstep planning phase), using a low-dimensional abstraction of the robot to account for its kinematic constraints~\cite{6078435, doi:10.1177/0278364910392608}.
+In \cite{doi:10.1177/0278364910392608}, a ``pose certificate'' is obtained by generating a whole-body configuration for each set through inverse kinematics, as done in $\mathcal{P}_2$.
+Then a
+motion is generated along the sequence through the use of optimization techniques. 
+The solutions proposed are designed for cyclic locomotion in quasi-flat scenarios, where the support polygon is a relevant method for equilibrium checks.
+They thus cannot be generalized to to multi contact locomotion.
+%For instance, the choice of the next effector in contact is often deterministic, the contacts are either coplanar or punctual, or the generation of the end-effector trajectory only works for ``ground'' obstacles, as it assumes that increasing the $z$ position of an effector drives it further away from obstacles~\cite{Kolter:2009:TTV:1703775.1703833}.
+However, some contributions on specific parts of the problem could be applied directly in our case. 
+For instance, learning ``terrain costs'', based on expert knowledge as proposed in ~\cite{Ratliff:2009:LSF:1569248.1569253}, could define good heuristics to compute the next contact location for an effector.
+Although we did not try to include such formulation in this paper, it would be straightforward to integrate such heuristic in our planner.
 
@@
 
-As far as robotics applications are concerned, none of the existing planners is \gls{interactive}(We define an interactive planner as one that outputs a solution in a few seconds or less) . That is 
+As far as robotics applications are concerned, none of the existing multi-contact planners is \gls{interactive}\footnote{We define an interactive planner as one for which the time to plan a motion is in the same order of magnitude as the time to execute it.
+For instance, computing one contact change should take about one or two seconds.}.
 However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
@@
 
-Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi contact motions from scratch.
+Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
 The novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.

### 2017-10-05 reviews_tro/reviewer_response.tex
Added intro to the letter.

@@
 \date {}
+
+%% \addtolength{\oddsidemargin}{-.6in}
+%% \linespread{1.5}
+%% \addtolength{\textwidth}{-1.5in}
+
 \begin{document}
 \maketitle
+
+\section{Summary of the modifications}
+
+First of all, we would like to thank all the reviewer and the associate editor (AE).
+We received three detailed and technical reviews that helped us to improve our paper.
+In general, we agreed we most of the reviews and take them into account in the revised version of the paper.
+The two main points that were raised by all reviewers was about i) the contributions of the paper with respect to our previous papers [20] and in a lesser extend [1] and ii) the organization of the paper, with some details postponed to appendix and some other deferred to cited papers.
+
+\textbf{About the contribution of this paper with respect to our previous papers [20] and [1]}.
+This paper is an extension of [20] (presented in the conference ISRR), that it aims at replacing.
+Like [20], the main contribution of our paper is an efficient and generic method for planning contact sequences of multi-contact locomotion, which leads to the first interactive contact planner in realistic \emph{robotic} scenarios.
+Our take-home message is  is that to address the hard problem of multi contact planning, it is necessary to use abstraction and simplification, through the decomposition of the problem into sub-problems and through dimensionality reduction.
+In particular, the planner (core algorithm) presented in this paper is almost identical to the one in [20].
+The main addition with respect to [20] is the empirical validation, with realistic robot model imposing true (difficult) kinematic  constraints.
+Specifically, the extension with respect to [20] consist in:
+\begin{itemize}
+\item the addition of the pseudo-code (the algorithm is not given in [20]);
+\item the introduction of a more efficient equilibrium criterion;
+\item the new experimental setup and analysis: validation with real robot models (while [20] was only using unrealistic avatars with are much less kinematically constrained); statistical analysis of the validity of the planner models and heuristics; generation of the complete movement through the implementation of a P3 solver; validation that the movements are feasible in a complete dynamic simulator and, for one of them, on a physical robot;
+\item the release of our source code: we provide the complete procedure to use our planner for any legged robot.
+\end{itemize}
+The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation.
+We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.
+
+\textbf{About the organization of the paper and the use of appendices and citation to reduce the core of the paper}.
+In order to try to make the paper more readable, the first version was mixing three styles: details given in the core of the paper; details given in the appendices; details deferred in cited papers.
+While all reviewers agreed that using all three styles together was a bad idea, we received contradictory reviews about how to modify the paper.
+Following the recommandation of the AE, we have worked to make the paper self-contained with all the contributions in the core of the paper.
+We have brought all the material from [20] in the revised paper so that no references to this preliminary paper is necessary anymore in the technical description, and removed trivial reference to [20] but in the introduction.
+We merged Appendices B.B (equilibrium criterion), D (pseudo code) and E (referernce to the source code) in the core of the paper.
+We kept in the appendices the manipulability computations which is necessary for completeness but is well-known.
+We also prefered to keep the trajectory solver P3 in the appendix, although some details are novel, in order to keep clear that the main contributions of the paper are on P1 and P2.
+
+\bigskip
+
+In the following, we list all the questions and remarks of the reviewers, answer to direct questions and point the modifications that were brought to the paper.
 

### 2017-10-05 reviews_tro/reviewer_response.tex
updated intro and finished text of detailed answer

--- 2017-10-04 reviews_tro/reviewer_response.tex
@@
 \date {}
+
+%% \addtolength{\oddsidemargin}{-.6in}
+%% \linespread{1.5}
+%% \addtolength{\textwidth}{-1.5in}
+
 \begin{document}
 \maketitle
+
+\section{Summary of the modifications}
+
+First of all, we would like to thank all the reviewers and the associate editor (AE).
+We received three detailed and technical reviews that helped improving our paper.
+Overall we agreed with most of the reviews and took them into account in the revised version of the paper.
+The main points of concern raised by the reviewers regarded:
+\begin{itemize}
+\item the diffficulty to understand the take-away message of the paper, and what was novel or not with respect to our previous paper [20] (and [1] to a lesser extend);
+\item the organization of the paper, with too much information relegated to the Appendix, or only refered to in other contributions.
+\end{itemize}
+
+\textbf{Regarding the contribution of this paper with respect to our previous papers}, this paper must be understood
+as an extension of [20] (presented at the ISRR conference).
+As a self-contained journal paper, the present paper aims at replacing [20].
+We present two contributions.
+The first one, as in [20], is an efficient and generic planning method for multi-contact locomotion (problems P1 and P2).
+The second one is a rigorous experimental validation of the approach on actual robot models, either in simulation or on the real HRP-2 robot.
+
+Regarding the first contribution, the algorithm presented here is the same as in [20], with three important novelties: a novel criterion for
+static equilibrium, the pseudo-code of the algorithm,
+as well as  the release of the source code of the planner.
+
+Regarding the second contribution, to validate the contact plans computed we introduced a complete framework for multi-contact motion synthesis.
+This framework additionally comprises an interpolation method to solve the problem P3, similar yet different than the one we proposed in [1].
+It also comprises an implementation of a state-of-the art dynamic simulator to check that the motions are physically plausible [28].
+These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}.
+The novelty lies in the validation of the method with real robot models, for which the planning is harder with respect to the avatars used in [20], because of stronger kinematic constraints.
+
+We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.
+To clarify these aspects in the paper, we rewrote the end of Section I.B to clearly state what is new and what is not in this paper.
+
+%~ This paper is an extension of [20] (presented at the ISRR conference), that it aims at replacing.
+%~ Like [20], the main contribution of our paper is an efficient and generic method for planning contact sequences of multi-contact locomotion, which leads to the first interactive contact planner in realistic \emph{robotic} scenarios.
+%~ Our take-home message is  is that to address the hard problem of multi contact planning, it is necessary to use abstraction and simplification, through the decomposition of the problem into sub-problems and through dimensionality reduction.
+%~ In particular, the planner (core algorithm) presented in this paper is almost identical to the one in [20].
+%~ The main addition with respect to [20] is the empirical validation, with realistic robot model imposing true (difficult) kinematic  constraints.
+%~ Specifically, the novelty with respect to [20] lies in:
+%~ \begin{itemize}
+%~ \item the addition of the pseudo-code (the algorithm is not given in [20]);
+%~ \item the introduction of a more efficient equilibrium criterion;
+%~ \item the new experimental setup and analysis: validation with real robot models (while [20] was only using unrealistic avatars with are much less kinematically constrained); statistical analysis of the validity of the planner models and heuristics; generation of the complete movement through the implementation of a P3 solver; validation that the movements are feasible in a complete dynamic simulator and, for one of them, on a physical robot;
+%~ \item the release of our source code: we provide the complete procedure to use our planner for any legged robot.
+%~ \end{itemize}
+%~ The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation.
+We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.
+
+\textbf{Regarding the organization of the paper}, in the initial version of the paper, we tried to synthesize as much as possible
+the core of the paper, while providing the most technical details in the appendix. Based on the reviewers comments, we modified
+the paper architecture in such a way that all the novel details are now directly included in the core of the paper. This should avoid the reader going back and forth
+in the paper, and also help him distinguishing between what is novel and what is not.
+As a result,  Appendices B.B (equilibrium criterion), D (pseudo code) and E (referernce to the source code) are now integrated in the core of the paper.
+The manipulability-based heuristics used for contact generation, as well as our solution to the problem P3 remain in the Appendix as they are not novel.
+However, as suggested by the reviewers, they are described in more details in the core.
+
+Following the recommandation of the associate editor who ruled over some contradictory opinions of the reviewers, we made the paper self-contained,
+in such a way that the paper [20] is now obsolete regarding this submission, and is only mentioned in the introduction, or to describe an improvement
+of the original algorithm. The missing information pointed out by the reviewers has been 
+included in the paper, as explained in the detailed answer that follows. As a result, we believe that in this new version the paper is self-contained and can be understood if read linearly.\\
+
+ %~ In order to try to make the paper more readable, the first version was mixing three styles: details given in the core of the paper; details given in the appendices; details deferred in cited papers.
+While all reviewers agreed that using all three styles together was a bad idea, we received contradictory reviews about how to modify the paper.
+Following the recommandation of the AE, we have worked to make the paper self-contained with all the contributions in the core of the paper.
+%~ We have brought all the material from [20] in the revised paper so that no references to this preliminary paper is necessary anymore in the technical description, and removed trivial reference to [20] but in the introduction.
+%~ We merged Appendices B.B (equilibrium criterion), D (pseudo code) and E (referernce to the source code) in the core of the paper.
+We kept in the appendices the manipulability computations which is necessary for completeness but is well-known.
+We also prefered to keep the trajectory solver P3 in the appendix, although some details are novel, in order to keep clear that the main contributions of the paper are on P1 and P2.
+
+We thus believe to have significantly improved the paper by addressing the majority of the comments and questions raised by the reviewers.
+
+\bigskip
+
+In the following, we list all the questions and remarks from the reviewers, answer to the questions, and detail the modifications that were brought to the paper.
+At the end of this letter, we also join a diff file between the original submission and the present one, that should help the reviewers visualizing the modifications 
+brought to the paper.
 
@@
 \as{Thanks for the comment.
-Indeed, all reviewers agreed that the contribution with respect to previous work was not clearly stated.
-We updated the paper to reflect the following.
-Firstly, our planner is almost identical to the one in [20], with important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
-for checking static equilibrium (see next question). The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation. The main
-novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].
-We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}
+As mentioned in the summary, our planner is almost identical to the one in [20], but the paper comprises important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
+for checking static equilibrium.
+The main novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].
+This is now explained in details in the introduction.}
 \qt{(Section I.B) The present paper is an extension of our conference paper[20]. As such, the motion planning algorithm presented here is the same as the one presented in the
@@
 We added this precision in the text.
+\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so the Fig.
+6 has been removed.
 \\Regarding the quality of our Figures,  our simulation Figures are taken from high resolution rendering in blender.
@@
 We assume that the issue encountered comes from the compression of the pdf file to fit the size constraints of the
-submission site.
-We tried to improve this compression, but if the paper is accepted, we believe that the uncompressed pdf obtained from the source of the paper will result in good quality pictures.
-If the reviewers think the paper would be better without these figures, we are willing to remove them.
-However, we believe that they would make the paper more dependent on the video, so we would prefer to leave them.
-\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so the Fig.
-6 has been removed.}\ndone
+submission site. We tried to improve this compression, but if the paper is accepted, we believe that the uncompressed pdf obtained from the source of the paper will result in good quality pictures.
+Regarding the interpretation of the results without the video, we agree with the reviewer and believe that the accompanying video is essential to the comprehension of the results.
+We considered three options: a) displaying several sequences of contact (as currently done); b) only displaying 2 or 3 representative consecutive contact stances; c)
+not providing any picture and referring to the video.
+We chose a) because we wanted the paper to be remain as independent from the video as possible, but are willing to opt for option b) or c) (or any other option) if preferred by the reviewer.}\done
 
@@
 environment to be described by some specific objects (e.g. polygons) ?
-What about the (unavoidable) uncertainty if a world model acquired via
-on-line sensing is considered ? Perhaps some of these issues (e.g.
-uncertainty) are not in the scope of the paper, but I can imagine that
-a robotic researcher who reads the paper and is about to use the open
-source software it advertises is very interested in the answers.}
+}
 
@@
 we agree that additional information must be added. In Section III we now indicate that the environment is represented as a polygon soup (or 3D mesh), and
-that we assume that it is fully known. Although the integration of on-line sensing and state estimation in the planner is an issue currently considered by our team, it is indeed
-out of the scope of this paper. We thus also indicate that we assume the environment and state of the robot are fully known. To perform the intersection computation, we use
+that we assume that it is fully known.  We thus also indicate that we assume the environment and state of the robot are fully known. To perform the intersection computation, we use
 the fcl collision detection library, which natively proposes the intersection method. We added a reference to the library in the paper. The performance of the method is included
@@
 (Legend of Table II)  The Collision column times includes the (negligible) octree intersection operation necessary to retrieve the candidate samples.}
-\ndone
+\done
+
+\quot{What about the (unavoidable) uncertainty if a world model acquired via
+on-line sensing is considered ? Perhaps some of these issues (e.g.
+uncertainty) are not in the scope of the paper, but I can imagine that
+a robotic researcher who reads the paper and is about to use the open
+source software it advertises is very interested in the answers.}
+\as{That is a good observation. Although the integration of on-line sensing and state estimation in the planner is an issue currently considered by our team, it is indeed
+out of the scope of this paper. As indicated in the previous answer we indicated it in Section III. We assume that our planner could work similarly with 
+a sensing system that would output data easily interpreted as a polygon soup (as for a LIDAR), otherwise additional treatment would be required.
+In any case additional work would be required to plan motions robust to the uncertainty of the sensing and the estimation.
+} \done
 
@@
 }
-\ndone
+\done
 

### 2017-10-06 reviews_tro/reviewer_response.tex
Review response to reviewer

@@
 % To print the list of change use: \listofchanges
-%~ \usepackage{changes}	% use this for the working version
+\usepackage{changes}	% use this for the working version
 %\usepackage[final]{changes} % use this for the final version
-%~ \definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
-%~ \newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
-%~ \newcommand{\addadp}[1]{\added[id=adp]{#1}}
+\definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
+\newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
+\newcommand{\addadp}[1]{\added[id=adp]{#1}}
 %~ \newcommand{\repadp}[2]{\replaced[id=adp]{#1}{#2}}
@@
 \begin{itemize}
-\item the diffficulty to understand the take-away message of the paper, and what was novel or not with respect to our previous paper [20] (and [1] to a lesser extend);
-\item the organization of the paper, with too much information relegated to the Appendix, or only refered to in other contributions.
+\item the diffficulty to understand the takeaway message of the paper, and what was novel with respect to our previous paper [20] (and [1] to a lesser extent);
+\item the organization of the paper, with too much information relegated to the Appendix, or only referred to in other contributions.
 \end{itemize}
@@
 
-Regarding the first contribution, the algorithm presented here is the same as in [20], with three important novelties: a novel criterion for
+Regarding the first contribution, the algorithm presented here is the same as in [20], but with three important novelties: a novel criterion for
 static equilibrium, the pseudo-code of the algorithm,
@@
 Regarding the second contribution, to validate the contact plans computed we introduced a complete framework for multi-contact motion synthesis.
-This framework additionally comprises an interpolation method to solve the problem P3, similar yet different than the one we proposed in [1].
-It also comprises an implementation of a state-of-the art dynamic simulator to check that the motions are physically plausible [28].
+This framework additionally comprises an interpolation method to solve the problem P3, similar but not identical to the one we proposed in [1].
+It also comprises an implementation of a state-of-the-art dynamic simulator to check that the motions are physically plausible [28].
 These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}.
-The novelty lies in the validation of the method with real robot models, for which the planning is harder with respect to the avatars used in [20], because of stronger kinematic constraints.
+The novelty lies in the validation of the method with real robot models, for which the planning is harder with respect to the avatars used in [20], because of stricter kinematic constraints.
 
 We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.
-To clarify these aspects in the paper, we rewrote the end of Section I.B to clearly state what is new and what is not in this paper.
+To clarify these aspects in the paper, we rewrote the end of Section I.B to clearly state what is new (and what is not) in this paper.
 
@@
 
-\textbf{Regarding the organization of the paper}, in the initial version of the paper, we tried to synthesize as much as possible
+\textbf{Regarding the organization of the paper}, in the initial version we tried to synthesize as much as possible
 the core of the paper, while providing the most technical details in the appendix. Based on the reviewers comments, we modified
@@
 in the paper, and also help him distinguishing between what is novel and what is not.
-As a result,  Appendices B.B (equilibrium criterion), D (pseudo code) and E (referernce to the source code) are now integrated in the core of the paper.
+As a result,  Appendices B.B (equilibrium criterion), D (pseudo code) and E (reference to the source code) are now integrated in the core of the paper.
 The manipulability-based heuristics used for contact generation, as well as our solution to the problem P3 remain in the Appendix as they are not novel.
@@
 
-Following the recommandation of the associate editor who ruled over some contradictory opinions of the reviewers, we made the paper self-contained,
-in such a way that the paper [20] is now obsolete regarding this submission, and is only mentioned in the introduction, or to describe an improvement
+Following the recommendation of the associate editor who ruled over some contradictory opinions of the reviewers, we made the paper self-contained,
+in such a way that the paper [20] is now obsolete with respect to this submission, and is only mentioned in the introduction, or to describe an improvement
 of the original algorithm. The missing information pointed out by the reviewers has been 
-included in the paper, as explained in the detailed answer that follows. As a result, we believe that in this new version the paper is self-contained and can be understood if read linearly.\\
+included in the paper, as explained in the detailed answers that follow. As a result, we believe that the new version of the paper is self-contained and can be understood if read linearly.\\
 
@@
 In the following, we list all the questions and remarks from the reviewers, answer to the questions, and detail the modifications that were brought to the paper.
-At the end of this letter, we also join a diff file between the original submission and the present one, that should help the reviewers visualizing the modifications 
+At the end of this letter, we also join a diff file between the original submission and the present one, that should help the reviewers see the modifications 
 brought to the paper.
@@
 To our knowledge, what is common in the literature is to reduce the friction coefficient to obtain more conservative bounds on the admissible forces, and thus a ``safety margin'' to the real cone boundary.
-An issue with this formulation is that the resulting margin may be too small for small forces, and too large for large contact forces.
-With our formulation, the obtained margin is constant, and thus guaranteed to be greater than the value $b_0$ we compute.
-We believe this formulation is new, but would be happy to cite any relevant reference if we happen to be wrong.
+An issue with this formulation is that the resulting margin may be too small for small forces, and too large for large forces.
+With our formulation instead, the margin is constant (i.e.
+the value $b_0$).
+We believe this formulation is novel, but would be happy to cite any relevant reference if we happen to be wrong.
 This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.}\done
@@
 In the new version of the paper, we do not consider that our solution to P3
-is a new contribution, although our formulation differs from [1] (our control variable is the acceleration
-of the center of mass, whereas in [1] the control variable is the contact forces). We provide it here for completeness, because it corresponds to our implementation, and 
+is a new contribution, although our formulation differs from [1] (our problem variables are the accelerations
+of the center of mass, whereas in [1] the problem variables are the contact forces). We provide it here for completeness, because it corresponds to our implementation, and 
 was required to validate our approach. 
@@
 
-Regarding the difference between our previous and current work, the postures computed in [20] were never validated in a simulator, nor the complete motions since 
+Regarding the difference between our previous and current work, the postures computed in [20] were not validated in a simulator, nor the complete motions since 
 we did not have a solution to P3 at the time. Then the main
@@
 
-\as{Thanks for the comments. We believe our new version of the papers answers to a). In particular, as mentionned above the novel content have been moved from the Appendix
-to the core of the paper. Regarding b), the take-away message is that to address the hard problem of multi-contact planning, it is necessary to use 
-abstraction and simplification, through the decomposition of the problem into three sub-problems, and through dimensionality reduction.
+\as{Thanks for the comments. We believe our new version of the papers answers to a). In particular, as mentioned above the novel content has been moved from the Appendix
+to the core of the paper. Regarding b), we updated the conclusion
+to include a take-away message. Our take-away message is that we can address the hard problem of multi-contact planning using 
+abstraction and simplification, i.e. the decomposition of the problem into three sub-problems, and  dimensionality reduction.
 We believe that the paper develops these aspects to explain the design of this planner. In the introduction, it is done
-with an extensive state of the art that explains how the community came up with the P1, P2, P3 decomposition to tackle a problem that is too hard. In the conclusion, it is explained 
-with an analysis of the results we obtain, that justify our pragmatic choice of going further in the abstraction to obtain interactive planning performances.
-To follow the reviewer's advice, we updated the conclusion
-to include a synthesized give away message.}
+with an extensive state of the art that explains how the community came up with the P1, P2, P3 decomposition to tackle a problem that is too hard otherwise. In the conclusion, it is explained 
+with an analysis of the results, which justify our pragmatic choice of going further in the abstraction to obtain interactive planning performance. }
 \qt{(Section X) In this paper we consider the multi-contact planning problem, formulated as three sub-problems  $\mathcal{P}_1$,  $\mathcal{P}_2$,  $\mathcal{P}_3$, addressed sequentially.
@@
 
-\as{We reorganized the paper such that the Table appear in the right order.}\ndone
+\as{We reorganized the paper such that the Table appears in the right place.}\ndone
 
@@
 \as{Thanks for the comment. On this particular point however, we wish to maintain our statement. For scenarios such as the car egress, a success rate of $68 \%$ means that the planner
-will fail at some point, but more importantly it means that if we restart the planning a few times with the same problem, there is a reasonable probability that it will find a feasible solution.
+will fail at some point, but more importantly it means that if we restart the planning a few times (e.g.
+3) with the same problem, there is a reasonable probability (e.g.
+97\%) that it will find a feasible solution.
 In this regard, the car egress scenario is effectively addressed, because we found a solution for it. \\
@@
 to compute a valid motion from scratch, even in simulation. The closest work to this achievement is the one from Escande et al., but the trajectories were manually interpolated
-after the contact plan was designed, and even so we do not have any idea on the success rate of the method. \\
+after the contact plan was designed, and moreover we cannot know the success rate of the method. \\
 Now regarding the heuristics, we guess that the reviewer is referring to the large number of repositionings that occurred in the contact plan. We believe that this issue
 is more related to the kinematic and environmental constraints than the heuristic for the choice of contacts. Developing heuristics that would take into account the environment to limit 
-contact repositioning is indeed out of the scope of this work.}
+contact repositioning is out of the scope of this work.}
 
@@
 The orientation of the end effector matters, but is not problematic regarding the way we generate contacts.
-Using an epsilon is an effective way to remove many unrelevant configurations, and
+Using an epsilon is an effective way to remove many irrelevant configurations, and
 we believe that more complex solutions aiming at taking into account the orientation would end up being counter productive.
@@
 the efficient 3D intersection query. Conversely, introducing a 6D octree rather than the current 3D one would make the problem much more complex: instead
-of using well known 3D operations to retrieve our candidate configurations, we would have to implement our own, new solutions. The added cost of doing this is probably 
-much higher than the computational cost of discarding a few configurations that could not be projected.  } \done
+of using well-known 3D operations to retrieve our candidate configurations, we would have to implement our own, new solutions. The added cost of doing this is probably 
+much higher than the computational cost of discarding a few configurations that could not be projected.  \addadp{Maybe add this discussion to the paper?}} 
+\done
 
@@
 \as{Thanks for the relevant comment.
-As observed by the reviewer, our state of the art voluntarily focuses on multi-contact planning related work, but it makes sense to cite the suggested papers.
+As observed by the reviewer, our state of the art voluntarily focuses on multi-contact planning, but it makes sense to cite the suggested papers.
 We added a paragraph at the end of the state of the art to discuss them.
@@
 The solutions proposed are designed for cyclic locomotion in quasi-flat scenarios, where the support polygon is a relevant method for equilibrium checks.
-They thus cannot be generalized to to multi contact locomotion.
+They thus cannot be generalized to multi contact locomotion.
 However, some contributions on specific parts of the problem could be applied directly in our case. 
@@
 We modified the text to explain this more clearly.
-As the Bi-RRT is a famous standard algorithm in the literature, we do not think it is necessary to include it in this paper, but we will gladly do so if required.
+As the Bi-RRT is a famous standard algorithm in the literature, we do not think it is necessary to explain it in this paper, but we will gladly do so if required.
 }
@@
 4 are blurred when they describe a previous state of the robots.
-We added this precision in the text.
-\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so the Fig.
+We added this information in the text.
+\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so Fig.
 6 has been removed.
@@
 
-\as{Thanks for the comment. The choice of not detailing too much the environment and octree description goes along with 
-our effort to maintain the paper as readable as possible, since it already contains various concepts. However,
+\as{Thanks for the comment. \deladp{The choice of not detailing too much the environment and octree description goes along with 
+our effort to maintain the paper as readable as possible, since it already contains various concepts. However,}
 we agree that additional information must be added. In Section III we now indicate that the environment is represented as a polygon soup (or 3D mesh), and
@@
 \as{That is a good observation. Although the integration of on-line sensing and state estimation in the planner is an issue currently considered by our team, it is indeed
-out of the scope of this paper. As indicated in the previous answer we indicated it in Section III. We assume that our planner could work similarly with 
+out of the scope of this paper. As reported in the previous answer we indicated it in Section III. We \deladp{assume}\addadp{think} that our planner could work similarly with 
 a sensing system that would output data easily interpreted as a polygon soup (as for a LIDAR), otherwise additional treatment would be required.
@@
 The suggestion of the reviewer to compare the simulation results with those on the real robot is interesting.
-However, our robot HRP-2 is not torque controlled, but position controlled, therefore as of today a comparison is not possible.}
+\addadp{However, in our simulations HRP-2 is torque controlled, whereas in our experiments HRP-2 is position controlled.
+This makes the usefulness of a comparison between simulation and experiments extremely doubtful, because we would be comparing motions obtained with different controllers.
+For these reasons, we prefer not to do this comparison.}}
 \qt{(Section VIII.A) To generate continuous movements from our contact plans we used either the framework proposed in [1], or our own implementation of a $\mathcal{P}_3$ solver (Appendix C).
@@
 novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].
-We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}
+\deladp{We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}}
 \qt{(Section I.B)  The present paper is an extension of our conference paper[20]. As such, the motion planning algorithm presented here is the same as the one presented in the
@@
 replaced by citations of [20] and [1] be removed).}
-\as{Thanks for the advice. This suggestion was in contradiction with the other reviewers, who thought that the paper should be self-contained.
+\as{Thanks for the advice.
+However, the suggestion of replacing some sections with references to [20] and [1] is in contradiction with the other reviewers, who thought that the paper should be self-contained.
 The associate editor decided that the paper must be self contained, so we maintained the indicated sections, but made clearer the fact that they are rephrasing of previous work. }\done
@@
 \as{We apologize for the confusion, that comes from an inaccurate use of English. What we meant by ``strong'' is that the hypothesis is constraining, because
-it defines a solution space different than the real one, and that we need to demonstrate that it makes sense.  We removed ``strong'' and replaced it by ``main'' to clarify. }\done
+it defines a solution space different from the real one, and that we need to demonstrate that it makes sense.  We removed ``strong'' and replaced it by ``main'' to clarify. }\done
 
@@
 conservative, but some clarification would be useful here.}
-\as{Indeed, the decimate tools removes vertices, and this was not explained clearly in the paper. We updated the text accordingly}
+\as{Indeed, the decimate tool removes vertices, and this was not explained clearly in the paper. We updated the text accordingly.}
 \qt{The resulting polytope can contain a very large number of faces.

### 2017-10-06 reviews_tro/reviewer_response.tex
added paragraph on 3d distance

@@
 % To print the list of change use: \listofchanges
-\usepackage{changes}	% use this for the working version
+%~ \usepackage{changes}	% use this for the working version
 %\usepackage[final]{changes} % use this for the final version
-\definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
-\newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
-\newcommand{\addadp}[1]{\added[id=adp]{#1}}
+%~ \definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
+\newcommand{\deladp}[1]{\textcolor{red}{#1}}
+\newcommand{\addadp}[1]{\textcolor{green}{#1}}
 %~ \newcommand{\repadp}[2]{\replaced[id=adp]{#1}{#2}}
@@
 \newcommand{\glslink}[2]{{#2}}
-\newcommand{\done}[0]{\textcolor{green}{DONE}}
+%~ \newcommand{\done}[0]{\textcolor{green}{DONE}}
+\newcommand{\done}[0]{}
 \newcommand{\ndone}[0]{\textcolor{red}{TODO}}
@@
 \as{
-The orientation of the end effector matters, but is not problematic regarding the way we generate contacts.
-Using an epsilon is an effective way to remove many irrelevant configurations, and
-we believe that more complex solutions aiming at taking into account the orientation would end up being counter productive.
- Indeed, using only 3D distance, we will end up with a few configurations
-that can't be projected correctly, and will end up being discarded. But the majority of irrelevant configurations would already have been rejected with
-the efficient 3D intersection query. Conversely, introducing a 6D octree rather than the current 3D one would make the problem much more complex: instead
-of using well-known 3D operations to retrieve our candidate configurations, we would have to implement our own, new solutions. The added cost of doing this is probably 
-much higher than the computational cost of discarding a few configurations that could not be projected.  \addadp{Maybe add this discussion to the paper?}} 
+We agree that the orientation of the end effector matters. Indeed, we could consider to further filter the set of candidates samples regarding orientation, to reduce the number of failed
+projections. However, because of the joint limits constraints, computing a relevant orientation distance is less trivial than in the 3D case. We tried to implement this
+in earlier versions of the planner, and did not obtain significant gains in the performance. We believe it is explained by the fact that the majority of unfeasible samples is already
+efficiently rejected by the simple ``epsilon distance'' check. We updated the text of the paper to explain this.
+}
+\qt{Because the distance computation $\epsilon$ does not account for the variation in orientation, several samples of $ C_{Contact}^{\epsilon}$  may turn out to be unfeasible at the time of projection.
+One could consider additionally filtering $ C_{Contact}^{\epsilon}$ based on the orientation with respect to the obstacle normal, but in our experience we did not notice
+any major improvement in the computational performances of the planner, so we do not perform this additional step. }
+%~ Using an epsilon is an effective way to remove many irrelevant configurations.  and
+%~ we believe that more complex solutions aiming at taking into account the orientation would end up being counter productive.
+ %~ Indeed, using only 3D distance, we will end up with a few configurations
+%~ that can't be projected correctly, and will end up being discarded. But the majority of irrelevant configurations would already have been rejected with
+%~ the efficient 3D intersection query. Introducing additional filtering . } 
 \done
@@
 Regarding the interpretation of the results without the video, we agree with the reviewer and believe that the accompanying video is essential to the comprehension of the results.
-We considered three options: a) displaying several sequences of contact (as currently done); b) only displaying 2 or 3 representative consecutive contact stances; c)
+We considered three options: a) displaying several sequences of contact (as currently done in most cases); b) only displaying 2 or 3 representative consecutive contact stances (currently done
+in the car scenario and in the real robot scenario, for which using a large number of pictures is not readable); c)
 not providing any picture and referring to the video.
-We chose a) because we wanted the paper to be remain as independent from the video as possible, but are willing to opt for option b) or c) (or any other option) if preferred by the reviewer.}\done
+We chose a mix between a) and b), because we wanted the paper to be remain as independent from the video as possible, but are willing to opt for any other option if preferred by the reviewer.}\done
 
@@
 
-\as{Thanks for the comment. \deladp{The choice of not detailing too much the environment and octree description goes along with 
-our effort to maintain the paper as readable as possible, since it already contains various concepts. However,}
-we agree that additional information must be added. In Section III we now indicate that the environment is represented as a polygon soup (or 3D mesh), and
+\as{Thanks for the comment. We agree that additional information must be added. In Section III we now indicate that the environment is represented as a polygon soup (or 3D mesh), and
 that we assume that it is fully known.  We thus also indicate that we assume the environment and state of the robot are fully known. To perform the intersection computation, we use
@@
 \as{That is a good observation. Although the integration of on-line sensing and state estimation in the planner is an issue currently considered by our team, it is indeed
-out of the scope of this paper. As reported in the previous answer we indicated it in Section III. We \deladp{assume}\addadp{think} that our planner could work similarly with 
+out of the scope of this paper. As reported in the previous answer we indicated it in Section III. We think that our planner could work similarly with 
 a sensing system that would output data easily interpreted as a polygon soup (as for a LIDAR), otherwise additional treatment would be required.
@@
 The suggestion of the reviewer to compare the simulation results with those on the real robot is interesting.
-\addadp{However, in our simulations HRP-2 is torque controlled, whereas in our experiments HRP-2 is position controlled.
+However, in our simulations HRP-2 is torque controlled, whereas in our experiments HRP-2 is position controlled.
 This makes the usefulness of a comparison between simulation and experiments extremely doubtful, because we would be comparing motions obtained with different controllers.
-For these reasons, we prefer not to do this comparison.}}
+For these reasons, we prefer not to do this comparison.}
 \qt{(Section VIII.A) To generate continuous movements from our contact plans we used either the framework proposed in [1], or our own implementation of a $\mathcal{P}_3$ solver (Appendix C).
@@
 for checking static equilibrium (see next question). The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation. The main
-novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].
-\deladp{We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}}
+novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].}
 \qt{(Section I.B)  The present paper is an extension of our conference paper[20]. As such, the motion planning algorithm presented here is the same as the one presented in the

### 2017-10-06 body/contact.tex
finished modifs on paper, now onto quest for typos
added paragraph on 3d distance

--- 2017-10-03 body/contact.tex
@@
 \centering
-  \begin{overpic}[width=0.9\linewidth]{figures/break_contact}
+  \begin{overpic}[width=1\linewidth]{figures/break_contact}
 		\put (0,4) {1} 
@@
 	\end{overpic}
-\caption{Contacts are maintained if joint limits and collisions constraints are respected (2). They are broken otherwise(3,4). The green line represents the root path.}
+\caption{Contacts are maintained if joint limits and collisions constraints are respected (2). They are broken otherwise(3,4). The green line represents the root path. The blurred character
+represents the previous contact configuration.}
 		   \label{fig:break_contact}
@@
 
-We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum distance 
+We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum 3D distance 
 between an effector and an obstacle is less than $\epsilon \in \mathbb{R}$.
@@
 %~ of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
+Because the distance computation $\epsilon$ does not account for the variation in orientation, several samples of $ C_{Contact}^{\epsilon}$  may turn out to be unfeasible at the time of projection.
+One could consider additionally filtering $ C_{Contact}^{\epsilon}$ based on the orientation with respect to the obstacle normal, but in our experience we did not notice
+any major improvement in the computational performances of the planner, so we do not perform this additional step.
+
 In all our experiments, the heuristic $h$ is implemented as a variation of a manipulability-based heuristic~\cite{Yoshikawa1984}. The manipulability is a real number that quantifies how 

### 2017-10-06 reviews_tro/reviewer_response.tex
typo

@@
 }
-\qt{Because the distance computation $\epsilon$ does not account for the variation in orientation, several samples of $ C_{Contact}^{\epsilon}$  may turn out to be unfeasible at the time of projection.
+\qt{Because the distance $\epsilon$ does not account for the variation in orientation, several samples of $ C_{Contact}^{\epsilon}$  may turn out to be unfeasible at the time of projection.
 One could consider additionally filtering $ C_{Contact}^{\epsilon}$ based on the orientation with respect to the obstacle normal, but in our experience we did not notice

### 2017-10-06 body/contact.tex
typo

@@
 %~ of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
-Because the distance computation $\epsilon$ does not account for the variation in orientation, several samples of $ C_{Contact}^{\epsilon}$  may turn out to be unfeasible at the time of projection.
+Because the distance $\epsilon$ does not account for the variation in orientation, several samples of $ C_{Contact}^{\epsilon}$  may turn out to be unfeasible at the time of projection.
 One could consider additionally filtering $ C_{Contact}^{\epsilon}$ based on the orientation with respect to the obstacle normal, but in our experience we did not notice

### 2017-10-09 reviews_tro/reviewer_response.tex
correct table order

@@
 
-\as{We reorganized the paper such that the Table appears in the right place.}\ndone
+\as{We reorganized the paper such that the Tables appears in the right place, by inverting the Computation times and success rates paragraphs. The order of appearance of the Tables thus remains
+the same, but the numbering is correct, and in agreement with the order of referencement in the text.}\done
 

### 2017-10-09 body/results.tex
correct table order

@@
 \subsubsection{HRP-2 -- Staircase with high steps (Figure~\ref{fig:stair_robust})}
+\newcommand{\widthValue}{0.13\linewidth}
 
-\begin{figure}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/stair}
+\begin{figure*}[h!t!]
+  \begin{center}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing1.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing2.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing3.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing4.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing5.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing6.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing7.jpeg}
   \caption{
@@
 		   \label{fig:stair_robust}
-\end{figure}
+  \label{fig:moviepicture2}
+  \end{center}
+  \vspace{-0.5cm}
+\end{figure*}
+
+%~ \begin{figure}
+  %~ \centering
+  %~ \includegraphics[width=1\linewidth]{figures/stair}
+  %~ \caption{
+           %~ HRP-2 in the stair climbing scenario. }
+		   %~ \label{fig:stair_robust}
+%~ \end{figure}
 
@@
 
-\begin{table*}
-\centering
-\footnotesize
-\begin{tabular}{ >{\centering\arraybackslash}m{37pt} | >{\centering\arraybackslash}m{57pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{70pt} | >{\centering\arraybackslash}m{73pt} | >{\centering\arraybackslash}m{80pt} | >{\centering\arraybackslash}m{10pt}}
-  Scenario (nb steps) &  Complete guide generation (ms) & Static equilibrium (ms) & Collision (ms) & Inverse Kinematics (ms) & Total generation time (ms) & Time per step (ms)\\
- \hline
-   Stairs (18) 	& 5 -- \textbf{6} --  18 		 & 13 --  \textbf{32} -- 329   	& 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{15} \\
-   Standing (24)& 65 -- \textbf{1086} --  5227   & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 371 --  \textbf{2257} -- 7671 & \textbf{94}  \\
-   Car (86)& 320 -- \textbf{6971} --  44002 & 409 --  \textbf{1766} -- 14752   	& 297 -- \textbf{1187} -- 8483 & 3154 --  \textbf{15323} -- 165541 & 5834 --  \textbf{31391} -- 281000 & \textbf{365}\\
-   Rubble (82)& 37 -- \textbf{573} --  1685 & 583 --  \textbf{2714} -- 9459 & 491 --  \textbf{1971} -- 6273 & 269 --  \textbf{706} -- 3118 & 1811 --  \textbf{7195} -- 23241 & \textbf{86} \\
-   Race (134)& 14 -- \textbf{51} --  125 & 455 --  \textbf{1359} -- 21045   & 397 --  \textbf{923} -- 9924 & 228 --  \textbf{471} -- 5415 & 1436 --  \textbf{3343} -- 41446 & \textbf{25}
- \end{tabular}
-\caption{minimum, \textbf{average} and worst time (in ms) spent in the generation process for each scenario and each critical part of the generation process (not all parts are timed,
-thus the average total computation time is higher than the sum of each part). The last
-column indicates the average time necessary to compute one contact transition.
-The Collision column times includes the (negligeable) octree intersection operation necessary to retrieve the candidate samples.}
-\label{tab:requestime}
-\quad
-\end{table*}
-
 \begin{table}
@@
 
+\begin{table*}[b!t!p!]
+\centering
+\footnotesize
+\begin{tabular}{ >{\centering\arraybackslash}m{37pt} | >{\centering\arraybackslash}m{57pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{70pt} | >{\centering\arraybackslash}m{73pt} | >{\centering\arraybackslash}m{80pt} | >{\centering\arraybackslash}m{10pt}}
+  Scenario (nb steps) &  Complete guide generation (ms) & Static equilibrium (ms) & Collision (ms) & Inverse Kinematics (ms) & Total generation time (ms) & Time per step (ms)\\
+ \hline
+   Stairs (18) 	& 5 -- \textbf{6} --  18 		 & 13 --  \textbf{32} -- 329   	& 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{15} \\
+   Standing (24)& 65 -- \textbf{1086} --  5227   & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 371 --  \textbf{2257} -- 7671 & \textbf{94}  \\
+   Car (86)& 320 -- \textbf{6971} --  44002 & 409 --  \textbf{1766} -- 14752   	& 297 -- \textbf{1187} -- 8483 & 3154 --  \textbf{15323} -- 165541 & 5834 --  \textbf{31391} -- 281000 & \textbf{365}\\
+   Rubble (82)& 37 -- \textbf{573} --  1685 & 583 --  \textbf{2714} -- 9459 & 491 --  \textbf{1971} -- 6273 & 269 --  \textbf{706} -- 3118 & 1811 --  \textbf{7195} -- 23241 & \textbf{86} \\
+   Race (134)& 14 -- \textbf{51} --  125 & 455 --  \textbf{1359} -- 21045   & 397 --  \textbf{923} -- 9924 & 228 --  \textbf{471} -- 5415 & 1436 --  \textbf{3343} -- 41446 & \textbf{25}
+ \end{tabular}
+\caption{minimum, \textbf{average} and worst time (in ms) spent in the generation process for each scenario and each critical part of the generation process (not all parts are timed,
+thus the average total computation time is higher than the sum of each part). The last
+column indicates the average time necessary to compute one contact transition.
+The Collision column times includes the (negligeable) octree intersection operation necessary to retrieve the candidate samples.}
+\label{tab:requestime}
+\quad
+\end{table*}
+
+\subsubsection{Success rates (Table~\ref{tab:sucess_planning})}
+%~ Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
+Despite the complexity of the scenarios and the approximations made in our formulation, our planner succeeded in the large majority of cases.
+
+Table~\ref{tab:requestpercent} presents the rate of successful contact generation.
+Note that a failure in contact generation for a root configuration is not equivalent to a failure in the contact plan.
+It simply means that another limb was tested for contact generation for the same root configuration.
+As expected, a more constrained scenario such as the car egress provides less satisfying results, despite the high success rate of the planner.
+
 \subsubsection{Computation times (Table~\ref{tab:requestime})}
@@
 
-\subsubsection{Success rates (Table~\ref{tab:sucess_planning})}
-%~ Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
-Despite the complexity of the scenarios and the approximations made in our formulation, our planner succeeded in the large majority of cases.
-
-Table~\ref{tab:requestpercent} presents the rate of successful contact generation.
-Note that a failure in contact generation for a root configuration is not equivalent to a failure in the contact plan.
-It simply means that another limb was tested for contact generation for the same root configuration.
-As expected, a more constrained scenario such as the car egress provides less satisfying results, despite the high success rate of the planner.
-
+\subsubsection*{Conclusion}
 These results confirm that our approach provides a satisfying compromise between completeness and efficiency, thus enabling online planning

### 2017-10-09 reviews_tro/reviewer_response.tex
typo in optim

@@
 
+\bigskip
+We now join a diff file comparing the original and current submissions.
+The text appearing in \textcolor{blue} denotes the additions. We used the latexdiff tools to generate this file, which is not fully compatible with the latex math package.
+For this reason, errors appear in the document, in particular regarding the subsections number. However, we think that this diff file will help the reviewers locate precisely
+where the modifications occured in the document. The reviewers will also notice that we reformulated the description our P3 solver in Appendix C, to better clarify what variables
+are used, and what variables are implicit (that is computed from the others).
+
 \end{document}

### 2017-10-09 body/appendix_optim.tex
typo in optim

@@
 The cost function $\ell$ is a weighted sum of the angular momentum and center-of-mass acceleration variation over the whole trajectory.
-The center-of-mass positions and velocities $\mathbf{c}(t)$, $\mathbf{\dot{c}}(t)$ are interval variables, obtained through the double integration of $\mathbf{\ddot{c}}(t)$.
+The center-of-mass positions and velocities $\mathbf{c}(t)$, $\mathbf{\dot{c}}(t)$ are internal variables, obtained through the double integration of $\mathbf{\ddot{c}}(t)$.
 Then $\mathbf{w}(t)$ is obtained directly from these variables. $\mathbf{c}_{\mathbf{q}_0}$ and $\mathbf{c}_{\mathbf{q}_1}$ are the center-of-mass positions for configurations

### 2017-10-09 reviews_tro/reviewer_response.tex
final version

--- 2017-10-09 reviews_tro/reviewer_response.tex
@@
 
+\section{Diff files between the original and current submission}
+We now join a diff file comparing the original and current submissions.
+The text appearing in blue denotes the additions. We used the latexdiff tools to generate this file, which is not fully compatible with all the latex packages that we use.
+For this reason, errors appear in the document, in particular regarding the subsections numbering. However, we think that this diff file will help the reviewers locate precisely
+where the modifications occured in the document. The reviewers will also notice that we reformulated the description our P3 solver in Appendix C, to better clarify what variables
+are used, and what variables are implicit (that is computed from the others).
+
+\includepdf[pages={-}]{diff.pdf}
+
 \end{document}

### 2017-10-09 reviews_tro/reviewer_response.tex
all done
rewrote I.B more like in the letter

@@
 \section{Answers to reviewer 1}
-\ndone reecriture p3 + enlever main video
+%~ \ndone reecriture p3 + enlever main video
 
@@
 This is now explained in details in the introduction.}
-\qt{(Section I.B) The present paper is an extension of our conference paper[20]. As such, the motion planning algorithm presented here is the same as the one presented in the
-previous work.
-Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
-We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section VI.
-Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
-all the technical information required to use the planner.\\
-To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
-We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work[1].
-The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
-Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
-We do not consider these additions as contributions per se, but describing them here is required to understand the experimental contribution of this paper.\\
-Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
-The novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
-As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are much more constrained kinematically than the virtual avatars used before,
-thus making the planning problem even harder. Through an extensive empirical analysis, we demonstrate that the method remains interesting in this context, qualitatively and efficiently.}
+\qt{(Section I.B) The present paper is an extension of our ISRR conference paper [20]. 
+As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section IV and V). 
+However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section V.B.2), a novel criterion for
+static equilibrium (Section VI), and the release of the source code of the planner (Section VII). \\
+The other contribution of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
+To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
+This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1].
+Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
+We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution. \\
+Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
+thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.}
 \done
@@
 of the paper because it is not required to understand the contributions to P1 and P2.}
-\qt{(Section I.B) We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work[1].
-The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
-Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
-We do not consider these additions as contributions per se, but describing them here is required to understand the experimental contribution of this paper.}\done
+\qt{(Section I.B) To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
+This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1].
+Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
+We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution.}\done
 
@@
 novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].}
-\qt{(Section I.B)  The present paper is an extension of our conference paper[20]. As such, the motion planning algorithm presented here is the same as the one presented in the
-previous work.
-Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
-We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section VI.
-Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
-all the technical information required to use the planner.\\
-To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
-We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work[1].
-The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
-Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
-We do not consider these additions as contributions per se, but describing them here is required to understand the experimental contribution of this paper.\\
-Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
-The novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
-As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are much more constrained kinematically than the virtual avatars used before,
-thus making the planning problem even harder. Through an extensive empirical analysis, we demonstrate that the method remains interesting in this context, qualitatively and efficiently.}\done
+\qt{(Section I.B) 
+The present paper is an extension of our ISRR conference paper [20]. 
+As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section IV and V). 
+However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section V.B.2), a novel criterion for
+static equilibrium (Section VI), and the release of the source code of the planner (Section VII). \\
+The other contribution of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
+To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
+This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1].
+Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
+We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution. \\
+Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
+thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.}\done
 

### 2017-10-09 body/intro.tex
Review response to reviewer
rewrote I.B more like in the letter

@@
 The solutions proposed are designed for cyclic locomotion in quasi-flat scenarios, where the support polygon is a relevant method for equilibrium checks.
-They thus cannot be generalized to to multi contact locomotion.
+They thus cannot be generalized to multi contact locomotion.
 %For instance, the choice of the next effector in contact is often deterministic, the contacts are either coplanar or punctual, or the generation of the end-effector trajectory only works for ``ground'' obstacles, as it assumes that increasing the $z$ position of an effector drives it further away from obstacles~\cite{Kolter:2009:TTV:1703775.1703833}.
@@
 \noindent \textbf{Comparison with our previous work:}
-The present paper is an extension of our conference paper~\cite{tonneauisrr15}. As such, the motion planning algorithm presented here is the same as the one presented in the
-previous work.
+The present paper is an extension of our ISRR conference paper~\cite{tonneauisrr15}. 
+As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section~\ref{rbprm} and~\ref{sec:contact}). 
+However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section~\ref{app:contact}), a novel criterion for
+static equilibrium (Section~\ref{sec:Equil}), and the release of the source code of the planner (Section~\ref{app:hpp}).
+
+The other contribution of the paper is a rigorous experimental validation of the approach on actual robot models (Section~\ref{sec:results}).
+To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
+This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work~\cite{Carpentier2016}.
+Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
+We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution.
+
+Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
+thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.
+%~ These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}.
+The novelty lies in the validation of the method with real robot models, for which the planning is harder with respect to the avatars used in [20], because of stricter kinematic constraints
+
+%~ As such, the motion planning algorithm presented here is the same as the one presented in the
+%~ previous work.
 Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper, with the exception of Section~\ref{app:contact}, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
 We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section~\ref{sec:Equil}.
-Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
-all the technical information required to use the planner.
-
-To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
+%~ Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
+%~ all the technical information required to use the planner.
+%~ 
+%~ To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
 We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work~\cite{Carpentier2016}.
@@
 Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
-We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand the experimental contribution of this paper.
+%~ We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand the experimental contribution of this paper.
 
-Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
-The novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
-As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are much more constrained kinematically than the virtual avatars used before,
-thus making the planning problem even harder. Through an extensive empirical analysis, we demonstrate that the method remains interesting in this context, qualitatively and efficiently.
-
-%~ This paper extends our previous conference paper~\citep{tonneauisrr15} in several ways.
-Our approach is now demonstrated on actual robots and in simulation, contrary to virtual avatars before, in a large variety of contexts.
-To achieve this we introduce a new criterion to efficiently compute the equilibrium of the robot while accounting for uncertainties of the real world.
-We provide the complete pseudo code of algorithm.
-We give the complete procedure to use our planner with any legged robot and provide our source code, allowing any user to directly use our method.
-We also propose a solution to $\mathcal{P}_3$, making our framework complete.
-Finally we provide an in-depth analysis of the results we obtain, and a theoretical discussion on the capabilities of the planner and its current limitations.
+%~ Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
+%~ Indeed, the novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
+%~ As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
+%~ thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method is remains efficient and qualitatively relevant in this context.

### 2017-10-09 body/conclusion.tex
final changes from rev 1

@@
 % !TEX root =  ../main_tro.tex
- \section{Discussion: validity and interest of our contact planner} 
+ \section{Discussion: validity and purpose of our contact planner} 
 \label{sec:discussion}
@@
 
-As demonstrated in the results section, the main interest of our method is the reduction of the algorithmic complexity of the problem, which leads to an interactive 
+As demonstrated in the results section, the main purpose of our method is the reduction of the algorithmic complexity of the problem, which leads to an interactive 
 application. This property is critical for 

### 2017-10-09 body/equilibrium.tex
final changes from rev 1

@@
 According to Coulomb's law, the non-slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
-As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays of unit length. We choose for instance:
+As classically done, we linearize the friction cone in a conservative fashion with a pyramid included in it, described by four generating rays of unit length. We choose for instance:
 \begin{equation*}

### 2017-10-09 reviews_tro/reviewer_response.tex
typos
nicolas comments

@@
 
-First of all, we would like to thank all the reviewers and the associate editor (AE).
+First of all, we would like to thank all the Reviewers and the associate editor (AE).
 We received three detailed and technical reviews that helped improving our paper.
 Overall we agreed with most of the reviews and took them into account in the revised version of the paper.
-The main points of concern raised by the reviewers regarded:
+The main points of concern raised by the Reviewers regarded:
 \begin{itemize}
@@
 \textbf{Regarding the organization of the paper}, in the initial version we tried to synthesize as much as possible
-the core of the paper, while providing the most technical details in the appendix. Based on the reviewers comments, we modified
+the core of the paper, while providing the most technical details in the appendix. Based on the Reviewers comments, we modified
 the paper architecture in such a way that all the novel details are now directly included in the core of the paper. This should avoid the reader going back and forth
@@
 The manipulability-based heuristics used for contact generation, as well as our solution to the problem P3 remain in the Appendix as they are not novel.
-However, as suggested by the reviewers, they are described in more details in the core.
+However, as suggested by the Reviewers, they are described in more details in the core.
 
-Following the recommendation of the associate editor who ruled over some contradictory opinions of the reviewers, we made the paper self-contained,
+Following the recommendation of the associate editor who ruled over some contradictory opinions of the Reviewers, we made the paper self-contained,
 in such a way that the paper [20] is now obsolete with respect to this submission, and is only mentioned in the introduction, or to describe an improvement
-of the original algorithm. The missing information pointed out by the reviewers has been 
+of the original algorithm. The missing information pointed out by the Reviewers has been 
 included in the paper, as explained in the detailed answers that follow. As a result, we believe that the new version of the paper is self-contained and can be understood if read linearly.\\
@@
  %~ In order to try to make the paper more readable, the first version was mixing three styles: details given in the core of the paper; details given in the appendices; details deferred in cited papers.
-While all reviewers agreed that using all three styles together was a bad idea, we received contradictory reviews about how to modify the paper.
+While all Reviewers agreed that using all three styles together was a bad idea, we received contradictory reviews about how to modify the paper.
 Following the recommandation of the AE, we have worked to make the paper self-contained with all the contributions in the core of the paper.
@@
 
-We thus believe to have significantly improved the paper by addressing the majority of the comments and questions raised by the reviewers.
+We thus believe to have significantly improved the paper by addressing the majority of the comments and questions raised by the Reviewers.
 
@@
 
-In the following, we list all the questions and remarks from the reviewers, answer to the questions, and detail the modifications that were brought to the paper.
-At the end of this letter, we also join a diff file between the original submission and the present one, that should help the reviewers see the modifications 
+In the following, we list all the questions and remarks from the Reviewers, answer to the questions, and detail the modifications that were brought to the paper.
+At the end of this letter, we also join a diff file between the original submission and the present one, that should help the Reviewers see the modifications 
 brought to the paper.
 
-\section{Answers to reviewer 1}
+\section{Answers to Reviewer 1}
 %~ \ndone reecriture p3 + enlever main video
@@
 for checking static equilibrium.
-The main novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].
+The main novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robot models that are kinematically more constrained than the virtual avatars from [20].
 This is now explained in details in the introduction.}
@@
 static equilibrium (Section VI), and the release of the source code of the planner (Section VII). \\
-The other contribution of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
+The other novelty of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
 To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
@@
 the value $b_0$).
+We updated the paper to explain this difference.
 We believe this formulation is novel, but would be happy to cite any relevant reference if we happen to be wrong.
-This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.}\done
+This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.
+}\done
+\qt{(Section VI.2) A simple approach to robustness consists in choosing a smaller friction coefficient, to constrain the forces to lie away from the boundaries of the real cone.
+However, this would result in a small safety margin for forces of low magnitude, and an excessively big safety margin for large forces as the boundaries grow more and more appart.
+In comparison, our margin $b_0$ is constant, and provides a helpful mean to compare the robustness of different contact configurations.}
 
@@
 If a sampled configuration is in $C_{Reach}^0$, but our method is unable to generate a contact configuration from it, as a result the sensitivity
-decreases. The sensitivity thus illustrates the confidence we have that any configuration in $C_{Reach}^0$ will effectively lead to a contact configuration.
+decreases. \textcolor{green}{The sensitivity thus illustrates the confidence we have that any configuration in $C_{Reach}^0$ will effectively lead to a contact configuration.}
 Conversely, the specificity refers to the percentage of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}.
@@
 decreases.
-The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to $C_{Reach}^0$ (or informally, the confidence that we are not missing valid solutions).
+\textcolor{green}{The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to $C_{Reach}^0$ (or informally, the confidence that we are not missing valid solutions).}
 We thus look for a compromise between sensitivity and specificity.} \done
@@
 
-\as{As also suggested by reviewer 4, we removed the references to the results of the previous work [20], including those.}\done
+\as{As also suggested by Reviewer 4, we removed the references to the results of the previous work [20], including those.}\done
 
@@
 after the contact plan was designed, and moreover we cannot know the success rate of the method. \\
-Now regarding the heuristics, we guess that the reviewer is referring to the large number of repositionings that occurred in the contact plan. We believe that this issue
+Now regarding the heuristics, we guess that the Reviewer is referring to the large number of repositionings that occurred in the contact plan. We believe that this issue
 is more related to the kinematic and environmental constraints than the heuristic for the choice of contacts. Developing heuristics that would take into account the environment to limit 
@@
 \as{We are suprised by this comment because after testing it, the link seems to be valid.
-In any case, in accordance with reviewer 6, we changed all the links source codes to provide a SHA link that will bring the user to the accurate revision of the projects considered.}
+In any case, in accordance with Reviewer 6, we changed all the links source codes to provide a SHA link that will bring the user to the accurate revision of the projects considered.}
 \qt{(Section VII) Both HPP and our planner can be simply downloaded and compiled by following the instructions on
@@
 \as{
-We agree that the orientation of the end effector matters. Indeed, we could consider to further filter the set of candidates samples regarding orientation, to reduce the number of failed
+We agree that the orientation of the end effector matters, but in our implementation considering it for filtering did not improve our approach.
+Indeed, we could consider to further filter the set of candidates samples regarding orientation, to reduce the number of failed
 projections. However, because of the joint limits constraints, computing a relevant orientation distance is less trivial than in the 3D case. We tried to implement this
@@
 One could consider additionally filtering $ C_{Contact}^{\epsilon}$ based on the orientation with respect to the obstacle normal, but in our experience we did not notice
-any major improvement in the computational performances of the planner, so we do not perform this additional step. }
+any significant improvement in the computational performances of the planner, so we do not perform this additional step. }
 %~ Using an epsilon is an effective way to remove many irrelevant configurations.  and
@@
 
-\quot{remaining remarks by reviewer 1}
-\as{The remaining comments from reviewer 1 were all addressed. Because they did not call for a discussion, as they are essentially related to style or typos, we do not list them here.}\done
+\quot{remaining remarks by Reviewer 1}
+\as{We thank the Reviewer for the consequent amount of remarks brought to the paper.
+The remaining comments that are not described here were all addressed, following the suggestions of the reviewer.
+Because they did not call for a discussion nor interpretation, as they are essentially related to style or typos mistakes, we do not list them here.}\done
 
@@
 
-\section{Answers to reviewer 4}
+\section{Answers to Reviewer 4}
 
@@
 \as{Thanks for the relevant comment.
-As observed by the reviewer, our state of the art voluntarily focuses on multi-contact planning, but it makes sense to cite the suggested papers.
+As observed by the Reviewer, our state of the art voluntarily focuses on multi-contact planning, but it makes sense to cite the suggested papers in the revision.
 We added a paragraph at the end of the state of the art to discuss them.
@@
 }
-\as{Thanks for the comment. As it is also suggested by reviewer 1 and the AE, we modified the paper to make it self contained. The references to previous work now solely indicate
+\as{Thanks for the comment. As it is also suggested by Reviewer 1 and the AE, we modified the paper to make it self contained. The references to previous work now solely indicate
 that a certain method is not a contribution of the present paper. \\ 
@@
 We added this information in the text.
-\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so Fig.
+\\ As proposed also by Reviewer 1, we removed references to previous results obtained in [20], so Fig.
 6 has been removed.
-\\Regarding the quality of our Figures,  our simulation Figures are taken from high resolution rendering in blender.
-The other Figures are drawn from manually designed powerpoint vectorial graphics (converted to png).
+\\Regarding the quality of our figures,  our simulation figures are taken from high resolution rendering in blender.
+The other figures are drawn from manually designed powerpoint vectorial graphics (converted to png).
 We assume that the issue encountered comes from the compression of the pdf file to fit the size constraints of the
-submission site. We tried to improve this compression, but if the paper is accepted, we believe that the uncompressed pdf obtained from the source of the paper will result in good quality pictures.
-Regarding the interpretation of the results without the video, we agree with the reviewer and believe that the accompanying video is essential to the comprehension of the results.
+submission site. We tried to improve this compression in this new submission. In any case if the paper is accepted, we believe that the uncompressed pdf could be used to obtain the desired quality.
+Regarding the interpretation of the results without the video, we agree with the Reviewer and believe that the accompanying video is essential to the comprehension of the results.
 We considered three options: a) displaying several sequences of contact (as currently done in most cases); b) only displaying 2 or 3 representative consecutive contact stances (currently done
@@
 not providing any picture and referring to the video.
-We chose a mix between a) and b), because we wanted the paper to be remain as independent from the video as possible, but are willing to opt for any other option if preferred by the reviewer.}\done
+We chose a mix between a) and b), because we wanted the paper to be remain as independent from the video as possible, but are willing to opt for any other option if preferred by the Reviewer.}\done
 
@@
 %~ \as{ThAll of our simulation figures are taken from high resolution rendering in blender, so we wouldn't know how to improve their quality.
-If the reviewers think the paper would be better without these figures, we are willing to remove them.
+If the Reviewers think the paper would be better without these figures, we are willing to remove them.
 However, we believe that they would make the paper more dependent on the video, so we would prefer to leave them.}
@@
 illustrations should be included in the paper. now they are only in the
-accompanying video.  also a comparison between the results of the same
+accompanying video.}
+\as{We agree with the remark, and included some illustrations of the experimental results on HRP-2 in the paper.}
+%~ We now elaborate on the validity of the simulation with respect to an experiment on the real robot.
+
+\done
+
+ \quot{also a comparison between the results of the same
 task(s) executed in dynamic simulation and on the real robot should
@@
 not depend too much on the idealized simulation environment. }
-\as{We agree with the remark, and included some illustrations of the experimental results on HRP-2 in the paper. \\
-We now elaborate on the validity of the simulation with respect to an experiment on the real robot. As also suggested by reviewer 1, we added a paragraph
+\as{As also suggested by Reviewer 1, we added a paragraph
 in the paper (Section VIII.A)  to provide more information on the simulation we use to validate our motions.
@@
 Of course, it does not guarantee that the motion could be executed on the robot, but it is used extensively by the community as a reasonable feasibility test.
-The suggestion of the reviewer to compare the simulation results with those on the real robot is interesting.
+The suggestion of the Reviewer to compare the simulation results with those on the real robot is interesting.
 However, in our simulations HRP-2 is torque controlled, whereas in our experiments HRP-2 is position controlled.
 This makes the usefulness of a comparison between simulation and experiments extremely doubtful, because we would be comparing motions obtained with different controllers.
-For these reasons, we prefer not to do this comparison.}
+For these reasons, we prefer not to do this comparison, although we plan to do it in the future, since we are currently working on making HRP-2 torque controlled.}
 \qt{(Section VIII.A) To generate continuous movements from our contact plans we used either the framework proposed in [1], or our own implementation of a $\mathcal{P}_3$ solver (Appendix C).
@@
 The companion video shows the obtained motions.}
-\done
 
-\section{Answers to reviewer 6}
+\section{Answers to Reviewer 6}
 \quot{ My primary concern, however,
@@
 \as{Thanks for the comment.
-Indeed, all reviewers agreed that the contribution with respect to previous work was not clearly stated.
-We updated the paper to reflect the following.
-Firstly, our planner is almost identical to the one in [20], with important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
-for checking static equilibrium (see next question). The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation. The main
-novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].}
-\qt{(Section I.B) 
+Indeed, the Reviewers agreed that the contribution with respect to previous work was not clearly stated.
+Here we reproduce the answer we made to Reviewer I: As mentioned in the summary, our planner is almost identical to the one in [20], but the paper comprises important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
+for checking static equilibrium.
+The main novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robot models that are kinematically more constrained than the virtual avatars from [20].
+This is now explained in details in the introduction.}
+\qt{(Section I.B, repeated from page 5) 
 The present paper is an extension of our ISRR conference paper [20]. 
@@
 \as{Thanks for the advice.
-However, the suggestion of replacing some sections with references to [20] and [1] is in contradiction with the other reviewers, who thought that the paper should be self-contained.
+However, the suggestion of replacing some sections with references to [20] and [1] is in contradiction with the other Reviewers, who thought that the paper should be self-contained.
 The associate editor decided that the paper must be self contained, so we maintained the indicated sections, but made clearer the fact that they are rephrasing of previous work. }\done
@@
 that approaches such as the CIO, if not initialized with an appropriate initial guess, might never get out of the local minima.
-We followed the reviewer's advice and simply removed the sentence.
+We followed the Reviewer's advice and simply removed the sentence.
 }\done
@@
 to better explain our claim.
-Indeed, as the reviewer states $B^*$  is the volume for every $q_0$.
+Indeed, as the Reviewer states $B^*$  is the volume for every $q_0$.
 If such a volume existed, we would know exactly whether or not a given root configuration can result in a valid contact configuration.
-We of course doubt that $B^*$ actually exists. }
+We of course doubt that the closed form representation of $B^*$ actually exists. }
 \qt{The ideal shape $B^*, W^0 \subset B^* \subset B^{\textrm{\it Suf}}$ would define a necessary \textbf{and} sufficient condition for contact creation. 
@@
 eventually delete the specific code I was citing. }
-\as{The reviewer is absolutely right. We indeed encounter the issue with other researchers that are currently using our planner. 
+\as{The Reviewer is absolutely right. We indeed encounter the issue with other researchers that are currently using our planner. 
 To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project, on a separate branch
 dedicated to the installation of the planner.
-Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the reviewer.}
+Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the Reviewer.}
 \qt{(Section VII) Both HPP and our planner can be simply downloaded and compiled by following the instructions on
@@
 The text appearing in blue denotes the additions. We used the latexdiff tools to generate this file, which is not fully compatible with all the latex packages that we use.
-For this reason, errors appear in the document, in particular regarding the subsections numbering. However, we think that this diff file will help the reviewers locate precisely
-where the modifications occured in the document. The reviewers will also notice that we reformulated the description our P3 solver in Appendix C, to better clarify what variables
+For this reason, errors appear in the document, in particular regarding the subsections numbering. However, we think that this diff file will help the Reviewers locate precisely
+where the modifications occured in the document. The Reviewers will also notice that we reformulated the description our P3 solver in Appendix C, to better clarify what variables
 are used, and what variables are implicit (that is computed from the others).

### 2017-10-09 body/contact.tex
nicolas comments

@@
 One could consider additionally filtering $ C_{Contact}^{\epsilon}$ based on the orientation with respect to the obstacle normal, but in our experience we did not notice
-any major improvement in the computational performances of the planner, so we do not perform this additional step.
+any significant improvement in the computational performances of the planner, so we do not perform this additional step.
 

### 2017-10-09 body/equilibrium.tex
equality between dual and primal optimum
nicolas comments

@@
 If $b_0$ is negative, the configuration is not in static equilibrium, and $b_0$ indicates ``how far'' from equilibrium the configuration is. We thus use $b_0$ as a measure of robustness.
+%~ One advantage of $b_0$ is that it gives a constant margin 
+A simple approach to robustness consists in choosing a smaller friction coefficient, to constrain the forces to lie away from the boundaries of the real cone.
+However, this would result in a small safety margin for forces of low magnitude, and an excessively big safety margin for large forces as the boundaries grow more and more appart.
+In comparison, our margin $b_0$ is constant, and provides a helpful mean to compare the robustness of different contact configurations.
 
@@
 \find \quad & \bm{\nu} \in \Rv{6}\\
-\maximize  \quad & -(\mathbf{D} \mathbf{c} + \mathbf{d})^T \nu \\
+\maximize  \quad & -(\mathbf{D} \mathbf{c} + \mathbf{d})^T \bm{\nu} \\
 \st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
@@
 Indeed, from Slater's conditions \citep{Boyd:2004:CO:993483}, we know that the optimal values of an LP and its dual are equal.
-Thus the optimal value of the LP~\eqref{eq:dual} is indeed the optimal $b_0$.
+Therefore the optimal value $\bm{\nu}^*$ gives the optimal value $b_0^*$ through the equality $b_0^* = (\mathbf{D} \mathbf{c} + \mathbf{d})^T \bm{\nu}^*$.
+ %~ Thus the optimal value of the LP~\eqref{eq:dual} is indeed the optimal $b_0$.

### 2017-10-09 body/equilibrium.tex
[review] Minor

@@
-% !TEX root =  ../main.tex
+% !TEX root =  ../main_tro.tex
 
@@
 A simple approach to robustness consists in choosing a smaller friction coefficient, to constrain the forces to lie away from the boundaries of the real cone.
-However, this would result in a small safety margin for forces of low magnitude, and an excessively big safety margin for large forces as the boundaries grow more and more appart.
+However, this would result in a small safety margin for forces of low magnitude, and an excessively large safety margin for large forces as the boundaries grow more and more apart.
 In comparison, our margin $b_0$ is constant, and provides a helpful mean to compare the robustness of different contact configurations.

### 2017-10-10 reviews_tro/reviewer_response.tex
andrea s mofis repercuted on letter

@@
 \qt{(Section VI.2) A simple approach to robustness consists in choosing a smaller friction coefficient, to constrain the forces to lie away from the boundaries of the real cone.
-However, this would result in a small safety margin for forces of low magnitude, and an excessively big safety margin for large forces as the boundaries grow more and more appart.
+However, this would result in a small safety margin for forces of low magnitude, and an excessively large safety margin for large forces as the boundaries grow more and more apart.
 In comparison, our margin $b_0$ is constant, and provides a helpful mean to compare the robustness of different contact configurations.}

### 2017-10-10 body/intro.tex
typos
Modif contribs.

@@
 movement.
-Our
+\textbf{Our
 main
@@
 a reduction property, the reachability condition.
+}
+%~ because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. 
+Compared to previous approaches of the litterature, our solution has two main novelties:
 
-%~ because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. 
-Compared to previous approaches, our solution has two main novelties:
-
-\noindent \textbf{Regarding $\mathcal{P}_1$}.
-We propose a fast guide path planning algorithm.
+\noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm.
 The key to its efficiency is that it does not sample directly the contact manifold, but an approximation of the \textit{contact reachable} space.
@@
 
-The reachability condition is the key to the strict separation between $\mathcal{P}_1$ and $\mathcal{P}_2$, hence to the low complexity of our planner, though it
-can result in failures.
-However, we demonstrate empirically its interest: the high success rate and low computation times allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
+\textbf{The reachability condition is the key to the strict separation between $\mathcal{P}_1$ and $\mathcal{P}_2$}, hence to the low complexity of our planner.
+However, the reduction may result in failures.
+We demonstrate empirically its interest, both in the high success rate and in the low computation times, by a deep experimental validation, with two realistic robot models (humanoid and quadruped) in various scenarios, involving statistics of the relevance of our hypotheses.
+These two properties allow us to plan, and re-plan upon failure, multi-contact sequences at \gls{interactive}$^1$ rates.
 
-To further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
+%% To further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
 For HRP-2, we detail the complete computation times to address sequentially the three problems, and compare them to related work, demonstrating that our method is orders of magnitude faster.
 
-Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem. \\
+%% Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem. \\
 
-\noindent \textbf{Comparison with our previous work:}
-The present paper is an extension of our ISRR conference paper~\cite{tonneauisrr15}. 
-As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section~\ref{rbprm} and~\ref{sec:contact}). 
-However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section~\ref{app:contact}), a novel criterion for
-static equilibrium (Section~\ref{sec:Equil}), and the release of the source code of the planner (Section~\ref{app:hpp}).
+%\noindent \textbf{Comparison with our previous work:}
+\subsection{Organization and comparison with preliminary paper~\cite{tonneauisrr15}}
+The present paper is an extension of our ISRR conference paper~\cite{tonneauisrr15}, that it aims at replacing.
+The core methodological part corresponds to the same motion planner, with important complements listed below.
+The present paper also proposes a novel experimental section, with statistical validation of the planner hypothesis using realistic robot models.
 
-The other contribution of the paper is a rigorous experimental validation of the approach on actual robot models (Section~\ref{sec:results}).
-To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
-This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work~\cite{Carpentier2016}.
-Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
-We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution.
+After presenting an overview of our method (Sec.~\ref{overview}) and introducing our notations (Sec.~\ref{notations}),
+we present our solution to address $\mathcal{P}_1$ (Sec.~\ref{rbprm}) and $\mathcal{P}_2$ (Sec.~\ref{sec:contact}) , which corresponds to the same motion planner as the one presented at \cite{tonneauisrr15}.
+The two sections hence are reformulation of the preliminary paper.
+Three important novelties have been added to the planner: the pseudo-code of the algorithm (Sec.~\ref{app:contact}), a novel criterion for
+static equilibrium (Sec.~\ref{sec:Equil}), and the release of the source code of the planner (Sec.~\ref{app:hpp}).
 
-Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
-thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.
+The other novelty of the paper is a rigorous experimental validation of the approach on actual robot models (Sec.~\ref{sec:results}).
+We had to introduce a complete framework for multi-contact motion synthesis, in order to validate our contact plans.
+It comprises an interpolation method to solve the problem $\mathcal{P}_3$, based on a reformulation of our previous work~\cite{Carpentier2016}.
+The generated motions are verified to be physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
+These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}.
+The novelty lies in the complete validation of the contact planner with real robot models, for which the planning is much harder with respect to the avatars used in \cite{tonneauisrr15}, because of stricter kinematic constraints.
+
+%We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution.
+
+%Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models.
+These models are more constrained kinematically than the virtual avatars used before, thus making the planning problem harder.
+Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.
+
 %~ These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}.

### 2017-10-10 reviews_tro/reviewer_response.tex
updated letter with simulator answer

@@
 
-Regarding the first contribution, the algorithm presented here is the same as in [20], but with three important novelties: a novel criterion for
+Regarding the first contribution, the algorithm presented here is the same as in [20], with three novelties: a novel criterion for
 static equilibrium, the pseudo-code of the algorithm,
@@
 Following the recommendation of the associate editor who ruled over some contradictory opinions of the Reviewers, we made the paper self-contained,
-in such a way that the paper [20] is now obsolete with respect to this submission, and is only mentioned in the introduction, or to describe an improvement
+in such a way that the paper [20] is now obsolete with respect to this submission, and is only mentioned in the introduction, or to emphasize an improvement
 of the original algorithm. The missing information pointed out by the Reviewers has been 
@@
 
-We thus believe to have significantly improved the paper by addressing the majority of the comments and questions raised by the Reviewers.
+We have thus addressed the majority of the comments and questions raised by the Reviewers. In our opinion, these modifications result in a significant improvement of the paper.
 
@@
 In the simulations we controlled the robot with a standard inverse-dynamics controller[31].
+The code source of the simulator is available at \url{https://github.com/andreadelprete/pinocchio_inv_dyn/releases/tag/rbprm}.
 This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
@@
 %~ \textcolor{red}{TODO ANDREA}.
+
+\quot{p.14 - not a good description of the simulation software used [\dots]Ref [38] focuses on robustness of simulations, but presumably, the testing in the present work is deterministic.
+ just using the same simulator?
+Not much description is given of how/whether contacts/slipping are modeled [\dots] Why isn't a 3rd party software used, or at least software with better documentation ?
+Better info is needed to understand what the simulation truly represents.}
+\as{Thanks for the comment.
+In summary, we had to implement our simulator because we were not satisfied with commercial simulators.
+Our simulator implements the paper ``Staggered projections for frictional contact in multibody systems'', by Kauffman, Sueda, James and Pai, which corresponds to a state-of-the-art solution to the problem.
+We considered other simulators, namely Mujoco, ODE, Gazebo, Bullet and Open-HRP.
+We discarded Mujoco because we work on Linux and it only provides a windows implementation.
+In our tests, we were not able to obtain coherent results with ODE and Gazebo, even for simple pendulum models.
+At the time of choosing which simulator to use, Bullet did not allow to express the dynamics
+equations using joint coordinates, but only cartesian coordinates, which made the formulation much more complex.
+Finally, Open-HRP is a great tool, but it only provides an interface for position-controlled models, and not torque-controlled.
+For these reasons, and other historical ones, we implemented our own simulator.
+We understand that because we use an in-house implementation, it is harder to trust the results we obtain.
+However, the implementation is from a state of the art LCP solver, used by many in the community.
+In this new submission, we also provide the source of the simulator.
+As for the paper itself and the rest of the code, we put a considerable effort in providing the reader with all the elements that will lead to an objective opinion of the paper.}
 
@@
 \quot{
+The first part of App.
+B is referred to within this Appendix as “new minor contributions derived from previous work”.
+These aren’t so novel, perhaps, and what is more useful to the reader is a better intuition for when to use each within the MAIN BODY of the paper, where they are originally described/associated to particular simulations.
+}
+
+\as{Agreed. We removed the incriminated text and provided an intuition of the heuristic in the text, as mentioned above.}\done
+
+\quot{
 p.6 six fingered hand is slightly misleading, since no rolling (as fully admitted in the text). I'd just cut this? \\
@@
 \done
-
-\quot{
-The first part of App.
-B is referred to within this Appendix as “new minor contributions derived from previous work”.
-These aren’t so novel, perhaps, and what is more useful to the reader is a better intuition for when to use each within the MAIN BODY of the paper, where they are originally described/associated to particular simulations.
-}
-
-\as{Agreed. We removed the incriminated text and provided an intuition of the heuristic in the text, as mentioned above.}\done
 

### 2017-10-10 body/intro.tex
nicolas comments

--- 2017-10-09 body/intro.tex
@@
 
-\noindent \textbf{Regarding $\mathcal{P}_1$}.
-We propose a fast guide path planning algorithm.
+\noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm.
 The key to its efficiency is that it does not sample directly the contact manifold, but an approximation of the \textit{contact reachable} space.
@@
 
-The reachability condition is the key to the strict separation between $\mathcal{P}_1$ and $\mathcal{P}_2$, hence to the low complexity of our planner, though it
+The reachability condition is the key to the strict separation between $\mathcal{P}_1$ and $\mathcal{P}_2$, hence to the low complexity of our planner. However the reduction
 can result in failures.
-However, we demonstrate empirically its interest: the high success rate and low computation times allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
+We demonstrate empirically its interest, through an extensive experimental validation with real robot models on a dynamic simulator.
+The high success rate and low computation times we obtain allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.
 
@@
 
-The other contribution of the paper is a rigorous experimental validation of the approach on actual robot models (Section~\ref{sec:results}).
+The other novelty of the paper is a rigorous experimental validation of the approach on actual robot models (Section~\ref{sec:results}).
 To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
@@
 Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
-We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution.
+These aspects of the framework are presented in details in the
+paper, but are not novel \textit{per se}. %Describing them here is required to understand our experimental contribution.
+The novelty lies in the complete
+validation of the contact planner with real robot models, for
+which the planning is much harder with respect to the avatars
+used in~\cite{tonneauisrr15}, because of more restrictive kinematic constraints.
 
-Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
-thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.
+%~ Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
+%~ thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.
 %~ These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}.

### 2017-10-10 reviews_tro/reviewer_response.tex
final version of main document .?

@@
 This is now explained in details in the introduction.}
-\qt{(Section I.B) The present paper is an extension of our ISRR conference paper [20]. 
-As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section IV and V). 
-However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section V.B.2), a novel criterion for
-static equilibrium (Section VI), and the release of the source code of the planner (Section VII). \\
-The other novelty of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
-To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
-This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1].
-Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
-We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution. \\
-Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
-thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.}
+\qt{(Section I.B) The present paper is
+an extension of our ISRR conference paper [20]. As such, our
+solution to address $\mathcal{P}_1$  and $\mathcal{P}_2$  is the same motion planner
+as the one presented at ISRR (reformulated in Section IV
+and V). However three important novelties have been added to
+the planner: the pseudo-code of the algorithm (Section V-B2),
+a novel criterion for static equilibrium (Section VI), and the
+release of the source code of the planner (Section VII). \\
+The other novelty of the paper is a rigorous experimental
+validation of the approach on actual robot models (Sec-
+tion VIII). To validate our contact plans we introduced a
+complete framework for multi-contact motion synthesis. This
+framework additionally comprises an interpolation method
+to solve the problem $\mathcal{P}_3$, based on a reformulation of our
+previous work [1]. Our solution to $\mathcal{P}_3$ allows us to verify that
+the synthesized motions are physically consistent, using our
+implementation of a state-of-the-art simulation algorithm [28].
+These aspects of the framework are presented in details in the
+paper, but are not novel per se. The novelty lies in the complete
+validation of the contact planner with real robot models, for
+which the planning is much harder with respect to the avatars
+used in [20], because of more restrictive kinematic constraints}
 \done
@@
 of Section I.B.}
-\qt{Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.}
+\qt{ our
+solution to address $\mathcal{P}_1$  and $\mathcal{P}_2$  is the same motion planner
+as the one presented at ISRR (reformulated in Section IV
+and V). However three important novelties have been added to
+the planner: the pseudo-code of the algorithm (Section V-B2),
+a novel criterion for static equilibrium (Section VI), and the
+release of the source code of the planner (Section VII).}
 \done
@@
 Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
-We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution.}\done
+These aspects of the framework are presented in details in the
+paper, but are not novel \textit{per se}.}\done
 
@@
 Better info is needed to understand what the simulation truly represents.}
-\as{Thanks for the comment.
-In summary, we had to implement our simulator because we were not satisfied with commercial simulators.
-Our simulator implements the paper ``Staggered projections for frictional contact in multibody systems'', by Kauffman, Sueda, James and Pai, which corresponds to a state-of-the-art solution to the problem.
-We considered other simulators, namely Mujoco, ODE, Gazebo, Bullet and Open-HRP.
-We discarded Mujoco because we work on Linux and it only provides a windows implementation.
-In our tests, we were not able to obtain coherent results with ODE and Gazebo, even for simple pendulum models.
-At the time of choosing which simulator to use, Bullet did not allow to express the dynamics
-equations using joint coordinates, but only cartesian coordinates, which made the formulation much more complex.
-Finally, Open-HRP is a great tool, but it only provides an interface for position-controlled models, and not torque-controlled.
-For these reasons, and other historical ones, we implemented our own simulator.
-We understand that because we use an in-house implementation, it is harder to trust the results we obtain.
-However, the implementation is from a state of the art LCP solver, used by many in the community.
-In this new submission, we also provide the source of the simulator.
-As for the paper itself and the rest of the code, we put a considerable effort in providing the reader with all the elements that will lead to an objective opinion of the paper.}
+\as{Thanks for the comment. In summary, we had to implement our
+simulator because we were not satisfied with off-the-shelf simulators. Our
+simulator implements the algorithm proposed by the paper ``Staggered projections for frictional
+contact in multibody systems'', by Kauffman, Sueda, James and Pai, which
+corresponds to a state-of-the-art solution to the problem.
+We considered other simulators, namely Mujoco, Gazebo and
+Open-HRP. We discarded Mujoco because it only provides a Windows 
+implementation, while we work on Linux. In our tests with Gazebo, we were not able to
+obtain coherent results, even for simple pendulum models. 
+Finally, Open-HRP is a great tool, but it only provides an interface for 
+position-controlled models, and not torque-controlled. For these reasons, 
+and other historical ones, we implemented our own simulator. 
+We understand that because we use an
+in-house implementation, it is harder to trust the results we obtain.
+However, we implemented a state of the art LCP algorithm, used
+by many in the community. In this new submission, we also provide the
+source of the simulator.}
 
@@
 This is now explained in details in the introduction.}
-\qt{(Section I.B, repeated from page 5) 
-The present paper is an extension of our ISRR conference paper [20]. 
-As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section IV and V). 
-However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section V.B.2), a novel criterion for
-static equilibrium (Section VI), and the release of the source code of the planner (Section VII). \\
-The other novelty of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
-To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
-This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1].
-Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
-We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution. \\
-Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
-thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.}\done
+\qt{(Section I.B, repeated from page 5) The present paper is
+an extension of our ISRR conference paper [20]. As such, our
+solution to address $\mathcal{P}_1$  and $\mathcal{P}_2$  is the same motion planner
+as the one presented at ISRR (reformulated in Section IV
+and V). However three important novelties have been added to
+the planner: the pseudo-code of the algorithm (Section V-B2),
+a novel criterion for static equilibrium (Section VI), and the
+release of the source code of the planner (Section VII). \\
+The other novelty of the paper is a rigorous experimental
+validation of the approach on actual robot models (Sec-
+tion VIII). To validate our contact plans we introduced a
+complete framework for multi-contact motion synthesis. This
+framework additionally comprises an interpolation method
+to solve the problem $\mathcal{P}_3$, based on a reformulation of our
+previous work [1]. Our solution to $\mathcal{P}_3$ allows us to verify that
+the synthesized motions are physically consistent, using our
+implementation of a state-of-the-art simulation algorithm [28].
+These aspects of the framework are presented in details in the
+paper, but are not novel per se. The novelty lies in the complete
+validation of the contact planner with real robot models, for
+which the planning is much harder with respect to the avatars
+used in [20], because of more restrictive kinematic constraints.}\done
 
@@
 We now join a diff file comparing the original and current submissions.
-The text appearing in blue denotes the additions. We used the latexdiff tools to generate this file, which is not fully compatible with all the latex packages that we use.
+The text appearing in blue denotes the editions. We used the latexdiff tools to generate this file, which is not fully compatible with all the latex packages that we use.
 For this reason, errors appear in the document, in particular regarding the subsections numbering. However, we think that this diff file will help the Reviewers locate precisely

### 2017-10-10 body/results.tex
rewrote I.B more like in the letter
link to simulator source code
final version of main document .?

@@
 In the simulations we controlled the robot with a standard inverse-dynamics controller~\cite{DelPrete2015b}.
+The code source of the simulator is available at \url{https://github.com/andreadelprete/pinocchio_inv_dyn/releases/tag/rbprm}.
 This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
@@
 %\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}
-\subsubsection{HRP-2 -- Staircase with high steps (Figure~\ref{fig:stair_robust})}
-\newcommand{\widthValue}{0.13\linewidth}
-
-\begin{figure*}[h!t!]
-  \begin{center}
-  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing1.jpeg}
-  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing2.jpeg}
-  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing3.jpeg}
-  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing4.jpeg}
-  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing5.jpeg}
-  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing6.jpeg}
-  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing7.jpeg}
-  \caption{
-           HRP-2 in the stair climbing scenario. }
-		   \label{fig:stair_robust}
-  \label{fig:moviepicture2}
-  \end{center}
-  \vspace{-0.5cm}
-\end{figure*}
-
-%~ \begin{figure}
-  %~ \centering
-  %~ \includegraphics[width=1\linewidth]{figures/stair}
-  %~ \caption{
-           %~ HRP-2 in the stair climbing scenario. }
-		   %~ \label{fig:stair_robust}
-%~ \end{figure}
-
-The goal is to climb three 15-cm high steps.
-
-\noindent\textbf{Contacts involved:} Feet and right arm.
-
-\noindent\textbf{Heuristics:} The manipulability $h_w$ is chosen for the feet; $h_{\textrm{\it EFORT}}$ is chosen for the right arm.
-%~ Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
-
-%~ \noindent\textbf{Observations:}
-%~ This scenario illustrates best the importance of the equilibrium-robustness criterion.
-%~ With a robust approach, more states are required to reach the last step (15 rather than 13 in average).
-%~ However, when the last step is reached by both feet, in the nonrobust case the contacts are extremely close to 
-%~ the cone limits (Figure~\ref{fig:stair_comp}).
-%~ 
-%~ 
-%~ The geometry of the environment is easily addressed by our planner, and the contact planning is several times faster than real time in this scenario.
-%~ 
-%~ Again, the interpolation motion between the contact steps is out of the scope of this paper.
-However it should be noted that the computed plan in this scenario has been executed successfully on the robot (\url{http://youtu.be/YjL-DBQgXwk#t=0m28s}).
-%~ 
-%~ \begin{figure}
-  %~ \centering
-  %~ \begin{overpic}[width=0.5\linewidth]{figures/stair_robust}
-		%~ \put (17,5) {\small{\color{red}$b_0 = 0.23$}} 
-		%~ \put (79,5) {\small{\color{green}$b_0 = 6.16$}} 
-	%~ \end{overpic}
-  %~ \caption{
-           %~ Evaluation of the robustness $b_0$ of two contact configurations. Although in equilibrium, the left configuration is on the verge of slipping.}
-		   %~ \label{fig:stair_comp}
-%~ \end{figure}
 
@@
 %~ While this scenario is the slowest to solve, the planner still computes a solution \glslink{interactive}{interactively}.
+
+\subsubsection{HRP-2 -- Staircase with high steps (Figure~\ref{fig:stair_robust})}
+\newcommand{\widthValue}{0.13\linewidth}
+
+\begin{figure*}[h!t!]
+  \begin{center}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing1.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing2.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing3.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing4.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing5.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing6.jpeg}
+  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
+    {./video/stairclimbing7.jpeg}
+  \caption{
+           HRP-2 in the stair climbing scenario. }
+		   \label{fig:stair_robust}
+  \end{center}
+  \vspace{-0.5cm}
+\end{figure*}
+
+%~ \begin{figure}
+  %~ \centering
+  %~ \includegraphics[width=1\linewidth]{figures/stair}
+  %~ \caption{
+           %~ HRP-2 in the stair climbing scenario. }
+		   %~ \label{fig:stair_robust}
+%~ \end{figure}
+
+The goal is to climb three 15-cm high steps.
+
+\noindent\textbf{Contacts involved:} Feet and right arm.
+
+\noindent\textbf{Heuristics:} The manipulability $h_w$ is chosen for the feet; $h_{\textrm{\it EFORT}}$ is chosen for the right arm.
+%~ Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
+
+%~ \noindent\textbf{Observations:}
+%~ This scenario illustrates best the importance of the equilibrium-robustness criterion.
+%~ With a robust approach, more states are required to reach the last step (15 rather than 13 in average).
+%~ However, when the last step is reached by both feet, in the nonrobust case the contacts are extremely close to 
+%~ the cone limits (Figure~\ref{fig:stair_comp}).
+%~ 
+%~ 
+%~ The geometry of the environment is easily addressed by our planner, and the contact planning is several times faster than real time in this scenario.
+%~ 
+%~ Again, the interpolation motion between the contact steps is out of the scope of this paper.
+However it should be noted that the computed plan in this scenario has been executed successfully on the robot (\url{http://youtu.be/YjL-DBQgXwk#t=0m28s}).
+%~ 
+%~ \begin{figure}
+  %~ \centering
+  %~ \begin{overpic}[width=0.5\linewidth]{figures/stair_robust}
+		%~ \put (17,5) {\small{\color{red}$b_0 = 0.23$}} 
+		%~ \put (79,5) {\small{\color{green}$b_0 = 6.16$}} 
+	%~ \end{overpic}
+  %~ \caption{
+           %~ Evaluation of the robustness $b_0$ of two contact configurations. Although in equilibrium, the left configuration is on the verge of slipping.}
+		   %~ \label{fig:stair_comp}
+%~ \end{figure}
 
@@
 In all scenarios, one can observe that the average computation time for one single step is largely below one second,
-thus allowing to consider \gls{interactive} applications and online autonomous planning of the robot motion.
+thus enabling \gls{interactive} applications and online autonomous planning of the robot motion.
 

### 2017-12-12 review_tro_2/reviewer_response.tex
r1 comment 2 LP

--- 2017-10-10 reviews_tro/reviewer_response.tex
@@
 \author {}
-\title {Cover letter for the resubmission of our paper ``An efficient acyclic contact planner for multiped robots''}
+\title {Cover letter for the resubmission of the conditionally accepted paper ``An efficient acyclic contact planner for multiped robots''}
 \date {}
@@
 
-First of all, we would like to thank all the Reviewers and the associate editor (AE).
-We received three detailed and technical reviews that helped improving our paper.
-Overall we agreed with most of the reviews and took them into account in the revised version of the paper.
-The main points of concern raised by the Reviewers regarded:
-\begin{itemize}
-\item the diffficulty to understand the takeaway message of the paper, and what was novel with respect to our previous paper [20] (and [1] to a lesser extent);
-\item the organization of the paper, with too much information relegated to the Appendix, or only referred to in other contributions.
-\end{itemize}
+First of all, we would like to thank all the Reviewers and the associate editor (AE) for considering our resubmission.
+We are glad to find them satisfied with our modifications, and made our best to answer to the remaining comments as well as possible.
 
-\textbf{Regarding the contribution of this paper with respect to our previous papers}, this paper must be understood
-as an extension of [20] (presented at the ISRR conference).
-As a self-contained journal paper, the present paper aims at replacing [20].
-We present two contributions.
-The first one, as in [20], is an efficient and generic planning method for multi-contact locomotion (problems P1 and P2).
-The second one is a rigorous experimental validation of the approach on actual robot models, either in simulation or on the real HRP-2 robot.
+In the following lines, we will quote all the remaining comments and answer to explain how exactly we updated the paper to address them.
 
-Regarding the first contribution, the algorithm presented here is the same as in [20], but with three important novelties: a novel criterion for
-static equilibrium, the pseudo-code of the algorithm,
-as well as  the release of the source code of the planner.
+\section{Answers to the associate editor}
 
-Regarding the second contribution, to validate the contact plans computed we introduced a complete framework for multi-contact motion synthesis.
-This framework additionally comprises an interpolation method to solve the problem P3, similar but not identical to the one we proposed in [1].
-It also comprises an implementation of a state-of-the-art dynamic simulator to check that the motions are physically plausible [28].
-These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}.
-The novelty lies in the validation of the method with real robot models, for which the planning is harder with respect to the avatars used in [20], because of stricter kinematic constraints.
+\quot {The C++/Java-style code listings are unnecessary and do not conform
+to journal standards.  The authors rewrite their pseudocode to use
+rigorous mathematical notation.  Suggest replacing "object.attribute"
+with "Attribute(object)", and describing the attributes of the objects
+in the text.
+}
 
-We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.
-To clarify these aspects in the paper, we rewrote the end of Section I.B to clearly state what is new (and what is not) in this paper.
+\as{TODO}
 
-%~ This paper is an extension of [20] (presented at the ISRR conference), that it aims at replacing.
-%~ Like [20], the main contribution of our paper is an efficient and generic method for planning contact sequences of multi-contact locomotion, which leads to the first interactive contact planner in realistic \emph{robotic} scenarios.
-%~ Our take-home message is  is that to address the hard problem of multi contact planning, it is necessary to use abstraction and simplification, through the decomposition of the problem into sub-problems and through dimensionality reduction.
-%~ In particular, the planner (core algorithm) presented in this paper is almost identical to the one in [20].
-%~ The main addition with respect to [20] is the empirical validation, with realistic robot model imposing true (difficult) kinematic  constraints.
-%~ Specifically, the novelty with respect to [20] lies in:
-%~ \begin{itemize}
-%~ \item the addition of the pseudo-code (the algorithm is not given in [20]);
-%~ \item the introduction of a more efficient equilibrium criterion;
-%~ \item the new experimental setup and analysis: validation with real robot models (while [20] was only using unrealistic avatars with are much less kinematically constrained); statistical analysis of the validity of the planner models and heuristics; generation of the complete movement through the implementation of a P3 solver; validation that the movements are feasible in a complete dynamic simulator and, for one of them, on a physical robot;
-%~ \item the release of our source code: we provide the complete procedure to use our planner for any legged robot.
-%~ \end{itemize}
-%~ The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation.
-We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.
+\quot {The URLs in the paper seem rather brittle and are not suitable for
+archival purposes.  Two URLs point to specific branches of various
+Github repositories, and another refers to the documentation of a
+Blender tool.  The authors should create a static web page for their
+software which is guaranteed to persist for years to come... }
 
-\textbf{Regarding the organization of the paper}, in the initial version we tried to synthesize as much as possible
-the core of the paper, while providing the most technical details in the appendix. Based on the Reviewers comments, we modified
-the paper architecture in such a way that all the novel details are now directly included in the core of the paper. This should avoid the reader going back and forth
-in the paper, and also help him distinguishing between what is novel and what is not.
-As a result,  Appendices B.B (equilibrium criterion), D (pseudo code) and E (reference to the source code) are now integrated in the core of the paper.
-The manipulability-based heuristics used for contact generation, as well as our solution to the problem P3 remain in the Appendix as they are not novel.
-However, as suggested by the Reviewers, they are described in more details in the core.
+\as{TODO}
 
-Following the recommendation of the associate editor who ruled over some contradictory opinions of the Reviewers, we made the paper self-contained,
-in such a way that the paper [20] is now obsolete with respect to this submission, and is only mentioned in the introduction, or to describe an improvement
-of the original algorithm. The missing information pointed out by the Reviewers has been 
-included in the paper, as explained in the detailed answers that follow. As a result, we believe that the new version of the paper is self-contained and can be understood if read linearly.\\
+\quot {... and to refer
+to the underlying method used by the Blender tool.}
 
- %~ In order to try to make the paper more readable, the first version was mixing three styles: details given in the core of the paper; details given in the appendices; details deferred in cited papers.
-While all Reviewers agreed that using all three styles together was a bad idea, we received contradictory reviews about how to modify the paper.
-Following the recommandation of the AE, we have worked to make the paper self-contained with all the contributions in the core of the paper.
-%~ We have brought all the material from [20] in the revised paper so that no references to this preliminary paper is necessary anymore in the technical description, and removed trivial reference to [20] but in the introduction.
-%~ We merged Appendices B.B (equilibrium criterion), D (pseudo code) and E (referernce to the source code) in the core of the paper.
-We kept in the appendices the manipulability computations which is necessary for completeness but is well-known.
-We also prefered to keep the trajectory solver P3 in the appendix, although some details are novel, in order to keep clear that the main contributions of the paper are on P1 and P2.
-
-We thus believe to have significantly improved the paper by addressing the majority of the comments and questions raised by the Reviewers.
-
-\bigskip
-
-In the following, we list all the questions and remarks from the Reviewers, answer to the questions, and detail the modifications that were brought to the paper.
-At the end of this letter, we also join a diff file between the original submission and the present one, that should help the Reviewers see the modifications 
-brought to the paper.
+\as{TODO}
 
@@
 
-\quot {As for novel contributions, this did not seem to be addressed/explained
-until page 3 (in the last paragraph of Sec. I). Much of the methodology
-has already been presented in earlier work, and the particular aspects
-that set THIS paper apart seem (as claimed by the authors) to be: 
+\quot {It looks like that sentence was not actually removed from the
+paper (referring to the sentence ``optimization-based approaches
+only converge locally'')
 }
 
-\as{Thanks for the comment.
-As mentioned in the summary, our planner is almost identical to the one in [20], but the paper comprises important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
-for checking static equilibrium.
-The main novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robot models that are kinematically more constrained than the virtual avatars from [20].
-This is now explained in details in the introduction.}
-\qt{(Section I.B) The present paper is an extension of our ISRR conference paper [20]. 
-As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section IV and V). 
-However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section V.B.2), a novel criterion for
-static equilibrium (Section VI), and the release of the source code of the planner (Section VII). \\
-The other novelty of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
-To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
-This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1].
-Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
-We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution. \\
-Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
-thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.}
-\done
+\as{Indeed, due to a misguided merge, the sentence was reintroduced against our will. We apologize for that, and guarantee that this time it was indeed removed!}
 
-%~ \as{We agree that our original submission is not clear on the novelty between this paper and previous contributions.
-%~ This paper must be understood as an extension of the ISRR paper [20], while the other cited contributions from our team 
-%~ are exploited to validate our approach in a complete pipeline.
-%~ In this extension, the algorithm used to compute a sequence of contacts is the same as in [20] (although this
-%~ algorithm is not explicitely given in [20]), with the introduction of a fastest test for verifying the robust quasi static equilibrium of a contact posture.
-%~ We reformulated the end of Section I.B to state clearly that the novelty of this paper essentially lies in the experimental setup and validation we achieved with real robot models, in particular
-%~ in the empirical justification of the method, and its validation in a dynamic simulation.
-This effort led to the proposal of the first open source, complete framework for multi contact motion synthesis, since
-%~ the validation required to write a solver for the problem P3. }\done
-
-\quot { (1) A criterion for robust static equilibrium, but it isn't
-described until page 12, in Appendix B, Section B. [\dots]  It would really help to describe this MUCH earlier in
-the paper (than page 13), if it is a significant novelty of your paper.}
-
-\as{We agree with the remark.
-To better distinguish between what is novel and not, we modified the paper to move the content of Appendix B.B (now Section VI), as well as D  (the pseudo code of our algorithm, now Section V.B.2) and E (reference to our source code, now Section VII) to the text itself.
-This results in the criterion being presented much earlier.}\done \jp
-
-\quot{It is a pretty intuitive/typical check for quantifying a
-margin for friction conesque violations (using an approximation for
-the friction cone, with x and y direction forces considered
-independently). }
+\quot {
+The dual reformulation of the LP in (11) in Sect. VI is fine, but is
+a perfectly standard procedure (that, in fact, many LP solvers will
+perform internally automatically). It would be sufficient to mention
+that the dual problem is solved without giving its full formulation.
+Whether the authors include the formulation is up to them.}
 
 \as{Thanks for the comment.
-To our knowledge, what is common in the literature is to reduce the friction coefficient to obtain more conservative bounds on the admissible forces, and thus a ``safety margin'' to the real cone boundary.
-An issue with this formulation is that the resulting margin may be too small for small forces, and too large for large forces.
-With our formulation instead, the margin is constant (i.e.
-the value $b_0$).
-We updated the paper to explain this difference.
-We believe this formulation is novel, but would be happy to cite any relevant reference if we happen to be wrong.
-This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.
-}\done
-\qt{(Section VI.2) A simple approach to robustness consists in choosing a smaller friction coefficient, to constrain the forces to lie away from the boundaries of the real cone.
-However, this would result in a small safety margin for forces of low magnitude, and an excessively large safety margin for large forces as the boundaries grow more and more apart.
-In comparison, our margin $b_0$ is constant, and provides a helpful mean to compare the robustness of different contact configurations.}
+Indeed, using the dual formulation is quite common for LP solvers.
+However, after computing the dual in our case, it turns out that some of the constraints can be solved trivially, which allows to reduce the dimensionality of the problem, resulting in (11).
+We experimentally verified that this form is faster to solve with respect to the one from equation (10).
+For this reason, we believe that equation (11) is worth mentioning, and prefer to leave it, since it was given as an option by the reviewer.
+Finally, we removed the reference to the Slater's conditions since we agree with the reviewer that the procedure is standard.}
 
-\quot{
-Provides complete pseudocode.  This is great! However,it's
-not fully clear how much of this is novel vs just re-presenting the
-same algorithms previously publilshed.	If only code in App. B and C
-are new, that should be clear. Conversely, if most of this presentation
-was NOT outlined in such detail before, that should also be clear (to
-the authors' credit). Either way, a brief clarification in the intro is
-recommended.}
-
-\as{We agree with the comment. Indeed, the pseudocode was not present in [20]. We included it in the core of the paper and clarified that it is novel at the end
-of Section I.B.}
-\qt{Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.}
-\done
-
-\quot{(3)	The contribution to provide a solution to P3 doesn't happen
-until page 13 (Appendix C) and as with contribution (1), it needs a
-better, intuitive summary (briefly) when first mentioned, as well as a
-roadmap for the reader as to where it will be described. It is also
-ambiguous in the paper whether to what extent (if any) this solution is
-novel, versus a presentation of work from [1]. "Provide a solution"
-seemed to imply novelty, in the context of this paragraph -- yet later,
-the authors (near the end of page 2) state they use a "state-of-the-art
-solution to P3", which implies the P3 solution is not a claim of
-novelty.}
-
-\as{The main contribution of the paper is the proposal and validation of an algorithm for P1 and P2. 
-In the new version of the paper, we do not consider that our solution to P3
-is a new contribution, although our formulation differs from [1] (our problem variables are the accelerations
-of the center of mass, whereas in [1] the problem variables are the contact forces). We provide it here for completeness, because it corresponds to our implementation, and 
-was required to validate our approach. 
-We rephrased the text to clearly indicate that the algorithm is not a contribution. We made the choice not to give an intuition nor describe the solver in the core
-of the paper because it is not required to understand the contributions to P1 and P2.}
-\qt{(Section I.B) To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
-This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1].
-Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
-We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution.}\done
-
-\quot{
-(4)	The last contribution is validation using an in-house
-simulator, as opposed to more approximate (unrealistic avatar) models
-used in previous work.	As with the previous contributions claimed, you
-should flesh out just a bit more detail, along with a roadmap to where
-a particular contribution appears in the paper, to prepare and guide
-the reader a bit more. More precisely, what are the differences between
-these two simulation approaches (previous and current)? You need a
-better description of what your in-house simulator does and does not
-do; a vague citation to another paper based on that simulator [38] is
-not sufficient.}
-
-\as{Thanks for the comment. Regarding the absence of details on the dynamic simulator, first of all we apologize because [38] is not the appropriate reference.
-We added a section VIII.A where we describe the simulation in more details.
-
-Regarding the difference between our previous and current work, the postures computed in [20] were not validated in a simulator, nor the complete motions since 
-we did not have a solution to P3 at the time. Then the main
-difference between the avatars of [20] and the real robots we now address are the joint limitations and auto-collisions probabilities.
-The robots are much more constrained in this regard, which makes the planning harder. We thus had to demonstrate that our algorithm is also working
-for these models, while maintaining good performance.
-We added this missing piece of information to the paper, at the end of Section I.B.}
-\qt{(Section VIII.A) To generate continuous movements from our contact plans we used either the framework proposed in [1], or our own implementation of a $\mathcal{P}_3$ solver (Appendix C).
-The resulting movements have been validated either on the real HRP-2 robot (details can be found in [1]), or with our dynamic simulator, based on a state-of-the-art algorithm [25].
-In the simulations we controlled the robot with a standard inverse-dynamics controller[31].
-This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
-The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
-The companion video shows the obtained motions.}
-\done
-%~ \textcolor{red}{TODO ANDREA}.
-
-\quot{[\dots] the authors need to better
-distinguish what the novel contributions are, and to highlight where
-these contributions will be presented in the paper.  Also, many
-important details are in the appendices. The paper would be improved
-with more intuitive comments/summaries on these details, rather than
-just writing (for example), we use a bunch of heuristics, and they are
-described in Appendix X.}
-
-\as{We agree. The novel details of the paper have been moved from the appendix to the core of the paper, and are explicitely detailed at the end of Section I.B.} \done
-
-\quot{
-Also, where you describe "sensitivity" vs "specificity", it would be
-useful to give a more intuitive explanation, example [\dots]}
-
-\as{We updated Section VIII.B.1 to illustrate the meaning of those terms.}
-\qt{(Section VIII.B.1) We then computed the sensitivity and specificity of the reachability condition.
-In this context, the sensitivity refers to the percentage of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}.
-If a sampled configuration is in $C_{Reach}^0$, but our method is unable to generate a contact configuration from it, as a result the sensitivity
-decreases. \textcolor{green}{The sensitivity thus illustrates the confidence we have that any configuration in $C_{Reach}^0$ will effectively lead to a contact configuration.}
-Conversely, the specificity refers to the percentage of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}.
-If a sampled configuration is \textbf{not} in $C_{Reach}^0$, but our method is able to generate a contact configuration from it, as a result the specificity
-decreases.
-\textcolor{green}{The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to $C_{Reach}^0$ (or informally, the confidence that we are not missing valid solutions).}
-We thus look for a compromise between sensitivity and specificity.} \done
-
-\quot{
-However, I suggest the paper will be improved if
-the presentation focuses more on highlighting (a) what is novel, and
-(b)
-what take-away messages you have, about why the whole framework is
-designed the way it is [within the main text; not hidden in
-appendices].}
-
-\as{Thanks for the comments. We believe our new version of the papers answers to a). In particular, as mentioned above the novel content has been moved from the Appendix
-to the core of the paper. Regarding b), we updated the conclusion
-to include a take-away message. Our take-away message is that we can address the hard problem of multi-contact planning using 
-abstraction and simplification, i.e. the decomposition of the problem into three sub-problems, and  dimensionality reduction.
-We believe that the paper develops these aspects to explain the design of this planner. In the introduction, it is done
-with an extensive state of the art that explains how the community came up with the P1, P2, P3 decomposition to tackle a problem that is too hard otherwise. In the conclusion, it is explained 
-with an analysis of the results, which justify our pragmatic choice of going further in the abstraction to obtain interactive planning performance. }
-\qt{(Section X) In this paper we consider the multi-contact planning problem, formulated as three sub-problems  $\mathcal{P}_1$,  $\mathcal{P}_2$,  $\mathcal{P}_3$, addressed sequentially.
-While we propose a global framework that handles all these problems, our contribution focuses on  $\mathcal{P}_1$ and  $\mathcal{P}_2$.
-The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
-the second problem $\mathcal{P}_2$ is the computation of a discrete sequence of whole-body configurations along the root path.
-We believe that this decomposition is currently the most promising approach towards
-a global resolution of the problem. We also claim to have achieved a significant step towards this objective thanks to the dimensionality reduction provided by
-the reachability condition. With our results and the release of our source code, we hope to inspire further research in this direction.}\done
-
-\quot{
-p.6 – this would be a great place to give more ``intuition'' for the heuristics.
-Hw seems to be used in essentially all foot contact, with $h_{efort}$ for hand/arm contacts, and ALSO $h_{vel}$ not seeming to be used at all.}
-\as{Thanks for the suggestion. The text was updated accordingly, and the heuristic $h_{vel}$ was removed since indeed it is not used.} 
-\qt{(Section V.C) In all our experiments, the heuristic $h$ is implemented as a variation of a manipulability-based heuristic [27]. The manipulability is a real number that quantifies how 
-``good'' a configuration is to perform a given task, based on the analysis of the Jacobian matrix.
-With such heuristics, a configuration can be chosen because it is far from singularities, and thus allows mobility in all directions.
-On the contrary, it can be chosen because it is particularly efficient to exert a force in a desired direction.
-In our experiments, the former solution is usually chosen for computing leg contacts, while the latter is used for computing hand contacts.
-We recall the manipulability measure and its derivatives in Appendix B.}\done
-
-\quot{
-p.6 six fingered hand is slightly misleading, since no rolling (as fully admitted in the text). I'd just cut this? \\
-p.9 Not sure if Fig. 14 is worth including. It looks like its just Fig. 9 from the ISRR paper (?) [20], and there is still no ``full solution'' here -
- just an illustration of potential future work. 
- (At a minimum, the figure/solution should probably be referenced as having already appeared in the previous ISSR paper?)
-}
-
-\as{As also suggested by Reviewer 4, we removed the references to the results of the previous work [20], including those.}\done
-
-\quot{
-p.8– somehow, “Table II” doesn’t appear until after “Table IV” – which is a frustrating choice for latex to make.
-Not sure what you can do to change that, but it would be preferable to have tables (and figures) appear in the correct order.
-}
-
-\as{We reorganized the paper such that the Tables appears in the right place, by inverting the Computation times and success rates paragraphs. The order of appearance of the Tables thus remains
-the same, but the numbering is correct, and in agreement with the order of referencement in the text.}\done
-
-\quot{
-p.8. Need to define ``discretization step'' more precisely, I think(?)… Min distance? Max distance? Nominal distance? In time? In Distance? In all 6 DOF??
-}
-
-\as{The distance considered is the distance travelled along the path, computed as a weighted 6D Euclidian variation along it. We updated the text to clarify this}.
-\qt{(Section V.B.1) $j$ depends on a user-defined variable, called the discretization step. It corresponds to the ratio between the length of the path $\mathbf{q}^0(t)$, and the number
-of configurations selected along it to create the contact configurations. The length of the path is computed as the weighted 6D Euclidian distance
-travelled along the it, with a weight of 0.7 for the translation part, and 0.3
-for the orientation part.}
-
-\quot{
-p.9 ‘addresses highly constrained environment’ – actually, not very well, correct?
-This was the “exception” case, for which results were rather poor, right?
-Also a case where “heuristics” are not so good for getting true (reliable) holds on grasp points, etc.?
-}
-
-\as{Thanks for the comment. On this particular point however, we wish to maintain our statement. For scenarios such as the car egress, a success rate of $68 \%$ means that the planner
-will fail at some point, but more importantly it means that if we restart the planning a few times (e.g.
-3) with the same problem, there is a reasonable probability (e.g.
-97\%) that it will find a feasible solution.
-In this regard, the car egress scenario is effectively addressed, because we found a solution for it. \\
-Furthermore, we believe that this result is to be compared with the current state of the art: we are not aware of any other method that was able
-to compute a valid motion from scratch, even in simulation. The closest work to this achievement is the one from Escande et al., but the trajectories were manually interpolated
-after the contact plan was designed, and moreover we cannot know the success rate of the method. \\
-Now regarding the heuristics, we guess that the Reviewer is referring to the large number of repositionings that occurred in the contact plan. We believe that this issue
-is more related to the kinematic and environmental constraints than the heuristic for the choice of contacts. Developing heuristics that would take into account the environment to limit 
-contact repositioning is out of the scope of this work.}
-
-\done
-
-\quot{
-The first part of App.
-B is referred to within this Appendix as “new minor contributions derived from previous work”.
-These aren’t so novel, perhaps, and what is more useful to the reader is a better intuition for when to use each within the MAIN BODY of the paper, where they are originally described/associated to particular simulations.
-}
-
-\as{Agreed. We removed the incriminated text and provided an intuition of the heuristic in the text, as mentioned above.}\done
-
-\quot{github link seems to be dead? (The link on p. 15 is OK\dots should they be the same?)}
-
-\as{We are suprised by this comment because after testing it, the link seems to be valid.
-In any case, in accordance with Reviewer 6, we changed all the links source codes to provide a SHA link that will bring the user to the accurate revision of the projects considered.}
-\qt{(Section VII) Both HPP and our planner can be simply downloaded and compiled by following the instructions on
-\url{https://humanoid-path-planner.github.io/hpp-doc/download.html?branch=rbprm}. \\
-(Appendix C) To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software.
-}\done
-
-\quot{p.5, on right, ``less than epsilon'' between end effector and world - but doesn’t the orientation of the end effector matter tremendously here, too?}
-\as{
-We agree that the orientation of the end effector matters, but in our implementation considering it for filtering did not improve our approach.
-Indeed, we could consider to further filter the set of candidates samples regarding orientation, to reduce the number of failed
-projections. However, because of the joint limits constraints, computing a relevant orientation distance is less trivial than in the 3D case. We tried to implement this
-in earlier versions of the planner, and did not obtain significant gains in the performance. We believe it is explained by the fact that the majority of unfeasible samples is already
-efficiently rejected by the simple ``epsilon distance'' check. We updated the text of the paper to explain this.
-}
-\qt{Because the distance $\epsilon$ does not account for the variation in orientation, several samples of $ C_{Contact}^{\epsilon}$  may turn out to be unfeasible at the time of projection.
-One could consider additionally filtering $ C_{Contact}^{\epsilon}$ based on the orientation with respect to the obstacle normal, but in our experience we did not notice
-any significant improvement in the computational performances of the planner, so we do not perform this additional step. }
-%~ Using an epsilon is an effective way to remove many irrelevant configurations.  and
-%~ we believe that more complex solutions aiming at taking into account the orientation would end up being counter productive.
- %~ Indeed, using only 3D distance, we will end up with a few configurations
-%~ that can't be projected correctly, and will end up being discarded. But the majority of irrelevant configurations would already have been rejected with
-%~ the efficient 3D intersection query. Introducing additional filtering . } 
-\done
-
-\quot{remaining remarks by Reviewer 1}
-\as{We thank the Reviewer for the consequent amount of remarks brought to the paper.
-The remaining comments that are not described here were all addressed, following the suggestions of the reviewer.
-Because they did not call for a discussion nor interpretation, as they are essentially related to style or typos mistakes, we do not list them here.}\done
-
-%~ \quot{p.14 – not a good description of the simulation software used \dots Ref [38] focuses on robustness of simulations, but presumably, the testing in the present work is deterministic.
- %~ just using the same simulator? Not much description is given of how/whether contacts/slipping are modeled\dots why isn’t a 3rd­‐part software used, or at least software with better 
- %~ documentation? Better info is needed to understand what the simulation truly represents.}
-%~ \as{Thanks for the comment. As mentionned before in this text, we added a paragraph (TODO location) where we describe more acccurately the simulator used to perform the validation.}
-%~ 
-%~ \begin{itemize}
-%~ \item \textbf{from avatars to real robots.} [20] presents a proof of concept applied to plan contact sequences for virtual avatars. The main
-%~ difference between these avatars and the real robots we now address are the joint limitations and auto-collisions probabilities.
-%~ The robots are much more constrained in this regard, which makes the planning harder. We thus had to demonstrate that our algortihm is also working
-%~ for these models, while maintaining decent performances.
-%~ \item \textbf{a more efficient verification for quasi-static equilibrium.} The criteria for robust equilibrium we use replaces the one proposed in [20]. The formulation is new,
-%~ but also much faster than before. Also, it provides a good way to differentiate between different contact candidates and select the most relevant one.
-%~ \item \textbf{statistical validation.} Compared to [20], our paper provides an extensive analysis of the efficiency of the algorithm. This empirical
-%~ study is required to justify that the simplifications made by our approach result in successful planning in general.
-%~ \item \textbf{validation in simulation}: Finally, using state of the art approaches, we validated our contact plans by first generating a continuous whole body motion to interpolate them (P3),
-%~ before validating them in a state of the art simulator or on the real robot. Although our formulation of P3 is novel regarding [1] (our control variable is the acceleration
-%~ of the center of mass, where in [1] the control variables are the contact forces), we do not claim a strong contribution.
-%~ \end{itemize}
-%~ 
-%~ It seems to us that this new material answers the T-RO guidelines regarding the novelty with respect to a conference paper.
-%~ This is not clearly stated in the original submission, and we thus rewrote the end of Section I to reflect this.
-
-\section{Answers to Reviewer 4}
-
-\quot{ I suggest that a few papers related to more general motion and contact planning for multi-legged robots can be added.
-Particularly, the problem of ``foothold selection'' for 4 or 6-legged robots travelling over uneven terrains is a problem related to the one tackled in the paper.
-It would be nice to select a few representative works (e.g.
-from the DARPA Learning Locomotion project or
-papers related to the ETHZ StarlETH and ANYmal robots) and explain the differences between those approaches and the problem solved in the paper.}
-\as{Thanks for the relevant comment.
-As observed by the Reviewer, our state of the art voluntarily focuses on multi-contact planning, but it makes sense to cite the suggested papers in the revision.
-We added a paragraph at the end of the state of the art to discuss them.
-As a matter of fact, we believe that the learning approaches
-proposed for computing relevant effector locations could be considered within
-our framework to provide good contact generation heuristics.}
-\qt{(Section I.A) Other contributions to legged locomotion, not directly related to multi-contact motion, are worth mentioning, as they rely on a similar decomposition of the problem.
-First a discrete sequence
-of contact sets is planned (at the so-called footstep planning phase), using a low-dimensional abstraction of the robot to account for its kinematic constraints [22,23].
-In [23], a ``pose certificate'' is obtained by generating a whole-body configuration for each set through inverse kinematics, as done in $\mathcal{P}_2$.
-Then a
-motion is generated along the sequence through the use of optimization techniques. 
-The solutions proposed are designed for cyclic locomotion in quasi-flat scenarios, where the support polygon is a relevant method for equilibrium checks.
-They thus cannot be generalized to multi contact locomotion.
-However, some contributions on specific parts of the problem could be applied directly in our case. 
-For instance, learning ``terrain costs'', based on expert knowledge as proposed in [24], could define good heuristics to compute the next contact location for an effector.
-Although we did not try to include such formulation in this paper, it would be straightforward to integrate such heuristic in our planner.}\done
-
-\quot{in my opinion a T-RO regular paper should be more self-containing.
-In several points the authors heavily rely on previous publications, e.g.
-when presenting the RRT-based planner (Section IV-C).
-This part should be expanded, describing in more detail what exactly has been changed with respect to the reference paper.
-}
-\as{Thanks for the comment. As it is also suggested by Reviewer 1 and the AE, we modified the paper to make it self contained. The references to previous work now solely indicate
-that a certain method is not a contribution of the present paper. \\ 
-In the particular case of Section IV-C, in fact the code of our implementation of the Bi-RRT with respect to the original one is exactly the same, with a single difference: when
-traditionally a collision detection predicate is called to validate a configuration, we replace this predicate with a call to the reachability condition.
-We modified the text to explain this more clearly.
-As the Bi-RRT is a famous standard algorithm in the literature, we do not think it is necessary to explain it in this paper, but we will gladly do so if required.
-}
-\qt{(Section IV-C) $C_{Reach}^0$ can be sampled efficiently thanks to Eq.~5, and can thus be used with any standard motion planner.
-Our current implementation uses the Bi-RRT planner [25] provided by the HPP software [26].
-Our implementation is exactly the same as the pseudo-code of the original planner (which does not detail the configuration validation method).
-With respect to a ``classic'' implementation, the only difference is that instead of validating a configuration using collision detection, we validate it with the \textit{reachability condition}.} \done
-
-\quot{I am a bit skeptical about the readability of the figures.
-Some of them seem to be too small (e.g.
-Fig.
-3, Fig.
-4), some are of quite poor quality (why the robot silhouettes in Fig.
-4 are blurred ?), and the purpose and meaning of Fig.
-6 is rather unclear.
-\\ In general, the simulation results presented qualitatively in the paper are a bit hard to interpret without the accompanying video, particularly if someone looks at a b/w print of the paper (as it may be looking in T-RO).}
-\as{Thanks for the comment, with which we agree in general.
-\\We increased the size of Fig.
-3 and 4 as much as possible to hold on one column, because we were not satisfied with the result obtained after trying to put them on a double column.
-The robot silhouettes in Fig.
-4 are blurred when they describe a previous state of the robots.
-We added this information in the text.
-\\ As proposed also by Reviewer 1, we removed references to previous results obtained in [20], so Fig.
-6 has been removed.
-\\Regarding the quality of our figures,  our simulation figures are taken from high resolution rendering in blender.
-The other figures are drawn from manually designed powerpoint vectorial graphics (converted to png).
-We assume that the issue encountered comes from the compression of the pdf file to fit the size constraints of the
-submission site. We tried to improve this compression in this new submission. In any case if the paper is accepted, we believe that the uncompressed pdf could be used to obtain the desired quality.
-Regarding the interpretation of the results without the video, we agree with the Reviewer and believe that the accompanying video is essential to the comprehension of the results.
-We considered three options: a) displaying several sequences of contact (as currently done in most cases); b) only displaying 2 or 3 representative consecutive contact stances (currently done
-in the car scenario and in the real robot scenario, for which using a large number of pictures is not readable); c)
-not providing any picture and referring to the video.
-We chose a mix between a) and b), because we wanted the paper to be remain as independent from the video as possible, but are willing to opt for any other option if preferred by the Reviewer.}\done
-
-%~ \quot{In general, the simulation results presented qualitatively in the paper are a bit hard to interpret without the accompanying video, particularly if someone looks at a b/w print of the paper (as it may be looking in T-RO).}
-%~ \as{ThAll of our simulation figures are taken from high resolution rendering in blender, so we wouldn't know how to improve their quality.
-If the Reviewers think the paper would be better without these figures, we are willing to remove them.
-However, we believe that they would make the paper more dependent on the video, so we would prefer to leave them.}
-%\as{\textbf{\textcolor{red}{TODO This is true, although I think he has a crappy printer because I don't think it's that bad.
-I don't know what to do about it though.
-I can't simply remove the Figures, because he tends to say he would like the paper to be more independent from the video...}}}\ndone
-
-\quot{ I also found the description of the environment model being
-used,
-and the way it is used in planning quite superficial. It is said, that
-``an octree data structure'' is used to keep the sampled the
-end-effector positions, and it is ``intersect'' with the environment.
-The idea is OK, I understand that the octree is a sample-based
-representation of the effector's workspace, and using some environment
-model it is possible to decide if the given configuration can provide a
-contact point or not. but how is the environment described ? How
-efficient is this intersection operation ? Does this method require the
-environment to be described by some specific objects (e.g. polygons) ?
-}
-
-\as{Thanks for the comment. We agree that additional information must be added. In Section III we now indicate that the environment is represented as a polygon soup (or 3D mesh), and
-that we assume that it is fully known.  We thus also indicate that we assume the environment and state of the robot are fully known. To perform the intersection computation, we use
-the fcl collision detection library, which natively proposes the intersection method. We added a reference to the library in the paper. The performance of the method is included
-in the ``collision'' column of Table II. We updated the legend of Table II to indicate that this time is negligible in comparison to the other collision detections that occur.
-} 
-\qt{(Section III) \textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains. $O$ is represented
-as a polygon soup (or mesh), where the normal of each surface is known. No further requirement is needed
-by our approach. In this work, we assume the environment is fully known. State uncertainty is out of the scope of the paper. \\
-(Legend of Table II)  The Collision column times includes the (negligible) octree intersection operation necessary to retrieve the candidate samples.}
-\done
-
-\quot{What about the (unavoidable) uncertainty if a world model acquired via
-on-line sensing is considered ? Perhaps some of these issues (e.g.
-uncertainty) are not in the scope of the paper, but I can imagine that
-a robotic researcher who reads the paper and is about to use the open
-source software it advertises is very interested in the answers.}
-\as{That is a good observation. Although the integration of on-line sensing and state estimation in the planner is an issue currently considered by our team, it is indeed
-out of the scope of this paper. As reported in the previous answer we indicated it in Section III. We think that our planner could work similarly with 
-a sensing system that would output data easily interpreted as a polygon soup (as for a LIDAR), otherwise additional treatment would be required.
-In any case additional work would be required to plan motions robust to the uncertainty of the sensing and the estimation.
-} \done
-
-\quot{I appreciate the thorough presentation of results in the paper, however
-it seems that this presentation is somewhat biased towards simulation
-results. Simulations are perfectly OK for planning algorithms, but if
-you decided to present the experimental results (HRP-2), then at least some
-illustrations should be included in the paper. now they are only in the
-accompanying video.}
-\as{We agree with the remark, and included some illustrations of the experimental results on HRP-2 in the paper.}
-%~ We now elaborate on the validity of the simulation with respect to an experiment on the real robot.
-
-\done
-
- \quot{also a comparison between the results of the same
-task(s) executed in dynamic simulation and on the real robot should
-convince the reader that the planning method's good performance does
-not depend too much on the idealized simulation environment. }
-\as{As also suggested by Reviewer 1, we added a paragraph
-in the paper (Section VIII.A)  to provide more information on the simulation we use to validate our motions.
-This simulation is an implementation of the state of the art approaches used by the community.
-Of course, it does not guarantee that the motion could be executed on the robot, but it is used extensively by the community as a reasonable feasibility test.
-The suggestion of the Reviewer to compare the simulation results with those on the real robot is interesting.
-However, in our simulations HRP-2 is torque controlled, whereas in our experiments HRP-2 is position controlled.
-This makes the usefulness of a comparison between simulation and experiments extremely doubtful, because we would be comparing motions obtained with different controllers.
-For these reasons, we prefer not to do this comparison, although we plan to do it in the future, since we are currently working on making HRP-2 torque controlled.}
-\qt{(Section VIII.A) To generate continuous movements from our contact plans we used either the framework proposed in [1], or our own implementation of a $\mathcal{P}_3$ solver (Appendix C).
-The resulting movements have been validated either on the real HRP-2 robot (details can be found in [1]), or with our dynamic simulator, based on a state-of-the-art algorithm [25].
-In the simulations we controlled the robot with a standard inverse-dynamics controller[31].
-This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
-The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
-The companion video shows the obtained motions.}
-
-\section{Answers to Reviewer 6}
-\quot{ My primary concern, however,
-is that the elements described in this paper have been described in
-previous published work, which makes the particular contributions of
-this work unclear. To be specific, the reachability-guided planner
-appears identical to the planner in [20] (as stated by the authors in
-the end of Section I.B.), and the optimal control formulation is from [1]
-(again as stated in I.B.).}
-\as{Thanks for the comment.
-Indeed, the Reviewers agreed that the contribution with respect to previous work was not clearly stated.
-Here we reproduce the answer we made to Reviewer I: As mentioned in the summary, our planner is almost identical to the one in [20], but the paper comprises important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
-for checking static equilibrium.
-The main novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robot models that are kinematically more constrained than the virtual avatars from [20].
-This is now explained in details in the introduction.}
-\qt{(Section I.B, repeated from page 5) 
-The present paper is an extension of our ISRR conference paper [20]. 
-As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section IV and V). 
-However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section V.B.2), a novel criterion for
-static equilibrium (Section VI), and the release of the source code of the planner (Section VII). \\
-The other novelty of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
-To validate our contact plans we introduced a complete framework for multi-contact motion synthesis.
-This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1].
-Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28].
-We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution. \\
-Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
-thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.}\done
-
-\quot{The authors highlight a particular
-contribution of this work to be the efficient robust balance criterion,
-but I had trouble understanding this particular contribution. Does this
-refer to the approximation of equilibrium feasibility by contact
-reachability? In that case, how does this differ from just applying the
-planner in [20] directly? Or does this refer to the robustness
-parameter $b_0$ in Appendix B Eq. (15)?}
-\as{Indeed, this part was also not clear. The robust balance criterion (renamed equilibrium criterion) indeed refers to the parameter $b_0$ in Appendix B (now Section VI). The solved
-LP replaces the static equilibrium test performed in [20], which was not computationally efficient, and took up to 4ms per candidate to compute. We updated the paper to explain
-this in details.}
-\qt{(Section I.B) We also improve the computational performance
-of the method by replacing the static equilibrium test with a more efficient test based on an
-new LP formulation presented in Section VI.}
-\done
-
-\quot{In addition, sections IV and V are very similar to the description of
-the planner in [20]. That's not surprising, of course, but there is no
-indication in those sections that the work being presented is a
-rephrasing of an earlier publication.[\dots] }
-\as{We agree with the comment, and indeed made this clearer in the new submission.}
-\qt{(Section I.B) Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.}\done
-
-\quot{I do, however,
-believe that the fact that the technical approaches are rephrasings of
-prior work should be made more clear (or that sections which can be
-replaced by citations of [20] and [1] be removed).}
-\as{Thanks for the advice.
-However, the suggestion of replacing some sections with references to [20] and [1] is in contradiction with the other Reviewers, who thought that the paper should be self-contained.
-The associate editor decided that the paper must be self contained, so we maintained the indicated sections, but made clearer the fact that they are rephrasing of previous work. }\done
-
-\quot{*I think I understand what the authors mean by solving at
-"interactive" rates, but this is worth defining. In my mind, it means
-something like "before the human operator gets bored", or roughly > 0.5
-Hz. But since the authors make the claim that this is the "first
-interactive implementation of a contact planner", then that statement
-should be made quantitative and verifiable}
-\as{We agree with the relevant comment, and updated the paper accordingly.}
-\qt{(End of Section I.A.) We define an interactive planner as one for which the time to plan a motion is in the same order of magnitude as the time to execute it (For instance, computing one contact change should take about one or two seconds).}\done
-
-\quot{* In Section I.A. the authors refer to Bretl's work on multi-contact
-planning as the inspiration for the P1,P2,P3 decomposition of the
-problem. But it is not, as the authors state, always necessary to
-decompose motion planning into choosing a series of quasi-static
-configurations. There are a variety of feasible motions with no static
-equilibrium configurations, so it is worth mentioning that this
-decomposition only solves a subset of motion planning problem.}
-\as{We apologize if this was not made clear enough in our first submission. In the conclusion, we indicate that the planner is limited to specific scenarios (where quasi-static motion
-is always an option) ``\textit{Our method applies to any scenario where at least one
-contact friction cone contains the direction opposed to the
-gravity (i.e. quasi-flat)[\dots]}''. We updated the state of the art to indicate that the decomposition in general is restricitve.}
-\qt{(End of before last paragraph of Section I.A) A limitation with
-this kind of approach, shared by our own method, is that most of the existing planners only address a subset of the problem, because their ability to find a solution depends
-on the existence of quasi-static equilibrium configurations along a feasible path, which is too restrictive in the general case.}\done
-
-\quot{* Also in Sect. I.A, the authors state that "optimization-based
-approaches only converge locally". This is not necessarily true, as
-convex or mixed-integer convex optimizations can (sometimes) be solved
-to a verifiable global optimum. I would suggest removing this sentence
-entirely.}
-\as{Thanks for the suggestion. Our phrasing was indeed misleading and inaccurate. Our intention was to explain that the geometry of the scene makes the problem highly non-convex, and
-that approaches such as the CIO, if not initialized with an appropriate initial guess, might never get out of the local minima.
-We followed the Reviewer's advice and simply removed the sentence.
-}\done
-
-\quot{* I am not sure why the $C_{reach}$ hypothesis is described as the ``strong''
-hypothesis in, e.g., Sect. IV. Is there a corresponding weak
-hypothesis?}
-\as{We apologize for the confusion, that comes from an inaccurate use of English. What we meant by ``strong'' is that the hypothesis is constraining, because
-it defines a solution space different from the real one, and that we need to demonstrate that it makes sense.  We removed ``strong'' and replaced it by ``main'' to clarify. }\done
-
-\quot{* In Sect. IV.B, the authors state that the ideal shape $B^*$ ``has no
-explicit definition''. For a particular configuration  $q_0$, $B^*$ is simply
-the surface of the robot, right? So I assume that this claim refers to
-there being no explicit definition of B* which is the volume around
-$B^*(q_0)$ for every $q_0$? I think I agree with the authors, but I found
-this claim hard to follow and would appreciate some clarification}
-\as{We updated the Section IV B.
-to better explain our claim.
-Indeed, as the Reviewer states $B^*$  is the volume for every $q_0$.
-If such a volume existed, we would know exactly whether or not a given root configuration can result in a valid contact configuration.
-We of course doubt that the closed form representation of $B^*$ actually exists. }
-\qt{The ideal shape $B^*, W^0 \subset B^* \subset B^{\textrm{\it Suf}}$ would define a necessary \textbf{and} sufficient condition for contact creation. 
-It would guarantee that any root configuration $\mathbf{q}^{0} \in B^*$ would result in a contact configuration, while any $\mathbf{q}^{0} \notin B^*$ could not.
-To our knowledge  $B^*$ has no explicit definition.
-Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.}\done
-
-\quot{* In Sect. V.A., the authors mention that the ability to plan a contact
-break and creation with the same limb allows the creation of dynamic
-motions. This is quite an important ability for an otherwise
-quasi-static planner, but this ability isn't described again until the
-appendix. It would be worth explaining why this is true in the paper
-itself. }
-\as{Thanks for the suggestion. We added a paragraph to better explain this in Section V.A.}
-\qt{This representation is sufficient to describe all the contact phases, because the single support phase is implicitely described.
-Furthermore it removes the need to compute a single-support quasi-static configuration as in the example.
-Indeed, there might be a case where no quasi-static solution exists for the single support phase (because of the environment), but
-there exists a dynamic motion connecting the two double support states.
-Such motion will be computed by our framework, because the quasi-static constraint is only required at the contact planning phase; as shown in the companion video, and explained in Appendix C, our framework is able to produce dynamic motions.
-}
-\done
-
-\quot{In Sect. V.B, the authors mention inserting an additional state
-between contact configurations if two or more contacts cannot be
-maintained. Shouldn't this step appear somewhere in Alg. 1?}
-\as{We apologize for the lack of clarity. This is actually handled by the method \textsc{RepositionContacts}. We now realize that the name is misleading. Indeed,
-\textsc{RepositionContacts} either adds a new contact, or tries to reposition one already existing. In Appendix D, we state:
-\textit{Otherwise, the method \textsc{RepositionContacts} is called. It repositions one end effector (either a free
-limb, or the oldest active contact) towards a new contact
-position if possible. }
-
-We changed the name to \textsc{IntermediateContactState}, with the hope that it makes things clearer.}\done
-
-\quot{* In Appendix A.A, it is mentioned that step 2 in the workspace
-computation is conservative, with reference to figure 17-right. But
-Fig. 17-right is the simplification of the convex hull, corresponding
-to step 3, and it's actually hard to tell from the figure that it is
-conservative. Not knowing anything about Blender's decimate tool, I
-assume it is operating by removing vertices, which would indeed be
-conservative, but some clarification would be useful here.}
-\as{Indeed, the decimate tool removes vertices, and this was not explained clearly in the paper. We updated the text accordingly.}
-\qt{The resulting polytope can contain a very large number of faces.
-A last step is thus to simplify it with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}).
-This tool removes a user-defined amount of vertices (and faces) of the polytope, thus resulting
-in a convervative approximation of the original shape. For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.}\done
-
-\quot{* Finally, I would suggest providing explicit git SHAs or tags for all
-of the github links. I've personally made the mistake of linking to a
-github project in a paper, only to have that project change and
-eventually delete the specific code I was citing. }
-\as{The Reviewer is absolutely right. We indeed encounter the issue with other researchers that are currently using our planner. 
-To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project, on a separate branch
-dedicated to the installation of the planner.
-Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the Reviewer.}
-\qt{(Section VII) Both HPP and our planner can be simply downloaded and compiled by following the instructions on
-\url{https://humanoid-path-planner.github.io/hpp-doc/download.html?branch=rbprm}. \\
-(Appendix C) To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software.
-}
-\done
-
-\section{Diff files between the original and current submission}
-We now join a diff file comparing the original and current submissions.
-The text appearing in blue denotes the additions. We used the latexdiff tools to generate this file, which is not fully compatible with all the latex packages that we use.
-For this reason, errors appear in the document, in particular regarding the subsections numbering. However, we think that this diff file will help the Reviewers locate precisely
-where the modifications occured in the document. The Reviewers will also notice that we reformulated the description our P3 solver in Appendix C, to better clarify what variables
-are used, and what variables are implicit (that is computed from the others).
-
-\includepdf[pages={-}]{diff.pdf}
+\section{Answers to Reviewer 2}
 

### 2017-12-12 body/equilibrium.tex
r1 comment 2 LP

@@
 
-Indeed, from Slater's conditions \citep{Boyd:2004:CO:993483}, we know that the optimal values of an LP and its dual are equal.
-Therefore the optimal value $\bm{\nu}^*$ gives the optimal value $b_0^*$ through the equality $b_0^* = (\mathbf{D} \mathbf{c} + \mathbf{d})^T \bm{\nu}^*$.
+The optimal value $\bm{\nu}^*$ gives the optimal value $b_0^*$ through the equality $b_0^* = (\mathbf{D} \mathbf{c} + \mathbf{d})^T \bm{\nu}^*$.
  %~ Thus the optimal value of the LP~\eqref{eq:dual} is indeed the optimal $b_0$.

### 2017-12-12 review_tro_2/reviewer_response.tex
reviewer 2 comment 2 fcl

@@
 
-\section{Summary of the modifications}
-
 First of all, we would like to thank all the Reviewers and the associate editor (AE) for considering our resubmission.
 We are glad to find them satisfied with our modifications, and made our best to answer to the remaining comments as well as possible.
-
 In the following lines, we will quote all the remaining comments and answer to explain how exactly we updated the paper to address them.
@@
 \as{TODO}
+\leavevmode \\
 
@@
 
+\leavevmode \\
+
 \quot {... and to refer
@@
 
+\quot{First of all, I would like to thank the authors for the significant
+improvements to the paper. The contributions are much clearer, the
+algorithms are better described, and I believe that my initial concerns
+about the paper have been addressed. In particular, I would like to
+thank the authors for the detailed inline responses.}
+
+\as{We in turn thank the reviewer, who significantly helped us improving the paper with his comments.}
+
+\leavevmode \\
 \quot {It looks like that sentence was not actually removed from the
@@
 
-\as{Indeed, due to a misguided merge, the sentence was reintroduced against our will. We apologize for that, and guarantee that this time it was indeed removed!}
+\as{Indeed, the sentence was somehow not removed from our latest version. We apologize for that, and guarantee that this time it was indeed removed!}
 
+\leavevmode \\
 \quot {
@@
 
+\leavevmode \\
 \section{Answers to Reviewer 2}
+\quot{All the issues I raised in the first review have been addressed in
+the revised version. I'm satisfied with most of the answers and the
+changes made to the paper. In particular the novelty of the paper with
+respect to the conference works is clarified. I appreciate also the new
+structure of the manuscript that integrates material previously
+relegated to the Appendices into the main body. Now, this work seems to
+be self contained enough for a journal. }
+\as{Thanks for the comment. We agree with the reviewer that the paper is better this way.}
+
+\leavevmode \\
+\quot{
+1. I understand the point of view of the authors who focus on
+multi-contact planning and would like to refer only briefly to previous
+work in motion planning for multi-legged, statically-stable robots. In
+principle that is OK, but I suggest to extend the paragraph added at
+the end of Section I.A by few sentences that clarify what are the
+advantages of your motion planning system with respect to these older
+approaches. You are focusing on humanoids, for which these gains are
+pretty obvious, but what about multi-legged robots? As an inspiration I
+recommend this youtube video showing a simulated hexapod that does
+something similar to your HyQ simulated example
+\url{https://www.youtube.com/watch?v=Xio1TbjAK80}}
+
+\as{TODO}
+\leavevmode \\
+
+\quot{
+While the answer about the environment model is satisfying, I think
+that you should no only refer to the FCL library in a footnote, but to
+cite the paper that explains how it works:
+http://ieeexplore.ieee.org/document/6225337/?reload=true
+}
+\as{Thanks for the suggestion. The citation has indeed been included, as reference TODO}
+\leavevmode \\
 

### 2017-12-12 body/contact.tex
reviewer 2 comment 2 fcl

@@
 \item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
-\item At runtime, when contact creation is required, intersect the octree and the environment\footnote{this operation is achieved natively by the fcl library https://flexible-collision-library.github.io/} to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
+\item At runtime, when contact creation is required, intersect the octree and the environment to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c)).
+This operation is achieved natively by the fcl library \cite{6225337};
 \item Use a user-defined heuristic $h$ to sort $S$;

### 2017-12-12 review_tro_2/reviewer_response.tex
reviewer 2 3 - performance

@@
 that you should no only refer to the FCL library in a footnote, but to
-cite the paper that explains how it works:
-http://ieeexplore.ieee.org/document/6225337/?reload=true
+cite the paper that explains how it works.
 }
@@
 
+\quot{
+In Tab. V you refer to the computation time in units - seconds,
+minutes and hours. If so, you should be more specific as to the
+source(s) of the numbers you give in that table. I understand the fact
+that it is practically impossible to have a fair comparison in that
+point, and the whole community is working toward changing this
+situation (as you explain). But you should write explicitly that you
+took the numbers from the respective papers, or from your
+re-implementation of the algorithms you consider (I think it is not the
+case). Well, the papers [8], [17], [18] are quite old, and comparing
+the computation times achieved on the computers available 10 years ago
+and nowadays isn't convincing. 
+}
+\as{Thanks for the comment.
+Indeed, we did not reimplement the solutions, so we modified the paper to state explicitely that the times indicated are those found in the original papers.
+Although we agree that the times do not stand as proofs, we still believe that the order of magnitude of the time computations remains significant, because an algorithm that took hours to compute 10 years ago is unlikely to perform in less than a few seconds on a modern computer, especially since our implementation is single threaded, and does not take any advantages of recent advances such as GPU programming.
+In any case, the reviewer is right when he says that this section should be clarified.
+Thus, we removed our over-statement ``\textit{Table V presents the computation times for these scenarios, clearly demonstrating that our approach is order of magnitude faster than previous works}'' and added the following comment in Section VIII E.}
+\qt{
+Table V presents the computation times for these scenarios.
+Since no implementation of the previous methods is available, we had no choice but to indicate the times directly taken from their corresponding papers.
+In the case of older contributions such as [8], [17], [18], because of the technological progress, it appears that our results would have been more meaningful if the benchmarks were run on a modern computer.
+However, we believe that the order of magnitudes between the performance of each method remain significant to suggest that our method is faster.
+}
+\leavevmode \\
+
 \end{document}

### 2017-12-12 body/results.tex
reviewer 2 3 - performance

@@
 
-Table~\ref{tab:compprev} presents the computation times for these scenarios, clearly demonstrating that our approach is order of magnitude faster than previous works.
+%~ Table~\ref{tab:compprev} presents the computation times for these scenarios, clearly demonstrating that our approach is order of magnitude faster than previous works.
+Table~\ref{tab:compprev} presents the computation times for these scenarios.
+Since no implementation of the previous methods is available, we had no choice but to indicate the times directly taken from their corresponding papers.
+In the case of older contributions such as~\cite{Hauser06usingmotion},~\cite{DBLP:conf/iser/EscandeKMG08} and~\cite{Bouyarmane2009}, because of the technological progress, it appears that our results would have been more meaningful if the benchmarks were run on a modern computer.
+However, we believe that the order of magnitudes between the performance of each method remain significant to suggest that our method is faster.
 

### 2017-12-12 review_tro_2/reviewer_response.tex
r2 typos

@@
 
+\quot{I found few typos, but I believe that these would be removed in the
+final editing process. 
+The issues raised above are not essentia to the quality of the
+contribution, but you can consider addressing these as an attempt to
+improve the paper. }
+\as{Thanks again for the comment.
+We made a last pass on the paper and indeed found a few remaining typos (negligeable -> negligible, implicitely -> implicitly, Euclidian -> Euclidean, travelled -> traveled, convervative -> conservative).
+More importantly we believe that we addressed all the remaining comments of the reviewer and hope that this will further improve the paper.}
 \end{document}

### 2017-12-12 body/appendix_roms.tex
r2 typos

@@
 This tool removes a user-defined amount of vertices (and faces) of the polytope, thus resulting
-in a convervative approximation of the original shape. For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
+in a conservative approximation of the original shape. For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
 \end{enumerate}

### 2017-12-12 body/contact.tex
r2 typos

@@
 %However, if the contact broken and the contact created sequentially are not on the same effector, our planner outputs two distinct states.
-This representation is sufficient to describe all the contact phases, because the single support phase is implicitely described.
+This representation is sufficient to describe all the contact phases, because the single support phase is implicitly described.
 Furthermore it removes the need to compute a single-support quasi-static configuration as in the example.
@@
 $j$ depends on a user-defined variable, called the discretization step.
-It corresponds to the ratio between the length of the path $\mathbf{q}^0(t)$\footnote{The length of the path is computed as the weighted 6D Euclidian distance
-travelled along the it, with a weight of $0.7$ for the translation part, and $0.3$ for the orientation part.}, and the number
+It corresponds to the ratio between the length of the path $\mathbf{q}^0(t)$\footnote{The length of the path is computed as the weighted 6D Euclidean distance
+traveled along the it, with a weight of $0.7$ for the translation part, and $0.3$ for the orientation part.}, and the number
 of configurations selected along it to create the contact configurations. 

### 2017-12-12 body/appendix_algo.tex
algo 1

@@
 
-First, we define an abstract structure State,
-that describes a contact configuration.
-The use of queues allows a FIFO approach regarding the order 
-in which contacts are tested: we try to replace older contacts first when necessary.
-Thus the algorithm is deterministic even though it can handle acyclic motions. \\
-
-\begin{lstlisting}]
-Struct Limb
-{
-    // Limb Configuration
-    Configuration qk;
-    // Effector position in
-    // world coordinates
-    vector6 pk;
-};
-
-Struct State
-{
-    // root location
-    Configuration q0;
-    // List of limbs not in contact
-    queue<Limb> freeLimbs;
-    // List of limbs in contact
-    queue<Limb> contactLimbs;
-};
-\end{lstlisting}
+%~ First, we define an abstract structure State,
+%~ that describes a contact configuration.
+%~ The use of queues allows a FIFO approach regarding the order 
+%~ in which contacts are tested: we try to replace older contacts first when necessary.
+%~ Thus the algorithm is deterministic even though it can handle acyclic motions. \\
+%~ 
+%~ \begin{lstlisting}]
+%~ Struct Limb
+%~ {
+    %~ // Limb Configuration
+    %~ Configuration qk;
+    %~ // Effector position in
+    %~ // world coordinates
+    %~ vector6 pk;
+%~ };
+%~ 
+%~ Struct State
+%~ {
+    %~ // root location
+    %~ Configuration q0;
+    %~ // List of limbs not in contact
+    %~ queue<Limb> freeLimbs;
+    %~ // List of limbs in contact
+    %~ queue<Limb> contactLimbs;
+%~ };
+%~ \end{lstlisting}
 
@@
 	\Function{Interpolate}{$s0$,$\mathbf{Q}^0$, $MAX\_TRIES$}
-		\State $list<$State$>$ $states = [s0]$
-		\State $nb\_fail = 0$ 
-		\State $i = 1;$ /*Current index in the list*/
+		\State $states \gets (s0)$ \Comment{List of states intialized with $s0$}
+		\State $nb\_fail \gets 0$ 
+		\State $i \gets 1;$ \Comment{Current index in the list}
 		\While {$i < length(\mathbf{Q}^0)$}
-			\State State $pState = last\_element(states)$
-			\State State $s =$ \textsc{GenFullBody}$(pState, \mathbf{Q}^0[i])$
-			\If {$s != NULL $}
-				\State $nb\_fail = 0$
-				\State $i += 1$
+			\State $pState \gets last\_element(states)$
+			\State $s \gets$ \textsc{GenFullBody}$(pState, element(\mathbf{Q}^0,i))$
+			\If {$s \not= 0 $}
+				\State $nb\_fail \gets 0$
+				\State $i \gets i+1$
 				\State \textbf{return} $\mathbf{q}^{0}$
 			\Else
-				\State $nb\_fail += 1$
+				\State $nb\_fail \gets nb\_fail + 1$
 				\If {$nb\_fail == MAX\_TRIES$}
@@
 				\EndIf				
-				\State $s = $\textsc{IntermediateContactState}$(pState)$
+				\State $s \gets $\textsc{IntermediateContactState}$(pState)$
 			\EndIf
-			\State $push\_back(states, s)$
+			\State $push_back(states, s)$
 		\EndWhile
@@
 This repositioning allows to increase the odds that the contact can be maintained at the next step.
+The method \textit{last\_element} returns the last element
+of the list, \textit{element} returns the element contained by a list at a given index, and \textit{push\_back} inserts an element at the end of a list.
 Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.

### 2017-12-12 body/appendix_algo.tex
algo 2

@@
 From the start configuration, given as an input by the user,
-we create the initial state $s0$.
+we create the initial state $s0$. A state is described by a whole-body configuration, as well as the list of currently active contacts.
 Algorithm~\ref{alg:interpolate}  is then called with $s0$, as well as the discretized path 
@@
 			\EndIf
-			\State $push_back(states, s)$
+			\State $push\_back(states, s)$
 		\EndWhile
@@
 	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
-		\State State $newState$
-		\State $newState.q0 = \mathbf{q}^0$
-		\State $newState.freeLimbs = pState.freeLimbs$
+		\State $newState \gets create\_state(q0, FreeLimbs(pState))$
 		\State /*First try to maintain previous contacts*/
-		\State $nbContactsBroken = 0$
-		\For {\textbf{each} Limb $k$ in $pState.contactLimbs$}
+		\State $nConBroken \gets 0$
+		\For {\textbf{each} $k$ in $ContactLimbs(pState)$}
 			\If {$!$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
-				\State $nbContactsBroken += 1$
-				\If {$nbContactsBroken > 1$}				
-					\State \textbf{return} $NULL$
+				\State $nConBroken \gets nConBroken +1$
+				\If {$nConBroken > 1$}				
+					\State \textbf{return} $0$
 				\EndIf				
-				\State $push(newState.freeLimbs,k)$
+				\State $add\_free\_limb(newState,k)$
 			\Else 					
-				\State $push(newState.contactLimbs,k)$
+				\State $add\_contact\_limb(newState,k)$
 			\EndIf
 		\EndFor
-		\For {\textbf{each} Limb $k$ in $pState.freeLimbs$}
+		\For {\textbf{each} $k$ in $FreeLimbs(pState)$}
 			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
-				\State $push(newState.contactLimbs,k)$
-				\State $remove(newState.freeLimbs,k)$		
+				\State $add\_contact\_limb(newState,k)$
 				\State \textbf{return} $newState$
@@
 		\Else
-			\State \textbf{return} $NULL$
+			\State \textbf{return} $0$
 		\EndIf
@@
 The method \textsc{IsInStaticEquilibrium} returns whether a given state is in static equilibrium.
+$ContactLimbs$ (respectively $FreeLimbs$) returns the list of limbs that are in contact (respectively \textit{not} in contact) at a given state.
+$add\_contact\_limb$ (respectively $add\_free\_limb$)  marks a limb as in contact (respectively \textit{not} in contact) at a given state.
+%~ $add\_contact\_limb(newState,k)$ marks a limb as in contact at a given state.
 

### 2017-12-12 body/appendix_algo.tex
uniform syntax

@@
 This repositioning allows to increase the odds that the contact can be maintained at the next step.
-The method \textit{last\_element} returns the last element
-of the list, \textit{element} returns the element contained by a list at a given index, and \textit{push\_back} inserts an element at the end of a list.
+The method \textit{length} returns the length of a list, \textit{last\_element} returns the last element
+of a list, \textit{element} returns the element contained by a list at a given index, and \textit{push\_back} inserts an element at the end of a list.
 Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.
@@
 	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
-		\State $newState \gets create\_state(q0, FreeLimbs(pState))$
-		\State /*First try to maintain previous contacts*/
+		\State $newState \gets CreateState(q0, ContactLimbs(pState))$
+		%~ \State /*First try to maintain previous contacts*/
 		\State $nConBroken \gets 0$
@@
 				\EndIf				
-				\State $add\_free\_limb(newState,k)$
+				\State $MarkFree(newState,k)$
 			\Else 					
-				\State $add\_contact\_limb(newState,k)$
+				\State $MarkContact(newState,k)$
 			\EndIf
@@
 			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
-				\State $add\_contact\_limb(newState,k)$
+				\State $MarkContact(newState,k)$
 				\State \textbf{return} $newState$
@@
 
-The method \textsc{IsInStaticEquilibrium} returns whether a given state is in static equilibrium.
-$ContactLimbs$ (respectively $FreeLimbs$) returns the list of limbs that are in contact (respectively \textit{not} in contact) at a given state.
-$add\_contact\_limb$ (respectively $add\_free\_limb$)  marks a limb as in contact (respectively \textit{not} in contact) at a given state.
+The method \textsc{IsInStaticEquilibrium} returns whether a given state is in static equilibrium. $CreateState$ creates a new state, given
+a configuration and the list of active contacts.
+$ContactLimbs$ (respectively $FreeLimbs$) returns the list of limbs that are in contact (respectively \textit{not} in contact).
+$MarkContact$ (respectively $MarkFree$)  marks a limb as in contact (respectively \textit{not} in contact).
 %~ $add\_contact\_limb(newState,k)$ marks a limb as in contact at a given state.

### 2017-12-12 body/appendix_algo.tex
pseudo code fini

@@
 				\State $i \gets i+1$
-				\State \textbf{return} $\mathbf{q}^{0}$
+				%~ \State \textbf{return} $\mathbf{q}^{0}$
 			\Else
 				\State $nb\_fail \gets nb\_fail + 1$
-				\If {$nb\_fail == MAX\_TRIES$}
+				%~ \If {$nb\_fail == MAX\_TRIES$}
+					%~ \State \textbf{return} $FAILURE$
+				%~ \EndIf				
+				\State $s \gets $\textsc{IntermediateContactState}$(pState)$
+				\If {$s == 0 \lor nb\_fail == MAX\_TRIES$}
 					\State \textbf{return} $FAILURE$
 				\EndIf		
-				\State $s \gets $\textsc{IntermediateContactState}$(pState)$
 			\EndIf
@@
 	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
-		\State $newState \gets CreateState(q0, ContactLimbs(pState))$
+		\State $newState \gets CreateState(\mathbf{q}^0, ContactLimbs(pState))$
 		%~ \State /*First try to maintain previous contacts*/
@@
 		\EndFor
-		\For {\textbf{each} $k$ in $FreeLimbs(pState)$}
-			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
+		\For {\textbf{each} $k$ in $FreeLimbs(newState)$}
+			\If {\textsc{GenerateContact}$(newState,k)$}	
 				\State $MarkContact(newState,k)$
@@
 $MarkContact$ (respectively $MarkFree$)  marks a limb as in contact (respectively \textit{not} in contact).
+These methods follow a FIFO approach: the first Limb chronologically marked as in contact (respectively not in contact) is returned first by $ContactLimbs$ (respectively $FreeLimbs$).
+This allows the algorithm to be deterministic even though it can handle acyclic motions.
 %~ $add\_contact\_limb(newState,k)$ marks a limb as in contact at a given state.
@@
 
-\textsc{GenerateContact}$(\mathbf{q}^0,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
- It generates a contact configuration in static equilibrium, and assigns the corresponding configuration to $k$.
+\textsc{GenerateContact}$(state,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
+It generates a contact configuration in static equilibrium, and assigns the corresponding configuration of the Limb $k$ to the state configuration.
 If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.
@@
 	\Function{IntermediateContactState}{$state$}        
-		\State $i=0$
-		\While {$i<length(states.freeLimbs)$}
-			\State Limb $k = pop(states.freeLimbs)$
-			\If {\textsc{GenerateContact}$(state.q0,k)$}	
-				\State $push(newState.contactLimbs,k)$			
-				\State \textbf{return}
-			\Else
-				\State $i+=1$
-				\State $push(states.freeLimbs,k)$		
+        \State $newState \gets state$
+		\For {\textbf{each} $k$ in $FreeLimbs(newState)$}
+			\If {\textsc{GenerateContact}$(newState,k)$}	
+				\State $MarkContact(newState,k)$			
+				\State \textbf{return} $newState$
+			%~ \Else
+				%~ \State $MarkFree(newState,k)$	\comment{Mark }
+				%~ \State $push(states.freeLimbs,k)$		
 			\EndIf
-		\EndWhile
-		\State $i=0$
-		\While {$i<length(states.contactLimbs)$}
-			\State Limb $k = pop(states.contactLimbs)$
-			\State Limb $copy = k$
-			\State $i+=1$
-			\If {\textsc{GenerateContact}$(state.q0,k)$}	
-				\State $push(newState.contactLimbs,k)$			
-				\State \textbf{return}
-			\Else
-				\State $push(newState.contactLimbs,copy)$	
+		\EndFor
+		\For {\textbf{each} $k$ in $ContactLimbs(newState)$}
+			\If {\textsc{GenerateContact}$(newState,k)$}	
+				\State /*Account for repositioning in FIFO queue*/		
+				\State $MarkContact(newState,k)$
+				\State \textbf{return} $newState$
+			%~ \Else
+				%~ \State $MarkFree(newState,k)$	\comment{Mark }
+				%~ \State $push(states.freeLimbs,k)$		
 			\EndIf
-		\EndWhile
+		\EndFor
+        \State
 		/*Fails if impossible to relocate any effector*/
-		\State \textbf{return} $FAILURE$
+		\State \textbf{return} $0$
+        %~ 
+        %~ 
+		%~ \State $i \gets 0$
+		%~ \While {$i<length(states.freeLimbs)$}
+			%~ \State Limb $k = pop(states.freeLimbs)$
+			%~ \If {\textsc{GenerateContact}$(state,k)$}	
+				%~ \State $push(newState.contactLimbs,k)$			
+				%~ \State \textbf{return}
+			%~ \Else
+				%~ \State $i+=1$
+				%~ \State $push(states.freeLimbs,k)$		
+			%~ \EndIf
+		%~ \EndWhile
+		%~ \State $i=0$
+		%~ \While {$i<length(states.contactLimbs)$}
+			%~ \State Limb $k = pop(states.contactLimbs)$
+			%~ \State Limb $copy = k$
+			%~ \State $i+=1$
+			%~ \If {\textsc{GenerateContact}$(state.q0,k)$}	
+				%~ \State $push(newState.contactLimbs,k)$			
+				%~ \State \textbf{return}
+			%~ \Else
+				%~ \State $push(newState.contactLimbs,copy)$	
+			%~ \EndIf
+		%~ \EndWhile
+		%~ /*Fails if impossible to relocate any effector*/
+		%~ \State \textbf{return} $FAILURE$
 	\EndFunction

### 2017-12-12 body/appendix_algo.tex
typos

@@
 From the start configuration, given as an input by the user,
-we create the initial state $s0$. A state is described by a whole-body configuration, as well as the list of currently active contacts.
+we create the initial state $s0$. A state is described by a whole-body configuration, as well as the list of currently active contacts and their associated 6D positions.
 Algorithm~\ref{alg:interpolate}  is then called with $s0$, as well as the discretized path 
@@
 	\Function{Interpolate}{$s0$,$\mathbf{Q}^0$, $MAX\_TRIES$}
-		\State $states \gets (s0)$ \Comment{List of states intialized with $s0$}
+		\State $states \gets (s0)$ \Comment{List of states initialized with $s0$}
 		\State $nb\_fail \gets 0$ 
@@
 The method \textsc{IsInStaticEquilibrium} returns whether a given state is in static equilibrium. $CreateState$ creates a new state, given
-a configuration and the list of active contacts.
+a root configuration and the list of active contacts.
 $ContactLimbs$ (respectively $FreeLimbs$) returns the list of limbs that are in contact (respectively \textit{not} in contact).

### 2017-12-13 review_tro_2/reviewer_response.tex
pseudo code fini
webpage and links

@@
 
-\as{TODO}
-\leavevmode \\
-
+\as{We rewrote Algorithm 1, 2 and 3 accordingly to the recommendations of the Associate Editor, which we thank for the comment that allowed us to spot
+a typo in pseudocode. Furthermore, it also allowed us to reduce the length of the pseudocode.}
+\leavevmode 
 \quot {The URLs in the paper seem rather brittle and are not suitable for
@@
 
-\as{TODO}
+\as{Thanks for the comment. In fact, the github urls are not pointing to branches, but to release versions of the code, presented as zip files, which are in fact permanent
+links. However, we agree with the Associate Editor that regrouping all the data and information on a single web page is relevant. So, we uploaded the source code to the first author
+website (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html}), and provided all the required information to install the material in it.
+The user has the choice to download the code either from the github permanent link or directly from the author's website directly.}
 
-\leavevmode \\
-
+\leavevmode 
 \quot {... and to refer
@@
 
-\as{TODO}
+\as{Regarding the decimate tool from Blender, we proceeded as follows: in the paper, we mention a SIGGRAPH paper as one option to reduce mesh complexity, and mention that equivalent methods are proposed by many 3D authoring tools, such as Blender which we used.
+Then, in the static web page, we remind this piece of information and provide a permanent link to the version of Blender we used, along with the documentation that explains how to use the decimate tool.}
+
+\qt{The resulting polytope can contain a very large number of faces.  A last step is thus to simplify it using an incremental decimation method[37].
+Variations of this method are commonly implemented in most authoring tools.
+For our experiments we used the Blender decimate tool.
+Details of its use can be found in the static webpage associated with this paper \url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#decimate}.\\}
 
@@
 }
-\as{Thanks for the suggestion. The citation has indeed been included, as reference TODO}
+\as{Thanks for the suggestion. The citation has indeed been included, as reference [31]TODO}
 \leavevmode \\

### 2017-12-13 body/appendix_optim.tex
final changes from rev 1
webpage and links

@@
 %~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
-To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software.
+To address the other scenarios, we propose a new implementation, entirely open source (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#optim} ), which can be integrated directly in our motion planner software.
 This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
@@
 
-where $\dot{L}$ is the angular momentum expressed at the com.
+\noindent where $\dot{L}$ is the angular momentum expressed at the com.
 Eq.~\ref{eq:new_eul_acc} defines a 6-dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,

### 2017-12-13 body/appendix_roms.tex
webpage and links

--- 2017-01-12 body/appendix_roms.tex
@@
 \item For each configuration, store the 3D position of the end effector joint relatively to the root of $R^k$; then compute the convex hull of the resulting point cloud;
-\item The resulting polytope can contain a very large number of faces.
-A last step is thus to simplify it in a conservative way with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}).
+\item The resulting polytope can contain a very large number of faces.  A last step is thus to simplify it using an incremental decimation method~\cite{Garland:1997:SSU:258734.258849}.
+Variations of this method are commonly implemented in most authoring tools.
+For our experiments we used the blender decimate tool.
+Details of its use can be found in the static webpage associated with this paper (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#decimate}).
+For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
+%~ (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}).
 For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.

### 2017-12-13 body/results.tex
r2 typos
webpage and links

@@
 In the simulations we controlled the robot with a standard inverse-dynamics controller~\cite{DelPrete2015b}.
-The code source of the simulator is available at \url{https://github.com/andreadelprete/pinocchio_inv_dyn/releases/tag/rbprm}.
+The code source of the simulator is available at \url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#simulator}.
 This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
@@
 column indicates the average time necessary to compute one contact transition.
-The Collision column times includes the (negligeable) octree intersection operation necessary to retrieve the candidate samples.}
+The Collision column times includes the (negligible) octree intersection operation necessary to retrieve the candidate samples.}
 \label{tab:requestime}

### 2017-12-13 review_tro_2/reviewer_response.tex
typos

@@
 
-\as{Thanks for the comment. In fact, the github urls are not pointing to branches, but to release versions of the code, presented as zip files, which are in fact permanent
+\as{Thanks for the comment. In fact, the github urls are not pointing to branches, but to release versions of the code, presented as zip files, which make them permanent
 links. However, we agree with the Associate Editor that regrouping all the data and information on a single web page is relevant. So, we uploaded the source code to the first author
 website (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html}), and provided all the required information to install the material in it.
-The user has the choice to download the code either from the github permanent link or directly from the author's website directly.}
+The user has the choice to download the code either from the github permanent link or directly from the author's website.}
 

### 2017-12-13 review_tro_2/reviewer_response.tex
r2 quadruped ? finished ?

@@
 
-\qt{The resulting polytope can contain a very large number of faces.  A last step is thus to simplify it using an incremental decimation method[37].
+\qt{The resulting polytope can contain a very large number of faces.  A last step is thus to simplify it using an incremental decimation method[38].
 Variations of this method are commonly implemented in most authoring tools.
@@
 
-\as{TODO}
+\as{Thanks for the comment. We added the following paragraph in the Section.}
+\qt{The following state of the art focuses on contributions proposed for humanoid robots, although our method is also demonstrated on quadruped robots such as HyQ, for which related work also exists [6].
+With robots using more than two legs for locomotion, different gait modes can be used to cross cluttered environments, allowing them to remain in a cyclic context.
+In these works collision avoidance is often treated as a height issue (assuming that all the obstacles are on the ``ground'', and so are the contacts, which prevents from going under a table, or using a wall as contact location for instance).
+While these approaches are efficient in many cases, we focus on the most generic case, where obstacles are not only on the ground, and cyclic locomotion might not be a solution.}
 \leavevmode \\
@@
 }
-\as{Thanks for the suggestion. The citation has indeed been included, as reference [31]TODO}
+\as{Thanks for the suggestion. The citation has indeed been included, as reference [32]}
 \leavevmode \\

### 2017-12-13 body/intro.tex
r1 - comment 1. remove optimization converges locally
r2 quadruped ? finished ?

@@
 \subsection{State of the art}
+The following state of the art focuses on contributions proposed for humanoid robots, although our method is also demonstrated on quadruped robots such as HyQ, for which related work also exists~\cite{BELTER20116918}.
+With robots using more than two legs for locomotion, different gait modes can be used to cross cluttered environments, allowing them to remain in a cyclic context.
+In these works collision avoidance is often treated as a height issue (assuming that all the obstacles are on the ``ground'', and so are the contacts, which prevents from going under a table, or using a wall as contact location for instance).
+While these approaches are efficient in many cases, we focus on the most generic case, where obstacles are not only on the ground, and cyclic locomotion might not be a solution.
 
@@
 the contacts manually for complex interactions.
-In addition to the limits of the current implementations, optimization-based approaches only converge locally.
 

### 2017-12-13 review_tro_2/reviewer_response.tex
praise the reviewer

@@
 
-\as{We in turn thank the reviewer, who significantly helped us improving the paper with his comments.}
+\as{We in turn thank the reviewer, who significantly helped us improving the paper with her / his comments.}
 
@@
 be self contained enough for a journal. }
-\as{Thanks for the comment. We agree with the reviewer that the paper is better this way.}
+\as{Thanks for the comment. We agree with the reviewer that the paper is clearer this way, thanks to her / his comments.}
 

### 2017-12-13 body/appendix_algo.tex
ongoing reformating algo

@@
 		\While {$i < length(\mathbf{Q}^0)$}
-			\State $pState \gets last\_element(states)$
+			\State $pState \gets lastElement(states)$
 			\State $s \gets$ \textsc{GenFullBody}$(pState, element(\mathbf{Q}^0,i))$
@@
 			\EndIf
-			\State $push\_back(states, s)$
+			\State $pushBack(states, s)$
 		\EndWhile
@@
 
-At each step, \textsc{GenFullBody} is called with the previous state as a parameter, as well
+At each step, \textsc{GenFullBody} (Algorithm~\ref{alg:pi}) is called with the previous state as a parameter, as well
 as a new root configuration. \textsc{GenFullBody} returns a new contact configuration, if it succeeded
 in computing a configuration with only one contact switch occurring.
-Otherwise, the method \textsc{IntermediateContactState} is called.
+Otherwise, the method \textsc{IntermediateContactState} (Algorithm~\ref{alg:repo}) is called.
 It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
 This repositioning allows to increase the odds that the contact can be maintained at the next step.
-The method \textit{length} returns the length of a list, \textit{last\_element} returns the last element
-of a list, \textit{element} returns the element contained by a list at a given index, and \textit{push\_back} inserts an element at the end of a list.
-Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.
+The method \textit{length} returns the length of a list, \textit{lastElement} returns the last element
+of a list, \textit{element} returns the element contained by a list at a given index, and \textit{pushBack} inserts an element at the end of a list.
+%~ Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.
 
@@
 	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
-		\State $newState \gets CreateState(\mathbf{q}^0, ContactLimbs(pState))$
+		\State $newState \gets CreateState(\mathbf{q}^0,$ 
+        \State $\quad \quad \quad \quad  \quad  \quad  \quad  \quad  \quad \quad  \quad ContactLimbs(pState))$
 		%~ \State /*First try to maintain previous contacts*/
@@
 		\For {\textbf{each} $k$ in $ContactLimbs(pState)$}
-			\If {$!$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
+			\If {$\neg$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
 				\State $nConBroken \gets nConBroken +1$
@@
 		\EndFor
-		\If {\textsc{IsInStaticEquilibrium}$(newState)$}
+		\If {$IsInStaticEquilibrium(newState)$}
 			\State \textbf{return} $newState$
@@
 
-The method \textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$ performs inverse kinematics to reach the previous contact position for the Limb.
+The procedure \textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$ performs inverse kinematics to reach the previous contact position for the limb.
 If it succeeds, the new limb configuration is assigned to $k$. If it fails, a random collision free configuration is assigned to $k$.
 
-The method \textsc{IsInStaticEquilibrium} returns whether a given state is in static equilibrium. $CreateState$ creates a new state, given
+The method $IsInStaticEquilibrium$ returns whether a given state is in static equilibrium. $CreateState$ creates a new state, given
 a root configuration and the list of active contacts.
@@
 $MarkContact$ (respectively $MarkFree$)  marks a limb as in contact (respectively \textit{not} in contact).
-These methods follow a FIFO approach: the first Limb chronologically marked as in contact (respectively not in contact) is returned first by $ContactLimbs$ (respectively $FreeLimbs$).
+These methods follow a FIFO approach: the first limb chronologically marked as in contact (respectively not in contact) is returned first by $ContactLimbs$ (respectively $FreeLimbs$).
 This allows the algorithm to be deterministic even though it can handle acyclic motions.
@@
 
-The pseudo code for the method \textsc{IntermediateContactState} is given by Algorithm~\ref{alg:repo}.
+%~ The pseudo code for the method \textsc{IntermediateContactState} is given by Algorithm~\ref{alg:repo}.
 
 \textsc{GenerateContact}$(state,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
-It generates a contact configuration in static equilibrium, and assigns the corresponding configuration of the Limb $k$ to the state configuration.
+It generates a contact configuration in static equilibrium, and assigns the corresponding configuration of the limb $k$ to the state configuration.
 If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.

### 2017-12-13 review_tro_2/reviewer_response.tex
ongoing reformating algo
andrea and nicolas comments

@@
 \newcommand\qt[1]{\begin{quote} \underline{added text in paper}: \textit{#1}\end{quote} \leavevmode \\ }
+\newcommand\rrt[1]{\begin{quote} \underline{\textbf{removed} text from the paper}: \textit{#1}\end{quote} \leavevmode \\ }
 \newcommand\jp{ \leavevmode \\}
@@
 First of all, we would like to thank all the Reviewers and the associate editor (AE) for considering our resubmission.
+In particular we thank them for the quick response they provided.
 We are glad to find them satisfied with our modifications, and made our best to answer to the remaining comments as well as possible.
@@
 
-\as{Regarding the decimate tool from Blender, we proceeded as follows: in the paper, we mention a SIGGRAPH paper as one option to reduce mesh complexity, and mention that equivalent methods are proposed by many 3D authoring tools, such as Blender which we used.
+\as{Regarding the decimate tool from Blender, we proceeded as follows: in the new version of paper, we mention a SIGGRAPH paper as one option to reduce mesh complexity, and mention that equivalent methods are proposed by many 3D authoring tools, such as Blender which we used.
 Then, in the static web page, we remind this piece of information and provide a permanent link to the version of Blender we used, along with the documentation that explains how to use the decimate tool.}
@@
 Finally, we removed the reference to the Slater's conditions since we agree with the reviewer that the procedure is standard.}
+\rrt{Indeed, from Slater’s conditions [33], we know that the optimal values of an LP and its dual are equal.}
 
-\leavevmode \\
 \section{Answers to Reviewer 2}
@@
 In the case of older contributions such as [8], [17], [18], because of the technological progress, it appears that our results would have been more meaningful if the benchmarks were run on a modern computer.
-However, we believe that the order of magnitudes between the performance of each method remain significant to suggest that our method is faster.
+However, since the computation times differ of several orders of magnitude, we believe that these results clearly show the computational benefits of our method.
 }

### 2017-12-13 body/results.tex
andrea and nicolas comments

@@
 In the case of older contributions such as~\cite{Hauser06usingmotion},~\cite{DBLP:conf/iser/EscandeKMG08} and~\cite{Bouyarmane2009}, because of the technological progress, it appears that our results would have been more meaningful if the benchmarks were run on a modern computer.
-However, we believe that the order of magnitudes between the performance of each method remain significant to suggest that our method is faster.
+However, since the computation times differ of several orders of magnitude, we believe that these results clearly show the computational benefits of our method.
 

### 2017-12-13 body/appendix_algo.tex
rewrote algo

@@
 		\State $i \gets 1;$ \Comment{Current index in the list}
-		\While {$i < length(\mathbf{Q}^0)$}
-			\State $pState \gets lastElement(states)$
-			\State $s \gets$ \textsc{GenFullBody}$(pState, element(\mathbf{Q}^0,i))$
+		\While {$i < Length(\mathbf{Q}^0)$}
+			\State $pState \gets LastElement(states)$
+			\State $s \gets$ \textsc{GenFullBody}$(pState, Element(\mathbf{Q}^0,i))$
 			\If {$s \not= 0 $}
@@
 			\EndIf
-			\State $pushBack(states, s)$
+			\State \textsc{PushBack}$(states, s)$
 		\EndWhile
@@
 This repositioning allows to increase the odds that the contact can be maintained at the next step.
-The method \textit{length} returns the length of a list, \textit{lastElement} returns the last element
-of a list, \textit{element} returns the element contained by a list at a given index, and \textit{pushBack} inserts an element at the end of a list.
+The method \textit{Length} returns the length of a list, \textit{LastElement} returns the last element
+of a list, \textit{Element} returns the element contained by a list at a given index, and \textsc{PushBack} inserts an element at the end of a list.
 %~ Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.
@@
 	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
-		\State $newState \gets CreateState(\mathbf{q}^0,$ 
+		\State $newState \gets \textsc{CreateState}(\mathbf{q}^0,$ 
         \State $\quad \quad \quad \quad  \quad  \quad  \quad  \quad  \quad \quad  \quad ContactLimbs(pState))$
@@
 				\EndIf				
-				\State $MarkFree(newState,k)$
+				\State \textsc{MarkFree}$(newState,k)$
 			\Else 					
-				\State $MarkContact(newState,k)$
+				\State \textsc{MarkContact}$(newState,k)$
 			\EndIf
@@
 			\If {\textsc{GenerateContact}$(newState,k)$}	
-				\State $MarkContact(newState,k)$
+				\State \textsc{MarkContact}$(newState,k)$
 				\State \textbf{return} $newState$
@@
 		\EndFor
-		\If {$IsInStaticEquilibrium(newState)$}
+		\If {\textsc{IsInStaticEquilibrium}$(newState)$}
 			\State \textbf{return} $newState$
@@
 If it succeeds, the new limb configuration is assigned to $k$. If it fails, a random collision free configuration is assigned to $k$.
-
-The method $IsInStaticEquilibrium$ returns whether a given state is in static equilibrium. $CreateState$ creates a new state, given
+The method \textsc{IsInStaticEquilibrium} returns whether a given state is in static equilibrium. \textsc{CreateState} creates a new state, given
 a root configuration and the list of active contacts.
 $ContactLimbs$ (respectively $FreeLimbs$) returns the list of limbs that are in contact (respectively \textit{not} in contact).
-$MarkContact$ (respectively $MarkFree$)  marks a limb as in contact (respectively \textit{not} in contact).
+\textsc{MarkContact} (respectively \textsc{MarkFree})  marks a limb as in contact (respectively \textit{not} in contact).
 These methods follow a FIFO approach: the first limb chronologically marked as in contact (respectively not in contact) is returned first by $ContactLimbs$ (respectively $FreeLimbs$).
@@
 			\If {\textsc{GenerateContact}$(newState,k)$}	
-				\State $MarkContact(newState,k)$			
+				\State \textsc{MarkContact}$(newState,k)$			
 				\State \textbf{return} $newState$
 			%~ \Else
-				%~ \State $MarkFree(newState,k)$	\comment{Mark }
+				%~ \State \textsc{MarkFree}$(newState,k)$	\comment{Mark }
 				%~ \State $push(states.freeLimbs,k)$		
@@
 				\State /*Account for repositioning in FIFO queue*/		
-				\State $MarkContact(newState,k)$
+				\State \textsc{MarkContact}$(newState,k)$
 				\State \textbf{return} $newState$
 			%~ \Else
-				%~ \State $MarkFree(newState,k)$	\comment{Mark }
+				%~ \State \textsc{MarkFree}$(newState,k)$	\comment{Mark }
 				%~ \State $push(states.freeLimbs,k)$		

### 2017-12-14 body/appendix_algo.tex
Minor revision of alg section.

@@
 			\EndIf
-			\State \textsc{PushBack}$(states, s)$
+			\State ${PushBack}(states, s)$
 		\EndWhile
@@
 
-At each step, \textsc{GenFullBody} (Algorithm~\ref{alg:pi}) is called with the previous state as a parameter, as well
+At each step of Algorithm~1, \textsc{GenFullBody} (Algorithm~\ref{alg:pi}) is called with the previous state as a parameter, as well
 as a new root configuration. \textsc{GenFullBody} returns a new contact configuration, if it succeeded
-in computing a configuration with only one contact switch occurring.
-Otherwise, the method \textsc{IntermediateContactState} (Algorithm~\ref{alg:repo}) is called.
-It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
+in computing a configuration with only one contact switch occurring~;
+otherwise, the method \textsc{IntermediateContactState} (Algorithm~\ref{alg:repo}) is called.
+It repositions an end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
 This repositioning allows to increase the odds that the contact can be maintained at the next step.
-The method \textit{Length} returns the length of a list, \textit{LastElement} returns the last element
-of a list, \textit{Element} returns the element contained by a list at a given index, and \textsc{PushBack} inserts an element at the end of a list.
+The method \textit{Length} returns the length of a list~; \textit{LastElement} returns the last element
+of a list~; \textit{Element} returns the element contained by a list at a given index~; and \textit{PushBack} inserts an element at the end of a list.
 %~ Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.
@@
 	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
-		\State $newState \gets \textsc{CreateState}(\mathbf{q}^0,$ 
-        \State $\quad \quad \quad \quad  \quad  \quad  \quad  \quad  \quad \quad  \quad ContactLimbs(pState))$
+		\State $newState \gets {CreateState}(\mathbf{q}^0,$ 
+        \State $\quad \quad \quad \quad  \quad  \quad  \quad  \quad  \quad \quad  \quad \quad\!\! ContactLimbs(pState))$
 		%~ \State /*First try to maintain previous contacts*/
@@
 		\For {\textbf{each} $k$ in $ContactLimbs(pState)$}
-			\If {$\neg$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
+			\If {$\neg$\textit{MaintainContact}$(pState,\mathbf{q}^0,k)$}
 				\State $nConBroken \gets nConBroken +1$
@@
 				\EndIf				
-				\State \textsc{MarkFree}$(newState,k)$
+				\State \textit{MarkFree}$(newState,k)$
 			\Else 					
-				\State \textsc{MarkContact}$(newState,k)$
+				\State \textit{MarkContact}$(newState,k)$
 			\EndIf
@@
 		\For {\textbf{each} $k$ in $FreeLimbs(newState)$}
-			\If {\textsc{GenerateContact}$(newState,k)$}	
-				\State \textsc{MarkContact}$(newState,k)$
+			\If {\textit{GenerateContact}$(newState,k)$}	
+				\State \textit{MarkContact}$(newState,k)$
 				\State \textbf{return} $newState$
@@
 		\EndFor
-		\If {\textsc{IsInStaticEquilibrium}$(newState)$}
+		\If {\textit{IsInStaticEquilibrium}$(newState)$}
 			\State \textbf{return} $newState$
@@
 
-The procedure \textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$ performs inverse kinematics to reach the previous contact position for the limb.
+In Alg.~\ref{alg:pi}, the procedure \textit{MaintainContact}$(pState,\mathbf{q}^0,k)$ performs inverse kinematics to reach the previous contact position for the limb.
 If it succeeds, the new limb configuration is assigned to $k$. If it fails, a random collision free configuration is assigned to $k$.
-The method \textsc{IsInStaticEquilibrium} returns whether a given state is in static equilibrium. \textsc{CreateState} creates a new state, given
+The method \textit{IsInStaticEquilibrium} returns whether a given state is in static equilibrium. \textit{CreateState} creates a new state, given
 a root configuration and the list of active contacts.
 $ContactLimbs$ (respectively $FreeLimbs$) returns the list of limbs that are in contact (respectively \textit{not} in contact).
-\textsc{MarkContact} (respectively \textsc{MarkFree})  marks a limb as in contact (respectively \textit{not} in contact).
+\textit{MarkContact} (respectively \textit{MarkFree})  marks a limb as in contact (respectively \textit{not} in contact).
 These methods follow a FIFO approach: the first limb chronologically marked as in contact (respectively not in contact) is returned first by $ContactLimbs$ (respectively $FreeLimbs$).
@@
 %~ $add\_contact\_limb(newState,k)$ marks a limb as in contact at a given state.
-
+%
 %~ The pseudo code for the method \textsc{IntermediateContactState} is given by Algorithm~\ref{alg:repo}.
-
-\textsc{GenerateContact}$(state,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
+%
+\textit{GenerateContact}$(state,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
 It generates a contact configuration in static equilibrium, and assigns the corresponding configuration of the limb $k$ to the state configuration.
 If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.
+
+Algorithm~\ref{alg:repo} generates an intermediate state by first trying to create a contact with one of the free limbs (trying in a FIFO order) then by repositioning one of the limbs in contact.
 
@@
 		\For {\textbf{each} $k$ in $FreeLimbs(newState)$}
-			\If {\textsc{GenerateContact}$(newState,k)$}	
-				\State \textsc{MarkContact}$(newState,k)$			
+			\If {\textit{GenerateContact}$(newState,k)$}	
+				\State \textit{MarkContact}$(newState,k)$			
 				\State \textbf{return} $newState$
 			%~ \Else
-				%~ \State \textsc{MarkFree}$(newState,k)$	\comment{Mark }
+				%~ \State \textit{MarkFree}$(newState,k)$	\comment{Mark }
 				%~ \State $push(states.freeLimbs,k)$		
@@
 		\For {\textbf{each} $k$ in $ContactLimbs(newState)$}
-			\If {\textsc{GenerateContact}$(newState,k)$}	
+			\If {\textit{GenerateContact}$(newState,k)$}	
 				\State /*Account for repositioning in FIFO queue*/		
-				\State \textsc{MarkContact}$(newState,k)$
+				\State \textit{MarkContact}$(newState,k)$
 				\State \textbf{return} $newState$
 			%~ \Else
-				%~ \State \textsc{MarkFree}$(newState,k)$	\comment{Mark }
+				%~ \State \textit{MarkFree}$(newState,k)$	\comment{Mark }
 				%~ \State $push(states.freeLimbs,k)$		
@@
 			%~ \State Limb $k = pop(states.freeLimbs)$
-			%~ \If {\textsc{GenerateContact}$(state,k)$}	
+			%~ \If {\textit{GenerateContact}$(state,k)$}	
 				%~ \State $push(newState.contactLimbs,k)$			
@@
 			%~ \State $i+=1$
-			%~ \If {\textsc{GenerateContact}$(state.q0,k)$}	
+			%~ \If {\textit{GenerateContact}$(state.q0,k)$}	
 				%~ \State $push(newState.contactLimbs,k)$			

### 2018-02-09 body/intro.tex
final modif from editor

@@
 For instance, computing one contact change should take about one or two seconds.}.
-However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}.
+However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Carpentier2016,Hauser2014, herzog2015trajectory, Park116}.
 In particular, our algorithm proposed in \cite{Carpentier2016} is \gls{interactive}.

### 2018-02-19 final.tex
aggregated document

%% bare_jrnl.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% see http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[journal]{IEEEtran}
%

%~ \setcounter{secnumdepth}{3}
\usepackage{moreverb,url}
\usepackage{makeidx}         % allows index generation
\usepackage[percent]{overpic}
\usepackage{array} 
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{listings}
\usepackage{color}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[Symbol]{upgreek}

\newcommand{\fratop}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\mx}[1]{\mathbf{\bm{#1}}} 				% Matrix symbol
\newcommand{\vc}[1]{\mathbf{\bm{#1}}} 					% Vector symbol
\newcommand{\degree}{\ensuremath{^\circ}}				% define the degree symbol
\newcommand{\pder}[2]{\frac{\partial#1}{\partial#2}}		% partial derivative
\newcommand{\ppder}[2]{\frac{\partial^2 #1}{\partial#2^2}}		% second partial derivative
\newcommand{\refframe}[1]{\mbox{\textless#1\textgreater}}	% to denote a reference frame
%\DeclareMathOperator*{\lexmin}{\text{lex}\!\min}			% lexmin
\DeclareMathOperator*{\argmin}{\arg\!\min}				% argmin
\DeclareMathOperator*{\argmax}{\arg\!\max}				% argmax
\DeclareMathOperator*{\st}{subject\,to}					% subject to
\DeclareMathOperator*{\minimize}{minimize}				% 
\DeclareMathOperator*{\maximize}{maximize}				% 
\DeclareMathOperator*{\find}{find}						% 
\DeclareMathOperator*{\dif}{\mathrm{d}}					% d
\DeclareMathOperator*{\half}{\frac{1}{2}}					% one half
\newcommand{\equilibriumfeasibility}[0]{\glslink{equilibrium feasible}{\textit{equilibrium feasibility}}}	% matrix
\newcommand{\contactreachability}{\textit{contact reachability}}	% matrix
\newcommand{\mat}[1]{\ensuremath{\begin{bmatrix}#1\end{bmatrix}}}	% matrix
\newcommand{\rank}[1]{\text{rank}(#1)}							% rank
\newcommand{\diag}[1]{\text{diag}(#1)}							% diag
\newcommand{\x}{\ensuremath{\times}}
\newcommand{\spac}{\ensuremath{\quad}}						% alias for space in math environment
\newcommand{\dx}[0]{\ensuremath{\delta x}}					% dx
\newcommand{\du}[0]{\ensuremath{\delta u}}					% du
\newcommand{\DX}[0]{\ensuremath{\Delta X}}						% DX
\newcommand{\DU}[0]{\ensuremath{\Delta U}}						% DU
\newcommand{\T}[0]{\ensuremath{\top}}							% transpose symbol
\newcommand{\pinv}[0]{\ensuremath{\dagger}}					% pseudoinverse symbol
\newcommand{\Rv}[1]{\ensuremath{\mathbb{R}^{#1}}}				% set of real-valued vectors
\newcommand{\R}[2]{\ensuremath{\mathbb{R}^{#1\times #2}}}		% set of real-valued matrices
\newcommand{\Spd}[1]{\ensuremath{\mathbb{S}_+^{#1}}}			% set of symmetric positive-definite matrices
\newcommand{\card}[1]{\ensuremath{\left\vert{#1}\right\vert}}			% cardinality of a set
\DeclareMathOperator{\Tr}{Tr}							% trace
\newcommand{\Expect}{{\rm I\kern-.3em E}}				% expectation
\newcommand{\Normal}{\mathcal{N}}					% normal distribution
\newcommand{\Prob}[1]{\text{P}(#1)}						% probability

\newcommand{\Y}{\ensuremath{\mathcal{Y}}}					% 
\newcommand{\Yin}{\ensuremath{\mathcal{Y}_{in}}}					% 
\newcommand{\Yout}{\ensuremath{\mathcal{Y}_{out}}}					%

%\algnewcommand{\algorithmicgoto}{\textbf{go to}}%
%%\algnewcommand{\Goto}[1]{\algorithmicgoto~\ref{#1}}%
%\algnewcommand{\Goto}{\algorithmicgoto\xspace}%
%\algnewcommand{\Label}{\State\unskip}

\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\commentt}[2]{\textcolor{#1}{\textbf{\textit{[#2]}}}} 	% make sure no comments left in the paper
%\newcommand{\commentt}[2]{}							% using this line rather than the previous one you remove all comments
\newcommand{\adnote}[1]{\commentt{blue}{AD: #1}}
\newcommand{\nmnote}[1]{\commentt{OliveGreen}{NM: #1}}

% Another method to track changes
% Examples of usage:
% - This is \added[id=per,remark={we need this}]{new} text.
% - This is \deleted[id=per,remark=obsolete]{unnecessary}text.
% - This is \replaced[id=per]{nice}{bad} text.
% To print the list of change use: \listofchanges
\usepackage{changes}	% use this for the working version
%\usepackage[final]{changes} % use this for the final version
\definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
\newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
\newcommand{\addadp}[1]{\added[id=adp]{#1}}
\newcommand{\repadp}[2]{\replaced[id=adp]{#1}{#2}}
\definechangesauthor[name={Steve Tonneau}, color=blue]{st}
\newcommand{\delst}[1]{\deleted[id=st]{#1}}
\newcommand{\addst}[1]{\added[id=st]{#1}}
\newcommand{\repst}[2]{\replaced[id=st]{#1}{#2}}
\newcommand{\gls}[1]{\textit{#1}}
\newcommand{\glslink}[2]{{#2}}

%~ \usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red,draft]{hyperref}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
%~ \usepackage{glossaries}

% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/pkg/ifpdf
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/pkg/algorithms
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/pkg/algorithmicx

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/array

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/pkg/subfig

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure.
% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
% corrections already built into the system in which case a warning will
% be issued if an attempt is made to load fixltx2e.sty as it is no longer
% needed.
% The latest version and documentation can be found at:
% http://www.ctan.org/pkg/fixltx2e

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/stfloats
% Do not use the stfloats baselinefloat ability as the IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that the IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/pkg/dblfloatfix

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
% Axel Sommerfeldt. This package may be useful when used in conjunction with 
% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
% submissions have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.4.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/pkg/endfloat
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/url
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{An efficient acyclic contact planner for multiped robots}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{Steve~Tonneau,
        Andrea~Del~Prete,~\IEEEmembership{Member,~IEEE,}
        ~Julien~Pettr\'e,~\IEEEmembership{Member,~IEEE,}
        ~Chonhyon~Park,~\IEEEmembership{Student Member,~IEEE,}
        ~Dinesh~Manocha,~\IEEEmembership{Member,~IEEE,}
        ~and~Nicolas~Mansard,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{S. Tonneau, A. Del Prete and N. Mansard are with LAAS-CNRS / Universit\'e de Toulouse, France e-mail: (pro@stevetonneau.fr)}% <-this % stops a space
\thanks{J. Pettr\'e is with Inria, Rennes, France}% <-this % stops a space
\thanks{C. Park and D. Manocha are with UNC, Chapel Hill, USA}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
%~ \markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
%~ {Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
% !TEX root =  ../main.tex
We present a contact planner for complex legged locomotion tasks: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car. The need for such a planner was shown at the Darpa Robotics Challenge, where such behaviors
could not be demonstrated (except for egress).

Current planners suffer from their prohibitive algorithmic complexity, because they deploy a tree of robot
configurations projected in contact with the environment.
%~ computation times, due to the difficulty
%~ of generating contact postures in a high-dimensional solution space that is impossible to sample.

We tackle this issue by introducing a reduction property: the reachability condition. This condition defines
a geometric approximation of the contact manifold, which is of low dimension, presents a Cartesian topology, and can be efficiently sampled and explored.
%~ Informally, the condition verifies that the root configuration of a robot is close, but not too close to obstacles: close to allow contact creation, not too close to avoid collision. Thanks to this condition we decompose 
The hard contact planning problem can then be decomposed into two
%~ sub-problems: first, planning a guide path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm. 
sub-problems: first, we plan a path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, we generate a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm.

The reduction breaks the algorithm complexity encountered
in previous works, resulting in the first interactive
implementation of a contact planner (open source). While
no contact planner has yet been proposed with theoretical
completeness, we empirically show the interest of our framework:
in a
few seconds, with high success rates, we generate complex contact plans for various scenarios and two robots, HRP-2 and HyQ. These plans are validated either in dynamic simulations, or on the real HRP-2 robot.

%~ Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically: in a
%~ few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ, and a dexterous hand. These plans are then validated either in dynamic simulations, or on the real HRP-2 robot, thus demonstrating 
%~ the first interactive multi-contact planner.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Multi contact locomotion, centroidal dynamics, Humanoid robots, legged robots, motion planning
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\newcommand{\citep}[1]{\cite{#1}}	% to denote a reference frame
\newcommand{\citeauthor}[1]{\cite{#1}}	% to denote a reference frame

\section{Introduction}
% !TEX root =  ../main_tro.tex
\newcommand{\Pa}{$\mathcal{P}_1$ }
\newcommand{\Pb}{$\mathcal{P}_2$ }

%~ Planning legged robot locomotion remains an open issue.
\IEEEPARstart{L}{egged} robots move by sequentially creating contacts with the environment.
After years of research, such robots can autonomously walk on flat ground, but struggle to navigate more complex environments.
%~ where
Deciding
where
to
create a 
contact with its feet and
possibly
its
hands
is
nontrivial,
e.g. to climb stairs using a
handrail.
%~ To climb stairs using a handrail, a humanoid robot must plan the contacts to create with its feet and hands. 
%~ Computing such motion is an open motion planning problem called acyclic contact planning.

Most of the complexity of this problem lies in the contact planning, i.e. the underlying decomposition
of the trajectory into contact phases where specific points
of
the
robot body are exerting forces on specific locations
of
the
environment. Tackling this complexity is the main
objective
of this paper.

Once the contact plan is known, efficient approaches exist to generate a dynamically feasible motion~\citep{Carpentier2016}.
 %~ However in general
%~ the contact plan is manually designed by a time consuming trial-and-error procedure. Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in cluttered, human-centered environments, which comprise constraining obstacles (stairs...).
In the specific (and simple) case of gaited biped locomotion on flat ground, 
%~ efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location. Furthermore, 
choosing the effector with which to create a contact is trivial because walking follows a cyclic pattern (the left foot always follows the right foot).
Efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
In the general case planning complex contact interactions is extremely challenging:
At any given time a contact must be chosen between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location). Furthermore, a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles; we say that the contact manifold is foliated~\citep{simeon-manipulation-04}. 
%~ Lastly, the contact manifold has a null measure and is thus impossible to sample. This restriction prevents the use of efficient sampling-based planners.
The
foliation prevents the use of efficient sampling-based
planners
for two reasons. (i) First, each sub-manifold of the         
foliation
has
a zero measure and cannot be directly sampled.         
A
sample
is
rather obtained by sampling a “free flying”         
configuration
and
explicitly projecting it in contact (which is a         
costly
numerical
operation).
(ii) Second, the foliated topology       
turns
the
exploration
by
spreading
a graph of configurations        
(probabilistic
road-map,
rapidly
exploring
random trees) into      
an
inefficient
random
process,
where
many
useless nodes       
are
sampled
on
parallel
sub-manifolds.
The
total
algorithmic       
complexity
of
classical
contact
planners
comes
from
both the        
number
of
graph
nodes
sampled
during
exploration
(ii) and
the cost of the projection when sampling new configurations
(i).

For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}. As a result at the DARPA Robotics Challenge, the participants stated that except for egress, the robots did not use multi contact strategies: they relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.

%~ Obtaining reasonable computation times is critical for a contact planner to be of any practical use.
 Our work aims at breaking the complexity of the acyclic contact planning problem.  To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorics of the contact selection problem. First we introduce a low-dimensional space, called the \textit{contact reachable} space, that can be sampled and mapped efficiently to the contact manifold. Then, given a path computed in the \textit{contact reachable} space, we propose a deterministic algorithm to generate a contact sequence along the path.
This decoupling presents pros and cons discussed in previous related literature, summarized in the following.
%~ whole-body trajectory and the contact location and timings, we decouple path and contact planning, as suggested by \citeauthor{Bouyarmane2009}.
%~ We first plan a path for the geometric root of the robot in a low dimensional space. We formulate this space as a computationally-efficient approximation
%~ of the \textit{equilibrium feasibile} space, that is the space of root configurations for which there exists a joint configuration such that the robot is in static equilibrium.
%~ We then use a deterministic algorithm
%~ This paper presents a geometrical approximation of this condition, and a concrete implementation of such a decoupled planner.
%~ This approximation results from a trade-off between a necessary and a sufficient condition for \textit{equilibrium feasibility}, allowing us to find solutions extremely rapidly, while preserving a high success rate in the demonstrated scenarios.

%~ In the remainder of this introduction, we discuss further the current state of the art. This allows us to situate more precisely our contributions.

\begin{figure*}
  \centering
  \begin{overpic}[width=0.8\linewidth]{figures/workflow}
    %~ \put (6.4,1.8) {\normalsize{Request}} 
    \put (30,10.8) {\large{\color{white}\Pa} }
    %~ \put (31.6,2) {\scriptsize{\color{white}RB-RRT}} 
    \put (66.4,10.8) {\large{\color{white}\Pb} }
    %~ \put (72.5,2) {\tiny{\color{white}Contact}} 
	\put (15,1) {a} 
	\put (50,1) {b} 
	\put (85,1) {c} 
  \end{overpic}
  \vspace{-1em}
  \caption{
    Overview of our two-stage framework. Given a path request between start and goal positions (left image), \Pa is the problem of computing a guide path in the space
    of \textit{equilibrium feasible} root configurations. We achieve this by defining a geometric condition, the \textit{reachability condition} (abstracted with the transparent cylinders on the middle image). \Pb is then the problem of extending the path into a discrete sequence of contact configurations using an iterative algorithm (right image).}
  \label{fig:framework}
\end{figure*}

\subsection{State of the art}
The following state of the art focuses on contributions proposed for humanoid robots, although our method is also demonstrated on quadruped robots such as HyQ, for which related work also exists~\cite{BELTER20116918}. With robots using more than two legs for locomotion, different gait modes can be used to cross cluttered environments, allowing them to remain in a cyclic context. In these works collision avoidance is often treated as a height issue (assuming that all the obstacles are on the ``ground'', and so are the contacts, which prevents from going under a table, or using a wall as contact location for instance). While these approaches are efficient in many cases, we focus on the most generic case, where obstacles are not only on the ground, and cyclic locomotion might not be a solution.

Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences, biomechanics, and virtual character animation.
Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}. These approaches are by definition not able to discover complex behaviors in unforeseen contexts.

The issue of planning acyclic contacts was first completely described by Bretl~\cite{Bretl:2006:MPM:1124573.1124585}. The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path. A third nontrivial problem, $\mathcal{P}_3$, 
%~ not addressed in this work, 
then consists in interpolating a complete motion between two postures of the contact sequence.  A key issue is to avoid combinatorial explosion when considering at the same time the possible contacts and the potential paths. Bretl's seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments. Following it, seve\-ral papers have applied this approach in specific situations, limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorics. Two main directions have been explored. \textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial aspect of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}. A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}. To get reasonable computation times, the method uses a simplified dynamic model for the avatar. Still, the method is far from real-time  (about 1 minute of computation for 20 contacts).  A similar approach has been considered for manipulation by \cite{gabicciniisrr15}. Deits and Tedrake proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered, and equilibrium is not considered~\cite{DBLP:conf/humanoids/DeitsT14}. 
Dai et al.~\cite{dai2014whole} extended the work of Posa et al.~\cite{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
the contacts manually for complex interactions.

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling has been shown by Escande et al. \cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e. an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, 
but at the cost of prohibitive computation times (up to several hours) for constraining scenarios. This approach is suboptimal because the quality of the motion depends on the quality of the guide path. Bouyarmane et al. ~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium. Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented). Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path. Chung and Khatib~\cite{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment. This planning phase does not account for collisions, implying that re-planning is required in case of failure. In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}. A limitation with
these approaches (including our method) is that the existing planners only address a subset of the problem, because their ability to find a solution depends
on the existence of quasi-static equilibrium configurations along a feasible path, which is too restrictive in the general case. Other contributions to legged locomotion, not directly related to multi-contact motion, are worth mentioning, as they rely on a similar decomposition of the problem. First a discrete sequence 
of contact sets is planned (at the so-called footstep planning phase), using a low-dimensional abstraction of the robot to account for its kinematic constraints~\cite{6078435, doi:10.1177/0278364910392608}. In \cite{doi:10.1177/0278364910392608}, a ``pose certificate'' is obtained by generating a whole-body configuration for each set through inverse kinematics, as done in $\mathcal{P}_2$. Then a
motion is generated along the sequence through the use of optimization techniques. 
The solutions proposed are designed for cyclic locomotion in quasi-flat scenarios, where the support polygon is a relevant method for equilibrium checks. They thus cannot be generalized to multi contact locomotion. 
%For instance, the choice of the next effector in contact is often deterministic, the contacts are either coplanar or punctual, or the generation of the end-effector trajectory only works for ``ground'' obstacles, as it assumes that increasing the $z$ position of an effector drives it further away from obstacles~\cite{Kolter:2009:TTV:1703775.1703833}. 
However, some contributions on specific parts of the problem could be applied directly in our case. 
For instance, learning ``terrain costs'', based on expert knowledge as proposed in ~\cite{Ratliff:2009:LSF:1569248.1569253}, could define good heuristics to compute the next contact location for an effector. Although we did not try to include such formulation in this paper, it would be straightforward to integrate such heuristic in our planner.

%~ For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
%~ used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
%~ at \gls{interactive} frame rates (a method is interactive when the computation time for planning is less than the
%~ execution time). While this new approach provides an exciting direction of research, currently the resulting motions
 %~ do not seem applicable to real robots.

As far as robotics applications are concerned, none of the existing multi-contact planners is \gls{interactive}\footnote{We define an interactive planner as one for which the time to plan a motion is in the same order of magnitude as the time to execute it. For instance, computing one contact change should take about one or two seconds.}.
However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Carpentier2016,Hauser2014, herzog2015trajectory, Park116}. In particular, our algorithm proposed in \cite{Carpentier2016} is \gls{interactive}.
Therefore, a planner capable of efficiently solving \Pa and \Pb could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
The main contribution of this paper is exactly this planner.

\subsection{Contributions}

Our solution belongs to the class of decoupled approaches, 
i.e. 
we
propose
specific
algorithms
to
efficiently
solve
both
\Pa
and \Pb
while 
relying 
on 
state-of-the-art 
solution 
to 
$\mathcal{P}_3$
to 
obtain
the
whole
movement.
Our
main
contribution
is
the
definition        
of
a reduction property, the reachability condition.

%~ because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. 
Compared to previous approaches, our solution has two main novelties:

\noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm. The key to its efficiency is that it does not sample directly the contact manifold, but an approximation of the \textit{contact reachable} space. The \textit{contact reachable} space is a low-dimensional space for which there exists a mapping to the contact manifold.

\noindent \textbf{Regarding $\mathcal{P}_2$},  we propose a fast method to extend a \textit{contact reachable} path into a sequence of whole-body configurations in static equilibrium. This  requires the explicit computation of contact configurations. It is guided by dedicated heuristics that quickly synthesize feasible configurations.

The reachability condition is the key to the strict separation between $\mathcal{P}_1$ and $\mathcal{P}_2$, hence to the low complexity of our planner. However the reduction
can result in failures. We demonstrate empirically its interest, through an extensive experimental validation with real robot models on a dynamic simulator. The high success rate and low computation times we obtain allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.

To further demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot. For HRP-2, we detail the complete computation times to address sequentially the three problems, and compare them to related work, demonstrating that our method is orders of magnitude faster.

Finally, we provide an extensive discussion on the consequences of our approach in terms of efficiency and completeness regarding the contact planning problem. \\

\noindent \textbf{Comparison with our previous work:}
The present paper is an extension of our ISRR conference paper~\cite{tonneauisrr15}. 
As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section~\ref{rbprm} and~\ref{sec:contact}). 
However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section~\ref{app:contact}), a novel criterion for
static equilibrium (Section~\ref{sec:Equil}), and the release of the source code of the planner (Section~\ref{app:hpp}).

The other novelty of the paper is a rigorous experimental validation of the approach on actual robot models (Section~\ref{sec:results}).
To validate our contact plans we introduced a complete framework for multi-contact motion synthesis. This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work~\cite{Carpentier2016}. Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}. 
These aspects of the framework are presented in details in the
paper, but are not novel \textit{per se}. %Describing them here is required to understand our experimental contribution.
The novelty lies in the complete
validation of the contact planner with real robot models, for
which the planning is much harder with respect to the avatars
used in~\cite{tonneauisrr15}, because of more restrictive kinematic constraints.

%~ Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
%~ thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.
%~ These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}. The novelty lies in the validation of the method with real robot models, for which the planning is harder with respect to the avatars used in [20], because of stricter kinematic constraints

%~ As such, the motion planning algorithm presented here is the same as the one presented in the
%~ previous work. Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper, with the exception of Section~\ref{app:contact}, where we additionally provide the pseudo-code of the algorithm absent from the original paper. We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section~\ref{sec:Equil}.
%~ Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
%~ all the technical information required to use the planner.
%~ 
%~ To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team. We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work~\cite{Carpentier2016}. The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework. Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
%~ We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand the experimental contribution of this paper.

%~ Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
%~ Indeed, the novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
%~ As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
%~ thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method is remains efficient and qualitatively relevant in this context.
% !TEX root =  ../main.tex
\section{Overview}
\label{overview}

Figure~\ref{fig:framework} illustrates our work flow.
\Pa and \Pb are addressed sequentially: From a given problem (a) we first plan a root guide path (b), before
extending it into a sequence of static equilibrium configurations (c). In the case where step (c) fails,
our framework invalidates the computed guide path, and restarts the planning from (b).

%

\subsection{Computation of a guide path --- \Pa (Section~\ref{rbprm})}
We first consider the problem of planning a root guide path (Figure~\ref{fig:framework}--$\mathcal{P}_1$).  The dimension of the path is equal to the number of degrees of freedom (DoFs) of the root of the robot.
Similarly to previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: there must exist a joint configuration that results in static equilibrium for each root configuration\footnote{Enforcing static equilibrium is a classical, conservative approach to reduce the search space and considering states  with non-zero accelerations and velocities, which can't be connected trivially. This does not mean that the final motion will necessary be quasi-static.}. Previous works verify \textit{equilibrium feasibility} by explicitly computing such a configuration. We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact reachability}, illustrated in the following.

An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' to the environment: close, because a contact surface must be partially included in the reachable workspace of the robot to allow contact creation; not too close, because the robot must avoid collision.
%~ \deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
More precisely, a root configuration is \textit{contact reachable} if the root scaled by a user-defined factor $s \geq 1$ is not in collision (Figure~\ref{fig:reach_cond} - red shape), while the reachable workspace is in collision (Figure~\ref{fig:reach_cond} - green shapes).
%~ Verifying the \textit{reachability condition} is very efficient in practice.
To plan a root path, we then use an RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).

\begin{figure}[t]
\centering
  \begin{overpic}[width=0.8\linewidth]{figures/reachability_cond}
	\end{overpic}
\caption{The reachability condition is verified by the right configuration: the trunk (red) is free of collisions while the limbs reachable workspace (green) intersect the environment.}
		   \label{fig:reach_cond}
\end{figure}

%
%~ Ideally we would like to sample root configurations that are \gls{equilibrium feasible}. A root configuration is \gls{equilibrium feasible} if and only if there exists at least one joint configuration such that:
%~ \begin{description}
%~ \item[C1]: the robot is not in collision with the environment, 
%~ \item[C2]: the robot is in static equilibrium. 
%~ \end{description}
%~ Verifying C2 is computationally expensive because it requires the projection of the whole-body configuration into a contact posture, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
%~ For this reason, we are interested in finding an approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} that is computationally efficient.
%~ A requirement for C2 is the condition:
%~ \begin{description}
%~ \item[C3]: at least one end-effector is in contact with the environment.
%~ \end{description}
%~ C1 and C3 define \glslink{contact reachable}{\textit{contact reachability}}, which is easier to verify than \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} because it does not require the computation of contact forces.
%~ We make the assumption that for the \textit{\gls{cluttered} contact planning problem}, most of the times \gls{contact reachable} root configurations are also \gls{equilibrium feasible}.
%~ For this reason we neglect C2 during the resolution of $\mathcal{P}_1$, and we consider it only when solving \Pb.
%~ However, even verifying \glslink{contact reachable}{\textit{contact reachability}} is computationally too expensive because it requires the projection of the whole-body configuration into a contact posture.
%~ 
%~ An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
%~ We approximate \glslink{contact reachable}{\textit{contact reachability}} with the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
%~ More precisely, a root configuration is \textit{reachable} if the root scaled by a factor $s \geq 1$ is not in collision, while the reachable workspace is in collision with the environment.
%~ Verifying the \textit{reachability condition} is very efficient in practice.
%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).

In the remainder of this paper, we use the terms \gls{contact reachable} and \gls{equilibrium feasible} to qualify either a root or a whole-body configuration, or a set
of such configurations.

\subsection{Generating a discrete sequence of contact configurations --- \Pb (Section~\ref{sec:contact})}
The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--$\mathcal{P}_2$).
To achieve this the root path is first decomposed into a sequence of discrete root configurations, according to a user-defined discretization step.
Each root configuration is then extended into a whole-body configuration in static equilibrium.
The algorithm thus proceeds iteratively, starting from the whole-body initial configuration of the robot.
It takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually. 
%~ Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database computed offline independently from the environment).
%~ Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.

%~ The details of the contact sequence generation are given in Section~\ref{sec:contact}.

%~ From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root configuration, an inverse-kinematics solver 
%~ is used to maintain the contacts that were previously existing. Possibly, some of these contacts cannot be maintained (because of joint limits or collisions).
%~ Then the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
%~ The algorithm is designed so that only one contact can be created or broken between two successive configurations. While it does not provide guarantees that the interpolation between two configurations is achievable, it appears empirically as a reasonable heuristic.

%~ To create a contact, we consider each limb of the robot as a manipulator arm attached to the root. We store a database of configurations for each manipulator. Figure~\ref{fig:contact_gen} -- 2 presents a few configurations for the right arm of the robot. The configurations in the database which are close to contact are considered (Figure~\ref{fig:contact_gen} -- 2 and 3). Among the candidates, 3 criteria are considered to choose the most relevant: the ideal candidate is free of collision, allows to maintain static balance if necessary, and  contributes to the motion in an optimal way. We call the measure of this contribution task efficiency: it represents the ability the limb has to exert a force compatible with the direction of motion. In Figure~\ref{fig:contact_gen} -- 4, the black arrow represents the direction of motion. Among the candidates, the most relevant to achieve a vertical motion is chosen, and projected on the contact surface using an inverse kinematics solver (Figure~\ref{fig:contact_gen} -- 5). The task efficiency is measured using the Extended FORce Transmission ratio (EFORT), proposed by ~\cite{Tonneau:2014:TEC:2619648.2619652}.

\section{Notation and definitions} \label{notations}

A vector  $\mathbf{x}$ is denoted with a bold lower-case letter.
A matrix $\mathbf{A}$ is denoted with a bold upper-case letter.
A set $C$ is denoted with an upper-case italic letter.
Scalar variables and functions are denoted with lower-case italic letters, such as
$r$ or $f(\textbf{x})$.

%\subsubsection{Robot description}
%\label{def}

%~ \begin{figure}
  %~ \centering
  %~ \begin{overpic}[width=0.8\linewidth]{figures/character}
    %~ \put (7,29) {\small{$\in \mathbf{q}^1$}}
    %~ \put (7,25) {\small{effector}}
  %~ \end{overpic}
  %~ \caption{
    %~ Left: Robot in a rest configuration. The right arm is denoted as the limb $R^1$. Each colored dot represents a degree of freedom around an axis. Right: Volumes of the robot. The red geometry denotes $W^0$ and must remain collision-free. The green spheres are the reachable workspace of each limb, the  $W^k$.}
  %~ \label{fig:character}
%~ \end{figure}

\medskip
\textbf{A robot} is a kinematic tree $R$ composed of: a root $R^0$, and $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
The root has $r \geq 6$ DoFs: for instance, HRP-2 has two extra DoFs in the torso, such that we have $r=8$.
Thus $R$ is fully described by a configuration (a vector of joint values) $\mathbf{q} \in  \mathbb{R}^{r+n}$, with $n$ the number of joint DoFs.
$\mathbf{q}$ is decomposed as follows:
\begin{itemize}
	\item $\mathbf{q}^k$ is a configuration  of the limb $R^k$; % (Figure~\ref{fig:character});
	\item $\mathbf{q}^{\overline{k}}$ is a vector of joint values of R \textbf{not} related to $R^k$. We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ is the world coordinates vector of $R^0$.
\end{itemize}
%Finally, $ \mathbf{q}_{start}$ and  $\mathbf{q}_{goal}$ are the start and goal configurations given as an input of our problem.

%\subsubsection{Environment and other volumes} 
%\label{sec:A}
\medskip
We then define a set of 3d volumes  $W^i, 0 \leq i \leq l$, each  attached to one joint of the root, such that 
$W^i(\mathbf{q}^{0})$ describes the world position of $W^i$ for the root configuration $\mathbf{q}^{0}$.

\noindent $W^0$ is a volume encompassing $R^0$ (Figure~\ref{fig:HyQ_roms}), or equal to it\footnote{$W^0$ is typically a low-polygonal bounding shape
of $R^0$ for performance.}.\\
 %~ (Figure~\ref{fig:character}-right: central cylinder). 
 $W^k$ is the reachable workspace of a limb $R^k$: %(Figure~\ref{fig:character}-right: the four ellipses). 
\begin{equation}
  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k \in C^k_{j.lim}, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\}
\end{equation}
%~ \adnote{in this definition you don't account for joint limits}
where $\mathbf{p}^k$ denotes the end-effector position (in the root frame) of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{j.lim}$ is the space
of admissible limb joint configurations. We also define \mbox{$W = \bigcup_{k=1}^{l}W^k$}.

%~ Each $W^i, 0 \leq i \leq l$ is  attached to one joint of the root, such that 
%~ $W^i(\mathbf{q}^{0})$ is the volume $W^i$ transformed the by root configuration $\mathbf{q}^{0}$.

\medskip
\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains. $O$ is represented
as a polygon soup (or mesh), where the normal of each surface is known. No further requirement is needed
by our approach. In this work, we assume the environment is fully known. State uncertainty is out of the scope of the paper.

\medskip
Finally we define some relevant subsets of the configuration space $C$.
$C_{Contact}$ is the set of whole body-configurations in contact and collision-free.
$C_{Contact}^k \subset C_{Contact}$ is the set of whole body-configurations where at least $R^k$ is in contact.

\medskip
$C_{Equil} \subset C_{Contact}$ is the set of whole body-configurations in static equilibrium and collision-free.

\medskip
For any set $C_{X}$, we define $C_{X}^0$:
\begin{equation*}
  C_{X}^0 = \left\{ \mathbf{q}^{0},  \exists \mathbf{q}^{\overline{0}}: \mathbf{q}^0  \oplus \mathbf{q}^{\overline{0}} \in C_{X} \right\}
\end{equation*}

%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.
\begin{figure}
  \centering
  \includegraphics[width=1\linewidth]{figures/HyQ_roms}
  \caption{
           Reachable workspace and torso bounding box of HyQ. Each green shape represent a reachable workspace $W^k$ of a limb. The red shape is $W^0$.}
		   \label{fig:HyQ_roms}
\end{figure}
% !TEX root =  ../main.tex
%~ \section{Computing a guide path in $C^0_{reach}$ ($\mathcal{P}_1$) }
\section{Root path planning in the contact reachable space}
\label{rbprm}

During the root path planning we only consider the root configuration $\mathbf{q}^0$ defined in the previous Section,
as well as the environment $O$.

Given start and goal configurations, we aim at computing a guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ $\mathbb{R}^r$ verifying:
\begin{equation*} \label{eq:path}
\forall t \in [0,1], \mathbf{q}^0(t)  \in C_{Equil}^0
\end{equation*}
This means that any root configuration must be extended into a whole-body, static equilibrium configuration.
$C_{Equil}^0$  cannot be described analytically.

The main hypothesis of this work is that for a large variety of locomotion tasks, we can define a space  $C_{Reach}^0 \simeq C_{Contact}^0$, such that 
\begin{equation} \label{eq:creach}
\forall t \in [0,1], \mathbf{q}^0(t) \in C_{Reach}^0 \Rightarrow \mathbf{q}^0(t)  \in C_{Equil}^0
\end{equation}
% for \gls{cluttered} problems.
We call  $C_{Reach}^0$ the \textit{contact reachable workspace}, and detail its construction in the following.
The validity of this hypothesis is discussed in depth in Section~\ref{sec:discussion}.
%~ We describe the construction of $C_{Reach}^0$ in the following.

\subsection{Conditions for contact reachability}
The contact reachable workspace is defined as a compromise between two necessary and a sufficient condition for contact creation.

\textbf{necessary conditions:}
For a contact to be possible, an obstacle $O_i \subset O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot. % (Figure~\ref{fig:contact_gen}--a).
 Also the torso of the robot $W^0(\mathbf{q}^{0})$ must necessarily be collision-free.
Therefore we can define an outer approximation  $ \mathcal{C}^0_{\textrm{\it Nec}} \supset$ $C_{Contact}^0$ as: 
\begin{equation}
\mathcal{C}^0_{\textrm{\it Nec}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
 %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
\end{equation}
%~ The inclusion $C_{Contact}^0$ $\subset \mathcal{C}^0_{\textrm{\it Nec}}$ is straightforward, but it is very important.
%~ Planning in $\mathcal{C}^0_{\textrm{\it Nec}}$ allows a strong reduction of the search space, while not discarding 
%~ any root configuration that could lead to a solution.

%~ The condition defining $\mathcal{C}^0_{\textrm{\it Nec}}$ is only necessary. This means that it might not be possible to extend a guide path planned 
%~ in $\mathcal{C}^0_{\textrm{\it Nec}}$ into a sequence of contact configurations.

\textbf{sufficient condition:}
Similarly we can define an inner approximation \mbox{$\mathcal{C}^0_{\textrm{\it Suf}} \subset $ $C_{Contact}^0$} by considering a bounding volume $B^{\textrm{\it Suf}}$ encompassing the whole robot in a given pose, except for the effector surfaces. 
\begin{equation}
\mathcal{C}^0_{\textrm{\it Suf}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ B^{\textrm{\it Suf}}(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
 %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
\end{equation}

%~ In general, the inclusion is strict, which means that we lose the completeness of the two-stage contact planner (i.e. the planner is not able to discover a path inside \mbox{$C_{reach} \setminus \mathcal{C}^0_{\textrm{\it Suf}}$}). However, the sufficient condition guarantees that any such path leads to a valid sequence of contacts.

\subsection{The compromising reachability condition}
\label{sec:scaling}
The ideal shape $B^*, W^0 \subset B^* \subset B^{\textrm{\it Suf}}$ would define a necessary \textbf{and} sufficient condition for contact creation. 
It would guarantee that any root configuration $\mathbf{q}^{0} \in B^*$ would result in a contact configuration, while any $\mathbf{q}^{0} \notin B^*$ could not.
To our knowledge  $B^*$ has no explicit definition.
Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.

%~ However, using a shape between $W^0$ and $B^{\textrm{\it Suf}}$ leads to a trade-off between a necessary and a sufficient condition. 
We define $W^0_s$ as the volume $W^0$ subject to a scaling transformation by a factor $s \in \mathbb{R}^+$.
%
We then consider the spaces $C_{s}^0$
 \begin{equation}
 \label{eq:reachc}
C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
 %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
\end{equation}
%
The parametrization of $s$ defines a trade-off:
If $s=1$, then $W^0_s$ = $W^0$, such that $C_1^0$ = $\mathcal{C}^0_{\textrm{\it Nec}}$.
 %~ We thus consider that $s \geq 1$, since smaller values would only worsen the approximation.
By increasing $s$, the condition can become sufficient, but less and less necessary.  
Eq.~\ref{eq:reachc} thus defines the \textit{reachability condition}. We fix a value $s^*$ for $s$ and define  $C_{Reach}^0 = C^0_{s^*}$.
The computation of $s^*$ is detailed in Section~\ref{sec:params}. 
In Appendix~\ref{app:rom}, we give a generic method to compute the $W$ volumes appearing in the definition of $C_{Reach}^0$.

\subsection{Computing the guide path in $C_{reach}^0$}
$C_{Reach}^0$ can be sampled efficiently thanks to Eq.~\ref{eq:reachc}, and can thus be used with any standard motion planner.
%~ The only significant change is to replace the collision checking with the \textit{reachability condition}.
Our current implementation uses the Bi-RRT planner \citep{770022} provided by the HPP software~\citep{7759083}.
Our implementation is exactly the same as the pseudo-code of the original planner (which does not detail the configuration validation method). With respect to a ``classic'' implementation, the only difference is that instead of validating a configuration using collision detection, we validate it with the \textit{reachability condition}.\\

%~ \delst{However, to improve the sampling efficiency 
%~ we bias the sampling process to generate near-obstacle configurations, similarly to cite{Amato98choosinggood}.}

This Section has presented a guide path planner for the geometric root of a robot, implemented as a low-dimensional sampling-based 
algorithm. Given start and goal configurations, it outputs a continuous path for the robot's root. 
%~ \deladp{Thanks to the \textit{reachability condition}, we assume that for any configuration in the path there exists a joint configuration that results in static equilibrium.}
%~ Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
%~ By doing this we can solve the problem with a sampling-based approach, in \gls{interactive} computation times.
% !TEX root =  ../main.tex
\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
\label{sec:contact}
In the second phase, we compute a discrete sequence of static equilibrium configurations $\mathbf{Q}^{\overline{0}}$ given a root path
$\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Reach}^0$}. This contact planner uses a contact generator, used to generate static equilibrium configurations. We first describe the contact planning algorithm, before describing
the contact generator.
%~ Our planner computes guide paths in $C_{Reach}^0$ , an approximation of \gls{$C_{Equil}^0$}.
%~ As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Equil}^0$}.
%~ If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section~\ref{sec:perf}.
%~ We now consider the second problem of computing a discrete sequence of equilibrium configurations $\mathbf{Q}^{\overline{0}}$ along $\mathbf{q}^0(t)$.

%~ In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a limb, given a
%~ root location.
%~ Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.

%~ Our criterion to assert efficiently the static equilibrium of the system
%~ is described in Appendix~\ref{sec:heuristics}.

\subsection{Definition of a contact sequence}
In previous contributions~\citep{DBLP:conf/iser/EscandeKMG08}, a contact plan is defined as a sequence of quasi-static equilibrium configurations
for each contact phase. For instance, a walk cycle would be described by three key configurations: a double-support configuration, a single-support configuration (a contact is broken), and another double-support configuration (a contact is created). 
Our definition of contact plan differs: between two consecutive configurations we allow both a contact break and a contact creation---if they are on the same effector. 
In the previous example, our contact plan would simply consist of the two double-support configurations. 
%However, if the contact broken and the contact created sequentially are not on the same effector, our planner outputs two distinct states.
This representation is sufficient to describe all the contact phases, because the single support phase is implicitly described. Furthermore it removes the need to compute a single-support quasi-static configuration as in the example. Indeed, there might be a case where no quasi-static solution exists for the single support phase (because of the environment), but
there exists a dynamic motion connecting the two double support states. Such motion will be computed by our framework, because the quasi-static constraint is only required at the contact planning phase; as shown in the companion video, and explained in Appendix~\ref{app:optim}, our framework is able to produce dynamic motions.

\subsection{Contact planning algorithm}
Starting from an initial whole-body configuration, we compute a sequence
of whole-body configurations  $\mathbf{Q}^{\overline{0}}$ along the root path $\mathbf{q}^0(t)$.
We first give an intuition of the algorithm, before providing its complete pseudo-code.
%~ The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
\subsubsection{Algorithm overview}
First, the root path $\mathbf{q}^0(t)$ is discretized into a sequence of $j$ key configurations:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*}

where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
$j$ depends on a user-defined variable, called the discretization step. It corresponds to the ratio between the length of the path $\mathbf{q}^0(t)$\footnote{The length of the path is computed as the weighted 6D Euclidean distance
traveled along the it, with a weight of $0.7$ for the translation part, and $0.3$ for the orientation part.}, and the number
of configurations selected along it to create the contact configurations. 
Each root configuration of $\mathbf{Q}^0$ is then extended into a whole-body configuration such that:
\begin{itemize} 
\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
\item At most one contact is added between two consecutive configurations.
\item Each configuration is in static equilibrium.
\item Each configuration is collision-free.
\end{itemize}

%~ We want to extend the configurations of $\mathbf{Q}^0$ in such a way that continuity is preserved regarding the contact transitions.
%~ To do so, we define an algorithm that, given the current root configuration, and the previous full body configuration, computes a full body configuration 
%~ in $C_{Equil}$ such that contacts are maintained if possible.
%~ The first full body configuration of the sequence is given by the initial state of the robot.

 %~ we propose a recursive mapping $\pi$, for any $0<i<j$:
%~ \begin{equation*}
    %~ \pi\colon\left\{
    %~ \begin{aligned}		
        %~ \mathbf{Q}^0 \in C_{Equil}^0 & \longrightarrow C_{Equil} \\
        %~ %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
        %~ \mathbf{q}^{0}_i &  \longrightarrow  g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
    %~ \end{aligned}
    %~ \right.
%~ \end{equation*} 
%~ $g$ is the method that extends a root configuration into a full-body configuration. At each step, it tries to generate a contact configuration that preserves as much as possible
%~ the previous contacts while allowing for static equilibrium. The objective is thus to characterize $g$.
%~ We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.

%~ The function $g$ is defined independently by $g^k$ for each limb $R^k$. In defining $g^k$, two aspects must be considered. Is the limb $R^k$ in contact? And which criteria is it optimizing?

\paragraph{Maintaining a contact in the sequence}

%~ Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
If kinematically possible, a limb in contact at step $i-1$ remains in contact at step $i$ (Figure~\ref{fig:break_contact}). 
%~ The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits. 
%~ The solver is directly provided by the HPP software.
Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
If two or more contacts can't be maintained between two consecutive configurations, one or more intermediate configurations are added, to ensure
that at most one contact is broken between two sequential configurations.
%~ For these steps the root configuration is the same as for the previous step, with the difference that
%~ one faulty contact is repositioned, in the hope that it will not be broken at the next step.

\begin{figure}[t]
\centering
  \begin{overpic}[width=1\linewidth]{figures/break_contact}
		\put (0,4) {1} 
		\put (25,4) {2} 
		\put (50,4) {3} 
		\put (76,4) {4} 
		%~ \put (68,58) {3.a)} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	\end{overpic}
\caption{Contacts are maintained if joint limits and collisions constraints are respected (2). They are broken otherwise(3,4). The green line represents the root path. The blurred character
represents the previous contact configuration.}
		   \label{fig:break_contact}
\end{figure}

%~ \begin{figure}[t]
%~ \centering
  %~ \begin{overpic}[width=0.6\linewidth]{figures/generate_contact}
		%~ \put (5,58) {1)} 
		%~ \put (37,58) {2)} 
		%~ \put (68,58) {3.a)} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	%~ \end{overpic}
%~ \caption{Contacts are generated when the configuration is not balanced.}
		   %~ \label{fig:generate_contact}
%~ \end{figure}

%\subsection{Generation of a contact configuration}  
\paragraph{Creating contacts}
Contacts are created using a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest. If the contact creation does not succeeds, the limb is pushed on top of the queue, and will only be tried again after the others. \\ \\
%~ \begin{enumerate}
%~ \item \deladp{At most one contact creation happens between two consecutive steps. }
%~ \item \deladp{A contact is validated if and only if the resulting configuration is in static equilibrium;}
%~ \item \deladp{We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
%~ was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
%~ \end{enumerate}

%~ \deladp{However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.}

% !TEX root =  ../main_tro.tex
\subsubsection{Pseudo-code of the Algorithm}
\label{app:contact}

%~ First, we define an abstract structure State,
%~ that describes a contact configuration.
%~ The use of queues allows a FIFO approach regarding the order 
%~ in which contacts are tested: we try to replace older contacts first when necessary.
%~ Thus the algorithm is deterministic even though it can handle acyclic motions. \\
%~ 
%~ \begin{lstlisting}]
%~ Struct Limb
%~ {
    %~ // Limb Configuration
    %~ Configuration qk;
    %~ // Effector position in
    %~ // world coordinates
    %~ vector6 pk;
%~ };
%~ 
%~ Struct State
%~ {
    %~ // root location
    %~ Configuration q0;
    %~ // List of limbs not in contact
    %~ queue<Limb> freeLimbs;
    %~ // List of limbs in contact
    %~ queue<Limb> contactLimbs;
%~ };
%~ \end{lstlisting}

From the start configuration, given as an input by the user,
we create the initial state $s0$. A state is described by a whole-body configuration, as well as the list of currently active contacts and their associated 6D positions.
Algorithm~\ref{alg:interpolate}  is then called with $s0$, as well as the discretized path 
$\mathbf{Q}^0$, as input parameters.

\begin{algorithm}[!tbp]
\caption{Discretization of a path} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{Interpolate}{$s0$,$\mathbf{Q}^0$, $MAX\_TRIES$}
		\State $states \gets (s0)$ \Comment{List of states initialized with $s0$}
		\State $nb\_fail \gets 0$ 
		\State $i \gets 1;$ \Comment{Current index in the list}
		\While {$i < Length(\mathbf{Q}^0)$}
			\State $pState \gets LastElement(states)$
			\State $s \gets$ \textsc{GenFullBody}$(pState, Element(\mathbf{Q}^0,i))$
			\If {$s \not= 0 $}
				\State $nb\_fail \gets 0$
				\State $i \gets i+1$
				%~ \State \textbf{return} $\mathbf{q}^{0}$
			\Else
				\State $nb\_fail \gets nb\_fail + 1$
				%~ \If {$nb\_fail == MAX\_TRIES$}
					%~ \State \textbf{return} $FAILURE$
				%~ \EndIf				
				\State $s \gets $\textsc{IntermediateContactState}$(pState)$
				\If {$s == 0 \lor nb\_fail == MAX\_TRIES$}
					\State \textbf{return} $FAILURE$
				\EndIf		
			\EndIf
			\State ${PushBack}(states, s)$
		\EndWhile
		\State \textbf{return} $states$
	\EndFunction
\end{algorithmic}
\label{alg:interpolate}
\end{algorithm}

At each step of Algorithm~1, \textsc{GenFullBody} (Algorithm~\ref{alg:pi}) is called with the previous state as a parameter, as well
as a new root configuration. \textsc{GenFullBody} returns a new contact configuration, if it succeeded
in computing a configuration with only one contact switch occurring~;
otherwise, the method \textsc{IntermediateContactState} (Algorithm~\ref{alg:repo}) is called.
It repositions an end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
This repositioning allows to increase the odds that the contact can be maintained at the next step.
The method \textit{Length} returns the length of a list~; \textit{LastElement} returns the last element
of a list~; \textit{Element} returns the element contained by a list at a given index~; and \textit{PushBack} inserts an element at the end of a list.
%~ Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.

\begin{algorithm}[!tbp]
\caption{Full body contact generation method} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
		\State $newState \gets {CreateState}(\mathbf{q}^0,$ 
        \State $\quad \quad \quad \quad  \quad  \quad  \quad  \quad  \quad \quad  \quad \quad\!\! ContactLimbs(pState))$
		%~ \State /*First try to maintain previous contacts*/
		\State $nConBroken \gets 0$
		\For {\textbf{each} $k$ in $ContactLimbs(pState)$}
			\If {$\neg$\textit{MaintainContact}$(pState,\mathbf{q}^0,k)$}
				\State $nConBroken \gets nConBroken +1$
				\If {$nConBroken > 1$}				
					\State \textbf{return} $0$
				\EndIf				
				\State \textit{MarkFree}$(newState,k)$
			\Else 					
				\State \textit{MarkContact}$(newState,k)$
			\EndIf
		\EndFor
		\For {\textbf{each} $k$ in $FreeLimbs(newState)$}
			\If {\textit{GenerateContact}$(newState,k)$}	
				\State \textit{MarkContact}$(newState,k)$
				\State \textbf{return} $newState$
			\EndIf
		\EndFor
		\If {\textit{IsInStaticEquilibrium}$(newState)$}
			\State \textbf{return} $newState$
		\Else
			\State \textbf{return} $0$
		\EndIf
	\EndFunction
\end{algorithmic}
\label{alg:pi}
\end{algorithm}

In Alg.~\ref{alg:pi}, the procedure \textit{MaintainContact}$(pState,\mathbf{q}^0,k)$ performs inverse kinematics to reach the previous contact position for the limb.
If it succeeds, the new limb configuration is assigned to $k$. If it fails, a random collision free configuration is assigned to $k$.
The method \textit{IsInStaticEquilibrium} returns whether a given state is in static equilibrium. \textit{CreateState} creates a new state, given
a root configuration and the list of active contacts.
$ContactLimbs$ (respectively $FreeLimbs$) returns the list of limbs that are in contact (respectively \textit{not} in contact).  \textit{MarkContact} (respectively \textit{MarkFree})  marks a limb as in contact (respectively \textit{not} in contact). These methods follow a FIFO approach: the first limb chronologically marked as in contact (respectively not in contact) is returned first by $ContactLimbs$ (respectively $FreeLimbs$). This allows the algorithm to be deterministic even though it can handle acyclic motions.
%~ $add\_contact\_limb(newState,k)$ marks a limb as in contact at a given state.
%
%~ The pseudo code for the method \textsc{IntermediateContactState} is given by Algorithm~\ref{alg:repo}.
%
\textit{GenerateContact}$(state,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
It generates a contact configuration in static equilibrium, and assigns the corresponding configuration of the limb $k$ to the state configuration.
If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.

Algorithm~\ref{alg:repo} generates an intermediate state by first trying to create a contact with one of the free limbs (trying in a FIFO order) then by repositioning one of the limbs in contact.

\begin{algorithm}[!tbp]
\caption{Adds or repositions a contact for one limb} \label{interpolate}
	\begin{algorithmic}[1]
	\Function{IntermediateContactState}{$state$}        
        \State $newState \gets state$
		\For {\textbf{each} $k$ in $FreeLimbs(newState)$}
			\If {\textit{GenerateContact}$(newState,k)$}	
				\State \textit{MarkContact}$(newState,k)$			
				\State \textbf{return} $newState$
			%~ \Else
				%~ \State \textit{MarkFree}$(newState,k)$	\comment{Mark }
				%~ \State $push(states.freeLimbs,k)$		
			\EndIf
		\EndFor
		\For {\textbf{each} $k$ in $ContactLimbs(newState)$}
			\If {\textit{GenerateContact}$(newState,k)$}	
				\State /*Account for repositioning in FIFO queue*/		
				\State \textit{MarkContact}$(newState,k)$
				\State \textbf{return} $newState$
			%~ \Else
				%~ \State \textit{MarkFree}$(newState,k)$	\comment{Mark }
				%~ \State $push(states.freeLimbs,k)$		
			\EndIf
		\EndFor
        \State
		/*Fails if impossible to relocate any effector*/
		\State \textbf{return} $0$
        %~ 
        %~ 
		%~ \State $i \gets 0$
		%~ \While {$i<length(states.freeLimbs)$}
			%~ \State Limb $k = pop(states.freeLimbs)$
			%~ \If {\textit{GenerateContact}$(state,k)$}	
				%~ \State $push(newState.contactLimbs,k)$			
				%~ \State \textbf{return}
			%~ \Else
				%~ \State $i+=1$
				%~ \State $push(states.freeLimbs,k)$		
			%~ \EndIf
		%~ \EndWhile
		%~ \State $i=0$
		%~ \While {$i<length(states.contactLimbs)$}
			%~ \State Limb $k = pop(states.contactLimbs)$
			%~ \State Limb $copy = k$
			%~ \State $i+=1$
			%~ \If {\textit{GenerateContact}$(state.q0,k)$}	
				%~ \State $push(newState.contactLimbs,k)$			
				%~ \State \textbf{return}
			%~ \Else
				%~ \State $push(newState.contactLimbs,copy)$	
			%~ \EndIf
		%~ \EndWhile
		%~ /*Fails if impossible to relocate any effector*/
		%~ \State \textbf{return} $FAILURE$
	\EndFunction
\end{algorithmic}
\label{alg:repo}
\end{algorithm}

%~ \section{Derivation of the manipulability ellipsoid}
%~ \label{app:manipulability}
%~ 
%~ Again, we assume that $\mathbf{J}$ is full rank. We discard the $k$ indices, and write the pseudo-inverse of $\mathbf{J}$ as $\mathbf{J}^{\dagger}$.
%~ \begin{eqnarray*}
%~ \mathbf{\dot{p}} & =  & \mathbf{J} \dot{\mathbf{q}} \\ 
%~ \mathbf{\dot{q}} & =  & \mathbf{J}^{\dagger} \dot{\mathbf{p}} \\ 
%~ \mathbf{\dot{q}}^T & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \\ 
%~ \mathbf{\dot{q}}^T\mathbf{\dot{q}} & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} \dot{\mathbf{p}}\\ 
%~ \end{eqnarray*}
%~ 
%~ Then, the equality $\mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} = (\mathbf{J}\mathbf{J}^T)^{-1}$ follows from the SVD decomposition of each term
%~ ~\citep{ben2003generalized}.

\subsection{Contact generator}
\label{sec:single_contact}

\begin{figure*}
  \centering
  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
		\put (1,1) {a} 
		\put (22,1) {b} 
		\put (42,1) {c} 
		\put (62,1) {d} 
		\put (83,1) {e} 
	\end{overpic}
  \caption{Generation of a contact configuration for the right leg of HRP-2. (a): Selection of reachable obstacles. (b): Entries of the limb samples database (with $N = 4$). (c): With a proximity query between the octree database and the obstacles, configurations too far from obstacles are discarded. (d): The best candidate according to a user-defined heuristic $h$ is chosen. (e): The final contact is achieved using inverse kinematics.}
  \label{fig:contact_gen}
\end{figure*}

Given a configuration of the root and the list of effectors that should be in contact, the contact generator computes the configuration of the limbs such that contacts are properly satisfied and the robot is in static equilibrium:

\begin{equation}
\label{eq:contact_gen}
	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \textrm{ \textbf{and}}\ \mathbf{q}^k \in  C_{Contact}^k 
\end{equation}

In previous works~\cite{DBLP:conf/iser/EscandeKMG08,Bouyarmane2009}, the generation of contact is typically implemented by randomly sampling configurations and projecting the whole robot configuration onto the closest surfaces with an inverse kinematics solver.
In case of failure of the projection, the process would randomly iterate.

We propose two modifications of this general algorithm principle.
First our contact generator handles each limb $R^k$ independently.
By handling each limb separately, we reduce the complexity of the generation of contact configurations.
This is made possible thanks to the reachability condition in $\mathcal{P}_1$ that produces a root path that we can afford not to modify in $\mathcal{P}_2$, and because we allow both a contact break and a contact creation between two consecutive configurations of the contact sequence.
Second, we rely on off-line generation of configuration candidates.
%~ Contact generation is typically addressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
%~ This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.
%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.

%~ Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
%~ $R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
%~ While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
%~ \begin{enumerate}
%~ \item Generate randomly a collision-free limb configuration;
%~ \item Project the end-effector onto the closest surface with inverse kinematics;
%~ \item If a valid solution is found, stop. Otherwise repeat from step 1.
%~ \end{enumerate}

%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.

We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum 3D distance 
between an effector and an obstacle is less than $\epsilon \in \mathbb{R}$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$ (We choose $N=10^4$);
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
\item At runtime, when contact creation is required, intersect the octree and the environment to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c)). This operation is achieved natively by the fcl library \cite{6225337};
\item Use a user-defined heuristic $h$ to sort $S$;
\item If $S$ is empty, stop (failure). Else select the first configuration of $S$. Project it onto contact using inverse kinematics. (Figure~\ref{fig:contact_gen} (d) and (e));
\item If Eq.~\ref{eq:contact_gen} is verified, stop (success). Otherwise remove the element from $S$ and go to step 5.
\end{enumerate}

%~ The reader is referred to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
%~ of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
Because the distance $\epsilon$ does not account for the variation in orientation, several samples of $ C_{Contact}^{\epsilon}$  may turn out to be unfeasible at the time of projection. One could consider additionally filtering $ C_{Contact}^{\epsilon}$ based on the orientation with respect to the obstacle normal, but in our experience we did not notice
any significant improvement in the computational performances of the planner, so we do not perform this additional step.

In all our experiments, the heuristic $h$ is implemented as a variation of a manipulability-based heuristic~\cite{Yoshikawa1984}. The manipulability is a real number that quantifies how 
``good'' a configuration is to perform a given task, based on the analysis of the Jacobian matrix. With such heuristics, a configuration can be chosen because it is far from singularities, and thus allows mobility in all directions. On the contrary, it can be chosen because it is particularly efficient to exert a force in a desired direction. In our experiments, the former solution is usually chosen for computing leg contacts, while the latter is used for computing hand contacts. We recall the manipulability measure and its derivatives in Appendix~\ref{sec:heuristics}.

Finally, to verify that a configuration is in static equilibrium, we use a new robust LP formulation. It replaces the computationally inefficient double description
approach used in our previous work~\cite{tonneauisrr15}, and presented in the following Section~\ref{sec:Equil}.

%~ TODO: $h$ in appendix, the static equilibrium is here ~\ref{sec:Equil}

% !TEX root =  ../main_tro.tex

\section{A criterion for robust static equilibrium}
\label{sec:Equil}
%~ The planner is designed so that any generated contact configuration is in static equilibrium.
%~ We are interested in a robust 
%~ criterion, that ensures that the robot remains in equilibrium in a real-world application, regardless of perception and control uncertainties.

We first give a linear program (LP) that verifies whether a contact configuration allows for static equilibrium. This LP is the same that was proposed in  \citep{Prete2016}.
From this formulation we derive a new LP that quantifies the robustness of the equilibrium to uncertainties in the contact forces.
In turn, from this value we can either choose the most robust candidate, or set a threshold on the required robustness. 
%~ While the presented LP is original, it is based on an analysis of the problem that we proposed in, where the interested reader can find more details.

\subsubsection{Conditions for static equilibrium}
We first define the variables of the problem, for $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
\item $m \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact point $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{\boldsymbol\upgamma}_{i1},\mathbf{\boldsymbol\upgamma}_{i2}$ form a local Cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{\boldsymbol\upgamma}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}

According to Coulomb's law, the non-slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
As classically done, we linearize the friction cone in a conservative fashion with a pyramid included in it, described by four generating rays of unit length. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{\boldsymbol\upgamma}_{i1} & \mathbf{n}_{i} -\mu \mathbf{\boldsymbol\upgamma}_{i1} & \mathbf{n}_{i} + \mu \mathbf{\boldsymbol\upgamma}_{i2} & \mathbf{n}_{i} - \mu \mathbf{\boldsymbol\upgamma}_{i2}}^T
\end{equation*}

Any force belonging to the linearized cone
can thus be expressed as a positive combination of its four generating rays.

\begin{equation*}
\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} : \bm{\beta}_i \ge 0 \text{ and } \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i,
\end{equation*}
where $\bm{\beta}_i$ contains the coefficients of the cone generators.
We can then stack all the constraints to obtain:
\begin{equation}\label{eq:gen}
\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \text{ and } \mathbf{f} = \mathbf{V} \bm{\beta},
\end{equation}
where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = (\mathbf{f}_0,...,\mathbf{f}_e)$.

From the Newton-Euler equations, to be in static equilibrium the contact forces have to compensate the gravitational forces:

\begin{align} \label{eq:new_eul}
\underbrace{
\mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
\hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
}_\mathbf{G} \bm{\beta}, = 
\underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
\underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
\end{align}
where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.

If there exists a $\bm{\beta}^*$ satisfying \eqref{eq:gen} and \eqref{eq:new_eul}, it means that the configuration is in static equilibrium.
The problem can then be formulated as an LP:

\begin{equation} \label{eq:lin_prog} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e} \\
%~ \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D} \mathbf{c} + \mathbf{d} \\
& \bm{\beta} \ge 0 \\
\end{aligned} \end{equation}

\subsubsection{Formulation of a robust LP}
Let $b_0 \in \mathbb{R}$ be a scalar value. We now define the following LP:

\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e}, b_0 \in \Rv{} \\
\minimize  \quad & -b_0 \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D} \mathbf{c} + \mathbf{d} \\
%~ \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge b_0 \bm{1}\\
\end{aligned} \end{equation}

We observe that if $b_0$ is positive then \eqref{eq:lin_prog} admits a solution, and $b_0$ is proportional to the minimum distance of the contact forces to the boundaries of the friction cones.
If $b_0$ is negative, the configuration is not in static equilibrium, and $b_0$ indicates ``how far'' from equilibrium the configuration is. We thus use $b_0$ as a measure of robustness.
%~ One advantage of $b_0$ is that it gives a constant margin 
A simple approach to robustness consists in choosing a smaller friction coefficient, to constrain the forces to lie away from the boundaries of the real cone.  However, this would result in a small safety margin for forces of low magnitude, and an excessively large safety margin for large forces as the boundaries grow more and more apart. In comparison, our margin $b_0$ is constant, and provides a helpful mean to compare the robustness of different contact configurations.

In our implementation, rather than solving directly \eqref{eq:lin_prog_rob}, we solve an equivalent problem of smaller dimension that we get by taking the dual of \eqref{eq:lin_prog_rob} and eliminating the Lagrange multipliers associated to the inequality constraints:
\begin{equation} \label{eq:dual} \begin{aligned}
\find \quad & \bm{\nu} \in \Rv{6}\\
\maximize  \quad & -(\mathbf{D} \mathbf{c} + \mathbf{d})^T \bm{\nu} \\
\st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
& \mathbf{1}^T \mathbf{G}^T \bm{\nu} = 1 \\
\end{aligned} \end{equation}

The optimal value $\bm{\nu}^*$ gives the optimal value $b_0^*$ through the equality $b_0^* = (\mathbf{D} \mathbf{c} + \mathbf{d})^T \bm{\nu}^*$.
 %~ Thus the optimal value of the LP~\eqref{eq:dual} is indeed the optimal $b_0$.

\section{Source code of our planner}
\label{app:hpp}
Our planner is implemented using the Humanoid Path Planner (HPP) software, introduced in~\cite{7759083}.
HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS.
HPP implements the standard tools and algorithms used in motion planning,
such as the Bi-RRT planner from which RB-RRT is derived.

The robot models used in our experiments are described using the standard urdf file format, compatible with HPP.

Our implementation of the planner is also open source.
Both HPP and our planner can be simply downloaded and compiled by following the instructions on
\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html}. 
% !TEX root =  ../main_tro.tex
\section{Results}
\label{sec:results}
In this section we present some of the results obtained with our planner. The complete sequences are shown in the companion video.
Specifically, we demonstrate the planner for two legged robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
%~ Finally, a last example suggests possible applications to dexterous manipulation.

Our contact plans are then interpolated with a dedicated solution to the interpolation problem $\mathcal{P}_3$. This allows us to validate the obtained
motions in a dynamic simulator. This validation is an important contribution as it increases the confidence that the contact plans we compute can effectively
result in feasible motions on the real robot. One motion is demonstrated on the real HRP-2 robot.
%~ The obtained motions can be seen at at the end of the video.

At the end of this section, we discuss the role of the parameters of our framework. We then provide the \textit{interactive} computation times obtained in each case.
We also compare the times obtained with HRP-2 with respect to previous works.

\subsection{Experimental validation of the contact plans} \label{sec:valid}
%~ \adnote{I added the description of the simulations here, but I wonder whether this part should not be moved to the beginning of the Results section.}
To generate continuous movements from our contact plans we used either the framework proposed in \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
The resulting movements have been validated either on the real HRP-2 robot (details can be found in~\cite{Carpentier2016}), or with our dynamic simulator, based on a state-of-the-art algorithm~\cite{Kaufman2008}.
In the simulations we controlled the robot with a standard inverse-dynamics controller~\cite{DelPrete2015b}. The code source of the simulator is available at \url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#simulator}.
This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
The companion video shows the obtained motions.

%~ These scenarios complement the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
%~ (Figure~\ref{fig:robots_old}).
 %~ We invite the interested reader to watch the ISRR video (\url{http://youtu.be/LmLAHgGQJGA}), and 
%~ to refer to the previous paper for a discussion on these results.

%~ We say that the planning is \gls{interactive} when the computation time for one step is lesser than the
%~ time to execute it. We arbitrarily approximate this time to one second.

%~ \begin{figure}[t]
%~ \centering
  %~ \begin{overpic}[width=1\linewidth]{figures/robots_old}
	%~ \end{overpic}
%~ \caption{Virtual avatars in various scenarios demonstrated in our conference paper.}
		   %~ \label{fig:robots_old}
%~ \end{figure}

\subsection{Description of the scenarios}
In all the scenarios considered, the formulation of the problem is always the same:
a start and goal root configuration are provided as input (except in the stair climbing scenario where the start whole body configuration is given).
The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
In each scenario we detail the contacts involved and the heuristics chosen (either $h_{\textrm{\it EFORT}}$, $h_{vel}$ or $h_{w}$, all of which are defined in the Appendix~\ref{sec:heuristics}).
 %~ and the constraints on the reachable workspaces (for instance in all the scenarios, the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 
%~ The companion video presents the complete contact sequence obtained in all these scenarios.

% \subsubsection*{Truck egress (Figure~\ref{res_truck_pres} and Figure~\ref{res_truck_bd}) -- Humanoid and insectoid robots.}
%\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}

\subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing})}
From a bent configuration, the robot has to stand up using a wall as support, and climbing a 25-cm high step.

\begin{figure}
  \centering
  \includegraphics[width=1\linewidth]{figures/standing}
  \caption{
           HRP-2 in the standing scenario. }
		   \label{fig:standing}
\end{figure}

\noindent\textbf{Contacts involved:} All (both feet and hands).

\noindent\textbf{Heuristics:} $h_w$ for the feet, $h_{\textrm{\it EFORT}}$  for the hands.

%~ \noindent\textbf{Observations:} The scenario illustrates well the acyclic aspect of the planning. For instance, in the four first frames of Figure~\ref{fig:standing}, we can see that the right foot
%~ is moved twice, with the left foot in between, before the configuration allows HRP-2 to move its hand.
%~ Because the contacts are tried in a FIFO manner, the fact that the output contact sequence is acyclic shows that a cyclic approach (with a finite state machine for instance) is not sufficient
%~ for the computed path. The reason for this is not reachability, but equilibrium. The planning is slower than for the stair scenario (because the contact generation fails more),
%~ though it remains compatible with \gls{interactive} performances. % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}

\subsubsection{HRP-2 -- Car egress (Figure~\ref{fig:car})}
In this scenario inspired from the DRC car egress HRP-2 has to step out of a car.

\begin{figure}
  \centering
  \includegraphics[width=1\linewidth]{figures/polaris}
  \caption{
           Selected frames from the car egress scenario. }
		   \label{fig:car}
\end{figure}

\noindent\textbf{Contacts involved:} All (both feet and hands).

\noindent\textbf{Heuristics:} $h_w$.

%~ \noindent\textbf{Observations:} The difficulty of this scenario lies in the strong reduction of the reachable workspace induced 
%~ by the extreme proximity of all obstacles. The planner is able to find a sequence, that consists in many steps.
%~ The proximity of the obstacles invalidate a large number of contact candidates because of collisions. To avoid breaking more than one contact between each step, the motion has to be decomposed into a large number of steps (61 in average).
%~ While this scenario is the slowest to solve, the planner still computes a solution \glslink{interactive}{interactively}.

\subsubsection{HRP-2 -- Staircase with high steps (Figure~\ref{fig:stair_robust})}
\newcommand{\widthValue}{0.13\linewidth}

\begin{figure*}[h!t!]
  \begin{center}
  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
    {./video/stairclimbing1.jpeg}
  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
    {./video/stairclimbing2.jpeg}
  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
    {./video/stairclimbing3.jpeg}
  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
    {./video/stairclimbing4.jpeg}
  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
    {./video/stairclimbing5.jpeg}
  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
    {./video/stairclimbing6.jpeg}
  \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
    {./video/stairclimbing7.jpeg}
  \caption{
           HRP-2 in the stair climbing scenario. }
		   \label{fig:stair_robust}
  \end{center}
  \vspace{-0.5cm}
\end{figure*}

%~ \begin{figure}
  %~ \centering
  %~ \includegraphics[width=1\linewidth]{figures/stair}
  %~ \caption{
           %~ HRP-2 in the stair climbing scenario. }
		   %~ \label{fig:stair_robust}
%~ \end{figure}

The goal is to climb three 15-cm high steps.

\noindent\textbf{Contacts involved:} Feet and right arm.

\noindent\textbf{Heuristics:} The manipulability $h_w$ is chosen for the feet; $h_{\textrm{\it EFORT}}$ is chosen for the right arm.
%~ Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).

%~ \noindent\textbf{Observations:}
%~ This scenario illustrates best the importance of the equilibrium-robustness criterion.
%~ With a robust approach, more states are required to reach the last step (15 rather than 13 in average).
%~ However, when the last step is reached by both feet, in the nonrobust case the contacts are extremely close to 
%~ the cone limits (Figure~\ref{fig:stair_comp}).
%~ 
%~ 
%~ The geometry of the environment is easily addressed by our planner, and the contact planning is several times faster than real time in this scenario.
%~ 
%~ Again, the interpolation motion between the contact steps is out of the scope of this paper. However it should be noted that the computed plan in this scenario has been executed successfully on the robot (\url{http://youtu.be/YjL-DBQgXwk#t=0m28s}).
%~ 
%~ \begin{figure}
  %~ \centering
  %~ \begin{overpic}[width=0.5\linewidth]{figures/stair_robust}
		%~ \put (17,5) {\small{\color{red}$b_0 = 0.23$}} 
		%~ \put (79,5) {\small{\color{green}$b_0 = 6.16$}} 
	%~ \end{overpic}
  %~ \caption{
           %~ Evaluation of the robustness $b_0$ of two contact configurations. Although in equilibrium, the left configuration is on the verge of slipping.}
		   %~ \label{fig:stair_comp}
%~ \end{figure}

\subsubsection{HyQ -- DRC-style rubble (Figure~\ref{fig:darpa})}
The quadruped robot must cross a rubble composed of bricks rotated at different angles and directions.

\begin{figure}
  \centering
  \includegraphics[width=1\linewidth]{figures/darpa}
  \caption{
           Robust crossing of rubbles by HyQ. }
		   \label{fig:darpa}
\end{figure}

\noindent\textbf{Contacts involved:} All (the 4 legs).

\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $20$.

%~ \noindent\textbf{Observations:} In this context, setting up a really important minimum value for $b_0$ is possible due to the high
%~ stability of the HyQ robot, and results in more contact switches, in exchange for safety. The guide path-planning in this scenario takes a few seconds in average, more than
%~ in any other scenarios. This is explained by the necessity of discovering a safe way to ``climb down'' the rubble. In this part of the planning, the constraint that the 4 reachable workspaces of all legs must collide with the environment at all times is hard to respect, but enforces the equilibrium of HyQ. %\adnote{Why do not relax this constraint and require only 3 legs?} 
%~ Again, the computation times remain however \gls{interactive}.

\subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs})}
In this long scene, HyQ has to cross a 55-cm large hole, followed by a narrow ``bridge'', only 25-cm large.

\begin{figure}
  \centering
  \includegraphics[width=0.4\linewidth]{figures/HyQ_bridge}
  \caption{
           HyQ crossing a narrow bridge. }
		   \label{fig:HyQ_bridge}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1\linewidth]{figures/HyQ_obs}
  \caption{
           Crossing a hole contact sequence for HyQ. }
		   \label{fig:HyQ_obs}
\end{figure}

\noindent\textbf{Contacts involved:} All (the 4 legs).

\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $10$.

%~ \noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is a hard case for a contact planner.
%~ While finding a guide path above the hole is easy for the guide planner, finding a sequence of contacts that allows for equilibrium is not trivial.
%~ Second, the narrow bridge is hard both for the planner and the contact generator: to make sure that equilibrium is preserved along the traversal,
%~ the bridge must be approached with the appropriate angle.
%~ The difficulty is illustrated in Figure~\ref{fig:HyQ_obs}, where several feet rearrangements are required to cross the hole (although the video shows this best).
%~ The planner however succeeds in finding a feasible sequence in the end, again with \gls{interactive} computation times.

\subsubsection{HRP-2 -- Path re-planning (Figure~\ref{fig:re-planning})}
In this long scene, HRP-2 plans a path through several obstacles. The scene is edited during the execution of the motion: a stair is added,
some stepping stones are removed, and part of the final staircase is deleted. All these modifications require re-planning.

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{figures/replanning}
  \caption{
           HRP-2 in the re-planning scenario. After the red step stones are removed, a new sequence of contacts is re-planned. Hand contacts
           are not presented here for readability.}
		   \label{fig:re-planning}
\end{figure}

\noindent\textbf{Contacts involved:} Feet and the right arm.

\noindent\textbf{Heuristics:} $h_w$  for all legs. $h_{\textrm{\it EFORT}}$  for the right arm. The robustness threshold is set to $2$.

%~ \noindent\textbf{Observations:} This scenario is designed to illustrate concretely the computation times of the planner.
%~ In the video, the footsteps indicating the contact sequence appear at the average speed of their computation (including the guide-path planning).

%~ \subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot})}
%~ This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
%~ our contact planner to compute a contact sequence for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
%~ Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences in less than 5 seconds.
%~ 
%~ \begin{figure*}
%~ \centering
  %~ \begin{overpic}[width=1\linewidth]{figures/penrot}
	%~ \end{overpic}
%~ \caption{Contact sequence found for a pen manipulation in a zero gravity environment.}
		   %~ \label{fig:penrot}
%~ \end{figure*}
%~ 
 %~ 
%~ \noindent\textbf{Contacts involved:} Three finger-tips.
%~ 
%~ \noindent\textbf{Heuristics:} $h_{\textrm{\it EFORT}}$ for all fingers.

\subsection{Role of the main parameters} \label{sec:influence}
We discuss the factors that influence the outcome of our planner: the root scaling factor $s$ (Section~\ref{sec:scaling}), the heuristics for
contact generation (Appendix~\ref{sec:heuristics}), and lastly, the discretization step for the guide path. The appropriate value for these parameters
is computed empirically based on use-case analysis or trials and errors.

\subsubsection{Choosing the scaling factor $s$} \label{sec:params}
%~ \deladp{To find a convenient value $s^*$ of $s$, we proceeded as follows. }
For several values of $s$, we generated $10 000$ configurations. 
%~ We then computed the sensitivity of the reachability condition (percentage of configurations in $C_{Reach}^0$ belonging to \gls{$C_{Contact}^0$}).
%~ Similarly we computed the specificity of the condition (percentage of configurations \textbf{not} in $C_{Reach}^0$ \textbf{not} belonging to \gls{$C_{Contact}^0$}).
We then computed the sensitivity and specificity of the reachability condition.  In this context, the sensitivity refers to the percentage of configurations in $C_{Reach}^0$, effectively belonging to \gls{$C_{Contact}^0$}. If a sampled configuration is in $C_{Reach}^0$, but our method is unable to generate a contact configuration from it, as a result the sensitivity
decreases. The sensitivity thus illustrates the confidence we have that any configuration in $C_{Reach}^0$ will effectively lead to a contact configuration.
Conversely, the specificity refers to the percentage of configurations \textbf{not} in $C_{Reach}^0$, effectively \textbf{not} belonging to \gls{$C_{Contact}^0$}.
If a sampled configuration is \textbf{not} in $C_{Reach}^0$, but our method is able to generate a contact configuration from it, as a result the specificity
decreases. The specificity thus illustrates the confidence we have that all configurations that allow contact creation belong to $C_{Reach}^0$ (or informally, the confidence that we are not missing valid solutions).
We thus look for a compromise between sensitivity and specificity.

The obtained results for HRP-2 are shown in Table~\ref{tab:scale}, averaged over all scenes (except for the car egress: in this scenario, 
statistical tests are not really conclusive since we are only interested in a small area of the environment).

As it can be expected, the scaling results in a high increase of the sensitivity, with a decrease of the specificity.
For HRP-2 we decided to set $s^*=1.2$.
%~ , and was also effective for the car egress scenario.

\begin{table}
\centering
\footnotesize
\begin{tabular}{c | c | c}
   Value of $s$ &  Sensitivity & Specificity\\
 \hline
   1   & 76\% & 100 \%\\
   1.1& 88\% & 96\% \\
   1.15& 93\% & 94\%\\
   1.2 & 97\% & 92.5\%\\
   1.25& 98\% & 91.7\%\\
   1.5 & 99\% & 90.5\%\\
 \end{tabular}
\caption{Sensitivity and specificity values of the reachability condition, depending on the scaling value $s$ of $W^0$.}
\label{tab:scale}
\quad
\end{table}

\subsubsection{Choosing the heuristics} \label{sec:heuristichoices}
In our conference paper~\citep{tonneauisrr15}, the computed motions were generated using the EFORT heuristic.
EFORT is designed for tasks requiring large magnitude contact forces (such as pushing / pulling / climbing). 
In locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
configurations close to singularities (and joint limits). While this did not significantly impact
the generation of the plan, the resulting interpolation turned out to be harder.
For this reason, we prefer to use our manipulability-based heuristic for the legs of the robot, but we still
use EFORT for the arms, which results in fewer contact repositionings.

%~ \subsubsection{On the robustness equilibrium criterion:} \label{sec:parrob}
%~ Robustness is really important when considering practical applications on the robot, to account 
%~ for the various uncertainties that result from environment and state estimation. However,
%~ maximizing the robustness criterion is often not optimal, because the resulting configurations may be too conservative, thus not favoring the motion. In our experiments, we choose not to maximize the robustness,
%~ but to empirically se	t a robustness threshold value under which a configuration is not considered to be in
%~ static equilibrium (If no candidate reaches the threshold, instead of failing, the algorithm can eventually return the ``more robust'' configuration
%~ found). Currently for HRP-2, a threshold value of 2 subjectively gives the best results in the considered scenarios, while 10 seems to be a good choice for HyQ. Both values do not
%~ significantly slow down the planning times.

\subsubsection{Discretization of the guide path} \label{sec:disc}
The discretization step is a user-defined, fixed parameter. The step
has an influence on the output of the planner: if too large steps are taken,
the planner may fail since we impose the constraint that only one contact change might occur
between two consecutive steps. On the other hand, a small step will not impact the success rate of the planner, 
but may generate unnecessary states. In most scenarios the torso of HRP-2 moves about 15 cm between two postures, but only 3 cm
for the car egress scenario to handle the geometry of the car.
For future work, we would like to automatically adapt the size of the discretization step to the complexity of the environment.

\subsection{Performance analysis} \label{sec:perf}
To analyze performance, we ran the planner 1000 times for each scenario.
We measured the computation time spent in each part of the algorithm, and analyzed success rate.

\begin{table}
\centering
\begin{tabular}{ l | c}
  &  Path planning success rate \\
 \hline
   Stairs     	& 100\% \\
   Standing			& 68\% 	\\
   Car 			& 77\% 	\\
   Rubble 				& 97\% 	\\
   Race        & 88.0\% 	\\
 \end{tabular}
\caption{ Percentage of successful complete contact planning rates for each scenario, rounded to the first decimal.}
\label{tab:sucess_planning}
\quad
\end{table}

\begin{table}
\centering
\begin{tabular}{ l | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{35pt} | >{\centering\arraybackslash}m{35pt} | c}
  &  Equilibrium success rate & Kinematic failure & Equilibrium failure \\
 \hline
   High stairs 	& 99.5\%  & 0.1\% 	& 0.4\% \\
   Standing up 		& 87.8\%  & 6.1\% 	& 6.1\% \\
   Car egress 		& 66.2\%  & 15.9\% 	& 17.9\% \\
   Rubble 			& 97.54\% & 0.16\% 	& 2.3\% \\
   Obstacle race 	& 92.4\%  & 0.15\% 	& 7.45\% \\
 \end{tabular}
\caption{Success rates obtained for the generation of static equilibrium contact configurations for each scenario, rounded to the first decimal. Column 1 indicates 
indicates the rate of contact generation that succeeded. In the cases where the generation fails, it can be
either a kinematic issue (column 2), or because no contact configuration led to a static equilibrium configuration (column 3). Note that a failure in the contact generation
is not equivalent to a failure of the contact planning algorithm.}
\label{tab:requestpercent}
\quad
\end{table}

\begin{table*}[b!t!p!]
\centering
\footnotesize
\begin{tabular}{ >{\centering\arraybackslash}m{37pt} | >{\centering\arraybackslash}m{57pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{70pt} | >{\centering\arraybackslash}m{73pt} | >{\centering\arraybackslash}m{80pt} | >{\centering\arraybackslash}m{10pt}}
  Scenario (nb steps) &  Complete guide generation (ms) & Static equilibrium (ms) & Collision (ms) & Inverse Kinematics (ms) & Total generation time (ms) & Time per step (ms)\\
 \hline
   Stairs (18) 	& 5 -- \textbf{6} --  18 		 & 13 --  \textbf{32} -- 329   	& 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{15} \\
   Standing (24)& 65 -- \textbf{1086} --  5227   & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 371 --  \textbf{2257} -- 7671 & \textbf{94}  \\
   Car (86)& 320 -- \textbf{6971} --  44002 & 409 --  \textbf{1766} -- 14752   	& 297 -- \textbf{1187} -- 8483 & 3154 --  \textbf{15323} -- 165541 & 5834 --  \textbf{31391} -- 281000 & \textbf{365}\\
   Rubble (82)& 37 -- \textbf{573} --  1685 & 583 --  \textbf{2714} -- 9459 & 491 --  \textbf{1971} -- 6273 & 269 --  \textbf{706} -- 3118 & 1811 --  \textbf{7195} -- 23241 & \textbf{86} \\
   Race (134)& 14 -- \textbf{51} --  125 & 455 --  \textbf{1359} -- 21045   & 397 --  \textbf{923} -- 9924 & 228 --  \textbf{471} -- 5415 & 1436 --  \textbf{3343} -- 41446 & \textbf{25}
 \end{tabular}
\caption{minimum, \textbf{average} and worst time (in ms) spent in the generation process for each scenario and each critical part of the generation process (not all parts are timed,
thus the average total computation time is higher than the sum of each part). The last
column indicates the average time necessary to compute one contact transition. The Collision column times includes the (negligible) octree intersection operation necessary to retrieve the candidate samples.}
\label{tab:requestime}
\quad
\end{table*}

\subsubsection{Success rates (Table~\ref{tab:sucess_planning})}
%~ Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
Despite the complexity of the scenarios and the approximations made in our formulation, our planner succeeded in the large majority of cases.

Table~\ref{tab:requestpercent} presents the rate of successful contact generation. Note that a failure in contact generation for a root configuration is not equivalent to a failure in the contact plan. It simply means that another limb was tested for contact generation for the same root configuration.
As expected, a more constrained scenario such as the car egress provides less satisfying results, despite the high success rate of the planner.

\subsubsection{Computation times (Table~\ref{tab:requestime})}
 %~ summarizes the computation times.

For HRP-2, most of the time was spent performing inverse kinematics.
This is not surprising considering the number of calls to the methods: IK projection is used intensively to maintain contact continuity between two postures; 
it is also applied every time a new candidate needs to be evaluated. In particular for the car egress scenario,
the kinematic constraints are very demanding to avoid collisions.

On the other hand for HyQ most of the time is spent testing the static equilibrium of the candidate configurations.

In all scenarios, one can observe that the average computation time for one single step is largely below one second,
thus enabling \gls{interactive} applications and online autonomous planning of the robot motion.

\subsubsection*{Conclusion}
These results confirm that our approach provides a satisfying compromise between completeness and efficiency, thus enabling online planning
while controlling the robot. Indeed, when the contact planning fails, it fails rapidly. This allows us to rapidly re-plan with a reasonable chance of success.
The most efficient (and immediate) approach to obtain a valid contact plan as fast as possible would be to launch in parallel several instances of the planner (our current implementation is single-threaded) and to use any successful result as a plan for solver $\mathcal{P}_3$.

\begin{table}
\centering
\begin{tabular}{ c | c | c }
 Scenario & Method  & Computation time \\
 \hline
   \multirow{3}{*}{Stair 20 cm} & Hauser~\cite{Hauser06usingmotion} &  5.42 min  \\							 
							  & Mordatch et al.\cite{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
							 & \textbf{Ours} + \cite{Carpentier2016}  & \textbf{$ <$ 2s} \\
 \hline
   \multirow{3}{*}{Stair 30 cm} & Hauser~\cite{Hauser06usingmotion} &  4.08 min  \\
							 & Mordatch et al.\cite{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
							 & \textbf{Ours}  & \textbf{$ <$ 2s}   \\
 \hline
   \multirow{3}{*}{Stair 40 cm} & Hauser \cite{Hauser06usingmotion} &  10.08 min  \\
							 & Mordatch et al.\cite{Mordatch:2012:DCB:2185520.2185539} & 2 to 10 min \\
							 & \textbf{Ours}   & \textbf{$ <$ 5s}   \\
 \hline
   \multirow{2}{*}{Table (car) egress} & Bouyarmane et al.~\cite{Bouyarmane2009, DBLP:conf/iser/EscandeKMG08} & 3.5 hours  \\
							 & \textbf{Ours}  & \textbf{$<$ 60 s} \\

 \end{tabular}
\caption{Comparison between the computation times obtained by our method and previous ones for addressing the whole problem.}
\label{tab:compprev}
\quad
\end{table}

\subsection{Comparison with previous work} \label{sec:compa}
%~ Few contributions provided computation times on the complete problem. To provide a completely fair comparison, it would be required to compare the methods on the same computers and exact same scenarios, which is not possible because 
%~ the source code of the other methods is often not available. %, necessary to provide comparison between our approaches. 
We did our best to provide a fair comparison of the computation complexity of our method with the state of the art. 
%Few contributions provided computation times on the complete problem. 
However existing benchmarks for motion planning algorithms~\cite{moll2014extensible} do not yet encompass contact planning.
Moreover, the source code of the previous methods of the state of the art is often not available.
Providing a fair comparison with the algorithms performing on the same computer and on the same scenarios is yet out of reach.
A step in this direction is the open-source release of our source code (see Section~\ref{app:hpp}) that allows any reader to reproduce our results.
Furthermore, $\mathcal{P}_3$ remains challenging in the presence of obstacles. The only valid scenarios addressed completely in previous works are thus the stair-climbing scenarios of different heights proposed by Hauser in \cite{Hauser06usingmotion}, and the table-egress scenario by Escande et al. in~\cite{DBLP:conf/iser/EscandeKMG08}, which we consider to be of similar complexity with respect to the car-egress scenario (we did not consider the stairs in the scene). Both scenarios are tested with HRP-2.

%~ Table~\ref{tab:compprev} presents the computation times for these scenarios, clearly demonstrating that our approach is order of magnitude faster than previous works.
Table~\ref{tab:compprev} presents the computation times for these scenarios. Since no implementation of the previous methods is available, we had no choice but to indicate the times directly taken from their corresponding papers.
In the case of older contributions such as~\cite{Hauser06usingmotion},~\cite{DBLP:conf/iser/EscandeKMG08} and~\cite{Bouyarmane2009}, because of the technological progress, it appears that our results would have been more meaningful if the benchmarks were run on a modern computer. However, since the computation times differ of several orders of magnitude, we believe that these results clearly show the computational benefits of our method.

 %~ Despite this observation and the fact that some results are old, we believe that the results show that our method
%~ performs better. Indeed, a difference in technology does not justify a difference as important as the one demonstrated in Table~\ref{tab:sucess_planning}.

% !TEX root =  ../main_tro.tex
 \section{Discussion: validity and purpose of our contact planner} 
\label{sec:discussion}

%~ \subsection{Accuracy and performance gain of the reachability condition}

As demonstrated in the results section, the main purpose of our method is the reduction of the algorithmic complexity of the problem, which leads to an interactive 
application. This property is critical for 
online applications with the robot and was not proposed by any of the previous contributions. Our method addresses highly constrained environments while improving the search time by orders of magnitude. This high performance is reached at the cost of some approximations that we discuss here. 
%~ In previous approaches, we identified 2 sources of high computational cost.

The first approximation is the verification of \contactreachability\ ($\mathbf{q}^0 \in C_{Contact}^0$).  Our \textit{reachability condition} ($\mathbf{q}^0 \in C_{Reach}^0$) is computationally efficient and provides an accurate approximation of $C_{Contact}^0$ (Section~\ref{sec:scaling}). This is demonstrated by the second column of Table~\ref{tab:requestpercent}, and illustrated by Figure~\ref{fig:dedefeas}. Indeed, in the large majority of cases, (84\% in the worst car egress case), we are able to find a contact configuration for any configuration in $C_{Reach}^0$.

%~ These results clearly demonstrate that the reachability condition, presented in Section~\ref{sec:scaling}, provides an accurate representation of the 
%~ contact we combine only-necessary and only-sufficient conditions which are trivial to estimate. Only-necessary conditions are appealing because they preserve the completeness of the search, while reducing 
%~ the search space: they provide an outer approximation of $C_{Contact}^0$. On the other hand, only-sufficient conditions provide the guarantee that any configuration that satisfies them is indeed \gls{contact reachable}: they provide an inner approximation of $C_{Contact}^0$.

Another source of computational cost identified in previous works is the verification of \equilibriumfeasibility. 
The main assumption of our work is that for the class of problems we consider \contactreachability\ implies \equilibriumfeasibility.
Our scenarios show that the  assumption is verified in the majority of cases when at least one contact surface is \textit{quasi flat}~\cite{Prete2016}, that is when
the friction cone of the contact surface contains the direction opposite to the gravity.
Figure~\ref{fig:dedefeas} illustrates this observation, demonstrated empirically by the third column of Table~\ref{tab:requestpercent}. In the worst case, in our experiment
the assumption was verified for 82\% of the total amount of trials that verified \textit{contact reachability}.
In the example of \cite{Bouyarmane2009}, the verification of \equilibriumfeasibility\ implies a constructive demonstration by exhibiting a valid $\mathbf{q}^{\overline{0}}$, requiring
several minutes of planning. Our method, in comparison, takes from a few milliseconds to several seconds.

These results clearly justify our pragmatic approach.

%~ \subsection{Addressing $\mathcal{P}_3$}
%~ \adnote{This subsection seems redundant to me}
%~ This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation
%~ . Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
%~ for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
%~ The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.
%~ In these scenarios the robot is able to execute our plans, while respecting the constraints inherent to the problem: joint kinematics and torque limits and
%~ collision avoidance.

\begin{figure}[t]
\centering
  \begin{overpic}[width=1\linewidth]{figures/2D_feas}
		\put (15,) {$C_{Equil}^0$      $\subset$} 
		\put (47,) {$C_{Contact}^0$ $\approx$ } 
		\put (76,) {$C_{Reach}^0$} 
	\end{overpic}
\caption{Illustration of several root configurations sets used in this paper in a 2D scene. Obstacles are violet, and units are in meters. To show the sets in a 2D representation, all the rotational joints of HRP-2 are locked in the shown configuration, such that a torso configuration
is only described by two positional parameters (x and y). The root of the robot is indicated with a black cross. To compute the reachable workspace, the point on the ankle indicated by a green cross was used. $C_{Equil}^0$ is included in $C_{Contact}^0$. $C_{Reach}^0$ approximates $C_{Contact}^0$. Depending on a parametrization, we can obtain $C_{Contact}^0 \subset C_{Reach}^0$. Considering the configurations around the top obstacle, we can observe a similarity between  $C_{Equil}^0$  and $C_{Contact}^0$ when the reachable workspace of the legs includes \textit{quasi-flat} surfaces.}
		   \label{fig:dedefeas}
\end{figure}

\section{Conclusion} 
\label{sec:conclusion}

In this paper we consider the multi-contact planning problem, formulated as three sub-problems  $\mathcal{P}_1$,  $\mathcal{P}_2$,  $\mathcal{P}_3$, addressed sequentially. While we propose a global framework that handles all these problems, our contribution focuses on  $\mathcal{P}_1$ and  $\mathcal{P}_2$.
The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
the second problem $\mathcal{P}_2$ is the computation of a discrete sequence of whole-body configurations along the root path.
We believe that this decomposition is currently the most promising approach towards
a global resolution of the problem. We also claim to have achieved a significant step towards this objective thanks to the dimensionality reduction provided by
the reachability condition. With our results and the release of our source code, we hope to inspire further research in this direction.

Our contribution to \Pa is the introduction of a low-dimensional space $C_{reach}^0$, an approximation of the space of \gls{equilibrium feasible} root configurations.
$C_{reach}^0$ can be efficiently sampled and has a low-dimension. For these reasons we are able to solve \Pa much faster than previous approaches.

%~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
Our contribution to \Pb is a fast contact generation scheme that can optimize user-defined criteria.

Our results demonstrate that our method allows a pragmatic compromise between three 
criteria that are hard to reconcile: generality, performance, and quality of the solution, making it the first acyclic contact
planner compatible with \gls{interactive} applications.

%
\textbf{Regarding generality}, the \textit{reachability condition}, coupled with an approach based on limb decomposition, 
allows the method to address automatically arbitrary legged robots.
%
\textbf{Regarding performance}, our framework is efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
%
\textbf{Regarding the quality of the paths}, we are able to compute
\gls{equilibrium feasible} paths in all the presented scenarios, with high success rates.
As for \cite{Bouyarmane2009}, failures can still occur, due to the approximate condition used to compute the guide path.
The low computational burden of our framework however allows for fast re-planning in case of failure.
Furthermore, because of this approximation, the guide search is not complete. The choice is deliberate, because we believe
that it is necessary to trade completeness for efficiency at all stages of the planner.
However, one direction for future work is to focus on a more accurate formulation of $C_{reach}^0$ to improve
the approximation.

Our method applies to any scenario where at least one contact friction cone contains
the direction opposed to the gravity (i.e. \textit{quasi-flat}). This class of scenarios include all the problems proposed at the DARPA Robotics Challenge.
One way to further extend its range of application, which we consider for future work, is to include the equilibrium criterion when solving $\mathcal{P}_1$.
Considering the set of obstacles intersecting with the reachable workspace for a given root configuration as candidate surfaces, we can use them to verify the equilibrium criterion.
This would give us a necessary condition for \equilibriumfeasibility. %, which we investigate in~\cite{fernbach:hal-01486933}.

While we have exhibited complete multi-contact locomotion obtained with our contact planner, our main concern for future work is to address the interpolation
between contact sequences ($\mathcal{P}_3$), which remains an open issue in highly-constrained scenarios.
Solving $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem in the interpolation phase, an issue 
not addressed by existing frameworks. We aim at providing our plans with transition certificates, that would define constraints on $\mathcal{P}_3$, under which
the transition between two contact configurations is feasible and collision-free.
%~ Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
%~ at the center of mass of the robot.
 %~ We are also
%~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
Finally, we aim at performing kinodynamic planning to remove the constraint that configurations be in static equilibrium. We believe that the most promising direction in this regard is to integrate the notion of Admissible Velocity Propagation \citep{DBLP:conf/rss/PhamCN13}.
Addressing these two issues is essential to bridge the gap between the planning and control aspects of legged locomotion.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.
% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
% labels, the optional argument to \subfloat must be present. If a
% subcaption is not desired, just leave its contents blank,
% e.g., \subfloat[].

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table and, given that table
% captions serve much like titles, are usually capitalized except for words
% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
% and up, which are usually not capitalized unless they are the first or
% last word of the caption. Table text will default to \footnotesize as
% the IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively. 
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

\appendices

%~ \clearpage
\section{Generating the $W$ volumes for HRP-2}
\label{app:rom}

We detail our method to generate the volumes $W$ used
in RB-RRT, with the example of HRP-2.
The kinematic tree is split into four limbs $R^k$.
The arms are connected to the shoulders, and the legs to the root.
The obtained volumes $W$ are shown in Figure~\ref{fig:hrp2_w}.

\begin{figure}
\centering
  \begin{overpic}[width=1\linewidth]{figures/hrp2_w}
	\end{overpic}
\caption{The $W$ volumes computed for HRP-2. The red shapes are $W^0$. The green shapes represent the $W^k$.}
		   \label{fig:hrp2_w}
\end{figure}

\subsection{Step 1: computing the reachable workspace $W^k$ of a limb}

To generate a volume $W^k$, we proceed as follows:
\begin{enumerate}
\item Generate randomly $N$ valid limb configurations for $R^k$, for $N$ really large (say $100000$);
\item For each configuration, store the 3D position of the end effector joint relatively to the root of $R^k$; then compute the convex hull of the resulting point cloud;
\item The resulting polytope can contain a very large number of faces.  A last step is thus to simplify it using an incremental decimation method~\cite{Garland:1997:SSU:258734.258849}.
Variations of this method are commonly implemented in most authoring tools. For our experiments we used the blender decimate tool. Details of its use can be found in the static webpage associated with this paper (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#decimate}). For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
%~ (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}). For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
\end{enumerate}

\begin{figure}
\centering
  \begin{overpic}[width=1\linewidth]{figures/roms}
	\end{overpic}
\caption{Different approximations of the range of motion of the right arm of HRP-2. Left: non convex-hull, computed with the powercrust algorithm~\citep{Amenta:2001:PC:376957.376986}. Middle:
convex hull of the reachable workspace. Right: Simplified hull used in our experiments.}
		   \label{fig:hrp2_roms}
\end{figure}

Figure~\ref{fig:hrp2_roms} illustrate the obtained $W^k$ for HRP-2.
Regarding the procedure, we can see that step 2 is conservative (Figure~\ref{fig:hrp2_roms}--right), which 
is acceptable, especially because the lost set essentially relates to configurations close to singularity (they are close to the boundaries of the reachable workspace, and
often not \gls{contact reachable}, as illustrated in Figure~\ref{fig:dedefeas}, where the exterior boundaries of the reachable workspace appear
red, thus not belonging to $C_{Contact}^0$). We choose again to be less complete but more efficient, regarding the number of collision tests to be performed by RB-RRT.
In step 1 on the other hand, selecting the convex hull (Figure~\ref{fig:hrp2_roms}--middle) instead of a minimum encompassing shape (Figure~\ref{fig:hrp2_roms}--left) may introduce false positives.
Concretely, because the false positive set intersects with $W^0$, the scaling volume of the robot torso, the induced error is compensated,
as verified by the results shown by Table~\ref{tab:requestpercent}.

\subsection{Step 2: computing the torso scaling workspace $W^0$ of the robot}
To define the volume $W^0$ of HRP-2, we proceed in an empirical manner.
First, we compute the bounding boxes of the robot torso, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
Then, we perform a scaling of these boxes by a factor $s$. 
The higher $s$ is, the more likely sampled configurations are to be feasible, but the less complete is the approach.
To compute the appropriate value of $s$, we proceed as described in Section~\ref{sec:params}, and choose empirically
$s^*=1.2$ as the appropriate value for HRP-2.
%~ \section{Heuristics and criteria for contact selection}
%~ \label{sec:heuristics}
% !TEX root =  ../main.tex
%~ \subsection{Manipulability-based heuristics for contact selection}
\section{Manipulability-based heuristics for contact selection}
\label{sec:heuristics}
This Section proposes two heuristics to select a contact that optimizes desired capabilities.
%~ These heuristics allow to locally optimize a contact location with respect to a particular task.
For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion, or to stay away from singular configurations.
We derive these heuristics from the work by \cite{Yoshikawa1984}, recalled here. %, that we present first.

\subsubsection{The force and velocity ellipsoids}

\begin{figure}[!tbp]
  \centering
	\begin{overpic}[width=1\linewidth]{figures/EFORT/ellipsoid}
		\put (8.5,1.2) {\small{Velocity ellipsoid}}
		\put (62.5,1.2) {\small{Force ellipsoid} \tiny{(scale 0.5)}}
	\end{overpic}
  \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 DoFs and 2 segments.
Only the horizontal and vertical speeds are shown (not the rotation speeds).}
 %~ since it would require being able to draw in four dimensions.}
		   \label{sec:efort_ellipsoid}
\end{figure}

We consider: a limb configuration $\mathbf{q}^k$; its end effector position $\mathbf{p}^k$; its Jacobian matrix
$\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
Yoshikawa~\cite{Yoshikawa1984} defines the velocity(~\ref{eq:vel}) and force~(\ref{eq:for}) ellipsoids:

 \begin{equation} 
 \label{eq:vel}
\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} \leq 1 
\end{equation}

 \begin{equation} 
 \label{eq:for}
\mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
\end{equation}

They describe the set of end-effector velocities (respectively forces) that can
be reached under the constraint $||\dot{\mathbf{q}}||^2 \leq 1$ for the current configuration.
The longer the axis of the ellipsoid is, the more important the velocity (resp. force) of the end-effector the direction of the axis can be (Figure~\ref{sec:efort_ellipsoid}).

\subsubsection{Manipulability-based heuristics}
From these definitions, we can derive two useful heuristics, that all account for the environment and the task being performed.
The first one, EFORT, was introduced by \cite{Tonneau2014}; the second one derives the manipulability measure proposed by Yoshikawa~\cite{Yoshikawa1984}.

With EFORT, we define the efficiency of a configuration as the ability of a limb to exert a force in a given direction.
We thus consider the force ellipsoid as a basis for our heuristic.
In a given direction $\uprho$, the length of the ellipsoid is given by the force-transmission ratio \citep{1087795}:
\begin{equation*}
f_\mathsf{T}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}	^{T})\uprho]^{-\frac{1}{2}}
\end{equation*}

In our problem, to compare candidate configurations, we include the quality of the contact surface, and choose $\uprho$ as the direction
opposite to the local motion (given by the difference between two consecutive root positions):

\begin{equation}
h_{\textrm{\it EFORT}}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
\end{equation}
where $\mu$ and $\mathbf{n}$ are respectively the friction coefficient and the normal vector of the contact surface.

%~ If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
%~ \begin{equation}
%~ h_{vel}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
%~ \end{equation}

%~ $h_{\textrm{\it EFORT}}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
$h_{\textrm{\it EFORT}}$ will favor contacts that allow large efforts.
$EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
In other less demanding cases, manipulability can also be considered to avoid singularities.
To do so, we can consider the manipulability measure $h_{w}$, also given by Yoshikawa~\cite{Yoshikawa1984}:

\begin{equation} \label{ellipsoid}
h_{w}(\mathbf{q}) = \sqrt{det(\mathbf{J}\mathbf{J}^T)}
\end{equation}
$h_{w}$ measures the ``distance'' of a given configuration to singularity. When $h_{w}$ is equal to 0, the configuration is singular;
the greater $h_{w}$ is, the further away the configuration is from singularity.

%~ \input{body/equilibrium}
% !TEX root =  ../main.tex
\section{Addressing $\mathcal{P}_3$}
\label{app:optim}
The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
%~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
To address the other scenarios, we propose a new implementation, entirely open source (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#optim} ), which can be integrated directly in our motion planner software. This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
The center-of-mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.

We rewrite Eq.~\ref{eq:new_eul} in the general case:

%~ \begin{align} \label{eq:new_eul}
%~ \underbrace{
%~ \mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
%~ \hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
%~ }_\mathbf{G} \bm{\beta}, = 
%~ \underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
%~ \underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
%~ \end{align}

\begin{align} \label{eq:new_eul_acc}
\mathbf{G} \bm{\beta} = 
\underbrace{\mat{m (\mathbf{\ddot{c}} - \mathbf{g}) \\m \mathbf{c} \times (\mathbf{\ddot{c}} - \mathbf{g}) + \dot{L}}}_{\mathbf{w}}
%~ \underbrace{m (\mathbf{\ddot{c}} - \mathbf{g})  \\ m (\mathbf{\ddot{dc}} - \mathbf{g}) + \mathbf{\dot{L}} }_{\mathbf{w}}
\end{align}

\noindent where $\dot{L}$ is the angular momentum expressed at the com.
Eq.~\ref{eq:new_eul_acc} defines a 6-dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
this cone determines all the admissible wrenches $\mathbf{w}$ that can be generated by contact forces inside their friction cones.
The face form of $\mathcal{K}$ can be computed using the double description method~\citep{Fukuda1996}, resulting in the following linear inequalities:

\begin{equation}
\label{eq:cone_k}
	\mathcal{K} =  \left\{ \mathbf{w}, \mathbf{A}\mathbf{w} \leq \mathbf{b	} \right\}
\end{equation}

The objective 
is then to plan a trajectory for the center of mass such that the generated $\mathbf{w}$ always verifies Eq.~\ref{eq:cone_k}. 
We now consider two contact configurations  $\mathbf{q}_0$ and $\mathbf{q}_1$ computed by our planner: in the general case one contact is broken and one created to get from
$\mathbf{q}_0$ to $\mathbf{q}_1$. We manually define the duration of each of the three contact phases.
In each phase $s$ the centroidal wrench $\mathbf{w}$ is constrained to lie inside a cone $\mathcal{K}_u, u =0 \dots 2$.
We call the total duration of the motion $T$, and formulate the following optimization problem:

\begin{equation} \label{eq:qp_rob} \begin{aligned}
%~ \find \quad & \mx{R} \in \Rv{3\times T}, \mxdd{C} \in  \Rv{6\times T} \\
%~ \minimize_{\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)}  \quad & \sum_{u=0}^{2} \int_{t_u}^{t_u + \Delta t_u}  \ell(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)) \text{dt}  \\
\minimize_{\mathbf{\ddot{c}}(t), \dot{L}(t)}  \quad & \sum_{u=0}^{2} \int_{t_u}^{t_u + \Delta t_u}  \ell(\mathbf{\ddot{c}}(t), \dot{L}(t)) \text{dt}  \\
%~ \st \quad & \mathbf{\ddot{c}}(t) = dyn(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{w}(t)) \\
%~ \st \quad & \mathbf{c}(t) =  dyn(\mathbf{\ddot{c}}, \mathbf{c}_{\mathbf{q}_0} )  \\
\st \quad & \mathbf{A}_u \mathbf{w}(t)  \leq \mathbf{b}_u , \forall t \in  \left[t_u, t_u + \Delta t_u \right[, \forall u \\
	\quad & \mathbf{Y}_u \mathbf{c}(t)  \leq \mathbf{y}_u , \forall t \in  \left[t_u, t_u + \Delta t_u \right[, \forall u \\
	\quad & \mathbf{c}(0)  = \mathbf{c}_{\mathbf{q}_0} \\
	\quad & \mathbf{c}(T)  = \mathbf{c}_{\mathbf{q}_1} \\
	\quad & \mathbf{c}(0)  = \mathbf{\dot{c}}(0)  = \mathbf{\ddot{c}}(0) = \mathbf{0}\\
	\quad & \mathbf{c}(T)  = \mathbf{\dot{c}}(T)  = \mathbf{\ddot{c}}(T) = \mathbf{0}\\
    %~ \quad & ({\vct{r}}\tra\vct{r}=1,  \forall t  ?) \\
    %~ \quad & \delta t \sum\limits_1^{T-1}{\vc{r}^1}\tra {\vc{r}^1} \ge t_{transit} \\
    %~ \quad & \mx{A}_{\int} \mxdd{C} \leq \vc{a}_{\int} \\
    %~ \quad & \mx{V}_{\int} \mxd{C} \leq \vc{v}_{\int} \\
    %~ \quad & \mx{B}_{\int}^{1} \mx{C} \leq \vc{b}_{\int} \\
    %~ \quad & [\vc{c}_0\tra, \vcd{c}_0\tra]\tra = \vc{x}_0 \\
    %~ \quad & \vc{c}_T = \vc{c}_T^{input} \\
\end{aligned} \end{equation}
%~ \adnote{Did you really use w as a variable of your problem? Or do you compute it inside your constraint function as a function of the CoM pos and acc? And what about dL? It appears in the definition of w but then it is no longer there in the problem definition!}
The cost function $\ell$ is a weighted sum of the angular momentum and center-of-mass acceleration variation over the whole trajectory.
The center-of-mass positions and velocities $\mathbf{c}(t)$, $\mathbf{\dot{c}}(t)$ are internal variables, obtained through the double integration of $\mathbf{\ddot{c}}(t)$.
Then $\mathbf{w}(t)$ is obtained directly from these variables. $\mathbf{c}_{\mathbf{q}_0}$ and $\mathbf{c}_{\mathbf{q}_1}$ are the center-of-mass positions for configurations
 $\mathbf{q}_0$ and $\mathbf{q}_1$ respectively.
%~ The system dynamics $dyn$ tie together the position, velocity and acceleration of the center of mass. 
$\mathbf{Y}_u$ and $\mathbf{y}_u$ denote stacked kinematic constraints on the center of mass position, determined by the active contact locations.
The inequalities for each contact are determined in the same way that the reachable workspace is computed in Appendix~\ref{app:rom}, with
the effector serving as root. % (a triangulated object is a polytope, and as such can be described as a set of linear inequalities).

This formulation can trivially be extended over the whole contact sequence.
In our implementation, the problem is discretized using time steps of 100 ms.

The output of this optimization problem is an admissible center-of-mass trajectory.
To compute the whole body motion, we use a two-step approach.

First, we plan a kinematic motion for the robot, subject to the contact constraints. We also constrain the center of mass to follow
the computed trajectory. This is achieved using a constraint-based RRT planner~\citep{7759083}. As a result we obtain a collision-free whole-body motion.

The entire resolution takes approximatively 1.5 seconds for a complete contact transition.

%This motion is then validated in our dynamic simulator~\citep{DelPrete2015b}, using the trajectory as a reference postural task for an inverse-dynamics
%controller that accounts for all the robot constraints. This final result is shown in the companion video.
%~ \input{body/appendix_algo}
%~ \input{body/appendix_hpp}

\section*{Acknowledgements}
This research is supported by Euroc (project under FP7 Grant Agreement  608849);  Entracte (ANR  grant  agreement  13-CORD-002-01);
the ARO Contract W911NF-14-1-0437; and the NSF award 1305286. %, and a grant from Boeing.

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
 %~ argument is your BibTeX string definitions and bibliography database(s)
\bibliography{biblio}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%~ \begin{thebibliography}{1}
%~ 
%~ \bibitem{IEEEhowto:kopka}
%~ H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
  %~ 0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%~ 
%~ \end{thebibliography}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

%~ \begin{IEEEbiographynophoto}{Steve Tonneau}
%~ TODO
%~ \end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

%~ \begin{IEEEbiographynophoto}{Jane Doe}
%~ Biography text here.
%~ \end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2018-02-19 cc.sh
cleaning figures and comrpessing images to fit limit

@@ -1,4 +1,3 @@
-pdflatex main_tro.tex
-bibtex main_tro.aux
-pdflatex main_tro.tex
-mv main_tro.pdf root.pdf
+pdflatex final.tex
+bibtex final.aux
+pdflatex final.tex

### 2018-02-19 final.tex
cleaning figures and comrpessing images to fit limit
removed 2 figures to make room

@@
 \newcommand{\refframe}[1]{\mbox{\textless#1\textgreater}}	% to denote a reference frame
-%\DeclareMathOperator*{\lexmin}{\text{lex}\!\min}			% lexmin
 \DeclareMathOperator*{\argmin}{\arg\!\min}				% argmin
@@
 
-%\algnewcommand{\algorithmicgoto}{\textbf{go to}}%
-%%\algnewcommand{\Goto}[1]{\algorithmicgoto~\ref{#1}}%
-%\algnewcommand{\Goto}{\algorithmicgoto\xspace}%
-%\algnewcommand{\Label}{\State\unskip}
-
 \newenvironment{definition}[1][Definition]{\begin{trivlist}
@@
 
-\newcommand{\commentt}[2]{\textcolor{#1}{\textbf{\textit{[#2]}}}} 	% make sure no comments left in the paper
-%\newcommand{\commentt}[2]{}							% using this line rather than the previous one you remove all comments
-\newcommand{\adnote}[1]{\commentt{blue}{AD: #1}}
-\newcommand{\nmnote}[1]{\commentt{OliveGreen}{NM: #1}}
-
-% Another method to track changes
-% Examples of usage:
-% - This is \added[id=per,remark={we need this}]{new} text.
-% - This is \deleted[id=per,remark=obsolete]{unnecessary}text.
-% - This is \replaced[id=per]{nice}{bad} text.
-% To print the list of change use: \listofchanges
-\usepackage{changes}	% use this for the working version
-%\usepackage[final]{changes} % use this for the final version
-\definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
-\newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
-\newcommand{\addadp}[1]{\added[id=adp]{#1}}
-\newcommand{\repadp}[2]{\replaced[id=adp]{#1}{#2}}
-\definechangesauthor[name={Steve Tonneau}, color=blue]{st}
-\newcommand{\delst}[1]{\deleted[id=st]{#1}}
-\newcommand{\addst}[1]{\added[id=st]{#1}}
-\newcommand{\repst}[2]{\replaced[id=st]{#1}{#2}}
 \newcommand{\gls}[1]{\textit{#1}}
@@
   \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing1.jpeg}
+    {./figures/stairclimbing1.jpeg}
   \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing2.jpeg}
+    {./figures/stairclimbing2.jpeg}
   \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing3.jpeg}
+    {./figures/stairclimbing3.jpeg}
   \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing4.jpeg}
+    {./figures/stairclimbing4.jpeg}
   \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing5.jpeg}
+    {./figures/stairclimbing5.jpeg}
   \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing6.jpeg}
+    {./figures/stairclimbing6.jpeg}
   \includegraphics[trim={7.0cm 0.0cm 20.0cm 0.0cm}, clip, width=\widthValue]
-    {./video/stairclimbing7.jpeg}
+    {./figures/stairclimbing7.jpeg}
   \caption{
@@
 
-\subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs})}
+\subsubsection{HyQ -- Obstacle race }
 In this long scene, HyQ has to cross a 55-cm large hole, followed by a narrow ``bridge'', only 25-cm large.
 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.4\linewidth]{figures/HyQ_bridge}
-  \caption{
-           HyQ crossing a narrow bridge. }
-		   \label{fig:HyQ_bridge}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/HyQ_obs}
-  \caption{
-           Crossing a hole contact sequence for HyQ. }
-		   \label{fig:HyQ_obs}
-\end{figure}
+%~ \begin{figure}
+  %~ \centering
+  %~ \includegraphics[width=0.4\linewidth]{figures/HyQ_bridge}
+  %~ \caption{
+           %~ HyQ crossing a narrow bridge. }
+		   %~ \label{fig:HyQ_bridge}
+%~ \end{figure}
+
+%~ \begin{figure}
+  %~ \centering
+  %~ \includegraphics[width=1\linewidth]{figures/HyQ_obs}
+  %~ \caption{
+           %~ Crossing a hole contact sequence for HyQ. }
+		   %~ \label{fig:HyQ_obs}
+%~ \end{figure}
 
@@
   \centering
-	\begin{overpic}[width=1\linewidth]{figures/EFORT/ellipsoid}
+	\begin{overpic}[width=1\linewidth]{figures/ellipsoid}
 		\put (8.5,1.2) {\small{Velocity ellipsoid}}

### 2018-02-19 reviews/reviews_tro/reviewer_response.tex
moved all review process to associated folder

--- 2017-10-10 reviews_tro/reviewer_response.tex
@@
 We now join a diff file comparing the original and current submissions.
-The text appearing in blue denotes the editions. We used the latexdiff tools to generate this file, which is not fully compatible with all the latex packages that we use.
-For this reason, errors appear in the document, in particular regarding the subsections numbering. However, we think that this diff file will help the Reviewers locate precisely
+The text appearing in blue denotes the editions.
+Most of the Figures have been removed, with the exception of the new Figure 8, to fit the size limit for the answer letter.
+We think that this diff file will help the Reviewers locate precisely
 where the modifications occured in the document. The Reviewers will also notice that we reformulated the description our P3 solver in Appendix C, to better clarify what variables

### 2018-02-19 code/data/config/conf_hrp2_polaris.py

@@ -1,107 +1,0 @@
-from math import sqrt
-import numpy as np
-
-''' *********************** USER-PARAMETERS *********************** '''
-SOLVER_ID       = 0;    # classic TSID formulation
-SOLVER_TO_INTEGRATE         = [SOLVER_ID];
-DATA_FILE_NAME              = 'data';
-TEXT_FILE_NAME              = 'results.txt';
-SAVE_DATA                   = True;
-
-''' INITIAL STATE PARAMETERS '''
-SMOOTH_FILTER_WINDOW_LENGTH = 41;
-MAX_TEST_DURATION           = -1;
-dt                          = 1e-3;
-FDR                         = '/home/adelpret/repos/20151101_contact_planner_steve/code/hpp_serialization/';
-INPUT_FILE_NAME             = [FDR+'polaris_2_compressed',FDR+'polaris_3_compressed',FDR+'polaris_4_compressed',
-                               FDR+'polaris_5_compressed',FDR+'polaris_6_compressed'];
-INPUT_FILE_NAME             = [FDR+'polaris_6_compressed'];
-INPUT_FILE_NAME             = [FDR+'polaris_5_compressed'];
-INPUT_FILE_NAME             = [FDR+'polaris_4_compressed'];
-INPUT_FILE_NAME             = [FDR+'polaris_3_compressed'];
-INPUT_FILE_NAME             = [FDR+'polaris_1_compressed',FDR+'polaris_2_compressed',FDR+'polaris_3_compressed',FDR+'polaris_4_compressed',
-                               FDR+'polaris_5_compressed'];
-#INPUT_FILE_NAME             = [FDR+'polaris_1_compressed'];
-model_path                  = ["/home/adelpret/devel/sot_hydro/install/share"];
-urdfFileName                = model_path[0] + "/hrp2_14_description/urdf/hrp2_14_reduced.urdf";
-sceneFileName               = "/home/adelpret/repos/20151101_contact_planner_steve/code/hpp_serialization/polaris.stl";
-freeFlyer                   = True;
-
-''' CONTROLLER CONFIGURATION '''
-ENABLE_CAPTURE_POINT_LIMITS     = False;
-ENABLE_TORQUE_LIMITS            = True; # param
-ENABLE_FORCE_LIMITS             = False; # param
[…]

### 2018-02-19 code/data/config/conf_hyq_bridge.py

@@ -1,97 +1,0 @@
-from math import sqrt
-import numpy as np
-
-''' *********************** USER-PARAMETERS *********************** '''
-SOLVER_ID       = 0;    # classic TSID formulation
-SOLVER_TO_INTEGRATE         = [SOLVER_ID];
-DATA_FILE_NAME              = 'data';
-TEXT_FILE_NAME              = 'results.txt';
-SAVE_DATA                   = True;
-
-''' INITIAL STATE PARAMETERS '''
-SMOOTH_FILTER_WINDOW_LENGTH = 41;
-MAX_TEST_DURATION           = -1;
-dt                          = 1e-3;
-#INPUT_FILE_NAME             = '/home_local/dev/hpp/src/2016_ijrr_traj/bridge_ok/test88_compressed';
-INPUT_FILE_NAME             = '/home/adelpret/repos/20151101_contact_planner_steve/code/data/hyq/bridge_hyq_t_var_04f_andrea';
-#model_path                  = ["/home_local/dev/hpp/install/share"];
-model_path                  = ["/home/adelpret/repos/20151101_contact_planner_steve/code/data/hyq"];
-urdfFileName                = model_path[0] + "/hyq_description/urdf/hyq.urdf";
-sceneFileName               = model_path[0] + "/hpp-rbprm-corba/meshes/groundcrouch.stl";
-freeFlyer                   = True;
-
-''' CONTROLLER CONFIGURATION '''
-ENABLE_CAPTURE_POINT_LIMITS     = False;
-ENABLE_TORQUE_LIMITS            = True; # param
-ENABLE_FORCE_LIMITS             = True; # param
-ENABLE_JOINT_LIMITS             = True; # param, modulate with 4 next (position viability false, velocity ok)
-IMPOSE_POSITION_BOUNDS          = True;
-IMPOSE_VELOCITY_BOUNDS          = True;
-IMPOSE_VIABILITY_BOUNDS         = True;
-IMPOSE_ACCELERATION_BOUNDS      = True;
-JOINT_POS_PREVIEW               = 1.5; # preview window to convert joint pos limits into joint acc limits
-JOINT_VEL_PREVIEW               = 1;   # preview window to convert joint vel limits into joint acc limits
[…]

### 2018-02-19 code/data/config/conf_hyq_darpa.py

@@ -1,100 +1,0 @@
-from math import sqrt
-import numpy as np
-
-''' *********************** USER-PARAMETERS *********************** '''
-SOLVER_ID       = 0;    # classic TSID formulation
-SOLVER_TO_INTEGRATE         = [SOLVER_ID];
-DATA_FILE_NAME              = 'data';
-TEXT_FILE_NAME              = 'results.txt';
-SAVE_DATA                   = True;
-
-''' INITIAL STATE PARAMETERS '''
-SMOOTH_FILTER_WINDOW_LENGTH = 41;
-MAX_TEST_DURATION           = -1;
-dt                          = 1e-3;
-FOLDER = '/home/adelpret/repos/20151101_contact_planner_steve/code/hpp_serialization/'
-INPUT_FILE_NAME             = [FOLDER+'seq_1',FOLDER+'seq_2',FOLDER+'seq_3',FOLDER+'seq_4',FOLDER+'seq_5',FOLDER+'seq_6'];
-#INPUT_FILE_NAME             = [FOLDER+'seq_5', FOLDER+'seq_6'];
-RECORD_FILE_NAME             = 'seq_all';
-RECORD                      = False
-CAMERA_TRANSFORM            = False #[-1.9623876810073853, -5.876553058624268, 3.1662111282348633, 0.8184708952903748, 0.5468565821647644, -0.09847398102283478, -0.14613758027553558];
-model_path                  = ["/home/adelpret/repos/20151101_contact_planner_steve/code/data/hyq"];
-urdfFileName                = model_path[0] + "/hyq_description/urdf/hyq.urdf";
-sceneFileName               ="/home/adelpret/repos/20151101_contact_planner_steve/code/hpp_serialization/darpareduced.stl";
-freeFlyer                   = True;
-
-''' CONTROLLER CONFIGURATION '''
-ENABLE_CAPTURE_POINT_LIMITS     = False;
-ENABLE_TORQUE_LIMITS            = True; # param
-ENABLE_FORCE_LIMITS             = True; # param
-ENABLE_JOINT_LIMITS             = True; # param, modulate with 4 next (position viability false, velocity ok)
-IMPOSE_POSITION_BOUNDS          = True;
-IMPOSE_VELOCITY_BOUNDS          = True;
-IMPOSE_VIABILITY_BOUNDS         = True;
[…]

### 2018-02-19 code/data/config/conf_hyq_hole.py

@@ -1,95 +1,0 @@
-from math import sqrt
-import numpy as np
-
-''' *********************** USER-PARAMETERS *********************** '''
-SOLVER_ID       = 0;    # classic TSID formulation
-SOLVER_TO_INTEGRATE         = [SOLVER_ID];
-DATA_FILE_NAME              = 'data';
-TEXT_FILE_NAME              = 'results.txt';
-SAVE_DATA                   = True;
-
-''' INITIAL STATE PARAMETERS '''
-SMOOTH_FILTER_WINDOW_LENGTH = 41;
-MAX_TEST_DURATION           = -1;
-dt                          = 1e-3;
-INPUT_FILE_NAME             = '/home_local/dev/hpp/src/2016_ijrr_traj/ground_crouch_ok/test34_compressed';
-model_path                  = ["/home_local/dev/hpp/install/share"];
-urdfFileName                = model_path[0] + "/hyq_description/urdf/hyq.urdf";
-sceneFileName               = model_path[0] + "/hpp-rbprm-corba/meshes/groundcrouch.stl";
-freeFlyer                   = True;
-
-''' CONTROLLER CONFIGURATION '''
-ENABLE_CAPTURE_POINT_LIMITS     = False;
-ENABLE_TORQUE_LIMITS            = True; # param
-ENABLE_FORCE_LIMITS             = True; # param
-ENABLE_JOINT_LIMITS             = True; # param, modulate with 4 next (position viability false, velocity ok)
-IMPOSE_POSITION_BOUNDS          = True;
-IMPOSE_VELOCITY_BOUNDS          = True;
-IMPOSE_VIABILITY_BOUNDS         = True;
-IMPOSE_ACCELERATION_BOUNDS      = True;
-JOINT_POS_PREVIEW               = 1.5; # preview window to convert joint pos limits into joint acc limits
-JOINT_VEL_PREVIEW               = 1;   # preview window to convert joint vel limits into joint acc limits
-MAX_JOINT_ACC                   = 30 #50.0;
-MAX_MIN_JOINT_ACC               = 30 #50.0;
-USE_JOINT_VELOCITY_ESTIMATOR    = False;
-ACCOUNT_FOR_ROTOR_INERTIAS      = True;
-
-# CONTROLLER GAINS
[…]

### 2018-02-19 code/data/config/test1_planar.py

@@ -1,88 +1,0 @@
-from math import sqrt
-import numpy as np
-
-''' *********************** USER-PARAMETERS *********************** '''
-SOLVER_ID       = 0;    # classic TSID formulation
-SOLVER_TO_INTEGRATE         = [SOLVER_ID];
-DATA_FILE_NAME              = 'data';
-TEXT_FILE_NAME              = 'results.txt';
-SAVE_DATA                   = True;
-
-''' INITIAL STATE PARAMETERS '''
-SMOOTH_FILTER_WINDOW_LENGTH = 51;
-MAX_TEST_DURATION           = 5000;
-dt                          = 1e-3;
-INPUT_FILE_NAME             = '../data/hrp2/polaris_hrp2_t_var_1f_andrea';
-model_path                  = ["/home/adelpret/devel/sot_hydro/install/share"];
-urdfFileName                = model_path[0] + "/hrp2_14_description/urdf/hrp2_14_reduced.urdf";
-freeFlyer                   = True;
-
-''' CONTROLLER CONFIGURATION '''
-ENABLE_CAPTURE_POINT_LIMITS     = False;
-ENABLE_TORQUE_LIMITS            = True;
-ENABLE_FORCE_LIMITS             = False;
-ENABLE_JOINT_LIMITS             = True;
-IMPOSE_POSITION_BOUNDS          = True;
-IMPOSE_VELOCITY_BOUNDS          = True;
-IMPOSE_VIABILITY_BOUNDS         = True;
-IMPOSE_ACCELERATION_BOUNDS      = True;
-JOINT_POS_PREVIEW               = 1.5; # preview window to convert joint pos limits into joint acc limits
-JOINT_VEL_PREVIEW               = 1;   # preview window to convert joint vel limits into joint acc limits
-MAX_JOINT_ACC                   = 10.0;
-MAX_MIN_JOINT_ACC               = 10.0;
-USE_JOINT_VELOCITY_ESTIMATOR    = False;
-ACCOUNT_FOR_ROTOR_INERTIAS      = True;
-
-# CONTROLLER GAINS
-kp_posture  = 100; #1.0;   # proportional gain of postural task
-kd_posture  = 2*sqrt(kp_posture);
-kp_constr   = 100.0;   # constraint proportional feedback gain
-kd_constr   = 2*sqrt(kp_constr);   # constraint derivative feedback gain
[…]

### 2018-02-19 code/python/analyze_test_1_results.py

@@ -1,134 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Tue Oct  6 13:36:34 2015
-
-@author: adelpret
-"""
-import numpy as np
-import plot_utils
-from plot_utils import plotQuantityPerSolver
-#from constraint_violations import ConstraintViolationType
-import matplotlib.pyplot as plt
-import os
-import pickle
-#from viewer_utils import Viewer
-#import viewer_utils
-from time import sleep
-import math
-
-            
-np.set_printoptions(precision=2, suppress=True);
-DATA_PATH = '../results/test_1/20160919_170307_batch_1/';
-
-print 'Analyze data in folder', DATA_PATH;
-
-DATA_FILE_NAME              = 'data';
-REPEAT_TESTS_WHERE_CAPTURE_POINT_DISAGREES_WITH_STABILITY_CRITERION = True;
-SAVE_RESULTS                = False;
-PLAY_SALIENT_MOMENTS        = False;
-DUMP_IMAGE_FILES            = False;
-VERBOSE                     = False;
-N_DIR = 0;
-if(N_DIR<=0):
-    N_DIR = len(os.listdir(DATA_PATH));
-N_SOLVERS = 2;
-TEXT_FILE_NAME = 'stats.txt';
-dt = 0.002;
-MAX_TIME = 4.0-dt;
-SALIENT_MOMENT_DURATION = MAX_TIME;
-SALIENT_MOMENT_SLOWDOWN_FACTOR = 1.0;
-
-line_styles     =["k-", "b--", "r:", "c-", "g-"];
-solver_names    = ['ID','CP'];
-
-time_to_fall        = np.zeros((N_DIR,N_SOLVERS));
-i_to_fall           = np.zeros((N_DIR,N_SOLVERS), np.int);
-scores = np.zeros((N_DIR,N_SOLVERS), np.int);
-falls = np.zeros(N_SOLVERS, np.int);
-if(PLAY_SALIENT_MOMENTS):
-    viewer_utils.ENABLE_VIEWER = True;
-    viewer = Viewer('viewer');
-    viewer.CAMERA_FOLLOW_ROBOT = False;
-#    viewer.activateMouseCamera();
-
-i = 0;
-info_tail = '';
-capture_point_true = 0;
-multi_contact_true = 0;
-controller_true = np.zeros(N_SOLVERS, np.int);
-capture_point_multi_contact_disagree = 0;
[…]

### 2018-02-19 code/python/hrp2014_wrapper.py

@@ -1,68 +1,0 @@
-from pinocchio.robot_wrapper import RobotWrapper
-import numpy as np
-import pinocchio as se3
-
-class Hrp2014Wrapper(RobotWrapper):
-
-    def __init__(self,*args):
-        filename = args[0]
-        if len(args) >= 2:
-          mesh_dir = args[1]
-          RobotWrapper.__init__(self,filename,mesh_dir,se3.JointModelFreeFlyer())
-        else:
-          RobotWrapper.__init__(self,filename,se3.JointModelFreeFlyer())
-#        RobotWrapper.__init__(self,filename, root_joint='ff')
-        self.q0 = np.matrix( [
-        0.0, 0.0, 0.648702, 0.0, 0.0 , 0.0, 1.0,                             # Free flyer 0-6
-        0.0, 0.0, 0.0, 0.0,                                                  # CHEST HEAD 7-10
-        0.261799388,  0.174532925, 0.0, -0.523598776, 0.0, 0.0, 0.174532925, # LARM       11-17
-        0.261799388, -0.174532925, 0.0, -0.523598776, 0.0, 0.0, 0.174532925, # RARM       18-24
-        0.0, 0.0, -0.453785606, 0.872664626, -0.41887902, 0.0,               # LLEG       25-30
-        0.0, 0.0, -0.453785606, 0.872664626, -0.41887902, 0.0,               # RLEG       31-36
-        ] ).T
-
-        self.ff = range(7)
-        self.chest = range(7,9)
-        self.head = range(9,11)
-        self.l_arm = range(11,18)
-        self.r_arm = range(18,25)
-        self.l_leg = range(25,31)
-        self.r_leg = range(31,37)
-
-        self.opCorrespondances = { "lh": "LARM_JOINT5",
-                                   "rh": "RARM_JOINT5",
-                                   "lf": "LLEG_JOINT5",
-                                   "rf": "RLEG_JOINT5",
-                                   }
-
-        for op,name in self.opCorrespondances.items():
-            idx = self.__dict__[op] = self.index(name)
-            #self.__dict__['_M'+op] = types.MethodType(lambda s,q: s.position(q,idx),self)
-
-    # --- SHORTCUTS ---
-    def Mrh(self,q):
-        return self.position(q,self.rh)
-    def Jrh(self,q):
-        return self.jacobian(q,self.rh)
-    def wJrh(self,q):
[…]

### 2018-02-19 code/python/track_locomotion_traj.py

@@ -1,552 +1,0 @@
-'''
-To do.
-'''
-import sys
-import os
-
-cwd = os.getcwd()[:-7]; # current path without "python"
-if cwd+'/data/config' not in sys.path:
-    sys.path += [cwd+'/data/config',];
-
-#import test1_planar as conf
-#~ import conf_hyq_hole as conf
-#import conf_hyq_bridge as conf
-import conf_hyq_darpa as conf
-#import conf_hrp2_polaris as conf
-
-from pinocchio_inv_dyn.standard_qp_solver import StandardQpSolver
-from pinocchio_inv_dyn.simulator import Simulator
-import pinocchio_inv_dyn.viewer_utils as viewer_utils
-from pinocchio_inv_dyn.inv_dyn_formulation_util import InvDynFormulation
-from pinocchio_inv_dyn.constraint_violations import ConstraintViolationType
-import pinocchio_inv_dyn.plot_utils as plot_utils
-from pinocchio_inv_dyn.geom_utils import plot_inequalities
-from pinocchio_inv_dyn.tasks import SE3Task, CoMTask, JointPostureTask
-from pinocchio_inv_dyn.trajectories import ConstantSE3Trajectory, ConstantNdTrajectory, SmoothedNdTrajectory, SmoothedSE3Trajectory
-from pinocchio_inv_dyn.multi_contact.utils import compute_com_acceleration_polytope, compute_GIWC
-
-from pinocchio.utils import XYZQUATToSe3
-import cProfile
-import pickle
-import numpy as np
-import numpy.matlib
-from numpy.linalg import norm
-import matplotlib.pyplot as plt
-from datetime import datetime
-#import matplotlib.pyplot as plt
-from time import sleep
-from math import sqrt
-
-EPS = 1e-4;
-
-def createListOfMatrices(listSize, matrixSize):
-    l = listSize*[None,];
-    for i in range(listSize):
-        l[i] = np.matlib.zeros(matrixSize);
-    return l;
-
-def createListOfLists(size1, size2):
-    l = size1*[None,];
-    for i in range(size1):
-        l[i] = size2*[None,];
-    return l;
-    
-
-def createInvDynFormUtil(q, v):
-    invDynForm = InvDynFormulation('hrp2_inv_dyn'+datetime.now().strftime('%m%d_%H%M%S'), 
-                                   q, v, conf.dt, conf.model_path, conf.urdfFileName, conf.freeFlyer);
[…]

### 2018-02-19 merge_tex.py
more cleaning

@@ -1,34 +1,0 @@
-#usage python merge_tex.py inFile.tex outFile.tex
-
-import re,sys
-
-inFile = sys.argv[1]
-outFile = sys.argv[2]
-
-infile = open(inFile, "r")
-filew = open(outFile, "w")
-
-
-def paste_text(filepath):
-    subfile = open(filepath,"r")
-    mergefiles(subfile)
-    subfile.close()
-
-def mergefiles(file):
-    for line in file.readlines():
-        #find input lines that are not commented
-        if(line.find("\\input") > -1  and (line.find("%") < 0 or line.find("%") > 1)):
-            m = re.search("\input{(.+?)}", line)
-            filepath = m.group(1)
-            #add tex extension is missing
-            if (filepath.find(".tex") < 0):
-                filepath+= ".tex"
-            paste_text(filepath)
-        else:
-            filew.write(line)
-
-
-mergefiles(infile)
-
-infile.close()
-filew.close()

### 2018-02-19 final.tex
16 pages with bio

@@
 \usepackage[Symbol]{upgreek}
+\usepackage[utf8]{inputenc}
 
@@
 few seconds, with high success rates, we generate complex contact plans for various scenarios and two robots, HRP-2 and HyQ.
-These plans are validated either in dynamic simulations, or on the real HRP-2 robot.
+These plans are validated in dynamic simulations or on the real HRP-2 robot.
 
@@
 planners
-for two reasons. (i) First, each sub-manifold of the         
+for two reasons. (i) Each sub-manifold of the         
 foliation
@@
 operation).
-(ii) Second, the foliated topology       
+(ii) The foliated topology       
 turns
@@
 
-Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences, biomechanics, and virtual character animation.
+Additionally to robotics, acyclic motion planning is also a problem of interest in biomechanics and virtual character animation.
 Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}.
@@
 %~ because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. 
-Compared to previous approaches, our solution has two main novelties:
+Our solution presents two main novelties:
 
@@
 	\end{overpic}
-\caption{The reachability condition is verified by the right configuration: the trunk (red) is free of collisions while the limbs reachable workspace (green) intersect the environment.}
+\caption{The reachability condition is verified if the trunk (red) is collision-free while the limbs reachable workspace (green) intersect the environment.}
 		   \label{fig:reach_cond}
@@
 The method \textit{Length} returns the length of a list~; \textit{LastElement} returns the last element
-of a list~; \textit{Element} returns the element contained by a list at a given index~; and \textit{PushBack} inserts an element at the end of a list.
+of a list~; \textit{Element} returns the element of a list at a given index~; and \textit{PushBack} inserts an element at the end of a list.
 %~ Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.
@@
 It generates a contact configuration in static equilibrium, and assigns the corresponding configuration of the limb $k$ to the state configuration.
-If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.
+If it fails, $k$ remains unchanged if collision free, else it is assigned a random collision free configuration.
 
@@
 According to Coulomb's law, the non-slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
-As classically done, we linearize the friction cone in a conservative fashion with a pyramid included in it, described by four generating rays of unit length. We choose for instance:
+Classically, we linearize the friction cone in a conservative fashion with a pyramid included in it, described by four generating rays of unit length. We choose for instance:
 \begin{equation*}
@@
 
-From the Newton-Euler equations, to be in static equilibrium the contact forces have to compensate the gravitational forces:
+From the Newton-Euler equations, to be in static equilibrium the contact forces have to compensate the gravity:
 
@@
 
-\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $20$.
+\noindent\textbf{Heuristics:} $h_w$. The robustness threshold $b_0$ is set to $20$.
 
@@
 
-\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $10$.
+\noindent\textbf{Heuristics:} $h_w$. The robustness threshold $b_0$ is set to $10$.
 
@@
 In this long scene, HRP-2 plans a path through several obstacles. The scene is edited during the execution of the motion: a stair is added,
-some stepping stones are removed, and part of the final staircase is deleted. All these modifications require re-planning.
+stepping stones are removed, as for parts of the final staircase. All these modifications require re-planning.
 
@@
 the generation of the plan, the resulting interpolation turned out to be harder.
-For this reason, we prefer to use our manipulability-based heuristic for the legs of the robot, but we still
+For this reason, we use our manipulability-based heuristic for the legs, but still
 use EFORT for the arms, which results in fewer contact repositionings.
@@
 for the car egress scenario to handle the geometry of the car.
-For future work, we would like to automatically adapt the size of the discretization step to the complexity of the environment.
+For future work, we would like to automatically adapt the size of the discretization step.
 
@@
 it is also applied every time a new candidate needs to be evaluated. In particular for the car egress scenario,
-the kinematic constraints are very demanding to avoid collisions.
+the collision avoidance constraints are demanding.
 
@@
 
-\subsection{Step 2: computing the torso scaling workspace $W^0$ of the robot}
+\subsection{Step 2: computing the torso scaling workspace $W^0$}
 To define the volume $W^0$ of HRP-2, we proceed in an empirical manner.
@@
 
-\begin{figure}[!tbp]
-  \centering
-	\begin{overpic}[width=1\linewidth]{figures/ellipsoid}
-		\put (8.5,1.2) {\small{Velocity ellipsoid}}
-		\put (62.5,1.2) {\small{Force ellipsoid} \tiny{(scale 0.5)}}
-	\end{overpic}
-  \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 DoFs and 2 segments.
-Only the horizontal and vertical speeds are shown (not the rotation speeds).}
- %~ since it would require being able to draw in four dimensions.}
-		   \label{sec:efort_ellipsoid}
-\end{figure}
-
 We consider: a limb configuration $\mathbf{q}^k$; its end effector position $\mathbf{p}^k$; its Jacobian matrix
 $\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
-Yoshikawa~\cite{Yoshikawa1984} defines the velocity(~\ref{eq:vel}) and force~(\ref{eq:for}) ellipsoids:
+Yoshikawa~\cite{Yoshikawa1984} defines the velocity (\ref{eq:vel}) and force~(\ref{eq:for}) ellipsoids:
 
@@
 be reached under the constraint $||\dot{\mathbf{q}}||^2 \leq 1$ for the current configuration.
-The longer the axis of the ellipsoid is, the more important the velocity (resp. force) of the end-effector the direction of the axis can be (Figure~\ref{sec:efort_ellipsoid}).
+The longer the axis of the ellipsoid is, the more important the velocity (resp. force) of the end-effector the direction of the axis can be.
+%~ (Figure~\ref{sec:efort_ellipsoid}).
 
 \subsubsection{Manipulability-based heuristics}
-From these definitions, we can derive two useful heuristics, that all account for the environment and the task being performed.
+From these definitions, we derive two heuristics that account for the environment and the task being performed.
 The first one, EFORT, was introduced by \cite{Tonneau2014}; the second one derives the manipulability measure proposed by Yoshikawa~\cite{Yoshikawa1984}.
@@
 With EFORT, we define the efficiency of a configuration as the ability of a limb to exert a force in a given direction.
-We thus consider the force ellipsoid as a basis for our heuristic.
 In a given direction $\uprho$, the length of the ellipsoid is given by the force-transmission ratio \citep{1087795}:
@@
 
-In our problem, to compare candidate configurations, we include the quality of the contact surface, and choose $\uprho$ as the direction
-opposite to the local motion (given by the difference between two consecutive root positions):
+To compare candidate configurations, we include the quality of the contact surface, and choose $\uprho$ as the direction opposite to the local motion (given by the difference between two consecutive root positions):
 
@@
 
-%~ If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
-%~ \begin{equation}
-%~ h_{vel}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
-%~ \end{equation}
-
-%~ $h_{\textrm{\it EFORT}}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
-$h_{\textrm{\it EFORT}}$ will favor contacts that allow large efforts.
-$EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
-In other less demanding cases, manipulability can also be considered to avoid singularities.
-To do so, we can consider the manipulability measure $h_{w}$, also given by Yoshikawa~\cite{Yoshikawa1984}:
+$h_{\textrm{\it EFORT}}$ will favor contacts that allow large efforts. Alternatively, the manipulability measure can be considered~\cite{Yoshikawa1984}:
 
@@
 \end{equation}
-$h_{w}$ measures the ``distance'' of a given configuration to singularity. When $h_{w}$ is equal to 0, the configuration is singular;
-the greater $h_{w}$ is, the further away the configuration is from singularity.
+%~ $h_{w}$ measures the ``distance'' of a given configuration to singularity. When $h_{w}$ is equal to 0, the configuration is singular;
+%~ the greater $h_{w}$ is, the further away the configuration is from singularity.
 
@@
 This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
-The center-of-mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.
+The obtained COM trajectory is then turned into a collision-free whole-body trajectory.
 
@@
 this cone determines all the admissible wrenches $\mathbf{w}$ that can be generated by contact forces inside their friction cones.
-The face form of $\mathcal{K}$ can be computed using the double description method~\citep{Fukuda1996}, resulting in the following linear inequalities:
+The face form of $\mathcal{K}$ can be computed using the double description method~\citep{Fukuda1996}:
 
@@
 
-%~ \begin{IEEEbiographynophoto}{Steve Tonneau}
-%~ TODO
-%~ \end{IEEEbiographynophoto}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/steve}}]{Steve Tonneau} is a post-doctorate in the Gepetto team at LAAS-CNRS.
+He received the joint M.Eng.
+and M.Sc.
+degrees from Insa Rennes in 2008.
+After 4 years in the industry he defended his Phd in 2015 at the INRIA/IRISA Mimetic research team.
+His research focuses on motion planning based for legged avatars in arbitrary situations.
+Applications include computer graphics animation and legged robotics.
+\end{IEEEbiography}
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/andrea}}]{Andrea Del Prete} received his degree in Computer
+Engineering (with honors) from the 2nd faculty of
+the University of Bologna (Italy) in 2009. In March
+2013 he got his PhD from the Cognitive Humanoids
+laboratory, Italian Institute of Technology, Genova
+(Italy). From 2014 to 2017 he had been a PostDoc
+at LAAS-CNRS in Toulouse (France), working on
+optimization-based control of the humanoid robot
+HRP-2. Since 2018 he is a senior researcher at
+the Max Planck Institute for Intelligent Systems, in
+Tuebingen (Germany).
+\end{IEEEbiography}
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/julien}}]{Julien Pettré}
+is a computer scientist.
+He is junior researcher at Inria, the French National Institute for Research in Computer Science and Control, in the Lagadic team.
+He received PhD from the University of Toulouse III in 2003, and Habilitation from the University of Rennes I in 2015.
+From 2004 to 2006, he was postdoctoral fellow at EPFL in Switzerland.
+He joined Inria in 2006.
+Julien Pettré is currently coordinator of the European H2020 Crowdbot project (2018-21), dedicated to the design of robot navigation techniques for crowded environments.
+He previously coordinated the national ANR JCJC Percolation project (2013-17) dedicated to the design of new microscopic crowd simulation algorithms, as well as the national ANR CONTINT Chrome project, dedicated to efficient and designer-friendly techniques for crowd animation.
+His research interests are crowd simulation, computer animation, virtual reality, robot navigation and motion planning. 
+\end{IEEEbiography}
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/chonhyon}}]{Chonhyon Park}
+received the B.S. degree and the M.S. degree in computer science and engineering from the Seoul National University, Seoul, South Korea 2005 and 2007, respectively.
+
+He received the Ph.D. degree in computer science at University of North Carolina at Chapel Hill in 2016.
+
+His current research interests include motion and path planning,  navigation of virtual characters, and many-core computing.
+\end{IEEEbiography}
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/dinesh}}]{Dinesh Manocha}
+is currently the Phi Delta
+Theta/Mason Distinguished Professor of Computer
+Science at the University of North Carolina at
+Chapel Hill. He received his Ph.D. in Computer
+Science at the University of California at Berkeley
+1992. Along with his students, Manocha has
+also received 15 best paper awards at the leading
+conferences. He has published more than 480
+papers and some of the software systems related
+to collision detection, GPU-based algorithms and
+geometric computing developed by his group have
+been downloaded by more than 200,000 users and are widely used in the
+industry. He has supervised 33 Ph.D. dissertations and is a fellow of ACM, AAAI,
+AAAS, and IEEE. He received Distinguished Alumni Award from Indian
+Institute of Technology, Delhi.
+\end{IEEEbiography}
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/nicolas}}]{Nicolas Mansard}
+received the joint M.Eng. and
+M.Sc. degrees in 2003, robotics and image process-
+ing from Nationale Supérieure d'Informatique et de
+Mathématiques Appliquées de Grenoble,
+France, and University Joseph Fourier, Grenoble,
+France, respectively, and the Ph.D. degree in robotics
+for his work with Lagadic Group, Institut National
+de Recherche en Informatique et en Automatique,
+Rennes, France, in 2006. He spent one year with
+Stanford University, Stanford, CA, with O. Khatib
+and one year with JRL-Japan with A. Kheddar. He
+is currently with the GEPETTO Group, LAAS- CNRS, Toulouse, France.
+He was an Invited Researcher with Emo Todorovs Group, University of
+Washington, in 2013. His research interest includes the generation of dynamic
+sensor-based robot movements. Dr. N. Mansard received the CNRS Bronze
+medal in 2015.
+\end{IEEEbiography}
 

### 2018-02-19 final.tex
removed all commentes

@@
-%% bare_jrnl.tex
-%% V1.4b
-%% 2015/08/26
-%% by Michael Shell
-%% see http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
-%% journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/pkg/ieeetran
-%% and
-%% http://www.ieee.org/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall the IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. The IEEE's font choices and paper sizes can   ***
-% *** trigger bugs that do not appear when using other class files.       ***                          ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
 \documentclass[journal]{IEEEtran}
-%
-
-%~ \setcounter{secnumdepth}{3}
+
 \usepackage{moreverb,url}
-\usepackage{makeidx}         % allows index generation
+\usepackage{makeidx}         %
 \usepackage[percent]{overpic}
@@
 \newcommand{\fratop}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
-\newcommand{\mx}[1]{\mathbf{\bm{#1}}} 				% Matrix symbol
-\newcommand{\vc}[1]{\mathbf{\bm{#1}}} 					% Vector symbol
-\newcommand{\degree}{\ensuremath{^\circ}}				% define the degree symbol
-\newcommand{\pder}[2]{\frac{\partial#1}{\partial#2}}		% partial derivative
-\newcommand{\ppder}[2]{\frac{\partial^2 #1}{\partial#2^2}}		% second partial derivative
-\newcommand{\refframe}[1]{\mbox{\textless#1\textgreater}}	% to denote a reference frame
-\DeclareMathOperator*{\argmin}{\arg\!\min}				% argmin
-\DeclareMathOperator*{\argmax}{\arg\!\max}				% argmax
-\DeclareMathOperator*{\st}{subject\,to}					% subject to
+\newcommand{\mx}[1]{\mathbf{\bm{#1}}} 				%
+\newcommand{\vc}[1]{\mathbf{\bm{#1}}} 					%
+\newcommand{\degree}{\ensuremath{^\circ}}				%
+\newcommand{\pder}[2]{\frac{\partial#1}{\partial#2}}		%
+\newcommand{\ppder}[2]{\frac{\partial^2 #1}{\partial#2^2}}		%
+\newcommand{\refframe}[1]{\mbox{\textless#1\textgreater}}	%
+\DeclareMathOperator*{\argmin}{\arg\!\min}				%
+\DeclareMathOperator*{\argmax}{\arg\!\max}				%
+\DeclareMathOperator*{\st}{subject\,to}					%
 \DeclareMathOperator*{\minimize}{minimize}				%
@@
 \DeclareMathOperator*{\find}{find}						%
-\DeclareMathOperator*{\dif}{\mathrm{d}}					% d
-\DeclareMathOperator*{\half}{\frac{1}{2}}					% one half
-\newcommand{\equilibriumfeasibility}[0]{\glslink{equilibrium feasible}{\textit{equilibrium feasibility}}}	% matrix
-\newcommand{\contactreachability}{\textit{contact reachability}}	% matrix
-\newcommand{\mat}[1]{\ensuremath{\begin{bmatrix}#1\end{bmatrix}}}	% matrix
-\newcommand{\rank}[1]{\text{rank}(#1)}							% rank
-\newcommand{\diag}[1]{\text{diag}(#1)}							% diag
+\DeclareMathOperator*{\dif}{\mathrm{d}}					%
+\DeclareMathOperator*{\half}{\frac{1}{2}}					%
+\newcommand{\equilibriumfeasibility}[0]{\glslink{equilibrium feasible}{\textit{equilibrium feasibility}}}	%
+\newcommand{\contactreachability}{\textit{contact reachability}}	%
+\newcommand{\mat}[1]{\ensuremath{\begin{bmatrix}#1\end{bmatrix}}}	%
+\newcommand{\rank}[1]{\text{rank}(#1)}							%
+\newcommand{\diag}[1]{\text{diag}(#1)}							%
 \newcommand{\x}{\ensuremath{\times}}
-\newcommand{\spac}{\ensuremath{\quad}}						% alias for space in math environment
-\newcommand{\dx}[0]{\ensuremath{\delta x}}					% dx
-\newcommand{\du}[0]{\ensuremath{\delta u}}					% du
-\newcommand{\DX}[0]{\ensuremath{\Delta X}}						% DX
-\newcommand{\DU}[0]{\ensuremath{\Delta U}}						% DU
-\newcommand{\T}[0]{\ensuremath{\top}}							% transpose symbol
-\newcommand{\pinv}[0]{\ensuremath{\dagger}}					% pseudoinverse symbol
-\newcommand{\Rv}[1]{\ensuremath{\mathbb{R}^{#1}}}				% set of real-valued vectors
-\newcommand{\R}[2]{\ensuremath{\mathbb{R}^{#1\times #2}}}		% set of real-valued matrices
-\newcommand{\Spd}[1]{\ensuremath{\mathbb{S}_+^{#1}}}			% set of symmetric positive-definite matrices
-\newcommand{\card}[1]{\ensuremath{\left\vert{#1}\right\vert}}			% cardinality of a set
-\DeclareMathOperator{\Tr}{Tr}							% trace
-\newcommand{\Expect}{{\rm I\kern-.3em E}}				% expectation
-\newcommand{\Normal}{\mathcal{N}}					% normal distribution
-\newcommand{\Prob}[1]{\text{P}(#1)}						% probability
+\newcommand{\spac}{\ensuremath{\quad}}						%
+\newcommand{\dx}[0]{\ensuremath{\delta x}}					%
+\newcommand{\du}[0]{\ensuremath{\delta u}}					%
+\newcommand{\DX}[0]{\ensuremath{\Delta X}}						%
+\newcommand{\DU}[0]{\ensuremath{\Delta U}}						%
+\newcommand{\T}[0]{\ensuremath{\top}}							%
+\newcommand{\pinv}[0]{\ensuremath{\dagger}}					%
+\newcommand{\Rv}[1]{\ensuremath{\mathbb{R}^{#1}}}				%
+\newcommand{\R}[2]{\ensuremath{\mathbb{R}^{#1\times #2}}}		%
+\newcommand{\Spd}[1]{\ensuremath{\mathbb{S}_+^{#1}}}			%
+\newcommand{\card}[1]{\ensuremath{\left\vert{#1}\right\vert}}			%
+\DeclareMathOperator{\Tr}{Tr}							%
+\newcommand{\Expect}{{\rm I\kern-.3em E}}				%
+\newcommand{\Normal}{\mathcal{N}}					%
+\newcommand{\Prob}[1]{\text{P}(#1)}						%
 
@@
 
-%~ \usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red,draft]{hyperref}
 \usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
-%~ \usepackage{glossaries}
-
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/pkg/ifpdf
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of the IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off
-% such as if a citation ever needs to be enclosed in parenthesis.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 5.0 (2009-03-20) and later if using hyperref.sty.
-% The latest version can be obtained at:
-% http://www.ctan.org/pkg/cite
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
+
 \ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
 \else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
 \fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation
-% can be obtained at: 
-% http://www.ctan.org/pkg/graphicx
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found at:
-% http://www.ctan.org/pkg/epslatex
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics.
-%
-% Note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/amsmath
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/pkg/algorithms
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/pkg/algorithmicx
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/array
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
-% for subfigure.sty, the latter of which is no longer maintained and is
-% incompatible with some LaTeX packages including fixltx2e. However,
-% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
-% which will override IEEEtran.cls' handling of captions and this will result
-% in non-IEEE style figure/table captions. To prevent this problem, be sure
-% and invoke subfig.sty's "caption=false" package option (available since
-% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
-% handling of captions.
-% Note that the Computer Society format requires a larger sans serif font
-% than the serif footnote size font used in traditional IEEE formatting
-% and thus the need to invoke different subfig.sty package options depending
-% on whether compsoc mode has been enabled.
-%
-% The latest version and documentation of subfig.sty can be obtained at:
-% http://www.ctan.org/pkg/subfig
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure.
-% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
-% corrections already built into the system in which case a warning will
-% be issued if an attempt is made to load fixltx2e.sty as it is no longer
-% needed.
-% The latest version and documentation can be found at:
-% http://www.ctan.org/pkg/fixltx2e
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/stfloats
-% Do not use the stfloats baselinefloat ability as the IEEE does not allow
-% \baselineskip to stretch. Authors submitting work to the IEEE should note
-% that the IEEE rarely uses double column equations and that authors should try
-% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
-% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
-% such ways.
-% Do not attempt to use stfloats with fixltx2e as they are incompatible.
-% Instead, use Morten Hogholm'a dblfloatfix which combines the features
-% of both fixltx2e and stfloats:
-%
-% \usepackage{dblfloatfix}
-% The latest version can be found at:
-% http://www.ctan.org/pkg/dblfloatfix
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
-% Axel Sommerfeldt. This package may be useful when used in conjunction with 
-% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
-% submissions have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.4.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/pkg/endfloat
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/url
-% Basically, \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
+
 \hyphenation{op-tical net-works semi-conduc-tor}
@@
 \begin{document}
-%
-% paper title
-% Titles are generally capitalized except for words such as a, an, and, as,
-% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
-% not capitalized unless they are the first or last word of the title.
-% Linebreaks \\ can be used within to get better formatting as desired.
-% Do not put math or special symbols in the title.
 \title{An efficient acyclic contact planner for multiped robots}
-%
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
 
@@
         ~Dinesh~Manocha,~\IEEEmembership{Member,~IEEE,}
-        ~and~Nicolas~Mansard,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
-\thanks{S. Tonneau, A. Del Prete and N. Mansard are with LAAS-CNRS / Universit\'e de Toulouse, France e-mail: (pro@stevetonneau.fr)}% <-this % stops a space
-\thanks{J. Pettr\'e is with Inria, Rennes, France}% <-this % stops a space
+        ~and~Nicolas~Mansard,~\IEEEmembership{Member,~IEEE,}%
+\thanks{S. Tonneau, A. Del Prete and N. Mansard are with LAAS-CNRS / Universit\'e de Toulouse, France e-mail: (pro@stevetonneau.fr)}%
+\thanks{J. Pettr\'e is with Inria, Rennes, France}%
 \thanks{C. Park and D. Manocha are with UNC, Chapel Hill, USA}}
 
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
-
-% The paper headers
-%~ \markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
-%~ {Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
 \maketitle
 
-% As a general rule, do not put math, special symbols or citations
-% in the abstract or keywords.
 \begin{abstract}
-% !TEX root =  ../main.tex
 We present a contact planner for complex legged locomotion tasks: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car.
@@
 configurations projected in contact with the environment.
-%~ computation times, due to the difficulty
-%~ of generating contact postures in a high-dimensional solution space that is impossible to sample.
 
@@
 a geometric approximation of the contact manifold, which is of low dimension, presents a Cartesian topology, and can be efficiently sampled and explored.
-%~ Informally, the condition verifies that the root configuration of a robot is close, but not too close to obstacles: close to allow contact creation, not too close to avoid collision.
-Thanks to this condition we decompose
 The hard contact planning problem can then be decomposed into two
-%~ sub-problems: first, planning a guide path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm.
 sub-problems: first, we plan a path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, we generate a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm.
@@
 
-%~ Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically: in a
-%~ few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ, and a dexterous hand.
-These plans are then validated either in dynamic simulations, or on the real HRP-2 robot, thus demonstrating
-%~ the first interactive multi-contact planner.
 \end{abstract}
 
-% Note that keywords are not normally used for peerreview papers.
 \begin{IEEEkeywords}
@@
 
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
 \IEEEpeerreviewmaketitle
 
-\newcommand{\citep}[1]{\cite{#1}}	% to denote a reference frame
-\newcommand{\citeauthor}[1]{\cite{#1}}	% to denote a reference frame
+\newcommand{\citep}[1]{\cite{#1}}	%
+\newcommand{\citeauthor}[1]{\cite{#1}}	%
 
 \section{Introduction}
-% !TEX root =  ../main_tro.tex
 \newcommand{\Pa}{$\mathcal{P}_1$ }
@@
 
-%~ Planning legged robot locomotion remains an open issue.
 \IEEEPARstart{L}{egged} robots move by sequentially creating contacts with the environment.
 After years of research, such robots can autonomously walk on flat ground, but struggle to navigate more complex environments.
-%~ where
 Deciding
@@
 handrail.
-%~ To climb stairs using a handrail, a humanoid robot must plan the contacts to create with its feet and hands. 
-%~ Computing such motion is an open motion planning problem called acyclic contact planning.
 
@@
 Once the contact plan is known, efficient approaches exist to generate a dynamically feasible motion~\citep{Carpentier2016}.
- %~ However in general
-%~ the contact plan is manually designed by a time consuming trial-and-error procedure.
-Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in cluttered, human-centered environments, which comprise constraining obstacles (stairs...).
 In the specific (and simple) case of gaited biped locomotion on flat ground, 
-%~ efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location. Furthermore, 
 choosing the effector with which to create a contact is trivial because walking follows a cyclic pattern (the left foot always follows the right foot).
@@
 Furthermore, a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles; we say that the contact manifold is foliated~\citep{simeon-manipulation-04}.
-%~ Lastly, the contact manifold has a null measure and is thus impossible to sample. This restriction prevents the use of efficient sampling-based planners.
 The
@@
 
-%~ Obtaining reasonable computation times is critical for a contact planner to be of any practical use.
  Our work aims at breaking the complexity of the acyclic contact planning problem.
@@
 This decoupling presents pros and cons discussed in previous related literature, summarized in the following.
-%~ whole-body trajectory and the contact location and timings, we decouple path and contact planning, as suggested by \citeauthor{Bouyarmane2009}.
-%~ We first plan a path for the geometric root of the robot in a low dimensional space. We formulate this space as a computationally-efficient approximation
-%~ of the \textit{equilibrium feasibile} space, that is the space of root configurations for which there exists a joint configuration such that the robot is in static equilibrium.
-%~ We then use a deterministic algorithm
-%~ This paper presents a geometrical approximation of this condition, and a concrete implementation of such a decoupled planner.
-%~ This approximation results from a trade-off between a necessary and a sufficient condition for \textit{equilibrium feasibility}, allowing us to find solutions extremely rapidly, while preserving a high success rate in the demonstrated scenarios.
-
-%~ In the remainder of this introduction, we discuss further the current state of the art. This allows us to situate more precisely our contributions.
 
@@
   \begin{overpic}[width=0.8\linewidth]{figures/workflow}
-    %~ \put (6.4,1.8) {\normalsize{Request}} 
     \put (30,10.8) {\large{\color{white}\Pa} }
-    %~ \put (31.6,2) {\scriptsize{\color{white}RB-RRT}} 
     \put (66.4,10.8) {\large{\color{white}\Pb} }
-    %~ \put (72.5,2) {\tiny{\color{white}Contact}} 
 	\put (15,1) {a} 
@@
 A third nontrivial problem, $\mathcal{P}_3$,
-%~ not addressed in this work, 
 then consists in interpolating a complete motion between two postures of the contact sequence.
@@
 an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide.
-A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning},
+A solution could then be found, %
 but at the cost of prohibitive computation times (up to several hours) for constraining scenarios.
@@
 They thus cannot be generalized to multi contact locomotion.
-%For instance, the choice of the next effector in contact is often deterministic, the contacts are either coplanar or punctual, or the generation of the end-effector trajectory only works for ``ground'' obstacles, as it assumes that increasing the $z$ position of an effector drives it further away from obstacles~\cite{Kolter:2009:TTV:1703775.1703833}.
 However, some contributions on specific parts of the problem could be applied directly in our case. 
@@
 Although we did not try to include such formulation in this paper, it would be straightforward to integrate such heuristic in our planner.
-
-%~ For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
-%~ used black-box physics simulators to perform Model Predictive Control for a humanoid character, and managed to obtain dynamically-consistent motions
-%~ at \gls{interactive} frame rates (a method is interactive when the computation time for planning is less than the
-%~ execution time). While this new approach provides an exciting direction of research, currently the resulting motions
- %~ do not seem applicable to real robots.
 
@@
 
-%~ because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. 
 Our solution presents two main novelties:
@@
 These aspects of the framework are presented in details in the
-paper, but are not novel \textit{per se}. %Describing them here is required to understand our experimental contribution.
+paper, but are not novel \textit{per se}. %
 The novelty lies in the complete
@@
 
-%~ Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
-%~ thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.
-%~ These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}.
-The novelty lies in the validation of the method with real robot models, for which the planning is harder with respect to the avatars used in [20], because of stricter kinematic constraints
-
-%~ As such, the motion planning algorithm presented here is the same as the one presented in the
-%~ previous work.
-Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper, with the exception of Section~\ref{app:contact}, where we additionally provide the pseudo-code of the algorithm absent from the original paper.
-We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section~\ref{sec:Equil}.
-%~ Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
-%~ all the technical information required to use the planner.
-%~ 
-%~ To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team.
-We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work~\cite{Carpentier2016}.
-The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework.
-Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
-%~ We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand the experimental contribution of this paper.
-
-%~ Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
-%~ Indeed, the novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
-%~ As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
-%~ thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method is remains efficient and qualitatively relevant in this context.
-% !TEX root =  ../main.tex
 \section{Overview}
@@
 our framework invalidates the computed guide path, and restarts the planning from (b).
-
-%
 
@@
 An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' to the environment: close, because a contact surface must be partially included in the reachable workspace of the robot to allow contact creation; not too close, because the robot must avoid collision.
-%~ \deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
 More precisely, a root configuration is \textit{contact reachable} if the root scaled by a user-defined factor $s \geq 1$ is not in collision (Figure~\ref{fig:reach_cond} - red shape), while the reachable workspace is in collision (Figure~\ref{fig:reach_cond} - green shapes).
-%~ Verifying the \textit{reachability condition} is very efficient in practice.
 To plan a root path, we then use an RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
-%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).
 
@@
 \end{figure}
-
-%
-%~ Ideally we would like to sample root configurations that are \gls{equilibrium feasible}.
-A root configuration is \gls{equilibrium feasible} if and only if there exists at least one joint configuration such that:
-%~ \begin{description}
-%~ \item[C1]: the robot is not in collision with the environment, 
-%~ \item[C2]: the robot is in static equilibrium. 
-%~ \end{description}
-%~ Verifying C2 is computationally expensive because it requires the projection of the whole-body configuration into a contact posture, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}.
-%~ For this reason, we are interested in finding an approximation of the \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} that is computationally efficient.
-%~ A requirement for C2 is the condition:
-%~ \begin{description}
-%~ \item[C3]: at least one end-effector is in contact with the environment.
-%~ \end{description}
-%~ C1 and C3 define \glslink{contact reachable}{\textit{contact reachability}}, which is easier to verify than \glslink{equilibrium feasible}{\textit{equilibrium feasibility}} because it does not require the computation of contact forces.
-%~ We make the assumption that for the \textit{\gls{cluttered} contact planning problem}, most of the times \gls{contact reachable} root configurations are also \gls{equilibrium feasible}.
-%~ For this reason we neglect C2 during the resolution of $\mathcal{P}_1$, and we consider it only when solving \Pb.
-%~ However, even verifying \glslink{contact reachable}{\textit{contact reachability}} is computationally too expensive because it requires the projection of the whole-body configuration into a contact posture.
-%~ 
-%~ An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
-%~ We approximate \glslink{contact reachable}{\textit{contact reachability}} with the \textit{reachability condition}, a geometrical criterion based on this intuitive description.
-%~ More precisely, a root configuration is \textit{reachable} if the root scaled by a factor $s \geq 1$ is not in collision, while the reachable workspace is in collision with the environment.
-%~ Verifying the \textit{reachability condition} is very efficient in practice.
-%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).
 
@@
 It takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually.
-%~ Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database computed offline independently from the environment).
-%~ Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.
-
-%~ The details of the contact sequence generation are given in Section~\ref{sec:contact}.
-
-%~ From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root configuration, an inverse-kinematics solver 
-%~ is used to maintain the contacts that were previously existing. Possibly, some of these contacts cannot be maintained (because of joint limits or collisions).
-%~ Then the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
-%~ The algorithm is designed so that only one contact can be created or broken between two successive configurations.
-While it does not provide guarantees that the interpolation between two configurations is achievable, it appears empirically as a reasonable heuristic.
-
-%~ To create a contact, we consider each limb of the robot as a manipulator arm attached to the root.
-We store a database of configurations for each manipulator.
-Figure~\ref{fig:contact_gen} -- 2 presents a few configurations for the right arm of the robot.
-The configurations in the database which are close to contact are considered (Figure~\ref{fig:contact_gen} -- 2 and 3).
-Among the candidates, 3 criteria are considered to choose the most relevant: the ideal candidate is free of collision, allows to maintain static balance if necessary, and  contributes to the motion in an optimal way.
-We call the measure of this contribution task efficiency: it represents the ability the limb has to exert a force compatible with the direction of motion.
-In Figure~\ref{fig:contact_gen} -- 4, the black arrow represents the direction of motion.
-Among the candidates, the most relevant to achieve a vertical motion is chosen, and projected on the contact surface using an inverse kinematics solver (Figure~\ref{fig:contact_gen} -- 5).
-The task efficiency is measured using the Extended FORce Transmission ratio (EFORT), proposed by ~\cite{Tonneau:2014:TEC:2619648.2619652}.
 
@@
 $r$ or $f(\textbf{x})$.
-
-%\subsubsection{Robot description}
-%\label{def}
-
-%~ \begin{figure}
-  %~ \centering
-  %~ \begin{overpic}[width=0.8\linewidth]{figures/character}
-    %~ \put (7,29) {\small{$\in \mathbf{q}^1$}}
-    %~ \put (7,25) {\small{effector}}
-  %~ \end{overpic}
-  %~ \caption{
-    %~ Left: Robot in a rest configuration.
-The right arm is denoted as the limb $R^1$.
-Each colored dot represents a degree of freedom around an axis.
-Right: Volumes of the robot.
-The red geometry denotes $W^0$ and must remain collision-free.
-The green spheres are the reachable workspace of each limb, the  $W^k$.}
-  %~ \label{fig:character}
-%~ \end{figure}
 
@@
 \begin{itemize}
-	\item $\mathbf{q}^k$ is a configuration  of the limb $R^k$; % (Figure~\ref{fig:character});
-	\item $\mathbf{q}^{\overline{k}}$ is a vector of joint values of R \textbf{not} related to $R^k$.
-We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
+	\item $\mathbf{q}^k$ is a configuration  of the limb $R^k$; %
+	\item $\mathbf{q}^{\overline{k}}$ is a vector of joint values of R \textbf{not} related to $R^k$. We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %
 	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ is the world coordinates vector of $R^0$.
 \end{itemize}
-%Finally, $ \mathbf{q}_{start}$ and  $\mathbf{q}_{goal}$ are the start and goal configurations given as an input of our problem.
-
-%\subsubsection{Environment and other volumes} 
-%\label{sec:A}
+
 \medskip
@@
 of $R^0$ for performance.}.\\
- %~ (Figure~\ref{fig:character}-right: central cylinder). 
- $W^k$ is the reachable workspace of a limb $R^k$: %(Figure~\ref{fig:character}-right: the four ellipses). 
+ $W^k$ is the reachable workspace of a limb $R^k$: %
 \begin{equation}
@@
 \end{equation}
-%~ \adnote{in this definition you don't account for joint limits}
 where $\mathbf{p}^k$ denotes the end-effector position (in the root frame) of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{j.lim}$ is the space
 of admissible limb joint configurations. We also define \mbox{$W = \bigcup_{k=1}^{l}W^k$}.
-
-%~ Each $W^i, 0 \leq i \leq l$ is  attached to one joint of the root, such that 
-%~ $W^i(\mathbf{q}^{0})$ is the volume $W^i$ transformed the by root configuration $\mathbf{q}^{0}$.
 
@@
 
-%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.
 \begin{figure}
@@
 \end{figure}
-% !TEX root =  ../main.tex
-%~ \section{Computing a guide path in $C^0_{reach}$ ($\mathcal{P}_1$) }
 \section{Root path planning in the contact reachable space}
@@
 \end{equation}
-% for \gls{cluttered} problems.
 We call  $C_{Reach}^0$ the \textit{contact reachable workspace}, and detail its construction in the following.
 The validity of this hypothesis is discussed in depth in Section~\ref{sec:discussion}.
-%~ We describe the construction of $C_{Reach}^0$ in the following.
 
@@
 \textbf{necessary conditions:}
-For a contact to be possible, an obstacle $O_i \subset O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot. % (Figure~\ref{fig:contact_gen}--a).
+For a contact to be possible, an obstacle $O_i \subset O$ necessarily intersects the reachable workspace $W(\mathbf{q}^{0})$ of the robot. %
  Also the torso of the robot $W^0(\mathbf{q}^{0})$ must necessarily be collision-free.
@@
 \begin{equation}
-\mathcal{C}^0_{\textrm{\it Nec}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ W^0(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
- %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
+\mathcal{C}^0_{\textrm{\it Nec}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ W^0(\mathbf{q}^{0}) \cap O = \emptyset \} %
 \end{equation}
-%~ The inclusion $C_{Contact}^0$ $\subset \mathcal{C}^0_{\textrm{\it Nec}}$ is straightforward, but it is very important.
-%~ Planning in $\mathcal{C}^0_{\textrm{\it Nec}}$ allows a strong reduction of the search space, while not discarding 
-%~ any root configuration that could lead to a solution.
-
-%~ The condition defining $\mathcal{C}^0_{\textrm{\it Nec}}$ is only necessary. This means that it might not be possible to extend a guide path planned 
-%~ in $\mathcal{C}^0_{\textrm{\it Nec}}$ into a sequence of contact configurations.
 
@@
 \begin{equation}
-\mathcal{C}^0_{\textrm{\it Suf}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ B^{\textrm{\it Suf}}(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
- %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
+\mathcal{C}^0_{\textrm{\it Suf}} = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ B^{\textrm{\it Suf}}(\mathbf{q}^{0}) \cap O = \emptyset \} %
 \end{equation}
-
-%~ In general, the inclusion is strict, which means that we lose the completeness of the two-stage contact planner (i.e.
-the planner is not able to discover a path inside \mbox{$C_{reach} \setminus \mathcal{C}^0_{\textrm{\it Suf}}$}).
-However, the sufficient condition guarantees that any such path leads to a valid sequence of contacts.
 
@@
 
-%~ However, using a shape between $W^0$ and $B^{\textrm{\it Suf}}$ leads to a trade-off between a necessary and a sufficient condition. 
 We define $W^0_s$ as the volume $W^0$ subject to a scaling transformation by a factor $s \in \mathbb{R}^+$.
-%
 We then consider the spaces $C_{s}^0$
@@
  \label{eq:reachc}
-C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} % \\ }
- %~ & \text{ and } & A_{torso}(\mathbf{q}^{root}) \cap W = \emptyset \}
+C^0_s = \{ \mathbf{q}^0 : W(\mathbf{q}^{0}) \cap O \neq \emptyset \textrm{ \textbf{and}}\ W^0_s(\mathbf{q}^{0}) \cap O = \emptyset \} %
 \end{equation}
-%
 The parametrization of $s$ defines a trade-off:
 If $s=1$, then $W^0_s$ = $W^0$, such that $C_1^0$ = $\mathcal{C}^0_{\textrm{\it Nec}}$.
- %~ We thus consider that $s \geq 1$, since smaller values would only worsen the approximation.
 By increasing $s$, the condition can become sufficient, but less and less necessary.  
@@
 $C_{Reach}^0$ can be sampled efficiently thanks to Eq.~\ref{eq:reachc}, and can thus be used with any standard motion planner.
-%~ The only significant change is to replace the collision checking with the \textit{reachability condition}.
 Our current implementation uses the Bi-RRT planner \citep{770022} provided by the HPP software~\citep{7759083}.
@@
 
-%~ \delst{However, to improve the sampling efficiency 
-%~ we bias the sampling process to generate near-obstacle configurations, similarly to cite{Amato98choosinggood}.}
-
 This Section has presented a guide path planner for the geometric root of a robot, implemented as a low-dimensional sampling-based 
 algorithm. Given start and goal configurations, it outputs a continuous path for the robot's root. 
-%~ \deladp{Thanks to the \textit{reachability condition}, we assume that for any configuration in the path there exists a joint configuration that results in static equilibrium.}
-%~ Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
-%~ By doing this we can solve the problem with a sampling-based approach, in \gls{interactive} computation times.
-% !TEX root =  ../main.tex
 \section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
@@
 the contact generator.
-%~ Our planner computes guide paths in $C_{Reach}^0$ , an approximation of \gls{$C_{Equil}^0$}.
-%~ As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Equil}^0$}.
-%~ If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section~\ref{sec:perf}.
-%~ We now consider the second problem of computing a discrete sequence of equilibrium configurations $\mathbf{Q}^{\overline{0}}$ along $\mathbf{q}^0(t)$.
-
-%~ In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a limb, given a
-%~ root location.
-%~ Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.
-
-%~ Our criterion to assert efficiently the static equilibrium of the system
-%~ is described in Appendix~\ref{sec:heuristics}.
 
@@
 In the previous example, our contact plan would simply consist of the two double-support configurations. 
-%However, if the contact broken and the contact created sequentially are not on the same effector, our planner outputs two distinct states.
 This representation is sufficient to describe all the contact phases, because the single support phase is implicitly described.
@@
 We first give an intuition of the algorithm, before providing its complete pseudo-code.
-%~ The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
 \subsubsection{Algorithm overview}
@@
 
-where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
+where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %
 $j$ depends on a user-defined variable, called the discretization step.
@@
 
-%~ We want to extend the configurations of $\mathbf{Q}^0$ in such a way that continuity is preserved regarding the contact transitions.
-%~ To do so, we define an algorithm that, given the current root configuration, and the previous full body configuration, computes a full body configuration 
-%~ in $C_{Equil}$ such that contacts are maintained if possible.
-%~ The first full body configuration of the sequence is given by the initial state of the robot.
-
- %~ we propose a recursive mapping $\pi$, for any $0<i<j$:
-%~ \begin{equation*}
-    %~ \pi\colon\left\{
-    %~ \begin{aligned}		
-        %~ \mathbf{Q}^0 \in C_{Equil}^0 & \longrightarrow C_{Equil} \\
-        %~ %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
-        %~ \mathbf{q}^{0}_i &  \longrightarrow  g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
-    %~ \end{aligned}
-    %~ \right.
-%~ \end{equation*} 
-%~ $g$ is the method that extends a root configuration into a full-body configuration. At each step, it tries to generate a contact configuration that preserves as much as possible
-%~ the previous contacts while allowing for static equilibrium. The objective is thus to characterize $g$.
-%~ We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
-
-%~ The function $g$ is defined independently by $g^k$ for each limb $R^k$. In defining $g^k$, two aspects must be considered. Is the limb $R^k$ in contact? And which criteria is it optimizing?
-
 \paragraph{Maintaining a contact in the sequence}
 
-%~ Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
 If kinematically possible, a limb in contact at step $i-1$ remains in contact at step $i$ (Figure~\ref{fig:break_contact}). 
-%~ The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits. 
-%~ The solver is directly provided by the HPP software.
 Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
@@
 that at most one contact is broken between two sequential configurations.
-%~ For these steps the root configuration is the same as for the previous step, with the difference that
-%~ one faulty contact is repositioned, in the hope that it will not be broken at the next step.
 
@@
 		\put (76,4) {4} 
-		%~ \put (68,58) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
 	\end{overpic}
@@
 
-%~ \begin{figure}[t]
-%~ \centering
-  %~ \begin{overpic}[width=0.6\linewidth]{figures/generate_contact}
-		%~ \put (5,58) {1)} 
-		%~ \put (37,58) {2)} 
-		%~ \put (68,58) {3.a)} 
-		%~ \put (5,27) {3.b)} 
-		%~ \put (37,27) {4.a)} 
-		%~ \put (68,27) {4.b)} 
-	%~ \end{overpic}
-%~ \caption{Contacts are generated when the configuration is not balanced.}
-		   %~ \label{fig:generate_contact}
-%~ \end{figure}
-
-%\subsection{Generation of a contact configuration}  
 \paragraph{Creating contacts}
@@
 \\ \\
-%~ \begin{enumerate}
-%~ \item \deladp{At most one contact creation happens between two consecutive steps. }
-%~ \item \deladp{A contact is validated if and only if the resulting configuration is in static equilibrium;}
-%~ \item \deladp{We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
-%~ was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
-%~ \end{enumerate}
-
-%~ \deladp{However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.}
-
-% !TEX root =  ../main_tro.tex
+
 \subsubsection{Pseudo-code of the Algorithm}
 \label{app:contact}
-
-%~ First, we define an abstract structure State,
-%~ that describes a contact configuration.
-%~ The use of queues allows a FIFO approach regarding the order 
-%~ in which contacts are tested: we try to replace older contacts first when necessary.
-%~ Thus the algorithm is deterministic even though it can handle acyclic motions. \\
-%~ 
-%~ \begin{lstlisting}]
-%~ Struct Limb
-%~ {
-    %~ // Limb Configuration
-    %~ Configuration qk;
-    %~ // Effector position in
-    %~ // world coordinates
-    %~ vector6 pk;
-%~ };
-%~ 
-%~ Struct State
-%~ {
-    %~ // root location
-    %~ Configuration q0;
-    %~ // List of limbs not in contact
-    %~ queue<Limb> freeLimbs;
-    %~ // List of limbs in contact
-    %~ queue<Limb> contactLimbs;
-%~ };
-%~ \end{lstlisting}
 
@@
 	\begin{algorithmic}[1]
-	%~ \Function{GenerateConfiguration}{}
 	\Function{Interpolate}{$s0$,$\mathbf{Q}^0$, $MAX\_TRIES$}
@@
 				\State $i \gets i+1$
-				%~ \State \textbf{return} $\mathbf{q}^{0}$
 			\Else
 				\State $nb\_fail \gets nb\_fail + 1$
-				%~ \If {$nb\_fail == MAX\_TRIES$}
-					%~ \State \textbf{return} $FAILURE$
-				%~ \EndIf				
 				\State $s \gets $\textsc{IntermediateContactState}$(pState)$
@@
 of a list~; \textit{Element} returns the element of a list at a given index~; and \textit{PushBack} inserts an element at the end of a list.
-%~ Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.
 
@@
 	\begin{algorithmic}[1]
-	%~ \Function{GenerateConfiguration}{}
 	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
@@
         \State $\quad \quad \quad \quad  \quad  \quad  \quad  \quad  \quad \quad  \quad \quad\!\! ContactLimbs(pState))$
-		%~ \State /*First try to maintain previous contacts*/
 		\State $nConBroken \gets 0$
@@
 This allows the algorithm to be deterministic even though it can handle acyclic motions.
-%~ $add\_contact\_limb(newState,k)$ marks a limb as in contact at a given state.
-%
-%~ The pseudo code for the method \textsc{IntermediateContactState} is given by Algorithm~\ref{alg:repo}.
-%
 \textit{GenerateContact}$(state,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
@@
 				\State \textbf{return} $newState$
-			%~ \Else
-				%~ \State \textit{MarkFree}$(newState,k)$	\comment{Mark }
-				%~ \State $push(states.freeLimbs,k)$		
 			\EndIf
@@
 		\State \textbf{return} $0$
-        %~ 
-        %~ 
-		%~ \State $i \gets 0$
-		%~ \While {$i<length(states.freeLimbs)$}
-			%~ \State Limb $k = pop(states.freeLimbs)$
-			%~ \If {\textit{GenerateContact}$(state,k)$}	
-				%~ \State $push(newState.contactLimbs,k)$			
-				%~ \State \textbf{return}
-			%~ \Else
-				%~ \State $i+=1$
-				%~ \State $push(states.freeLimbs,k)$		
-			%~ \EndIf
-		%~ \EndWhile
-		%~ \State $i=0$
-		%~ \While {$i<length(states.contactLimbs)$}
-			%~ \State Limb $k = pop(states.contactLimbs)$
-			%~ \State Limb $copy = k$
-			%~ \State $i+=1$
-			%~ \If {\textit{GenerateContact}$(state.q0,k)$}	
-				%~ \State $push(newState.contactLimbs,k)$			
-				%~ \State \textbf{return}
-			%~ \Else
-				%~ \State $push(newState.contactLimbs,copy)$	
-			%~ \EndIf
-		%~ \EndWhile
-		%~ /*Fails if impossible to relocate any effector*/
-		%~ \State \textbf{return} $FAILURE$
 	\EndFunction
@@
 \end{algorithm}
-
-%~ \section{Derivation of the manipulability ellipsoid}
-%~ \label{app:manipulability}
-%~ 
-%~ Again, we assume that $\mathbf{J}$ is full rank. We discard the $k$ indices, and write the pseudo-inverse of $\mathbf{J}$ as $\mathbf{J}^{\dagger}$.
-%~ \begin{eqnarray*}
-%~ \mathbf{\dot{p}} & =  & \mathbf{J} \dot{\mathbf{q}} \\ 
-%~ \mathbf{\dot{q}} & =  & \mathbf{J}^{\dagger} \dot{\mathbf{p}} \\ 
-%~ \mathbf{\dot{q}}^T & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \\ 
-%~ \mathbf{\dot{q}}^T\mathbf{\dot{q}} & =  & \dot{\mathbf{p}}^T \mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} \dot{\mathbf{p}}\\ 
-%~ \end{eqnarray*}
-%~ 
-%~ Then, the equality $\mathbf{J}^{\dagger T} \mathbf{J}^{\dagger} = (\mathbf{J}\mathbf{J}^T)^{-1}$ follows from the SVD decomposition of each term
-%~ ~\citep{ben2003generalized}.
 
@@
 Second, we rely on off-line generation of configuration candidates.
-%~ Contact generation is typically addressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
-%~ This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.
-%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
-
-%~ Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
-%~ $R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
-%~ While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
-%~ \begin{enumerate}
-%~ \item Generate randomly a collision-free limb configuration;
-%~ \item Project the end-effector onto the closest surface with inverse kinematics;
-%~ \item If a valid solution is found, stop. Otherwise repeat from step 1.
-%~ \end{enumerate}
-
-%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
 
@@
 
-%~ The reader is referred to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
-%~ of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
 Because the distance $\epsilon$ does not account for the variation in orientation, several samples of $ C_{Contact}^{\epsilon}$  may turn out to be unfeasible at the time of projection.
@@
 
-%~ TODO: $h$ in appendix, the static equilibrium is here ~\ref{sec:Equil}
-
-% !TEX root =  ../main_tro.tex
-
 \section{A criterion for robust static equilibrium}
 \label{sec:Equil}
-%~ The planner is designed so that any generated contact configuration is in static equilibrium.
-%~ We are interested in a robust 
-%~ criterion, that ensures that the robot remains in equilibrium in a real-world application, regardless of perception and control uncertainties.
 
@@
 In turn, from this value we can either choose the most robust candidate, or set a threshold on the required robustness.
-%~ While the presented LP is original, it is based on an analysis of the problem that we proposed in, where the interested reader can find more details.
 
@@
 \find \quad & \bm{\beta} \in \Rv{4e} \\
-%~ \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
 \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D} \mathbf{c} + \mathbf{d} \\
@@
 \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D} \mathbf{c} + \mathbf{d} \\
-%~ \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
 & \bm{\beta} \ge b_0 \bm{1}\\
@@
 If $b_0$ is negative, the configuration is not in static equilibrium, and $b_0$ indicates ``how far'' from equilibrium the configuration is. We thus use $b_0$ as a measure of robustness.
-%~ One advantage of $b_0$ is that it gives a constant margin 
 A simple approach to robustness consists in choosing a smaller friction coefficient, to constrain the forces to lie away from the boundaries of the real cone.
@@
 The optimal value $\bm{\nu}^*$ gives the optimal value $b_0^*$ through the equality $b_0^* = (\mathbf{D} \mathbf{c} + \mathbf{d})^T \bm{\nu}^*$.
- %~ Thus the optimal value of the LP~\eqref{eq:dual} is indeed the optimal $b_0$.
 
@@
 \url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html}. 
-% !TEX root =  ../main_tro.tex
 \section{Results}
@@
 Specifically, we demonstrate the planner for two legged robots, in a large variety of environments: the humanoid HRP-2 and the quadruped HyQ.
-%~ Finally, a last example suggests possible applications to dexterous manipulation.
 
@@
 result in feasible motions on the real robot. One motion is demonstrated on the real HRP-2 robot.
-%~ The obtained motions can be seen at at the end of the video.
 
@@
 \subsection{Experimental validation of the contact plans} \label{sec:valid}
-%~ \adnote{I added the description of the simulations here, but I wonder whether this part should not be moved to the beginning of the Results section.}
 To generate continuous movements from our contact plans we used either the framework proposed in \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
@@
 
-%~ These scenarios complement the ones demonstrated on virtual avatars (Figure~\ref{fig:robots_old}) in our previous ISRR paper~\citep{tonneauisrr15} and video\footnote{\url{http://youtu.be/LmLAHgGQJGA}}.
-%~ (Figure~\ref{fig:robots_old}).
- %~ We invite the interested reader to watch the ISRR video (\url{http://youtu.be/LmLAHgGQJGA}), and 
-%~ to refer to the previous paper for a discussion on these results.
-
-%~ We say that the planning is \gls{interactive} when the computation time for one step is lesser than the
-%~ time to execute it. We arbitrarily approximate this time to one second.
-
-%~ \begin{figure}[t]
-%~ \centering
-  %~ \begin{overpic}[width=1\linewidth]{figures/robots_old}
-	%~ \end{overpic}
-%~ \caption{Virtual avatars in various scenarios demonstrated in our conference paper.}
-		   %~ \label{fig:robots_old}
-%~ \end{figure}
-
 \subsection{Description of the scenarios}
@@
 In each scenario we detail the contacts involved and the heuristics chosen (either $h_{\textrm{\it EFORT}}$, $h_{vel}$ or $h_{w}$, all of which are defined in the Appendix~\ref{sec:heuristics}).
- %~ and the constraints on the reachable workspaces (for instance in all the scenarios, the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 
-%~ The companion video presents the complete contact sequence obtained in all these scenarios.
-
-% \subsubsection*{Truck egress (Figure~\ref{res_truck_pres} and Figure~\ref{res_truck_bd}) -- Humanoid and insectoid robots.}
-%\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}
 
@@
 
-%~ \noindent\textbf{Observations:} The scenario illustrates well the acyclic aspect of the planning. For instance, in the four first frames of Figure~\ref{fig:standing}, we can see that the right foot
-%~ is moved twice, with the left foot in between, before the configuration allows HRP-2 to move its hand.
-%~ Because the contacts are tried in a FIFO manner, the fact that the output contact sequence is acyclic shows that a cyclic approach (with a finite state machine for instance) is not sufficient
-%~ for the computed path. The reason for this is not reachability, but equilibrium. The planning is slower than for the stair scenario (because the contact generation fails more),
-%~ though it remains compatible with \gls{interactive} performances. % \adnote{Maybe define what you mean by 'interactive' and 'real-time' at some point.}
-
 \subsubsection{HRP-2 -- Car egress (Figure~\ref{fig:car})}
@@
 \noindent\textbf{Heuristics:} $h_w$.
-
-%~ \noindent\textbf{Observations:} The difficulty of this scenario lies in the strong reduction of the reachable workspace induced 
-%~ by the extreme proximity of all obstacles. The planner is able to find a sequence, that consists in many steps.
-%~ The proximity of the obstacles invalidate a large number of contact candidates because of collisions.
-To avoid breaking more than one contact between each step, the motion has to be decomposed into a large number of steps (61 in average).
-%~ While this scenario is the slowest to solve, the planner still computes a solution \glslink{interactive}{interactively}.
 
@@
 
-%~ \begin{figure}
-  %~ \centering
-  %~ \includegraphics[width=1\linewidth]{figures/stair}
-  %~ \caption{
-           %~ HRP-2 in the stair climbing scenario. }
-		   %~ \label{fig:stair_robust}
-%~ \end{figure}
-
 The goal is to climb three 15-cm high steps.
@@
 \noindent\textbf{Heuristics:} The manipulability $h_w$ is chosen for the feet; $h_{\textrm{\it EFORT}}$ is chosen for the right arm.
-%~ Regarding equilibrium, the video demonstrates two sequences computed for two different threshold values of $b_0$: $0$ and $2$ (Figure~\ref{fig:stair_robust}).
-
-%~ \noindent\textbf{Observations:}
-%~ This scenario illustrates best the importance of the equilibrium-robustness criterion.
-%~ With a robust approach, more states are required to reach the last step (15 rather than 13 in average).
-%~ However, when the last step is reached by both feet, in the nonrobust case the contacts are extremely close to 
-%~ the cone limits (Figure~\ref{fig:stair_comp}).
-%~ 
-%~ 
-%~ The geometry of the environment is easily addressed by our planner, and the contact planning is several times faster than real time in this scenario.
-%~ 
-%~ Again, the interpolation motion between the contact steps is out of the scope of this paper.
-However it should be noted that the computed plan in this scenario has been executed successfully on the robot (\url{http://youtu.be/YjL-DBQgXwk#t=0m28s}).
-%~ 
-%~ \begin{figure}
-  %~ \centering
-  %~ \begin{overpic}[width=0.5\linewidth]{figures/stair_robust}
-		%~ \put (17,5) {\small{\color{red}$b_0 = 0.23$}} 
-		%~ \put (79,5) {\small{\color{green}$b_0 = 6.16$}} 
-	%~ \end{overpic}
-  %~ \caption{
-           %~ Evaluation of the robustness $b_0$ of two contact configurations. Although in equilibrium, the left configuration is on the verge of slipping.}
-		   %~ \label{fig:stair_comp}
-%~ \end{figure}
 
@@
 
-%~ \noindent\textbf{Observations:} In this context, setting up a really important minimum value for $b_0$ is possible due to the high
-%~ stability of the HyQ robot, and results in more contact switches, in exchange for safety. The guide path-planning in this scenario takes a few seconds in average, more than
-%~ in any other scenarios.
-This is explained by the necessity of discovering a safe way to ``climb down'' the rubble.
-In this part of the planning, the constraint that the 4 reachable workspaces of all legs must collide with the environment at all times is hard to respect, but enforces the equilibrium of HyQ.
-%\adnote{Why do not relax this constraint and require only 3 legs?}
-%~ Again, the computation times remain however \gls{interactive}.
-
 \subsubsection{HyQ -- Obstacle race }
@@
 
-%~ \begin{figure}
-  %~ \centering
-  %~ \includegraphics[width=0.4\linewidth]{figures/HyQ_bridge}
-  %~ \caption{
-           %~ HyQ crossing a narrow bridge. }
-		   %~ \label{fig:HyQ_bridge}
-%~ \end{figure}
-
-%~ \begin{figure}
-  %~ \centering
-  %~ \includegraphics[width=1\linewidth]{figures/HyQ_obs}
-  %~ \caption{
-           %~ Crossing a hole contact sequence for HyQ. }
-		   %~ \label{fig:HyQ_obs}
-%~ \end{figure}
-
 \noindent\textbf{Contacts involved:} All (the 4 legs).
@@
 \noindent\textbf{Heuristics:} $h_w$. The robustness threshold $b_0$ is set to $10$.
-
-%~ \noindent\textbf{Observations:} Despite the apparent simplicity of the scene, this scenario is a hard case for a contact planner.
-%~ While finding a guide path above the hole is easy for the guide planner, finding a sequence of contacts that allows for equilibrium is not trivial.
-%~ Second, the narrow bridge is hard both for the planner and the contact generator: to make sure that equilibrium is preserved along the traversal,
-%~ the bridge must be approached with the appropriate angle.
-%~ The difficulty is illustrated in Figure~\ref{fig:HyQ_obs}, where several feet rearrangements are required to cross the hole (although the video shows this best).
-%~ The planner however succeeds in finding a feasible sequence in the end, again with \gls{interactive} computation times.
 
@@
 
-%~ \noindent\textbf{Observations:} This scenario is designed to illustrate concretely the computation times of the planner.
-%~ In the video, the footsteps indicating the contact sequence appear at the average speed of their computation (including the guide-path planning).
-
-%~ \subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot})}
-%~ This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
-%~ our contact planner to compute a contact sequence for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
-%~ Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences in less than 5 seconds.
-%~ 
-%~ \begin{figure*}
-%~ \centering
-  %~ \begin{overpic}[width=1\linewidth]{figures/penrot}
-	%~ \end{overpic}
-%~ \caption{Contact sequence found for a pen manipulation in a zero gravity environment.}
-		   %~ \label{fig:penrot}
-%~ \end{figure*}
-%~ 
- %~ 
-%~ \noindent\textbf{Contacts involved:} Three finger-tips.
-%~ 
-%~ \noindent\textbf{Heuristics:} $h_{\textrm{\it EFORT}}$ for all fingers.
-
 \subsection{Role of the main parameters} \label{sec:influence}
@@
 \subsubsection{Choosing the scaling factor $s$} \label{sec:params}
-%~ \deladp{To find a convenient value $s^*$ of $s$, we proceeded as follows. }
 For several values of $s$, we generated $10 000$ configurations. 
-%~ We then computed the sensitivity of the reachability condition (percentage of configurations in $C_{Reach}^0$ belonging to \gls{$C_{Contact}^0$}).
-%~ Similarly we computed the specificity of the condition (percentage of configurations \textbf{not} in $C_{Reach}^0$ \textbf{not} belonging to \gls{$C_{Contact}^0$}).
 We then computed the sensitivity and specificity of the reachability condition.
@@
 For HRP-2 we decided to set $s^*=1.2$.
-%~ , and was also effective for the car egress scenario.
 
@@
 use EFORT for the arms, which results in fewer contact repositionings.
-
-%~ \subsubsection{On the robustness equilibrium criterion:} \label{sec:parrob}
-%~ Robustness is really important when considering practical applications on the robot, to account 
-%~ for the various uncertainties that result from environment and state estimation. However,
-%~ maximizing the robustness criterion is often not optimal, because the resulting configurations may be too conservative, thus not favoring the motion.
-In our experiments, we choose not to maximize the robustness,
-%~ but to empirically se	t a robustness threshold value under which a configuration is not considered to be in
-%~ static equilibrium (If no candidate reaches the threshold, instead of failing, the algorithm can eventually return the ``more robust'' configuration
-%~ found). Currently for HRP-2, a threshold value of 2 subjectively gives the best results in the considered scenarios, while 10 seems to be a good choice for HyQ. Both values do not
-%~ significantly slow down the planning times.
 
@@
 \subsubsection{Success rates (Table~\ref{tab:sucess_planning})}
-%~ Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
 Despite the complexity of the scenarios and the approximations made in our formulation, our planner succeeded in the large majority of cases.
@@
 \subsubsection{Computation times (Table~\ref{tab:requestime})}
- %~ summarizes the computation times.
 
@@
 \subsection{Comparison with previous work} \label{sec:compa}
-%~ Few contributions provided computation times on the complete problem.
-To provide a completely fair comparison, it would be required to compare the methods on the same computers and exact same scenarios, which is not possible because
-%~ the source code of the other methods is often not available. %, necessary to provide comparison between our approaches. 
 We did our best to provide a fair comparison of the computation complexity of our method with the state of the art. 
-%Few contributions provided computation times on the complete problem. 
 However existing benchmarks for motion planning algorithms~\cite{moll2014extensible} do not yet encompass contact planning.
@@
 
-%~ Table~\ref{tab:compprev} presents the computation times for these scenarios, clearly demonstrating that our approach is order of magnitude faster than previous works.
 Table~\ref{tab:compprev} presents the computation times for these scenarios.
@@
 
- %~ Despite this observation and the fact that some results are old, we believe that the results show that our method
-%~ performs better. Indeed, a difference in technology does not justify a difference as important as the one demonstrated in Table~\ref{tab:sucess_planning}.
-
-% !TEX root =  ../main_tro.tex
  \section{Discussion: validity and purpose of our contact planner} 
 \label{sec:discussion}
-
-%~ \subsection{Accuracy and performance gain of the reachability condition}
 
@@
 This high performance is reached at the cost of some approximations that we discuss here.
-%~ In previous approaches, we identified 2 sources of high computational cost.
 
@@
 Indeed, in the large majority of cases, (84\% in the worst car egress case), we are able to find a contact configuration for any configuration in $C_{Reach}^0$.
-
-%~ These results clearly demonstrate that the reachability condition, presented in Section~\ref{sec:scaling}, provides an accurate representation of the 
-%~ contact we combine only-necessary and only-sufficient conditions which are trivial to estimate.
-Only-necessary conditions are appealing because they preserve the completeness of the search, while reducing
-%~ the search space: they provide an outer approximation of $C_{Contact}^0$.
-On the other hand, only-sufficient conditions provide the guarantee that any configuration that satisfies them is indeed \gls{contact reachable}: they provide an inner approximation of $C_{Contact}^0$.
 
@@
 These results clearly justify our pragmatic approach.
-
-%~ \subsection{Addressing $\mathcal{P}_3$}
-%~ \adnote{This subsection seems redundant to me}
-%~ This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation
-%~ . Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
-%~ for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
-%~ The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.
-%~ In these scenarios the robot is able to execute our plans, while respecting the constraints inherent to the problem: joint kinematics and torque limits and
-%~ collision avoidance.
 
@@
 
-%~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
 Our contribution to \Pb is a fast contact generation scheme that can optimize user-defined criteria.
@@
 
-%
 \textbf{Regarding generality}, the \textit{reachability condition}, coupled with an approach based on limb decomposition, 
 allows the method to address automatically arbitrary legged robots.
-%
 \textbf{Regarding performance}, our framework is efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
-%
 \textbf{Regarding the quality of the paths}, we are able to compute
@@
 Considering the set of obstacles intersecting with the reachable workspace for a given root configuration as candidate surfaces, we can use them to verify the equilibrium criterion.
-This would give us a necessary condition for \equilibriumfeasibility. %, which we investigate in~\cite{fernbach:hal-01486933}.
+This would give us a necessary condition for \equilibriumfeasibility. %
 
@@
 the transition between two contact configurations is feasible and collision-free.
-%~ Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
-%~ at the center of mass of the robot.
- %~ We are also
-%~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
 Finally, we aim at performing kinodynamic planning to remove the constraint that configurations be in static equilibrium.
@@
 
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation results for the network.}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that the IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command,
-% and the \label for the overall figure must come after \caption.
-% \hfil is used as a separator to get equal spacing.
-% Watch out that the combined width of all the subfigures on a 
-% line do not exceed the text width or a line break will occur.
-%
-%\begin{figure*}[!t]
-%\centering
-%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
-%\label{fig_second_case}}
-%\caption{Simulation results for the network.}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat[]), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
-% labels, the optional argument to \subfloat must be present. If a
-% subcaption is not desired, just leave its contents blank,
-% e.g., \subfloat[].
-
-% An example of a floating table. Note that, for IEEE style tables, the
-% \caption command should come BEFORE the table and, given that table
-% captions serve much like titles, are usually capitalized except for words
-% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
-% and up, which are usually not capitalized unless they are the first or
-% last word of the caption. Table text will default to \footnotesize as
-% the IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that the IEEE does not put floats in the very first column
-% - or typically anywhere on the first page for that matter. Also,
-% in-text middle ("here") positioning is typically not used, but it
-% is allowed and encouraged for Computer Society conferences (but
-% not Computer Society journals). Most IEEE journals/conferences use
-% top floats exclusively. 
-% Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the
-% \fnbelowfloat command of the stfloats package.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
-
 \appendices
 
-%~ \clearpage
 \section{Generating the $W$ volumes for HRP-2}
@@
 For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
-%~ (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}).
-For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
 \end{enumerate}
@@
 $s^*=1.2$ as the appropriate value for HRP-2.
-%~ \section{Heuristics and criteria for contact selection}
-%~ \label{sec:heuristics}
-% !TEX root =  ../main.tex
-%~ \subsection{Manipulability-based heuristics for contact selection}
 \section{Manipulability-based heuristics for contact selection}
@@
 This Section proposes two heuristics to select a contact that optimizes desired capabilities.
-%~ These heuristics allow to locally optimize a contact location with respect to a particular task.
 For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion, or to stay away from singular configurations.
-We derive these heuristics from the work by \cite{Yoshikawa1984}, recalled here. %, that we present first.
+We derive these heuristics from the work by \cite{Yoshikawa1984}, recalled here. %
 
@@
 The longer the axis of the ellipsoid is, the more important the velocity (resp. force) of the end-effector the direction of the axis can be.
-%~ (Figure~\ref{sec:efort_ellipsoid}).
 
@@
 \end{equation}
-%~ $h_{w}$ measures the ``distance'' of a given configuration to singularity. When $h_{w}$ is equal to 0, the configuration is singular;
-%~ the greater $h_{w}$ is, the further away the configuration is from singularity.
-
-%~ \input{body/equilibrium}
-% !TEX root =  ../main.tex
+
 \section{Addressing $\mathcal{P}_3$}
@@
 The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
-%~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
 To address the other scenarios, we propose a new implementation, entirely open source (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#optim} ), which can be integrated directly in our motion planner software.
@@
 
-%~ \begin{align} \label{eq:new_eul}
-%~ \underbrace{
-%~ \mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
-%~ \hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
-%~ }_\mathbf{G} \bm{\beta}, = 
-%~ \underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
-%~ \underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
-%~ \end{align}
-
 \begin{align} \label{eq:new_eul_acc}
@@
 \underbrace{\mat{m (\mathbf{\ddot{c}} - \mathbf{g}) \\m \mathbf{c} \times (\mathbf{\ddot{c}} - \mathbf{g}) + \dot{L}}}_{\mathbf{w}}
-%~ \underbrace{m (\mathbf{\ddot{c}} - \mathbf{g})  \\ m (\mathbf{\ddot{dc}} - \mathbf{g}) + \mathbf{\dot{L}} }_{\mathbf{w}}
 \end{align}
@@
 \begin{equation} \label{eq:qp_rob} \begin{aligned}
-%~ \find \quad & \mx{R} \in \Rv{3\times T}, \mxdd{C} \in  \Rv{6\times T} \\
-%~ \minimize_{\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)}  \quad & \sum_{u=0}^{2} \int_{t_u}^{t_u + \Delta t_u}  \ell(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{\ddot{c}}(t), \mathbf{w}(t)) \text{dt}  \\
 \minimize_{\mathbf{\ddot{c}}(t), \dot{L}(t)}  \quad & \sum_{u=0}^{2} \int_{t_u}^{t_u + \Delta t_u}  \ell(\mathbf{\ddot{c}}(t), \dot{L}(t)) \text{dt}  \\
-%~ \st \quad & \mathbf{\ddot{c}}(t) = dyn(\mathbf{c}(t), \mathbf{\dot{c}}(t), \mathbf{w}(t)) \\
-%~ \st \quad & \mathbf{c}(t) =  dyn(\mathbf{\ddot{c}}, \mathbf{c}_{\mathbf{q}_0} )  \\
 \st \quad & \mathbf{A}_u \mathbf{w}(t)  \leq \mathbf{b}_u , \forall t \in  \left[t_u, t_u + \Delta t_u \right[, \forall u \\
@@
 	\quad & \mathbf{c}(T)  = \mathbf{\dot{c}}(T)  = \mathbf{\ddot{c}}(T) = \mathbf{0}\\
-    %~ \quad & ({\vct{r}}\tra\vct{r}=1,  \forall t  ?) \\
-    %~ \quad & \delta t \sum\limits_1^{T-1}{\vc{r}^1}\tra {\vc{r}^1} \ge t_{transit} \\
-    %~ \quad & \mx{A}_{\int} \mxdd{C} \leq \vc{a}_{\int} \\
-    %~ \quad & \mx{V}_{\int} \mxd{C} \leq \vc{v}_{\int} \\
-    %~ \quad & \mx{B}_{\int}^{1} \mx{C} \leq \vc{b}_{\int} \\
-    %~ \quad & [\vc{c}_0\tra, \vcd{c}_0\tra]\tra = \vc{x}_0 \\
-    %~ \quad & \vc{c}_T = \vc{c}_T^{input} \\
 \end{aligned} \end{equation}
-%~ \adnote{Did you really use w as a variable of your problem?
-Or do you compute it inside your constraint function as a function of the CoM pos and acc?
-And what about dL?
-It appears in the definition of w but then it is no longer there in the problem definition!}
 The cost function $\ell$ is a weighted sum of the angular momentum and center-of-mass acceleration variation over the whole trajectory.
@@
  $\mathbf{q}_0$ and $\mathbf{q}_1$ respectively.
-%~ The system dynamics $dyn$ tie together the position, velocity and acceleration of the center of mass. 
 $\mathbf{Y}_u$ and $\mathbf{y}_u$ denote stacked kinematic constraints on the center of mass position, determined by the active contact locations.
 The inequalities for each contact are determined in the same way that the reachable workspace is computed in Appendix~\ref{app:rom}, with
-the effector serving as root. % (a triangulated object is a polytope, and as such can be described as a set of linear inequalities).
+the effector serving as root. %
 
@@
 
-%This motion is then validated in our dynamic simulator~\citep{DelPrete2015b}, using the trajectory as a reference postural task for an inverse-dynamics
-%controller that accounts for all the robot constraints. This final result is shown in the companion video.
-%~ \input{body/appendix_algo}
-%~ \input{body/appendix_hpp}
-
 \section*{Acknowledgements}
 This research is supported by Euroc (project under FP7 Grant Agreement  608849);  Entracte (ANR  grant  agreement  13-CORD-002-01);
-the ARO Contract W911NF-14-1-0437; and the NSF award 1305286. %, and a grant from Boeing.
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
+the ARO Contract W911NF-14-1-0437; and the NSF award 1305286. %
+
 \bibliographystyle{IEEEtran}
- %~ argument is your BibTeX string definitions and bibliography database(s)
 \bibliography{biblio}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-%~ \begin{thebibliography}{1}
-%~ 
-%~ \bibitem{IEEEhowto:kopka}
-%~ H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-  %~ 0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-%~ 
-%~ \end{thebibliography}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
 
@@
 
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-%~ \begin{IEEEbiographynophoto}{Jane Doe}
-%~ Biography text here.
-%~ \end{IEEEbiographynophoto}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
 \end{document}
